### 2020-03-29 pde/Notes/CourseNotes.tex
Several new files.

\documentclass{article}
\usepackage{amsmath, amsthm, amssymb, amsfonts, graphicx, color}
\usepackage{mathtools}
\usepackage{hyperref}
%\renewcommand{\rmdefault}{ppl}
%\usepackage[small, euler-digits]{eulervm}

\newcommand{\td}[2]{\frac{d{#1}}{d{#2}}}
\newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cvar}[1]{\mathrm{char}_\alpha(#1)}
\newcommand{\cce}{\subset\subset}
\newcommand{\fint}{\mbox{--}\mkern-16mu\int}
\DeclarePairedDelimiter\norm{\lVert}{\rVert}
\DeclareMathOperator*{\esssup}{ess\,sup}
\DeclareMathOperator*{\Int}{Int}
\DeclareMathOperator{\dive}{div}

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\numberwithin{thm}{section}

\theoremstyle{plain}
\newtheorem{prop}{Proposition}
\numberwithin{prop}{section}

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\numberwithin{defn}{section}

\theoremstyle{remark}
\newtheorem*{rem}{Remark}

\newtheorem*{cor}{Corollary}

\numberwithin{equation}{section}
\begin{document}
\title{Partial Differential Equations\\A Course by Prof. A. K. Nandakumaran}
\author{Amey Joshi}
\date{07-Aug-2018}
\maketitle

\section{Notation}\label{s1}
A point $x = (x_1, \ldots, x_n)$ will usually be a point in $\mathbb{R}^n$. By default, all vectors 
are column vectors. The set of natural numbers will include $0$. If $\alpha = (\alpha_1, \ldots, 
\alpha_n)$, where $\alpha_i$, for $1 \le i \le n$ are integers then the symbol $|\alpha| = \alpha_1 + 
\cdots + \alpha_n$. $U$ is usually a domain (that is, an open set) in $\mathbb{R}^n$. The symbol 
$D^\alpha$ is defined as
\begin{equation}\label{s1e1}
D^\alpha = \frac{\partial^{|\alpha|}}{\partial x_1^{\alpha_1}\cdots\partial x_n^{\alpha^n}}.
\end{equation}
We define
\begin{equation}\label{s1e2}
D^0f = f.
\end{equation}
If $e_i \in \mathbb{R}^n$ is the standard basis of the vector space $\mathbb{R}^n$ then
\[
D^{e_i}f = \pd{f}{x_i},
\]
where $f: \mathbb{R}^n \rightarrow \mathbb{R}$ is a function differentiable in a certain $U$. 
Using this notation, the gradient in $n$ dimensions is
\begin{equation}\label{s1e3}
Df = (D^{e_1}f, \ldots, D^{e_n}f).
\end{equation}

We will also use the $D$ notation for total derivatives. Thus the total derivative of a function
$u$ with respect to a variable $\sigma$ is denoted by
\begin{equation}\label{s1e4}
D_\sigma u = \td{u}{\sigma}.
\end{equation}

\begin{defn}\label{s1d1}
A partial differential equation (PDE) of order $k$, where $k$ is a positive integer, is an equation
of the form
\begin{equation}\label{s1e5}
F(x, D^0u, \ldots, D^ku) = 0.
\end{equation}
where $F:U \times \mathbb{R} \times \cdots \times \mathbb{R}^{n^k} \rightarrow \mathbb{R}$ and
$u: \mathbb{R}^n \rightarrow \mathbb{R}$.
\end{defn}
\begin{rem}
Recall that there are $n^k$ partial derivatives of order $k$ of a function of $n$ variables.
\end{rem}

\begin{defn}\label{s1d2}
A PDE is linear if it is of the form
\[
\sum_{|\alpha| \le k}a_\alpha(x) D^\alpha u = f(x),
\]
where the functions $f, a_\alpha: \mathbb{R}^n \rightarrow \mathbb{R}$ are arbitrary functions.
\end{defn}

\begin{defn}\label{s1d3}
A PDE is semi-linear if it is of the form
\[
\sum_{|\alpha|=k}a_\alpha(x)D^ku + b(x, D^0u, \ldots, D^{k-1}u) = f(x),
\]
where the functions $f, a_\alpha: \mathbb{R}^n \rightarrow \mathbb{R}$ are arbitrary functions and 
$b$ are arbitrary real valued functions over their domain.
\end{defn}

\begin{defn}\label{s1d4}
A PDE is quasilinear if it is of the form
\[
\sum_{|\alpha|=k}a_\alpha(x, D^0u, \ldots, D^{k-1}u)D^ku + b(x, D^0u, \ldots, D^{k-1}u) = f(x),
\]
where the functions $f, a_\alpha: \mathbb{R}^n \rightarrow \mathbb{R}$ are arbitrary functions and 
$b$ are arbitrary real valued functions over their domain.
\end{defn}

\begin{defn}\label{s1d5}
A PDE is nonlinear if it is not linear, semi-linear or quasilinear.
\end{defn}

\begin{defn}\label{s1d6}
A function $v:\mathbb{R}^n \rightarrow \mathbb{R}$ is a classical solution of the partial 
differential equation \eqref{s1e5} in a domain $U \subset \mathbb{R}^n$ if $v$ is defined over 
$U$, if it has all the partial derivatives appearing in $F$ defined over $U$ and if 
\[
F(x, D^0v, \ldots, D^kv) = 0,
\]
is an identity for all points $x \in U$. A solution $v$ is also called the \emph{integral 
surface} of the PDE \eqref{s1e5}.
\end{defn}

Sometimes we will use the following alternative notation for partial derivatives.
\begin{equation}\label{s1e6}
\pd{u}{x} = u_{x}
\end{equation}
so that the gradient of a function of $n$ variables is
\[
Df = (f_{ x_1}, \ldots, f_{ x_2}).
\]
In order to avoid multiple levels of subscripts, we will sometimes have a comma before the subscript
$x$ in \eqref{s1e6} so that the divergence of a vector field $B = (B_1, \ldots, B_n)$ is
\[
\text{div}\;B = B_{1, x_1} + \ldots + B_{n,x_n} = B_{i, x_i},
\]
where we extreme right hand side uses the Einstein summation convention.

\section{First order PDE}\label{s2}
Let $e_1 = (1, 0)$ and $e_2 = (0, 1)$ denote the basis of the vector space $\mathbb{R}^2$. Using 
definition \ref{s1d1}, a PDE of first order has the general form
\begin{equation}\label{s2e1}
F\left(x_1, x_2, u, \pd{u}{x_1}, \pd{u}{x_2}\right) = F(x_1, x_2, u, D^{e_1}u, D^{e_2}u) = 0.
\end{equation}
A linear first order PDE has the form
\begin{equation}\label{s2e2}
a_1(x_1, x_2)D^{e_1}u + a_2(x_1, x_2)D^{e_2}u = b(x_1, x_2)u + d(x_1, x_2),
\end{equation}
where $a_1, a_2, b, c$ are arbitrary functions $\mathbb{R}^2 \mapsto \mathbb{R}$. A semi-linear first
order PDE has the form
\begin{equation}\label{s2e3}
a_1(x_1, x_2)D^{e_1}u + a_2(x_1, x_2)D^{e_2}u = b(x_1, x_2, u),
\end{equation}
where $a_1, a_2$ are arbitrary functions $\mathbb{R}^2 \mapsto \mathbb{R}$ while $b:\mathbb{R}^3
\rightarrow \mathbb{R}$. A quasilinear first order PDE has the form
\begin{equation}\label{s2e4}
a_1(x_1, x_2, u)D^{e_1}u + a_2(x_1, x_2, u)D^{e_2}u = b(x_1, x_2, u),
\end{equation}
where $a_1, a_2, b$ are arbitrary functions $\mathbb{R}^3 \mapsto \mathbb{R}$.

\subsection{Cauchy Problem}\label{s2s1}
Let $I$ be an interval in $\mathbb{R}$. For let $X_1, X_2, U$ be arbirtary function $I \mapsto 
\mathbb{R}$ such that
\begin{enumerate}
\item $X_1, X_2$ are piecewise continuous in $I$,
\item $(X_1^\prime)^2 + (X_2^\prime)^2 \ne 0$ in $I$.
\end{enumerate}
The functions $X_1, X_2$ define the parametric equations of a curve in the plane $\mathbb{R}^2$. The
pair $(X_1(\eta), X_2(\eta))$ is called the \emph{initial curve} or the \emph{datum curve} $\Gamma$. 
$U$ is called the \emph{initial value} of the solution of the PDE on the curve $\Gamma$. The Cauchy 
problem consists in finding a solution of \eqref{s2e1} in the neighborhood of $\Gamma$ such that its 
value on $\Gamma$ is identical to $U(\eta)$. The reason it is called an initial value problem will
be clear in the following section.

Formally, the \emph{Cauchy problem} for the PDE \eqref{s2e1} consists in finding a domain $U 
\subset \mathbb{R}^2$ and a function $u:\mathbb{R}^2 \rightarrow \mathbb{R}$ such that
\begin{enumerate}
\item $(X_1(\eta), X_2(\eta)) \in U$, 
\item $u(X_1(\eta), X_2(\eta)) = U(\eta)$ for all $\eta \in I$ and
\item $u$ is a solution of \eqref{s2e1}.
\end{enumerate}

\subsection{Cauchy Problem for the semi-linear PDE}\label{s2s2}
We will solve the Cauchy problem for the semi-linear \eqref{s2e3}. The linear case of \eqref{s2e2}
is just the special case of \eqref{s2e3}. We will assume that the functions $a_1, a_2$ are 
continuously differentiable in a domain of $\mathbb{R}^2$ and $b$ is continuously differentiable
in a domain of $\mathbb{R}^3$. We can write the equation \eqref{s2e3} as
\[
(a_1, a_2)\cdot Du = b(x_1, x_2, u),
\]
where the left hand side is written as the scalar product of the vector $(a_1, a_2)$ and the
gradient $Du$. We can then interpret the left hand side of the equation \eqref{s2e3} as the projection
of the gradient $Du$ in the direction $(a_1, a_2)$. Let us consider a family of one-parameter curves
whose tangents are along the direction $(a_1, a_2)$. Then they satisfy the system of autonomous 
ordinary differential equations (ODE),
\begin{eqnarray}
D_\sigma x_1 &=& a_1(x_1, x_2) \label{s2e5} \\
D_\sigma x_2 &=& a_2(x_1, x_2), \label{s2e6}
\end{eqnarray}
where $\sigma$ is a parameter of the curves. We have deliberately chosen to denote it differently
from the parameter of the initial curve $\Gamma$. Since $u$ is a function of $x_1, x_2$, we can
as well assume that it is also a function of the same parameter $\sigma$ so that
\begin{equation}\label{s2e7}
D_\sigma u = D^{e_1}u D_\sigma x_1 + D^{e_2}u D_\sigma x_2.
\end{equation}
Using \eqref{s2e5} and \eqref{s2e6},
\begin{equation}\label{s2e8}
D_\sigma u = a_1(x_1, x_2)D^{e_1}u +  a_2(x_1, x_2)D^{e_2}u = b(x_1, x_2, u),
\end{equation}
where the last equality follows from the PDE \eqref{s2e3}. Equation \eqref{s2e8} is called the
\emph{compatibility condition}.

Equations \eqref{s2e5} and \eqref{s2e6} are called the \emph{characteristic equations} and their
solution is called the \emph{characteristic curve}. The theory of autonomous ODE assures us that
any point of the plane $\mathbb{R}^2$ lies on a unique characteristic curve. The solution of equations 
\eqref{s2e5} and \eqref{s2e6} can be written as
\begin{equation}\label{s2e9}
f(x_1, x_2, c) = 0,
\end{equation}
where $c$ is a constant of integration. Equation \eqref{s2e9} thus defines a one-parameter family
of characteristic curves where $c$ is the parameter. (We get a one-parameter family of curves because
equations \eqref{s2e5} amd \eqref{s2e6} can be combined to $D_{x_1}x_2 = a_2/a_1$. Upon integration,
we get $x_2 = g(x_1) + c$. The single constant of integration, $c$, makes it a one-parameter family
of curves.)

Now consider a point $\eta_0$ in the interval $I$ of the Cauchy problem. It determines the point
$(X_1(\eta_0), X_2(\eta_0))$ in the plane $\mathbb{R}^2$. There is a unique characteristic curve
passing through $(X_1(\eta_0), X_2(\eta_0)$. The Cauchy problem for \eqref{s2e3} consists in solving
the ODE \eqref{s2e5} and \eqref{s2e6} subject to the initial conditions $x_1(\sigma = 0) = 
X_1(\eta_0)$ and $x_2(\sigma = 0) = X_2(\eta_0)$. Using these initial values, we get a unique
characteristic curve 
\begin{equation}\label{s2e10}
f(x_1, x_2, \sigma) = 0.
\end{equation} 
Let 
\begin{equation}\label{s2e11}
U_0 = U(x_1 = X_1(\eta_0), x_2 = X_2(\eta_0)).
\end{equation}
Then $u_0$ is the initial value of the ODE \eqref{s2e8}. The solutions of equations \eqref{s2e5}, 
\eqref{s2e6} and \eqref{s2e8} can be written as
\begin{eqnarray}
x_1(\sigma) &=& x_1(X_1(\eta_0), X_2(\eta_0), \sigma) \label{s2e12} \\
x_2(\sigma) &=& x_2(X_1(\eta_0), X_2(\eta_0), \sigma) \label{s2e13} \\
u(\sigma)   &=& u(X_1(\eta_0), X_2(\eta_0), U_0, \sigma) \label{s2e14}
\end{eqnarray}
Equation \eqref{s2e14} gives the solution of \eqref{s2e3} along the characteristic curve passing
through the point $(X_1(\eta_0), X_2(\eta_0))$ on the initial curve $\Gamma$. However, we want a 
solution $u$ in terms of $x_1, x_2$ and not just $\sigma$. Let us write equations \eqref{s2e12} and
\eqref{s2e13} as 
\begin{eqnarray}
x_1 &=& x_1(\sigma, \eta_0) \label{s2e15} \\
x_2 &=& x_2(\sigma, \eta_0). \label{s2e16}
\end{eqnarray}
If we can invert them to get
\begin{eqnarray}
\sigma &=& \sigma(x_1, x_2) \label{s2e17} \\
\eta_0 &=& \eta_0(x_1, x_2) \label{s2e18}
\end{eqnarray}
then we can substitute \eqref{s2e17} and \eqref{s2e18} in \eqref{s2e14} to get $u$ in the desired
form. The inversion of \eqref{s2e17} and \eqref{s2e18} is possible if 
\begin{equation}\label{s2e19}
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta)} \ne 0 \text{ at } \eta = \eta_0.
\end{equation}
This condition is equivalent to
\[
\frac{D_\sigma x_2}{D_\sigma x_1} \ne \frac{D_\eta x_2}{D_\eta x_1} \text{ at } \eta = \eta_0.
\]
It means that the characteristic curve and the normal curve are not parallel to each other. Therefore,
equation \eqref{s2e18} is also called the \emph{transversality condition}.

\section{Examples of Semi-linear PDE}\label{s3}
\begin{enumerate}
\item Consider the problem (example 2.1 of \cite{pprr}) $2D^{e_1}u + 3D^{e_2}u = 1$ with Cauchy data
on $\alpha x - \beta y = 0$, where $\alpha, \beta \in \mathbb{R}$. The parametric equations 
\eqref{s2e5}, \eqref{s2e6} and \eqref{s2e8} for this problem are
\begin{eqnarray}
D_\sigma x_1 &=& 2 \label{s3e1} \\
D_\sigma x_2 &=& 3 \label{s3e2} \\
D_\sigma u &=& 1 \label{s3e3} 
\end{eqnarray}
The initial conditions are provided by the Cauchy data. If $\eta$ is the parameter describing the
initial curve $\Gamma$ then $x = \eta, y = \alpha\eta/\beta$ are its parametric equations. Let the
function $u$ take a value $f(\eta)$ on the curve $\Gamma$. Then the initial values of the system of 
ODE \eqref{s3e1}, \eqref{s3e2} and \eqref{s3e3} are $x_1(\sigma = 0) = \eta_0$, $x_2(\sigma = 0) = 
\alpha\eta_0/\beta, u(\sigma = 0) = f(\eta_0)$ so that its solution is
\begin{eqnarray}
x_1(\sigma) &=& 2\sigma + \eta_0 \label{s3e4} \\
x_2(\sigma) &=& 3\sigma + \frac{\alpha}{\beta}\eta_0 \label{s3e5} \\
u(\sigma) &=& \sigma + f(\eta_0) \label{s3e6}
\end{eqnarray}
The characteristic curves defined by \eqref{s3e4} and \eqref{s3e5} are straight lines of slope $3/2$ 
and varying $y$-intercept.

We want a solution of the problem in terms of $x_1, x_2$ and not $\sigma, \eta_0$. Therefore, we 
invert \eqref{s3e4} and \eqref{s3e5} to get
\begin{eqnarray}
\sigma &=& \frac{\beta x_2 - \alpha x_1}{3\beta - 2\alpha} \label{s3e7} \\
\eta_0 &=& \beta\frac{3x_1 - 2x_2}{3\beta - 2\alpha} \label{s3e8}
\end{eqnarray}
so that the solution of the problem is
\begin{equation}\label{s3e9}
u(x_1, x_2) = \frac{\beta x_2 - \alpha x_1}{3\beta - 2\alpha} +
 f\left(\beta\frac{3x_1 - 2x_2}{3\beta - 2\alpha}\right)
\end{equation}
Note that the inversion of \eqref{s3e7} and \eqref{s3e8} is possible only when $3\beta \ne 2\alpha$,
that is $3/2 \ne \alpha/\beta$, that is when the characteristic curve and the initial curve do not
coincide.

In the event $3\beta = 2\alpha$ the two curves coincide and are described by the same parameter 
$\sigma$. Further $\eta_0 = 0$ so that the equations \eqref{s3e4} and \eqref{s3e6} give $\sigma = 
x_1/2$ or $x_2/3$. Further, $\eta_0 = 0$ is equivalent to $\beta x - \alpha y = 0$ or $(3x - 2y) = 0$.
The solution of the problem then becomes
\begin{equation}\label{s3e10}
u(x_1, x_2) = \frac{x_1}{2} + g(3x - 2y) + f(0)
\end{equation}
or
\begin{equation}\label{s3e11}
u(x_1, x_2) = \frac{x_2}{3} + g(3x - 2y) + f(0),
\end{equation}
where $g$ is an arbitrary $C^1$ function such that $g(0) = 0$.

\item (Problem Set 1, Ex. 1) $(x_1 + 2)D^{e_1}u + 2x_2D^{e_2}u = 2u$ with initial condition 
$u(-1, x_2) = \sqrt{x_2}$. The initial curve is the line $x_1 = -1$. Its parametric form is $x_1 = -1,
x_2 = \eta$. The parametric equations of the characteristic curve are
\begin{eqnarray}
D_\sigma x_1 &=& x_1 + 2 \label{s3e12} \\
D_\sigma x_2 &=& 2x_2 \label{s3e13} \\
D_\sigma u &=& 2u \label{s3e14}
\end{eqnarray}
with the initial conditions $x_1(\sigma = 0) = -1, x_2(\sigma = 0) = \eta, u(\sigma = 0) = 
\sqrt{\eta}$. Their solutions are
\begin{eqnarray}
x_1 &=& e^\sigma - 2 \label{s3e15} \\ 
x_2 &=& \eta_0 e^{2\sigma} \label{s3e16} \\
u &=& \sqrt{\eta_0}e^{2\sigma} \label{s3e17}
\end{eqnarray}
Inverting \eqref{s3e15} and \eqref{s3e16} we get 
\begin{eqnarray}
e^\sigma &=& x_1 + 2 \label{s3e18} \\
\eta_0 &=& x_2/(x_1 + 2)^2 \label{s3e19}
\end{eqnarray} 
so that
\begin{equation}\label{s3e20}
u(x) = \sqrt{x_2}(x_1 + 2)
\end{equation}
is the solution of the problem. The equation for the characteristic curve is obtained by eliminating
the parameter $\sigma$ from equations \eqref{s3e15} and \eqref{s3e16}. It is $x_2 = \eta_0(x_1 + 2)^2$.
Figure \ref{f1} shows the characteristic curves of the problem.
\begin{figure}
\centering
\includegraphics{ss1_p1a}
\caption{Characteristic curves $x_2 = \eta_0(x_1 + 2)^2$.}
\label{f1}
\end{figure}
Figure \ref{f2} shows the integral surface.
\begin{figure}
\centering
\includegraphics{ss1_p1b}
\caption{The integral surface $u(x) = \sqrt{x_2}(x_1 + 2)$.}
\label{f2}
\end{figure}

\item (Problem Set 1, Ex. 2) $x_1D^{e_1}u + x_2D^{e_2}u = 2u, x_1 > 0, x_2 > 0$ with initial 
conditions (a) $u = 1$ on the hyperbola $x_1x_2 = 1$ and (b) $u = 1$ on the circle $x_1^2 + x_2^2 = 
1$.

The parametric representation of the rectangular hyperbola $x_1x_2 = 1$ is $x_1 = \eta, x_2 = 
\eta^{-1}$. The parametric equation of the characteristic curves is
\begin{eqnarray}
D_\sigma x_1 &=& x_1 \label{s3e21} \\
D_\sigma x_2 &=& x_2 \label{s3e22} \\
D_\sigma u &=& 2u \label{s3e23},
\end{eqnarray}
subject to the initial conditions $x_1(\sigma = 0) = \eta_0, x_2(\sigma = 0) = \eta_0^{-1}$ and
$u(\sigma = 0) = 1$. Their solutions are
\begin{eqnarray}
x_1 &=& \eta_0 e^{\sigma} \label{s3e24} \\
x_2 &=& \eta_0^{-1} e^{\sigma} \label{s3e25} \\
u &=& e^{2\sigma} \label{s3e26}
\end{eqnarray}
The equation of characteristic curve is $x_2 = \eta_0^{-2}x_1$. It is a straight line passing
through the origin with slope $\eta_0^{-2}$. The first two of the above equations can be inverted 
to get
\begin{eqnarray}
e^\sigma &=& \sqrt{x_1x_2} \label{s3e27} \\
\eta_0 &=& \sqrt{\frac{x_1}{x_2}} \label{s3e28}
\end{eqnarray}
so that the solution of the problem is
\begin{equation}\label{s3e29}
u(x) = x_1x_2.
\end{equation}
Further, the Jacobian of transformation is
\[
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta_0)} = -\frac{2e^\sigma}{\eta_0} \ne 0
\]
so that the inversion is always possible.

The parametric representation of the circle $x_1^2 + x_2^2 = 1$ is $x_1 = \cos\eta$ and $x_2 = 
\sin\eta$. The parametric equations of the characteristic curves and $u$ are \eqref{s3e21}, 
\eqref{s3e22} and \eqref{s3e23} but now the initial conditions are $x_1(\sigma = 0) = \cos\eta_0$,
$x_2(\sigma = 0) = \sin\eta_0$ and $u(\sigma = 0) = 1$. Therefore, their solutions are
\begin{eqnarray}
x_1 &=& e^{\sigma}\cos\eta_0  \label{s3e30} \\
x_2 &=& e^{\sigma}\sin\eta_0 \label{s3e31} \\
u &=& e^{2\sigma} \label{s3e32}
\end{eqnarray}
The equation of the characteristic curve is $x_2 = x_1\tan\eta_0$. It is a straight line passing
through the origin with a slope $\tan\eta_0$. The first two of the above equations can be inverted 
to get
\begin{eqnarray}
e^\sigma &=& \sqrt{x_1^2 + x_2^2} \label{s3e33} \\
\eta_0 &=& \tan^{-1}\left(\frac{x_2}{x_1}\right) \label{s3e34}
\end{eqnarray}
Since $x_1 > 0, x_2 > 0$, these inversions are always possible and the solution of the problem is
\begin{equation}\label{s3e35}
u(x) = x_1^2 + x_2^2.
\end{equation}
Further, the Jacobian of transformation is
\[
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta_0)} = e^{2\sigma} \ne 0
\]
so that the inversion is always possible.

Now suppose that the initial curve was $x_2 = e^{x_1}$. Then we can parametrize it as $x_1 = \eta$
and $x_2 = e^\eta$. The initial conditions for the equations \eqref{s3e21}, \eqref{s3e22} and 
\eqref{s3e23} are $x_1(\sigma = 0) = \eta_0, x_2(\sigma = 0)$ and $u(\sigma = 0) = 1$ and hence
their solutions are
\begin{eqnarray}
x_1 &=& e^{\sigma}\eta_0  \label{s3e36} \\
x_2 &=& e^{\sigma + \eta_0} \label{s3e37} \\
u &=& e^{2\sigma}. \label{s3e38}
\end{eqnarray}
Once again, the characteristic curve is a straight line through the origin with equation 
\[
x_2 = \frac{e^{\eta_0}}{\eta_0}x_1.
\]
However, the Jacobian of transformation is
\[
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta_0)} = e^{2\sigma + \eta_0}(\eta_0 - 1) 
\]
so that the inversion is possible if $\eta_0 \ne 1$.

\item (Problem Set 1, Ex. 3a) $x_1D^{e_1}u + x_2D^{e_2}u = ku, x \in \mathbb{R}, 0 < \alpha \le y$
with the initial condition $u(x, \alpha) = F(x)$, where $F$ is a smooth function. $k$ and $\alpha$ 
are constants.

The initial curve is a straight line parallel to the $x_1$ axis with parametric equation $x = \eta, y = 
\alpha$. The parametric equations of the characteristic curve and $u$ are
\begin{eqnarray}
D_\sigma x_1 &=& x_1 \label{s3e39} \\
D_\sigma x_2 &=& x_2 \label{s3e40} \\
D_\sigma u &=& ku \label{s3e41},
\end{eqnarray}
with initial conditions $x_1(\sigma = 0) = \eta_0, x_2(\sigma = 0) = \alpha, u(\sigma = 0) = 
F(\eta_0)$. The solutions of these equations are
\begin{eqnarray}
x_1 &=& \eta_0 e^\sigma \label{s3e42} \\
x_2 &=& \alpha e^\sigma \label{s3e43} \\
u &=& F(\eta_0)e^{k\sigma}. \label{s3e44}
\end{eqnarray}
The characteristics are straight lines with the equation 
\begin{equation}\label{s3e45}
x_2 = \frac{\alpha}{\eta_0}x_1
\end{equation}
Equations \eqref{s3e42} and \eqref{s3e43} can be inverted to get
\begin{eqnarray}
e^\sigma &=& \frac{x_2}{\alpha} \label{s3e46} \\
\eta_0 &=& \frac{\alpha x_1}{x_2} \label{s3e47}
\end{eqnarray}
Since the Jacobian of the transformation is $-\alpha e^{2\sigma}$, it is never zero and hence the
inversion of equations \eqref{s3e46} and \eqref{s3e47} is always possible. The solution of the problem
is then
\begin{equation}\label{s3e48}
u(x) = F\left(\frac{\alpha x_1}{x_2}\right)\left(\frac{x_2}{\alpha}\right)^k.
\end{equation}

\item (Problem Set 1, Ex. 3b) $(x_1 + 2)D^{e_1}u + 2x_2D^{e_2}u = \alpha u$ with initial condition 
$u(-1, x_2) = \sqrt{x_2}$. The initial curve is the line $x_1 = -1$. Its parametric form is $x_1 = -1,
x_2 = \eta$. The parametric equations of the characteristic curve along with the compatibility 
condition are
\begin{eqnarray}
D_\sigma x_1 &=& x_1 + 2 \label{s3e49} \\
D_\sigma x_2 &=& 2x_2 \label{s3e50} \\
D_\sigma u &=& \alpha u \label{s3e51}
\end{eqnarray}
with the initial conditions $x_1(\sigma = 0) = -1, x_2(\sigma = 0) = \eta, u(\sigma = 0) = 
\sqrt{\eta}$. Their solutions are
\begin{eqnarray}
x_1 &=& e^\sigma - 2 \label{s3e52} \\ 
x_2 &=& \eta_0 e^{2\sigma} \label{s3e53} \\
u &=& \sqrt{\eta_0}e^{\alpha\sigma} \label{s3e54}
\end{eqnarray}
The characteristic curve is $x_2 =\eta_0(x_1 + 2)^2$ and the Jacobian of transformation is $e^{2\sigma}
$ which is never zero so that the first two of the above equations can be inverted to get
\begin{eqnarray}
e^\sigma &=& x_1 + 2 \label{s3e55} \\
\eta_0 &=& \frac{x_2}{(x_1 + 2)^2}. \label{s3e56}
\end{eqnarray}
The solution of the problem is
\begin{equation}\label{s3e57}
u(x) = \sqrt{x_2}(x_1 + 2)^\alpha
\end{equation}

\item (Problem Set 1, Ex. 3c) $x_2D^{e_1}u - x_1D^{e_2}u = 0$ with $u(x_1, 0) = x_1^2$. The initial
curve is the line $x_2 = 0$ whose parametric form is $x_1 = \eta, x_2 = 0$. The parametric equations
of the characteristics along with the compatibility condition are
\begin{eqnarray}
D_\sigma x_1 &=& x_2 \label{s3e58} \\
D_\sigma x_2 &=& -x_1 \label{s3e59} \\
D_\sigma u &=& 0 \label{s3e60}
\end{eqnarray}
with initial conditions $x_1(\sigma = 0) = \eta_0, x_2(\sigma = 0) = 0, u(\sigma = 0) = 0$. 
Differentiating equation \eqref{s3e58} with respect to $\sigma$, $D_{\sigma\sigma} = -x_1$, whose
solution is $x_1 = a\cos\sigma + b\sin\sigma$. Substituting it in \eqref{s3e58} we immediately get
$x_2 = -a\sin\sigma + b\cos\sigma$. The initial conditions give $a = \eta$ and $b = 0$ so that
the solutions of the first two equations are
\begin{eqnarray}
x_1 &=& \eta\cos\sigma \label{s3e61} \\
x_2 &=& -\eta\sin\sigma \label{s3e62}
\end{eqnarray}
The solution of equation \eqref{s3e60} is $u(\sigma) = c$, which after using the initial conditions
gives $u(\sigma) = \eta^2$. Using equations \eqref{s3e61} and \eqref{s3e62} in the expression for $u$
we get the solution
\begin{equation}\label{s3e63}
u(x) = x_1^2 + x_2^2.
\end{equation}

\item (Problem Set 1, Ex. 3d) $x_1^2D^{e_1}u - x_2^2D^{e_2}u = 0$ with $u(1, x_2) = F(x_2)$. The 
initial curve is the line $x_1 = 1$ whose parametric form is $x_1 = 1, x_2 = \eta$. The parametric
equations of the characteristics along with the compatibility condition are
\begin{eqnarray}
D_\sigma x_1 &=& x_1^2 \label{s3e64} \\
D_\sigma x_2 &=& -x_2^2 \label{s3e65} \\
D_\sigma u &=& 0 \label{s3e66}
\end{eqnarray}
with initial conditions $x_1(\sigma = 0) = 1, x_2(\sigma = 0) = \eta, u(\sigma = 0) = F(x_2)$. The 
solution to the first two equations are
\begin{eqnarray*}
-\frac{1}{x_1} &=& \sigma + a \\
\frac{1}{x_2} &=& \sigma + b.
\end{eqnarray*}
Using the initial conditions, we get $a = -1$ and $b = \eta^{-1}$ so that
\begin{eqnarray}
1 -\frac{1}{x_1} &=& \sigma \label{s3e67} \\
\frac{1}{x_2} &=& \sigma + \frac{1}{\eta} \label{s3e68}.
\end{eqnarray}
The solution of \eqref{s3e66} is $u = c$. With the initial conditions $u(\sigma) = F(\eta)$. Using
equations \eqref{s3e67} and \eqref{s3e68} we get the solution
\begin{equation}\label{s3e69}
u(x) = F\left(\frac{x_1x_2}{x_1 + x_2 - x_1x_2}\right).
\end{equation}

\item (Problem Set 1, Ex. 4) Find the characteritic curves of
\begin{enumerate}
\item $(x_1^2 - x_2^2 + 1)D^{e_1}u + 2x_1x_2D^{e_2}u = 0$. The equations of characteristics are
\[
\frac{dx_1}{x_1^2 - x_2^2 + 1} = \frac{dx_2}{2x_1x_2}.
\]
This is an exact differential of
\[
f(x_1, x_2) = x_1^2x_2 - \frac{x_2^3}{3} + x_2 = c,
\]
where $c$ is a constant. This is also the equation of characteristics.
\item $2x_1x_2D^{e_1}u - (x_1^2 + x_2^2)D^{e_2}u = 0$. The equations of characteristics are
\[
\frac{dx_1}{2x_1x_2} = -\frac{dx_2}{x_1^2 + x_2^2}.
\]
This is an exact differential of
\[
f(x_1, x_2) = x_1x_2^2 + \frac{x_1^3}{3} + x_2 = c,
\]
where $c$ is a constant. This is also the equation of characteristics.
\end{enumerate}
\end{enumerate}

\section{First Order Quasilinear PDE}\label{s4}
Continuing to work with the simple case of two independent variables, the general form of a 
quasilinear PDE is given by \eqref{s2e4}
\[
a_1(x_1, x_2, u)D^{e_1}u + a_2(x_1, x_2, u)D^{e_2}u = a_3(x_1, x_2, u).
\]
If we define $f = u(x_1, x_2) - x_3$ then the above equation can be written as
\begin{equation}\label{s4e1}
(a_1, a_2, a_3)\cdot Df = 0
\end{equation}
The gradient vector $Df$ is normal to the surface $f = 0$. Therefore, the vector $(a_1, a_2, a_3)$ is
tangent to it. The direction of $(a_1, a_2, a_3)$ at any point $(x_1, x_2, x_3)$ is called the 
\emph{Monge direction}. A surface $x_3 = u(x_1, x_2)$ is an integral surface if and only if at every
point on it, the tangent plane at the point contains the Monge direction. A space curve whose tangent
at every point coincides with the Monge direction is called the \emph{Monge curve}. Its parametric
equations are
\begin{eqnarray}
D_\sigma x_1 &=& a_1(x_1, x_2, u) \label{s4e2} \\
D_\sigma x_2 &=& a_2(x_1, x_2, u) \label{s4e3} \\
D_\sigma x_3 &=& a_3(x_1, x_2, u) \label{s4e4}
\end{eqnarray}
A solution to this system of ODEs is a $2$-parameter family of curves. (We can combine them to get
$D_{x_1}x_2 = a_2/a_1$ and $D_{x_1}x_3 = a_3/a_1$. Solving them gives one constant for each one of
them.) A particular solution of these equations will need two conditions at $\sigma = 0$ for each
one of $x_1, x_2, x_3$. They are provided by conditions of the form $x_1 = X_1(\eta_1, \eta_2), 
x_2 = X_2(\eta_1, \eta_2), x_3 = X_3(\eta_1, \eta_2)$, defining the `initial surface' of the problem.
The Monge curves, thus, are a $2$-parameter family of curves with parametric equations $x_1 = x_1(
\sigma, \eta_1, \eta_2), x_2 = x_2(\sigma, \eta_1, \eta_2), x_3 = x_3(\sigma, \eta_1, \eta_2)$.

The projection of the Monge curve on the $x_1x_2$ plane is called the characteristic curve of the
problem.

We argued that the Monge curves are a $2$-parameter family of curves. However, we can always fix
one of the parameters to a constant, say $0$, and consider them as a $1$-parameter family of
curves. We will show that the $1$-parameter sub-family of Monge curves is equally good to find 
solutions of a quasi-linear PDE.
\begin{thm}\label{s4t1}
Every surface generated by a $1$-parameter sub-family of Monge curves is an integral surface.
\end{thm}
\begin{proof}
Consider a surface generated by a $1$-parameter sub-family of Monge curves obtained by setting the 
second parameter to a constant, say $\eta_0$. At every point on this surface the Monge direction 
coincides with the tangent to the surface. Therefore, the tangent plane at every point has the
Monge vector, making it an integral surface.
\end{proof}

The converse of the above theorem is also true.
\begin{thm}\label{s4t2}
Every integral surface can be considered to be generated by a $1$-parameter sub-family of Monge
curves.
\end{thm}
\begin{proof}
Let $x_3 = u(x_1, x_2)$ be an integral surface of the quasi-linear PDE
\begin{equation}\label{s4e5}
a_1(x_1, x_2, u)D^{e_1}u + a_2(x_1, x_2, u)D^{e_2}u = a_3(x_1, x_2, u)
\end{equation}
and let $x_1 = X_1(\eta), x_2 = X_2(\eta)$ and
$x_3 = u(X_1(\eta), X_2(\eta))$ be the space curve on it. Further, let us assume that the curve does 
not coincide with a Monge curve. Consider a solution of the system of equations
\begin{eqnarray*}
D_\sigma x_1 &=& a_1(x_1, x_2, x_3(x_1, x_2)) \label{s4e6} \\
D_\sigma x_2 &=& a_2(x_1, x_2, x_3(x_1, x_2)) \label{s4e7}
\end{eqnarray*}
with initial conditions $x_1(\sigma = 0) = X_1(\eta), x_2(\sigma = 0) = X_2(\eta)$. Since $x_3 = 
u(x_1, x_2)$ on the surface,
\begin{equation}\label{s4e8}
D_\sigma x_3 = D_\sigma x_1D^{e_1}u + D_\sigma x_1D^{e_1}u.
\end{equation}
Using the first three equations in the fourth one we get
\begin{equation}\label{s4e9}
D_\sigma x_3 = a_3(x_1, x_2, u).
\end{equation}
Equations \eqref{s4e6}, \eqref{s4e7} and \eqref{s4e9} tell that the curve $(x_1(\sigma, \eta), 
x_2(\sigma, \eta), x_3(\sigma, \eta))$ passing through a point $X_1(\eta), X_2(\eta), X_3(\eta)$ on the 
integral surface is a Monge curve. Thus, any integral surface can be generated by a $1$-parameter
family of Monge curves.
\end{proof}

The following theorem suggests a way to solve the Cauchy problem for a first-order quasi-linear PDE.
\begin{thm}\label{s4t3}
Let $X_i: [0, 1] \rightarrow \mathbb{R}$, $i = 1, 2, 3$, be continuously differentiable functions. 
Let $a_1, a_2, a_3: U_2 \rightarrow \mathbb{R}$, where $U_2 \subset \mathbb{R}^3$, have
continuous first order partial derivatives throughout $U_2$. Let the initial curve 
\[
\Gamma = \{(X_1(\eta), X_2(\eta), X_3(\eta): \eta \in [0, 1]\},
\]
be entirely in $U_2$ such that
\[
D_\eta X_2(\eta) a_1(X_1(\eta), X_2(\eta), X_3(\eta)) \ne 
D_\eta X_1(\eta) a_2(X_1(\eta), X_2(\eta), X_3(\eta)).
\]
Then there exists a solution $u(x_1, x_2)$ of the quasi-linear PDE
\[
a_1(x_1, x_2, u)D^{e_1}u + a_2(x_1, x_2, u)D^{e_2}u = a_3(x_1, x_2, u)
\]
in the neighborhood of the datum curve 
\[
\gamma = \{(X_1(\eta), X_2(\eta): \eta \in [0, 1]\},
\]
such that $u(\eta) = u(X_1(\eta), X_2(\eta))$.
\end{thm}
\begin{proof}
Since $a_1, a_2, a_3: U_2 \rightarrow \mathbb{R}$, where $U_2 \subset \mathbb{R}^3$, have
continuous first order partial derivatives throughout $U_2$, the system of equations
\begin{eqnarray*}
D_\sigma x_1 &=& a_1 \\
D_\sigma x_2 &=& a_2 \\
D_\sigma x_3 &=& a_3,
\end{eqnarray*}
where we have used $x_3 = u$, has a unique, continuously differentiable solution satisfying the
initial conditions
\begin{eqnarray*}
x_1(0, \eta) &=& X_1(\eta) \\
x_2(0, \eta) &=& X_2(\eta) \\
x_3(0, \eta) &=& X_3(\eta).
\end{eqnarray*}
Therefore, the functions $X_1, X_2, X_3$ too are continuously differentiable functions of $\eta$.
We assumed that the Jacobian
\[
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta)} = \pd{x_1}{\sigma}\pd{x_2}{\eta} - 
\pd{x_1}{\eta}\pd{x_2}{\sigma} = a_1\pd{x_2}{\eta} - a_2\pd{x_1}{\eta} \ne 0.
\]
Therefore, in the neighborhood of $\sigma = 0$, we can invert the relations $x_1 = x_1(\sigma, \eta)$
and $x_2 = x_2(\sigma, \eta)$ to get $\sigma = \sigma(x_1, x_2)$ and $\eta = \eta(x_1, x_2)$ and
substitute them in $x_3(\sigma, \eta)$ to get the solution in the form $x_3 = u(x_1, x_2)$. Further,
at any point on the datum curve $u(X_1(\eta), X_2(\eta)) = u(0, \eta) = x_3(0, \eta) = X_3(\eta)$
so that the initial condition on $D_\sigma x_2 = a_3$ is also satisfied.

We will now prove that the solution $u$ is unique. Let there be two integral surfaces $S_1$ and $S_2$
passing through the initial curve $\Gamma$. Let $(X_1(\eta_0), X_2(\eta_0), X_3(\eta_0)$ be a point 
common to both surfaces. Then the Monge curve through it is also common to both surfaces. Thus, both
surfaces are generated by the same family of Monge curves, which means that the two surfaces are 
identical.
\end{proof}

\section{Examples of Quasi-linear PDE}\label{s5}
\begin{enumerate}
\item (Example 1.10.2 of \cite{ta}) $uD^{e_1}u + D^{e_2}u = 1$ with the initial data $x_0 = s, y_0 = s,
u_0 = s/2$ for $s \in [0, 1]$. The parametric equations of the characteristics are
\begin{eqnarray}
D_\sigma x_1 &=& x_3 \label{s5e1} \\
D_\sigma x_2 &=& 1 \label{s5e2} \\
D_\sigma x_3 &=& 1 \label{s5e3}.
\end{eqnarray}
From equations \eqref{s5e3} and \eqref{s5e1},
\begin{equation}\label{s5e4}
D^2_\sigma x_1 = 1
\end{equation}
Therefore,
\begin{eqnarray}
x_1 &=& \frac{\sigma^2}{2} + c_1\sigma + c_2 \label{s5e5} \\
x_2 &=& \sigma + c_3 \label{s5e6} \\
x_3 &=& \sigma + c_4. \label{s5e7}
\end{eqnarray}
The initial conditions are $x_1(0, \eta) = s, x_2(0, \eta) = s, x_3(0, \eta) = s/2$. Therefore, the
solutions are
\begin{eqnarray}
x_1 &=& \frac{\sigma^2}{2} + c_1\sigma + s \label{s5e8} \\
x_2 &=& \sigma + s \label{s5e9} \\
x_3 &=& \sigma + \frac{s}{2}. \label{s5e10}
\end{eqnarray}
From equations \eqref{s5e1}, \eqref{s5e8} and \eqref{s5e10} we conclude that $c_1 = s/2$ so that
\begin{equation}\label{s5e11}
x_1 = \frac{\sigma^2}{2} + \frac{\sigma}{2}s + s
\end{equation}
We obtained the solution $u = x_3$ in terms of $s$ and $\sigma$. We need it in terms of $x_1, x_2$.
From equations \eqref{s5e11} and \eqref{s5e9},
\[
x_1 - x_2 = \frac{\sigma^2}{2} + \frac{\sigma}{2}s - \sigma = \frac{\sigma}{2}(\sigma + s) - \sigma
= \frac{\sigma}{2}x_2 - \sigma = \sigma\left(\frac{x_2}{2} - 1\right),
\]
so that
\begin{equation}\label{s5e12}
\sigma = \frac{x_1 - x_2}{x_2/2 - 1}.
\end{equation}
Using \eqref{s5e9}, we get
\begin{equation}\label{s5e13}
s = \frac{x_2^2/2 - x_1}{x_2/2 - 1}
\end{equation}
so that the solution of the problem is 
\begin{equation}\label{s5e14}
x_3 = u(x_1, x_2) = \frac{x_1 - x_2}{x_2/2 - 1} + \frac{1}{2}\frac{x_2^2/2 - x_1}{x_2/2 - 1} = 
\frac{2x_1 - 4x_2 + x_2^2}{2(x_2 - 2)}.
\end{equation}

\item (Example 2.3 of \cite{pprr}) $uD^{e_1}u + D^{e_2}u = 0$ with intial data (a) $u(x_1, 0) = x_1$
for $x_1 \in [0, 1]$ and (b) $u(x_1, 0) = 1/2$ for $x_1 \in [0, 1]$.

The parametric form of the data are (a) $\{x_1 = \eta, x_2 = 0, u (=x_3) = \eta\}$ and (b) $\{x_1 = 
\eta, x_2 = 0, x_3 = 1/2\}$.

The parametric equations of the Monge curves are
\begin{eqnarray*}
D_\sigma x_1 &=& u \\
D_\sigma x_2 &=& 1 \\
D_\sigma x_3 &=& 0
\end{eqnarray*}
Their solutions for the two data sets are
\begin{eqnarray*}
x_1 &=& \eta\sigma + \eta \\
x_2 &=& \sigma \\
x_3 &=& \eta
\end{eqnarray*}
and
\begin{eqnarray*}
x_1 &=& \frac{\sigma}{2}\sigma + \eta\\
x_2 &=& \sigma \\
x_3 &=& \frac{1}{2}
\end{eqnarray*}
respectively so that the characteristic curves are the lines $x_1 = \eta(x_2 + 1)$ and $x_1 = 
x_2/2 + \eta$. They are shown in figure \ref{f3}.
\begin{figure}
\centering
\includegraphics[scale=0.75]{cnf1}
\caption{Characteristic curves for the initial data.}
\label{f3}
\end{figure}

The characteristics $x_1 = \eta(x_2 + 1)$ all meet at the point $(0, -1)$. It happens because of the
form of the solution
\[
u = x_3 = \eta = \frac{x_1}{x_2 + 1}.
\]
Clearly, $u(x_1, x_2)$ has a singularity at $x_2 = -1$.

\item (Problem Set 1. Ex. 5b) $uD^{e_1}u + D^{e_2}u = 1$, $u(x, x) = x/2$ for $x \in [0, 1]$. The
parametric equations of the Monge curves are
\begin{eqnarray}
D_\sigma x_1 &=& u \label{s5e15} \\
D_\sigma x_2 &=& 1 \label{s5e16} \\
D_\sigma x_3 &=& 1 \label{s5e17}.
\end{eqnarray}
The general solution of \eqref{s5e17} is 
\begin{equation}\label{s5e18}
x_3 = \sigma + c_1
\end{equation}
so that the general solution of \eqref{s5e15} becomes 
\begin{equation}\label{s5e19}
x_1 = \frac{\sigma^2}{2} + c_1\sigma + c_2.
\end{equation}
Finally, the general solution of \eqref{s5e16} is
\begin{equation}\label{s5e20}
x_2 = \sigma + c_3.
\end{equation}
The initial data is $x_1(\sigma = 0, \eta) = \eta, x_2(\sigma = 0, \eta) = \eta \rightarrow 
x_3(\sigma = 0, \eta) = \eta/2$. Therefore $c_1 = \eta/2$, $c_2 = \eta$ and $c_3 = \eta$. Therefore
\begin{eqnarray}
x_1 &=& \frac{\sigma^2}{2} + \frac{\eta}{2}\sigma + \eta \label{s5e21} \\
x_2 &=& \sigma + \eta \label{s5e22} \\
x_3 &=& \sigma + \frac{\eta}{2} \label{s5e23}
\end{eqnarray}
The transversality condition is
\[
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta)} = \pd{x_1}{\sigma}\pd{x_2}{\eta} - 
\pd{x_1}{\eta}\pd{x_2}{\sigma} = \left(\sigma + \frac{\eta}{2}\right) - 
\left(\frac{\sigma}{2} + 1\right) = \frac{\sigma + \eta}{2} - 1
\]
Thus,
\[
\frac{\partial(x_1, x_2)}{\partial(\sigma, \eta)} \ne 1
\]
as long as $\sigma + \eta \ne 2$. This is definitely true for $x_1, x_2 \in [0, 1]$. Inverting
equations \eqref{s5e21} and \eqref{s5e22} we get
\begin{eqnarray}
\sigma &=& \frac{2(x_2 - x_1)}{2 - x_1} \label{s5e24} \\
\eta &=& \frac{x_2(2 - x_1)}{2 - x_1} \label{s5e25}
\end{eqnarray}
Using \eqref{s5e23}, \eqref{s5e24} and \eqref{s5e25} we get
\begin{equation}\label{s5e26}
u(x_1, x_2) = \frac{2(x_2 - x_1)}{2 - x_1} + \frac{1}{2}\frac{x_2(2 - x_1)}{2 - x_1}
\end{equation}

\item (Problem Set 1, Ex. 6) $uD^{e_1}u + D^{e_2}u = 0$ with several different initial conditions.
The parametric equations of the Monge curve are
\begin{eqnarray}
D_\sigma x_1 &=& x_3 \label{s5e27} \\
D_\sigma x_2 &=& 1 \label{s5e28} \\
D_\sigma x_3 &=& 0 \label{s5e29}
\end{eqnarray}
their general solution of which is
\begin{eqnarray}
x_1 &=& c_1\sigma + c_2 \label{s5e30} \\
x_2 &=& \sigma + c_3 \label{s5e31} \\
x_3 &=& c_1 \label{s5e32}
\end{eqnarray}
Let us now examine the various initial conditions.
\begin{enumerate}
\item \begin{equation}\label{s5e33}
u(x, 0) = \begin{cases}
0 \text{ if } x < 0 \\
1 \text{ if } x \ge 0
\end{cases}
\end{equation}
The parametric equations of the Monge curves are
\begin{equation}\label{s5e34}
x_1 = \begin{cases}
\eta \text{ if } x_1 < 0 \\
\sigma + \eta \text{ if } x_1 \ge 0
\end{cases}
\end{equation}
\begin{equation}\label{s5e35}
x_2 = \sigma
\end{equation}
We can combine equations \eqref{s5e34} and \eqref{s5e35} to get
\begin{equation}\label{s5e36}
x_1 = \begin{cases}
\eta \text{ if } x_1 < 0 \\
x_2 + \eta \text{ if } x_1 \ge 0
\end{cases}
\end{equation}
Thus, the characteristic curve is the negative $x_1$ axis and the parallel lines $x_2 = x_1 - \eta$
with $x_1 \ge 0$ for all $\eta$.

\item \begin{equation}\label{s5e37}
u(x, 0) = \begin{cases}
1 \text{ if } x < 0 \\
0 \text{ if } x \ge 0
\end{cases}
\end{equation}
The parametric equations of the Monge curves are
\begin{equation}\label{s5e38}
x_1 = \begin{cases}
\sigma + \eta \text{ if } x_1 < 0 \\
\eta \text{ if } x_1 \ge 0
\end{cases}
\end{equation}
\begin{equation}\label{s5e39}
x_2 = \sigma
\end{equation}
We can combine equations \eqref{s5e34} and \eqref{s5e35} to get
\begin{equation}\label{s5e40}
x_1 = \begin{cases}
\sigma + \eta \text{ if } x_1 < 0 \\
\eta \text{ if } x_1 \ge 0
\end{cases}
\end{equation}
Thus, the characteristic curve is the positive $x_1$ axis and the parallel lines $x_2 = x_1 - \eta$
with $x_1 \ge 0$ for all $\eta$.

\item \begin{equation}\label{s5e41}
u(x, 0) = \begin{cases}
0 \text{ if } x < 0 \\
1 \text{ if } x \ge 1
\end{cases}
\end{equation}
and $u(x, 0)$ is smooth and increasing in $[0, 1]$. Let us denote it by $\phi(x)$. Thus,
\begin{equation}\label{s5e42}
u(x, 0) = \begin{cases}
0 \text{ if } x < 0 \\
\phi(x) \text{ if } x \in [0, 1] \\
1 \text{ if } x \ge 1
\end{cases}
\end{equation}
for a smooth and increasing $\phi$. Thus the parametric equations of the Monge curves are $x_2 = 
\sigma$ and 
\begin{equation}\label{s5e43}
x_1 = \begin{cases}
\eta \text{ if } x_1 < 0 \\
\phi(\eta) \text{ if } x_1 \in [0, 1] \\
\sigma + \eta \text{ if } x_1 \ge 1
\end{cases}
\end{equation}
or
\begin{equation}\label{s5e44}
x_1 = \begin{cases}
\eta \text{ if } x_1 < 0 \\
\phi(\eta) \text{ if } x_1 \in [0, 1] \\
x_2 + \eta \text{ if } x_1 \ge 1
\end{cases}
\end{equation}
The characteristic curves is the negative $x_1$ axis, the function $\phi(\eta)$ for $x_1 \in [0, 1]$
and the lines $x_2 = x_1 - \eta$ for $x_1 > 1$.
\end{enumerate}
\end{enumerate}

\section{On envelopes of surfaces}\label{s6}\cite{fj}
Consider a one-parameter family of surfaces $S_\lambda$ defined by
\begin{equation}\label{s6e1}
x_3 = G(x_1, x_2, \lambda),
\end{equation}
where $\lambda$ is the parameter. Differentiating both sides with respect to $\lambda$, we get
\begin{equation}\label{s6e2}
0 = \pd{G}{\lambda}(x_1,x_2,\lambda).
\end{equation}
In principle, we can solve this equation to get $\lambda = g(x, y)$ so that the equation of the 
surface can be written as 
\begin{equation}\label{s6e3}
x_3 = G(x_1, x_2, g(x_1, x_2)).
\end{equation}
Equation \eqref{s6e3} defines the envelope $E$ of all one-parameter surfaces defined by \eqref{s6e1}.
It touches all surfaces $S_\lambda$ alonga curve $\gamma_\lambda$. Consider a point $(X_1, X_2, X_3)$
on one such curve $\gamma_\lambda$. Since it lies on a curve with parameter $\lambda$, we naturally
have $g(X_1, X_2) = \lambda$. The direction numbers of the normal to the envelope $E$ are
\[
D(G(x_1, x_2, g(x_1, x_2)) - x_3) = 
\left(\pd{G}{x_1} + \td{G}{\lambda}\pd{g}{x_1}, \pd{G}{x_2} + \td{G}{\lambda}\pd{g}{x_2}, -1\right)
\]
Using \eqref{s6e2}, we have
\[
D(G(x_1, x_2, g(x_1, x_2)) - x_3) = \left(\pd{G}{x_1}, \pd{G}{x_2}, -1\right)
\]
Naturally, these are same as the direction numbers of a normal to $S_\lambda$ given by \eqref{s6e1}. 
In the future, we will find it advantageous to use the differential forms of equations \eqref{s6e1}
\begin{equation}\label{s6e4}
dx_3 = D^{e_1}Gdx_1 + D^{e_2}Gdx_2
\end{equation}
and
\[
0 = \frac{\partial^2 G}{\partial x_1 \partial\lambda}dx_1 +
\frac{\partial^2 G}{\partial x_2 \partial\lambda}dx_2 + 
\frac{\partial^2 G}{\partial\lambda^2}d\lambda.
\]
Now on a curve $\gamma_\lambda$, $\lambda$ is fixed so that
\begin{equation}\label{s6e5}
0 = \frac{\partial^2 G}{\partial x_1 \partial\lambda}dx_1 + 
\frac{\partial^2 G}{\partial x_2 \partial\lambda}dx_2.
\end{equation}

\section{First order nonlinear PDE}\label{s7}
The general form of a first order nonlinear PDE is given by \eqref{s2e1}
\begin{equation}\label{s7e1}
F(x_1, x_2, u, p, q) = 0
\end{equation}
where $F:U^2 \times \mathbb{R} \times U^2 \rightarrow \mathbb{R}$, $D^{e_1}u = p$ and  
$D^{e_2}u = q$. If $f = u(x_1, x_2) - x_3$ then $Df = (p, q, -1)$ is a normal to the integral surface.
A plane normal to $Df$ will be tangent to the integral surface. Its general equation is
\begin{equation}\label{s7e2}
p(x_1 - x_1^0) + q(x_2 - x_2^0) - 1(x_3 - x_3^0) = 0,
\end{equation}
where $(x_1^0, x_2^0, x_3^0)$ is an arbitrary, but fixed, point on it. As yet, we do not know the 
solution $x_3 = u(x_1, x_2)$. Therefore, we do not know $p$ and $q$. Yet, we know that they are not
both arbitrary because they are related by equation \eqref{s7e1}. If we fix one of them, say $p$,
then \eqref{s7e1} gives the other and therefore equation \eqref{s7e2} is a one-parameter family of
planes passing through the point $(x_1^0, x_2^0, x_3^0)$. An envelope of this family of planes is
a cone, called the \emph{Monge cone}. There is a Monge cone on every point of the integral surface.
Equation \eqref{s7e1} can as well be interpreted as defining a field of Monge cones. The integral 
surface of the problem is tangent to all cones. Therefore, only one plane in the envelope of a given
Monge cone is tangential to the integral surface. Its generator defines the \emph{Monge direction} at 
that point.

For a fixed $p$ (and therefore also a fixed $q$ by virtue of \eqref{s7e1}), differentiating \
\eqref{s7e2} gives
\begin{equation}\label{s7e3}
pdx_1 + qdx_2 = dx_3.
\end{equation}
We can also differentiate \eqref{s7e2} with respect to the `parameter' $p$ to get the relation
\[
(x_1 - x_1^0) + \td{q}{p}(x_2 - x_2^0) = 0.
\]
If we differentiate it once more and retain only the first order terms we get
\begin{equation}\label{s7e4}
dx_1 + \td{q}{p}dx_2 = 0
\end{equation}
which can be written as 
\begin{equation}\label{s7e5}
\td{q}{p} = -\frac{dx_1}{dx_2}.
\end{equation}
From equation \eqref{s7e1} we get
\[
F_{p}dp + F_{q}dq = 0
\]
which is same as
\begin{equation}\label{s7e6}
\td{q}{p} = -\frac{F_{p}}{F_{q}}.
\end{equation}
From equations \eqref{s7e5} and \eqref{s7e6} we get
\begin{equation}\label{s7e7}
\frac{dx_1}{F_{p}} = \frac{dx_2}{F_{q}}.
\end{equation}
If $\sigma$ is a parameter then the previous equation also means
\begin{eqnarray}
\td{x_1}{\sigma} &=& F_{p} \label{s7e8} \\
\td{x_2}{\sigma} &=& F_{q} \label{s7e9}
\end{eqnarray}
so that if we also differentiate equation \eqref{s7e3} with respect to $\sigma$ we get
\begin{equation}\label{s7e10}
\td{x_3}{\sigma} = p\td{x_1}{\sigma} + q\td{x_2}{\sigma}.
\end{equation}
Equation \eqref{s7e10} is also called the \emph{strip condition}. In fact, we can combine equations 
\eqref{s7e8}, \eqref{s7e9} and \eqref{s7e10} to get
\begin{equation}\label{s7e11}
\frac{dx_1}{F_{p}} = \frac{dx_2}{F_{q}} = \frac{dx_3}{pF_{p} + qF_{q}}
\end{equation}
This is the parametric equation of the Monge curves.

For a quasi-linear equation $F = a_1(x_1, x_2, u)p + a_2(x_1, x_2, u)q - a_3(x_1, x_2, u)$ so that 
equation \eqref{s7e11} gives
\[
\frac{dx_1}{a_1} = \frac{dx_2}{a_2} = \frac{dx_3}{pa_1 + qa_2}.
\]
Since $pa_1 + qa_2 = a_2$, we get
\[
\frac{dx_1}{a_1} = \frac{dx_2}{a_2} = \frac{dx_3}{a_3}
\]
which is just another form of equations the parametric equations \eqref{s4e2} to \eqref{s4e4} of its
Monge curve.

Coming back to equation \eqref{s7e11}, we have derived three equations for three unknowns $x_1, x_2,
x_3$ as functions of $\sigma$. We have two more unknowns $p$ and $q$. We need a differential equation
for each one of them in terms of the parameter $\sigma$. To get them we begin with the definition 
$p = D^{e_1}u$ so that
\begin{equation}\label{s7e12}
\td{p}{\sigma} = \frac{\partial^2 u}{\partial x_1^2}\td{x_1}{\sigma} + 
\frac{\partial^2 u}{\partial x_1\partial x_2}\td{x_2}{\sigma} 
\end{equation}
Using \eqref{s7e11}, we get
\begin{equation}\label{s7e13}
\td{p}{\sigma} = \pd{p}{x_1}F_{p} + \pd{p}{x_2}F_{q}
\end{equation}
Since $F(x_1, x_2, u, p, q) = 0$, we also have
\begin{equation}\label{s7e14}
\pd{F}{x_1} + \pd{F}{u}\pd{u}{x_1} + F_{p}\pd{p}{x_1} + F_{q}\pd{q}{x_1} = 0
\end{equation}
From equations \eqref{s7e13} and \eqref{s7e14} we get the parametric equation of $p$ 
\begin{equation}\label{s7e15}
D_\sigma p = -F_{x_1} - F_up.
\end{equation}
Similarly, from equation
\begin{equation}\label{s7e16}
\td{q}{\sigma} = \frac{\partial^2 u}{\partial x_1\partial x_2}\td{x_1}{\sigma} +
\frac{\partial^2 u}{\partial x_2^2}\td{x_2}{\sigma} = \pd{q}{x_1}F_{p} + \pd{q}{x_2}F_{q}
\end{equation}
and
\begin{equation}\label{s7e17}
\pd{F}{x_2} + \pd{F}{u}\pd{u}{x_2} + F_{p}\pd{q}{x_1} + F_{q}\pd{q}{x_2} = 0
\end{equation}
we get the parametric equation for $q$ as
\begin{equation}\label{s7e18}
D_\sigma q = -F_{x_2} - F_uq.
\end{equation}
Note that
\[
\pd{q}{x_1} = \pd{q}{x_2}.
\]
makes equations \eqref{s7e14} and \eqref{s7e17} identical. The equations of the Monge curves are 
\begin{equation}\label{s7e19}
\frac{dx_1}{F_{p}} = \frac{dx_2}{F_{q}} = \frac{dx_3}{pF_{p} + qF_{q}} = 
\frac{dp}{-F_{x_1} - F_{u}p} = \frac{dq}{-F_{x_2} - F_{u}q}
\end{equation}
Using these equations it is easy to conclude that
\begin{equation}\label{s7e20}
D_\sigma F = 0
\end{equation}
or that $F$ is constant along a Monge curve. Equation \eqref{s7e19} is a set of five ordinary 
differential equations. We need five initial conditions to solve them. Three are given by the initial
data. The initial values of $p$ and $q$ are obtained as follows. Let $p(\sigma = 0) = p_0$ and 
$q(\sigma = 0) = q_0$. Then the following equations are true
\begin{eqnarray}
F(x_1^0, x_2^0, u, p, q) &=& 0 \label{s7e21} \\
\td{x_3}{\sigma} = p_0\td{x_1}{\sigma} + q_0\td{x_2}{\sigma}. \label{s7e22} 
\end{eqnarray}
The second equation is just the strip condition \eqref{s7e10} at $\sigma = 0$.

\section{Examples of nonlinear PDE}\label{s8}
\begin{enumerate}
\item (Problem Set 1, Ex. 7) $q = p^3$ with the initial condition $u = 2x_1^{3/2}$ in the $x_1$ axis. 
Let us first get the initial data to solve the system of equations \eqref{s7e19}. Let it be $(x_1^0,
x_2^0, x_3^0 = u_0, p_0, q_0)$. The parameteric equation of the initial curve is $x_1 = \eta, x_2 = 0,
x_3 = 2\eta^{3/2}$. $p_0$ and $q_0$ can be chosen from the conditions
\begin{eqnarray}
q_0 &=& p_0^{3} \label{s8e1} \\
\td{x_3}{\eta} &=& p_0\td{x_1}{\eta} + q_0\td{x_2}{\eta} \label{s8e2}
\end{eqnarray}
The second equation reduces to
\[
3\eta^{1/2} = p_0 \cdot 1
\]
Using this equation in \eqref{s8e1} we get
\begin{eqnarray}
p_0 &=& 3\eta^{1/2} \label{s8e3} \\
q_0 &=& 3\eta^{3/2} \label{s8e4}
\end{eqnarray}
If we write the PDE as $F = p^3 - q = 0$ so that the equations of the characteristic strip are
\[
\frac{dx_1}{3p^2} = \frac{dx_2}{-1} = \frac{dx_3}{3p^3 - q} = \frac{dp}{} = \frac{dq}{}
\]
The system of ODE is
\begin{eqnarray}
D_\sigma x_1 &=& 3p^2 \label{s8e5} \\
D_\sigma x_2 &=& -1 \label{s8e6} \\
D_\sigma x_3 &=& 3p^3 - q \label{s8e7} \\
D_\sigma p &=& 0 \label{s8e8} \\
D_\sigma q &=& 0. \label{s8e9}
\end{eqnarray}
subject to the initial conditions $x_1(\sigma = 0) = \eta, x_2(\sigma = 0) = 0, x_3(\sigma = 0) =
 2\eta^{3/2}, p_0 = 3\eta^{1/2}, q_0 = 27\eta^{3/2}$.

The solutions of the last two equations are $p = p_0= 3\eta^{1/2}, q = q_0 = 27\eta^{3/2}$. The
solutions of first two equations are $x_1 = 3p_0^2\sigma + c_1$ and $x_2 = -\sigma + c_2$. Since the
initial data is on the $x_1$ axis, $x_1 = \eta$ and $x_2 = 0$ when $\sigma = 0$ so that we have
\begin{eqnarray}
x_1 &=& 27\eta\sigma + \eta \label{s8e10} \\
x_2 &=& -\sigma. \label{s8e11}
\end{eqnarray}
These equations can be inverted to get
\begin{eqnarray}
\sigma &=& -x_2 \label{s8e12} \\
\eta &=& \frac{x_1}{1 - 27 x_2} \label{s8e13}
\end{eqnarray}
The solution of \eqref{s8e7} subject to the initial condition $x_3(\sigma = 0) = 2\eta^{3/2}$ is
\begin{equation}\label{s8e14}
u = x_3 = 78\eta^{3/2}\sigma + 2\eta^{3/2} = 2\eta^{3/2}(27\sigma + 1).
\end{equation}
From equations \eqref{s8e13} and \eqref{s8e14} we get the solution
\begin{equation}\label{s8e15}
u(x_1, x_2) = \frac{2x_1^{3/2}}{(1 - 27x_2)^{1/2}}.
\end{equation}
It is easy to verify that
\begin{eqnarray*}
p &=& \frac{3x_1^{1/2}}{(1 - 27x_2)^{1/2}} \\
q &=& \frac{27x_1^{3/2}}{(1 - 27x_2)^{3/2}}
\end{eqnarray*}
so that $p^3 = q$.

\item (Problem Set 1, Ex. 8) $x_1p^2 + x_2q = u$ passing through the line $x_2 = 1$, $x_1 + x_3 = 0$.
Let $(x_1^0, x_2^0, x_3^0, p_0, q_0)$ be the initial data for the system of equations \eqref{s7e19}.
The parametric equations of the initial curve are $x_2 = 1, x_1 = \eta, x_3 = -\eta$. The initial
values of the first derivatives can be obtained from the equations
\begin{eqnarray*}
\eta p_0^2 + 1\cdot q_0 &=& -\eta \\
\td{x_3}{\eta} &=& p_0\td{x_1}{\eta} + q_0\td{x_2}{\eta}
\end{eqnarray*}
The second equation reduces to
\[
-1 = p_0\cdot 1 + q_0\cdot 0.
\]
From these equations we get
\begin{eqnarray}
p_0 &=& -1 \label{s8e16} \\
q_0 &=& -2\eta. \label{s8e17}
\end{eqnarray}

If we write the PDE as $F = x_1p^2 + x_2q - u = 0$ then the system of ODEs is
\begin{eqnarray}
D_\sigma x_1 &=& 2px_1 \label{s8e18} \\
D_\sigma x_2 &=& x_2 \label{s8e19} \\
D_\sigma x_3 &=& 2p^2x_1 + qx_2 \label{s8e20} \\
D_\sigma p &=& -p^2 + p \label{s8e21} \\
D_\sigma q &=& -q + q = 0 \label{s8e22}
\end{eqnarray}
The solution of the last equation is 
\begin{equation}\label{s8e23}
q(\sigma) = -2\eta.
\end{equation}
The solution of equation \eqref{s8e21} is
\begin{equation}\label{s8e24}
p(\sigma) = \frac{e^\sigma}{e^\sigma - 2}.
\end{equation}
Substituting it in \eqref{s8e18} we get
\begin{equation}\label{s8e25}
x_1(\sigma) = \eta(e^\sigma - 2)^2.
\end{equation}
The solution of \eqref{s8e19} is 
\begin{equation}\label{s8e26}
x_2(\sigma) = e^\sigma.
\end{equation}
Using \eqref{s8e23}, \eqref{s8e24},
\eqref{s8e25} in equation \eqref{s8e20} we get
\[
D_\sigma x_3 = 2\eta e^{2\sigma} - 2\eta e^\sigma
\]
so that
\[
x_3(\sigma) = \eta e^{2\sigma} - 2\eta e^\sigma + c,
\]
where $c$ is a constant of integration. Since $x_3(\sigma = 0) = -\eta$, we have
\begin{equation}\label{s8e27}
x_3(\sigma) = \eta e^{2\sigma} - 2\eta e^\sigma.
\end{equation}
Inverting \eqref{s8e24} and \eqref{s8e25}, we get
\begin{eqnarray}
e^\sigma &=& x_2 \label{s8e28} \\
\eta &=& \frac{x_1}{(x_2 - 2)^2} \label{s8e29}.
\end{eqnarray}
Substituting \eqref{s8e27} and \eqref{s8e28} in \eqref{s8e26} we get
\begin{equation}\label{s8e30}
u(x_1, x_2) = \frac{x_1x_2}{x_2 - 2}.
\end{equation}
We can readily verify that
\begin{eqnarray*}
p &=& \frac{x_2}{x_2 - 2} \\
q &=& \frac{-2x_1}{(x_2 - 2)^2}
\end{eqnarray*}
so that
\[
x_1 p^2 + x_2 q = \frac{x_1x_2}{x_2 - 2} = u.
\]

\item (Problem Set 1, Ex. 9) $p^2 + q^2 = 1$ with $u(x, y) = 0$ on $x + y = 1$. Let $(x_1^0, x_2^0, 
x_3^0, p_0, q_0)$ be the initial data for the system of equations \eqref{s7e19}. The parametric
form of the initial data is $x_1 = \eta, x_2 = 1 - \eta, x_3 = 0$. The initial values of the first
derivatives can be obtained from the equations
\begin{eqnarray*}
p_0^2 + q_0^2 &=& 1 \\
p_0\td{x_1}{\eta} + q_0\td{x_2}{\eta} &=& \td{x_3}{\eta},
\end{eqnarray*}
the second equation is
\[
p_0 - q_0 = 0
\]
Thus we have
\begin{equation}\label{s8e31}
p_0 = q_0 = \pm\frac{1}{\sqrt{2}}.
\end{equation}
If we write the PDE as $F = p^2 + q^2 - 1 = 0$ then the system of ODEs is
\begin{eqnarray}
D_\sigma x_1 &=& 2p \label{s8e32} \\
D_\sigma x_2 &=& 2q \label{s8e33} \\
D_\sigma x_3 &=& 2(p^2 + q^2) \label{s8e34} \\
D_\sigma p &=& 0 \label{s8e35} \\
D_\sigma q &=& 0 \label{s8e36}
\end{eqnarray}
The solutions of the first three equations is
\begin{eqnarray}
x_1 &=& 2p_0\sigma + \eta \label{s8e37} \\
x_2 &=& 2p_0\sigma + 1 - \eta \label{s8e38} \\
x_3 &=& 4p_0^2 \sigma \label{s8e39}
\end{eqnarray}
From equations \eqref{s8e37} and \eqref{s8e38}
\[
\sigma = \frac{x_1 + x_2 - 1}{4p_0}
\]
so that the solution is 
\begin{equation}\label{s8e40}
u(x_1, x_2) = \pm\frac{x_1 + x_2 - 1}{\sqrt{2}}.
\end{equation}
We can readily verify that
\begin{eqnarray*}
p &=& \pm\frac{1}{\sqrt{2}} \\
q &=& \pm\frac{1}{\sqrt{2}} 
\end{eqnarray*}
so that
\[
p^2 + q^2 = 1.
\]
\end{enumerate}

\section{General first order PDE}\label{s9}
If $x = (x_1, \ldots, x_n) \in \mathbb{R}^n$, $u: \mathbb{R}^n \rightarrow \mathbb{R}$. Then
$F(x, u, Du) = 0$, where $F:\mathbb{R}^n \times \mathbb{R} \times \mathbb{R}^n \rightarrow \mathbb{R}$ 
in a domain $U \subset \mathbb{R}^n$ is a general first order partial differential equation in 
$n$-variables. Let $S$ be a hypersurface in $\mathbb{R}^n$. Then the initial condition is $u = g$ on $S$.

Let $u$ be a solution of the problem and let $x$ be a fixed point in $U$. We want to find $u$
along a curve joining $x$ to a point $x_0$ on $S$. Since $u = g$ on $x_0$, we know the value of
$u$ at one end of this curve. Let us suppose that this curve is described by a parameter $\sigma
\in I \subset \mathbb{R}$. That is, $x(\sigma) = (x_1(\sigma), \ldots, x_n(\sigma))$ so that we
also have
\begin{equation}\label{s9e1}
z(\sigma) := u(x(\sigma))
\end{equation}
and
\begin{equation}\label{s9e2}
p(\sigma) := D(u(x(\sigma))).
\end{equation}
Thus, $z$ gives the value of the solution along the curve $x(s)$ and $p$ gives the value of its 
gradient along the same curve. In component form, equation \eqref{s9e2} is
\begin{equation}\label{s9e3}
p_i(\sigma) = \pd{u(x(\sigma))}{x_i}
\end{equation}
so that
\begin{equation}\label{s9e4}
\td{p_i}{\sigma} = \sum_{j=1}^n\frac{\partial^2 u}{\partial x_i \partial x_j}\td{x_i}{\sigma}.
\end{equation}
In order to get rid of the second derivatives in the above equation, let us differentiate the 
functional form of the PDE $F(p, z, x) = 0$ with respect to $x_i$.
\[
\td{F}{\sigma} = \sum_{j=1}^n\pd{F}{p_j}\td{p_j}{x_i} + \pd{F}{z}\td{z}{x_i} + \pd{F}{x_i} = 0.
\]
Using \eqref{s9e3} in the above equation we get
\begin{equation}\label{s9e5}
\td{F}{\sigma} = \sum_{j=1}^n\pd{F}{p_j}\frac{\partial^2 u}{\partial x_i \partial x_j} + 
\pd{F}{z}\td{z}{x_i} + \pd{F}{x_i} = 0.
\end{equation}
Let us choose the curve $x(\sigma)$ so that the coefficients of the second derivatives in
equations \eqref{s9e4} and \eqref{s9e5} agree with each other, that is, let us set
\begin{equation}\label{s9e6}
\td{x_i}{\sigma} = \pd{F}{p_j}.
\end{equation}
Let us write \eqref{s9e5} in the form
\[
\sum_{j=1}^n\pd{F}{p_j}(x, u, Du)\frac{\partial^2 u}{\partial x_i \partial x_j} + 
\pd{F}{z}(x, u, Du)\td{z}{x_i} + \pd{F}{x_i}(x, u, Du) = 0
\]
which emphasizes that it is valid throughout $U$. If we restrict ourselves to the curve $x = 
x(\sigma)$ then, recalling equations \eqref{s9e1} and \eqref{s9e2}, $u = z$ and $Du = p$ so that
\begin{equation}\label{s9e7}
\sum_{j=1}^n\pd{F}{p_j}(x, z, p)\frac{\partial^2 u}{\partial x_i \partial x_j} + 
\pd{F}{z}(x, z, p)\td{z}{x_i} + \pd{F}{x_i}(x, z, p) = 0
\end{equation}
Substituting \eqref{s9e6} into \eqref{s9e4} we get
\begin{equation}\label{s9e8}
\td{p_i}{\sigma} = \sum_{j=1}^n\frac{\partial^2 u}{\partial x_i \partial x_j}\pd{F}{p_j}.
\end{equation}
From equations \eqref{s9e7} and \eqref{s9e8} we get
\begin{equation}\label{s9e9}
\td{p_i}{\sigma} = -\pd{F}{z}(x, z, p)\td{z}{x_i} - \pd{F}{x_i}(x, z, p)
\end{equation}
Differentiating \eqref{s9e1} with respect to $\sigma$ we get
\begin{equation}\label{s9e10}
\td{z}{\sigma} = \sum_{i=1}^n\pd{u}{x_i}\td{x_i}{\sigma}.
\end{equation}
Using \eqref{s9e3} and \eqref{s9e6} in the above equation, we get
\begin{equation}\label{s9e11}
\td{z}{\sigma} = \sum_{i=1}^n p_i\pd{F}{p_i}.
\end{equation}
Equations \eqref{s9e6}, \eqref{s9e9} and \eqref{s9e11} define the curve along which we can obtain
the solution. It is called the \emph{characteristic curve} and we summarize its equations in 
conventional form as
\begin{eqnarray}
D_\sigma x_i &=& F_{p_j} \label{s9e12} \\
D_\sigma z   &=& \sum_{j=1}^n p_j F_{p_j} \label{s9e13} \\
D_\sigma p_i &=& -F_{x_i} - F_z z_{x_i} \label{s9e14}
\end{eqnarray}
The treatment in this section follows Evan's book \cite{evans}.

\section{Hamilton-Jacobi equations - I}\label{s10}
It is an equation of the form
\begin{equation}\label{s10e1}
u_t + H(x, Du) = 0,
\end{equation}
where $x \in \mathbb{R}^n$, $t \in \mathbb{R}$ and $H: \mathbb{R}^{2n} \rightarrow \mathbb{R}$.
Let us define $y = (x, t) \in \mathbb{R}^{n+1}$ and $q = (u, u_t) \in \mathbb{R}^{n+1}$ so that
we can write \eqref{s10e1} as
\begin{equation}\label{s10e2}
q_{n+1} + H(x, p) = 0.
\end{equation}
Let us write this equation as $F = q_{n+1} + H(x, p) = 0$ so that
\[
F_{q_j} = H_{q_j}
\]
But $q_j = p_j$ for $1 \le j \ne n$ so that
\begin{equation}\label{s10e3}
F_{p_j} = H_{p_j}
\end{equation}
and
\begin{equation}\label{s10e4}
F_{q_{n+1}} = 1.
\end{equation}
Similarly,
\begin{eqnarray}
F_{x_i} &=& H_{x_i} \label{s10e5} \\
F_{t} &=& 0 \label{s10e6}
\end{eqnarray}
Since $F$ does not explicitly depend on $u$ (but only on its derivatives $u_t$ and $Du$) we have
\begin{equation}\label{s10e7}
F_z = 0.
\end{equation}
Let $y(\sigma)$ be the characteristic curve. Using equations \eqref{s9e12} to \eqref{s9e14} and
the partial derivatives evaluated in \eqref{s10e3} to \eqref{s10e7}, we get
\begin{eqnarray}
D_\sigma x_i &=& H_{p_i} \label{s10e8} \\
D_\sigma t &=& 1 \label{s10e9} \\
D_\sigma z &=& \sum_{i=1}^{n+1} q_j F_{q_j} = p \cdot D_p H + q_{n+1}\label{s10e10} \\
D_\sigma p_i &=& -H_{x_i} \label{s10e11} \\
D_\sigma q_{n+1} &=& 0 \label{s10e12}
\end{eqnarray}
From equation \eqref{s10e9} it is clear that we can as well identify the variable $t$ as the
parameter $\sigma$ so that equations \eqref{s10e8} and \eqref{s10e10} can be written in the more
familiar form
\begin{eqnarray}
\dot{x} &=& D_p H \label{s10e13} \\
\dot{p} &=& -D_x H \label{s10e14}
\end{eqnarray}
of the canonical equations of classical mechanics, also called the Hamilton equations. For future
reference, we will write \eqref{s10e10} as
\begin{equation}\label{s10e15}
\dot{u} = p \cdot D_p H - H,
\end{equation}
where we have used \eqref{s10e2} to write $q_{n+1} = -H$ and \eqref{s9e1} to write $z(\sigma) = 
u(y(\sigma)) = u(x(\sigma), t(\sigma))$.

\section{Calculus of variations}\label{s11}
Let $w: [0, T] \rightarrow \mathbb{R}^n$ be vector-valued functions. Define a class
\begin{equation}\label{s11e1}
\mathcal{A} = \left\{w \in C^2([0, T] \rightarrow \mathbb{R}^n : w(0) = a, w(T) = b\right\},
\end{equation}
where $a$ and $b$ are fixed points in $\mathbb{R}^n$. Let $L: \mathbb{R}^n \times \mathbb{R}^n 
\rightarrow \mathbb{R}$ be a given function of $w$ and $\dot{w}$, generically called the lagrangian.
Let
\begin{equation}\label{s11e2}
J(w) = \int_0^T L(w(s), \dot{w}(s))ds,
\end{equation}
where $\dot{w} = dw/ds$. The problem of calculus of variations is to find a $u \in 
\mathbb{R}^n$ such that
\begin{equation}\label{s11e3}
J(u) \le J(w) \;\forall w \in \mathcal{A}.
\end{equation}
Let us assume that such a $u$ exists and that we can express a neighboring `point' $w$ as
\begin{equation}\label{s11e4}
w = u + \tau v,
\end{equation}
where $\tau \in \mathbb{R}$ and $v \in C^2([0, T] \rightarrow \mathbb{R}^n)$ such that $v(0) = v(T)
= 0$. Therefore,
\begin{equation}\label{s11e5}
J(u) \le J(u + \tau v), \forall v \in C^2([0, T] \rightarrow \mathbb{R}^n).
\end{equation}
For a fixed $v$, $J$ is a function of a real variable $\tau$ that has a minimum at $\tau = 0$. Let
is call that function $h$. That is,
\begin{equation}\label{s11e6}
h(\tau) = J(u + \tau v) = \int_0^T L(u + \tau v, \dot{u} + \tau\dot{v})ds.
\end{equation}
Derivative of this function with respect to $\tau$ is
\[
h^\prime(\tau) = \int_0^T(v\cdot D_u L + \dot{v}\cdot D_{\dot{u}}L)ds.
\]
so that
\[
h^\prime(\tau) = 
\int_0^T v\cdot D_uL ds + D_{\dot{u}}L v\Big|_0^T - \int_0^T v\cdot \frac{d}{ds}D_{\dot{u}}Lds.
\]
Since $v(0) = v(T) = 0$, the second term on the right hand side vanishes and 
\[
h^\prime(\tau) = \int_0^Tv\cdot\left(D_uL - \frac{d}{ds}D_{\dot{u}}L\right)ds.
\]
At the extremum of $h$, $h^\prime(\tau) = 0$ so that
\[
\int_0^Tv\cdot\left(D_uL - \frac{d}{ds}D_{\dot{u}}L\right)ds = 0
\]
for all $v$, which is possible only if 
\begin{equation}\label{s11e7}
D_uL - \frac{d}{ds}D_{\dot{u}}L = 0.
\end{equation}
Equation \eqref{s11e7} is called the Euler-Lagrange equation. A member $u$ of $\mathcal{A}$ which
saitsfies \eqref{s11e7} is called its \emph{critical point}. Every minimum of the functional $J$ is
its critical point but the converse is not true.

\section{Hamilton equations}\label{s12}
We choose different symbols for the arguments of the lagrangian. We write it as $L(x,v,t)$, where
$x \in \mathbb{R}$, $v = \dot{x}$ and $t \in \mathbb{R}$. The Euler-Lagrange equations of 
\eqref{s11e7} now appear as
\begin{equation}\label{s12e1}
D_x L - \frac{d}{dt}\left(D_v L\right) = 0.
\end{equation}
The canonical momentum is defined as
\begin{equation}\label{s12e2}
p = D_{v}L
\end{equation}
and the hamiltonian associated with $L$ is defined as
\begin{equation}\label{s12e3}
H(x, p) = p \cdot v - L(x, v).
\end{equation}
Now,
\[
D_x H = \sum_{i=1}^n \pd{H}{x_i} = \sum_{i=1}^n \frac{\partial}{\partial x_i} (v_jp_j) - 
\sum_{i=1}^n \pd{L}{x_j}\pd{x_j}{x_i} - \sum_{i=1}^n\pd{L}{v_j}\pd{v_j}{x_i}.
\]
Using \eqref{s12e2} in the last term on the right hand side of the above equation,
\[
D_x H = \sum_{i=1}^n \frac{\partial}{\partial x_i} \left(\sum_{j=1}^nv_jp_j\right) - 
\sum_{i=1}^n \sum_{j=1}^n \pd{L}{x_j}\pd{x_j}{x_i} - \sum_{i=1}^n \sum_{j=1}^n p_j\pd{v_j}{x_i}.
\]
Since $x_j$ and $p_j$ are independent variables,
\[
D_x H = \sum_{i=1}^n \frac{\partial}{\partial x_i} \left(\sum_{j=1}^nv_jp_j\right) - 
\sum_{i=1}^n \sum_{j=1}^n \pd{L}{x_j}\delta_{ij} - 
\sum_{i=1}^n \frac{\partial}{\partial x_i} \left(\sum_{j=1}^nv_jp_j\right)
\]
or
\[
D_x H = -\sum_{j=1}^n \pd{L}{x_j} = -D_x L.
\]
From equations \eqref{s12e1} and \eqref{s12e2} we get
\begin{equation}\label{s12e4}
\dot{p} = -D_x H.
\end{equation}
Similarly,
\[
D_p H = \sum_{i,j=1}^n \frac{\partial}{\partial p_i} (v_jp_j) = \sum_{i=1}^n v_i.
\]
By definition, $v = \dot{x}$ so that
\begin{equation}\label{s12e5}
\dot{x} = D_p H.
\end{equation}
Equations \eqref{s12e4} and \eqref{s12e5} are the same as the pair \eqref{s10e13} and \eqref{s10e14}.
They are Hamilton's equations.

\section{Hamilton-Jacobi equations - II}\label{s13}
We recall that the Hamilton-Jacobi equation is of the form \eqref{s10e1},
\[
u_t + H(x, Du) = 0,
\]
where $x \in \mathbb{R}^n$, $t \in \mathbb{R}$ and $H: \mathbb{R}^{2n} \rightarrow \mathbb{R}$. Let 
us now consider an equation of the form,
\begin{equation}\label{s13e1}
u_t + H(Du) = 0,
\end{equation}
where $u: \mathbb{R}^n \times [0, \infty) \rightarrow \mathbb{R}$ and $H: \mathbb{R}^n \rightarrow
\mathbb{R}$ is an \emph{arbitrary} function, not necessarily a classical hamiltonian. Further, we 
require that $u(x,t = 0) = g(x)$.

We know that in classical mechanics, the function $u$ is the action
\[
u = \int_0^t L(x(s), v(s))ds,
\]
where $v = \dot{x}$ and the function $H(x, p) = p\cdot v - L(x, v)$. Can we use this analogy to
get a solution of \eqref{s13e1}?

\subsection{Fenchel transform}
To get a solution that looks like the classical action, we need something that is analogous to the 
Legendre transformation. Since $H$ is now a function of only $Du$, that is $H: \mathbb{R}^n 
\rightarrow \mathbb{R}$, we will also consider an $L: \mathbb{R}^n \rightarrow \mathbb{R}$.
\begin{defn}\label{s13d1}
If $L: \mathbb{R}^n \rightarrow \mathbb{R}$ is such that
\begin{itemize}
\item The mapping $v \mapsto L(v)$ is convex and
\item $L$ is coercive, that is
\[
\lim_{|v| \rightarrow \infty} \frac{L(v)}{|v|} = +\infty.
\]
\end{itemize}
then the Fenchel transform of $L$ is 
\[
L^\ast(p) = \sup_{v \in \mathbb{R}^n}\{p\cdot v - L(v)\},
\]
where $p \in \mathbb{R}^n$.
\end{defn}
\begin{rem}
For a given $p$, $L^\ast(p)$ exists because of the coercivity of $L$. For as $|v|$ tends to $\infty$,
$L(v)$ will rise faster than $|v|$ and terms of the form $p\cdot v - L(v)$ will start becoming 
negative. However, for small enough $|v|$, they might be positive. The set $\{p\cdot v - L(v)\}$ is
a subset of $\mathbb{R}$ that is bounded above. Therefore, it has a supremum.
\end{rem}

\begin{rem}
Let $v^\ast$ be that value of $v$ such that $p \cdot v^\ast - L(v^\ast)$ is the supremum of the
set 
\[
\sup_{v \in \mathbb{R}^n}\{p\cdot v - L(v)\}.
\]
In other words, the mapping $f(v) = p \cdot v - L(v)$ has a maximum at $v^\ast$. That can happen only
when $D_v f = p - D_v L = 0$ at $v = v^\ast$ or that $p = D_v L|_{v^\ast}$. (We assumed that L is
indeed differentiable at $v = v^\ast$.) This means that we can write $v^\ast = v(p)$. Therefore the 
supremum $p \cdot v^\ast - L(v^\ast)$ is indeed a function of $p$ alone. Therefore, $L^\ast$ is 
indeed like the hamiltonian defined by Legendre transformation.
\end{rem}

Therefore, we define
\begin{defn}\label{s13d2}
The hamiltonian function corresponding to the lagrangian defined in \ref{s13d1} is $H(p) = L^\ast(p)$.
\end{defn}

We will first prove that
\begin{prop}\label{s13l1}
If $A$ and $B$ are two non-empty subsets of $\mathbb{R}$ then $\sup(A + B) \le \sup(A) + \sup(B)$,
where the set $A + B = \{a + b: a \in A, b \in B\}$.
\end{prop}
\begin{proof}
Since $a = a + b - b$, $a \le \sup(A + B) - b$. Therefore, $\sup(A) \le \sup(A + B) - b$ or $b \le
\sup(A + B) - \sup(A)$ or $\sup(B) \le \sup(A + B) - \sup(A)$.
\end{proof}

We will now show that
\begin{prop}\label{s13l2}
$H$ is a convex function of $p$.
\end{prop}
\begin{proof}
Consider $H(\lambda p_1 + (1 - \lambda)p_2) = \sup_{v \in \mathbb{R}^n}\{(\lambda p_1 + 
(1 - \lambda)p_2)\cdot v - L(v)\}$ so that, using proposition \ref{s13l1}, 
\[
H(\lambda p_1 + (1 - \lambda)p_2) \le 
\lambda\sup_{v \in \mathbb{R}^n}\{p_1\cdot v - L(v)\} + 
(1-\lambda)\sup_{v \in \mathbb{R}^n}\{p_2\cdot v - L(v)\}.
\]
That is $H(\lambda p_1 + (1 - \lambda)p_2) \le \lambda H(p_1) + (1 - \lambda)H(p_2)$.
\end{proof}

\begin{prop}\label{s13l3}
$H$ is coercive.
\end{prop}
\begin{proof}
Since $H(p) = \sup_{v \in \mathbb{R}^n}\{v \cdot p - L(v)\}$, for a particular choice of $v = \lambda
p/|p|$, we have
\[
H(p) \ge \lambda p \cdot \frac{p}{|p|} - L\left(\lambda\frac{p}{|p|}\right) = \lambda |p| 
- L\left(\lambda\frac{p}{|p|}\right).
\]
Therefore,
\[
\frac{H(p)}{|p|} \ge \lambda - \frac{1}{|p|}L\left(\lambda\frac{p}{|p|}\right).
\]
Because $L$ is coercive, we have
\[
\frac{H(p)}{|p|} \ge \lambda.
\]
But this is true for any $\lambda > 0$. Therefore, even $H$ is coercive.
\end{proof}

\begin{prop}\label{s13l4}
If $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is convex then for every $x \in \mathbb{R}^n$, there
exists an $r \in \mathbb{R}^n$ such that $f(y) \ge f(x) + r\cdot(y - x)$, for all $y \in 
\mathbb{R}^n$.
\end{prop}
\begin{proof}
Since $f$ is convex
\[
f(\lambda y + (1 - \lambda)x) \le \lambda f(y) + (1 - \lambda) f(x)
\]
for any $x, y$ in the domain of $f$ and $\lambda \in [0, 1]$. By a simple rearrangement of terms,
\[
\frac{f(x + \lambda(y - x)) - f(x)}{\lambda} \le f(y) - f(x).
\]
When we take the limit of the above equation as $\lambda \rightarrow 0$, the left hand side is 
just the directional derivative of $f$ in the direction $y - x$, that is
\[
Df\cdot(y-x) \le f(y) - f(x).
\]
\end{proof}

Finally,
\begin{prop}\label{s13l5}
$L = H^\ast$, where
\[
H^\ast(v) = \sup_{p \in \mathbb{R}^n}\{v \cdot p - H(p)\}.
\]
\end{prop}
\begin{proof}
Since $H(p) = L^\ast(v)$, $H(p) \ge p\cdot v - L(v)$ for any $p, v$ so that $L(v) \ge v \cdot p - 
H(p)$ and hence 
\begin{equation}\label{s13e2}
L(v) \ge H^\ast(v).
\end{equation}

Now,
\[
H^\ast(v) = \sup_{p \in \mathbb{R}^n}\{v \cdot p - H(p)\}
= \sup_{p \in \mathbb{R}^n}\left\{v\cdot p - \sup_{u\in\mathbb{R}^n}\{p\cdot u - L(u)\}\right\}
\]
so that
\[
H^\ast(v) = \sup_{p \in \mathbb{R}^n}\inf_{u \in \mathbb{R}^n}\left\{p\cdot(v-u) + L(u)\right\}.
\]
Using proposition \ref{s13l4},
\begin{equation}\label{s13e3}
H^\ast(v) \le \sup_{p \in \mathbb{R}^n}\{ L(v) - L(u) + L(u)\} = L(v),
\end{equation}
the latter equality follows from the fact that what remains does not depend on $p$. The proposition 
follows from equations \eqref{s13e2} and \eqref{s13e3}.
\end{proof}

Thus, the Fenchel transform works exactly like the Legendre transform. The fact that $H = L^\ast$ 
and $L = H^\ast$ means the equivalence of the following equations,
\begin{eqnarray}
p \cdot v &=& H(p) + L(v) \label{s13e4} \\
p &=& D_v L \label{s13e5} \\
v &=& D_p H \label{s13e6},
\end{eqnarray}
where the last equation follows if we write remarks just after definition \ref{s13d1} customized 
for $H^\ast$.

\subsection{Hopf-Lax formula}
Recall that we are interested in solving \eqref{s13e1}. Its characteristic equations, which follow
immediately from \eqref{s10e13}, \eqref{s10e14} and \eqref{s10e15}, are
\begin{eqnarray}
\dot{x} &=& D_p H \label{s13e7} \\
\dot{p} &=& 0 \label{s13e8} \\
\dot{u} &=& p\cdot D_p H - H. \label{s13e9}
\end{eqnarray}
Using equation \eqref{s13e6} in equations \eqref{s13e7} and \eqref{s13e8}, the above three equations
become
\begin{eqnarray}
\dot{x} &=& v \label{s13e10} \\
\dot{p} &=& 0 \label{s13e11} \\
\dot{u} &=& p\cdot v - H \label{s13e12}
\end{eqnarray}
Using \eqref{s13e4} in \eqref{s13e12},
\begin{equation}\label{s13e13}
\dot{u} = L(v) = L(\dot{x})
\end{equation}
where the last equation follows from \eqref{s13e10}. The solution of this equation is
\begin{equation}\label{s13e14}
u(x, t) = \int_0^t L(\dot{x})ds + g(x).
\end{equation}
We got a solution similar to the one in classical mechanics, $u$ is the analog of classical action.
But we also know that $x, \dot{x}$ are the ones that minimize the action. Therefore, we now explore 
the solutions that are
\begin{equation}\label{s13e15}
u(x, t) := \inf\left\{\int_0^t L(\dot{w}(s))ds + g(w(0)) : w(t) = x(t)\right\}.
\end{equation}
We next show that the above equation has a simpler form.
\begin{thm}[Hopf-Lax formula]\label{s13t1}
If $x \in \mathbb{R}^n$ and $t > 0$ then the solution of the minimization problem of \eqref{s13e15}
is
\[
u(x, t) = \min_{y \in \mathbb{R}^n}\left\{tL\left(\frac{x - y}{t}\right) + g(y)\right\}
\]
if $g$ is a Lipshitz function.
\end{thm}
\begin{proof}
For a fixed $y \in \mathbb{R}^n$, let 
\[
w(s) := y + \frac{s}{t}(x - y),
\]
so that $w(t) = x, w(0) = y$ and $\dot{w} = (x - y)/t$ and 
\[
\int_0^t L(\dot{w}(s))ds = \int_0^t L\left(\frac{x - y}{t}\right)ds = tL\left(\frac{x - y}{t}\right).
\]
From \eqref{s13e15},
\begin{equation}\label{s13e16}
u(x, t) \le tL\left(\frac{x - y}{t}\right) + g(y).
\end{equation}

Now, since $L$ is a convex function,by Jensen's inequality,
\[
L\left(\frac{1}{t}\int_0^t \dot{w}(s)ds\right) \le \frac{1}{t}\int_0^t L(\dot{w}(s))ds.
\]
If $y = w(0)$,
\[
L\left(\frac{1}{t}\int_0^t \dot{w}(s)ds\right) + g(y) \le \frac{1}{t}\int_0^t L(\dot{w}(s))ds + g(y).
\]
For our choice of $w$, the left hand side is
\[
tL\left(\frac{x - y}{t}\right) + g(y) \le \frac{1}{t}\int_0^t L(\dot{w}(s))ds + g(y).
\]
From \eqref{s13e15},
\begin{equation}\label{s13e17}
tL\left(\frac{x - y}{t}\right) + g(y) \le u(x, t).
\end{equation}
From equations \eqref{s13e15} and \eqref{s13e16} we get
\[
u(x, t) = \inf_{y \in \mathbb{R}^n}\left\{tL\left(\frac{x - y}{t}\right) + g(y)\right\}.
\]

\noindent Since $g$ is a Lipshitz function, $g(y) \le \alpha |y|$. Since $L$ is coersive $L(t^{-1}(x - y))$
rises faster than $|y|$ for fixed $x, t$. Therefore, the infimum is the minimum.
\end{proof}

\noindent We next prove the following functional identity.
\begin{prop}\label{s13p6}
For each $x \in \mathbb{R}^n$ and $0 \le s < t$, we have
\[
u(x, t) = \min_{y \in \mathbb{R}^n}\left\{(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s)\right\}
\]
where $u$ is defined by the Hopf-Lax formula.
\end{prop}
\begin{proof}
Fix $y \in \mathbb{R}^n$ and let $s \in [0, t)$. Select $z \in \mathbb{R}^n$ for which Hopf-Lax
formula attains a minimum, that is
\begin{equation}\label{s13e18}
u(y, s) = sL\left(\frac{y - z}{s}\right) + g(z).
\end{equation}
For any $x, y, t$,
\[
\frac{x - z}{t} = \left(1 - \frac{s}{t}\right)\frac{x - y}{t - s} + \frac{s}{t}\frac{y - z}{s}
\]
so that convexity of $L$ gives,
\[
L\left(\frac{x - z}{t}\right) \le \left(1 - \frac{s}{t}\right)L\left(\frac{x - y}{t - s}\right)
+ \frac{s}{t}L\left(\frac{y - z}{s}\right)
\]
or
\[
tL\left(\frac{x - z}{t}\right) \le (t - s)L\left(\frac{x - y}{t - s}\right) + 
sL\left(\frac{y - z}{s}\right)
\]
Adding $g(z)$ to either sides,
\[
tL\left(\frac{x - z}{t}\right) + g(z) \le (t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s),
\]
where we used \eqref{s13e17} to get the seconf term on the right hand side. From theorem \ref{s13t1},
\[
u(x, t) \le tL\left(\frac{x - z}{t}\right) + g(z)
\]
for all $z \in \mathbb{R}^n$ so that
\[
u(x, t) \le  (t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s).
\]
Since this is true for all $y \in \mathbb{R}^n$, 
\begin{equation}\label{s13e19}
u(x, t) \le 
\min_{y \in \mathbb{R}^n}\left\{(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s)\right\}.
\end{equation}

\noindent Now choose $w$ such that
\[
u(x, t) = tL\left(\frac{x - w}{t}\right) + g(w).
\]
If 
\begin{equation}\label{s13e20}
y = \frac{s}{t}x + \left(1 - \frac{s}{t}\right)w
\end{equation}
so that
\[
x - y =  \left(1 - \frac{s}{t}\right)(x - w)
\]
or
\begin{equation}\label{s13e21}
\frac{x - y}{t - s} = \frac{x - w}{t}.
\end{equation}
Likewise, from \eqref{s13e20},
\[
y - w = \frac{s}{t}(x - w)
\]
so that
\[
\frac{y - w}{s} = \frac{x - w}{t}
\]
and therefore from equation \eqref{s13e21}
\begin{equation}\label{s13e22}
\frac{x - y}{t - s} = \frac{x - w}{t} = \frac{y - w}{s}.
\end{equation}
Therefore,
\[
(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s) = 
(t - s)L\left(\frac{x - w}{t}\right) + u(y, s)
\]
or
\[
(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s) = tL\left(\frac{x - w}{t}\right) + u(y, s)
- sL\left(\frac{x - w}{t}\right).
\]
Using \eqref{s13e22}
\[
(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s) = tL\left(\frac{x - w}{t}\right) + u(y, s)
- sL\left(\frac{y - w}{s}\right)
\]
But,
\[
u(y, s) \le sL\left(\frac{y - w}{s}\right) + g(w)
\]
so that
\[
(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s) \le tL\left(\frac{x - w}{t}\right) + g(w).
\]
But from our choice of $w$, the right hand side is $u(x, t)$ so that
\begin{equation}\label{s13e23}
(t - s)L\left(\frac{x - y}{t - s}\right) + u(y, s) \le u(x, t).
\end{equation}
The proposition follows immediately from equations \eqref{s13e19} and \eqref{s13e23}.
\end{proof}

\begin{prop}\label{s13p7}
The function $u$ defined by the Hopf-Lax formula in theorem \ref{s13t1} is Lipshitz continuous.
\end{prop}
\begin{proof}
Recall the Hopf-Lax formula,
\[
u(x, t) = \min_{y \in \mathbb{R}^n}\left\{tL\left(\frac{x - y}{t}\right) + g(y)\right\},
\]
where the function $g$ is assumed to be Lipshitz continuous. Fix $x, x_1, t$ and let $y$ be
such that
\[
u(x, t) = tL\left(\frac{x - y}{t}\right) + g(y)
\]
so that
\[
u(x_1, t) - u(x, t) = \min_{z \in \mathbb{R}^n}\left\{tL\left(\frac{x_1 - z}{t}\right) + g(z)\right\} 
- tL\left(\frac{x - y}{t}\right) - g(y).
\]
Choose $z = x_1 - x + y$ so that
\[
u(x_1, t) - u(x, t) \le tL\left(\frac{x_1 - (x_1 - x + y)}{t}\right) + g(x_1 - x + y) -
tL\left(\frac{x - y}{t}\right) + g(y)
\]
or
\[
u(x_1, t) - u(x, t) \le g(x_1 - x + y) - g(y)
\]
Since $g$ is Lipshitz continuous, there exists $\alpha \in \mathbb{R}$ such that
\[
\sup_{x, y \in \mathbb{R}^n}\frac{|g(x) - g(y)|}{x - y} \le \alpha.
\]
Therefore, $|g(x_1 - x + y) - g(y)| \le \alpha|x_1 - x + y - y| = \alpha|x_1 - x|$ and hence
\[
u(x_1, t) - u(x, t) \le \alpha|x_1 - x|.
\]
Equivalently,
\begin{equation}\label{s13e24}
|u(x, t) - u(x_1, t)| \le \alpha |x - x_1|.
\end{equation}
This means that $u$ is Lipshitz continuous in $x$. Let us now consider continuity in $t$. If we 
choose $y = x$ in Hopf-Lax formula we get
\begin{equation}\label{s13e25}
u(x, t) \le tL(0) + g(x).
\end{equation}
Since $g$ is Lipshitz continuous, $|g(x) - g(y)| \le \alpha |x - y|$ for all $x, y \in \mathbb{R}^n$.
Therefore,
\[
-\alpha|x - y| \le g(x) - g(y) \le \alpha|x - y|
\]
or $g(y) \ge g(x) - \alpha|x - y|$ so that
\[
u(x, t) = \min_{y \in \mathbb{R}^n}\left\{tL\left(\frac{x - y}{t}\right) + g(y)\right\}
\]
implies that
\[
u(x, t) \ge \min_{y \in \mathbb{R}^n}
\left\{tL\left(\frac{x - y}{t}\right) + g(x) - \alpha|x - y|\right\}
\]
or
\[
u(x, t) \ge g(x) + \min_{y \in \mathbb{R}^n}
\left\{tL\left(\frac{x - y}{t}\right) - \alpha|x - y|\right\}
\]
If we let $zt = x - y$ then the above equation becomes
\[
u(x, t) \ge g(x) + t \min_{z \in \mathbb{R}^n}\{L(z) - \alpha |z|\}
\]
which is same as
\[
u(x, t) \ge g(x) - t \max_{z \in \mathbb{R}^n}\{\alpha |z| - L(z)\}
\]
or
\[
u(x, t) \ge g(x) - t \max_{w \in B(0, \alpha)}\max{z \in \mathbb{R}^n}\{w \cdot z - L(z)\}.
\]
or
\begin{equation}\label{s13e26}
u(x, t) \ge g(x) - t \max_{w \in B(0, \alpha)}H.
\end{equation}

We can write equations \eqref{s13e25} and \eqref{s13e26} as
\begin{eqnarray*}
u(x, t) - g(x) &\le& tL(0) \\
g(x) - u(x, t) &\le& t\max_{w \in B(0, \alpha)}H
\end{eqnarray*}
or that
\begin{equation}\label{s13e27}
|u(x, t) - g(x)| \le Ct,
\end{equation}
where 
\[
C := \max\left(L(0), \max_{w \in B(0, \alpha)}H\right).
\]
This means that $u$ is Lipshitz continuous at $t = 0$. From \eqref{s13e25} we also have
\[
u(x, t) - u(x, s) \le (t - s)L(0) \Rightarrow \frac{|u(x, t) - u(x, s)|}{|t - s|} \le |L(0)|
\]
which tells us that $L$ is Lipshitz continuous for all $t$.
\end{proof}

Rademacher's theorem states that a Lipshitz continuous function is differentiable almost everywhere
in its domain. We will now show that wherever it is differentiable, the function $u$ defined by the
Hopf-Lax formula is a solution of Hamilton-Jacobi equations.

\section{Green's theorem}\label{s14}
It is related to Stokes theorem
\begin{equation}\label{s14e1}
\iint (\nabla \times \vec{F})\cdot \hat{n}da = \oint\vec{F} \cdot d\vec{l}.
\end{equation}
If $\vec{F} = (L, M, 0)$ and if $L$ and $M$ are functions of $x, y$ alone then $\nabla\times\vec{F} = 
\hat{e}_z (M_x - L_y)$. The line integral on the right hand is then a curve entirely in the $xy$-plane
and $d\vec{l} = \hat{e}_xdx + \hat{e}_ydy$ so that
\begin{equation}\label{s14e2}
\iint \left(\pd{M}{x} - \pd{L}{y}\right)dxdy = \oint \left(Ldx + Mdy\right).
\end{equation}

Gauss theorem in three dimensions reads
\begin{equation}\label{s14e3}
\iiint \nabla\cdot\vec{F}dx_1dx_2dx_3 = \iint (\vec{F}\cdot\hat{n})da.
\end{equation}
We can extend it to $n$ dimensions as
\begin{equation}\label{s14e4}
\int \sum_{i=1}^n\pd{F_i}{x_i}dV = \int \left(\sum_{i=1}^n F_in_i\right) da.
\end{equation}
If $F = (0, \ldots, u, \ldots, 0)$, where only the $i$th component is non-zero then
\begin{equation}\label{s14e5}
\int \pd{u}{x_i}dV = \int u n_i da.
\end{equation}
where $n_i$ is the $i$th component of the normal. Applying the above equation to $uv$ instead,
\[
\int \left(\pd{u}{x_i}v + u\pd{v}{x_i}\right)dV = \int uv n_i dA
\]
or 
\begin{equation}\label{s14e6}
\int u\pd{v}{x_i} dV = -\int v \pd{u}{x_i} dV + \int (uv) n_i dA.
\end{equation}

\section{Conservation laws}\label{s15}
Consider the initial value problem
\begin{eqnarray}
u_t + (F(u))_x &=& 0 \text{ in } \mathbb{R} \times (0, \infty) \label{s15e1} \\
u &=& g \text{ on } \mathbb{R} \times {t=0} \nonumber,
\end{eqnarray}
where $F, g: \mathbb{R} \rightarrow \mathbb{R}$ are given functions and $u:\mathbb{R} \times 
\mathbb{R} \rightarrow \mathbb{R}$ is an unknown.

\begin{defn}\label{s15d1}
A function $v: \mathbb{R} \times [0, \infty) \rightarrow \mathbb{R}$ is called a test function if it
is a smooth function with compact support.
\end{defn}
Multiply the PDE by a smooth function $v$ and integrate it over the domain
of $u$ to get
\[
\int_0^\infty \int_{-\infty}^\infty \left(vu_t + v(F(u))_x\right)dxdt = 0.
\]
Applying \eqref{s14e6} to each term separately,
\[
-\int_0^\infty \int_{-\infty}^\infty v_t u dxdt + \int_{-\infty}^\infty \left[vu\right]_0^\infty dx
-\int_0^\infty \int_{-\infty}^\infty v_x F(u) dxdt + 
\int_0^\infty \left[vF(u)\right]_{-\infty}^\infty = 0.
\]
Since $v$ has compact support $[vF(u)]_{-\infty}^\infty = 0$ and $[uv]_0^\infty = -u(x, 0)v(x, 0)
= -g(x)v(x, 0)$ so that
\[
-\int_0^\infty \int_{-\infty}^\infty v_t u dxdt - \int_{-\infty}^\infty  g(x)v(x, 0) dx - 
\int_0^\infty \int_{-\infty}^\infty v_x F(u) dxdt = 0
\]
or
\begin{equation}\label{s15e2}
-\int_0^\infty \int_{-\infty}^\infty \left(v_t u + v_x F(u)\right)dxdt = 
\int_{-\infty}^\infty  g(x)v(x, 0) dx.
\end{equation}
\begin{defn}\label{s15d2}
A function $u\mathbb{R} \times \mathbb{R} \rightarrow \mathbb{R}$ is an integral solution of the
initial value problem \eqref{s15e1} if it satisfies equation \eqref{s14e2} for all test functions 
$v$.
\end{defn}

We will now show how discontinuities propagate. Consider an open region $V \subset \mathbb{R} 
\times (0, \infty)$ and let $C$ be a curve in it dividing the region $V_l$ on it left and $V_r$
on its right. Let us now consider a test function $v$ which is compact in $V_l$ and $u$ is an
integral solution. Then from equation \eqref{s15e2}, we get
\[
-\int_0^\infty \int_{-\infty}^\infty \left(v_t u + v_x F(u)\right)dxdt = 0
\]
Applying the `integration by parts' \eqref{s14e6} to the left hand side,
\begin{equation}\label{s15e3}
\iint_{V_l}v\left(u_t + (F(u))_x\right)dxdt = 0
\end{equation}
from which we conclude that in $V_l$,
\begin{equation}\label{s15e4}
u_t + (F(u))_x = 0.
\end{equation}
We will reach the same conclusion for $V_r$ if we consider a test function $v$ with a compact
support in $V_r$.

Let us now consider a test function which is compact in $V$ but it may not vanish on $C$. Once again,
equation \eqref{s15e2} gives
\begin{equation}\label{s15e5}
-\iint_V \left(v_t u + v_x F(u)\right)dxdt = 0.
\end{equation}
Writing it as a sum of two integrals,
\begin{equation}\label{s15e6}
-\iint_{V_l} \left(v_t u + v_x F(u)\right)dxdt -\iint_{V_r} \left(v_t u + v_x F(u)\right)dxdt = 0.
\end{equation}
Applying `integration by parts' \eqref{s14e6} to the first of these integrals we get
\[
\iint_{V_l} \left(v u_t + v (F(u))_x\right)dxdt - \int_{C_1} v(un_t + F(u)n_x) dl - 
\int_C v(un_t + F(u)n_x) dl,
\]
where the curve $C_1 = \partial V_l - C$ and $(n_t, n_x)$ is the `outward normal' to $\partial V_l$. 
The second of these integrals vanishes because of compact support of $v$ in $V$. Therefore,
\begin{equation}\label{s15e7}
\iint_{V_l} \left(v_t u + v_x F(u)\right)dxdt = \iint_{V_l} v\left(u_t + (F(u))_x\right)dxdt
- \int_C v(un_t + F(u)n_x) dl.
\end{equation}
Repeating this exercise for $V_r$ we get
\begin{equation}\label{s15e8}
\iint_{V_r} \left(v_t u + v_x F(u)\right)dxdt = \iint_{V_r} v\left(u_t + (F(u))_x\right)dxdt
- \int_C v(un_t^\prime + F(u)n_x^\prime) dl,
\end{equation}
where $(n_t^\prime, n_x^\prime)$ are the `outward normal' to $\partial V_r$. Equation \eqref{s15e4}
is true in both $V_l$ and $V_r$ so that equations \eqref{s15e5} and \eqref{s15e6} become
\begin{eqnarray}
\iint_{V_l}\left(v_t u + v_x F(u)\right)dxdt&=&-\int_C v(un_t + F(u)n_x) dl \label{s15e9} \\
\iint_{V_r}\left(v_t u + v_x F(u)\right)dxdt&=&-\int_C v(un_t^\prime + F(u)n_x^\prime)dl 
\label{s15e10} 
\end{eqnarray}
Adding equations \eqref{s15e9} and \eqref{s15e10} and using equation \eqref{s15e6} we get
\[
0 = \int_C v(un_t + F(u)n_x) dl + \int_C v(un_t^\prime + F(u)n_x^\prime)dl.
\]
Now note that $(n_t, n_x) = (-n_t, -n_x)$ so that
\begin{equation}\label{s15e11}
\int_C v\left([[u_t]]n_t + [[F(u)]]n_x\right) dl = 0
\end{equation}
where $[[\cdot]]$ is the jump in the argument. Since this is true for all test functions $v$ with
compact support in $V$, we get
\begin{equation}\label{s15e12}
[[u_t]]n_t + [[F(u)]]n_x = 0.
\end{equation}
If the parametric equations of $C$ are $(x, t) = (s(t), t)$ then a tangent to $C$ is $(\dot{s}, 1)$
and therefore a normal is $(-1, \dot{s})$ so that a unit normal is 
\begin{equation}\label{s15e13}
(n_x, n_t) = \left(-\frac{1}{\sqrt{1 + \dot{s}^2}}, \frac{\dot{s}}{\sqrt{1 + \dot{s}^2}}\right)
\end{equation}
From equations \eqref{s15e12} and \eqref{s15e13} we get the \emph{Rankine-Hugonoit condition}
\begin{equation}\label{s15e14}
[[u_t]] = \dot{s}[[F(u)]].
\end{equation}

\section{Laplace equation}\label{s16}
If $u: \mathbb{R}^n \rightarrow \mathbb{R}$ then Laplace equation is
\begin{equation}\label{s16e1}
\sum_{i=1}^n\frac{\partial^2 u}{\partial x_i^2} = 0.
\end{equation}

The following proposition is problem 2 in chapter 2 of Evan's book.
\begin{prop}\label{s16p1}
Laplace equation is invariant under rotation.
\end{prop}
\begin{proof}
We have to show that if $A$ is an orthogonal matrix then $\Delta u(x) = 0 \Rightarrow \Delta u(y) = 0$ 
where $y = Ax$. When $A$ is an orthogonal matrix then $AA^T = I$ or
\begin{equation}\label{s16e2}
\sum_{j = 1}^n a_{ij}a_{kj} = \delta_{ik}.
\end{equation}
Further $y = Ax$ means that
\begin{equation}\label{s16e3}
y_i = \sum_{j = 1}^n a_{ij} x_j.
\end{equation}
Therefore,
\[
\pd{u}{x_i} = \sum_{j=1}^n\pd{u}{y_j}\pd{y_j}{x_i} = \sum_{j = 1}^n\pd{u}{y_j}\sum_{k=1}^n a_{jk}
\pd{x_k}{x_i} = \sum_{j = 1}^n\pd{u}{y_j}\sum_{k=1}^n a_{jk}\delta_{ki} = 
\sum_{j = 1}^n\pd{u}{y_j}a_{ji}
\]
Differentiating once more with respect to $x_i$,
\[
\frac{\partial^2 u}{\partial x_i^2} = 
\sum_{j,k = 1}^n \frac{\partial^2 u}{\partial y_j \partial y_k} a_{ji}a_{ki}
\]
so that
\[
\Delta u(x) = \sum_{i=1}^n \sum_{j,k = 1}^n \frac{\partial^2 u}{\partial y_j \partial y_k} a_{ji}
a_{ki} = \sum_{j,k = 1}^n \frac{\partial^2 u}{\partial y_j \partial y_k} \sum_{i=1}^n a_{ji}a_{ki}
\]
Using the definition \eqref{s16e2} of orthogonality,
\[
\Delta u(x) = 
\sum_{j,k = 1}^n \frac{\partial^2 u}{\partial y_j \partial y_k} \delta_{jk} = 
\sum_{j=1}^n \frac{\partial^2 u}{\partial y_j} = \Delta u(y)
\]
\end{proof}

\noindent Proposition \ref{s16p1} suggests that Laplace's equation may have radial solutions.
\begin{prop}\label{s16p2}
$\Delta u = 0$ if and only if 
\begin{equation}\label{s16e4}
v^{\prime\prime} + \frac{n-1}{r}v^\prime = 0.
\end{equation}
where $r = (x_1^2 + \cdots + x_n^2)^{1/2} = \norm{x}$ and $u(x) = v(r)$.
\end{prop}
\begin{proof}
We have
\[
\pd{r}{x_i} = \frac{x_i}{r}
\]
so that
\[
\pd{v}{x_i} = \pd{v}{r}\pd{r}{x_i} = \frac{x_i}{r}v^\prime
\]
and
\[
\frac{\partial^2 v}{\partial x_i^2} = \frac{1}{r}v^\prime - \frac{x_i^2}{r^3}v^\prime + 
\frac{x_i^2}{r^2}v^{\prime\prime}
\]
Thus,
\[
\Delta v = v^{\prime\prime} + \frac{n}{r}v^\prime - \frac{1}{r}v^\prime = 
v^{\prime\prime} + \frac{n - 1}{r}v^\prime
\]
\end{proof}

\noindent We can write equation \eqref{s16e4} as
\[
\td{v^\prime}{r} = -(n-1)\frac{v^\prime}{r}
\]
so that
\[
\ln v^\prime = -(n-1)\ln r + \ln a = \ln ar^{-(n-1)}
\]
or $v^\prime = ar^{-(n-1)}$. One more integration gives
\begin{equation}\label{s16e5}
v(r) = \begin{cases}
a\ln r + b \text{ when } n = 2 \\
cr^{-(n-2)} + b \text { when } n \ge 2
\end{cases}
\end{equation}
where the constant $c = a/(n - 2)$.

For a particular choice of constants, the above equations are written as
\begin{equation}\label{s16e6}
\Phi(r) = \begin{cases}
-(2\pi)^{-1}\ln r \text{ when } n = 2 \\
[n(n-2)V(n, 1)]^{-1} r^{-(n - 2)} \text{ when } n > 2,
\end{cases}
\end{equation}
where
\begin{equation}\label{s16e7}
V(n, 1) = \frac{\pi^{n/2}}{\Gamma(1 + n/2)}
\end{equation}
is the volume of a unit ball. $\Phi(\cdot)$ are called \emph{fundamental solutions} and it is easy
to see that
\begin{eqnarray}
\norm{D\Phi(r)} &\le& C\norm{x}^{-(n-1)} \label{s16e8} \\
\norm{D^2\Phi(r)} &\le& C\norm{x}^{-n} \label{s16e9},
\end{eqnarray}
where $C$ is a constant. Unlike equation \eqref{s16e6} these relations are true for all integers $n$.

Before proceeding, we note the formulae for volume and surface area of an $n$-sphere or radius $r$.
\begin{eqnarray}
V(n, r) &=& \frac{\pi^{n/2}}{\Gamma(1 + n/2)}r^n \label{s16e10} \\
S(n, t) &=& \frac{n}{r}V(n, r). \label{s16e11}
\end{eqnarray}

\section{Problems}\label{s17}
\begin{enumerate}
\item Define $L(v) := 1 + v^2/4$, where $v$ is a scalar. Let $x(\cdot)$ be the trajectory with
initial value $x(t) = x$. Then the \emph{value function} is defined as 
\begin{equation}\label{s17e1}
u(t, x) = \min\left\{\int_t^\tau L(s, x(s), \dot{x}(s))ds : x \text{ is Lipshitz}\right\}.
\end{equation}
Show that $u$ is given by $u(t, x) = \min\{(t - \tau)L(v)\}$.

\noindent Solution: Let $w(s) = vs$. Then $w(t) = vt$, $w(\tau) = v\tau$ and $\dot{w}(s) = v$ so
that
\[
\int_t^\tau L(\dot{w}(s))ds = \int_t^\tau L(v)ds = (\tau - t)L(v).
\]
Therefore, 
\begin{equation}\label{s17e2}
u(t, x) = \min_{v \in \mathbb{R}}\{(\tau - t)L(v)\}.
\end{equation}

\item Continuing the previous problem, show that the minimizing solution is 
\[
v^\ast = \begin{cases}
2 & \text{ if } x \ge t \\
0 & \text{ if } |x| < t \\
-2 & \text{ if } x \le -t
\end{cases}
\]

\noindent Solution: Recall that $\tau$ is the exit time. It is easy to see that
\[
\tau = \begin{cases}
1 & \text{ if } |x| < t \\
v/x = v/1 = v & \text{ otherwise. }
\end{cases}
\]
so that equation \eqref{s17e2} becomes
\[
u(t, x) = \begin{cases}
(v - t)L(v) & \text{ if } |x| \ge t \\
(1 - t)L(v) & \text{ if } |x| < t.
\end{cases}
\]
Let us write the left hand side of the above equation as $f(v)$. Assuming that the minimum exists in 
\eqref{s17e2}, we can find it by differentiating the $f$ with respect to $v$ and setting the result
to zero to get $v^\ast$. If we do that, we get
\[
v^\ast = \begin{cases}
\pm 2 & \text{ if } |x| \ge t \\
0 & \text{ if } |x| < t.
\end{cases}
\]
Since $t$ is restricted to $[0, 1]$ while $x$ is in $[-1, 1]$, $v = x(s)/s$ implies that the first of 
the cases above can be further split as
\[
v^\ast = \begin{cases}
 2 & \text{ if } x \ge t \\
-2 & \text{ if } x \le -t \\
0 & \text{ if } |x| < t.
\end{cases}
\]
The corresponding value of $u(t, x)$ is
\[
u(t, x) = \begin{cases}
2(\tau - t) & \text{ if } x \ge t \\
-2(\tau - t) & \text{ if } x \ge -t \\
(1 - t) & \text{ if } |x| < t.
\end{cases}
\]
Since $\pm 2\tau = \pm 1$ be definition of the exit time, and as $\pm 2 t = \pm x$, we can 
write the above equation as
\[
u(t, x) = \begin{cases}
1 - |x| & \text{ if } |x| \ge t \\
1 - t & \text{ if } |x| < t.
\end{cases}
\]

\item Continuing with the previous problem, find the differentiable region and show $u$ satisfies the 
following Hamilton-Jacobi equation wherever it is differentiable:
\[
-u_t(t, x) + (u_x(t, x))^2 - 1 = 0.
\]
and satisfies the conditions
\[
u(t, 1) = u(t, -1) = 0, t \in [0, 1]; u(1, x) = 0, x \in [-1, 1].
\]

\noindent Solution: In the region $|x| \ge t$, except when $x = 0$,
\begin{eqnarray*}
u_t(t, x) &=& 0 \\
u_x(t, x) &=& \pm 1
\end{eqnarray*}
so that $-u_t(t, x) + (u_x(t, x))^2 - 1 = 0$. Likewise, in the region $|x| < t$,
\begin{eqnarray*}
u_t(t, x) &=& -1 \\
u_x(t, x) &=& 0 
\end{eqnarray*}
so that once again $-u_t(t, x) + (u_x(t, x))^2 - 1 = 0$. In order to find out the regions of 
differentiability, let us write the definition of $u$ more explicitly as
\[
u(t, x) = \begin{cases}
1 - x & \text{if } x \ge t \\
1 - t & \text{if } x < t \\
1 + x & \text{if } x \le -t
\end{cases}
\]
I think, the function is not differentiable along the line $x = -t$.

\item Consider a lagrangian $L(q) = 1 + |q|^2/4$, $q \in \mathbb{R}^n$ and derive the hamiltonian via
the Legendre transformation.

\noindent Solution: Let us write the lagrangian as
\[
L = 1 + \frac{1}{4}\sum_{i=1}^n q_i^2.
\]
I am assuming that $q = \dot{x}$ so that the `canonical momentum' is
\[
p = D_qL = \frac{1}{2}q.
\]
Thus the equation $p = D_q L$ is solvable for $q$ in terms of $p$ and hence the Hamiltonian is
\[
H(p) = \frac{1}{2}q\cdot q - L = |p|^2 - 1.
\]
We would have got the same result by considering the Legendre-Fenchel transform
\[
H(p) = \sup_{v \in \mathbb{R}^n}\left\{p \cdot q - L(q)\right\}.
\]
For, assuming that the supremum exists, we just need to find the maximum of the function $p \cdot q -
L(q)$, which is just $q^\ast = 2p$.

\item Continuing the above problem, define
\[
u(x, t) = \min\left\{\int_0^t\left(1 + \frac{\dot{w}(s)^2}{4}\right)ds + \frac{w(0)^2}{2}\right\},
\]
where the minimum is taken over all smooth trajectories $w$ satisfying $w(t) = x$.

\noindent Solution: Using Hopf-Lax formula we have
\[
u(x, t) = \min_{y \in \mathbb{R}^n}\left\{tL\left(\frac{x - y}{t}\right) + \frac{y^2}{2}\right\}.
\]
Substituting for $L$, we want to find the minimum of 
\[
f(y) = t + \frac{1}{4t}|x - y|^2 + \frac{y^2}{2},
\]
which is
\[
y^\ast = \frac{x}{1 + 2t}.
\]
Therefore,
\[
u(x, t) = t + \frac{|x|^2}{2(1 + 2t)}
\]
so that
\begin{eqnarray*}
u_t &=& 1 - \frac{|x|^2}{(1 + 2t)^2} \\
u_x &=& \frac{x}{1 + 2t}
\end{eqnarray*}
so that, since $H(p) = p^2 - 1$,
\[
H(u_x) = \frac{|x|^2}{(1 + 2t)^2} - 1
\]
Thus, we have $u_t + H(u_x) = 0$. The initial condition is $u(x, 0) = |x|^2/2$.

\item Consider the Burger's equation $u_t + uu_x = 0$ for $x \in \mathbb{R}, t > 0$ with initial
condition $u(x, 0) = u_0(x)$ where $u_0$ is given by
\[
u_0(x) = \begin{cases}
1 & \text{ if } x \le 0 \\
1 - x \text{ if } x \in [0, 1] \\
0 \text{ if } x \ge 1.
\end{cases}
\]
\begin{itemize}
\item Show that the characteristic curves do not meet till $t = 1$ by constructing them. Draw the
characteristics and solve the problem for $u(x, t)$ for all $x \in \mathbb{R}$ and $t \in (0, 1)$.

\noindent Solution: The parametric equations of the Monge curve are
\begin{eqnarray}
D_\sigma t &=& 1 \\
D_\sigma x &=& u \\
D_\sigma u &=& 0 
\end{eqnarray}
and their general solutions are
\begin{eqnarray}
t(\sigma) &=& \sigma + c_1 \\
x(\sigma) &=& u\sigma + c_2 \\
u(\sigma) &=& c_3.
\end{eqnarray}

The parametric equations of the initial curve are $t(\sigma = 0) = 0, x(\sigma = 0) = \eta$ 
so that the particular solutions of the differential equations of Monge curve are
\begin{eqnarray}
t(\sigma) &=& \sigma \label{s17e9} \\
x(\sigma) &=& u\sigma + \eta \label{s17e10}
\end{eqnarray}
and 
\begin{equation}\label{s17e11}
u(\sigma = 0) = \begin{cases}
1 & \text{ if } \eta \le 0 \\
1 - \eta & \text{ if } \eta \in [0, 1] \\
0 & \text{ if } \eta \ge 1
\end{cases}
\end{equation}
Combining equations \eqref{s17e9} to \eqref{s17e11}, we get
\begin{equation}\label{s17e12}
x(\sigma) = \begin{cases}
\sigma + \eta & \text{if } \eta \le 0 \\
(1 - \eta)\sigma + \eta & \text{if } \eta \in [0, 1] \\
\eta & \text{if } \eta \ge 1.
\end{cases}
\end{equation}
Therfore, the inverse relations are
\begin{equation}\label{s17e13}
\sigma = t
\end{equation}
and
\begin{equation}\label{s17e14}
\eta = \begin{cases}
x - t & \text{ if } \eta \le 0 \\
(x - t)(1 - t)^{-1} & \text{ if } \eta \in [0, 1] \\
x  & \text{ if } \eta \ge 1
\end{cases}
\end{equation}
Using \eqref{s17e14} in \eqref{s17e11}, we get, for $t \in (0, 1)$,
\begin{equation}\label{s17e15}
u(x, t) = \begin{cases}
1 & \text{ if } x \le t \\
(1 - x)(1 - t)^{-1} & \text{ if } t \le x \le 1 \\
0 & \text{ if } x \ge 1.
\end{cases}
\end{equation}

Now consider equation \eqref{s17e12} for the cases $\eta_1, \eta_2 \in [0, 1]$
\begin{eqnarray}
x(\sigma, \eta_1) &=& (1 - \eta_1)\sigma + \eta_1 \label{s17e16} \\
x(\sigma, \eta_2) &=& (1 - \eta_2)\sigma + \eta_2 \label{s17e17}
\end{eqnarray}
The two curves meet when $x(\sigma, \eta_1) = x(\sigma, \eta_2)$, which happens when $\sigma = 1$.
From equation \eqref{s17e13}, $\sigma = 1$ whrn $t = 1$. Thus, the characteristic curves cross when
$t = 1$ and therefore equation \eqref{s17e15} is valid only in $(0, 1)$.

\item Construct a curve of discontinuity $\sigma(t)$ for $t \ge 1$ with $\sigma(1) = 1$, define a
discontinuous solution $u(x, t)$ for all $x < \sigma(t)$, $x > \sigma(t)$ and $1 \le t$ that 
satisfies Rankine-Hugoniot condition.

\noindent Solution: Let 
\begin{equation}
\sigma(t) = \frac{\alpha + t}{\alpha + 1},
\end{equation}
where $\alpha \ge 0$ and let 
\begin{equation}
u(x, t) = \begin{cases}
\sqrt{2/(\alpha + 1)} & \text{if } x < \sigma(t) \\
0 & \text{if } x > \sigma(t)
\end{cases}
\end{equation}
The function $F$ when Burger's equation is written in the form of conservation law $u_t + (F(u))_x 
= 0$ is $F(u) = u^2/2$. Therefore in the region to the left of $x = \sigma(t)$, $F(u_l) = 
1/(\alpha + 1)$ and on the right it is $F(u_r) = 0$ so that the jump in $F$ is $1/(\alpha + 1)$. The 
jump is exactly equal to $\dot{\sigma}$.

However, $u(x, t)$ should also agree with \eqref{s17e15}, we require $u(x, t) = 1$ for $x < 
\sigma(t)$, which is possible only if $\alpha = 1$.
\end{itemize}
\end{enumerate}

\section{Lax-Oleinik formula}\label{s18}
We are interested in finding an analog of the Hopf-Lax formula for the initial value problem of
the conservation law \eqref{s15e1}
\begin{eqnarray*}
u_t + (F(u))_x &=& 0 \text{ in } \mathbb{R} \times (0, \infty) \\
u &=& g \text{ on } \mathbb{R} \times {t = 0}
\end{eqnarray*}

Let us consider an equation of the form
\begin{eqnarray}
w_t + F(w_x) &=& 0 \text{ in } \mathbb{R} \times (0, \infty) \label{s18e1} \\
w &=& h \text{ on } \mathbb{R} \times {t = 0} \nonumber
\end{eqnarray}
Equation \eqref{s18e1} is the Hamilton-Jacobi equation for the hamiltonian function $F$. Then the
solution $w$ given by Hopf-Lax formula is
\[
w(x, t) = \min_{y \in \mathbb{R}}\left\{tL\left(\frac{x - t}{t}\right) + h(y)\right\},
\]
where $L = F^\ast$. How is all this related to the conservation law? To find it out, let us 
differentiate equation \eqref{s18e1} with respect to $x$ so that
\begin{eqnarray}
(w_x)_t + (F(w_x))_x &=& 0 \text{ in } \mathbb{R} \times (0, \infty) \label{s18e2} \\
w_x &=& h_x \text{ on } \mathbb{R} \times {t = 0}. \nonumber
\end{eqnarray}
Equation \eqref{s18e2} is the conservation law for $u = w_x$. Since $h_x = g$ we need
\[
h(x) = \int_0^x g(x^\prime)dx^\prime.
\]
Finally, we conjecture that the solution of the conservation law \eqref{s15e1} is
\[
u = w_x = \frac{\partial}{\partial x}
\left[\min_{y \in \mathbb{R}}\left\{tL\left(\frac{x - t}{t}\right) + h(y)\right\}\right].
\]
Lax-Oleinik formula is a justification of this conjecture.

\subsection{$L^p$ spaces}
\begin{defn}\label{s18d1}
Let $1 \le p < \infty$ and let $(S, \Sigma, \mu)$ be a measure space. $L^p$ space is the set of all
measurable functions $f:S \rightarrow \mathbb{R}$ such that the $L^p$ norm,
\[
\norm{f}_p = \left(\int_S |f|^p d\mu\right)^{1/p} < \infty.
\]
\end{defn}

$L^\infty$ space is a generalization of $L^p$ spaces for which $p \rightarrow \infty$. The 
corresponding norm is defined as
\begin{equation}\label{s18e3}
\norm{f}_\infty = \esssup_{x \in S} |f|.
\end{equation}
The essential supremum is a generalization of the maximum to measurable functions. If $f:S 
\rightarrow \mathbb{R}$, the essential supremum of $f$ is the smallest number $\alpha$ such that
the set $\{x \in S : f(x) > \alpha\}$ has measure zero. If no such number exists then the essential
supremum is $\infty$. Thus we have
\begin{defn}\label{s18d2}
Let $(S, \Sigma, \mu)$ be a measure space. $L^\infty$ space is the set of all measurable functions
$f: S \rightarrow \mathbb{R}$ such that $\norm{f}_\infty < \infty$.

The following theorem gives the solution of the conservation law.
\begin{thm}[Lax-Oleinik formula]\label{s18t1}
Assume that $F: \mathbb{R} \rightarrow \mathbb{R}$ in \eqref{s15e1} is smooth an uniformly convex
and $g \in L^\infty(\mathbb{R})$. Then,
\begin{itemize}
\item[(i)] For each time $t > 0$, there exists for all but at most countably many values of $x \in
\mathbb{R}$ a unique point $y(x, t)$ sich that
\[
\min_{y \in \mathbb{R}}\left\{tL\left(\frac{x - y}{t}\right) + h(y)\right\} = 
tL\left(\frac{x - y(x,t)}{t}\right) + h(y(x, t)).
\]
\item[(ii)] The mapping $x \mapsto y(x, t)$ is non-decreasing.
\item[(iii)] For each time $t > 0$, the solution of \eqref{s15e1} is
\[
u(x, t) = G\left(\frac{x - y(x, t)}{t}\right),
\]
where the function $G = (F^\prime)^{-1}$.
\end{itemize}
\end{thm}
\end{defn}

\section{Classification of linear PDEs}\label{s19}
A general linear PDE of order $k$ is of the form
\begin{equation}\label{s19e1}
L(u) = \sum_{|\alpha| \le k} a_\alpha(x) D^\alpha(u) = f,
\end{equation}
where $x \in \mathbb{R}^n$ and $a_\alpha, f: \mathbb{R}^n \rightarrow \mathbb{R}$ are given 
functions.

\begin{defn}\label{s19d1}
The principal part of a linear differential operator $L$ of degree $k$ of the form
\[
L = \sum_{|\alpha| \le k} a_\alpha(x) D^\alpha,
\]
is 
\[
L_0 = \sum_{|\alpha| = k} a_\alpha(x) D^\alpha.
\]
\end{defn}

Since $x \in \mathbb{R}^n$, the multi-index $\alpha$ is of the form $(\alpha_1, \ldots, \alpha_n)$.
\begin{defn}\label{s19d2}
The characteristic form of associated with a linear differential operator $L$ of degree $k$ is
\[
\chi_L(x, \xi) = \sum_{|\alpha| = k} a_\alpha(x) \xi^\alpha,
\]
where $\xi^\alpha = (\xi_1^{\alpha_1}, \ldots, \xi_n^{\alpha_n})$.
\end{defn}

\begin{defn}\label{s19d3}
The characteristic variety associated with a linear differential operator $L$ of degree $k$ is
\[
\cvar{L} = \{\xi \in \mathbb{R}^n : \xi \ne 0 \text{ and } \chi_L(x, \xi) = 0\}.
\]
\end{defn}

Examples of characteristic forms for linear differential operators over $x \in \mathbb{R}^n$.
\begin{enumerate}
\item $L = D^{e_i}$ has $\chi_L = \xi_i$ and $\cvar{L} = \varnothing$, where $1 \le i \le n$.
\item $L = D^{(1,1)}$ has $\chi_L = \xi_1\xi_2$ and $\cvar{L}$ is the set of all mutually orthogonal
vectors.
\item $L = D^{e_1} + \imath D^{e_2}$ has $\chi_L = \xi_1 + \imath\xi_2$ and $\cvar{L} = \varnothing$.
\item $L = \Delta$, the Laplacian, has $\chi_L = \xi_1^2 + \cdots + \xi_n^2$ and $\cvar{L} = 
\varnothing$.
\item If 
\[
L = \frac{\partial}{\partial x_1} - \sum_{i=2}^n\frac{\partial^2}{\partial x_i^2}
\]
then 
\[
\chi_L = \xi_1 - \sum_{i=2}^n\xi_i^2
\]
and $\cvar{L} = \{\xi \in \mathbb{R}^n : \xi = \xi_2^2 + \cdots + \xi_n^2\}$.
\item If 
\[
L = \frac{\partial^2}{\partial x_1^2} - \sum_{i=2}^n\frac{\partial^2}{\partial x_i^2}
\]
then 
\[
\chi_L = \xi_1^2 - \sum_{i=2}^n\xi_i^2
\]
and $\cvar{L} = \{\xi \in \mathbb{R}^n : \xi^2 = \xi_2^2 + \cdots + \xi_n^2\}$.
\end{enumerate}

\begin{defn}\label{s19d4}
A linear differential operator $L$ is elliptic of $\cvar{L} = \varnothing$.
\end{defn}

Let us now focus on second order linear differential operators, first on $\mathbb{R}^2$. In this
case we simplify our notation by letting $(x, y)$ be a general vector in $\mathbb{R}^2$ and 
writing the operator as 
\begin{equation}\label{s19e2}
Lu = a(x, y)u_{xx} + 2b(x, y)u_{xy} + c(x, y)u_{yy} + \text{ lower order terms}.
\end{equation}
The principal part of $L$ is 
\begin{equation}\label{s19e3}
L_0u = a(x, y)u_{xx} + 2b(x, y)u_{xy} + c(x, y)u_{yy}
\end{equation}
and its characteristic form is
\begin{equation}\label{s19e4}
Q(\xi, \eta) = a\xi^2 + 2b\xi\eta + c\eta^2 = \eta^2\left(a\zeta^2 + 2b\zeta + c\right) = q(\zeta),
\end{equation}
where $\zeta = \xi/\eta$. The quadratic form $Q$ is
\begin{itemize}
\item Definite if $Q(\xi, \eta) = 0 \iff (\xi, \eta) = 0$.
\item Semi-definite if $Q(\xi, \eta)$ is either non-negative for all $(\xi, \eta)$ or non-positive
for all $(\xi, \eta)$.
\item Indefinite if $Q(\xi, \eta)$ changes sign as $(\xi, \eta)$ varies over $\mathbb{R}^2$.
\end{itemize}

Recall that $(\xi, \eta) \in \mathbb{R}^2$ so that \emph{only in the case of linear differential
operator of second order over two variables},
\begin{itemize}
\item $Q$ is definite $\Rightarrow q(\zeta) = 0$ has complex roots or $b^2 - ac < 0$ and $L$ is 
called elliptic.
\item $Q$ is semi-definite $\Rightarrow q(\zeta) = 0$ has real and equal roots or $b^2 - ac = 0$ and 
$L$ is called parabolic.
\item $Q$ is indefinite $\Rightarrow q(\zeta) = 0$ has distinct real roots or $b^2 - ac > 0$ and $L$ 
is called hyperbolic.
\end{itemize}

If $a, b, c$ are constants then the linear differential operators retain their elliptic, parabolis of 
hyperbolic nature throughout the real plane. Otherwise, their nature depends on where we examine them.

There is more to this scheme of classification than the nature of roots of the characteristic forms.
Refer to section 1, chapter 2 of Fritz John's book \cite{fj} for a relation between the roots and the 
characteristic curves.

\section{Fundamental solutions and Poisson equation}\label{s20}
Elliptic equations have no characteristics and therefore have no initial value problem as well. 
Instead, they have boundary value problems of at least the following types:
\begin{itemize} 
\item $u(x) = g(x)$ on $\partial U$, called the Dirichlet boundary condition,
\item $n\cdot Du = g$ on $\partial U$, where $n$ is the outward normal to the boundary, called
the Neumann boundary condition and
\item Mixed boundary conditions which are a combination of Dirichlet and Neumann boundary conditions.
\end{itemize}

\noindent We next define the convolution of functions $\Phi, f: \mathbb{R}^n \rightarrow \mathbb{R}$
\begin{equation}\label{s20e1}
(\Phi \star f)(x) = \int_{\mathbb{R}^n}\Phi(x - y)f(y)dy
\end{equation}
We will not show that if $u(x) = (\Phi \star f)(x)$ then $\Delta u = -f$. We first note that
\[
\Delta \int_{\mathbb{R}^n}\Phi(x - y)f(y)dy \ne \int_{\mathbb{R}^n}\Delta \Phi(x - y)f(y)dy
\]
because $\Delta\Phi(x - y)$ blows up at $x = y$. Computing the laplacian of $u$ needs more care. In
the proof that follows, we will write $\Delta_x$ and $\Delta_y$ to differentiate the laplacians with
respect to $x$ and $y$ respectively.

\begin{thm}\label{s20t1}
If $u(x) = (\Phi \star f)(x)$ then $\Delta u = - f$ where $f$ is a $C^2(\mathbb{R}^n)$ function with a 
compact support.
\end{thm}
\begin{proof}
In the relation 
\[
u(x) = (\Phi \star f)(x) = \int_{\mathbb{R}^n}\Phi(x - y)f(y)dy
\]
put $x - y = z$ for a fixed $x$. Since the Lebesgue measure is translation invariant, $dy = dz$ 
and therefore
\[
u(x) = \int_{\mathbb{R}^n}\Phi(z)f(x - z)dz = \int_{\mathbb{R}^n}\Phi(y)f(x - y)dy.
\]
Consider
\[
\frac{u(x + h_m e_i) - u(x)}{h_m} = \int_{\mathbb{R}^n}\Phi(y)\frac{f(x+h_me_i-y) - f(x-y)}{h_m}dy,
\]
where $e_i$ is the $i$th member of the standard basis of $\mathbb{R}^n$ and $h_m \in \mathbb{R}$. The
series $\{g_m\}$ where
\[
g_m = \frac{f(x+h_me_i-y) - f(x-y)}{h_m}
\]
converges to $u_{x_i}$ uniformly. Therefore, the integral and the derivative can be interchanged
and hence
\begin{eqnarray}
u_{x_i} &=& \int_{\mathbb{R}^n}\Phi(y)f_{x_i}(x - y)dy \\
u_{x_ix_j} &=& \int_{\mathbb{R}^n}\Phi(y)f_{x_ix_j}(x - y)dy.
\end{eqnarray}
From the above two equations it is clear that
\[
\Delta_x u = \int_{\mathbb{R}^n}\Phi(y)\Delta_xf(x - y)dy.
\]
Since $\Phi$ blows up at the origin, we will isolate this singularity by splitting the integral
into two as
\begin{equation}\label{s20e4}
\Delta_x u = I(\epsilon) + J(\epsilon)
\end{equation}
where
\begin{eqnarray}
I(\epsilon) &=& \int_{B(0, \epsilon)}\Phi(y)\Delta_xf(x - y)dy \label{s20e5} \\
J(\epsilon) &=& \int_{\mathbb{R}^n \setminus B(0, \epsilon)}\Phi(y)\Delta_xf(x - y)dy \label{s20e6} 
\end{eqnarray}
Consider the modulus of $I(\epsilon)$,
\[
|I(\epsilon)| \le \int_{B(0, \epsilon)}|\Phi(y)||\Delta_x f(x - y)|dy 
\]
Since $f$ is a $C^2$ function with a compact support, over the domain of integration $|\Delta_x
f(x-y)| \le \norm{D^2f}_\infty$, so that
\[
|I(\epsilon)| \le \norm{D^2f}_\infty \int_{B(0, \epsilon)}|\Phi(y)|dy.
\]
In $B(0, \epsilon)$, 
\[
|\Phi(y)| \le \begin{cases}
C_1\ln\epsilon & \text{ when } n = 2 \\
C_1\epsilon^{n-2} & \text{ when } n > 2,
\end{cases}
\]
for some constant $C_1$ while the volume of the ball itself is 
\[
V(n, \epsilon) = \frac{\pi^{n/2}}{\Gamma(1 + n/2)}\epsilon^n
\]
so that
\[
|I(\epsilon)| \le \begin{cases}
C_2\epsilon^2\ln\epsilon & \text{ when } n = 2 \\
C_2\epsilon^2 & \text{ when } n > 2
\end{cases}
\]
for some constant $C_2$. In either case, 
\begin{equation}\label{s20e7}
\lim_{\epsilon \rightarrow 0}|I(\epsilon)| = 0.
\end{equation}
Since $\Delta_x f(x - y) = \Delta_y f(x - y)$, we can write \eqref{s20e6} as
\[
\int_{\mathbb{R}^n \setminus B(0, \epsilon)}\Phi(y)\Delta_yf(x - y)dy.
\]
Our equations will be more transparent if we write $\Delta_y = D_y^2$ so that
\[
J(\epsilon) = \int_{\mathbb{R}^n \setminus B(0, \epsilon)}\Phi(y)D_y(D_yf(x - y))dy.
\]
Using integration by parts \eqref{s14e6},
\begin{equation}\label{s20e8}
J(\epsilon) =  K(\epsilon) + L(\epsilon)
\end{equation}
where
\begin{eqnarray}
K(\epsilon) &=& -\int_{\mathbb{R}^n \setminus B(0, \epsilon)} D_y\Phi D_yf(x-y) dy \label{s20e9} \\
L(\epsilon) &=& \int_{\partial B(0, \epsilon)} \Phi(y) D_y f(x - y) \cdot n dS(y) \label{s20e10}
\end{eqnarray}
and $dS(y)$ is a surface measure. In the surface integral, we ignore the outer boundary of the
region because $f$ has a compact support. Analogous to $I(\epsilon)$,
\[
|L(\epsilon)| \le C_3\norm{Df}_\infty \int_{\partial B(0, \epsilon)} \Phi(y) dS(y)
\]
Using the fact that the surface area of an $n$-ball of radius $r$ is $nV(n, r)/r$ we get
\[
|L(\epsilon)| \le \begin{cases}
C_4\norm{Df}_\infty \epsilon\ln\epsilon & \text{ when } n = 2 \\
C_4\norm{Df}_\infty \epsilon & \text{ when } n > 2
\end{cases}
\]
for some constant $C_4$. In either case,
\begin{equation}\label{s20e11}
\lim_{\epsilon \rightarrow 0}|L(\epsilon)| = 0.
\end{equation}
Using integration by parts once more on the right hand side of \eqref{s20e9},
\[
K(\epsilon) = \int_{\mathbb{R}^n \setminus B(0, \epsilon)} D^2_y\Phi f(x - y) dy -
\int_{\partial B(0, \epsilon)} f(x - y) D_y\Phi(y) \cdot n dS(y).
\] 
In the domain of integration $\Delta_y\Phi(y) = 0$. Therefore the first integral vanishes and
\[
K(\epsilon) = -\int_{\partial B(0, \epsilon)} f(x - y) D_y\Phi(y) \cdot n  dS(y).
\]
From \eqref{s16e6},
\[
D_y\Phi(y) = C_5 \frac{y}{\norm{y}^n},
\]
for some constant $C_5$ all $n \ge 2$, while $n = -y/\norm{y}$ so that
\[
D_y\Phi(y) \cdot n = C_5\frac{\norm{y}^2}{\norm{y}^{n+1}} = C_5\frac{1}{\norm{y}^{n-1}}.
\]
On the surface $\partial B(0, \epsilon)$,
\[
D_y\Phi(y) \cdot n = C_5\frac{1}{\epsilon^{n-1}}
\]
so that
\[
K(\epsilon) = -\frac{C_5}{\epsilon^{n-1}}\int f(x - y)dy.
\]
The constant $C_5$ is just the normalizing constant in equations\eqref{s16e6} so that the above
integral is just the average of $f$ over the surface $\partial B(0, \epsilon)$. In the limit
$\epsilon \rightarrow 0$, it just becomes the value of $f$ at $x$. Therefore,
\[
\lim_{\epsilon \rightarrow 0}K(\epsilon) = -f(x).
\]
Using this equation along with\eqref{s20e4}, \eqref{s20e7} and \eqref{s20e11}, we get $\Delta u = -f$.
\end{proof}

\section{Mean value property}\label{s21}
\begin{defn}\label{s21d1}
A function $u$ is called sub-harmonic (super-harmonic) if $\Delta u \ge 0$ ($\Delta u \le 0$).
\end{defn}

\begin{defn}\label{s21d2}
A set $A$ is compactly embedded in a domain $U$ if $A \subset \bar{A} \subset \Int{U} \subset U$,
where $\bar{A}$ is the closure of $A$ and $\Int{U}$ is the interior of $U$.
\end{defn}

\noindent If $A$ is compactly embedded in $U$, we express it as $A \cce U$.

\begin{thm}\label{s21t1}
Let $u \in C^2(U)$ and $B(y, R) \cce U$. If $u$ is sub-harmonic then
\begin{eqnarray*}
u(y) &\le& \frac{1}{|B(y, R)|}\int u(x)dx \\
 &\le& \frac{1}{|\partial B(y, R)|}\int u(x)d\sigma(x),
\end{eqnarray*}
where $B(y, R)$ is a ball of radius $R$ centered at $y$.
\end{thm}
\begin{proof}
Let $0 < \rho < R$. Since $u$ is sub-harmonic,
\[
0 \le \int_{B(y,\rho)}\Delta u dx = \int_{\partial B(y,\rho)} Du \cdot nd\sigma(x).
\]
The integral on the extreme right hand side depends on $\rho$. Let us define a function $g: \mathbb{R}^+ 
\rightarrow \mathbb{R}^+$ as
\[
g(\rho) = \int_{\partial B(y,\rho)} Du \cdot nd\sigma(x) \ge 0.
\]
The gradient involves derivatives only with respect to $r = \norm{x}$. However, $r$ varies over the surface 
$\partial B(y,\rho)$ because the sphere $B(y, R)$ is not centered at the origin. Therefore, we transform $x$ so that
the integration is over the surface of a unit sphere centered at the origin. Let 
\begin{equation}\label{s21e1}
w = \frac{x - y}{\rho}
\end{equation}
so that $d\sigma(x) = \rho^{n-1}d\sigma(w)$ and
\[
g(\rho) = \rho^{n-1}\int_{B(0,1)} Du(w)\cdot nd\sigma(w) = \rho^{n-1}\int_{B(0,1)} \pd{u}{r} d\sigma(w).
\]
In this configuration $Du$ is everywhere parallel to $n$ and hence $Du \cdot n = |Du|$. Since $Du$ involves just 
the radial derivative, $Du = \partial u/\partial r$. We can now pull out the derivative outside the integral and
get
\[
g(\rho) = \rho^{n-1}\frac{\partial}{\partial r}\int_{B(0,1)} u(w)d\sigma(w).
\]
If we reverse the transformation \eqref{s21e1} we get
\[
g(\rho) = \frac{\partial}{\partial\rho}\int_{B(y, R)} u(x)d\sigma(x) \ge 0
\]
Thus $g$ is a monotonically increasing function. Since $\rho > 0$, $g(\rho)/n\alpha_n\rho^{n-1}$, where $\alpha_n
= \pi^{n/2}/\Gamma(1 + n/2)$ is also monotonically increasing function. Let $0 \le \rho_1 < \rho_2 \le R$ so that
\[
\frac{1}{n\alpha_n\rho^{n-1}}\int_{B(y,\rho_1)} u(x)d\sigma(x) \le 
\frac{1}{n\alpha_n\rho^{n-1}}\int_{B(y,\rho_2)} u(x)d\sigma(x)
\]
Take the limit $\rho_1 \rightarrow 0$ so that the left hand side becomes $u(y)$ and
\begin{equation}\label{s21e2}
u(y) \le \frac{1}{n\alpha_n\rho^{n-1}}\int_{B(y,\rho_2)} u(x)d\sigma(x).
\end{equation}
If we take the limit of the right hand side as $\rho_2 \rightarrow R$ we get
\[
u(y) \le \frac{1}{|B(y, R)|}\int_{B(y,\rho_2)} u(x)d\sigma(x).
\]
If, instead, we write \eqref{s21e2} as
\[
n\alpha_n\rho^{n-1} u(y) \le \int_{\partial B(y,\rho_2)} u(x)d\sigma(x)
\]
and integrate with respect to $\rho$ from $0$ to $R$,
\[
\alpha_n \rho^n u(y) \le \int_{B(y, R)} u(x) dx
\]
which is same as
\[
u(y) \le \frac{1}{|B(y, R)|}\int_{B(y, R)} u(x) dx.
\]
\end{proof}

\begin{rem}
Recall that $|B(y, R)|$ is the volume of the $n$-sphere of radius $R$ centered at $y$ and $|\partial B(y, R)|$ is 
its surface area.
\end{rem}

We can analogously prove 
\begin{thm}\label{s21t2}
Let $u \in C^2(U)$ and $B(y, R) \cce U$. If $u$ is super-harmonic then
\begin{eqnarray*}
u(y) &\ge& \frac{1}{|B(y, R)|}\int u(x)dx \\
 &\ge& \frac{1}{|\partial B(y, R)|}\int u(x)d\sigma(x),
\end{eqnarray*}
where $B(y, R)$ is a ball of radius $R$ centered at $y$.
\end{thm}

An immediate consequence of theorems \ref{s21t1} and \ref{s21t2} is
\begin{thm}\label{s21t3}
Let $u \in C^2(U)$ and $B(y, R) \cce U$. If $u$ is harmonic then
\begin{eqnarray*}
u(y) &=& \frac{1}{|B(y, R)|}\int u(x)dx \\
 &=& \frac{1}{|\partial B(y, R)|}\int u(x)d\sigma(x),
\end{eqnarray*}
where $B(y, R)$ is a ball of radius $R$ centered at $y$.
\end{thm}

It is remarkable that the converse of the above theorem is also true.
\begin{thm}\label{s21t4}
If $u \in C^2(\mathbb{R})$ and $u$ satisfies the mean value property of theorem \ref{s21t3} for all $x \in U$
and for all $B(y, R) \cce U$. Then $\Delta u = 0$ in $U$.
\end{thm}
\begin{proof}
Let $x \in U$ and $B(y,r) \cce U$. From theorem \ref{s21t3},
\[
h(r) = \frac{1}{|\partial B(y,r)|}\int u(x)d\sigma(x) = u(y).
\]
At least locally, $h$ is independent of $r$ so that $h^\prime(r) = 0$. Thus,
\[
\frac{d}{dr}\left(\frac{1}{|\partial B(y,r)|}\int u(x)d\sigma(x)\right) = 0.
\]
Using transformation of \eqref{s21e1}, 
\[
\frac{d}{dr}\left(\frac{r^{n-1}}{|\partial B(y,r)|}\int u(y + rw)d\sigma(w)\right) = 0.
\]
Since $|\partial B(y,r)| = r^{n-1}|\partial B(0,1)|$,
\[
\frac{d}{dr}\left(\frac{1}{|\partial B(0,1)|}\int u(y + rw)d\sigma(w)\right) = 0.
\]
The derivative can now be pulled inside the integral so that
\[
\frac{1}{|\partial B(0,1)|} \int w\cdot Du d\sigma(w) = 0.
\]
Reversing the transformation,
\[
\frac{1}{|\partial B(y,r)|} \int \frac{x - y}{r}\cdot Du d\sigma(x) = 0.
\]
But $(x - y)/r$ is just the unit normal $n$ to the sphere $B(y,r)$ so that
\[
\frac{1}{|\partial B(y,r)|} \int n\cdot Du d\sigma(x) = 0 \Rightarrow \int n\cdot Du d\sigma(x) = 0
\]
Using Gauss theorem we get
\[
\int_{B(y,r)} \Delta u dx = 0.
\]
If this is true for all spheres compactly contained in $U$, $\Delta u = 0$.
\end{proof}

\begin{thm}[Strong Maximum Principle]\label{s21t5}
Let $U$ be an open, connected set, $u \in C^2(U) \cap C(\bar{U})$ and $\Delta u = 0$. Then 
$\max_{\bar{U}}u = \max_{\partial U}u$.
\end{thm}
\begin{proof}
Let $M := \max_{\bar{U}}u$ and let $x_0 \in U$ be such that $u(x_0) = M$. Let $\delta$ be the least 
distance of $x_0$ from $\partial U$ and let $0 < r < \delta$. Using theorem \ref{s21t3} we get
\[
M = \frac{1}{|B(x_0,r)|}\int u(x)dx \le \frac{M}{|B(x_0,r)|}\int dx \le M
\]
If for any $x \in B(x_0,r)$ $u(x) = M$ then the last inequality will be strict, leading to a contradiction. 
Therefore $u(x) = M$ for all $x \in B(x_0,r)$. If $A = \{x \in U : u(x) = M\}$ then $A$ is an open set.

To prove that $A$ is a closed set, consider a Cauchy sequence $\{x_i\}$ in $A$. Since $u$ is continuous, the limit
of this sequence will also be in $A$, making it a closed set.

Since $U$ is connected, a subset $A$ of $U$ will be both open and closed only if $A = \varnothing$ or 
$A = U$. But $A$ cannot be empty because a continuous function in $U$ will have at least one maximum
in $U$. Therefore $A = U$, that is $u$ is a constant over $U$ if it attains a maximum in its 
interior.

If, therefore, $u$ is not constant in $U$ then it must attain its maximum only on $\partial U$.
\end{proof}

We can analogously prove
\begin{thm}[Strong Minimum Principle]\label{s21t6}
Let $U$ be an open, connected set, $u \in C^2(U) \cap C(\bar{U})$ and $\Delta u = 0$. Then 
$\min_{\bar{U}}u = \min_{\partial U}u$.
\end{thm}

\begin{thm}[Uniqueness of solution of Poisson equation]\label{s21t7}
Let $g \in C(\partial U), f \in C(U)$. Then there exists at most one solution $u \in C^2(U)\cap 
C(\bar{U})$ of the boundary value problem $-\Delta u = f$ in $U$ and $u = g$ on $\partial U$.
\end{thm}
\begin{proof}
Let, if possible, there be two solutions $u$ and $v$ of the boundary-value problem. Then $w_1 = u - v$ and $w_2 =
v - u$ are both solutions of the Laplace's equation with the boundary condition being that $w_1 = 0$ and $w_2 = 0$
on $\partial U$. By the strong maximum principle, the maximum values taken by both $w_1$ and $w_2$ are zero.
Further, by construction $w_1 = -w_2$. Therefore, both are identically zero.
\end{proof}

\section{Ideas from functional analysis}\label{s22}
\begin{defn}\label{s22d1}
A pair $(X, d)$ where $X$ is a non-empty set and $d: X \times X \rightarrow \mathbb{R}$ is called a metric space if
\begin{itemize}
\item $d(x, y) = d(y, x)$ for all $x, y \in X$.
\item $d(x, y) = 0$ if and only if $x = 0$.
\item $d(x, z) \le d(x, y) + d(y, z)$ for all $x, y, z \in X$.
\end{itemize}
\end{defn}

\begin{prop}\label{s22p1}
If $(X, d)$ is a metric space then $d(x, y) \ge 0$ for all $x, y \in X$.
\end{prop}
\begin{proof}
$d(x, y) + d(y, x) \ge d(x, x) \Rightarrow d(x, y) + d(x, y) \ge 0 \Rightarrow 2d(x, y) \Rightarrow 0$.
\end{proof}

\begin{defn}\label{s22d2}
A mapping $\norm{\cdot}: X \rightarrow \mathbb{R}$ is called a norm if
\begin{itemize}
\item $\norm{x + y} \le \norm{x} + \norm{y}$ for all $x, y \in X$,
\item $\norm{\lambda x} = |\lambda|\norm{x}$ for all $x \in X, \lambda \in \mathbb{R}$,
\item $\norm{x} = 0$ if and only if $x = 0$.
\end{itemize}
\end{defn}

\begin{rem}
The second point in the above definition is called the \emph{homogeneity condition}.
\end{rem}

\begin{defn}\label{s22d3}
A set $X$ equipped with a norm $\norm{\cdot}$ is called a normed space.
\end{defn}

A metric space need not be a normed space. Consider a metric space $(X, d)$ where $d(x, y) = \norm{x - y}$. Then
$d(\lambda x, \lambda y) = \norm{\lambda(x - y)} = |\lambda|\norm{x - y}$. If $d$ is a discrete metric, that is, 
$d(x, y) = $ if $x \ne y$ then clearly $d(\lambda x, \lambda y) \ne 1$ is $x \ne y$.

\begin{defn}\label{s22d4}
A metric is said to be homogeneous if $d(\lambda x, \lambda y) = |\lambda|d(x, y)$ for all $\lambda \in \mathbb{R}$
and translational invariant if $d(x + a, y + a) = d(x, y)$ for all $x, y, a \in X$.
\end{defn}

\begin{prop}\label{s22p2}
A metric space $(X, d)$ with a homogeneous and translational invariant metric is a normed space with the norm
$\norm{x} = d(x, 0)$.
\end{prop}
\begin{proof}
$\norm{x} = d(x, 0) = 0$ if an only if $x = 0$ follows from the property of $d$. $\norm{\lambda x} = d(\lambda x,
0) = |\lambda|d(x, 0)$ from the homogeneity of $d$. Therefore, $\norm{\lambda x} = |\lambda|\norm{x}$. Finally,
$\norm{x + y} = d(x + y, 0) = d(x, -y)$ by translational invariance so that $\norm{x + y} \le d(x, 0) + d(-y, 0)
\le d(x, 0) + d(y, 0)$, where we used homogeneity. Finally, using the definition of norm $\norm{x + y} \le \norm{x}
+ \norm{y}$.
\end{proof}

\begin{defn}\label{s22d5}
A sequence $\{x_n\}$ of elements in $X$ is a Cauchy sequence when for any $\epsilon > 0$ there exists $N \in 
\mathbb{N}$ such that for all $m, n \ge N$, $\norm{x_n - x_m} \le \epsilon$.
\end{defn}

\begin{defn}\label{s22d6}
A set $X$ is complete if every Cauchy sequence in $X$ converges to a point in $X$.
\end{defn}

\begin{defn}\label{s22d7}
A complete, normed linear space is called a Banach space.
\end{defn}

\begin{rem}
A linear space is the same as a vector space.
\end{rem}

\begin{defn}\label{s22d8}
Let $U \subset \mathbb{R}^n$ be an open set and let $0 < \gamma \le 1$. A function $u: U \rightarrow 
\mathbb{R}$ is said to be H\"{o}lder continuous with exponent $\gamma$ if there is a constant $C$ such that
\[
|u(x) - u(y)| \le C|x - y|^\gamma,
\]
for all $x, y \in U$.
\end{defn}

Unless stated otherwise $U \subset \mathbb{R}^n$ in what follows in this section.

\begin{defn}\label{s22e9}
Let $u: U \rightarrow \mathbb{R}$ be bounded and continuous then 
\[
\norm{u}_{C(\bar{U})} = \sup_{x \in U}|u(x)|.
\]
\end{defn}

\begin{rem}
The $C$ in the subscript on the left hand side indicates that we are dealing with continuous functions.
\end{rem}

\begin{rem}
If $\norm{u}_{C(\bar{U})}$ is finite then the function $u$ is bounded in $U$.
\end{rem}

\begin{defn}\label{s22e10}
The $\gamma$-th H\"{o}lder seminorm of $u: U \rightarrow \mathbb{R}$ is
\[
[u]_{C^{0, \gamma}(\bar{U})} = 
\sup_{x, y \in U, x \ne y}\left\{\frac{|u(x) - u(y)|}{|x - y|^\gamma}\right\}
\]
and the $\gamma$-th H\"{o}lder norm is 
\[
\norm{u}_{C^{0, \gamma}(\bar{U})} = \norm{u}_{C(\bar{U})} + [u]_{C^{0, \gamma}(\bar{U})}.
\]
\end{defn}

\begin{rem}
If a function $u$ has a finite $\gamma$-th H\"{o}lder seminorm then it is H\"{o}lder continuous with exponent
$\gamma$. If $u$ has a finite $\gamma$-th H\"{o}lder norm then it is bounded and H\"{o}lder continuous with 
exponent $\gamma$.
\end{rem}

\begin{defn}\label{s22e11}
The H\"{o}lder space $C^{k, \gamma}(\bar{U})$ consists of all functions $u \in C^k(\bar{U})$ for which
the norm 
\[
\norm{u}_{C^{k, \gamma}(\bar{U})} = \sum_{|\alpha| \le k}\norm{D^\alpha u}_{C(\bar{U})} + 
\sum_{|\alpha| \le k}[D^\alpha u]_{C^{0, \gamma}(\bar{U})}
\]
is finite.
\end{defn}

H\"{o}lder space $C^{k, \gamma}(\bar{U})$ has functions 
\begin{itemize}
\item That are continuously differentiable $k$ times.
\item All $k$ derivatives are bounded (first term on RHS) and H\"{o}lder continuous with exponent $\gamma$ (second
term on RHS).
\end{itemize}

\begin{prop}\label{s22p3}
The norm $\norm{u}_{C^{k, \gamma}(\bar{U})}$ satisfies the three properties in definition \ref{s22d2}.
\end{prop}
\begin{proof}
Since both terms of the norm are defined in terms of a supremum, it is clear that for $u, v \in 
C^{k, \gamma}(\bar{U})$,
\begin{eqnarray*}
u(x) &\le& \norm{u}_{C^{k, \gamma}(\bar{U})} \\
v(x) &\le& \norm{v}_{C^{k, \gamma}(\bar{U})}
\end{eqnarray*}
so that
\[
u(x) + v(x) \le \norm{u}_{C^{k, \gamma}(\bar{U})} + \norm{v}_{C^{k, \gamma}(\bar{U})}
\]
and hence
\[
\norm{u+v}_{C^{k, \gamma}(\bar{U})} \le \norm{u}_{C^{k, \gamma}(\bar{U})} + 
\norm{v}_{C^{k, \gamma}(\bar{U})}.
\]
Consider
\[
\norm{\lambda u}_{C^{k, \gamma}(\bar{U})} = \norm{\lambda u}_{C(\bar{U})} + 
[\lambda u]_{C^{0, \gamma}(\bar{U})} = \sup_{x\in SU}|\lambda u(x)| +
\sup_{x, y \in U, x \ne y}\left\{\frac{|\lambda u(x) - \lambda u(y)|}{|x - y|^\gamma}\right\}, 
\]
or
\[
\norm{\lambda u}_{C^{k, \gamma}(\bar{U})} = |\lambda|
\left(\sup_{x \in U}|u(x)| + \sup_{x, y \in U, x \ne y}\left\{\frac{|u(x) - u(y)|}{|x - y|^\gamma} \right\}\right) 
\]
The statement $\norm{u}_{C^{k, \gamma}(\bar{U})} = 0$ if and only if $u = 0$ follows immediately from the
definition of the norm.
\end{proof}

\begin{prop}\label{s22p4}
The H\"{o}lder space $C^{k, \gamma}(\bar{U})$ is complete with respect to the $\gamma$-th H\"{o}lder norm.
\end{prop}
\begin{proof}
If $\{u_n\}$ is a Cauchy sequence in $C^{k, \gamma}(\bar{U})$ then we have to show that it converges to a 
limit in the same set. Let $u_n(x) \rightarrow u(x)$ as $n \rightarrow \infty$. Since elements of
$C^{k, \gamma}(\bar{U})$ belong to the closed set $C^k(\bar{U})$, 
$\norm{D^\alpha u_n}_{C(\bar{U})} \rightarrow \norm{D^\alpha u}_{C(\bar{U})}$. Therefore,
\begin{equation}
\sum_{|\alpha| \le k}\norm{D^\alpha u_n}_{C(\bar{U})} \rightarrow 
\sum_{|\alpha| \le k}\norm{D^\alpha u}_{C(\bar{U})}
\end{equation}

Now fix an $\alpha$ such that $|\alpha| \le k$ and $x, y \in U$ such that $x \ne y$. Let $\gamma \in (0, 1]$,
 $D^\alpha u_m(x) = v_m(x)$ and $D^\alpha u(x) = v(x)$.
\[
\frac{|v(x) - v(y)|}{|x - y|^\gamma} = 
\frac{|v(x)- v_m(x) + v_m(x) - v_m(y) + v_m(y) - v(y)|}{|x - y|^\gamma}
\]
or
\[
\frac{|v(x) - v(y)|}{|x - y|^\gamma} \le
\frac{|v(x)- v_m(x)|}{|x - y|^\gamma} + \frac{|v_m(x)- v_m(y)|}{|x - y|^\gamma} + 
\frac{|v_m(y)- v(y)|}{|x - y|^\gamma}
\]
Since $v_m(x) \rightarrow v(x)$ for large enough $m$ the first and the third terms on the right hand side can be
made as small as possible. Further, the H\"{o}lder continuity of $v_m$ guarantees that the second term is less
than a constant independent of $x$ and $y$. Therefore,
\[
\frac{|v(x) - v(y)|}{|x - y|^\gamma} \le M_\alpha,
\]
where the subscript $\alpha$ reminds that $D^\alpha u_m(x) = v_m(x)$ and $D^\alpha u(x) = v(x)$. The above equation
tells that $D^\alpha u(x) = v(x)$ is H\"{o}lder continuous for chosen $\alpha$. Therefore, it is clear that
\[
[v]_{C^{0, \gamma}(\bar{U})} = 
\sup_{x, y \in U, x \ne y}\left\{\frac{|v(x) - v(y)|}{|x - y|^\gamma}\right\}
\]
exists and so does the sum of these for all $|\alpha| \le k$. Therefore the function $u$ is also a member of the
H\"{o}lder space.
\end{proof}

\begin{thm}\label{s22t1}
The space of functions $C^{k, \gamma}(\bar{U})$ is a Banach space.
\end{thm}
\begin{proof}
Follows immediately from propositions \ref{s22p3} and \ref{s22p4}.
\end{proof}

\section{Green's function and representation theorem}\label{s23}
\begin{thm}\label{s23t1}
Let $u$ be a harmonic function in $U$. Then
\begin{eqnarray*}
|u_{x_i}(x_0)| &\le& \frac{1}{V(n, 1)}\left(\frac{2}{r}\right)^n\norm{u}_{L^1(\partial B(x_0, r/2))} \\
 &\le& \frac{2n}{r}\norm{u}_{L^\infty(\partial B(x_0, r/2))}
\end{eqnarray*}
and $V(n, 1)$ is the volume of a unit $n$-ball given by equation \eqref{s16e10}.
\end{thm}
\begin{proof}
We will prove by induction on $k$. The case $k = 0$ follows immediately from the mean value theorem \ref{s21t3},
that is,
\begin{equation}\label{s23e1}
u(x_0) = \frac{1}{B(x_0, r/2)}\int u(x)dx
\end{equation}
where we use the more convenient notation $B(x_0, r/2)$ instead of $B_{r/2}(x_0)$. Since $u$ is harmonic, so is 
$u_{x_i}$ so that from \eqref{s23e1} we get
\[
u_{x_i}(x_0) = \frac{1}{B(x_0, r/2)}\int u_{x_i}(x)dx
\]
Using \eqref{s14e6},
\[
u_{x_i}(x_0) = \frac{1}{B(x_0, r/2)}\int un_i d\sigma(x) = \frac{2^n}{V(n, 1)r^n}\int un_i d\sigma(x).
\]
Therefore,
\[
|u_{x_i}(x_0)| \le \frac{2^n}{V(n, 1)r^n}\int |u|n_i d\sigma(x)
\]
so that
\[
|u_{x_i}(x_0)| \le \frac{1}{V(n, 1)}\left(\frac{2}{r}\right)^n\norm{u}_{L^1(\partial B(x_0, r/2))}.
\]
If instead we estimate the integral,
\[
|u_{x_i}(x_0)| \le \frac{2^n}{V(n, 1)r^n}\frac{nV(n, 1)r^{n-1}}{2^{n-1}}\norm{u}_{L^\infty(\partial B(x_0, r/2))},
\]
or
\[
|u_{x_i}(x_0)| \le \frac{2n}{r}\norm{u}_{L^\infty(\partial B(x_0, r/2))}.
\]
\end{proof}

\begin{thm}[Lioville's theorem]\label{s23t2}
If $u$ is a harmonic function defined on all $\mathbb{R}^n$ which is bounded above or below then $u$ is constant.
\end{thm}
\begin{proof}
Assume that $u$ is non-negative and bounded above. For $x, y \in \mathbb{R}^n$, let $r = R + d(x, y)$ where 
$d(x, y)$ is the euclidean distance between $x$ and $y$. Consider balls $B(x, R)$ and $B(y, r)$. Clearly, $B(x, r)$
is contained in $B(y, r)$ so that $B(x, R) \le B(y, r)$. Using the mean value theorem \ref{s21t3} at $x$ we have
\[
u(x) = \frac{1}{B(x, R)} \int_{B(x, R} u(z)dz \le \frac{1}{B(x, R)}\int_{B(y, r} u(z)dz.
\]
The last inequality comes from the fact that we are integrating over a larger volume but are still dividing it by
$B(x, R)$. Similarly,
\[
u(y) = \frac{1}{B(y, r)}\int_{B(y, r} u(z)dz.
\]
From the above two equations,
\[
u(x) \le \frac{B(y, r)}{B(x, R)}u(y) \le u(y).
\]
If we reverse the roles of $x$ and $y$ we will get $u(y) \le u(x)$ so that $u(x) = u(y)$.

If instead $u$ is non-positive and bounded below, we apply the same analysis for $-u$.
\end{proof}

\begin{proof}
Consider points $x, y \in \mathbb{R}^n$ and balls $B(x, r), B(y, r)$. For large enough $r$ the two balls will
coincide except for a small proportion of volume. If $u$ is bounded then the average over the two balls will be
within a small value of each other. By making $r$ large enough, since $u$ is bounded throughout $\mathbb{R}^n$,
we can make the two averages as close to each other as desired so that $u(x) = u(y)$.

This proof comes from \href{https://en.wikipedia.org/wiki/Harmonic_function#Liouville%27s_theorem}{Edward Nelson}.
\end{proof}

Theorem \ref{s20t1} showed that functions $u$ of the form $u(x) = (\Phi \star f)(x)$ is a solution of the Poisson's
equation if $f$ is a $C_c^2(\mathbb{R}^n)$ function. We will now show that the converse is also true.
\begin{thm}[Representation Formula]\label{s23t3}
If $f \in C^2_c(\mathbb{R}^n)$, $n \ge 3$, then any bounded solution of $-\Delta u = f$ in $\mathbb{R}^n$ has the
form
\[
u(x) = \int_{R^n}\Phi(x - y)f(y)dy + C,
\]
for some constant $C$ and $x \in \mathbb{R}^n$.
\end{thm}
\begin{proof}
Consider
\[
v(x) = \int_{R^n}\Phi(x - y)f(y)dy.
\]
Since $\Phi(x) \rightarrow 0$ as $|x| \rightarrow \infty$, $v$ is a bounded function. Further, theorem \ref{s20t1}
assures us that it solves the Poisson equation. Let if possible $u$ be any other bounded function. Then their
difference $w = u - v$ is a harmonic function which is also bounded. But Liouville's theorem \eqref{s23t2} assures
us that $w$ is a constant. Therefore, $u(x) = v(x) + C$.
\end{proof}

Consider an expression of the form
\[
\int_{V_\epsilon} \left(u(y)\Delta\Phi(y - x) - \Phi(y - x)\Delta u(y)\right)dy,
\]
where $V_\epsilon = U - B(x, \epsilon)$. The integral is over entire $U$ except for a small ball of 
radius $\epsilon$ around $x$. Using equation \eqref{s14e6},
\begin{eqnarray*}
\int_{V_\epsilon} \left(u(y)\Delta\Phi(y - x) - \Phi(y - x)\Delta u(y)\right)dy &=& 
\int_{\partial U} \left(u(y)\pd{\Phi}{n}(y - x) - \Phi(y - x)\pd{u}{n}(y)\right)d\sigma(y) + \\
 & & \int_{\partial B} \left(u(y)\pd{\Phi}{n}(y - x) - \Phi(y - x)\pd{u}{n}(y)\right)d\sigma(y)
\end{eqnarray*}
where the normal $n$ is appropriate to the surface of integration. In the domain $V_\epsilon$, $y$ is never equal 
to $x$. Therefore, $\Delta\Phi(y - x)$ always vanishes so that
\begin{eqnarray*}
-\int_{V_\epsilon} \Phi(y - x)\Delta u(y) dy &=& 
\int_{\partial U} \left(u(y)\pd{\Phi}{n}(y - x) - \Phi(y - x)\pd{u}{n}(y)\right)d\sigma(y) + \\
 & & \int_{\partial B} \left(u(y)\pd{\Phi}{n}(y - x) - \Phi(y - x)\pd{u}{n}(y)\right)d\sigma(y)
\end{eqnarray*}
Consider the first of the integrals around $\partial B$, that is
\[
I_1 = \int_{\partial B} u(y)\pd{\Phi}{n}(y - x)d\sigma(y).
\]
Since
\[
\pd{\Phi}{n} = n\cdot D\Phi = -\frac{y - x}{|y - x|} \cdot \frac{-1}{nV(n, 1)}\frac{y - x}{|y - x|^{n+1}} = 
\frac{1}{nV(n, 1)\epsilon^n} = \frac{1}{|B(x, \epsilon)|}
\]
\[
I_1 = \frac{1}{|B(x, \epsilon)|}\int_{\partial B} u(y)d\sigma(y) = u(x),
\]
where we used the mean value theorem to get the last equality. Now consider the modulus of
\[
I_2 = \int_{\partial B}\Phi(y - x)\pd{u}{n}(y)d\sigma(y)
\]
that is
\[
|I_2| \le \max|\Phi| \int_{\partial B}\pd{u}{n}(y)d\sigma(y) = \max|\Phi|\int_B udy = 
(\max|\Phi|) u(x) nV(n, 1)\epsilon^n,
\]
where we once again used the mean value theorem to get the last equality. As $\epsilon \rightarrow 0$, $|I_2|
\rightarrow 0$. Thus, in the limit $\epsilon \rightarrow 0$, we have
\[
-\int_{V_\epsilon} \Phi(y - x)\Delta u(y) dy = 
\int_{\partial U} \left(u(y)\pd{\Phi}{n}(y - x) - \Phi(y - x)\pd{u}{n}(y)\right)d\sigma(y) + u(x)
\]
Since $\Delta u = -f$ in $U$ and $u = g$ on $\partial U$,
\[
\int_{V_\epsilon} \Phi(y - x)f(y) dy = 
\int_{\partial U} \left(g(y)\pd{\Phi}{n}(y - x) - \Phi(y - x)\pd{u}{n}(y)\right)d\sigma(y) + u(x)
\]
or
\begin{equation}\label{s23e2}
u(x) = \int_U \Phi(y - x)f(y) dy + \int_{\partial U}\Phi(y - x)\pd{u}{n}(y)d\sigma(y)
- \int_{\partial U} g(y)\pd{\Phi}{n}(y - x) d\sigma(y),
\end{equation}
where we also used the fact that as $\epsilon \rightarrow 0$, $V_\epsilon \rightarrow U$. Although this equation 
is true it is of no use in solving the boundary value problem because we are not prescribed the normal derivative
of $u$ on $\partial U$. This is where Green's functions come into picture.

Now consider the boundary-value problem $\Delta\phi^x = 0$ in $U$ and $\phi^x = \Phi(y - x)$ on $\partial U$. Then,
applying Green's theorem for $u$ and $\phi^x$,
\[
\int_U(u\Delta\phi^x - \phi^x\Delta u)dy = 
\int_{\partial U}\left(u\pd{\phi^x}{n} - \phi^x\pd{u}{n}\right)d\sigma(y).
\]
Using the boundary value problem for $\phi^x$, 
\[
-\int_U \phi^x\Delta u dy = \int_{\partial U}\left(u\pd{\phi^x}{n} - \Phi(y - x)\pd{u}{n}\right)d\sigma(y).
\]
Using the fact that $\Delta u = -f$ and rearranging,
\begin{equation}\label{s23e3}
\int_{\partial U} \Phi(y - x)\pd{u}{n} d\sigma(y) = 
\int_{\partial U} g(y)\pd{\phi^x}{n} d\sigma(y) - \int_U \phi^x f(y) dy
\end{equation}
Combining equations \eqref{s23e2} and \eqref{s23e3},
\begin{equation}\label{s23e4}
u(x) = \int_U G(x, y)f(y)dy - \int_{\partial U} g(y)\pd{G}{n}d\sigma(y),
\end{equation}
where
\begin{equation}\label{s23e5}
G(x, y) = \Phi(y - x) - \phi^x(y)
\end{equation}
is called the Green's function.

{\color{red}\noindent To do:

Exercises:
\begin{enumerate}
\item Show that the Green's function is symmetric, that is, $G(x, y) = G(y, x)$.
\item If $K(x, y)$ is the Poisson kernel then show that $K \in C^\infty(\mathbb{R}^n)$, $\Delta_x K = 0$ in
$\mathbb{R}^n_+$ and 
\[
\int_{\mathbb{R}^n_+}K(x, y)dy = 1.
\]
\item Theorem 14 from Evans' book.
\end{enumerate}
}

\section{Perron's theorem}\label{s24}

\section{Sobolev space}\label{s25}

\section{Weak solutions of elliptic equations}\label{s26}

\section{The heat equation}\label{s27}
We will consider the heat equation of the form $u_t - \Delta u = 0$ where $u: U \times [0, \infty) \rightarrow
\mathbb{R}$, $U$ being a domain in $\mathbb{R}^n$. We first observe that
\begin{prop}\label{s27p1}
If $u(x, t)$ is a solution of the heat equation then so is $u(\lambda x, \lambda^2 t)$ for any $\lambda \in 
\mathbb{R}$.
\end{prop}
\begin{proof}
Let $v = u(\lambda x, \lambda^2 t)$. Then $v_t = \lambda^2 u_t$ and $\Delta u = \lambda^2 \Delta x$.
\end{proof}

This suggests that $|x|^2/t$ is a constant for the heat equation. We therefore examine solutions of the form
\begin{equation}\label{s27e1}
u(x, t) = w(t)v\left(\frac{|x|^2}{t}\right).
\end{equation}
so that
\begin{eqnarray*}
u_t &=& w^\prime(t)v\left(\frac{|x|^2}{t}\right) - w(t)v^\prime\left(\frac{|x|^2}{t}\right)\frac{|x|^2}{t^2} \\
u_{x_i} &=& w(t)v^\prime\left(\frac{|x|^2}{t}\right)\frac{2x_i}{t} \\
u_{x_ix_i}&=& w(t)v^{\prime\prime}\left(\frac{|x|^2}{t}\right)\frac{4x_i^2}{t^2} + 
 w(t)v^\prime\left(\frac{|x|^2}{t}\right)\frac{2}{t} \\
\Delta u &=& w(t)v^{\prime\prime}\left(\frac{|x|^2}{t}\right)\frac{4|x|^2}{t^2} + 
 w(t)v^\prime\left(\frac{|x|^2}{t}\right)\frac{2n}{t}
\end{eqnarray*}
and the heat equation becomes
\begin{equation}\label{s27e2}
w^\prime(t)v\left(\frac{|x|^2}{t}\right) - w(t)v^\prime\left(\frac{|x|^2}{t}\right)\frac{2n}{t} - 
w(t)\frac{|x|^2}{t^2}\left(v^\prime + 4v^{\prime\prime}\right) = 0.
\end{equation}
We now choose $v$ such that $v^\prime + 4v^{\prime\prime} = 0 \Rightarrow v + 4v^\prime + c_1 = 0$. If 
$z = |x|^2/t$ then $\ln|v + c_1| = -z/4$ or that $v = \exp(-z/4) - c_1$. We therefore choose
\begin{equation}\label{s27e3}
v\left(\frac{|x|^2}{t}\right) = \exp\left(-\frac{|x|^2}{4t}\right).
\end{equation}
Substituting \eqref{s27e3} into \eqref{s27e2} we get
\[
\left(w^\prime(t) + \frac{n}{2t}w(t)\right)\exp\left(-\frac{|x|^2}{4t}\right) = 0
\]
which is possible only if $w^\prime(t) + nw(t)/(2t) = 0$ or $w(t) = c_2t^{-n/2}$. Thus, the solution \eqref{s27e1}
becomes
\begin{equation}\label{s27e4}
u(x, t) = \frac{c}{t^{n/2}}\exp\left(-\frac{|x|^2}{4t}\right).
\end{equation}
where $c$ is a constant. We choose $c$ such that
\begin{equation}\label{s27e5}
\int_{\mathbb{R}^n}u(x, t)dx = 1
\end{equation}
and denote the resulting solution by $\phi(x, t)$. Now,
\[
\frac{c}{t^{n/2}}\int_{\mathbb{R}^n}\exp\left(-\frac{|x|^2}{4t}\right)dx = 
\frac{c}{t^{n/2}}\prod_{i=1}^n\int_{-\infty}^\infty \exp\left(-\frac{x_i^2}{4t}\right)dx_i = 
\frac{c}{t^{n/2}}(4\pi t)^{n/2}
\]
so that the normalisation condition \eqref{s27e5} gives $c = (4\pi)^{-n/2}$ or that the fundamental solution is
\begin{equation}\label{s27e6}
\Phi(x, t) = \frac{1}{(4\pi t)^{n/2}}\exp\left(-\frac{|x|^2}{4t}\right).
\end{equation}

\begin{prop}\label{s27p2}
If $\Phi(x, t)$ given by \eqref{s27e6} is a solution of the heat equation so is $\Phi(x - y)$ for a fixed $y \in
\mathbb{R}^n$.
\end{prop}
\begin{proof}
\[
\Phi(x - y, t) = \frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right).
\]
\[
\Phi_{x_j} = 
\frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right)\left(-\frac{(x_j - y_j)}{2t}\right)
\]
\[
\begin{split}
\Phi_{x_jx_j} = 
\frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right)
\left(\frac{(x_j - y_j)^2}{2t^2}\right) - \\
\frac{1}{2t}\frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right).
\end{split}
\]
\[
\begin{split}
\Delta\Phi = 
\frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right)
\left(\frac{|x - y|^2}{2t^2}\right) - \\                 
\frac{n}{2t}\frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right).
\end{split}
\]
\[
\begin{split}
\Phi_t = \frac{n}{2t}\frac{1}{(4\pi t)^{n/2}}\prod_{i=1}^n\exp\left(-\frac{(x_i - y_i)^2}{4t}\right) \\
- \frac{1}{(4\pi t)^{n/2}}\exp\left(-\frac{|x - y|^2}{4t}\right)\left(\frac{|x - y|^2}{2t^2}\right)
\end{split}
\]
so that
\[
\Phi_t - \Delta\Phi = 0.
\]
\end{proof}

We will use the fundamental solution $\Phi$ to get solve the initial value problem 
\begin{equation}\label{s27e7}
\begin{split}
u_t - \Delta u &= 0 \text{ in } \mathbb{R}^n \times (0, \infty) \\
u &= g \text{ on } \mathbb{R}^n \times \{t = 0\}
\end{split}
\end{equation}
Carrying forward the analogy with Laplace's equation we will check if 
\begin{equation}\label{s27e8}
u(x, t) = \Phi \star g
\end{equation}
solves \eqref{s27e7}. To that end, we prove
\begin{thm}\label{s27t1}
Assume that $g \in C^\infty(\mathbb{R}^n) \cap L^\infty(\mathbb{R}^n)$ and let $u$ be defined by \eqref{s27e8}. 
Then
\begin{enumerate}
\item $u \in C^\infty(\mathbb{R}^n \times (0, \infty))$,
\item $u_t - \Delta u = 0$ for all $x \in \mathbb{R}^n, t > 0$,
\item For all $x \in \mathbb{R}^n, t > 0$,
\[
\lim_{(x, t) \rightarrow (x_0, 0)}u(x, t) = g(x_0, 0),
\]
for each point $x_0 \in \mathbb{R}^n$.
\end{enumerate}
\end{thm}
\begin{proof}
From equations \eqref{s27e6} and \eqref{s27e8},
\begin{equation}\label{s27e9}
u(x, t) = \frac{1}{(4\pi n)^{t/2}}\int_{\mathbb{R}^n} \exp\left(-\frac{|x - y|^2}{4t}\right)g(y)dy.
\end{equation}
The function
\[
\frac{1}{(4\pi n)^{t/2}}\exp\left(-\frac{|x - y|^2}{4t}\right)g(y)
\]
is infinitely differentiable with respect to both $x$ and $t$. Therefore, so is $u$.

The proof for $u_t - \Delta u = 0$ is very similar to the proof of proposition \ref{s27p2}.

We now prove claim (iii) of the theorem. Since $g$ is continuous, we can fix an $x_0 \in \mathbb{R}^n$ and an
$\epsilon > 0$ to find $\delta > 0$ such that $|x_0 - y| < \delta \Rightarrow |g(x_0) - g(y)| < \epsilon$. Now
choose an $x \in \mathbb{R}^n$ such that $|x - x_0| < \delta/2$. For such an $x$, consider
\begin{equation}\label{s27e10}
|u(x, t) - g(x_0)| = \Big|\int_{\mathbb{R}^n}\Phi(x - y, t)g(y)dy - g(x_0)\Big|.
\end{equation}
Since $\Phi$ is normalised, we can write the term $g(x_0)$ on the right hand side as
\[
\int_{\mathbb{R}^n}\Phi(x - y, t)g(x_0)dy
\]
so that
\[
|u(x, t) - g(x_0)| \le \int_{\mathbb{R}^n}\Big|\Phi(x - y, t)(g(y) - g(x_0))\Big|dy.
\]
We now split the domain of the integral into a ball of radius $\delta$ around $x_0$ and the rest. Thus,
\begin{equation}\label{s27e11}
|u(x, t) - g(x_0)| = I + J
\end{equation}
where
\begin{eqnarray*}
I &=& \int_{B(x_0, \delta)}\Big|\Phi(x - y, t)(g(y) - g(x_0))\Big|dy \\ 
J &=& \int_{B^c(x_0, \delta)}\Big|\Phi(x - y, t)(g(y) - g(x_0))\Big|dy,
\end{eqnarray*}
and $B^c(x_0, \delta) = \mathbb{R}^n - B(x_0, \delta)$ is the complement of the ball. In the domain of the $I$,
$|g(y) - g(x_0)| < \epsilon$ so that
\[
I \le \epsilon \int_{B(x_0, \delta)}\Big|\Phi(x - y, t))\Big|dy < \epsilon\int_{\mathbb{R}^n}\Phi(x - y, t)dy,
\]
as $\Phi$ takes positive values throughout $\mathbb{R}^n$. Since $\Phi$ is normalised,
\begin{equation}\label{s27e12}
|x - x_0| < \delta/2 \Rightarrow I < \epsilon.
\end{equation}
Now,
\[
J = \int_{B^c(x_0, \delta)}\Big|\Phi(x - y, t)(g(y) - g(x_0))\Big|dy \le
2\norm{g}_{L^\infty}\int_{B^c(x_0, \delta)} \Phi(x - y, t)dy.
\]
The factor of $2$ on the right hand side comes because there are two $g$'s in the integrand. We can pull out the
factor with $t^{-n/2}$ outside the integral so that
\[
J \le \frac{C}{t^{n/2}}\int_{B^c(x_0, \delta)}\exp\left(-\frac{|x - y|^2}{4t}\right)dy,
\]
for some constant $C$. Since
\[
|y - x_0| = |y - x + x - x_0| \le |y - x| + |x - x_0| < |y - x| + |y - x| \Rightarrow \frac{1}{2}|y-x_0| < |y-x|.
\]
we have
\[
\frac{|y - x_0|^2}{4} < |y - x|^2 \Rightarrow -\frac{|y - x_0|^2}{4} > -|y - x|^2 \Rightarrow
\exp\left(-\frac{|y - x_0|^2}{16t}\right) > \exp\left(-\frac{|x - y|^2}{4t}\right)
\]
and hence
\[
J \le \frac{C}{t^{n/2}}\int_{B^c(x_0, \delta)}\exp\left(-\frac{|y - x_0|^2}{16t}\right)dy
\]
In the domain of the $J$, $y$ is guaranteed to be not equal to $x_0$. Therefore, we can consider a limit $t 
\rightarrow 0$ in which the integrand tends to zero and hence
\begin{equation}\label{s27e13}
|x - x_0| < \frac{\delta}{2}, t \rightarrow 0 \Rightarrow J \rightarrow 0.
\end{equation}
Claim (iii) follows immediately from \eqref{s27e11}, \eqref{s27e12} and \eqref{s27e13}.
\end{proof}

We will now illustrate how to find a solution of the non-homogeneous heat conduction equation in one space
dimension. The problem to be solved is
\begin{eqnarray}
u_t - u_{xx} &=& f(x, t) \label{s27e14} \\
u(x, 0) &=& 0 \nonumber,
\end{eqnarray}
where $x \in \mathbb{R}$ and $t \in (0, \infty)$. We construct the solution to this problem using the solution
of the problem
\begin{eqnarray}
v_t - v_{xx} &=& 0 \label{s27e15} \\
v(x, 0) = f(x, \tau) \nonumber,
\end{eqnarray}
where $x \in \mathbb{R}$ and $0 < \tau < t$. Using theorem \ref{s27t1}, we know that
\[
v(x, t; \tau) = 
\frac{1}{\sqrt{4\pi(t - \tau)}}\int_{\mathbb{R}}\exp\left(-\frac{(x - y)^2}{4(t - \tau)}\right)f(y, \tau)dy.
\]
We now show that
\begin{equation}\label{s27e16}
u(x, t) = \int_0^t v(x, t; \tau) d\tau
\end{equation}
solves \eqref{s27e14}. We will need the proposition,
\begin{prop}\label{s27p3} If
\[
g(t) = \int_{\alpha(t)}^{\beta(t)} f(s, t)ds
\]
then
\[
g^\prime(t) = \int_{\alpha(t)}^{\beta(t)} \td{f}{t}ds + f(t, \beta(t))\beta^\prime(t) - 
f(t, \alpha(t))\alpha^\prime(t).
\]
\end{prop}
\begin{proof}
\[
g^\prime(t) = \lim_{h \rightarrow 0}\frac{1}{h}
\left(\int_{\alpha(t+h)}^{\beta(t+h)} f(s, t+h)ds - \int_{\alpha(t)}^{\beta(t)} f(s, t)ds\right)
\]
Write the first integral as 
\[
\int_{\alpha(t)}^{\beta(t)} f(s, t+h)ds + \int_{\beta(t)}^{\beta(t+h)} f(s, t+h)ds - 
\int_{\alpha(t)}^{\alpha(t+h)} f(s, t+h)ds
\]
from which the result follows immediately.
\end{proof}
Using proposition \ref{s27p3} to find derivative of $u(x, t)$ given by \eqref{s27e16} with respect to $t$, we get
\[
u_t(x, t) = \int_0^t v_t(x, t; \tau) d\tau + v(x, t; t)
\]
Using the initial value of the problem \eqref{s27e15}, we have
\begin{equation}\label{s27e17}
u_t(x, t) = \int_0^t v_t(x, t; \tau) d\tau + F(x, t).
\end{equation}
We also have
\[
u_{xx}(x, t) = \int_0^t v_{xx}(x, t; \tau)
\]
so that
\[
u_t - u_{xx} = \int_0^t(v_t- v_{xx})d\tau + F(x,t) = F(x, t)
\]
by virtue of \eqref{s27e15}. Further, $u(x, 0) = 0$ because the limits of the integral coincide. This is 
\emph{Duhamel's principle} for the heat equation. We will now prove it for arbitrary dimensions and with more
rigour.

\begin{thm}\label{s27t2}
A solution of the inhomogeneous heat equation
\begin{eqnarray*}
u_t - \Delta u &=& f \text{ in } \mathbb{R}^n \times (0, \infty) \\
u(x, 0) &=& 0 \text{ on } \mathbb{R}^n \times {0} 
\end{eqnarray*}
is 
\begin{equation}\label{s27e18}
u(x, t) = \int_0^t\int_{\mathbb{R}^n}\Phi(x - y,t - s)f(y, s)dyds,
\end{equation}
where $\Phi$ is the fundamental solution given by \eqref{s27e6}. $u$ has the following properties:
\begin{enumerate}
\item $u \in C^{2, 1}(\mathbb{R}^n \times (0, \infty))$.
\item $u_t - \Delta u = f$ for all $x \in \mathbb{R}^n, t > 0$.
\item For all $x_0 \in \mathbb{R}^n$,
\[
\lim_{(x,t)\rightarrow(x_0,0)}u(x, t) = 0.
\]
\end{enumerate}
\end{thm}
\begin{proof}
In order to prove the first claim we have to integrate $u$ with respect to $t$. The fundamental solution $\Phi$ has
a singularity at $t = 0$. Therefore we cannot interchange the integral and the derivative. Let us introduce the 
variables $y \mapsto x - y^\prime, s \mapsto t - s^\prime$ so that $dy = -dy^\prime, ds = -ds^\prime$ and
\[
u(x, t) = -\int_t^0\int_{\infty}^{-\infty}\cdots\int_{\infty}^{-\infty}\Phi(y^\prime, s^\prime)f(x - y^\prime,
t - s^\prime)dy^\prime ds^\prime
\]
or
\[
u(x,t) = \int_0^t\int_{\mathbb{R}^n}\Phi(y^\prime, s^\prime)f(x - y^\prime, t - s^\prime)dy^\prime ds^\prime
= \int_0^t\int_{\mathbb{R}^n}\Phi(y, s)f(x - y, t - s)dyds.
\]
In this form the $t$-dependence is transferred from $\Phi$ to $f$ and hence we can interchange the integral and
the derivative with respect to $t$. It is now easy to see that $u \in C^{2, 1}(\mathbb{R}^n \times (0, \infty))$.

Let us now prove the second claim. We first calculate the partial derivative of $u$ with respect to $t$.
\[
u_t = \int_0^t\int_{\mathbb{R}^n}\Phi(y, s)\frac{\partial}{\partial t}f(x - y, t - s)dyds +
\int_0^t\int_{\mathbb{R}^n}\Phi(y, t)f(x - y, 0)dy + 0,
\]
where we have used proposition \ref{s27p3}. Similarly,
\[
\Delta_x u = \int_0^t\int_{\mathbb{R}^n}\Phi(y, s)\Delta_x f(x - y, t - s)dyds,
\]
where we have used the subscript to indicate that the laplacian is with respect to $x$. There are no additional
terms because $\Phi \rightarrow 0$ as $x_i \rightarrow \pm\infty$ for all $i = 1, \ldots, n$. Thus,
\begin{eqnarray*}
u_t - \Delta_x u &=& 
\int_0^t\int_{\mathbb{R}^n}\Phi(y, s)\left(\frac{\partial}{\partial t} - \Delta_x\right)f(x - y, t - s)dyds + \\
& & \int_0^t\int_{\mathbb{R}^n}\Phi(y, t)f(x - y, 0)dy
\end{eqnarray*}

Since $\Phi(y, s)$ has a singularity at $s = 0$, we split the domain
$[0, t]$ of the first integral as $[0, \epsilon]$ and $[\epsilon, t]$. Thus,
\begin{eqnarray*}
u_t - \Delta_x u &=& 
\int_0^\epsilon\int_{\mathbb{R}^n}\Phi(y, s)\left(\frac{\partial}{\partial t}-\Delta_x\right)f(x - y, t - s)dyds+\\
& & 
\int_\epsilon^t\int_{\mathbb{R}^n}\Phi(y, s)\left(\frac{\partial}{\partial t}-\Delta_x\right)f(x - y, t - s)dyds+\\
& & \int_0^t\int_{\mathbb{R}^n}\Phi(y, s)f(x - y, 0)dy \\
&:=& I + J + K
\end{eqnarray*}
Now,
\begin{eqnarray*}
I &\le& (\norm{f_t}_{L^\infty} + \norm{D^2f}_{L^\infty})\int_0^\epsilon\int_{\mathbb{R}^n}\Phi(y, s)dyds \\
&\le& C\int_0^\epsilon\int_{\mathbb{R}^n}\Phi(y, 0)dyds = C\epsilon\int_{\mathbb{R}^n}\Phi(y, 0)dy = C\epsilon,
\end{eqnarray*}
where $C = (\norm{f_t}_{L^\infty} + \norm{D^2f}_{L^\infty})$. Thus,
\begin{equation}\label{s27e19}
\lim_{\epsilon\rightarrow 0} I = 0.
\end{equation}
We can write the integral $J$ as
\[
J = 
-\int_\epsilon^t\int_{\mathbb{R}^n}\Phi(y, s)\left(\frac{\partial}{\partial s}-\Delta_y\right)f(x-y, t-s)dyds
\]
so that we can use integration by parts and let the differential operator act on $\Phi$. Thus,
\begin{eqnarray*}
J &=& 
\int_\epsilon^t\int_{\mathbb{R}^n}f(x-y, t-s)\left(\frac{\partial}{\partial s}-\Delta_y\right)\Phi(y, s)dyds + \\
& & \int_{\mathbb{R}^n}\Phi(y, \epsilon)f(x-y,t-\epsilon)dy - \\
& & \int_{\mathbb{R}^n}\Phi(y,t)f(x-y,0)dy.
\end{eqnarray*}
The boundary terms of the $y$ integral vanish because $\Phi \rightarrow 0$ as $x_i \rightarrow \pm\infty$ for all
$i = 1, \ldots, n$. The first term on the right hand side is zero because $\Phi$ solves the heat equation. The
third term is identical to $K$ so that
\begin{equation}\label{s27e20}
J = \int_{\mathbb{R}^n}\Phi(y, \epsilon)f(x-y,t-\epsilon)dy - K
\end{equation}
so that using equations \eqref{s27e19} and \eqref{s27e20} we get
\[
u_t - \Delta u = \lim_{\epsilon\rightarrow 0}\int_{\mathbb{R}^n}\Phi(y, \epsilon)f(x-y,t-\epsilon)dy.
\]
Under a transformation $y \mapsto x - z$ with a fixed $x$, the integral in the above equation becomes
\[
u_t - \Delta u = \lim_{\epsilon\rightarrow 0}\int_{\mathbb{R}^n}\Phi(x - z, \epsilon)f(z,t-\epsilon)dz.
\]
{\color{red}This part of the proof is weak. In the limit $\epsilon \rightarrow 0$ the most dominant part of the
integrand will be the one in which $x - z$ is very close to zero or $f(z, t - \epsilon)$ is very close to 
$f(x, t)$. We can the pull $f(x, t)$ out of the integral and evaluate the rest to $1$ following the normalization
property of $\Phi$.} Therefore,
\[
u_t - \Delta u = f(x, t).
\]

Lastly to prove the third claim, we note that
\begin{eqnarray*}
\norm{u(x, t)}_{L^\infty} &=& \int_0^t\int_{\mathbb{R}^n}\Phi(x - y,t - s)f(y, s)dyds \\
&=& \int_0^t\int_{\mathbb{R}^n}\Phi(y, s)f(x - y, t - s)dyds \\
&=& \norm{f}_{L^\infty}\int_0^t\int_{\mathbb{R}^n}\Phi(y, s)dyds \\
&=& \norm{f}_{L^\infty}\int_0^tds = t\norm{f}_{L^\infty}
\end{eqnarray*}
so that
\[
\lim_{t \rightarrow \infty}u(x, t) = 0.
\]
\end{proof}

Let $U$ be the domain over which heat equation is defined. Then the \emph{parabolic cylinder} is the set
\begin{equation}\label{s27e21}
U_T := U \times (0, T].
\end{equation}
The parabolic boundary of $U_T$ is
\begin{equation}\label{s27e22}
\Gamma_T := \bar{U}_T - U_T.
\end{equation}
Since $U_T$ includes the points $U \times {T}$, they are excluded from $\Gamma_T$. If we imagine $U$ to be closed
curve then $U_T$ is a cylinder with base $U \times {0}$. The set $\Gamma_T$ thus has the base of this cylinder and
its sides but not its top. Let us now define the set
\begin{defn}
For fixed $x \in \mathbb{R}^n, t \in \mathbb{R}, r > 0$ we define the heat ball
\[
E(x, t; r) = \{(y, s) \in \mathbb{R}^{n+1}: s \le t, \Phi(x - y, t - s) \ge r^{-n}\}
\]
\end{defn}
The boundary of $E(x, t; r)$ is the set $\{(y, s) \in \mathbb{R}^{n+1}; s \le t, \Phi(x - y,t - s) = r^{-n}$. Note 
that $r$ and $n$ are fixed, the latter being the spatial dimensions of the domain. The equation $\Phi(x - y, t - s)
= r^{-n}$ is a level set of $\Phi$. The rearranged form of the equation $\Phi(x - y, t - s)r^n = 1$ can also be
interpreted as a product of $\Phi$ and volume of a $n$-ball being constant. Thus $\Phi$ is some sort of density
although we do not know of what quantity.

\begin{prop}\label{s27p4}
The origin of $\mathbb{R}^{n+1}$ is a member of the heat ball $E(x, t; r)$.
\end{prop}
\begin{proof}
For a fixed $r$ and $n$, we can always find $x$ and $t > 0$ such that $\Phi(x, t) \ge r^{-n}$ because of continuity
of $\Phi$.
\end{proof}

The boundary of $E(0, 0; r)$ is $\Phi(-y, -s) = r^{-n}$ or $\ln\Phi(-y, -s) = -n\ln r$ or
\[
-\frac{n}{2}\ln(-4\pi s) - \frac{|y|^2}{4(-s)} = -n \ln r
\]
or
\[
-\frac{n}{2}\ln(-4\pi s) + \frac{|y|^2}{4s} + n\ln r = 0.
\]
Let us define a function
\begin{equation}\label{s27e23}
\psi = -\frac{n}{2}\ln(-4\pi s) + \frac{|y|^2}{4s} + n\ln r
\end{equation}
so that the boundary of $E(0, 0; r)$ contains $\psi = 0$.

We now prove the unintuitive result
\begin{prop}\label{s27p5}
For $y \in \mathbb{R}^n, s \le 0$, if we denote $E(0, 0; 1)$ as $E(1)$,
\[
\iint_{E(1)}\frac{|y|^2}{s^2}dyds = 4.
\]
\end{prop}
\begin{proof}
In order to carry out the integration, we need the limits of the integral which, in turn, follow from the boundary
of $E(1)$. Recall that
\[
E(1) = \{(y, s) \in \mathbb{R}^{n+1} : s \le 0, \Phi(-y, -s) \ge 1\}.
\]
This definition gives the upper limit of $s$. We get the lower limit from the condition $\Phi(-y, -s) \ge 1$ or
\[
\frac{1}{(-4\pi s)^{n/2}}\exp\left(-\frac{|y|^2}{4(-s)}\right) \ge 1 \Rightarrow \exp\left(\frac{|y|^2}{4s}\right) 
\ge (-4\pi s)^{n/2}.
\]
Now $\inf\{\exp(x): x > 0\} = 1$ so that the above condition implies
\[
1 \ge (-4\pi s)^{n/2} \Rightarrow 1^{2/n} \ge -4\pi s \Rightarrow -s \le \frac{1}{4\pi} \Rightarrow s \ge 
-\frac{1}{4\pi}
\] 
Thus the limits of the $s$ integral are from $-1/4\pi$ to $0$. Let us now find the limits on $y$.
\[
\frac{1}{(-4\pi s)^{n/2}}\exp\left(\frac{-|y|^2}{4(-s)}\right) \ge 1 \Rightarrow -\frac{n}{2}\ln(-4\pi s) -
\frac{|y|^2}{4(-s)} \ge 0 \Rightarrow |y|^2 \le 2ns\ln(-4\pi s).
\]
Since $|y|^2$ is non-negative, the limits of the $y$ integral are of the form
\[
0 \le |y|^2 \le 2ns\ln(-4\pi s).
\]
Thus,
\[
\iint_{E(1)}\frac{|y|^2}{s^2}dyds = \int_{-1/4\pi}^0\int_{|y|^2 \le 2ns\ln(-4\pi s)}\frac{|y|^2}{s^2}dyds.
\]
Since the limit of the $y$ integral contains $s$, we must first solve it. The domain of the $y$ integral is the
$n$-sphere of radius $\sqrt{2ns\ln(-4\pi s)}$. We note that the integrand is a function of $|y|^2$, that is, it 
has a radial symmetry. Therefore, we use spherical polar coordinates to solve the $y$ integral. Let us define
\begin{equation}\label{s27e24}
\rho_0 = 2ns\ln(-4\pi s)
\end{equation}
and write $dy = dS(\rho)d\rho$, where $dS(\rho)$ is the surface measure. Thus,
\[
\int_{|y|^2 \le 2ns\ln(-4\pi s)}\frac{|y|^2}{s^2}dy = 
\int_0^{\rho_0}\int_{\partial B(0, \rho_0)} \rho^2 dS(\rho)d\rho.
\]
From equation \eqref{s16e11},
\[
\int_{\partial B(0, \rho_0)} dS(\rho) = \frac{n\pi^{n/2}}{\Gamma(1 + n/2)}\rho^{n-1}
\]
so that
\begin{equation}\label{s27e25}
\int_{|y|^2 \le 2ns\ln(-4\pi s)}\frac{|y|^2}{s^2}dy = \alpha_n\int_0^{\rho_0} \rho^{n+1}d\rho = 
\alpha_n\frac{\rho_0^{n+2}}{n+2}
\end{equation}
where
\begin{equation}\label{s27e26}
\alpha_n = \frac{n\pi^{n/2}}{\Gamma(1 + n/2)}
\end{equation}
Thus,
\[
\iint_{E(1)}\frac{|y|^2}{s^2}dyds = \frac{\alpha_n}{n+2}\int_{-1/4\pi}^0\frac{(2ns\ln(-4\pi s))^{(n+2)/2}}{s^2}ds
\]
or
\begin{equation}\label{s27e27}
\iint_{E(1)}\frac{|y|^2}{s^2}dyds = \frac{\alpha_n(2n)^{(n+2)/2}}{n+2} I
\end{equation}
where
\[
I = \int_{-1/4\pi}^0\frac{(s\ln(-4\pi s))^{(n+2)/2}}{s^2}ds
\]
Let us introduce the variable $u = -4\pi s$ so that the limits of integral go from $1$ to $0$ and
\[
I = \frac{1}{2^n\pi^{n/2}}\int_0^1\frac{(-u\ln(u))^{(n+2)/2}}{u^2}du
\]
Now let $v = -\ln u$ so that the limits of integral go from $\infty$ to $0$. $dv = -du/u$ or $e^{v}dv = -du$ and
\[
I = \frac{1}{2^n\pi^{n/2}}\int_{\infty}^0 \frac{(e^{-v}v)^{(n+2)/2}}{e^{-2v}}e^{-v}(-dv)
= \frac{1}{2^n\pi^{n/2}}\int_0^\infty v^{(n+2)/2}e^{-vn/2}dv
\]
Lastly, let $w = vn/2$ so that
\[
I = \frac{1}{2^n\pi^{n/2}}\int_0^\infty \left(\frac{2}{n}\right)^{(n+2)/2}w^{(n+2)/2}e^{-w}\frac{2}{n}dw
\]
or
\[
I = \frac{1}{2^n\pi^{n/2}}\left(\frac{2}{n}\right)^{(n+4)/2}\int_0^\infty w^{1 + n/2}e^{-w}dw
\]
or
\[
I = \frac{1}{2^n\pi^{n/2}}\left(\frac{2}{n}\right)^{(n+4)/2}\Gamma\left(2 + \frac{n}{2}\right) =
\]
or
\begin{equation}\label{s27e28}
I =  
\frac{1}{2^n\pi^{n/2}}\left(\frac{2}{n}\right)^{(n+4)/2}\left(1 + \frac{n}{2}\right)
\Gamma\left(1 + \frac{n}{2}\right)
\end{equation}
From equations \eqref{s27e26}, \eqref{s27e27} and \eqref{s27e28}, we have
\[
\iint_{E(1)}\frac{|y|^2}{s^2}dyds = 
\frac{1}{2^n\pi^{n/2}}
\frac{n\pi^{n/2}}{\Gamma(1 + n/2)}\frac{(2n)^{(n+2)/2}}{n+2}\left(\frac{2}{n}\right)^{(n+4)/2}
\left(\frac{n+2}{2}\right)\Gamma\left(1 + \frac{n}{2}\right)
\]
Let us first cancel the most obvious factors so that
\[
\iint_{E(1)}\frac{|y|^2}{s^2}dyds =
\frac{n}{2^n}2^{(n+2)/2}n^{(n+2)/2}\left(\frac{2}{n}\right)^{(n+4)/2}\frac{1}{2} = 4.
\]
\end{proof}

We now have enough results to prove
\begin{thm}[Mean value theorem]\label{s27t3}
Let $u \in C^{1, 2}(U_T)$, where $U_T$ is a parabolic cylinder, solve the heat equation. Then,
\[
u(x, t) = \frac{1}{4r^n}\iint_{E(x,t;r)} u(y, s)\frac{|x - y|^2}{(t - s)^2}dyds
\]
for each $E(x,t;r) \subset U_T$.
\end{thm}
\begin{proof}
Shift the coordinate axes so that $x = 0$ and $t = 0$. Let $E(0, 0; r)$ be denoted by $E(r)$ and let
\[
\phi(r) = \frac{1}{r^n}\iint_{E(r)} u(y, s)\frac{|y|^2}{s^2}dyds.
\]
Change the variables of integration as $y = r\bar{y}$ and $s = r^2\bar{s}$. Then $dyds = r^{n+2}d\bar{y} 
d\bar{s}$. Further, $E(r)$ is transformed to
\[
\left\{(\bar{y}, \bar{s}) \in \mathbb{R}^{n+1}: \bar{s} \le \frac{t}{r^2}, 
\Phi\left(-r\bar{y}, -r^2\bar{s}\right) \ge \frac{1}{r^n}\right\}
\]
Using the definition of $\Phi$,
\[
\Phi\left(-r\bar{y}, -r^2\bar{s}\right) = \frac{1}{r^n}\Phi(-\bar{y}, -\bar{s})
\]
so that the region $E(r)$ is transformed to 
\[
\left\{(\bar{y}, \bar{s}) \in \mathbb{R}^{n+1}: \bar{s} \le \frac{t}{r^2}, 
\Phi(-\bar{y}, -\bar{s}) \ge \right\} = E(1).
\]
Therefore,
\[
\phi(r) = \frac{1}{r^n}\iint_{E(1)} u(r\bar{y}, r^2\bar{s})\frac{|\bar{y}|^2}{{\bar{s}}^2} r^{n+2}
d\bar{y} d\bar{s}.
\]
or,
\[
\phi(r) = \iint_{E(1)} u(r\bar{y}, r^2\bar{s})\frac{|\bar{y}|^2}{{\bar{s}}^2}d\bar{y} d\bar{s}
\]
Now that the domain of integration is independent of $r$, we can interchange the integral and the derivative to
get
\[
\phi^\prime(r) = \iint_{E(1)}\left(\sum_{i=1}^n\pd{u}{\bar{y}_i}\bar{y}_i +
\pd{u}{\bar{s}}(2r\bar{s})\right)\frac{|\bar{y}|^2}{\bar{s}^2}d\bar{y} d\bar{s}
\]
Reverse the transformation so that
\[
\phi^\prime(r) = \frac{1}{r^{n+1}}\iint_{E(1)}\left(\sum_{i=1}^n\pd{u}{y_i}y_i +
2s\pd{u}{s}\right)\frac{|y|^2}{s^2}dy ds := A + B.
\]
From \eqref{s27e23},
\[
\pd{\psi}{y_i} = \frac{y_i}{s}
\]
so that
\[
\sum_{i=1}^n y_i\pd{\psi}{y_i} = \frac{|y|^2}{s}
\]
and hence
\[
B = \frac{4}{r^{n+1}}\sum_{i=1}^n \iint_{E(1)}\pd{u}{s} y_i\pd{\psi}{y_i} dyds.
\]
Integrating by parts,
\[
B = \frac{4}{r^{n+1}}\sum_{i=1}^n\left(\int_{E(1)} \left[u_sy_i\psi\right]_{\partial E(1)} ds -
\iint_{E(1)} \psi\left(y_i\frac{\partial^2u}{\partial s\partial y_i} + \pd{u}{s}\right)dyds\right)
\]
The first term vanishes because $\psi$ is zero on $E(1)$. The second term in the second integral is independent
of $i$ so that summing it is just the same term multiplied by $n$.
\[
B = 
-\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\pd{u}{s}dyds - \frac{4}{r^{n+1}}\sum_{i=1}^n
\iint_{E(1)} y_i\psi\frac{\partial^2u}{\partial s\partial y_i} dyds.
\]
Integrating the second term by parts with respect to $s$,
\begin{eqnarray*}
B &=& -\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\pd{u}{s}dyds - \\
 & & \frac{4}{r^{n+1}}\sum_{i=1}^n\left(\int_{E(1)} \left[y_i\psi\pd{u}{y_i}
\right]_{\partial E(1)}dy - \iint_{E(1)} y_i\pd{\psi}{s}\pd{u}{y_i} dyds\right).
\end{eqnarray*}
Since $\psi$ vanishes on the $\partial E(1)$,
\[
B = -\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\pd{u}{s}dyds + 
\frac{4}{r^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\pd{\psi}{s}\pd{u}{y_i} dyds
\]
From equation \eqref{s27e23},
\[
\pd{\psi}{s} = -\frac{n}{2s} - \frac{|y|^2}{4s^2},
\]
so that, 
\[
B = -\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\pd{u}{s}dyds - 
\frac{4}{r^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\left(\frac{n}{2s} + \frac{|y|^2}{4s^2}\right)\pd{u}{y_i} dyds
\]
or
\begin{eqnarray*}
B &=& -\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\pd{u}{s}dyds - 
\frac{2n}{sr^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\pd{u}{y_i}dyds - \\
& & \frac{1}{r^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\frac{|y|^2}{s^2}\pd{u}{y_i} dyds
\end{eqnarray*}
The last integral is $A$ so that
\[
A + B = -\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\pd{u}{s}dyds - 
\frac{2n}{sr^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\pd{u}{y_i}dyds
\]
Since $u$ solves the heat equation,
\[
A + B = -\frac{4}{r^{n+1}}\iint_{E(1)}n\psi\Delta u dyds - 
\frac{2n}{sr^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\pd{u}{y_i}dyds
\]
Integrating the first term by parts with respect to $y$,
\begin{eqnarray*}
A + B &=& -\frac{4}{r^{n+1}}\left(\sum_{i=1}^n\int_{E(1)} \left[n\psi\pd{u}{y_i}\right]_{\partial E(1)}ds 
+ \sum_{i=1}^n\iint_{E(1)}n\pd{\psi}{y_i}\pd{u}{y_i} dyds\right) - \\ 
& & \frac{2n}{sr^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\pd{u}{y_i}dyds
\end{eqnarray*}
The first term vanishes because $\psi = 0$ on the boundary. In the second term we substitute
\[
\pd{\psi}{y_i} = \frac{y_i}{2s}
\]
to get
\[
A + B = -\frac{2n}{sr^{n+1}}\sum_{i=1}^n\iint_{E(1)}y_i\pd{u}{y_i} dyds + 
\frac{2n}{sr^{n+1}}\sum_{i=1}^n\iint_{E(1)} y_i\pd{u}{y_i}dyds = 0
\]
That is $\phi^\prime(r) = 0$ or that $\phi(r)$ is a constant. Therefore,
\[
\phi(r) = \lim_{u \rightarrow 0}\phi(u) = 
\lim_{\rho \rightarrow 0}\frac{1}{\rho^n}\iint_{E(\rho)} u(y, s)\frac{|y|^2}{s^2}dyds.
\]
Since the domain of integration depends on $\rho$ we introduce the transformation $y = \rho\bar{y}, 
s = \rho^2\bar{s}$ to
get
\[
\phi(r) = \lim_{\rho \rightarrow 0} \iint_{E(1)} u(\rho y, \rho^2 s)\frac{|\bar{y}|^2}{\bar{s}^2}d\bar{y}d\bar{s}
= u(0, 0)\iint_{E(1)} \frac{|\bar{y}|^2}{\bar{s}^2}d\bar{y}d\bar{s}
\]
Using proposition \ref{s27e5}, we get
\[
\phi(r) = 4u(0, 0).
\]
Therefore, from the definition of $\phi$,
\[
u(0, 0) = \phi(r) = \frac{1}{4r^n}\iint_{E(r)} u(y, s)\frac{|y|^2}{s^2}dyds.
\]
Shifting the origin back to its original position,
\[
u(x, t) = \frac{1}{4r^n}\iint_{E(r)} u(y, s)\frac{|x - y|^2}{(t - s)^2}dyds.
\]
\end{proof}

\section{Wave equation}\label{s28}
The general form of the wave equation is $\Box u = 0$, where the d'Alembertian operator $\Box = \partial_{tt} - 
\Delta$. We will work in units that make the wave velocity $1$. Unlike Laplace equation and the heat equation,
the method of solution of the wave equation depends on whether the number of spatial dimensions is odd or even.
Therefore, we shall start with the simplest case, 
\begin{equation}\label{s28e1}
u_{tt} - u_{xx} = 0.
\end{equation}
Let us introduce new coordinates
\begin{eqnarray*}
\xi &=& x - t \\
\eta &=& x + t.
\end{eqnarray*}
Inverting them, we get
\begin{eqnarray*}
x &=& \frac{1}{2}(\eta + \xi) \\
t &=& \frac{1}{2}(\eta - \xi)
\end{eqnarray*}
so that
\begin{eqnarray*}
u_x &=& u_{\xi} + u_{\eta} \\
u_t &=& -u_{\xi} + u_{\eta} \\
u_{xx} &=& u_{\xi\xi} + 2u_{\xi\eta} + u_{\eta\eta} \\
u_{tt} &=& u_{\xi\xi} - 2u_{\xi\eta} + u_{\eta\eta} \\
\Box u &=& 4u_{\xi\eta}
\end{eqnarray*}
and hence the wave equation in $(\xi, \eta)$ coordinates is $u_{\xi\eta} = 0$. Its solution is $u(\xi, \eta) = 
g(\xi) + h(\eta)$ or, reverting to $(x, t)$ coordinates
\begin{equation}\label{s28e2}
u(x, t) = g(x - t) + h(x + t).
\end{equation}
Consider the diagram in figure \ref{s28f1}. The sides $AB$ and $CD$ are along the curves $x - t = c_1$, a constant
and $x + t = c_2$, another constant. Therefore,
\begin{eqnarray*}
x_A - t_A &=& x_D - t_D \\
x_B - t_B &=& x_C - t_C \\
x_A + t_A &=& x_B + t_B \\
x_D + t_D &=& x_C + t_C
\end{eqnarray*}
so that $u(A) = u(x_A, t_A) = g(x_A - t_A) + h(x_A + t_A)$, $u(C) = g(x_C - t_C) + h(x_C + t_C)$ and hence
\begin{eqnarray}
u(A) + u(C) &=& g(x_A - t_A) + h(x_A + t_A) + g(x_C - t_C) + h(x_C + t_C) \nonumber \\
 &=& g(x_D - t_D) + h(x_B + t_B) + g(x_B - t_B) + h(x_D + t_D) \nonumber \\
 &=& g(x_B - t_B) + h(x_B + t_B) + g(x_D - t_D) + h(x_D + t_D) \nonumber \\
 &=& u(B) + u(D). \label{s28e3}
\end{eqnarray}
This equation is called as the \emph{parallelogram law}.
\begin{figure}
\centering
\includegraphics[scale=0.5]{s28_f1}
\caption{The parallelogram law.}
\label{s28f1}
\end{figure}

We will now solve the initial value problem $u_{tt} - u_{xx} = 0$ such that $u(x 0) = \varphi(x)$ and 
$u_t(x, 0) = \psi(x)$, where $\varphi \in C^2(\mathbb{R})$ and $\psi \in C^1(\mathbb{R})$. From equation 
\eqref{s28e2}, 
\begin{eqnarray}
u(x, 0) &=& g(x) + h(x) = \varphi(x) \label{s28e4} \\
u_t(x, 0) &=& -g^\prime(x) + h^\prime(x)  = \psi(x) \label{s28e5}
\end{eqnarray}
so that
\begin{eqnarray}
h^\prime(x) &=& \frac{1}{2}(\varphi^\prime(x) + \psi(x)) \label{s28e6} \\
g^\prime(x) &=& \frac{1}{2}(\varphi^\prime(x) - \psi(x)) \label{s28e7}
\end{eqnarray}
Integrating the above equations, we get
\begin{equation}\label{s28e8}
h(x) = \frac{1}{2}\left(\varphi(x) + \int_0^x\psi(s)ds\right) + c,
\end{equation}
where $c$ is a constant of integration. Substituting this equation in \eqref{s28e4} we get
\begin{equation}\label{s28e9}
g(x) = \frac{1}{2}\left(\varphi(x) - \int_0^x\psi(s)ds\right) - c.
\end{equation}
Using equations \eqref{s28e8} and \eqref{s28e9} in \eqref{s28e2}, we get
\[
u(x) = \frac{1}{2}\left(\varphi(x + t) + \int_0^{x+t}\psi(s)ds\right) + c +
\frac{1}{2}\left(\varphi(x - t) - \int_0^{x-t}\psi(s)ds\right) - c
\]
or
\begin{equation}\label{s28e10}
u(x) = \frac{1}{2}(\varphi(x+t) + \varphi(x-t)) + \frac{1}{2}\int_{x-t}^{x+t}\psi(s)ds.
\end{equation}
Equation \eqref{s28e10} is called the d'Alembert's solution of the one dimensional wave equation.

\begin{defn}\label{s28d1}
The spherical mean of a function $f : \mathbb{R}^n \rightarrow \mathbb{R}$ at a point $x \in \mathbb{R}^n$ is
defined as
\[
F(x; r) = \frac{1}{S(r, n)}\fint_{\partial B(x, r)} f(y)dS(y)
\]
where $S(r, n)$ is the surface area of the ball $B(x, r)$ or
\[
F_1(x; r) = \frac{1}{V(r, n)}\fint_{B(x, r)} f(y)dy
\]
\end{defn}

We will now show 
\begin{prop}\label{s28p1}
Let $u$ be a solution of the initial value problem $u_{tt} - u_{xx} = 0$ subject to the conditions $u = g, u_t = h$
on $\mathbb{R}^n \times \{t = 0\}$. Define
\begin{equation}\label{s28e11}
U(x; r, t) = \fint_{\partial B(x,r)}u(y, t)dS(y),
\end{equation}
where $x \in \mathbb{R}^n, r > 0, t > 0$. Then $U \in C^m(\bar{\mathbb{R}}_+ \times [0, \infty))$ and $U$ 
satisfies the Euler-Poisson-Darboux equation 
\begin{equation}\label{s28e12}
U_{tt} - U_{rr} - \frac{n - 1}{r}U_r = 0
\end{equation} 
in $\mathbb{R}_+ \times (0, \infty)$ and $U = G, U_t = H$ on $\mathbb{R}_+ \times \{t = 0\}$, where $G$ and $H$
are defined as
\begin{eqnarray}
G(x; r, t) &=& \fint_{\partial B(x,r)}g(y, t)dS(y) \label{s28e13}\\
H(x; r, t) &=& \fint_{\partial B(x,r)}h(y, t)dS(y) \label{s28e14}
\end{eqnarray}
\end{prop}
\begin{proof}
Let us first differentiate $U$ with respect to $r$. Since the domain of integration depends on $r$, we tranform the
problem so that it is over a ball of radius $1$ centred at the origin. Let $z = (y - x)/r$ or $y = x + rz$. Then
$dS(y) = r^{n-1}dS(z)$ and
\begin{eqnarray*}
U(x; r, t) &=& \fint_{\partial B(x,r)} u(y,t)dS(y) \\
 &=& \frac{1}{S(n, r)}\int_{\partial B(x,r)} u(y, t)dS(y) \\
 &=& \frac{1}{S(n, r)}\int_{\partial B(0,1)} u(x + rz, t)r^{n-1} dS(z) \\
 &=& \frac{1}{S(n, 1)}\int_{\partial B(0,1)} u(x + rz, t) dS(z) \\
 &=& \fint_{\partial B(0,1)} u(x + rz, t)dS(z)
\end{eqnarray*}
so that
\begin{eqnarray*}
U_r(x; r, t) &=& \fint_{\partial B(0, 1)}Du(x + rz; r, t)zdS(z) \\
 &=& \fint_{\partial B(x, r)} Du(y)\frac{y - x}{r}dS(y)
\end{eqnarray*}
The vector $(y - x)/r$ is a unit normal to the surface so that
\[
U_r(x; r, t) = \fint_{\partial B(x, r)} \pd{u}{n}dS(y) = \frac{1}{S(r, n)}\int \Delta u dy 
= \frac{r}{n}\frac{1}{V(r, n)}\int\Delta u dy
\]
or
\begin{equation}\label{s28e15}
U_r(x; r, t) = \frac{r}{n}\fint_{B(x,r)} \Delta u dy.
\end{equation}
We now proceed to find the second derivative of $U$ with respect to $r$. Before we do so, let us write $U_r$ in
full as
\[
U_r(x; r, t) = \frac{r}{n}\frac{1}{\alpha(n)r^n}\int_{B(x, r)}\Delta u dy,
\]
where we wrote $V(r, n) = \alpha(n)r^n$, $\alpha(n) = \pi^{n/2}/\Gamma(1 + n/2)$. Thus,
\begin{equation}\label{s28e16}
U_r(x; r, t) = \frac{1}{n\alpha(n)r^{n-1}}\int_{B(x, r)}\Delta u dy.
\end{equation}
Let us write the volume integral on the right hand side as
\[
U_r(x; r, t) = \frac{1}{n\alpha(n)r^{n-1}}\int_0^r d\rho\int_{\partial B(x,\rho)} \Delta u dy
\]
Now,
\[
\frac{1}{n\alpha(n)\rho^{n-1}}\int_{\partial B(x,\rho)} \Delta u dy = \fint \Delta u(y)dy
\]
so that
\[
r^{n-1}U_r(x; r, t) = \int_0^r \rho^{n-1} d\rho \fint_{\partial B(x,\rho)} \Delta u(y)dS(y)
\]
Differentiating with respect to $r$,
\[
(n-1)r^{n-2}U_r + r^{n-1}U_{rr} = 
\lim_{h\rightarrow 0} \frac{1}{h} \int_r^{r+h} \rho^{n-1} \fint_{\partial B(x,\rho)} \Delta u(y)dS(y)d\rho
\]
or
\[
(n-1)r^{n-2}U_r + r^{n-1}U_{rr} = r^{n-1} \fint_{\partial B(x,r)} \Delta u(y)dS(y)
\]
or
\[
(n-1)\frac{U_r}{r} + U_{rr} = \fint_{\partial B(x,r)} \Delta u(y)dS(y)
\]
Using \eqref{s28e15}, we get
\begin{equation}\label{s28e17}
U_{rr}(x;r,t) = \left(\frac{1}{n} - 1\right)\fint_{B(x,r)} \Delta u dy + \fint_{\partial B(x,r)} \Delta u(y)dS(y).
\end{equation}
In order to find out $U_{tt}$, let us start from \eqref{s28e16} and rearrange it as
\[
r^{n-1}U_r = \frac{1}{n\alpha(n)}\int_{B(x,r)}\Delta u dy
\]
Differentiating with respect to $r$, we get
\[
(n-1)r^{n-2}U_r + r^{n-1}U_{rr} = \frac{1}{n\alpha(n)}\frac{\partial}{\partial r}\int_{B(x,r)}\Delta u dy
\]
or
\[
(n-1)r^{n-2}U_r + r^{n-1}U_{rr} = \frac{1}{n\alpha(n)}\int_{\partial B(x,r)}\Delta u dS(y)
\]
Since $u$ solves the wave equation
\[
(n-1)r^{n-2}U_r + r^{n-1}U_{rr} = \frac{r^{n-1}}{n\alpha(n)r^{n-1}}\int_{\partial B(x,r)}u_{tt} dS(y)
\]
or
\begin{equation}\label{s28e18}
(n-1)r^{n-2}U_r + r^{n-1}U_{rr} = r^{n-1}\fint_{\partial B(x,r)}u_{tt}dS(y) = r^{n-1}U_{tt}
\end{equation}
Dividing both sides by $r^{n-1}$ we get the Euler-Poisson-Darboux equation.
\end{proof}

Before proceeding to use the Euler-Poisson-Darboux equation let us solve the wave equation for the half plane.
\begin{eqnarray}
u_{tt} - u_{xx} &=& 0  \text{ in } \mathbb{R}_+ \times (0, \infty) \label{s28e19} \\
u &=& g  \text{ on } \mathbb{R}_+ \times \{t = 0\} \label{s28e20} \\
u_t &=& h \text{ on } \mathbb{R}_+ \times \{t = 0\} \label{s28e21} \\
u &=& 0 \text{ on } \{x = 0\} \times (0, \infty) \label{s28e22}.
\end{eqnarray}
Let us extend functions $u, g, h$ to entire $\mathbb{R}$ by odd reflection. That is, define
\begin{equation}\label{s28e23}
\tilde{u}(x, t) = \begin{cases}
u(x, t) & (x \ge 0, t \ge 0) \\
-u(-x, t) & (x \le 0, t \ge 0)
\end{cases}
\end{equation}
\begin{equation}\label{s28e24}
\tilde{g}(x) = \begin{cases}
g(x) & (x \ge 0) \\
-g(-x) & (x \le 0)
\end{cases}
\end{equation}
\begin{equation}\label{s28e25}
\tilde{h}(x) = \begin{cases}
h(x) & (x \ge 0) \\
-h(-x) & (x \le 0)
\end{cases}
\end{equation}
Why do we need an odd extension? Because we want to satisfy \eqref{s28e22}. If we had considered an even extension 
then we would have had $u(x, t) = u(-x, t)$ at $x = 0$ but that would not have guaranteed us $u = 0$ on $x = 0$. 
If, instead, we were working with Neumann boundary condition $u_x = 0$ on $\{x = 0\} \times (0, \infty)$ we would 
have been forced to consider even extension.

Using the definitions \eqref{s28e24} to \eqref{s28e25} it is easy to conclude that the $\tilde{u}$ is a solution of 
the initial value problem,
\begin{eqnarray}
\tilde{u}_{xx} &=& \tilde{u}_{tt} \text{ in } \mathbb{R} \times (0, \infty) \label{s28e26} \\
\tilde{u} &=& \tilde{g} \text{ on } \mathbb{R} \times \{t = 0\} \label{s28e27} \\
\tilde{u} &=& \tilde{h} \text{ on } \mathbb{R} \times \{t = 0\} \label{s28e28}.
\end{eqnarray}
Note that after extending $u$ to the entire real line, we do not need the special condition at $x = 0$. Our choice 
of odd extension automatically gives it to us. We thus observe that $\tilde{u}$ satisfies the boundary value 
condition for which we have found d'Alembert's solution. Thus, using \eqref{s28e10} we get
\begin{equation}\label{s28e29}
\tilde{u}(x, t) = \frac{1}{2}(\tilde{g}(x+t) + \tilde{g}(x-t)) + \frac{1}{2}\int_{x-t}^{x+t}\tilde{h}(s)ds.
\end{equation}
If $x > t > 0$ then \eqref{s28e29}, written in terms of $u$ becomes
\begin{equation}\label{s28e30}
u(x, t) = \frac{1}{2}(g(x+t) + g(x-t)) + \frac{1}{2}\int_{x-t}^{x+t}h(s)ds.
\end{equation}
If $x < t$, $\tilde{g}(x-t) = -g(t-x)$ and $\tilde{h}(x-t) = -h(t-x)$ so that \eqref{s28e29} becomes
\begin{eqnarray*}
u(x,t) &=& \frac{1}{2}(g(x+t) - g(t-x)) + \frac{1}{2}\int_{x-t}^{x+t}\tilde{h}(s)ds \\
 &=& \frac{1}{2}(g(x+t) - g(t-x)) + \frac{1}{2}\int_{x-t}^{0}\tilde{h}(s)ds + 
 \frac{1}{2}\int_{0}^{x+t}\tilde{h}(s)ds
\end{eqnarray*}
In the term on the right hand side, we put $\tilde{h}(s) = -h(-s)$ and in the third term we put $\tilde{h}(s) =
h(s)$ so that 
\[
u(x,t) = \frac{1}{2}(g(x+t) - g(t-x))-\frac{1}{2}\int_{x-t}^{0}h(-s)ds+\frac{1}{2}\int_{0}^{x+t}h(s)ds.
\]
Using the transformation $s \mapsto -s$ in the second term we get
\[
u(x,t) = \frac{1}{2}(g(x+t)-g(t-x))+\frac{1}{2}\int_{t-x}^{0}h(-s)ds+\frac{1}{2}\int_{0}^{x+t}h(s)ds
\]
or
\[
u(x,t) = \frac{1}{2}(g(x+t)-g(t-x))-\frac{1}{2}\int_{0}^{t-x}h(-s)ds+\frac{1}{2}\int_{0}^{x+t}h(s)ds
\]
or
\begin{equation}\label{s28e31}
u(x,t) = \frac{1}{2}(g(x+t)-g(t-x)) + \frac{1}{2}\int_{t-x}^{x+t}h(s)ds
\end{equation}

We will now solve the Euler-Poisson-Darboux equation for $n=3$. We recall that the $U$ in it is related to the 
solution of the initial value problem $u_{tt} - \Delta u = 0$ in $\mathbb{R}^n \times (0, \infty)$ subject to the 
conditions $u = g$ and $_t = h$ on $\mathbb{R}^n \times \{t=0\}$. Let us consider the transformation $\tilde{U} := 
rU, \tilde{G} := rG, \tilde{H} := rH$, where $U, G, H$ are defined in equations \eqref{s28e11}, \eqref{s28e13} and 
\eqref{s28e14}. Differentiating $\tilde{U}$ with respect to $r$ twice we get
\begin{eqnarray*}
\tilde{U}_r &=& U + rU_r \\
\tilde{U}_{rr} &=& 2U_r + rU_{rr}
\end{eqnarray*}
Substituting in \eqref{s28e12} for $n=3$ we get $\tilde{U}_{tt} - \tilde{U}_{rr} = 0$ in $\mathbb{R}_+ \times 
(0, \infty)$. Since $U(x; r, 0) = G(x; r, 0)$ (because $u(\cdot, 0) = g(\cdot)$), we have $\tilde{U} = \tilde{G}$ 
on $\mathbb{R}_+ \times \{t=0\}$. For similar reason, we also have $\tilde{U}_t = \tilde{H}$ on $\mathbb{R}_+ 
\times\{t=0\}$. Finally, when $r = 0$, $\tilde{U}$ vanishes. Thus, $\tilde{U}$ solves the initial value problem
\begin{eqnarray*}
\tilde{U}_{tt} - \tilde{U}_{rr} &=& 0 \text{ in } \mathbb{R}_+ (0, \infty) \\
\tilde{U} &=& \tilde{G} \text{ on } \mathbb{R}_+ \times \{t=0\} \\
\tilde{U}_t &=& \tilde{H} \text{ on } \mathbb{R}_+ \times \{t=0\} \\
\tilde{U} &=& 0 \text{ on } \{r = 0\} \times (0, \infty)
\end{eqnarray*}
Comparing these equations with \eqref{s28e19} to \eqref{s28e22}, we see that $\tilde{U}$ is a solution of the one-
dimensional wave equation in the half-plane and therefore its solution for $0 \le r \le t$, by analogy with 
\eqref{s28e31} is
\begin{equation}\label{s28e32}
\tilde{U}(x; r, t) = \frac{1}{2}(\tilde{G}(t+r)-\tilde{G}(t-r)) + \frac{1}{2}\int_{t-r}^{t+r}\tilde{H}(s)ds
\end{equation}

Since $U(x; r, t)$ is the average of $u$ at $x$ over a ball of radius $r$, in the limit $r$ tends to $0$, 
$U(x; r,t)$ will become $u(x)$, that is
\begin{equation}\label{s28e33}
\lim_{r \rightarrow 0^+}U(x; r, t) = u(x, t)
\end{equation}
so that\begin{equation}\label{s28e34}
\lim_{r \rightarrow 0^+}\frac{\tilde{U}(x; r, t)}{r} = u(x, t).
\end{equation}
Using \eqref{s28e32} in \eqref{s28e34} we get
\[
u(x,t) = \frac{1}{2}\lim_{r \rightarrow 0^+}\frac{\tilde{G}(t+r)-\tilde{G}(t-r)}{r} + 
\lim_{r \rightarrow 0^+}\left(\frac{1}{2r}\int_{t-r}^{t+r}\tilde{H}(s)ds\right)
\]
or
\[
u(x,t) = \tilde{G}^\prime(x;t) + \tilde{H}(x;t)
\]
Since $\tilde{G} = tG$ and $\tilde{H} = tH$,
\begin{equation}\label{s28e35}
u(x,t) = \frac{\partial}{\partial t}\left(t\fint_{\partial B(x,t)}g(y, t)dS(y)\right) + 
t\fint_{\partial B(x,t)}h(y, t)dS(y)
\end{equation}
Since
\begin{equation}\label{s28e36}
\fint_{\partial B(x,t)}g(y, t)dS(y) = \fint_{\partial B(0,1)}g(x+tz, t)dS(z),
\end{equation}
the first term on the right hand side of \eqref{s28e36} is
\begin{eqnarray*}
\frac{\partial}{\partial t}\left(t\fint_{\partial B(0,1)}g(x+tz, t)dS(z)\right) &=& 
\fint_{\partial B(0,1)}g(x+tz, t)dS(z) \\
& & + t\fint_{\partial B(0,1)}zDg(x+tz, t)dS(z)
\end{eqnarray*}
Reverting back to $y$, and substituting in \eqref{s28e35}, we get
\[
u(x,t) = \fint_{\partial B(x,t)}g(y, t)dS(y) + \fint_{\partial B(0,1)} Dg(y)\cdot(y-x)dS(y) + 
t\fint_{\partial B(x,t)}h(y, t)dS(y)
\]
or
\begin{equation}\label{s28e37}
u(x,t) = \fint_{\partial B(x,t)} (th(y) + g(y) + Dg\cdot(y-x))dS(y).
\end{equation}
Equation \eqref{s28e37} is called \emph{Kirchhoff's formula} for the solution of the initial-value problem.

There is no transformation that can transform Euler-Poisson-Darboux equation for $n=2$ into a one-dimensional wave 
equation. Jacque Hadamard devised an ingenious technique, called the \emph{method of descent}, that uses the 
solution for $n=3$ dimensions to $n=2$. Let $u(x_1, x_2, t)$ solve $u_{tt} - \Delta u = 0$ in $\mathbb{R}^2 \times 
(0, \infty)$ subject to $u = g$ and $u_t = h$ on $\mathbb{R}^2 \times \{t=0\}$. Define $\tilde{u}: \mathbb{R}^3 
\times (0, \infty) \mapsto \mathbb{R}$ such that
\begin{equation}\label{s28e38}
\tilde{u}(x_1, x_2, x_3, t) := u(x_1, x_2, t).
\end{equation}
From the above definition it is clear that $\tilde{u}_{tt} = u_{tt}, \tilde{u}_{x_1x_1} = u_{x_1x_1}, 
\tilde{u}_{x_2x_2} = u_{x_2x_2}$. Further, $\tilde{u}_{x_3}$ and $\tilde{u}_{x_3x_3}$ are both zero. Therefore, if 
$u$ is a solution of the two dimensional wave equation then $\tilde{u}_{tt} - \Delta\tilde{u} = 0$ in $\mathbb{R}^3 
\times (0, \infty)$ and $\tilde{u} = \tilde{g}, \tilde{u}_t = \tilde{h}$ on $\mathbb{R}^3 \times \{t = 0\}$. Thus, 
$\tilde{u}$ is a solution of three dimensional wave equation and we can write it using \eqref{s28e35} as
\[
\tilde{u}(x,t) = \frac{\partial}{\partial t}\left(t\fint_{\partial B(x,t)}\tilde{g}(y, t)dS(y)\right) + 
t\fint_{\partial B(x,t)}\tilde{h}(y, t)dS(y)
\]
In particular, if we choose $x_3=0$ and write $\bar{x} = (x_1, x_2, 0)$ then
\begin{equation}
u(x_1, x_2, t) = \frac{\partial}{\partial t}\left(t\fint_{\partial B(\bar{x},t)}\tilde{g}(y_1, y_2, 0, t)
d\bar{S}(\bar{y})\right) + t\fint_{\partial B(\bar{x},t)}\tilde{h}(y_1,y_2,0, t)d\bar{S}(\bar{y}),
\end{equation}
where $d\bar{S}(\bar{y})$ is the two-dimensional surface measure on the ball centred at $\bar{x}$ and with radius 
$t$. We now transform the surface integral to a ``volume" integral on a disc centred around $\bar{x}$. A point 
$(y_1, y_2)$ on the disc maps to a point $f(y_1, y_2) = (y_1, y_2, \sqrt{t^2 - (x_1 - y_1)^2 - (x_1 - y_2)^2})$. We 
will compute the area element on the sphere in terms of the ``volume" element on the disc. It is the cross product 
$D_1f \times D_2f$
\[
\left(1, 0, \frac{x_1 - y_1}{\sqrt{t^2 - (x_1 - y_1)^2 - (x_1 - y_2)^2}}\right) \times 
\left(0, 1, \frac{x_1 - y_1}{\sqrt{t^2 - (x_1 - y_1)^2 - (x_1 - y_2)^2}}\right)
\]
or
\[
\left(-\frac{(x_1 - y_1)}{\sqrt{t^2 - |x - y|^2}}, -\frac{(x_2 - y_2)}{\sqrt{t^2 - |x - y|^2}}, 1\right)
\]
The magnitude of the cross product is
\[
|D_1f \times D_2f| = \left(1 + \frac{|x - y|}{\sqrt{t^2 - |x - y|^2}}\right)^{1/2} = \frac{t}{\sqrt{t^2 - 
|x - y|^2}}
\]
so that
\[
d\bar{S}(\bar{y}) = \frac{t}{\sqrt{t^2 - |x - y|^2}}dy_1dy_2 = \frac{t}{\sqrt{t^2 - |x - y|^2}}dy
\]
However, since both hemispheres map to the same disc, we must have
\begin{equation}\label{s28e40}
d\bar{S}(\bar{y}) = \frac{2t}{\sqrt{t^2 - |x - y|^2}}dy_1dy_2 = \frac{2t}{\sqrt{t^2 - |x - y|^2}}dy
\end{equation}
As an aside, we note that if $\gamma(y) = (t^2 - |y - x|^2)^{1/2}$ then we observe that
\[
|D_1f \times D_2f| = (1 + |D\gamma(y)|^2)^{1/2}.
\]
Coming back to \eqref{s28e40},
\[
\fint_{\partial B(\bar{x},t)}\bar{g}d\bar{S}(\bar{y}) = 
\frac{1}{4\pi t^2}\int_{\partial\bar{B}(\bar{x},t)}\bar{g}d\bar{S}(\bar{y}) = 
\frac{1}{2\pi t}\int_{B(x,t)}\frac{g(y)}{\sqrt{t^2 - |x - y|^2}}dy,
\]
where we recall that $x = (x_1, x_2)$ and $y = (y_1, y_2)$. Since the ``volume" of the ball $B(x,t)=\pi t^2$, we
can write the last integral as an average to get
\begin{equation}\label{s28e41}
\fint_{\partial B(\bar{x},t)}\bar{g}d\bar{S}(\bar{y}) = \frac{t}{2}\fint_{B(x,t)} \frac{g(y)}{\sqrt{t^2 - 
|x - y|^2}}dy.
\end{equation}
Using equation \eqref{s28e41} and the corresponding equation for $h$ in \eqref{s28e35}, we get
\begin{equation}\label{s28e42}
u(x,t)=\frac{\partial}{\partial t}\left(\frac{t^2}{2}\fint_{B(x,t)} \frac{g(y)}{\sqrt{t^2 - |x - y|^2}}dy\right) + 
\frac{t^2}{2}\fint_{B(0,1)} \frac{h(x)}{\sqrt{t^2 - |x - y|^2}}dy.
\end{equation}
In order to find the derivative in the first term we transform the variable of integration to $z = (y - x)/t$ to
get
\[
\frac{t^2}{2}\fint_{B(x,t)} \frac{g(y)}{\sqrt{t^2 - |x - y|^2}}dy = 
\frac{t}{2}\fint_{B(0,1)} \frac{g(x+tz)}{\sqrt{1 - |z|^2}}dz
\]
so that
\[
\frac{\partial}{\partial t}\left(\frac{t^2}{2}\fint_{B(x,t)} \frac{g(y)}{\sqrt{t^2 - |x - y|^2}}dy\right) =
\frac{1}{2}\fint_{B(0,1)} \frac{g(x+tz)}{\sqrt{1 - |z|^2}}dz + 
\frac{t}{2}\fint_{B(0,1)} \frac{z\cdot Dg(x+tz)}{\sqrt{1 - |z|^2}}dz.
\]
Reverting back to $y$,
\begin{eqnarray*}
\frac{\partial}{\partial t}\left(\frac{t^2}{2}\fint_{B(x,t)} \frac{g(y)}{\sqrt{t^2 - |x - y|^2}}dy\right) &=&
\frac{t}{2}\fint_{B(x,t)}\frac{g(y)}{\sqrt{t^2 - |x - y|^2}}dy + \\
\frac{t}{2}\fint_{B(x,t)}\frac{(y-x)\cdot Dg}{\sqrt{t^2 - |x - y|^2}}dy
\end{eqnarray*}
Using this equation in \eqref{s28e42},
\begin{equation}\label{s28e43}
u(x,t) = \frac{1}{2}\fint_{B(x,t)}\frac{tg(y) + t(y-x)\cdot Dg(y) + t^2h(y)}{\sqrt{t^2 - |x - y|^2}}dy.
\end{equation}
Equation \eqref{s28e43} is called \emph{Poisson's formula} for the two dimensional wave equation.

We will now solve the Euler-Poisson-Darboux equation for arbitrary dimensions treating even and odd space 
dimensions separately.

\subsection{Odd space dimensions}
We will find the following proposition useful.
\begin{prop}\label{s28p2}
Let $\phi: \mathbb{R} \rightarrow \mathbb{R}$ be $C^{k+1}$. Then
\begin{enumerate}
\item \begin{equation}\label{s28e44}
\left(\frac{d}{dr}\right)^2\left(\frac{1}{r}\frac{d}{dr}\right)^{k-1}\left(r^{2k-1}\phi(r)\right) = 
\left(\frac{1}{r}\frac{d}{dr}\right)^k\left(r^{2k}\td{\phi}{r}\right).
\end{equation}
\item \begin{equation}\label{s28e45}
\left(\frac{1}{r}\frac{d}{dr}\right)^{k-1}\left(r^{2k-1}\phi(r)\right) = 
\sum_{j=0}^{k-1}\beta^k_j r^{j+1}\left(\frac{d}{dr}\right)^{j}\phi,
\end{equation}
where $\beta^k_j$ are constants, independent of $r$.
\item $\beta^k_0 = (2k-1)!!$.
\end{enumerate}
\end{prop}
\begin{proof}
We will prove \eqref{s28e44} by induction on $k$. When $k=1$,
\[
\text{LHS}= \left(\frac{d}{dr}\right)^2(r\phi(r)) = 2\phi^\prime(r) + r\phi^{\prime\prime}(r).
\]
and
\[
\text{RHS} = \left(\frac{1}{r}\frac{d}{dr}\right)(r^2\phi^\prime(r)) = 2\phi^\prime(r) + r\phi^{\prime\prime}(r).
\]
Let the claim be true for $1 < m < k$. Consider
\begin{equation}\label{s28e46}
\left(\frac{d}{dr}\right)^2\left(\frac{1}{r}\frac{d}{dr}\right)^{m}\left(r^{2m+1}\phi(r)\right) =
\left(\frac{d}{dr}\right)^2\left(\frac{1}{r}\frac{d}{dr}\right)^{m-1}\left(r^{2m-1}\psi(r)\right),
\end{equation}
where
\begin{equation}\label{s28e47}
\psi(r) = (2m+1)\phi(r) + r\phi^\prime(r).
\end{equation}
Using induction hypothesis on the right hand side of \eqref{s28e46} we get
\begin{equation}\label{s28e48}
\left(\frac{d}{dr}\right)^2\left(\frac{1}{r}\frac{d}{dr}\right)^{m}\left(r^{2m+1}\phi(r)\right) =
\left(\frac{1}{r}\frac{d}{dr}\right)^m\left(r^{2m}\td{\psi}{r}\right).
\end{equation}
Using the definition of $\psi$,
\[
r^{2m}\td{\psi}{r} = \left(\frac{1}{r}\frac{d}{dr}\right)\left(r^{2m+1}\td{\phi}{r}\right)
\]
so that equation \eqref{s28e48} becomes
\[
\left(\frac{d}{dr}\right)^2\left(\frac{1}{r}\frac{d}{dr}\right)^{m}\left(r^{2m+1}\phi(r)\right) =
\left(\frac{1}{r}\frac{d}{dr}\right)^{m+1}\left(r^{2(m+1)-1}\td{\phi}{r}\right).
\]

Equation \eqref{s28e45} can also be proved by induction on $k$. For $k=1$, LHS = $r\phi(r)$ while
\[
\text{RHS} = \sum_{j=0}^0 \beta^1_0 r\phi(r) = r\phi(r),
\]
if $\beta^1_0 = 1$, which it is as will be shown in part three of the proposition. Let the claim be true for $1 < m 
< k$. Consider
\begin{equation}\label{s28e49}
\left(\frac{1}{r}\frac{d}{dr}\right)^m(r^{2m+1}\phi(r)) = 
\left(\frac{1}{r}\frac{d}{dr}\right)\left[\left(\frac{1}{r}\frac{d}{dr}\right)^{m-1}(r^{2m-1}\psi(r))\right],
\end{equation}
where $\psi(r) = r^2\phi(r)$. Using the induction hypothesis for the term in square brackets on the right hand
side of the above equation we get
\begin{equation}\label{s28e50}
\left(\frac{1}{r}\frac{d}{dr}\right)^m(r^{2m+1}\phi(r)) = 
\left(\frac{1}{r}\frac{d}{dr}\right)\left[\sum_{j=0}^{m-1}\beta^m_j r^{j+1}\left(\frac{d}{dr}\right)^j\psi\right]
\end{equation}
From the definition of $\psi$, we can conclude that 
\[
\left(\frac{d}{dr}\right)^j\psi = \begin{cases}
2r\phi(r) + r^2\phi^\prime(r) & \text{ for } j = 1 \\ 
2\alpha_j \left(\frac{d}{dr}\right)^{j-2}\phi + 
2jr\left(\frac{d}{dr}\right)^{j-1}\phi + r^2\left(\frac{d}{dr}\right)^j\phi & \text{ for } j \ge 2.
\end{cases}
\]
The numbers $\alpha_j$ are defined recursively as $\alpha_j = \alpha_{j-1} + j$ with $\alpha_1 = 1$. Thus,
\begin{equation}\label{s28e51}
\left(\frac{1}{r}\frac{d}{dr}\right)\left(r^{j+1}\frac{d^j\psi}{dr^j}\right) = 
6r\phi(r) + 6r^2\phi^{\prime}(r) + r^3\phi^{\prime\prime}(r)
\end{equation}
when $j = 1$ and 
\begin{eqnarray}
\left(\frac{1}{r}\frac{d}{dr}\right)\left(r^j\frac{d^j\psi}{dr^j}\right) &=&
2(j+1)\alpha_j r^{j-1}\frac{d^{j-2}\phi}{dr^{j-2}} + 
r^j(2\alpha_j + 2j(j+2))\frac{d^{j-1}\phi}{dr^{j-1}} + \nonumber \\
& & 3r^{j+1}(j+1)\frac{d^j\phi}{dr^j} + r^{2+j}\frac{d^{j+1}\phi}{dr^{j+1}} \label{s28e52}
\end{eqnarray}
when $j \ge 2$. Substituting equations \eqref{s28e51} and \eqref{s28e52} in \eqref{s28e50},
\begin{eqnarray*}
\left(\frac{1}{r}\frac{d}{dr}\right)^m(r^{2m+1}\phi(r)) &=& 
\beta^m_0 (6r\phi(r) + 6r^2\phi^{\prime}(r) + r^3\phi^{\prime\prime}(r)) + \\
& & \sum_{j=1}^{m-1}\beta^m_j\left(2(j+1)\alpha_j r^{j-1}\frac{d^{j-2}\phi}{dr^{j-2}} + 
r^j(2\alpha_j + 2j(j+2))\frac{d^{j-1}\phi}{dr^{j-1}}\right.\\ 
& & \left. + r^{j+1}(j+1)\frac{d^j\phi}{dr^j} + r^{2+j}\frac{d^{j+1}\phi}{dr^{j+1}}\right) \\
&=& \sum_{j=0}^{m+1}\beta^{m+1}_j r^{j+1}\frac{d^j\phi}{dr^j},
\end{eqnarray*}
where the constants $\beta^{m+1}_j$ are independent of $r$.

\noindent Lastly, in order to prove that $\beta_0^k = (2k-1)!!$ we just note that it emerges out of the term
\[
\left(\frac{1}{r}\frac{d}{dr}\right)^{k-1} r^{2k-1}.
\]
\end{proof}

Returning back to the wave equation, let $u$ solve the equation
\begin{eqnarray}
u_{tt} &=& \Delta u \text{ in } \mathbb{R}^n \times (0, \infty) \label{s28e53} \\
u &=& g \text{ on } \mathbb{R}^n \times \{t = 0\} \nonumber \\
u_t &=& h \text{ on } \mathbb{R}^n \times \{t = 0\} \nonumber
\end{eqnarray}
where $n = 2k + 1$ and $k \ge 2$. Let $u \in C^{2k+1}(\mathbb{R}^n \times (0,\infty))$ solve this initial value
problem. Then the function $U$ defined by \eqref{s28e11} is also in $C^{2k+1}(\mathbb{R}^2)$. Define
\begin{eqnarray}
\tilde{U}(r, t) &:=& \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}(r^{2k-1}U(x; r,t)) \label{s28e54} \\
\tilde{G}(r) &:=& \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}(r^{2k-1}G(x;r)) \label{s28e55} \\
\tilde{H}(r) &:=& \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}(r^{2k-1}H(x;r)) \label{s28e56}
\end{eqnarray}
then by definition of $U, G, H$ and the above three equations
\begin{eqnarray}
\tilde{U}(r, 0) &=& \tilde{G}(r) \label{s28e57} \\
\tilde{U}_t(r, 0) &=& \tilde{H}(r) \label{s28e58}
\end{eqnarray}

Analogous to the case of three dimensions, we show that
\begin{prop}\label{s28p3}
$\tilde{U}$ solves the one-dimensional wave equation. That is $\tilde{U}_{tt} - \tilde{U}_{rr} = 0$ in 
$\mathbb{R}_+ \times (0, \infty)$, $\tilde{U} = \tilde{G}$ and $\tilde{U}_t = \tilde{H}$ on $\mathbb{R}_+ \times
\{t = 0\}$ and $\tilde{U}(0, t) = 0$.
\end{prop}
\begin{proof}
From equation \eqref{s28e54},
\[
\tilde{U}_{rr} = 
\frac{\partial^2}{\partial r^2}\left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}(r^{2k-1}U(x; r,t))
\]
From equation \eqref{s28e44},
\begin{eqnarray*}
\tilde{U}_{rr} &=& \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^k\left(r^{2k}U_r\right) \\
 &=& \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}\left(2kr^{2k-2}U_r + r^{2k-1}U_{rr}\right) \\
 &=& \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}\left[r^{2k-1}\left(U_{rr} + 
 \frac{n-1}{r}U_r\right)\right],
\end{eqnarray*}
where we used the fact that $2k = n - 1$. From the Euler-Poisson-Darboux equation \eqref{s28e12}, 
\[
\tilde{U}_{rr} = \left(\frac{1}{r}\frac{\partial}{\partial r}\right)^{k-1}(r^{2k-1}U_{tt})
\]
From \eqref{s28e54}, we see that the right hand side is $\tilde{U}_{tt}$. Thus $\tilde{U}$ satisfies the one-
dimensional wave equation. Let us now check if it also satisfies the boundary conditions. Using \eqref{s28e45},
we can write $\tilde{U}(r,t)$ as
\begin{equation}\label{s28e59}
\tilde{U}(r,t) = \sum_{j=0}^{k-1}\beta^k_j r^{j+1}\frac{\partial^j U}{\partial r^j}.
\end{equation}
Since $\beta_k^j$ are all constants and the $U$ is in $C^{2k+1}$ , $U(0, t) = 0$. The other two boundary conditions
follow from the definitions of $U, G, H$ and equations \eqref{s28e54} to \eqref{s28e56}.
\end{proof}

Since $\tilde{U}$ satisfies the one-dimensional wave equation, it can be expressed as d'Alembert's solution
\eqref{s28e10} 
\begin{equation}\label{s28e60}
\tilde{U}(r,t) = \frac{\tilde{G}(t+r) - \tilde{G}(t-r)}{2} + \frac{1}{2}\int_{t-r}^{t+r}\tilde{H}(y)dy.
\end{equation}
We are interested in getting a solution to equation \eqref{s28e53}. Therefore, we need a relationship between
$\tilde{U}$ and $u$. From equation \eqref{s28e59},
\[
\frac{\tilde{U}(r,t)}{r} = \sum_{j=0}^{k-1}\beta^k_j r^{j}\frac{\partial^j U}{\partial r^j}.
\]
so that
\[
\lim_{r \rightarrow 0}\frac{\tilde{U}(r,t)}{r} = \lim_{r\rightarrow 0} \beta^k_0 U(x; r, t) = \beta^k_0 u(x,t)
\]
or
\begin{equation}\label{s28e61}
u(x,t) = \lim_{r \rightarrow 0} \frac{\tilde{U}(r,t)}{\beta^k_0 r}.
\end{equation}
From equations \eqref{s28e60} and \eqref{s28e61},
\[
u(x,t) = \frac{\tilde{G}^\prime(t) + \tilde{H}(t)}{\beta^k_0}
\]
From part 3 of proposition \ref{s28p2}, $\beta_0^k = (2k-1)!!$. Since $n = 2k+1$, $\beta_0^k = (n-2)!!$. From
\eqref{s28e56}, 
\[
\tilde{H}(t) = \left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}(t^{n-2}H(x,t))
\]
so that using \eqref{s28e13}, we get
\begin{equation}\label{s28e62}
\tilde{H}(t) = \left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}
\left(t^{n-2}\fint_{\partial B(x,r)}h(y, t)dS(y)\right)
\end{equation}
Similarly,
\begin{equation}\label{s28e63}
\tilde{G}^\prime(t) = \frac{\partial}{\partial t}\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}
\left(t^{n-2}\fint_{\partial B(x,r)}h(y, t)dS(y)\right)
\end{equation}
Substituting equations \eqref{s28e62} and \eqref{s28e63} in \eqref{s28e61} we get
\begin{eqnarray}
u(x,t) &=& \frac{1}{(n-2)!!}\left[\frac{\partial}{\partial t}\left(\frac{1}{t}\frac{\partial}{\partial t}
\right)^{(n-3)/2}\left(t^{n-2}\fint_{\partial B(x,t)}gdS\right)\right. + \nonumber \\
& & \left.\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}\left(t^{n-2}
\fint_{\partial B(x,r)}h(y, t)dS(y)\right)\right] \label{s28e64}
\end{eqnarray}
Equation \eqref{s28e64} is the solution to the initial value problem of \eqref{s28e53} when $n=2k+1$.

\subsection{Even space dimensions}
We will use Hadamard's method of descent to derive the explicit solution of the wave equation for even space
dimensions. Once again, define 
\begin{equation}\label{s28e65}
\bar{u}(x_1, \ldots, x_{n+1}, t) := u(x_1, \ldots, x_n, t).
\end{equation}
Define $\bar{g}$ and $\bar{h}$ similarly. Assume that $\bar{u}$ solves the wave equation in $\mathbb{R}^{n+1}
\times (0, \infty)$ with initial conditions $\bar{u} = \bar{g}$ and $\bar{u}_t = \bar{h}$ on $\mathbb{R}^{n+1}
\times \{0\}$. Now fix $x \in \mathbb{R}^n$ and $\bar{x} = (x, 0)$. Using \eqref{s28e64},
\begin{eqnarray}
u(x,t) &=& \frac{1}{(n-2)!!}\left[\frac{\partial}{\partial t}\left(\frac{1}{t}\frac{\partial}{\partial t}
\right)^{(n-2)/2}\left(t^{n-1}\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S}\right)\right. + \nonumber \\
& & \left.\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-2)/2}\left(t^{n-1}
\fint_{\partial\bar{B}(x,t)}\bar{h}(y, t)d\bar{S}(y)\right)\right]. \label{s28e66}
\end{eqnarray}
Note that $\bar{B}(x,t)$ is a ball in $\mathbb{R}^{n+1}$ with centre $x$ and radius $t$ and $d\bar{S}$ is the
$n$-dimensional surface measure on $\partial\bar{B}(x,t)$. From equations \eqref{s16e7}, \eqref{s16e10} and
\eqref{s16e11}, $S(n+1, t) = (n+1)V(n+1,1)t^n$ so that
\[
\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S} = \frac{1}{(n+1)V(n+1,1)t^n}\int_{\partial\bar{B}(x,t)}\bar{g}d\bar{S}.
\]
Transforming the integral to the $n$-disk,
\[
\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S} = \frac{2}{(n+1)V(n+1,1)t^n}
\int_{\partial\bar{B}(x,t)}\bar{g}(1 + |D\gamma(y)|^2)^{1/2}dy,
\]
where $\gamma(y) = \sqrt{t^2 - |x - y|^2}$ and the factor of $2$ arising due to contribution from each hemisphere.
Since 
\[
D\gamma(y) = \frac{t}{\sqrt{t^2 - |x - y|^2}}
\]
we have
\[
\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S} = \frac{2}{(n+1)V(n+1,1)t^{n-1}}
\int_{\partial\bar{B}(x,t)}\frac{\bar{g}}{\sqrt{t^2 - |x - y|^2}}dy.
\]
Writing the integral on the right hand side and an average,
\[
\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S} = \frac{2V(n,1)t^n}{(n+1)V(n+1,1)t^{n-1}}
\fint_{\partial\bar{B}(x,t)}\frac{\bar{g}}{\sqrt{t^2 - |x - y|^2}}dy.
\]
or
\begin{equation}\label{s28e67}
\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S} = \frac{2V(n,1)t}{(n+1)V(n+1,1)}
\fint_{\partial\bar{B}(x,t)}\frac{\bar{g}}{\sqrt{t^2 - |x - y|^2}}dy.
\end{equation}
Since $V(n, 1) = \pi^{n/2}/\Gamma(1 + n/2)$, 
\[
\frac{V(n,1)}{V(n+1,1)} = \frac{1}{\sqrt{\pi}}\frac{\Gamma((n+3)/2)}{\Gamma((n+2)/2)}
\]
Since $n$ is even, $n+3$ is odd and therefore the expansion of the numerator will end with $\Gamma(1/2) = 
\sqrt{\pi}$. Using the fact that $\Gamma(z) = (z-1)\Gamma(z-1)$, we get
\[
\frac{V(n,1)}{V(n+1,1)} = \frac{(n+1)(n-1) \cdots 3}{n(n-2)(n-4) \cdots 2}
\]
and hence
\begin{equation}\label{s28e68}
\frac{V(n,1)}{(n+1)V(n+1)} = \frac{(n-1)!!}{n!!}.
\end{equation}
From equations \eqref{s28e67} and \eqref{s28e68}, we get
\begin{equation}\label{s28e69}
\fint_{\partial\bar{B}(x,t)}\bar{g}d\bar{S} = \frac{2t(n-1)!!}{n!!}
\fint_{\partial B(x,t)}\frac{g}{\sqrt{t^2 - |x - y|^2}}dy.
\end{equation}
Substituting \eqref{s28e69} and the analogous equation for $\bar{h}$ in \eqref{s28e66}, we get the explicit form
of the solution of the wave equation in $n$ dimensions as
\begin{eqnarray}
u(x,t) &=& \frac{1}{n!!}\left[\frac{\partial}{\partial t}\left(\frac{1}{t}\frac{\partial}{\partial t}
\right)^{(n-2)/2}\left(t^n\fint_{B(x,t)}\frac{g}{\sqrt{t^2 - |x - y|^2}}dy\right)\right. + \nonumber \\
& & \left.\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-2)/2}\left(t^n
\fint_{B(x,t)}\frac{h}{\sqrt{t^2 - |x - y|^2}}dy\right)\right]. \label{s28e70}
\end{eqnarray}
Comparing equations \eqref{s28e64} and \eqref{s28e70} brings out the key difference between the solution of the
wave equation in odd and even dimensions. When the number of spatial dimensions is odd the data $g$ and $h$ at
$x \in \mathbb{R}^n$ determine the solution $u$ only on the boundary $\partial B(x, t)$. On the other hand, when
the number of spatial dimensions is even, data in the entire ball $B(x, t)$ determine the solution on its boundary.
Thus, Huygen's principle is valid in odd spatial dimensions but not in even spatial dimensions.

Equations \eqref{s28e64} and \eqref{s28e70} give the explicit solutions of the wave equations in any dimensions. 
Yet we haven't confirmed that they indeed solve the wave equation and have the right continuity properties. The
next two theorems will accomplish this task.

\begin{thm}\label{s28t1}
Let $n \ge 3$ be an odd integer, $g \in C^{m+1}(\mathbb{R}^n), h \in C^n(\mathbb{R}^n)$ for $m = (n + 1)/2$. Then
$u$ defined by \eqref{s28e64} has the following properties.
\begin{enumerate}
\item $u \in C^2(\mathbb{R}^n \times (0, \infty)$,
\item $u_{tt} - \Delta u = 0$ in $\mathbb{R}^n \times (0, \infty)$,
\item For all $x \in \mathbb{R}^n, t > 0$, $u(x, t) \rightarrow g(x_0)$ and $u_t(x, t) \rightarrow h(x_0)$ in the
limit $x \rightarrow x_0, t \rightarrow 0+$.
\end{enumerate}
\end{thm}
\begin{proof}
Using equations \eqref{s28e14} and \eqref{s28e15}, we can write \eqref{s28e64} as
\begin{equation}\label{s28e71}
u(x,t) = \frac{1}{(n-2)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}(t^{n-2}G(x; t,t)) +
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}(t^{n-2}H(x; t,t))\right]
\end{equation}
Using \eqref{s28e44},
\begin{equation}\label{s28e72}
u_{tt}(x,t) = \frac{1}{(n-2)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-1)/2}(t^{n-1}G_t(x; t,t)) +
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-1)/2}(t^{n-1}H_t(x; t,t))\right]
\end{equation}
Let us now calculate $G_t$ and $H_t$. From \eqref{s28e13},
\[
G(x;t,t) = \fint_{\partial B(x,t)}g(y,t)dS(y) = \frac{1}{S(n,t)}\int_{\partial B(x,t)} g(y,t)dS(y)
\]
Introduce a new variable $z = (y - x)/t$ so that $dS(z) = dS(y)/t^{n-1}$ and
\[
G(x;t,t) = \frac{1}{S(n,1)}\int_{\partial B(0,1)} g(x + tz) dS(z)
\]
so that
\[
G_t(x;t,t) = \frac{1}{S(n,1)}\int_{\partial B(0,1)} z\cdot Dg dS(z) = 
\frac{1}{S(n,r)}\int_{\partial B(x,t)} \frac{(y - x)}{r}\cdot Dg dS(y)
\]
Since $(y - x)/r$ is a unit normal to the ball $\partial B(x,r)$, we can use Gauss' theorem to get
\[
G_t(x;t,t) = \frac{t}{nV(n,t)}\int_{B(x,t)} \Delta g dy
\]
or
\begin{equation}\label{s28e73}
G_t(x;t,t) = \frac{t}{n}\fint_{B(x,t)} \Delta g dy
\end{equation}
Similarly,
\begin{equation}\label{s28e74}
H_t(x;t,t) = \frac{t}{n}\fint_{B(x,t)} \Delta h dy
\end{equation}
From \eqref{s28e73},
\[
t^{n-1}G_t = \frac{t^n}{n}\fint_{B(x,t)} \Delta g dy = \frac{t^n}{nV(n,t)}\int_{B(x,t)}\Delta g dy =
\frac{1}{nV(n,1)}\int_{B(x,t)}\Delta g dy
\]
so that
\[
\frac{\partial}{\partial t}(t^{n-1}G_t) = \frac{1}{nV(n,1)}\frac{\partial}{\partial t}\int_{B(x,t)}\Delta g dy
\]
Split the volume integral on the right hand side as
\[
\frac{\partial}{\partial t}(t^{n-1}G_t) = \frac{1}{nV(n,1)}\frac{\partial}{\partial t}
\left(\int_0^t\int_{\partial B(x,\rho)} \Delta g dS(y)d\rho\right)
\]
Using the first principles to compute the derivative on the right hand side,
\[
\frac{\partial}{\partial t}(t^{n-1}G_t) = \frac{1}{nV(n,1)}\lim_{h\rightarrow 0}
\left(\int_t^{t+h}\int_{\partial B(x,\rho)} \Delta g dS(y)d\rho\right)
\]
or
\[
\frac{\partial}{\partial t}(t^{n-1}G_t) = \frac{1}{nV(n,1)}\int_{\partial B(x,t)} \Delta g dS(y)
\]
or
\[
\frac{1}{t}\frac{\partial}{\partial t}(t^{n-1}G_t) = \frac{1}{nV(n,1)t}\int_{\partial B(x,t)} \Delta g dS(y)
\]
and hence
\begin{equation}\label{s28e75}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-1)/2}(t^{n-1}G_t(x; t,t)) = 
\frac{1}{nV(n,1)}\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}
\frac{1}{t}\int_{\partial B(x,t)} \Delta g dS(y)
\end{equation}
Similarly,
\begin{equation}\label{s28e76}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-1)/2}(t^{n-1}H_t(x; t,t)) = 
\frac{1}{nV(n,1)}\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}
\frac{1}{t}\int_{\partial B(x,t)} \Delta h dS(y)
\end{equation}
Substituting \eqref{s28e75} and \eqref{s28e76} in \eqref{s28e72},
\begin{eqnarray}
u_{tt}(x,t) &=& \frac{1}{nV(n,1)(n-2)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}\frac{1}{t}\int_{\partial B(x,t)} \Delta g dS(y) 
\right. \nonumber \\
& & + \left.
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}\frac{1}{t}\int_{\partial B(x,t)}\Delta hdS(y)\right]
\label{s28e77}
\end{eqnarray}
From \eqref{s28e13}, once again,
\[
\Delta G(x; t,t) = \Delta_x \frac{1}{S(n,t)}\int_{\partial B(x,t)}g(y,t)dS(y) 
\]
Introduce a new variable $z = y - x$ so that
\[
\Delta G(x;t,t) = \frac{1}{S(n,t)}\Delta_x\int_{\partial B(0,t)} g(x + z, t)dS(z) = 
\frac{1}{S(n,t)}\int_{\partial B(0,t)} \Delta_x g(x + z, t)dS(z) 
\]
or
\[
\Delta G(x;t,t) = \frac{1}{S(n,t)}\int_{\partial B(0,t)} \Delta_z g(x + z, t)dS(z) = 
\fint_{\partial B(x,t)} \Delta_y g(y, t)dS(y)
\]
We can carry out the same steps for $H$ to get
\begin{eqnarray}
\Delta G(x;t,t) &=& \fint_{\partial B(x,t)} \Delta g(y, t)dS(y) \label{s28e78} \\
\Delta H(x;t,t) &=& \fint_{\partial B(x,t)} \Delta h(y, t)dS(y) \label{s28e79}
\end{eqnarray}
Substituting equations \eqref{s28e78} and \eqref{s28e79} in \eqref{s28e77} we get
\begin{eqnarray*}
u_{tt}(x,t) &=& \frac{1}{nV(n,1)(n-2)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}\frac{S(n,t)}{t}\Delta G(x;t,t) 
\right.  \\
& & + \left.
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}\frac{S(n,t)}{t}\Delta H(x;t,t)\right]
\end{eqnarray*}
Now,
\[
S(n,t) = \frac{n}{t}V(n,t) = \frac{n}{t}V(n,1)t^n
\]
so that
\begin{eqnarray*}
u_{tt}(x,t) &=& \frac{1}{(n-2)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}t^{n-2}\Delta G(x;t,t) 
\right.  \\
& & + \left.
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}t^{n-2}\Delta H(x;t,t)\right]
\end{eqnarray*}
or,
\begin{eqnarray*}
u_{tt}(x,t) &=& \Delta\left\{\frac{1}{(n-2)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}t^{n-2} G(x;t,t) 
\right.\right.  \\
& & + \left.\left.
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{(n-3)/2}t^{n-2} H(x;t,t)\right]\right\}
\end{eqnarray*}
From equation \eqref{s28e71} we get $u_{tt} = \Delta u$. This also shows that $u \in C^2(\mathbb{R}^n \times 
(0,\infty))$.

Since $n=2k+1$, let us write \eqref{s28e71} in terms of $k$ so that
\[
u(x,t) = \frac{1}{(2k-1)!!}\left[\frac{\partial}{\partial t}
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{k-1}(t^{2k-1}G(x; t,t)) +
\left(\frac{1}{t}\frac{\partial}{\partial t}\right)^{k-1}(t^{2k-1}H(x; t,t))\right]
\]
Using equation \eqref{s28e45}, 
\begin{equation}\label{s28e80}
u(x,t) = \frac{1}{(2k-1)!!}\left[\frac{\partial}{\partial t}\sum_{j=0}^{k-1}\beta_j^kt^{j+1}\frac{d^jG}{dt^j}
+ \sum_{j=0}^{k-1}\beta_j^kt^{j+1}\frac{d^jH}{dt^j}\right]
\end{equation}
or, using part (3) of proposition \ref{s28p2},
\[
u(x,t) = G + t\td{G}{t} + \frac{1}{(2k-1)!!}\left[
\frac{\partial}{\partial t}\sum_{j=1}^{k-1}\beta_j^kt^{j+1}\frac{d^jG}{dt^j}
+ \sum_{j=0}^{k-1}\beta_j^kt^{j+1}\frac{d^jH}{dt^j}\right]
\]
In the limit $t \rightarrow 0$, only the first term on the right hand side survives. This is because all the 
derivatives of $G$ and $H$ up to order $k-1$ are known to exist. The limit of the right hand side is thus $g(x)$. 
Taking the derivative of \eqref{s28e80} with respect to $t$,
\[
\frac{1}{(2k-1)!!}\left[\frac{\partial^2}{\partial t^2}\sum_{j=0}^{k-1}\beta_j^kt^{j+1}\frac{d^jG}{dt^j}
+ \frac{\partial}{\partial t}\sum_{j=0}^{k-1}\beta_j^kt^{j+1}\frac{d^jH}{dt^j}\right]
\]
Once again, the only surviving term is $H$, whose limit as $t \rightarrow 0$ is $h(x)$. Note that the term $G_t$
also tends to zero because of \eqref{s28e73}.
\end{proof}

The analogous theorem for even spatial dimensions is
\begin{thm}\label{s28t2}
Let $n \ge 3$ be an even integer, $g \in C^{m+1}(\mathbb{R}^n), h \in C^n(\mathbb{R}^n)$ for $m = (n + 2)/2$. Then
$u$ defined by \eqref{s28e70} has the following properties.
\begin{enumerate}
\item $u \in C^2(\mathbb{R}^n \times (0, \infty)$,
\item $u_{tt} - \Delta u = 0$ in $\mathbb{R}^n \times (0, \infty)$,
\item For all $x \in \mathbb{R}^n, t > 0$, $u(x, t) \rightarrow g(x_0)$ and $u_t(x, t) \rightarrow h(x_0)$ in the
limit $x \rightarrow x_0, t \rightarrow 0+$.
\end{enumerate}
\end{thm}

\subsection{Nonhomogeneous problem}
Consider the Cauchy problem for the nonhomogeneous wave equation,
\begin{equation}\label{s28e81}
u_{tt} - \Delta u = f \text{ in } \mathbb{R}^n \times (0, \infty)
\end{equation}
subject to the initial conditions
\begin{equation}\label{s28e82}
u = 0, u_t = 0 \text{ on } \mathbb{R}^n \times \{0\}.
\end{equation}
Introduce the function $v = u(x,t;s)$ satisfying the initial value problem $v_{tt} - \Delta v = 0$ in $\mathbb{R}^n
\times [s, \infty)$ subject to the initial conditions $v(x,t;s) = 0, v_t(x,t;s) = f(s)$ on $\mathbb{R}^n \times
\{t=s\}$. If 
\begin{equation}\label{s28e83}
u(x, t) = \int_0^t v(x,t;s)ds
\end{equation}
then \emph{Duhamel's principle} asserts that $u$ solves the problem of equations \eqref{s28e81} and \eqref{s28e82}.
We will prove it in the following theorem.
\begin{thm}\label{s28t3}
Assume that $n \ge 2$, $f \in C^{\lceil{n/2}+1\rceil}(\mathbb{R}^n \times (0, \infty)$. For $u$ defined by 
\eqref{s28e83},
\begin{enumerate}
\item $u \in C^2(\mathbb{R}^n$,
\item $u_{tt} - \Delta u = f$ in $\mathbb{R}^n \times (0, \infty)$,
\item $u(x,t) \rightarrow 0$ and $u_t(x,t) \rightarrow 0$ as $(x,t) \rightarrow (x_0, 0+)$ for every $x_0 \in
\mathbb{R}^n$.
\end{enumerate}
\end{thm}
\begin{proof}
Irrespective of whether $n$ is even or odd, theorems \ref{s28t1} and \ref{s28t2}, $v$ is twice differentiable in $
[s, \infty)$ for all $s > 0$ and $x \in \mathbb{R}^n$. Since the first derivative of $u$ is $v$, $u$ is also 
guaranteed to be twice differentiable.

From equation \eqref{s28e83}, using proposition \ref{s27p3}
\[
u_t(x,t) = v(x,t;t) + \int_0^t v_t(x,t;s)ds 
\]
Since $v(x,t;s) = 0$ when $t=s$ (refer to the initial conditions on the $v$-problem), we have
\begin{equation}\label{s28e84}
u_t(x,t) = \int_0^t v_t(x,t;s)ds
\end{equation}
Similarly,
\[
u_{tt}(x,t) = v_t(x,t;t) + \int_0^t v_{tt}(x,t;s)ds = f(x,t) + \int_0^t v_{tt}(x,t;s)ds
\]
Taking the laplacian of $u$ is quite straightforward,
\[
\Delta u = \int_0^t \delta v(x,t;s)ds = \int_0^t v_{tt}(x,t;s)ds
\]
so that $u_{tt} - \Delta u = f(x,t)$.

Finally, $u(x,0) = 0$ by its definition and $u_t(x,0) = 0$ be \eqref{s28e84}.
\end{proof}

Lastly, we show how the formula for retarted potential naturally arises in the Duhamel's principle applied to the
inhomogeneous wave equation in three dimensions. The solution $v$ of the corresponding homogeneous equation is
given by Kirchhoff's formula \eqref{s28e37} but with initial conditions $v(x, 0) = 0$ and $v_t(x,t) = f$. Thus,
in Kirchhoff's formula, $g = 0$ and $h = f$. The solution of the homogeneous problem is
\[
v(x,t;s)=\fint_{\partial B(x,t-s)}(t-s)fdS(y)
\]
in $\mathbb{R}^n \times (s,\infty)$. Therefore, the solution of the homogeneous equation is
\[
u(x,t) = \int_0^t\fint_{\partial B(x,t-s)}(t-s)f(y,s)dS(y)ds = 
\frac{1}{4\pi}\int_0^t\int_{\partial B(x,t-s)}\frac{f(y,s)}{t - s}dS(y)ds
\]
Introduce the variable $r = t - s$ so that
\[
u(x,t) = \frac{1}{4\pi}\int_0^t\int_{\partial B(x,r)}\frac{f(y, t-r)}{r}dS(y)dr
= \frac{1}{4\pi}\int_{B(x,t)}\frac{f(y, t - |x - y|)}{|y - x|}dy.
\]

\section{Problems}\label{s29}
\begin{enumerate}
\setlength{\leftmargin}{0pt}
\item Let $u, v \in C^2(\bar{U})$ prove the following identities using the Green-Gauss theorem \ref{s30t1}
\begin{equation}\label{s29e1}
\int_U v_{x_i} dx = \int_{\partial U} vn_i dS(x)
\end{equation}
If we apply this formula to $uv$, we get
\[
\int_U (uv)_{x_i} dx = \int_{\partial U} uv n_i dS(x)
\]
or
\[
\int_U uv_{x_i} dx + \int_U u_{x_i}v dx = \int_{\partial U} uv n_i dS(x)
\]
from which we get the integration by parts formula
\begin{equation}\label{s29e2}
\int_U u_{x_i} v dx = -\int_U uv_{x_i}dx + \int_{\partial U} uv n_i dS(x).
\end{equation}
\begin{enumerate}
\setlength{\leftmargin}{0pt}
\item 
\begin{equation}\label{s29e3}
\int_U\Delta u dx = \int_{\partial U}n \cdot Du dS(x) 
\end{equation}
Use $v = u_{x_i}$ in \eqref{s29e1} and the fact that $\Delta u = \sum_{i=1}^nu_{x_ix_i}$. 
\item 
\begin{equation}\label{s29e4}
\int_U v\Delta u dx = -\int_U Du\cdot Dv dx + \int_{\partial U}v n\cdot Du dS(x)
\end{equation}
Use the mapping $u \mapsto u_{x_i}$ in \eqref{s29e2} to get
\[
\int_U u_{x_ix_i} v dx = -\int_U u_{x_i}v_{x_i}dx + \int_{\partial U} u_{x_i}v n_i dS(x).
\]
Summing $i$ from $1$ to $n$ gives \eqref{s29e4}.
\item
\begin{equation}\label{s29e5}
\int_U(u \Delta v - v \Delta u) dx = \int_{\partial U}(u\;Dv - v\;Du)\cdot n dS.
\end{equation}
Interchanging $u$ and $v$ in equation \eqref{s29e4},
\begin{equation}\label{s29e6}
\int_U u\Delta v dx = -\int_U Du\cdot Dv dx + \int_{\partial U}u n\cdot Dv dS(x)
\end{equation}
Subtracting \eqref{s29e4} from \eqref{s29e6} gives \eqref{s29e5}.
from which we 
\end{enumerate}

\item Let $\Phi$ be the fundamental solution of the Laplace equation. Prove that $\Phi$ and $\Phi_{x_i}$ are
locally integrable but $\Phi_{x_ix_j}$ is not.

Solution: We will prove the assertion separately for $n=2$ and $n > 2$. For $n=2$,
\begin{eqnarray*}
\Phi(r) &=& -\frac{1}{2\pi}\ln r \\
\Phi_{x_i}(r) &=& -\frac{1}{2\pi}\frac{x_i}{r} \\
\Phi_{x_ix_j}(r) &=& -\frac{1}{2\pi}\left(\frac{\delta_{ij}}{r} - \frac{x_ix_j}{r^3}\right)
\end{eqnarray*}
The only singularity of these functions is $r = 0$. Therefore, we consider the integral of each one of these over
a disk of radius $R$ and centred at the origin. Thus,
\[
\int_0^R\int_0^{2\pi}\Phi(r) rdrd\theta = -\int_0^Rr\ln r dr = \frac{R^2}{2}\left(\frac{1}{2} - \ln R\right)
\]
and
\begin{eqnarray*}
\int_0^R\int_0^{2\pi}\Phi_{x_1}(r) rdrd\theta &=& 0 \\
\int_0^R\int_0^{2\pi}\Phi_{x_2}(r) rdrd\theta &=& 0
\end{eqnarray*}
However,
\[
\int_0^R\int_0^{2\pi}\Phi_{x_ix_j}(r) rdrd\theta
\]
does not exist because integral of the term $x_ix_j/r^3$ diverges for all $i, j$.

For $n > 2$, 
\begin{eqnarray*}
\Phi(r) &=& C_n r^{-(n-2)} \\
\Phi_{x_i}(r) &=& -(n-2)C_n r^{-n} x_i \\
\Phi_{x_ix_j}(r) &=& n(n-2)C_n \frac{x_i}{r^{n+1}} - (n-2)\frac{\delta_{ij}}{r^n},
\end{eqnarray*}
where $C_n = (n(n-2)V(n, 1))^{-1}$ is independent of $r$.

We will use hyperspherical coordinates $(r, \phi_1, \ldots, \phi_n)$ to carry out the integration. The angular
coordinates $\phi_1, \ldots, \phi_{n-2}$ vary over $[0, \pi]$ and $\phi_{n-1}$ varies over $[0, 2\pi)$. The 
relation between them and the euclidean coordinates is 
\begin{eqnarray*}
x_1 &=& r\cos\phi_1 \\
x_2 &=& r\sin\phi_1\cos\phi_2 \\
x_3 &=& r\sin\phi_1\sin\phi_2\cos\phi_3 \\
\vdots &=& \vdots \\
x_n &=& r\sin\phi_1\ldots\sin\phi_{n-2}\cos\phi_{n-1}
\end{eqnarray*}
If $x = (x_1, \ldots, x_n)$ then the volume element in terms of the hyperspherical coordinates is
\[
dx = r^{n-1}\sin^{n-1}\phi_1\sin^{n-2}\phi_2 \ldots \sin\phi_{n-2} drd\phi_1d\phi_2\ldots d\phi_{n-1}
\]
so that
\[
\int_U \Phi dx = C_n\int_0^R rdr \int_0^{\pi} \cdots \int_0^{2\pi}  
\sin^{n-1}\phi_1 \ldots \sin\phi_{n-2} drd\phi_1\ldots d\phi_{n-1}
\]
exists. Similarly,
\[
\int_U \Phi_{x_i}dx = -(n-2)C_n \int_U \frac{\sin\phi_1\ldots\cos\phi_{i}}{r^{n-1}}dx
\]
also exists because the $r$ integral is
\[
\int_0^R  dr = R
\]
However, the $r$-integral in
\[
I = \int_U \Phi_{x_ix_j}dx
\]
is
\[
\int_0^R \frac{r^{n-1}}{r^n}dr = \int_0^R \frac{dr}{r}
\]
because of which it does not exist.

\item Let $\Omega$ be a domain in $\mathbb{R}^2$ symmetric about the $x$ axis and let $\Omega^+ = \{(x,y):y>0\}$
be its upper part. Assume that $u \in C^2(\bar{\Omega}^+)$ be harmonic in $\Omega^+$ with $u = 0$ on 
$\partial\Omega^+ \cap \{y = 0\}$. Define 
\[
v(x,y) = \begin{cases}
u(x, y) & \text{ if } y \ge 0, (x, y) \in \Omega \\
-u(x, -y) & \text{ if } y < 0, (x, y) \in \Omega
\end{cases}
\]
Show that $v$ is harmonic.

Solution: The function $v$ has two portions. The one in $\Omega^+$ coincides with $u$ while the one in $\Omega^-$ 
is a reflection of $u$, that is, it coincides with $u(x, -y)$ but has the opposite sign. On the $x$ axis, $y = 0$ 
and we observe that $v$ is continuous on $\partial\Omega^+ \cap \{y = 0\}$ because $u$ is continuous in $\Omega^+$.

Since $u$ is harmonic in $\Omega^+$, $v$ too is harmonic in $\Omega^+$. In $\Omega^-$, $v_x = -u_x, v_{xx} = 
u_{xx}, v_y = -u_y$ and $v_{yy} = u_{yy}$ so that $u_{xx} + u_{yy} = 0 \Rightarrow v_{xx} + v_{yy} = 0$.

\item Let $u \in C^2(\Omega) \cap C^0(\bar{\Omega})$ be the solution of
\[
\Delta u + \sum_k a_k(x)u_{x_k} + c(x)u = 0 \text{ in } \Omega,
\]
with $c(x) < 0$ in $\Omega$, $u = 0$ on $\partial\Omega$ and $a_k$'s being smooth. Show that $u \equiv 0$ in 
$\Omega$.

Solution: Since $u \in C^2(\Omega) \cap C^0(\bar{\Omega})$ it must have an extremum in $\bar{\Omega}$. Let $x_0$ be
the extremum. At the extremum, $u$ satisfies the equation $\Delta u + c(x_0)u(x_0) = 0$. Since $c(x) < 0$ for all
$x \in \Omega$, we have $\Delta u > 0$ at $x_0$ if $u(x_0) > 0$. Therefore $x_0$ is a minimum and $u(x_0) > 0$.
This means that $u$ is positive in $\bar{\Omega}$ contradicting the condition that $u = 0$ on $\partial\Omega$.
If, on the other hand, $u(x_0) < 0$ then $\Delta u < 0$ and hence $x_0$ is a maximum. If the maximum $u(x_0) < 0$
then the function should take only lesser values in $\bar{\Omega}$ contradicting the condition $u=0$ on
$\partial\Omega$.

\item Consider the PDE $-\Delta u = \lambda u$ in $\Omega$ with $u = 0$ on $\partial\Omega$ where $\lambda \in
\mathbb{R}$ and $\Omega$ an open set. If $\lambda \le 0$ prove that $u \equiv 0$ and that there is no non-trivial
solution.

Solution: We can write the PDE as $\Delta u = -\lambda u$ with $\lambda \le 0$. This is same as the previous 
problem with $c(x) = \lambda$.

\item Prove the following proposition.
\begin{prop}\label{s29p1}
Show that if $u$ is sub-harmonic in $U$ and $u \le M$ on $\partial U$ then $u \le M$ everywhere in $U$.
\end{prop}
\begin{proof}
$u$ is sub-harmonic in $U$ means that $\Delta u \ge 0$ in $U$. Construct a function $v := u + 
\epsilon\norm{x}^2$ for $\epsilon > 0$. Then 
\begin{equation}\label{s29e7}
\Delta v = \Delta u + 2n\epsilon \ge 0.
\end{equation}
Now $v$ can attain a maximum in $\bar{U}$ only on $\partial U$ and not in $\Int U$. For if there indeed was a point 
$x_0 \in \Int U$ where $v$ attained its maximum, $v_{xx}(x_0) \le 0, v_{yy}(x_0) \le 0$ and $\Delta v(x_0) \le 0$ 
contracticting equation \eqref{s29e7}. $u \le M$ on $\partial U$ implies that $v \le M + \epsilon\norm{x}^2$ on 
$\partial M$. Since $u \le v$ , this also means that 
\begin{equation}\label{s29e8}
u \le M + \epsilon\norm{x}^2 \text{ in } U.
\end{equation}
If at any point in $U$, $u > M$ then we can choose an $\epsilon$ small enough to violate \eqref{s29e8}.
\end{proof}

\item Prove the following proposition.
\begin{prop}\label{s29p2}
Show that if $u$ is super-harmonic in $U$ and $u \ge m$ on $\partial U$ then $u \ge m$ everywhere in $U$.
\end{prop}
\begin{proof}
Let $u$ be super-harmonic in $U$, that is, $\Delta u \le 0$ in $U$. Construct a function $v := u - 
\epsilon\norm{x}^2$, where $\epsilon > 0$, so that 
\begin{equation}\label{s29e9}
\Delta v = \Delta u - 2n\epsilon \le 0.
\end{equation} 
Let, if possible, $v$ attain a minimum at a point $x_0 \in \Int U$. Then $v_{xx}(x_0) \ge 0, v_{yy}(x_0) \ge 0$ so 
that $\Delta v \ge 0$, contraticting \eqref{s29e8}. Thus, the minimum of $v$ has to be on $\partial U$. $u \ge m$
on $\partial U$ implies that $v \ge m - \epsilon\norm{x}^2$ on $\partial U$. But $u \ge v$ is also true from the 
way we defined $v$. Therefore, 
\begin{equation}\label{s29e10}
u \ge m - \epsilon\norm{x}^2.
\end{equation}
If at any point in $U$, $u < m$, we can choose an 
$\epsilon$ small enough to violate \eqref{s29e10}.
\end{proof}

\item Prove the following proposition.
\begin{prop}\label{s29p3}
Show that if $u$ is harmonic in a bounded region $U$ and $u$ is continously differentiable in $\bar{U}$ then
$|Du|^2$ attains its maximum on $\partial U$.
\end{prop}
\begin{proof}
$Du$ is continuous in $\bar{U}$ as $u \in C^1(U)$. In particular, $Du$ is continuous in the subset 
$\partial U$ of $\bar{U}$. Since $U$ is bounded, so is $\partial U$ and hence continuity of $Du$ on $\partial U$
implies that $|Du|^2$ attains its extrema on $\partial{U}$. Let $M$ be the maximum of $|Du|^2$ on $\partial U$. If
we can show that $v = |Du|^2$ is sub-harmonic the using proposition \ref{s29p1} we can show that $|Du|^2 
\le M$ in all of $U$.

Let us now show that $v = |Du|^2$ is sub-harmonic.
\begin{eqnarray*}
v &=& \sum_{i=1}^n u_{x_i}^2 \\
Dv &=& \sum_{j=1}^n\left(2\sum_{i=1}^n u_{x_i}u_{x_ix_j}\right)e_j,  
\end{eqnarray*}
where $e_i$ is the standard basis vector of $\mathbb{R}^n$. Therefore,
\[
\Delta v = 2\sum_{j=1}^n\sum_{i=1}^n \left( u_{x_ix_j}^2 + u_{x_i}u_{x_ix_jx_j}\right)
\]
The second sum in the above equation can be split as
\begin{eqnarray*}
\sum_{j=1}^n\sum_{i=1}^n u_{x_i}u_{x_ix_jx_j} &=& 
\sum_{i=1}^n u_{x_i}u_{x_ix_ix_i} + \sum_{i=1}^n\sum_{j=1,j n\ne 1}^n u_{x_i}u_{x_ix_jx_j} \\
&=& \sum_{i=1}^n u_{x_i}\left(u_{x_ix_i} + \sum_{j=1}^n u_{x_jx_j}\right)_{x_i}
\end{eqnarray*}
The sum in the bracket is $\Delta u$, which is zero because $u$ is harmonic. Thus,
\[
\Delta v = 2\sum_{j=1}^n\sum_{i=1}^n u_{x_ix_j}^2 \ge 0
\]
making $v$ a sub-harmonic function.
\end{proof}

\item Prove the following propositions.
\begin{prop}\label{s29p4}
Let $u$ be continuous in $\bar{U}$ where $U \subset \mathbb{R}^n$ is a bounded set. Let $L$ be a linear operator
defined as 
\begin{equation}\label{s29e11}
L(u) = \Delta u + \sum_{i=1}^n b_i(x)u_{x_i},
\end{equation}
where $b_i(x)$ are continuous in $U$. For a function $f \ge 0$ in $U$, if $L(u) = f$ then $u$ attains its
maximum on $\partial U$.
\end{prop}
\begin{proof}
Since $u$ is continuous in $\bar{U}$ and $U$ is bounded, $u$ attains is maximum in $\bar{U}$. Let the maximum
be at $x_0 \in \Int U$. Then $L(u)(x_0) = (\Delta u)(x_0) = f \ge 0$. But this means that $u_{x_ix_i}(x_0) \ge 0$ 
for all $i = 1, \ldots, n$, which cannot be the case if $x_0$ is a maximum. Therefore, $x_0$ must be on 
$\partial U$ where $L(u) = f$ is not applicable.
\end{proof}

\begin{prop}\label{s29p5}
Continuing with the same definitions as in proposition \ref{s29p4}, if $L(u) \ge 0$ and if $u \le M$ on $\partial
U$ then $u \le M$ in $U$.
\end{prop}
\begin{proof}
Consider 
\[
L(e^{\alpha x_1}) = \alpha^2 e^{\alpha x_1} + \alpha b_1(x) e^{\alpha x_1} = \alpha e^{\alpha x_1}(\alpha +b_1(x))
\]
Since $b_1$ is continuous in the bounded set $U$, $b_1$ is itself bounded. Therefore, we can always choose
$\alpha$ such that $L(e^{\alpha x_1}) \ge 0$. For such an $\alpha$, define the function
\[
v = u + e^{\alpha x_1}
\]
so that $L(v) \ge 0$. Since $v$ too is continuous in the bounded set $U$, it has a maximum, say $N$. Let, if
possible, $v$ attain its maximum at $x_0 \in \Int U$. Then $L(v)(x_0) = (\Delta v)(x_0) \ge 0$, which is possible
only if $v_{x_ix_i} \ge 0$ for all $i = 1, \ldots, n$. But this contradicts with the fact that $v$ attains its
maximum at $x_0$. Therefore $v$ attains maximum $N$ on $\partial U$ and hence, 
\[
N = M + \max\{e^{\alpha x_1} : x_1 \in \partial U\}.
\]
Now $v \le N$ in $U$ also means that $u \le N$ in $U$. Therefore,
\[
u \le M + \max\{e^{\alpha x_1} : x_1 \in \partial U\} \in U.
\]
This also means that $u \le M$ in $U$. For if it was not so then we can choose an $\alpha$ so that the above
inequality is violated.
\end{proof}
\begin{rem}
We chose $\alpha$ at two places in the above proof. Once at the beginning and other at the end. In either cases,
choosing a greater $\alpha$ is ``better'' and hence the two choices cannot contradict each other.
\end{rem}

\item Prove the following proposition
\begin{prop}\label{s29p6}
The problem $\Delta u = f$ in $U$ and $u = g$ on $\partial U$ is translational invariant.
\end{prop}
\begin{proof}
Let $y = x + a$ where $a \in \mathbb{R}^n$ is a constant. Then
\[
\pd{u}{x} = \pd{u}{y}\pd{y}{x} \Rightarrow u_x = u_y.
\]
\end{proof}
\begin{rem}
Proposition \ref{s16p1} showed that the problem is also rotationally invariant.
\end{rem}

\item Prove the following propositions
\begin{prop}\label{s29p7}
Let $u$ solve the problem $L(u) = f$ in a bounded region $U$ and $u = g$ on $\partial U$, where $L$ is the operator
defined in equation \eqref{s29e11}. If $v$ satisfies $L(v) \le f$ in $U$ and $v \ge g$ on $\partial U$ then show
that $u \le v$ in $U$.
\end{prop}
\begin{proof}
$L(u - v) = L(u) - L(v) = f - L(v) \ge 0$. Thus $u - v$ is sub-harmonic. On the boundary $u - v = g - v \le 0$. 
Therefore, by proposition \ref{s29p1}, $u - v \le 0$ in $U$.
\end{proof}

\begin{prop}\label{s29p8}
Let $u$ solve the problem $L(u) = f$ in a bounded region $U$ and $u = g$ on $\partial U$, where $L$ is the operator
defined in equation \eqref{s29e11}. If $w$ satisfies $L(w) \ge f$ in $U$ and $w \le g$ on $\partial U$ then show
that $u \ge w$ in $U$.
\end{prop}
\begin{proof}
$L(w - u) = L(w) - L(u) = L(w) - f \ge 0$. Thus $w - u$ is sub-harmonic. On the boundary $w - u = w - g \le 0$. 
Therefore, by proposition \ref{s29p1}, $w - u \le 0$ in $U$.
\end{proof}

\item Prove the following proposition
\begin{prop}\label{s29p9}
Let $u$ satisfy the boundary value problem $\Delta u = f$ in a bounded region $U$ of $\mathbb{R}^n$ and $u = g$ on 
$\partial U$. Show that
\[
|u| \le \max_{x \in \partial U}|g| + M \max_{x \in \bar{U}}|f|,
\]
in $\bar{U}$, where $M$ is a constant.
\end{prop}
\begin{proof}
Translate the region $U$ so that $x_1 \ge 0$ in $U$. This is possible following proposition \ref{s29p6}. Since
$U$ is bounded, there exists a positive constant $a$ such that $0 \le x_1 \le a$ for all $x \in U$. Let $c_1 = 
\max_{x \in \partial U}|g|$ and $c_2 = \max_{x \in U}|f|$.

Define $v = c_1 + (e^a - e^{x_1})c_2$. Since $0 \le x_1 \le a$, $v \ge c_1$ and hence $v \ge g$ on $\partial U$. 
Further, $\Delta v = -c_2e^{x_1} \le f$ in $U$. Using proposition \ref{s29p7} we have 
\begin{equation}\label{s29e12}
u \le v \text{ in } \bar{U}.
\end{equation}

Likewise, if $w = -v$ then $w = -v \le g$ on $\partial U$ and $\Delta w = -\Delta v \ge f$ in $U$. Using 
proposition \ref{s29p8} we have
\begin{equation}\label{s29e13}
u \ge w = -v \text{ in } \bar{U} \Rightarrow -u \le v \text{ in } \bar{U}.
\end{equation}
From equations \eqref{s29e12} and \eqref{s29e13}, $|u| \le v = c_1 + (e^a - e^{x_1})c_2 \le c_1 + (e^1 - 1)c_2$.
The proposition follows from the definitions of $c_1$ and $c_2$ and with $M = e^a - 1$.
\end{proof}

\item An immediate corollary of the above proposition is
\begin{cor}
Let $u \in C^2(\overline{B(0, 1)})$ solve $-\Delta u = f$ in $B(0, 1)$, $u = 0$ on $\partial B(0,1)$. Show 
that there exists $C > 0$ such that 
\[
\max_{x \in B(0,1)} |u(x)| \le C \max_{x \in B(0,1)}|f|.
\]
\end{cor}

\item Show that the mapping
\[
x \mapsto \frac{x - e_n}{|x + e_n|},
\]
where $e_1, \ldots, e_n$ is the standard basis of $\mathbb{R}^n$, is a $C^\infty$ function mapping the upper half
space $\mathbb{R}^n_+$ to the unit ball $B(0, 1)$ in a one-to-one fashion. Further, show that the boundary
$\{x \in \mathbb{R}^n: x_n = 0\}$ is mapped to the unit sphere $\partial B(0, 1)$.

Solution: If $x = (x_1, \ldots, x_n)$ then $\mathbb{R}^n_+ = \{x \in \mathbb{R}^n : x_n \ge 0\}$ and the 
transformation is $x \mapsto y$ is
\begin{eqnarray*}
y_i &=& \frac{x_i}{\sqrt{x_1^2 + \cdots + (x_n + 1)^2}} \text{ for } i = 1, \ldots, n - 1 \\
y_n &=& \frac{x_n - 1}{\sqrt{x_1^2 + \cdots + (x_n + 1)^2}}  
\end{eqnarray*}
so that
\begin{equation}\label{s29e14}
y^2 = \frac{x^2 - 2x_n + 1}{x^2 + 2x_n + 1}.
\end{equation}
If $x_n \ge 0$, $y^2 \le 1$ so that $\mathbb{R}^n_+$ is mapped to $B(0, 1)$. Equation \eqref{s29e14} also shows 
that when $x_n = 0$, $y^2 = 1$. Thus, the boundary of $\mathbb{R}^n_+$ maps to $\partial B(0, 1)$. The mapping is
also in $C^\infty$ because the only point where the denominator vanishes is $-e_n$ which is not in 
$\mathbb{R}^n_+$.

In order to show that the mapping is one-to-one, I will show that it is invertible. When none if $x_i$ are
zero, we can express the mapping in matrix form as
\[
\frac{1}{|x + e_n|}\begin{pmatrix}
1 & 0 & \ldots & 0 \\
0 & 1 & \ldots & 0 \\
\vdots & \vdots & \ldots & \vdots \\
-((n-1)x_1)^{-1} & -((n-1)x_2)^{-1} & \ldots &1 
\end{pmatrix}\begin{pmatrix}
x_1 \\
x_2 \\
\vdots \\
x_n
\end{pmatrix} = \begin{pmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{pmatrix}
\]
or
\[
Tx = y
\]
The transformation matrix $T$ is non-singular and hence the mapping is invertible. If one or more $x_i$ are zero 
then we set $y_i = 0$ and use the same transformation for the other components of $x$.

We also note that
\[
T^{-1} = \beta \begin{pmatrix}
1 & 0 & \ldots & 0 \\
0 & 1 & \ldots & 0 \\
\vdots & \vdots & \ldots & \vdots \\
((n-1)\beta y_1)^{-1} & ((n-1)\beta y_2)^{-1} & \ldots &1 
\end{pmatrix}
\]
where $\beta = |x + e_n|$. However, to find the inverse transformation, we also need to write $\beta$ in terms of
$y$. From 
\eqref{s29e14},
\[
\beta^2 y^2 = x^2 - 2x_n + 1 = x^2 -2(\beta y_n + 1) + 1
\]
so that
\begin{equation}\label{s29e15}
x^2 = \beta^2 y^2 + 2\beta y_n + 1
\end{equation}
Since 
\begin{equation}\label{s29e16}
2x_n = 2\beta y_n + 2
\end{equation}
from equation \eqref{s29e15} and \eqref{s29e16} we have
\[
\beta^2 = x^2 + 2x_n + 1 =  \beta^2 y^2 + 4\beta y_n + 4
\]
or
\[
\beta^2(1 - y^2) -4y_n\beta - 4 = 0
\]
or
\[
\beta = \frac{4y_n \pm \sqrt{16y_n^2 + 16(1 - y^2)}}{2(1 - y^2)} = \frac{y_n \pm \sqrt{y_n^2+(1-y^2)}}{(1-y^2)}
\]
Since $\beta$ is required to be positive, 
\begin{equation}\label{s29e17}
\beta = \frac{y_n + \sqrt{y_n^2+(1-y^2)}}{(1-y^2)}
\end{equation}

\item Let $U$ be an open bounded set in $\mathbb{R}^n$ and $u \in C^2(U) \cap C^0(\bar{U})$ satisfy $\Delta u = -1$
in $U$ and $u = 0$ on $\partial U$. Show that $u(x) \ge d^2(x, \partial U)/(2n)$ for all $x \in U$.

Solution: Let $v(x) = u(x) + |x - a|^2/(2n)$ for a fixed $a \in U$. Then $\Delta v = 0$ in $U$ and $v(x) = 
|x - a|^2/(2n)$ on $\partial U$. For a fixed $a \in U$, $v(x) \ge d^2(a, \partial U)/(2n)$ on $\partial U$. Using
the strong minimum principle, 
\[
v(a) \ge \min_{x \in \bar{U}}v(x) = \min_{x \in \partial U}v(x) \ge \frac{d^2(a, \partial U)}{2n}
\]
or
\[
u(a) \ge \frac{d^2(a, \partial U)}{2n}
\]
for all $a \in U$.

\item If $u$ is a harmonic function in $\mathbb{R}^n$ satisfying $|u(x)| \le C(1 + |x|^k)$ for some non-negative
integer $k$ and all $x \in \mathbb{R}^n$ show that $u$ is a polynomial of degree at most $k$.

Solution: The Poisson formula for $B(0,r)$ is
\[
u(x) = \frac{r^2 - |x|^2}{nV(n, 1)r}\int_{\partial B(0,r)} \frac{g(y)}{|x - y|^n}dS(y)
\]
for $x \in \mathbb{R}^n$ and $u$ be the solution of $\Delta u = 0$ in $B(0,1)$ and $u = g$ on $\partial B(0,1)$. It
therefore follows that
\[
D^\alpha_x u(x) = \frac{1}{nV(n,1)r}\int_{\partial B(0,r)} g(y)D^\alpha_x \frac{r - |x|^2}{|x - y|^n} dS(y)
\]
The function $(r - |x|)^2/|x - y|^n$ goes as $1/|x|^{n-1}$ as $|x| \rightarrow \infty$. Therefore its $\alpha$-th
derivative goes as $1/|x|^{|\alpha|+n-1}$. We are now given that $|u(x)| = |g(x)| \le C(1 + |x|^k)$ so that 
\[
D^\alpha u(x) \le C\frac{r^{k}}{r^{|\alpha|}}
\]
Thus, in the limit $r \rightarrow \infty$, $D^\alpha(x) \rightarrow 0$ if $|alpha| > k$. Since all partial 
derivatives of order $|alpha| > k$ vanish as $|x| \rightarrow \infty$, $u$ must be a polynomial of order at most
$k$,
\end{enumerate}

\section{Miscellaneous}\label{s30}
We will prove the Green-Gauss theorem for convex regions. A proof for arbitrary regions is best given in terms of
Stokes theorem expressed using differential forms.
\begin{thm}\label{s30t1}
Let $U$ be a convex region in $\mathbb{R}^n$ and let $u: \mathbb{R}^n \rightarrow \mathbb{R}$ be $C^1(\bar{U})$.
Then 
\[
\int_U u_{x_i}dx = \int_{\partial U} un_i dS,
\]
where $n$ is the outward normal to $\partial U$.
\end{thm}
\begin{proof}
Since $U$ is convex, we can find $f_1, f_2 \in C^1(\mathbb{R}^n)$ such that
\[
f_1(x_1, \ldots, \hat{x}_i, \ldots, x_n) \le x_i \le f_2(x_1, \ldots, \hat{x}_i, \ldots, x_n)
\]
for all $x \in U$. For every $x \in \mathbb{R}^n$, $(x_1, \ldots, \hat{x}_i, \ldots, x_n)$ denotes a vector in 
$\mathbb{R}^{n-1}$ with the $i$th component removed. We can write the volume integral as
\[
\int_U u_{x_i}dx = \int_{D}\int_{f_1}^{f_2} u_{x_i}dx_i dS,
\]
or
\begin{equation}\label{s30e1}
\int_U u_{x_i}dx = 
\int_{D}\left(u(x_1, \ldots, f_2, \ldots, x_n) - u(x_1, \ldots, f_1, \ldots, x_n)\right) dS
\end{equation}
where we emphasize that $f_1$ and $f_2$ are not constants but functions of $n-1$ variables and $D$ is the 
projection of $U$ on the hyperplane $x_1 \ldots \hat{x}_i \ldots x_n$. Let $S_1, S_2, S_3$ be portions of the
surface $\partial U$ on which $n_i$ is positive, zero and negative. Then
\begin{eqnarray}
\int_{S_1} u n_i dS &=& +\int_D u(x_1,\ldots,f_2(x_1,\ldots,\hat{x}_i,\ldots,x_n),\ldots,x_n)dS \label{s30e2} \\
\int_{S_2} u n_i dS &=& 0 \label{s30e3} \\
\int_{S_3} u n_i dS &=& -\int_D u(x_1,\ldots,f_1(x_1,\ldots,\hat{x}_i,\ldots,x_n),\ldots,x_n)dS \label{s30e4}
\end{eqnarray}
If we sum equations \eqref{s30e2} to \eqref{s30e4}, the left hand side is the surface integral of $u n_i$ over
$\partial U$ and the right hand side is equal to the right hand side of \eqref{s30e1}.
\end{proof}

There is another way to prove theorem \ref{s30t1} using Gauss Divergence theorem.
\begin{proof}
Applying Gauss divergence theorem for a vector field $F = \alpha u(x)$, where $\alpha$ is a constant vector with
no component vanishing and $u : \mathbb{R}^n \rightarrow \mathbb{R}$ is $C^{1}(U)$, we get
\[
\int_U \alpha_i u_{x_i}dx = \int_{\partial U}\alpha_i u n_i dS
\]
The theorem follows from the fact that $\alpha_i$ is non-zero. 
\end{proof}
\bibliographystyle{plain}
\bibliography{CourseNotes}
\end{document}

### 2020-03-29 p3/cubic_spline.py

@@ -1,0 +1,42 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import math
+
+N = 128
+
+def mu(x):
+  y = np.ones(len(x))
+  # Select indices of x where x[i] < 0.
+  si = np.where(x < 0)[0]
+  y[si] = 0
+  return y
+
+lo = -2.5
+hi = 2.5
+n_samples = N
+
+x = np.linspace(lo, hi, n_samples)
+# We will call the five terms of \beta^3(x) b_0, b_1, b_2, b_3.
+b_0 = np.multiply(pow(x + 2, 3), mu(x + 2))/6
+b_1 = -2/3 * np.multiply(pow(x + 1, 3), mu(x + 1))
+b_2 = np.multiply(pow(x, 3), mu(x))
+b_3 = -2/3 * np.multiply(pow(x - 1, 3), mu(x - 1))
+b_4 = np.multiply(pow(x - 2, 3), mu(x - 2))/6
+
+beta_3 = b_0 + b_1 + b_2 + b_3 + b_4
+alpha = math.sqrt(3) - 2
+K = -6.0*alpha/(1 - alpha*alpha)
+
+truncate_at = 20
+eta = np.zeros(len(x) - 2*truncate_at)
+
+for i in range(0, len(eta)):
+    sum = 0
+    for j in range(-truncate_at, truncate_at):
+        sum += pow(alpha, np.abs(j)) * beta_3[i - j]
+
+    eta[i] = sum
+
+eta = K * eta
+plt.plot(x[truncate_at:(128-truncate_at)], eta)
+plt.show()

### 2020-03-29 p3/dbs.py

@@ -1,0 +1,37 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+# We want to consider signals for n = -5, -4, ..., 4, 5.
+# Note that we want to simulate negative indexes. We do
+# so by translating an index to its array position. An
+# index i of the signal translates to (i + EXTENT) in the
+# array. Thus, the signal indexes -5, -4, ..., 4, 5 map to
+# array indexes 0, 1, ..., 10.
+EXTENT = 5
+
+# The vectors b_1^0 and c_1^0.
+b = np.zeros(2 * EXTENT + 1)
+c = np.zeros(2 * EXTENT + 1)
+
+# Initial values
+b[0 + EXTENT] = 1
+c[0 + EXTENT] = 1
+
+def get_b1(k):
+    i = k + 5 # index in the array
+    return (k + 1) * c[i] + (1 - k) * c[i - 3]
+
+def get_c1(k):
+    i = k + 5
+    return (k + 3/2) * b[i + 1] + (1/2 - k)*b[i]
+
+# The vectors b_1^1 and c_1^1.
+B = np.zeros(2 * EXTENT + 1)
+C = np.zeros(2 * EXTENT + 1)
+for k in range(-3, 3):
+    B[k + EXTENT] = get_b1(k)
+    C[k + EXTENT] = get_c1(k)
+
+print(B)
+print(C)
+              

### 2020-03-29 p3/fft.py

@@ -1,0 +1,32 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+N = 128
+
+def mu(x):
+  y = np.ones(len(x))
+  # Select indices of x where x[i] < 0.
+  si = np.where(x < 0)[0]
+  y[si] = 0
+  return y
+
+lo = -2.5
+hi = 2.5
+n_samples = N
+
+x = np.linspace(lo, hi, n_samples)
+# We will call the five terms of \beta^3(x) b_0, b_1, b_2, b_3.
+b_0 = np.multiply(pow(x + 2, 3), mu(x + 2))/6
+b_1 = -2/3 * np.multiply(pow(x + 1, 3), mu(x + 1))
+b_2 = np.multiply(pow(x, 3), mu(x))
+b_3 = -2/3 * np.multiply(pow(x - 1, 3), mu(x - 1))
+b_4 = np.multiply(pow(x - 2, 3), mu(x - 2))/6
+
+beta_3 = b_0 + b_1 + b_2 + b_3 + b_4
+bf = np.abs(np.fft.fft(beta_3))
+bf1 = np.fft.fftshift(bf)/np.max(bf)
+plt.plot(x[44:84], bf1[44:84], label = r'$\tilde{\beta}^3$')
+##plt.plot(x, beta_3, label = r'$\beta^3(x)$')
+plt.title('Cubic spline and its Fourier transform')
+plt.legend()
+plt.show()

### 2020-03-29 p3/splines.py

@@ -1,0 +1,56 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+def mu(x):
+  y = np.ones(len(x))
+  # Select indices of x where x[i] < 0.
+  si = np.where(x < 0)[0]
+  y[si] = 0
+  return y
+
+lo = -2.5
+hi = 2.5
+n_samples = 1001
+
+x = np.linspace(lo, hi, n_samples)
+
+# Compute the two terms of \beta^0(x)
+b_0 = mu(x + 1/2)
+b_1 = mu(x - 1/2)
+
+beta_0 = b_0 - b_1
+
+# Computer the three terms of \beta^1(x)
+b_0 = np.multiply((x+1), mu(x+1))
+b_1 = np.multiply(x, mu(x))
+b_2 = np.multiply((x-1), mu(x-1))
+
+beta_1 = b_0 - 2*b_1 + b_2
+
+# Compute the four terms of \beta^2(x)
+b_0 = np.multiply(pow(x + 3/2, 2), mu(x + 3/2))/2
+b_1 = -3 * np.multiply(pow(x + 1/2, 2), mu(x + 1/2))/2
+b_2 = 3 * np.multiply(pow(x - 1/2, 2), mu(x - 1/2))/2
+b_3 = -np.multiply(pow(x - 3/2, 2), mu(x - 3/2))/2
+
+beta_2 = b_0 + b_1 + b_2 + b_3
+
+# We will call the five terms of \beta^3(x) b_0, b_1, b_2, b_3.
+b_0 = np.multiply(pow(x + 2, 3), mu(x + 2))/6
+b_1 = -2/3 * np.multiply(pow(x + 1, 3), mu(x + 1))
+b_2 = np.multiply(pow(x, 3), mu(x))
+b_3 = -2/3 * np.multiply(pow(x - 1, 3), mu(x - 1))
+b_4 = np.multiply(pow(x - 2, 3), mu(x - 2))/6
+
+beta_3 = b_0 + b_1 + b_2 + b_3 + b_4
+
+p = plt.plot(x, beta_0, label = r'$\beta^0(x)$')
+p = plt.plot(x, beta_1, label = r'$\beta^1(x)$')
+p = plt.plot(x, beta_2, label = r'$\beta^2(x)$')
+p = plt.plot(x, beta_3, label = r'$\beta^3(x)$')
+p = plt.xlabel('x')
+p = plt.ylabel(r'$\beta^n(x)$')
[…]

### 2020-07-06 p5/proposal.tex
Initial commit.

\documentclass{article}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{bm}
\begin{document}
\title{Verifying research using ODEs in NLP}
\maketitle
The difference equation to be examined is
\begin{equation}\label{e1}
h_t = (1 - z_{t-1})\odot h_{t-1} + z_t \odot \hat{h}_t,
\end{equation}
where
\begin{eqnarray}
\hat{h}_t &=& \tanh(W \odot (r_t \odot h_{t - 1} + x_t) \label{e2} \\
r_t &=& \sigma\left(w^l x_t + U^l h_{t - 1}\right) \label{e3} \\
z_t &=& \sigma\left(w^z x_t + U^z h_{t - 1}\right) \label{e4}
\end{eqnarray}
Now suppose I want to investigate equation \eqref{e1} and I start writing a
program to do so. It is clear that I need the following:
\begin{enumerate}
\item $h_0$, the initial value. Now thankfully, this difference equation
is a discrete version of a differential equation. We hope that the ODE is
stable and therefore where we start does not make a difference. (Actually
it does make a difference when $h$ has more than one dimension. But for
the moment, let's gloss it over.) Let us choose $h_0 = 0$.
\item $z_0$. We cannot use equation \eqref{e4} to proceed because it will
then demand $h_{-1}$ from us and there is no $h$ before `initial time'. Let
us make another assumption $z_0 = 0$ to permit us to proceed.
\item $x_t$, there is nothing that tells us what this is.
\item $w^l, w^z, U^l, U^z, W$, we need to know what these objects are before
we proceed.
\end{enumerate}
You now see that we have $5$ unknowns $w^l, w^z, U^l, U^z, W$. We do not 
even know whether they are vectors or matrices, constants or variables. 
More seriously, we do not know that $x_t$ is. We need it for all $t \ge 0$.

I think we are not in a stage where we can conduct a mathematical analysis
of difference equations. We must define our objects precisely before we
proceed.

\end{document}

### 2020-07-06 p5/stability_rnn.tex
Initial commit.

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Dan Will}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to derive criteria for
the stability of recurrent neural networks.

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol.

\section{Ordinary Differential Equations of the first order}\label{s3}
An ordinary differential equation, abbreviated as ODE, of the first order 
is an equation of the form
\begin{equation}\label{s3e1}
\frac{dy}{dx} = f(x, y),
\end{equation}
where $x \in \mathbb{R}$, $y$ is the real-valued unknown function, $f: 
\mathbb{R}^2 \mapsto \mathbb{R}$. Note that $f$ depends on both $x$ and 
$y$. It called a first order ODE because the highest order derivative
of the unknown function $y$ with respect to $x$ is of the first order.
This equation is usually accompanied with the initial value of $y$ at a 
certain $x$, say $x = x_0$. Equation \eqref{s3e1} and $y(x_0) = y_0$ is
called an \emph{inital value problem}.

The oldest method to solve equation \eqref{s3e1} numerically was devised 
by Euler in the eighteenth century. It follows straight from the definition
of the first derivative of a function. If we know the value of $y$ at $x_0$,
we can approximate the value of $y$ at a neighbouring point $x_0 + h$ as
\begin{equation}\label{s3e2}
y(x_0 + h) = y(x_0) + h\left[\frac{dy}{dx}\right]_{x_0}.
\end{equation}
But the value of the derivative of $y$ with respect to $x$ is given by
the differential equation itself. Thus,
\begin{equation}\label{s3e3}
y(x_0 + h) = y(x_0) + h f(x_0, y(x_0)).
\end{equation}
In general, $n$ steps beyond the first, one can write
\begin{equation}\label{s3e4}
y(x_n + h) = y(x_{n}) + h f(x_{n}, y(x_{n})).
\end{equation}
Euler's algorithm is not the best numerical algorithm to solve ODE's but
it provides the germ for the the rest.

\section{RNN with a Single Layer}\label{s4}
Recurrent neural networks, abbreviated as RNN, feed a part of their output
to their input. The output if a step $n$ thus depends on both the input
at that step and the output at the previous step. This makes them similar
to the ordinary differential equations described in the previous section.
Suppose we have a single layer neural network that is updated using the
following rule:
\begin{equation}\label{s4e1}
y_{n+1} = \sigma(w_x x_{n} + w_y y_n),
\end{equation}
where $w_x$ and $w_y$ are weights and $\sigma$ is an activation function. 
If we now identify the correspondence
\begin{eqnarray}
y_{n+1} &\rightarrow& y(x_n + h) \\
y_{n} &\rightarrow& y(x_n) \\
\sigma(w_x x_n + w_y y_n) &\rightarrow& y(x_n) + hf(x_n, y(x_n))
\end{eqnarray}
then we can spot a striking resemblance between Euler's algorithm and an
RNN. The idea that an RNN corresponds to an ODE and therefore the stability
of an RNN can be studied using the stability of the underlying ODE is the
topic of this article.

We need not interpret equation \eqref{s3e1} as a scalar equation. We can
as well let $y$ and $f$ be vector functions. That is, $y:\mathbb{R}^p
\mapsto \mathbb{R}^q$, $x \in \mathbb{R}^p$ and $f: \mathbb{R}^{p+q}
\mapsto \mathbb{R}^q$. In this case, equation \eqref{s3e4} becomes
\begin{equation}\label{s4e5}
y(x_n + h) = y(x_n) + \nabla y(x_n)\cdot h.
\end{equation}
Here
\begin{equation}\label{s4e6}
\nabla y = \begin{pmatrix} y_{1, x_1} & \ldots & y_{1, x_p} \\
	   \vdots \\
	   y_{q, x_1} & \ldots & y_{q, x_p}
	   \end{pmatrix}
\end{equation}
is an $q \times p$ matrix and $h \in \mathbb{R}^p$ so that their product
$\nabla y(x_n) \cdot h$ is in $\mathbb{R}^q$. An RNN corresponding to this
equation is
\begin{equation}\label{s4e7}
y_{n+1} = \sigma\left(W_x x_n + W_y y_n\right),
\end{equation}
where $y_n$ and $y_{n+1}$ are $q$-dimensional vectors, $x_n \in \mathbb{R}
^p$, $W_x$ is an $q \times p$ weight matrix for the input and $W_y$ is an
$q \times q$ weight matrix for the output.

\section{Higher order ODE}\label{s5}
A system of first order ODE is a set of equations of the form
\begin{eqnarray*}
\frac{dy_1}{dx} &=& f_1(x, y_1, \ldots, y_k) \\
\vdots \\
\frac{dy_k}{dx} &=& f_k(x, y_1, \ldots, y_k) \\
\end{eqnarray*}
It can be written in a vector form as
\begin{equation}\label{s5e1}
\frac{dy}{dx} = f,
\end{equation}
where
\[
y = \begin{pmatrix}y_1 \\ \vdots \\ y_k\end{pmatrix},\;\text{and}\;
f = \begin{pmatrix}f_1 \\ \vdots \\ f_k\end{pmatrix}
\]
A system of first order ODE is interesting because they are equivalent to
a higher order differential equation. For example, one can write the $k$th
order ODE
\begin{equation}\label{s5e2}
\frac{d^ky}{dx^k} = f\left(x, y, \frac{dy}{dx}, \ldots, 
	\frac{d^{k-1}y}{dx^{k-1}}\right)
\end{equation}
as the system
\begin{eqnarray*}
\frac{dy_{k-1}}{dx} &=& f(x, y, y_1, \ldots, y_{k-1}) \\
\frac{dy_{k-2}}{dx} &=& y_{k-1} \\
\vdots \\
\frac{dy}{dx} &=& y_1
\end{eqnarray*}
If this equation is also accompanied by an initial value then we can solve
it numerically using equations \eqref{s4e5}, \eqref{s4e6} and \eqref{s4e7}.

\section{Runge-Kutta methods}\label{s6}
Runge-Kutta methods are a collection of numerical algorithms developed by
the German mathematicians Carl Runge and Martin Kutta in the late 
nineteenth centuries\cite{lambert1991numerical}. We will describe the most
commonly used among them, called the `RK4' method. If $y: \mathbb{R}
\mapsto \mathbb{R}$ is the unknown function that obeys the ODE
\begin{equation}\label{s6e1}
\frac{dy}{dt} = f(t, y), y(t_0) = y_0
\end{equation}
then one can get the value oy $y$ at $t_n + h$ using the formula
\begin{equation}\label{s6e2}
y(t_n + h) = y(t_n) + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4).
\end{equation}
The constants $k_1, \ldots, k_4$ are given by
\begin{eqnarray}
k_1 &=& f(t_n, y(t_n)) \label{s6e3} \\
k_2 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_1}{2}\right) 
 \label{s6e4} \\
k_3 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_2}{2}\right) 
 \label{s6e5} \\
k_4 &=& f(t_n + h, y_n + hk_3). \label{s6e6}
\end{eqnarray}
Let us study these equations carefully. The value of $y$ at $t_n + h$ 
depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Of these,
$k_1$ depends on $y(t_n)$ and $t_n$, $k_2$ on $y(t_n)$, $t_n$ and $k_1$,
$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. 
Runge-Kutta methods perform much better then Euler methods when $f$ is 
a non-linear function of $x$. The number of intermediate values, $4$ in 
this case, is called the \emph{the number of stages} of the algorithm.

\section{RNN with Hidden Layers}\label{s7}
Let us try to interpret the `RK4' algorithm as a neural network. It is 
clearly a recurrent neural network because the output at `time' $t_n + h$
depends on that at `time' $t_n$. If we write $y(t_n) = y_n$ then we can
say that $y_{n+1}$ depends on $y_n$. However, it does not depend on $y_n$
alone but on four other numbers which must be computed sequentially. $k_1$
is computed from $t_n$ and $y_n$. The rest from $t_n$, $y_n$ and the 
`previous' $k$. We can thus construct a neural network with four hidden 
layers, each computing $k_1$ to $k_4$ and the output later combining
all of them to get the final result. This is a $4$-stage, $1$st order
RNN.

We now define a general RNN\cite{niu2019recurrent}.
\begin{defn}\label{s7d1}
An ODE-RNN of $n$th order in non-linearity and $t$th order in gradient,
called $(n,t)$-ODERNN, where $n, t \ge 1$ are integers is described by
the following update rule.
\begin{eqnarray}
K_1 &=& \sigma_1(W_1Y_l + b_1) \label{s7e1} \\
K_q &=& \gamma_q K_{q-1} + \kappa_1\sigma_q\left(
 W_qY_{l-t_{q-1}}+b_q+h\sum_{k=1}^{q-1}\alpha_{q,k}K_k\right)\label{s7e2}\\
Y_{l+1} &=& \gamma_{n+1}Y_l + \kappa_{n+1}\sigma_{q+1}\left(
 W_{n+1}Y_{l-t_{n-1}}+b_{n+1}+h\sum_{k=1}^n\beta_kK_k
 \right)\label{s7e3},
\end{eqnarray}
where
\begin{enumerate}
\item $q = 1, \ldots, n$. This explains the $n$ in the $(n,t)$-ODERNN.
\item The right hand side of equations \eqref{s7e1} to \eqref{s7e3} 
depend on $Y_l$ to $Y_{l - t_{n-1}}$, a total of $n$ previous values.
Here each $t_k = \lfloor tk/n\rfloor$. Because $t_k$ for each hidden layer
is defined in terms of $t$, we also specify it in the name of the ODE. 
This explain $t$ in $(n, t)$-ODERNN.
\item $Y_l \in \mathbb{R}^s$, $l \in \mathbb{Z}$;
\item $W_1,\ldots, W_{n+1}$ are $p \times s$ weight matrices;
\item $b_1,\ldots, b_{n+1} \in \mathbb{R}^p$ are bias vectors;
\item $\alpha_{q, k}, \beta_k$, for $q = 2, \ldots, n$ and $k = 1, \ldots,
n$, $\gamma_2, \ldots, \gamma_{n}, \kappa_2, \ldots, \kappa_{n}$ are 
$p \times p$ matrices;
\item $\gamma_{n+1}$ is an $s \times s$ matrix and $\kappa_{n+1}$ is a 
$s \times p$ matrix.
\item $h \in \mathbb{R}$ is a constant and
\item $\sigma_1, \ldots, \sigma_{q+1}$ are activation function defined as
\[
\sigma_i\begin{pmatrix}x_1 \\ \vdots \\ x_p\end{pmatrix} = 
\begin{pmatrix}\sigma_i(x_1) \\ \vdots \\ \sigma_i(x_p)\end{pmatrix},
i = 1, \ldots, q+1.
\]
\end{enumerate}
\end{defn}
\noindent Before proceeding, a few remarks are in order.
\begin{rem}
$K_1, \ldots, K_q$ are all $p$-dimensional vectors. That is, they 
are all members of $\mathbb{R}^p$.
\end{rem}
\begin{rem}
We can readily confirm that the arguments to the activation functions
$\sigma_1, \ldots, \sigma_{q+1}$ are all $p$-dimensional vectors.
\end{rem}
\begin{rem}
The number of hidden layers is $n$. The RNN thus has $n + 2$ layers. 
\end{rem}

\noindent We also define
\begin{defn}\label{s7d2}
The Burrage-Butcher tensor corresponding to the $(n,t)$-ODERNN of 
definition \ref{s7d1} is
\begin{equation}\label{x}
Q_{i, j} = \beta_i\alpha_{i, j} + \beta_j\alpha_{j, i} - \beta_i\beta_j^T.
\end{equation}
\end{defn}

\noindent Let us consider a few examples of $(n,t)$-ODERNN.
\begin{enumerate}
\item $(2,2)$-ODERNN. It has the composition rules
\begin{eqnarray*}
K_1 &=& \sigma_1(W_1 Y_l + b_1) \\
K_2 &=& \gamma_2K_1 + \kappa_2\sigma_2(W_2Y_{l-1}+b_2+h\alpha_{2,1}K_1) \\
Y_{l+1} &=& \gamma_3Y_l + \kappa_3\sigma_3\left(
 W_3 Y_{l-1} + b_3 + h(\beta_1K_1 + \beta_2K_2)\right)
\end{eqnarray*}
\item $(4, 2)$-ODERNN. It has $4$ hidden layers and it uses $2$ previous
inputs. It has the composition rules
\begin{eqnarray*}
K_1 &=& \sigma_1(W_1 Y_l + b_1) \\
K_2 &=& \gamma_2K_1 + \kappa_2\sigma_2(W_2Y_{l-1}+b_2+h\alpha_{2,1}K_1) \\
K_3 &=& \gamma_3K_2 + \kappa_3\sigma_3\left(W_3Y_{l-1} + b_3 + 
    h\sum_{k=1}^{2}\alpha_{3, k}K_k\right) \\
K_4 &=& \gamma_4K_2 + \kappa_4\sigma_4\left(W_4Y_{l-2} + b_4 + 
    h\sum_{k=1}^{3}\alpha_{4, k}K_k\right) \\
Y_{l+1} &=& \gamma_5Y_l + \kappa_5\sigma_5\left(W_5Y_2 + b_5 + 
    h\sum_{k=1}^{4}\alpha_{5, k}K_k\right) 
\end{eqnarray*}
\end{enumerate}

\section{Gated Recurrent Unit as ODERNN}\label{s8}
Consider a single layer of a Gated Recurrent Unit (GRU). Its composition
rule is\cite{gereon2018hands}
\begin{eqnarray}
\begin{pmatrix} r_t \\ z_t \end{pmatrix} &=&
\sigma
\begin{pmatrix}
    W_r X_t + U_r Y_{t-1} + b_z \\ 
    W_rX_t + U_rY_{t-1} + b_r
\end{pmatrix} \label{s8e1} \\
Y_t &=& (1 - z_t) \circ Y_{t-1} + 
 z_t \circ \tanh(W_hX_t + U_h(r_t \circ Y_{t-1} + b_h)) \label{s8e2}
\end{eqnarray}
where $\circ$ denotes the element-wise product, also called the Hadamard
product of vectors. There are some differences between this composition
rule and the one mentioned in the paper\cite{niu2019recurrent} by Niu and 
others. We prefer these because the equations they are more general than
the equations (22) and (23) of Niu's paper.

We now introduce the functions
\begin{eqnarray}
\sigma_t^\prime(X) &=& r_t \circ X \label{s8e3} \\
\sigma_t^{\prime\prime}(X) &=& z_t \circ \tanh(X). \label{s8e4}
\end{eqnarray}
The functions $\sigma^\prime$ and $\sigma^{\prime\prime}$ allow us to 
write the composition rule of equation \eqref{s8e2} as
\begin{equation}\label{s8e5}
Y_t = (1 - z_t) \circ Y_{t-1} + \sigma_t^{\prime\prime}\left(
	W_tY_{t-1} + U_h(\sigma_t^\prime(Y_{t-1}) + b_h)\right).
\end{equation}
If $x$ and $y$ are vectors then we can always express their Hadamard 
product as a matrix operation
\begin{equation}\label{s8e6}
x \circ y = D[x]y,
\end{equation}
where $D[x]$ is a diagonal matrix whose diagonal entries as same as $x$.
Using this transformation, we observe that equation \eqref{s8e5} has the
same structure as \eqref{s7e3}. Further, there are two hidden layers, 
one each for $r_t$ and $z_t$. This makes the GRU of equations \eqref{s8e1}
and \eqref{s8e2} a $(2, 1)$-ODERNN.

\section{Stability analysis}\label{s9}
RNNs frequently have a problem of `exploding gradients'
\cite{gereon2018hands}. In the context of ODEs this is called an 
instability. Because of a correspondence between and RNN and an ODE, we 
can study the problem of exploding gradients in the former as the 
instability in the latter. In this section we will review certain notions
of stability of ODE.

Consider the initial value problem of equation \eqref{s3e1}. The general
$k$-step method to get its approximate numerical solution is defined by
the formula
\begin{equation}\label{s9e1}
\sum_{i=0}^k \alpha_i x_{n+i} = h\sum_{i=0}^k \beta_i f_{n+i},
\end{equation}
where $\alpha_i, \beta_i$ are real constants.
\begin{defn}\label{s9d1}
The $k$-step method defined above is said to be A-stable
\cite{dahlquist1963special} if all solutions of \eqref{s9e1} tend to zero 
as $n \rightarrow \infty$ when the method is applied to the equation
\[
    \frac{dy}{dx} = qx
\]
where $q \in \mathbb{C}$ with $\Re(q) < 0$ and a fixed $h$ in equation
\eqref{s9d1}.
\end{defn}

In order to introduce the idea of B-stability and BN-stability, we need
the definition of contractive solutions\cite{lambert1991numerical}.
\begin{defn}\label{s9d2}
Let $y_1$ and $y_2$ be any two solutions of \eqref{s3e1} satisfying the
initial conditions $y_1(a) = \eta_1, y_2(a) = \eta_2, \eta_1 \ne \eta_2$.
Then if $||y_1(x_2) - y_2(x_2)|| \le ||y_1(x_1) - y_2(x_1)||$ for all
$a \le x_1 \le x_2 \le b$ then the solutions are said to be contractive
in $[a, b]$.
\end{defn}
A contractive solutions get closer to each other as one gets away from
the initial point.

\begin{defn}\label{s9d3}
An ODE of the form \eqref{s3e1} is said to be autonomous if the right hand
side is independent of $x$.
\end{defn}

We are now in a position to define B and BN stability
\cite{lambert1991numerical}.
\begin{defn}\label{s9d4}
If a Runge-Kutta method is applied to an autonomous system of ODE such that
\[
    \langle f(x, y_1) - f(x, y_2), y_1 - y_2\rangle \le 0
\]
generates contractive numerical solutions then the method is said to be
B-stable. If the same is true for a non-autonomous system then the method
is said to be $BN$-stable.
\end{defn}
\begin{rem}
$\langle v_1, v_2 \rangle$ is just the inner product of the two vectors.
\end{rem}

It is also mentioned in Lambert's book \cite{lambert1991numerical} that
\begin{thm}\label{s9t1}
Runge-Kutta method is $BN$-stable if the matrices $\beta_n$ (equation
\eqref{s7e3}) and the Burrage-Butcher tensor are positive semi-definite.
\end{thm}

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because there
is only a single hidden layer. The Burrage-Butcher tensor is thus $-\beta_1
\beta_1^T$, which is \emph{not} positive semidefinite unless $\beta_1 = 0$.
Thus the GRU is not a stable algorithm, an observation also corroborated by
elsewhere\cite{kanai2017preventing}.
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}

### 2020-07-06 p3/direct_spline.py

@@ -1,4 +1,7 @@
 def b(n, m, k):
+    if m == 0:
+        raise ValueError('m cannot be zero')
+
     rv = 0; # Return value.
     if n == 0:
         if -m/2 <= k and k <= m/2:
@@ -7,11 +10,12 @@ def b(n, m, k):
         rv = (k/m + (n+1)/2)*c(n-1, m, k) + ((n+1)/2 - k/m)*c(n-1,m,k-m)
         rv = rv/n
 
-    print(f'b({n},{m},{k}) = {rv}')
-
     return rv
 
 def c(n, m, k):
+    if m == 0:
+        raise ValueError('m cannot be zero')
+
     rv = 0; # Return value.
     if n == 0:
         if m - 2 <= k and k <= 0:
@@ -20,18 +24,5 @@ def c(n, m, k):
         rv = (-k/m + (n+2)/2)*b(n-1, m, k-m) + (n/2 - k/m)*b(n-1,m,k)
         rv = rv/n
 
-    print(f'c({n},{m},{k}) = {rv}')
-
     return rv
 
-
-
-##print("b(0,1,0) = ", b(0,0,0))
-##print("c(0,1,0) = ", c(0,1,0))
-##print("b(1,1,0) = ", b(1,1,0))
-##print("c(1,1,0) = ", c(1,1,0))
-##print("c(1,1,-1) = ", c(1,1,-1))
-##print("c(1,1,1) = ", c(1,1,1))
-##print("b(2,1,1) = ", b(2,1,1))
-##print("b(2,1,0) = ", b(2,1,0))
-##print("b(2,1,-1) = ", b(2,1,-1))

### 2020-07-06 p5/stability_rnn.tex
First review

@@
 We follow the contemporary mathematical practice of not using separate
-notation for vectors and scalars. Instead, we define every symbol.
+notation for vectors and scalars. Instead, we define every symbol. Whenver
+a symbol stands for a vector, it denotes a column vector.
 
@@
 Euler's algorithm is not the best numerical algorithm to solve ODE's but
-it provides the germ for the the rest.
+it provides the germ for the rest.
 
@@
 Recurrent neural networks, abbreviated as RNN, feed a part of their output
-to their input. The output if a step $n$ thus depends on both the input
-at that step and the output at the previous step. This makes them similar
+to their input. The output of a step $n$ thus depends on both the input
+at that step and the output of the previous step. This makes them similar
 to the ordinary differential equations described in the previous section.
@@
 is an $q \times p$ matrix and $h \in \mathbb{R}^p$ so that their product
-$\nabla y(x_n) \cdot h$ is in $\mathbb{R}^q$. An RNN corresponding to this
-equation is
+$\nabla y(x_n) \cdot h$ is in $\mathbb{R}^q$. The symbol $y_{i, x_j}$ 
+stands for the partial derivative of the $i$th component of $y$, denoted by
+$y_i$ with respect to $x_j$. Thus,
+\[
+    y_{i, x_j} = \frac{\partial y_i}{\partial x_j}.
+\]
+An RNN corresponding to this equation is
 \begin{equation}\label{s4e7}
@@
 \section{Higher order ODE}\label{s5}
-A system of first order ODE is a set of equations of the form
+We can interpret the vector form of equation \eqref{s3e1} as a system of 
+first order ODE
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-It can be written in a vector form as
+Its equivalent vector form is
 \begin{equation}\label{s5e1}
-\frac{dy}{dx} = f,
+\frac{d}{dx}\begin{pmatrix}y_1 \\ \vdots \\ y_k\end{pmatrix} = 
+\begin{pmatrix}f_1 \\ \vdots \\ f_k\end{pmatrix}.
 \end{equation}
-where
-\[
-y = \begin{pmatrix}y_1 \\ \vdots \\ y_k\end{pmatrix},\;\text{and}\;
-f = \begin{pmatrix}f_1 \\ \vdots \\ f_k\end{pmatrix}
-\]
-A system of first order ODE is interesting because they are equivalent to
+A system of first order ODE is interesting because it is equivalent to
 a higher order differential equation. For example, one can write the $k$th
@@
 the German mathematicians Carl Runge and Martin Kutta in the late 
-nineteenth centuries\cite{lambert1991numerical}. We will describe the most
+nineteenth century\cite{lambert1991numerical}. We will describe the most
 commonly used among them, called the `RK4' method. If $y: \mathbb{R}
@@
 Let us study these equations carefully. The value of $y$ at $t_n + h$ 
-depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Of these,
+depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Each of
+these is the value of the derivative of $y$ at different points. Of these,
 $k_1$ depends on $y(t_n)$ and $t_n$, $k_2$ on $y(t_n)$, $t_n$ and $k_1$,
-$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. 
+$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. If
+one were to implement this algorithm, every iteration of it would consist
+of using the output of the previous iteration and a calculation of four
+constants, one after another, to get the value of the current iteration.
+This is indeed like a recurrent neural network with four hidden layers.
+
 Runge-Kutta methods perform much better then Euler methods when $f$ is 
@@
 layers, each computing $k_1$ to $k_4$ and the output later combining
-all of them to get the final result. This is a $4$-stage, $1$st order
-RNN.
+all of them to get the final result. We call a recurrent neural network
+implementing an algorithm to solve an ordinary differential equation an
+ODERNN.
 
-We now define a general RNN\cite{niu2019recurrent}.
+We now define a general ODERNN\cite{niu2019recurrent}.
 \begin{defn}\label{s7d1}
-An ODE-RNN of $n$th order in non-linearity and $t$th order in gradient,
+An ODE-RNN of $n$-th order in non-linearity and $t$-th order in gradient,
 called $(n,t)$-ODERNN, where $n, t \ge 1$ are integers is described by

### 2020-07-06 p3/bspline_filter.py
Clean up

@@ -1,0 +1,61 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+def mu(x):
+    y = np.ones(len(x))
+    # Select indices of x such that x[i] < 0.
+    si = np.where(x < 0)[0]
+    y[si] = 0
+
+    return y
+
+lo = -2.5
+hi = +2.5
+n_samples = 101
+
+x = np.linspace(lo, hi, n_samples)
+
+def beta_0(x):
+    b0 = mu(x + 1/2)
+    b1 = mu(x - 1/2)
+
+    return b0 - b1
+
+def convolve_and_normalize(x, y):
+    z = np.convolve(x, y)
+    z = z/np.max(z)
+
+    return z
+
+def add_to_plot(y, m, n):
+    l = len(y)
+    x = np.linspace(0, l, l) - l//2
+
+    left = l//4
+    right = l - left
+
+    plt.plot(x[left:right], y[left:right], label = rf'$\beta_{m}^{n}$'.format(m, n))
+
+
+beta_0_1 = beta_0(x)
+beta_1_1 = convolve_and_normalize(beta_0_1, beta_0_1)
+beta_2_1 = convolve_and_normalize(beta_1_1, beta_0_1)
+beta_3_1 = convolve_and_normalize(beta_2_1, beta_0_1)
+
+def b_m_0(m):
+    x = np.zeros(n_samples)
+    left = n_samples//2 - m//2
+    right = n_samples//2 + m//2
+
+    if m % 2 == 1:
+        right = right + 1
+
+    for i in range(left, right):
+        x[i] = 1
+
+    return x
+
+b_2_0 = b_m_0(2)
+b_3_2 = np.convolve(np.convolve(np.convolve(b_2_0, b_2_0), b_2_0), beta_3_1)
+b_3_2 = b_3_2/8
+

### 2020-07-06 p3/downloaddata.py

@@ -1,0 +1,270 @@
+import hashlib
+import sys
+import os
+import json
+
+import errno
+import warnings
+
+# http://stackoverflow.com/questions/2028517/python-urllib2-progress-hook
+
+def url_download_report(bytes_so_far, url_download_size, total_size):
+    percent = float(bytes_so_far) / total_size
+    percent = round(percent * 100, 2)
+    if bytes_so_far > url_download_size:
+        # Note that the carriage return is at the begining of the
+        # string and not the end. This accomodates usage in 
+        # IPython usage notebooks. Otherwise the string is not
+        # displayed in the output.
+        sys.stdout.write("\rDownloaded %d of %d bytes (%0.2f%%)" %
+                         (bytes_so_far, total_size, percent))
+        sys.stdout.flush()
+    if bytes_so_far >= total_size:
+        sys.stdout.write("\rDownloaded %d of %d bytes (%0.2f%%)\n" %
+                         (bytes_so_far, total_size, percent))
+        sys.stdout.flush()
+        
+ 
+def url_download_read(url, outputfile, url_download_size=8192 * 2, report_hook=None):
+    # Use the urllib2 to download the data. The Requests package, highly
+    # recommended for this task, doesn't support the file scheme so we opted
+    # for urllib2 which does.  
+
+    try:
+        # Python 3
+        from urllib.request import urlopen, URLError, HTTPError
+    except ImportError:
+        from urllib2 import urlopen, URLError, HTTPError
+    from xml.dom import minidom
+    
+    # Open the url
+    try:
+        url_response = urlopen(url)
+    except HTTPError as e:
+        return "HTTP Error: {0} {1}\n".format(e.code, url)
+    except URLError as e:
+        return "URL Error: {0} {1}\n".format(e.reason, url)
+    # MIDAS is a service and therefor will not generate the expected URLError
+    # when given a nonexistent url. It does return an error message in xml.
+    # When the response is xml then we have an error, we read the whole message
+    # and return the 'msg' attribute associated with the 'err' tag.
+    # The URLError above is not superfluous as it will occur when the url 
+    # refers to a non existent file ('file://non_existent_file_name') or url
+    # which is not a service ('http://non_existent_address').    
+    try:
+        # Python 3
+        content_type = url_response.info().get("Content-Type")
+    except AttributeError:
+        content_type = url_response.info().getheader("Content-Type")
+    # MIDAS error message in json format
[…]

### 2020-07-06 p3/z_transform.py

@@ -1,0 +1,18 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from mpl_toolkits import mplot3d
+
+x1 = np.linspace(-5, 5, 11)
+y1 = x1
+
+R = np.concatenate((np.linspace(-3, -1, 15), np.linspace(1, 3, 15)))
+x = np.outer(R, R)
+y = np.outer(R, R)
+z_im = -np.divide(y, (np.multiply(x - 1, x - 1) - np.multiply(y, y)))
+z_re = np.divide(np.multiply(x, x-1) + np.multiply(y, y), np.multiply(x-1, x-1) - np.multiply(y,y))
+
+fig = plt.figure()
+ax = plt.axes(projection = '3d')
+ax.plot_surface(x, y, z_re, cmap = 'viridis', edgecolor = 'none')
+plt.show()
+

### 2020-07-06 p4/generate-images.py

@@ -1,0 +1,40 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Jun 23 23:30:07 2020
+
+@author: ameyjoshi
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+from PIL import Image
+
+f = np.zeros(36)
+f[14] = f[15] = f[20] = f[21] =  f[35] = 128
+ref = Image.new('1', (6, 6))
+ref.putdata(f)
+
+tst = ref.transform(ref.size, Image.AFFINE, (1, 0, 1, 0, 1, 1))
+m = np.array(tst.getdata())
+
+plt.figure()
+plt.suptitle('Original and transformed images')
+p1 = plt.subplot(121)
+p1.set_title('Original')
+plt.imshow(ref)
+
+p2 = plt.subplot(122)
+p2.set_title('Transformed')
+plt.imshow(tst)
+
+plt.show()
+
+def mse(ref, tst):
+    r = np.array(ref.getdata())
+    t = np.array(tst.getdata())
+    
+    return np.pow(r - t, 2)
+
+MAX_ITER = 200
+mu_0 = np.ones(6) * 0.1
+

### 2020-07-06 p4/grad_desc.py

@@ -1,0 +1,28 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+def gradz_dot_dh(r, dh):
+    gradz = [2*r[0] + r[1], r[0] + 2*r[1]]
+
+    return np.dot(gradz, dh)
+
+def euclidean_dist(r1, r2):
+    return np.sqrt((r1[0] - r2[0])**2 + (r1[1] - r2[1])**2)
+
+threshold = 0.000001
+max_iter = 10000
+step = [0.01, 0.01]
+
+prev = [0.1, 0.1]
+n = 0
+while n < max_iter:
+    n = n + 1
+    curr = prev - gradz_dot_dh(prev, step)
+    if euclidean_dist(prev, curr) < threshold:
+        print('Found extremum.')
+        break
+    else:
+        prev = curr
+
+print(n)
+print(curr)

### 2020-07-06 p4/lm.py

@@ -1,0 +1,35 @@
+import numpy as np
+
+def gradz_dot_dh(r, dh):
+    gradz = [4 * r[0]**3 + r[1], r[0] + 4 * r[1]**3]
+
+    return np.array([gradz[0] * dh[0], gradz[1] * dh[1]])
+    
+def hessian(r, lmbd):
+    return (1 - lmbd) * 12 * r[0]**2 * r[1]**2 - 1
+
+def euclidean_dist(r1, r2):
+    return np.sqrt((r1[0] - r2[0])**2 + (r1[1] - r2[1])**2)
+
+threshold = 0.00001
+max_iter = 10000
+step = [0.01, 0.01]
+
+prev = [2, 1]
+n = 0
+lmbd = 2
+
+while n < max_iter:
+    n = n + 1
+    hp = hessian(prev, lmbd)
+    delta = gradz_dot_dh(prev, step)
+    curr = prev + delta/hp
+    if euclidean_dist(prev, curr) < threshold:
+        print('Found an extremum.')
+        break
+    else:
+        prev = curr
+
+print(n)
+print(curr)
+

### 2020-07-06 p5/stability_rnn.tex
Second revision.

@@
 notation for vectors and scalars. Instead, we define every symbol. Whenver
-a symbol stands for a vector, it denotes a column vector.
+a symbol stands for a vector, it denotes a column vector. $\mathbb{R}$ is 
+the set of real numbers. $\mathbb{Z}$ is the set of integers. If $z$ is
+a complex number then $\Re(z)$ denotes its real part and $\Im(z)$ its
+imaginary part.
 
@@
 \item $b_1,\ldots, b_{n+1} \in \mathbb{R}^p$ are bias vectors; 
-\item $\alpha_{q, k}, \beta_k$, for $q = 2, \ldots, n$ and $k = 1, \ldots,
-n$, $\gamma_2, \ldots, \gamma_{n}, \kappa_2, \ldots, \kappa_{n}$ are 
-$p \times p$ matrices;
+\item $\alpha_{q, k}$, for $q = 2, \ldots, n$ and $k = 1, \ldots, n - 1$
+are $p \times p$ matrices. So are $\gamma_2, \ldots, \gamma_{n}, \kappa_2, 
+\ldots, \kappa_{n}$;
 \item $\gamma_{n+1}$ is an $s \times s$ matrix and $\kappa_{n+1}$ is a 
@@
 \begin{rem}
-$K_1, \ldots, K_q$ are all $p$-dimensional vectors. That is, they 
+There is probably a typo in equation (3) of section 2.1 Niu and other's
+paper\cite{niu2019recurrent}. I believe that the range of the sum should
+be $1$ to $q - 1$ as given here instead of $1$ to $n$ as mentioned in the
+paper.
+\end{rem}
+\begin{rem}
+$K_1, \ldots, K_n$ are all $p$-dimensional vectors. That is, they 
 are all members of $\mathbb{R}^p$.
@@
 \cite{gereon2018hands}. In the context of ODEs this is called an 
-instability. Because of a correspondence between and RNN and an ODE, we 
+instability. Because of a correspondence between an RNN and an ODE, we 
 can study the problem of exploding gradients in the former as the 
@@
 $k$-step method to get its approximate numerical solution is defined by
-the formula
+the formula\cite{dahlquist1963special}
 \begin{equation}\label{s9e1}
@@
 \end{defn}
-A contractive solutions get closer to each other as one gets away from
+Contractive solutions get closer to each other as one gets away from
 the initial point.
@@
 
-The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because there
-is only a single hidden layer. The Burrage-Butcher tensor is thus $-\beta_1
-\beta_1^T$, which is \emph{not} positive semidefinite unless $\beta_1 = 0$.
-Thus the GRU is not a stable algorithm, an observation also corroborated by
-elsewhere\cite{kanai2017preventing}.
+The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because the
+hidden layers are not connected to each other. The Burrage-Butcher tensor 
+is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semidefinite 
+unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
+observation also corroborated by elsewhere\cite{kanai2017preventing}.
 \bibliographystyle{plain}

### 2020-07-11 p5/stability_rnn.tex
Third revision.
Fixed a typo.
A better filler.
Updated with LTSM architecture

@@
     using Ordinary Differential Equations}
-\author{Dan Will}
+\author{Author's name}
 \maketitle
@@
 to solve ordinary differential equations, a topic that has been studied
-intensively for several decades. We use this analogy to derive criteria for
-the stability of recurrent neural networks.
+intensively for several decades. We use this analogy to bank upon the 
+vast research about ordinary differential equations to derive criteria 
+for the stability of recurrent neural networks.
 
@@
 \end{equation}
-But the value of the derivative of $y$ with respect to $x$ is given by
+The square bracket around the derivative and the subscript $x_0$ to the
+right bracket indicates that the derivative is to be evaluated at $x = x_0$.
+The value of the derivative of $y$ with respect to $x$ is given by
 the differential equation itself. Thus,
@@
 \end{eqnarray}
-then we can spot a striking resemblance between Euler's algorithm and an
+then we can spot a resemblance between Euler's algorithm and an
 RNN. The idea that an RNN corresponds to an ODE and therefore the stability
@@
 where $q \in \mathbb{C}$ with $\Re(q) < 0$ and a fixed $h$ in equation
-\eqref{s9d1}.
+\eqref{s9d1}. Here $n$ is the number of steps in the numerical algorithm.
 \end{defn}
@@
 observation also corroborated by elsewhere\cite{kanai2017preventing}.
+
+\section{GRU with an Attention Layer}\label{s10}
+We will now analyze GRU with attention layer. We assume that the 
+architecture of the network is same as shown in figure 4 of the paper
+\cite{ran2019lstm} by Xiangdong and others, except that the LTSM is 
+replaced with a GRU. There are $n$ such GRU's, each with an output
+$h^i_l \in \mathbb{R}^s$. The combined output of all of them is the $s
+\times n$ matrix $Y_l$ defined by
+\begin{equation}\label{s10e1}
+H_l = [h^1_l, \ldots, h^n_l].
+\end{equation}
+The matrix $H_l$ is scaled with an $n \times 1$  weight matrix $W_H$. 
+Similarly, if $A$ is an $s \times n$ embedding matrix and $W_v$ is the
+corresponding weight matrix, then
+\begin{equation}\label{s10e2}
+M_l = \tanh\begin{pmatrix} H_lW_H \\ AW_v \end{pmatrix}
+\end{equation}
+is a member of $\mathbb{R}^{2s}$. If $W= [w_1, \ldots, w_n]$ is a $(2s) 
+\times n$ weight matrix then
+\begin{equation}\label{s10e3}
+a_i = \frac{\exp(w_i^T M_l)}{\sum_{j=1}^n \exp(w_j^T M_l)}
+\end{equation}
+are the $n$ components of the vector $a = (a_1, \ldots, a_n)
+\in \mathbb{R}^n$. Finally, the product
+\begin{equation}\label{s10e4}
+Y_l = H_l a
+\end{equation}
+is a member of $\mathbb{R}^s$. Note that the output of the individual
+GRUs is also an $s$-dimensional vector.
+
+Equations \eqref{s10e1} to \eqref{s10e4} where $Y_l$ is gived by 
+equations \eqref{s8e1} to \eqref{s8e5} defined GRU with an attention layer.
+We will analyse the stability of this architecture in this section.
+
+Equation \eqref{s10e4} tells that $Y_l$ depends $H_l$ but not the $H_{l-1}$.
+Thus, it is not a recurrence relation. $Y_l$ is not computed from the
+`present' $H_l$, not a `previous' one. Therefore, this attention layer 
+does not play a role in determining the stability of the network in the
+architecture shown in figure 4 of the paper \cite{ran2019lstm} by 
+Xiangdong and others.
+
 \bibliographystyle{plain}

### 2020-07-12 p5/stability_rnn.tex
First draft of attention layer.

@@
 \usepackage{amsmath, amssymb, amsfonts, amsthm}
+\DeclareMathOperator*{\argmax}{argmax}
+\DeclareMathOperator*{\smax}{softmax}
 \numberwithin{equation}{section}
@@
 definition \ref{s7d1} is
-\begin{equation}\label{x}
+\begin{equation}\label{s7e4}
 Q_{i, j} = \beta_i\alpha_{i, j} + \beta_j\alpha_{j, i} - \beta_i\beta_j^T.
@@
 \section{GRU with an Attention Layer}\label{s10}
-We will now analyze GRU with attention layer. We assume that the 
-architecture of the network is same as shown in figure 4 of the paper
-\cite{ran2019lstm} by Xiangdong and others, except that the LTSM is 
-replaced with a GRU. There are $n$ such GRU's, each with an output
-$h^i_l \in \mathbb{R}^s$. The combined output of all of them is the $s
+We will now analyze GRU with attention layer used in three separate
+architectures. 
+\subsection{Architecture proposed Ran and others} 
+We assume that the architecture of the network is same as shown in figure 4 
+of the paper \cite{ran2019lstm} by Ran and others, except that the 
+Long Short-Term Model (LTSM) is replaced with a GRU. There are $n$ such 
+GRU's, each with an output $h^i_l \in \mathbb{R}^s$. The combined output 
+of all of them is the $s
 \times n$ matrix $Y_l$ defined by
@@
 architecture shown in figure 4 of the paper \cite{ran2019lstm} by 
-Xiangdong and others.
+Ran and others.
+
+\subsection{Architecture proposed by Zang and others}
+We now consider the model proposed by Zang and others\cite{zhang2017gru}.
+The model is described by equations (11) to (17) in their paper. We
+reproduce them for an easy reference.
+\begin{eqnarray}
+z^\prime_t &=& \sigma(W_{yz}Ey_{t-1}+U_{sz}s_{t-1}+C_{cz}c_t)\label{s10e5}\\
+r^\prime_t &=& \sigma(W_{yr}Ey_{t-1}+U_{sr}s_{t-1}+C_{cr}c_t)\label{s10e6}\\
+\tilde{s}_t &=& \tanh\left(W_{ys}Ey_{t-1} + U_{rs}(r^\prime_t \circ
+	    s_{t-1} + C_{cs}c_t\right) \label{s10e7} \\
+s_t &=& (1 - z^\prime_t)\circ s_{t-1} + z^\prime_t \circ \tilde{s}_t
+    \label{s10e8} \\
+e_{ti} &=& \nu^T_{\text{att}}\tanh(W_{\text{att}}s_{t-1} +
+    U_\text{att}a_i \label{s10e9} \\
+\alpha_{ti} &=& 
+    \frac{\exp(e_{ti})}{\sum_{k=1}^L\exp(e_{tk})}\label{s10e10}\\
+c_t &=& \sum_{i=1}^L \alpha_{ti}a_i. \label{s10e11}
+\end{eqnarray}
+In these equations the lower case symbols are either vectors or scalars;
+the upper case symbols are matrices. In particular:
+\begin{itemize}
+\item $y_i \in \mathbb{R}^K$;
+\item $a_i \in \mathbb{R}^D$;
+\item $s_i, \tilde{s}_i, z_i, r_t \in \mathbb{R}^n$;
+\item $W_{yz}, W_{yr}, W_{ys}, U_{sz}, U_{sr}, U_{rs}, C_{cz}, C_{cr}, 
+C_{cs}$ are weight matrices of appropriate dimensions. $E$ is an embedding
+matrix;
+\item If $n^\prime$ is the attention dimension then $\nu_{\text{att}}^{n
+^\prime}$, $W_\text{att}$ is an $n \times n^\prime$ matrix and $U_a$ is an 
+$n^\prime \times D$ matrix.
+\end{itemize}
+The attention layer determines the vector $c_t$, which is analogous to
+the bias vectors $b_q$, $q = 1, \ldots, n + 1$ in equations \eqref{s7e1}
+to \eqref{s7e3} in definition \ref{s7d1}. The bias vectors \emph{do not}
+contribute to the Burrage-Butcher tensor defined in \eqref{s7e4}. Therefore,
+we conclude that the attention layer has no influence on the stability of
+the underlying network.
+
+\subsection{Architecture proposed by Liu and others}
+In this section we consider the neural network used in table-to-text
+generation by Liu and others \cite{liu2018table}. The goal of their
+network is to generate a sequence
+\begin{equation}\label{s10e12}
+w^\ast_{1:p} = \argmax_{w_{1:p}}\prod_{t=1}^p\mathrm{P}(w_t | w_{0:t-1}, 
+	R_{1:n}),
+\end{equation}
+where $w_i$ are generated tokens and $R_n$ are field value records in a
+table. The probability on the right hand side is defined by
+\begin{equation}\label{s10e13}
+\mathrm{P}(w_t| w_{0:t-1}, R_{1:n}) = \smax(W_s \circ g_t),
+\end{equation}
+where $W_s$ is a weight matrix and
+\begin{equation}\label{s10e14}
+g_t = \tanh(W_t[s_t, a_t])
+\end{equation}
+$s_t$ being the output of an LTSM neural network and $a_t$ being the 
+attention vector. It is built as
+\begin{equation}\label{s10e15}
+a_t = \sum_{i=1}^L \alpha_{t_i}h_i,
+\end{equation}
+where
+\begin{equation}\label{s10e16}
+\alpha_{t_i} = \frac{\exp(g(s_t, h_i))}{\sum_{j=1}^N\exp(g(s_t, h_j))},
+\end{equation}
+and
+\begin{equation}\label{s10e17}
+g(s_t, h_i) = \tanh(W_p h_i) \circ \tanh(W_q s_t).
+\end{equation}
+Here $s_t$ is the output of the LTSM at time $t - 1$ and 
+\begin{equation}\label{s10e18}
+h_i = o_t \circ \tanh(c_t).
+\end{equation}
+The variables in this equation are boolean vectors of size $n$. They 
+are defined as
+\begin{equation}\label{s10e19}
+\begin{pmatrix}i_t \\ f_t \\ o_t \\ \hat{c}_t \end{pmatrix} = 
+\begin{pmatrix}\sigma \\ \sigma \\ \sigma \\ \tanh\end{pmatrix}
+W_{4n, 2n}^c\begin{pmatrix}d_t \\ h_{t-1} \end{pmatrix}
+\end{equation}
+and
+\begin{equation}\label{s10e20}
+c_t = f_t \circ c_{t-1} + i_t \circ \hat{c}_t.
+\end{equation}
+Analogous to the case of the architecture proposed by \cite{ran2019lstm}
+the attention layer $a_t$ is used solely that the final stage. It is not
+fed back to the input and therefore it has no influence on the stability
+of the LTSM.
+
+\subsection{Remarks on stability}
+Recall theorem \ref{s9t1} that the stability of the Runge-Kutta method 
+depends on the whether or not the matrices $\beta_n$ \emph{and} the 
+Burrage-Butcher tensor are positive semi-definite. An attention layer will
+influence the stability of an algorithm if it influences either $\beta_i$
+of $\alpha_{i,j}$ which together form the matrices that determine the
+stability. The architectures of Ran et al. and Liu et al. use the 
+attention layer only at the last stage. They do not feed the attention
+later at time $t$ to get the output at time $t + 1$. On the other hand, 
+the architecture of Zang et al. uses the attention layer in bias vectors
+which do not determine the matrices that decide the stability of the
+algorithm.
+
+The stability of Runge-Kutta methods depends on their ability to generate
+contractive solutions in an interval. If a method is not stable it does not
+mean that it cannot generate \emph{any} contractive pair of solutions. All
+it means is that \emph{not all} pairs are contractive in the sense of 
+definition \ref{s9d2}. Existence of one, or a few, solutions to the 
+algorithm is not a contradiction to it being unstable. Instability means
+that \emph{not all} solutions are stable, and stability means that
+\emph{all} solutions are stable. An existence of a few coverging solutions
+does not tell us anything about the stability of the algorithm. However, 
+an existence of a single unstable solution renders an algorithm unstable.
 

### 2020-07-13 p5/stability_rnn.tex
Final.

@@
 GRU's, each with an output $h^i_l \in \mathbb{R}^s$. The combined output 
-of all of them is the $s
-\times n$ matrix $Y_l$ defined by
+of all of them is the $s \times n$ matrix $H_l$ defined by
 \begin{equation}\label{s10e1}
@@
 
-Equations \eqref{s10e1} to \eqref{s10e4} where $Y_l$ is gived by 
-equations \eqref{s8e1} to \eqref{s8e5} defined GRU with an attention layer.
-We will analyse the stability of this architecture in this section.
+Equations \eqref{s10e1} to \eqref{s10e4}, where the columns of $H_l$ can be
+constructed from equations \eqref{s8e1} to \eqref{s8e5}, define GRU with 
+an attention layer.  (Note that equations \eqref{s8e1} to \eqref{s8e5} use 
+$Y_t$. To use them here replace $Y$ with $H$. We want to reserve $Y$ for 
+the final output and we want to treat the intermediate results as hidden 
+layers, denoted by $H$.)
 
-Equation \eqref{s10e4} tells that $Y_l$ depends $H_l$ but not the $H_{l-1}$.
-Thus, it is not a recurrence relation. $Y_l$ is not computed from the
-`present' $H_l$, not a `previous' one. Therefore, this attention layer 
-does not play a role in determining the stability of the network in the
-architecture shown in figure 4 of the paper \cite{ran2019lstm} by 
-Ran and others.
+In order to analyse the stability of this architecture note that in equation
+\eqref{s10e4} $Y_l$ depends on $H_l$, but not $H_{l-1}$, and $a$. Further,
+$a$ is computed from the variables pertaining to step $l$ alone.
+Thus, it is not a recurrence relation. $Y_l$ is computed from the
+`present' $H_l$ and the `present' attention layer $a$, not the `previous' 
+ones. Therefore, this attention layer does not play a role in determining 
+the stability of the network in the architecture shown in figure 4 of the 
+paper \cite{ran2019lstm} by Ran and others. In mathematical terms, the
+attention layer does not get into any of the terms forming the 
+Burrage-Butcher tensor and therefore it does not influence the network's 
+stability.
 
@@
 \tilde{s}_t &=& \tanh\left(W_{ys}Ey_{t-1} + U_{rs}(r^\prime_t \circ
-	    s_{t-1} + C_{cs}c_t\right) \label{s10e7} \\
+	    s_{t-1} + C_{cs}c_t)\right) \label{s10e7} \\
 s_t &=& (1 - z^\prime_t)\circ s_{t-1} + z^\prime_t \circ \tilde{s}_t
@@
 e_{ti} &=& \nu^T_{\text{att}}\tanh(W_{\text{att}}s_{t-1} +
-    U_\text{att}a_i \label{s10e9} \\
+    U_\text{att}a_i)\label{s10e9} \\
 \alpha_{ti} &=& 
@@
 \end{equation}
-where $w_i$ are generated tokens and $R_n$ are field value records in a
+where $w_i$ are the generated tokens and $R_n$ are field-value records in a
 table. The probability on the right hand side is defined by
@@
 \end{equation}
-Here $s_t$ is the output of the LTSM at time $t - 1$ and 
+Here $s_t$ is the output of the LTSM at time $t$ and 
 \begin{equation}\label{s10e18}
@@
 
+The paper also defines a dual attention layer $a_t^\prime$. However it
+does not give a mathematical expression telling how it is used to compute
+the final output. Nevertheless, we can refer to figure 3 of the paper
+\cite{liu2018table} to confirm that it is not fed back to the network 
+and conclude that it will not affect its stability.
+
 \subsection{Remarks on stability}
@@
 Burrage-Butcher tensor are positive semi-definite. An attention layer will
-influence the stability of an algorithm if it influences either $\beta_i$
-of $\alpha_{i,j}$ which together form the matrices that determine the
+influence the stability of an algorithm if it determines either $\beta_i$
+or $\alpha_{i,j}$, which together form the matrices that decide the
 stability. The architectures of Ran et al. and Liu et al. use the 
@@
 
-The stability of Runge-Kutta methods depends on their ability to generate
+The stability of Runge-Kutta methods refers to their ability to generate
 contractive solutions in an interval. If a method is not stable it does not
-mean that it cannot generate \emph{any} contractive pair of solutions. All
+imply that it cannot generate \emph{any} contractive pair of solutions. All
 it means is that \emph{not all} pairs are contractive in the sense of 
-definition \ref{s9d2}. Existence of one, or a few, solutions to the 
-algorithm is not a contradiction to it being unstable. Instability means
-that \emph{not all} solutions are stable, and stability means that
-\emph{all} solutions are stable. An existence of a few coverging solutions
-does not tell us anything about the stability of the algorithm. However, 
-an existence of a single unstable solution renders an algorithm unstable.
+definition \ref{s9d2}. Existence of one, or a few, convergent solutions to 
+the algorithm is not a contradiction to it being unstable. An instability 
+means that \emph{not all} solutions are stable. On the other hand stability 
+means that \emph{all} solutions are stable. An existence of a few coverging 
+solutions does not guarantee the stability of the algorithm. However, an 
+existence of a single unstable solution renders an algorithm unstable.
+
+Translating this argument to the world of neural networks, adding an 
+attention layer might improve the performance of a stable solution but it
+will not make an unstable solution stable.
 

### 2020-08-22 la/c1.tex
Initial commit.

\chapter{Spaces}\label{c1}
\section{Fields}\label{c1s1}
\begin{defn}\label{c1s1d1}
A set $F$ with binary operations $+$ and $\cdot$ defined on it is a
field is
\begin{enumerate}
\item (A) Properties of the operation `$+$':
\begin{enumerate}
\item For every $\alpha, \beta \in F$, $\alpha + \beta \in F$,
\item For every $\alpha, \beta \in F$, $\alpha + \beta = \beta +
\alpha$,
\item For all $\alpha, \beta, \gamma \in F$, $\alpha + (\beta + 
\gamma) = (\alpha + \beta) + \gamma$,
\item There exists a member $0$ such that $\alpha + 0 = \alpha$ for
all $\alpha \in F$,
\item For every $\alpha \in F$, there exists a member $-\alpha$
such that $\alpha + (-\alpha) = 0$.
\end{enumerate}
\item (B) Properties of the operation `$\cdot$':
\begin{enumerate}
\item For every $\alpha, \beta \in F$, $\alpha \cdot \beta \in F$,
\item For every $\alpha, \beta \in F$, $\alpha \cdot \beta = \beta 
\cdot \alpha$,
\item For all $\alpha, \beta, \gamma \in F$, $\alpha \cdot (\beta 
\cdot \gamma) = (\alpha \cdot \beta) \cdot \gamma$,
\item There exists a member $0$ such that $\alpha \cdot 1 = \alpha$ for
all $\alpha \in F$,
\item For every $\alpha \ne 0 \in F$, there exists a member 
$\alpha^{-1}$ such that $\alpha \cdot \alpha^{-1} = 1$.
\end{enumerate}
\item The operation $\cdot$ is distributive over $+$. That is $\alpha
\cdot (\beta + \gamma) = \alpha \cdot \beta + \alpha \cdot \gamma$
for all $\alpha, \beta, \gamma \in F$.
\end{enumerate}
\end{defn}

\begin{rem}
Members of a field are called scalars.
\end{rem}

\subsection{Exercises}
\begin{enumerate}
\item We will prove the following assertions.
\begin{enumerate}
\item $0 + \alpha$ by commutativity of `+' is the same as $\alpha + 0
= \alpha$.
\item $\alpha + \beta = \alpha + \gamma \Rightarrow -\alpha + (\alpha
+ \beta) = -\alpha + (\alpha + \gamma) \Rightarrow (-\alpha + \alpha)
+ \beta = (-\alpha + \alpha) + \gamma$. Now, by commutativity of `+',
$\alpha + (-\alpha) = 0 \Rightarrow -\alpha + \alpha = 0$. Therefore,
$0 + \beta = 0 + \gamma \Rightarrow \beta = \gamma$.
\item $\alpha + (\beta - \alpha) = \alpha + (-\alpha + \beta) = 
\alpha + (-\alpha) + \beta = 0 + \beta = \beta$.
\item Since $0$ is an additive identity, $0 + 0 = 0 \Rightarrow
\alpha\cdot(0 + 0) = \alpha\cdot 0 \Rightarrow \alpha\cdot 0 + 
\alpha\cdot 0 = \alpha\cdot 0$. Since $\alpha\cdot 0$ is a member of
$F$, it has an additive inverse $-\alpha\cdot 0$. Thus,
$-\alpha\cdot 0 + (\alpha\cdot 0 + \alpha\cdot 0) = 
-\alpha\cdot 0 + \alpha\cdot 0 \Rightarrow
(-\alpha\cdot 0 + \alpha\cdot 0) + \alpha\cdot 0 = 0 \Rightarrow
0 + \alpha\cdot 0 = 0 \Rightarrow \alpha\cdot 0 = 0.$
The claim $0\cdot\alpha = 0$ follows from commutativity of $\cdot$.
\item We observe that $(-1)\alpha + \alpha = \alpha(-1) + \alpha(1)
= \alpha(-1 + 1) = \alpha 0 = 0$. Thus, $(-1)\alpha$ is the additive
inverse of $\alpha$. $(-1)\alpha = -\alpha$ follows from the
uniqueness of the additive inverse.
\item $(-\alpha)(-\beta) + (-\alpha)(\beta) = -\alpha(-\beta + \beta)
= -\alpha \cdot 0 = 0$. Thus, $(-\alpha)(-\beta)$ is the additive 
inverse of $(-\alpha)(\beta)$. We also observe that $\alpha\beta + 
(-\alpha)\beta = (\alpha - \alpha)\beta = 0\cdot\beta = 0$. Thus,
$\alpha\beta$ is also an additive inverse of $(-\alpha)(\beta)$. The
claim follows from the uniqueness of additive inverse.

This claim used the fact that $(\alpha + \beta)\gamma = \alpha\gamma
+ \beta\gamma$. It follows from commutativity of the two operations.

\item Let $\alpha\beta = 0$. If $\alpha \ne 0$ then $\alpha^{-1}$
exists. Multiply both sides of the equation by $\alpha^{-1}$ to get
$\beta = 0$. Similarly, $\beta \ne 0$ gives $\alpha = 0$. The equality
is also true when both $\alpha$ and $\beta$ are zero.
\end{enumerate}
\end{enumerate}

### 2020-08-30 la/c1.tex
WIP 31Aug2020

@@
 \end{enumerate}
+
+\item
+\begin{enumerate}
+\item The set of all integers is not a field because the inverse
+under addition does not exist.
+\item The set of all integers is a group. It is not a field because
+multiplicative inverse does not exist.
 \end{enumerate}
+
+\item Given $m \in \soi$ and $Z_m = \{0, 1, \ldots, m - 1\}$, the 
+field operations are defined as $\alpha \oplus \beta = (\alpha + 
+\beta)\mod n$ and $\alpha \odot \beta = (\alpha\cdot\beta)\mod n$.
+The set $(Z_m, \oplus)$ is an abelian group. However, $(Z_m, \odot)$
+has an inverse if and only if $m$ is prime.
+\end{enumerate}

### 2020-08-30 optim/or-tools/lp/diet.py

@@ -1,0 +1,68 @@
+#!/bin/python
+
+from ortools.linear_solver import pywraplp
+
+# 'Hello world' of linear programming. We are given a list of food items
+# with their calorie content and nutrition content. We are also given the
+# minimum nutrition needed per day. We are asked to design a diet that meets
+# all nutritional needs and supplies at least a certain number of calories.
+
+# There are three nutrients, vitamins A, B and C.
+Nutrition = [2000, 300, 430]
+# Every food has calorie and vitamin content data.
+FoodData = [['Trout', 600, 203, 92, 100],
+            ['CB Sandwich', 350, 90, 84, 230],
+            ['Burrito', 250, 270, 80, 512],
+            ['Hamburger', 500, 500, 90, 210]]
+# Minimum consumption
+MinimumCalories = 2500
+
+solver= pywraplp.Solver.CreateSolver('diet', 'GLOP')
+
+# Create decision variables.
+consumption = [] # How much food should one consume?
+objf = solver.Objective()
+for i in range(0, len(FoodData)):
+    dvar = solver.NumVar(0, solver.infinity(), FoodData[i][0])
+    objf.SetCoefficient(dvar, FoodData[i][1])
+    consumption.append(dvar)
+
+objf.SetMinimization()
+
+# Add constraints
+constraints = []
+# Meet nutrition needs.
+for i in range(0, len(Nutrition)):
+    cn = solver.Constraint(Nutrition[i], solver.infinity())
+    for j in range(0, len(FoodData)):
+        cn.SetCoefficient(consumption[j], FoodData[i][2 + i])
+
+    constraints.append(cn)
+
+# Eat all foods.
+for i in range(0, len(FoodData)):
+    cn = solver.Constraint(0, solver.infinity())
+    cn.SetCoefficient(consumption[i], 1)
+    constraints.append(cn)
+
+# Minimal calories must be eaten.
+for i in range(0, len(FoodData)):
+    cn = solver.Constraint(MinimumCalories, solver.infinity())
+    for j in range(0, len(FoodData)):
+        cn.SetCoefficient(consumption[j], FoodData[i][1])
+
+# Solve the problem
+status = solver.Solve()
+
+if status == solver.OPTIMAL:
+    calories_consumed = 0
+    for i in range(0, len(consumption)):
+       calories_consumed += consumption[i].solution_value() * FoodData[i][1]
+
+    print(f'Calories consumed = {calories_consumed}') 
[…]

### 2020-08-30 optim/or-tools/lp/toy-lp-1.py

@@ -1,0 +1,40 @@
+#!/bin/python
+
+from ortools.linear_solver import pywraplp
+
+solver = pywraplp.Solver.CreateSolver('a_simple_program', 'GLOP')
+
+x = solver.NumVar(0, solver.infinity(), 'x') # x >= 0
+y = solver.NumVar(0, solver.infinity(), 'y') # y >= 0
+
+# x + 2y <= 14
+cn0 = solver.Constraint(-solver.infinity(), 14)
+cn0.SetCoefficient(x, 1)
+cn0.SetCoefficient(y, 2)
+
+# 3x - y >= 0
+cn1 = solver.Constraint(0, solver.infinity())
+cn1.SetCoefficient(x, 3)
+cn1.SetCoefficient(y, -1)
+
+# x - y <= 2
+cn2 = solver.Constraint(-solver.infinity(), 2)
+cn2.SetCoefficient(x, 1)
+cn2.SetCoefficient(y, -1)
+
+# The objective function f(x, y) = 3x + 4y
+objf = solver.Objective()
+objf.SetCoefficient(x, 3)
+objf.SetCoefficient(y, 4)
+objf.SetMaximization()
+
+# Solve the problem
+solver.Solve()
+opt_value = 3 * x.solution_value() + 4 * y.solution_value()
+
+print('Solution:')
+print('x = ', round(x.solution_value(), 4))
+print('y = ', round(y.solution_value(), 4))
+print('f(x, y) = ', round(opt_value, 4))
+
+

### 2020-08-30 optim/or-tools/lp/toy-lp-2.py

@@ -1,0 +1,38 @@
+#!/bin/python
+
+from ortools.linear_solver import pywraplp
+
+solver = pywraplp.Solver.CreateSolver('Ex 1 from Mosek', 'GLOP')
+
+# The decision variables
+x = solver.NumVar(0, solver.infinity(), 'x')
+y = solver.NumVar(0, solver.infinity(), 'y')
+z = solver.NumVar(0, solver.infinity(), 'z')
+
+# The constraint x + y + z = 1
+cn1 = solver.Constraint(1, solver.infinity())
+cn1.SetCoefficient(x, 1)
+cn1.SetCoefficient(y, 1)
+cn1.SetCoefficient(z, 1)
+
+cn2 = solver.Constraint(-solver.infinity(), 1)
+cn2.SetCoefficient(x, 1)
+cn2.SetCoefficient(y, 1)
+cn2.SetCoefficient(z, 1)
+
+# The objective function x + 2y - z.
+objf = solver.Objective()
+objf.SetCoefficient(x, 1)
+objf.SetCoefficient(y, 2)
+objf.SetCoefficient(z, -1)
+objf.SetMinimization()
+
+# Solve the problem
+status = solver.Solve()
+if status == solver.OPTIMAL:
+    optimal_value = x.solution_value() + 2*y.solution_value() - z.solution_value()
+    print(f'Optimal value = {optimal_value}')
+    print(f'x = {x.solution_value()}, y = {y.solution_value()}, z = {z.solution_value()}')
+else:
+    print('The solver failed to find an optimal value.')
+

### 2020-08-30 optim/or-tools/qp/portfolio.py

@@ -1,0 +1,29 @@
+#!/bin/python
+
+from cvxopt import matrix
+from cvxopt.blas import dot
+from cvxopt.solvers import qp, options
+import numpy as np
+
+# Number of assets
+n = 4
+
+# Variance covariance matrix
+S = matrix([[0.04, 0.006, -0.004, 0],
+               [0.006, 0.01, 0, 0],
+               [-0.004, 0, 0.0025, 0],
+               [0, 0, 0, 0.1]])
+# Mean returns
+pbar = matrix([0.12, 0.1, 0.07, 0.03])
+G = matrix(-1.0 * np.eye(n))
+h = matrix(0.0, (n, 1))
+A = matrix(1.0, (1, n))
+b = matrix(1.0)
+
+mu = 1000
+
+solution = qp(mu * S, -pbar, G, h, A, b)
+print(solution)
+print(solution['x'])
+print(sum(solution['x']))
+

### 2020-08-30 optim/or-tools/qp/portfolio.py
Added a reference.

@@ -1,5 +1,8 @@
 #!/bin/python
 
+# Example from http://cvxopt.org/examples/book/portfolio.html.
+# Refer to 'Convex Optimization' by Boyd and Vandenberghe
+
 from cvxopt import matrix
 from cvxopt.blas import dot
 from cvxopt.solvers import qp, options

### 2020-08-31 optim/or-tools/lp/toy-lp-1.py
Significant simplification.

@@ -8,29 +8,20 @@ x = solver.NumVar(0, solver.infinity(), 'x') # x >= 0
 y = solver.NumVar(0, solver.infinity(), 'y') # y >= 0
 
 # x + 2y <= 14
-cn0 = solver.Constraint(-solver.infinity(), 14)
-cn0.SetCoefficient(x, 1)
-cn0.SetCoefficient(y, 2)
+solver.Add(x + 2*y <= 14)
 
 # 3x - y >= 0
-cn1 = solver.Constraint(0, solver.infinity())
-cn1.SetCoefficient(x, 3)
-cn1.SetCoefficient(y, -1)
+solver.Add(3*x - y >= 0)
 
 # x - y <= 2
-cn2 = solver.Constraint(-solver.infinity(), 2)
-cn2.SetCoefficient(x, 1)
-cn2.SetCoefficient(y, -1)
+solver.Add(x - y <= 2)
 
 # The objective function f(x, y) = 3x + 4y
-objf = solver.Objective()
-objf.SetCoefficient(x, 3)
-objf.SetCoefficient(y, 4)
-objf.SetMaximization()
+solver.Maximize(3*x + 4*y)
 
 # Solve the problem
 solver.Solve()
-opt_value = 3 * x.solution_value() + 4 * y.solution_value()
+opt_value = solver.Objective().Value()
 
 print('Solution:')
 print('x = ', round(x.solution_value(), 4))

### 2020-08-31 optim/or-tools/lp/toy-lp-2.py
Significant simplification.

@@ -10,27 +10,16 @@ y = solver.NumVar(0, solver.infinity(), 'y')
 z = solver.NumVar(0, solver.infinity(), 'z')
 
 # The constraint x + y + z = 1
-cn1 = solver.Constraint(1, solver.infinity())
-cn1.SetCoefficient(x, 1)
-cn1.SetCoefficient(y, 1)
-cn1.SetCoefficient(z, 1)
-
-cn2 = solver.Constraint(-solver.infinity(), 1)
-cn2.SetCoefficient(x, 1)
-cn2.SetCoefficient(y, 1)
-cn2.SetCoefficient(z, 1)
+solver.Add(x + y + z >= 1)
+solver.Add(x + y + z <= 1)
 
 # The objective function x + 2y - z.
-objf = solver.Objective()
-objf.SetCoefficient(x, 1)
-objf.SetCoefficient(y, 2)
-objf.SetCoefficient(z, -1)
-objf.SetMinimization()
+solver.Minimize(x + 2*y - z)
 
 # Solve the problem
 status = solver.Solve()
 if status == solver.OPTIMAL:
-    optimal_value = x.solution_value() + 2*y.solution_value() - z.solution_value()
+    optimal_value = solver.Objective().Value()
     print(f'Optimal value = {optimal_value}')
     print(f'x = {x.solution_value()}, y = {y.solution_value()}, z = {z.solution_value()}')
 else:

### 2020-08-31 optim/or-tools/lp/toy-lp-2.py
Used an equality constraint instead of two inequalities.

@@ -10,8 +10,7 @@ y = solver.NumVar(0, solver.infinity(), 'y')
 z = solver.NumVar(0, solver.infinity(), 'z')
 
 # The constraint x + y + z = 1
-solver.Add(x + y + z >= 1)
-solver.Add(x + y + z <= 1)
+solver.Add(x + y + z == 1)
 
 # The objective function x + 2y - z.
 solver.Minimize(x + 2*y - z)

### 2020-08-31 optim/or-tools/binp/0-1-knapsack.py
WIP 01Sep2020

@@ -1,0 +1,50 @@
+#!/bin/python
+
+# A 0-1 knapsack problem consists in choosing a subset of items with maximal
+# value subject the a constraint of weight.
+#
+# Suppose that there are N items i_1, ..., i_N, each with weight w_1, ..., w_N
+# and value v_1, ..., v_N. The problem consists in choosing the items in such
+# a way that a bag of capacity W can be filled such the value of the items is
+# maximal.
+#
+# It is called a 0-1 knapsack problem because each item can either be chosen
+# or not.
+
+from ortools.algorithms import pywrapknapsack_solver
+from ortools.algorithms.pywrapknapsack_solver import KnapsackSolver
+
+# The problem data.
+values = [
+    360, 83, 59, 130, 431, 67, 230, 52, 93, 125, 670, 892, 600, 38, 48, 147,
+    78, 256, 63, 17, 120, 164, 432, 35, 92, 110, 22, 42, 50, 323, 514, 28,
+    87, 73, 78, 15, 26, 78, 210, 36, 85, 189, 274, 43, 33, 10, 19, 389, 276,
+    312
+]
+weights = [[
+    7, 0, 30, 22, 80, 94, 11, 81, 70, 64, 59, 18, 0, 36, 3, 8, 15, 42, 9, 0,
+    42, 47, 52, 32, 26, 48, 55, 6, 29, 84, 2, 4, 18, 56, 7, 29, 93, 44, 71,
+    3, 86, 66, 31, 65, 0, 79, 20, 65, 52, 13
+]]
+capacity = [850]
+
+solver = KnapsackSolver(KnapsackSolver.KNAPSACK_MULTIDIMENSION_BRANCH_AND_BOUND_SOLVER, '0-1 knapsack')
+
+solver.Init(values, weights, capacity)
+solution = solver.Solve()
+
+selection = []
+item_wt = []
+total_weight = 0
+
+print(f'Total value = {solution}')
+for i in range(0, len(values)):
+    if solver.BestSolutionContains(i):
+        selection.append(i)
+        item_wt.append(weights[0][i])
+        total_weight += weights[0][i]
+
+print(f'Total weight = {total_weight}')
+print(f'Selection = {selection}')
+print(f'Item weights = {item_wt}')
+

### 2020-08-31 optim/or-tools/binp/multi-knapsack.py

@@ -1,0 +1,64 @@
+#!/bin/python
+
+# Knapsack problem is an example of an NP-hard problem. 
+# 
+# We are given a knapsack with a certain capacity in terms of the weight it can
+# carry. We are also given a set of items. Each item has a weight, a value and
+# a quantity. The quantity is the number of items of that kind.
+ 
+# The goal of the knapsack problem is to fill it with items that maximize its 
+# value. The problem described here is slightly different. It is called the 
+# multi-knapsack problem. Here an item is also associated with a certain number
+# of resources. Taking an item results in a consumption of these resources. We 
+# must maximize the value of the knapsack while still being controlled by the
+# availability of the resources.
+
+from ortools.linear_solver import pywraplp
+from ortools.linear_solver.pywraplp import Solver
+
+# The data.
+nResources = 7
+nItems = 12
+resourceAvailability = [18209, 7692, 1333, 924, 26638, 61188, 13360]
+itemValue = [96, 76, 56, 11, 86, 10, 66, 86, 83, 12, 9, 81]
+resourceUse = [
+      [19,   1,  10,  1,   1,  14, 152, 11,  1,   1, 1, 1],
+      [ 0,   4,  53,  0,   0,  80,   0,  4,  5,   0, 0, 0],
+      [ 4, 660,   3,  0,  30,   0,   3,  0,  4,  90, 0, 0],
+      [ 7,   0,  18,  6, 770, 330,   7,  0,  0,   6, 0, 0],
+      [ 0,  20,   0,  4,  52,   3,   0,  0,  0,   5, 4, 0],
+      [ 0,   0,  40, 70,   4,  63,   0,  0, 60,   0, 4, 0],
+      [ 0,  32,   0,  0,   0,   5,   0,  3,  0, 660, 0, 9]]
+
+solver = Solver.CreateSolver('multi-knapsack', 'CBC')
+
+# How many numbers of each items should be packed? Each one of these is a 
+# decision variable.
+take = [None] * nItems
+for i in range(0, nItems):
+    take[i] = solver.IntVar(0, solver.infinity(), f'take[{i}]')
+
+# Build the objective function.
+objf = solver.Objective()
+for i in range(0, nItems):
+    objf.SetCoefficient(take[i], itemValue[i])
+
+objf.SetMaximization()
+
+for i in range(0, nResources):
+    cn = solver.Constraint(0, resourceAvailability[i])
+    for j in range(0, nItems):
[…]

### 2020-08-31 optim/or-tools/binp/wagons.py

@@ -1,0 +1,75 @@
+#!/bin/python
+
+# Bin-packing problem involves using the minimal number of bins to carry a
+# certain number of items.
+#
+# Suppose that we are given items of a certain weight. They should be 
+# transported through wagons of a certain capacity. What is the least number
+# of wagons that we should hire to carry all goods?
+
+itemWts = [48, 30, 19, 36, 36, 27, 42, 42, 36, 24, 30]
+maxwt = 100
+
+# We ensure that every item can be fitted in a wagon. That is, none of them
+# has a weight more than the wagon's capacity.
+
+from ortools.linear_solver import pywraplp
+from ortools.linear_solver.pywraplp import Solver
+
+solver = Solver.CreateSolver('bin-packing', 'CBC')
+
+# The decision variables are of the form x_{ij}. They take a value 1 if
+# item i is put in wagon j, 0 otherwise.
+nItems = len(itemWts)
+nWagons = nItems
+
+x = {}
+for i in range(0, nItems):
+    for j in range(0, nWagons):
+        vname = f'x[{i}, {j}]'
+        x[(i, j)] = solver.IntVar(0, 1, vname)
+
+# Recall that we need at most as many wagons as we have items. w[i] is 1
+# is wagon i is used.
+w = [None] * nWagons
+for i in range(0, nWagons):
+    w[i] = solver.IntVar(0, 1, f'w[{i}]')
+
+# We now add the constraints.
+# Each item must be in exactly one bin.
+for i in range(0, nItems):
+    solver.Add(sum(x[(i, j)] for j in range(0, nWagons)) == 1)
+
+# We cannot load a wagon beyond its capacity.
+for j in range(0, nWagons):
+    cn = sum(x[(i, j)]*itemWts[i] for i in range(0, nItems)) <= w[j]*maxwt
+    solver.Add(cn)
+
+# The objective function
+solver.Minimize(solver.Sum([w[j] for j in range(0, nWagons)]))
+
+status = solver.Solve()
+
+if status == Solver.OPTIMAL:
+    wagons_used = 0
+    for j in range(0, nWagons):
+        if w[j].solution_value() == 1:
+            items = []
+            loading = 0
+            for i in range(0, nItems):
+                if x[(i, j)].solution_value() != 0:
+                    items.append(i)
+                    loading += itemWts[i]
+
+            if loading > 0:
+                wagons_used += 1
+                print(f'Wagon number {j}:')
+                print(f'    Items: {items}')
[…]

### 2020-08-31 optim/or-tools/lp/diet.py

@@ -35,7 +35,7 @@ constraints = []
 for i in range(0, len(Nutrition)):
     cn = solver.Constraint(Nutrition[i], solver.infinity())
     for j in range(0, len(FoodData)):
-        cn.SetCoefficient(consumption[j], FoodData[i][2 + i])
+        cn.SetCoefficient(consumption[j], FoodData[j][2 + i])
 
     constraints.append(cn)
 
@@ -46,10 +46,9 @@ for i in range(0, len(FoodData)):
     constraints.append(cn)
 
 # Minimal calories must be eaten.
+cn = solver.Constraint(MinimumCalories, solver.infinity())
 for i in range(0, len(FoodData)):
-    cn = solver.Constraint(MinimumCalories, solver.infinity())
-    for j in range(0, len(FoodData)):
-        cn.SetCoefficient(consumption[j], FoodData[i][1])
+    cn.SetCoefficient(consumption[i], FoodData[i][1])
 
 # Solve the problem
 status = solver.Solve()

### 2020-08-31 optim/or-tools/mip/basic-mip.py

@@ -1,0 +1,32 @@
+#!/bin/python
+
+from ortools.linear_solver import pywraplp
+
+# We solve the following problem.
+# Maximize x + 10y subjec to:
+# x + 7y <= 17.5; x <= 3.5; x, y >= 0
+# where x, y are integers.
+
+solver = pywraplp.Solver.CreateSolver('basic-mip', 'CBC')
+
+# Integer decision variables
+x = solver.IntVar(0, solver.infinity(), 'x')
+y = solver.IntVar(0, solver.infinity(), 'y')
+
+# Add constraints
+solver.Add(x + 7*y <= 17.5)
+solver.Add(x <= 3.5)
+
+# Set up the objective function
+solver.Maximize(x + 10*y)
+
+status = solver.Solve()
+
+if status == pywraplp.Solver.OPTIMAL:
+    print('Solution:')
+    print(f'Max value: {solver.Objective().Value()}.')
+    print(f'x = {x.solution_value()}')
+    print(f'y = {y.solution_value()}')
+else:
+    print('The solver did not find an optimal solution.')
+

### 2020-08-31 optim/or-tools/mip/diet-ip.py

@@ -1,0 +1,57 @@
+#!/bin/python
+
+from ortools.linear_solver import pywraplp
+from ortools.linear_solver.pywraplp import Solver
+
+# A person needs three nutrients. Let's assume they are vitamins A, B and C.
+Nutrition = [2000, 300, 430]
+# They can be supplied by four foods. The first number if the calories supplied
+# by a food type. The remaining three numbers are the nutrients.
+Foods = [['Trout', 600, 203, 92, 100],
+         ['CB Sandwich', 350, 90, 84, 230],
+         ['Burrito', 250, 270, 80, 512],
+         ['Hamburger', 500, 500, 90, 210]]
+# A person needs a certain minimum calories.
+MinCalories = 2500
+
+solver = Solver.CreateSolver('diet', 'CBC')
+# The decision variables. How much quantity of each food typu should a person
+# consume?
+consumption = [None] * len(Foods)
+for i in range(0, len(Foods)):
+  consumption[i] = solver.IntVar(1, solver.infinity(), Foods[i][0])
+
+# The objective function is to minimize the number of calories.
+objf = solver.Objective()
+for i in range(0, len(Foods)):
+  objf.SetCoefficient(consumption[i], Foods[i][1])
+
+objf.SetMinimization()
+
+# Add constraints. Unlike the previous examples where we had our constraints
+# given to us as inequalities, here we will have to build them from the given
+# data.
+
+# The constraint that the nutrient needs must be met.
+for i in range(0, len(Nutrition)):
+  cn = solver.Constraint(Nutrition[i], solver.infinity()) # This is the rhs.
+  for j in range(0, len(Foods)): # We build the lhs.
+    cn.SetCoefficient(consumption[j], Foods[j][i + 2]) 
+
+# The calorie needs should be met.
+cn = solver.Constraint(MinCalories, solver.infinity())
+for i in range(0, len(Foods)):
+  cn.SetCoefficient(consumption[i], Foods[i][1])
+
+status = solver.Solve()
+if status == Solver.OPTIMAL:
+  calories = solver.Objective().Value()
+  for i in range(0, len(Foods)):
+    amt = round(consumption[i].solution_value(), 4)
+    print(f'{Foods[i][0]}: {amt}.')
+
+  print(f'Calories: {calories}.')
+else:
+  print('The solver could not find an optimum.')
+
+del solver, consumption, Nutrition, Foods, objf, cn

### 2020-08-31 optim/or-tools/mip/mip-array.py

@@ -1,0 +1,62 @@
+#!/bin/python 
+
+from ortools.linear_solver import pywraplp
+from ortools.linear_solver.pywraplp import Solver
+
+def create_data_model():
+  """Stores the data for the problem."""
+  data = {}
+  data['constraint_coeffs'] = [
+      [5, 7, 9, 2, 1],
+      [18, 4, -9, 10, 12],
+      [4, 7, 3, 8, 5],
+      [5, 13, 16, 3, -7],
+  ]
+  data['bounds'] = [250, 285, 211, 315]
+  data['obj_coeffs'] = [7, 8, 2, 9, 6]
+  data['num_vars'] = 5
+  data['num_constraints'] = 4
+
+  return data
+
+data = create_data_model()
+
+solver = Solver.CreateSolver('mip', 'CBC')
+# Quite like in the case of the 'Diet problem' the decision variables are 
+# created as elements of a list.
+x = [None] * data['num_vars']
+for i in range(0, len(x)):
+  x[i] = solver.IntVar(0, solver.infinity(), f'x[{i}]')
+
+print(f'Number of variables = {solver.NumVariables()}')
+
+# Build the constraints.
+for i in range(0, data['num_constraints']):
+  cn = solver.Constraint(0, data['bounds'][i], f'cn[{i}]')
+  for j in range(0, data['num_vars']):
+    cn.SetCoefficient(x[j], data['constraint_coeffs'][i][j])
+print(f'Number of constraints = {solver.NumConstraints()}')
+
+# Build the objective function.
+objf = solver.Objective()
+for i in range(0, data['num_vars']):
+  objf.SetCoefficient(x[i], data['obj_coeffs'][i])
+
+objf.SetMaximization()
+
+status = solver.Solve()
+if status == Solver.OPTIMAL:
+  maxval = solver.Objective().Value()
+  for i in range(0, data['num_vars']):
+    xv = round(x[i].solution_value(), 4)
+    print(f'x[{i}]: {xv}', end=' ')
+
+  print()
+  print(f'maxval: {maxval}.')
+
+  print('Problem solved in %f milliseconds' % solver.wall_time())
+  print('Problem solved in %d iterations' % solver.iterations())
+  print('Problem solved in %d branch-and-bound nodes' % solver.nodes())
+else:
+  print('The solver could not find an optimum.')
+  print(f'status = {status}')

### 2020-09-01 optim/or-tools/scheduling/assignment.py
CP examples.

@@ -1,0 +1,80 @@
+#!/bin/python
+
+from ortools.sat.python import cp_model
+
+class SolutionPrinter(cp_model.CpSolverSolutionCallback):
+    def __init__(self, shifts, nDays, nShifts, nPeople, solutions):
+        cp_model.CpSolverSolutionCallback.__init__(self)
+        self._shifts = shifts
+        self._nDays = nDays
+        self._nShifts = nShifts
+        self._nPeople = nPeople
+        self._solutions = set(solutions)
+        self._solnCount = 0
+
+    def on_solution_callback(self):
+        if self._solnCount in self._solutions:
+            print(f'Solution {self._solnCount}')
+
+            for d in range(self._nDays):
+                print(f'Day {d}')
+                for p in range(self._nPeople):
+                    assigned = False
+                    for s in range(self._nShifts):
+                        if self.Value(self._shifts[(d,s,p)]):
+                            assigned = True
+                            print(f'    Person {p} works shift {s}')
+                    if not assigned:
+                        print(f'    Person {p} does not work.')
+            print()
+
+        self._solnCount += 1
+
+    def solnCount(self):
+        return self._SolnCount
+
+nPeople = 4
+nShifts = 3
+nDays   = 3
+
+allPeople = range(nPeople)
+allShifts = range(nShifts)
+allDays   = range(nDays)
+
+model = cp_model.CpModel()
+
+# Create the decision variables.
+shifts = {}
+for d in allDays:
+    for s in allShifts:
+        for p in allPeople:
+            shifts[(d, s, p)] = model.NewBoolVar(f'shift[({d},{s},{p})]')
+
+# Add constraints.
+# Each shift, every day should have exactly one person assigned to it.
+for d in allDays:
+    for s in allShifts:
+        model.Add(sum(shifts[(d,s,p)] for p in allPeople) == 1)
+
+# Each person works is assigned at most one shift every day
+for p in allPeople:
+    for d in allDays:
+        model.Add(sum(shifts[(d,s,p)] for s in allShifts) <= 1)
+
+minShifts = (nShifts * nDays) // nPeople
+maxShifts = minShifts + ((nShifts * nDays) % nPeople)
+
+for p in allPeople:
+    nShiftsAssigned = 0
+    for d in allDays:
+        for s in allShifts:
+            nShiftsAssigned += shifts[(d,s,p)]
+
[…]

### 2020-09-01 optim/or-tools/scheduling/n-queens.py

@@ -1,0 +1,63 @@
+#!/bin/python
+
+import sys
+from ortools.sat.python import cp_model
+
+def main(size):
+    model = cp_model.CpModel()
+    # The decision variables. This is an array of 'size' number of queens,
+    # q1, q2, .... If qi = j then it means that a queen is located on the
+    # j-th row of the i'th column.
+    queens = [model.NewIntVar(0, size-1, f'q{i}') for i in range(size)]
+
+    # No more than one queen can occupy a position. Therefore,
+    model.AddAllDifferent(queens)
+
+    # We now add the diagonal constraints.
+    for i in range(size):
+        d1 = [] # One diagonal.
+        d2 = [] # Another diagonal.
+
+        for j in range(size):
+            # Decision variables for queens on ascending diagonal.
+            q1 = model.NewIntVar(0, 2*size, f'd1_{i}')
+            d1.append(q1)
+            model.Add(q1 == queens[j] + j) # Recall, ascending diagonal
+
+            # Decision variables for queens on descending diagonal.
+            q2 = model.NewIntVar(-size, size, f'd1_{i}')
+            d2.append(q2)
+            model.Add(q2 == queens[j] - j) # Recall, descending diagonal
+
+        model.AddAllDifferent(d1)
+        model.AddAllDifferent(d2)
+
+    solver = cp_model.CpSolver()
+    solutionPrinter = SolutionPrinter(queens)
+    status = solver.SearchForAllSolutions(model, solutionPrinter)
+    print()
+    print(f'Found {solutionPrinter.SolutionCount()} solutions.')
+
+class SolutionPrinter(cp_model.CpSolverSolutionCallback):
+    def __init__(self, variables):
+        cp_model.CpSolverSolutionCallback.__init__(self)
+        self._variables = variables
+        self._solnCount = 0
+
+    def on_solution_callback(self):
+        self._solnCount += 1
+
+        for v in self._variables:
+            print(f'{v}={self.Value(v)}', end = ' ')
+        print()
+
+    def SolutionCount(self):
+        return self._solnCount
+
+if __name__ == '__main__':
+    size = 8
+    if len(sys.argv) > 1:
+        size = int(sys.argv[1])
+
+    main(size)
+

### 2020-09-01 optim/or-tools/scheduling/n-queens.py
Updated comments

@@ -15,19 +15,19 @@ def main(size):
 
     # We now add the diagonal constraints.
     for i in range(size):
-        d1 = [] # One diagonal.
-        d2 = [] # Another diagonal.
+        d1 = [] # Diagonal rising from bottom left to top right.
+        d2 = [] # Diagonal falling from top left to bottom right.
 
         for j in range(size):
-            # Decision variables for queens on ascending diagonal.
+            # Decision variables for queens on rising diagonal.
             q1 = model.NewIntVar(0, 2*size, f'd1_{i}')
             d1.append(q1)
-            model.Add(q1 == queens[j] + j) # Recall, ascending diagonal
+            model.Add(q1 == queens[j] + j) # Recall, rising diagonal
 
-            # Decision variables for queens on descending diagonal.
+            # Decision variables for queens on falling diagonal.
             q2 = model.NewIntVar(-size, size, f'd1_{i}')
             d2.append(q2)
-            model.Add(q2 == queens[j] - j) # Recall, descending diagonal
+            model.Add(q2 == queens[j] - j) # Recall, falling diagonal
 
         model.AddAllDifferent(d1)
         model.AddAllDifferent(d2)

### 2020-09-01 optim/or-tools/cp/cp-optim.py
Some more CP problems.

@@ -1,0 +1,27 @@
+#!/bin/python
+
+from ortools.sat.python import cp_model
+
+model = cp_model.CpModel()
+solver = cp_model.CpSolver()
+
+# Decision variables
+ub = max([50, 45, 37])
+x = model.NewIntVar(0, ub, 'x')
+y = model.NewIntVar(0, ub, 'y')
+z = model.NewIntVar(0, ub, 'z')
+
+model.Add(2*x + 7*y + 3*z <= 50)
+model.Add(3*x - 5*y + 7*z <= 45)
+model.Add(5*x + 2*y - 6*z <= 37)
+
+model.Maximize(2*x + 2*y + 3*z)
+
+status = solver.Solve(model)
+if status == cp_model.OPTIMAL:
+    print(f'Objective function: {solver.ObjectiveValue()}')
+    print('Assignment:', end = ' ')
+    print(f'x={solver.Value(x)}, y={solver.Value(y)}, z={solver.Value(z)}')
+else:
+    print('Optimal solution could not be found.')
+

### 2020-09-01 optim/or-tools/cp/cryptarithmetic.py

@@ -1,0 +1,57 @@
+#!/bin/python
+
+from ortools.sat.python import cp_model
+
+# Is there an assignment of digits to the letter such that the equation
+# CP + IS + FUN = TRUE is true?
+
+line = 'CP + IS + FUN = TRUE'
+chars = set(c for c in line if c.isalpha())
+
+base = 10 # We are looking at decimals
+if len(chars) > base:
+    print('No assignment is possible')
+    exit(0)
+
+model = cp_model.CpModel()
+
+# Decision variables
+c = model.NewIntVar(1, base - 1, 'C')
+p = model.NewIntVar(0, base - 1, 'P')
+i = model.NewIntVar(1, base - 1, 'I')
+s = model.NewIntVar(0, base - 1, 'S')
+f = model.NewIntVar(1, base - 1, 'F')
+u = model.NewIntVar(0, base - 1, 'U')
+n = model.NewIntVar(0, base - 1, 'N')
+t = model.NewIntVar(1, base - 1, 'T')
+r = model.NewIntVar(0, base - 1, 'R')
+e = model.NewIntVar(0, base - 1, 'E')
+
+allVars = [c, p, i, s, f, u, n, t, r, e]
+# All letters need to be assigned a different number.
+model.AddAllDifferent(allVars)
+
+# Add the constraint in the line
+model.Add(c * base + p + i * base + s + f * base * base + u * base +
+              n == t * base * base * base + r * base * base + u * base + e)
+solver = cp_model.CpSolver()
+
+class SolutionPrinter(cp_model.CpSolverSolutionCallback):
+    def __init__(self, variables):
+        cp_model.CpSolverSolutionCallback.__init__(self)
+        self._variables = variables
+        self._solnCount = 0
+
+    def on_solution_callback(self):
+        self._solnCount += 1
+        for v in self._variables:
+            print(f'{v}={self.Value(v)}', end=' ')
+        print()
+
+    def solution_count(self):
+        return self._solnCount
+
+solutionPrinter = SolutionPrinter(allVars)
+status = solver.SearchForAllSolutions(model, solutionPrinter)
+print(f'status = {solver.StatusName(status)}')
+

### 2020-09-01 optim/or-tools/cp/feasibility.py

@@ -1,0 +1,39 @@
+#!/bin/python
+
+# Is there a feasible assignment to a constraint programming solution?
+
+from ortools.sat.python import cp_model
+
+class SolutionPrinter(cp_model.CpSolverSolutionCallback):
+    def __init__(self, variables):
+        cp_model.CpSolverSolutionCallback.__init__(self)
+        self._variables = variables
+        self._solnCount = 0
+
+    def on_solution_callback(self):
+        self._solnCount += 1
+        for v in self._variables:
+            print(f'{v}={self.Value(v)}', end = ' ')
+        print()
+
+    def solution_count(self):
+        return self._solnCount
+
+model = cp_model.CpModel()
+
+allowedVals = range(2)
+x = model.NewIntVar(min(allowedVals), max(allowedVals), 'x')
+y = model.NewIntVar(min(allowedVals), max(allowedVals), 'y')
+z = model.NewIntVar(min(allowedVals), max(allowedVals), 'z')
+
+model.Add(x != y)
+model.Add(y != z)
+model.Add(x + y + z <= 5)
+
+solver = cp_model.CpSolver()
+solutionPrinter = SolutionPrinter([x, y, z])
+status = solver.SearchForAllSolutions(model, solutionPrinter)
+
+print(f'Status = {solver.StatusName(status)}')
+print(f'Found {solutionPrinter.solution_count()} solutions.')
+

### 2020-09-01 optim/or-tools/cp/work-alloc-size.py

@@ -1,0 +1,70 @@
+#!/bin/python
+
+from ortools.sat.python import cp_model
+
+# A worker can do a task at a particular cost. The problem consists in 
+# allocating all workers to a task they can do at the minimal cost. There is
+# additional constraint, Each task has a size and a worker cannot be assigned
+# more tasks if the sum of their size exceeds a certain limit.
+
+# This problem can be solved with an MIP solver but a CP-SAT solver is a
+# better option because the decision variables are boolean.
+
+model = cp_model.CpModel()
+
+# The data for the problem described above can be represented as a 
+# bi-partite graph. The adjacency matrix for the graph is:
+bg = [[90, 76, 75, 70, 50, 74, 12, 68],
+      [35, 85, 55, 65, 48, 101, 70, 83],
+      [125, 95, 90, 105, 59, 120, 36, 73],
+      [45, 110, 95, 115, 104, 83, 37, 71],
+      [60, 105, 80, 75, 59, 62, 93, 88],
+      [45, 65, 110, 95, 47, 31, 81, 34],
+      [38, 51, 107, 41, 69, 99, 115, 48],
+      [47, 85, 57, 71, 92, 77, 109, 36],
+      [39, 63, 97, 49, 118, 56, 92, 61],
+      [47, 101, 71, 60, 88, 109, 52, 90]]
+task_sizes = [10, 7, 3, 12, 15, 4, 11, 5]
+total_size_max = 15
+
+n_workers = len(bg) # The number of rows.
+n_tasks   = len(bg[0]) # The number of columns.
+
+# The decision variable x_{i,j} is true is worker i is assigned task j and 
+# false otherwise. We now create the boolean decision variables.
+x = [False] * n_workers 
+for i in range(n_workers):
+    t = [False] * n_tasks
+    for j in range(n_tasks):
+        t[j] = model.NewBoolVar(f'x[{i}, {j}]')
+    x[i] = t
+
+# Add the constraints.
+# Each worker is assigned at most one task.
+for i in range(n_workers):
+    model.Add(sum(x[i][j] * task_sizes[j] for j in range(n_tasks)) <= total_size_max)
+# A task is allocated to exactly one worker.
+for j in range(n_tasks):
+    model.Add(sum(x[i][j] for i in range(n_workers)) == 1)
+
+# Build the objective function.
+terms = []
+for i in range(n_workers):
+    for j in range(n_tasks):
+        terms.append(bg[i][j] * x[i][j])
[…]

### 2020-09-01 optim/or-tools/cp/work-alloc-team.py

@@ -1,0 +1,71 @@
+#!/bin/python
+
+from ortools.sat.python import cp_model
+
+# A worker can do a task at a particular cost. The problem consists in 
+# allocating all workers to a task they can do at the minimal cost. There is
+# additional constraint, The workers are divided into two teams and each
+# team can work on at most a certain number of tasks.
+
+# This problem can be solved with an MIP solver but a CP-SAT solver is a
+# better option because the decision variables are boolean.
+
+model = cp_model.CpModel()
+
+# The data for the problem described above can be represented as a bi-partite
+# graph. The adjacency matrix for the graph is:
+bg = [[90, 76, 75, 70],
+      [35, 85, 55, 65],
+      [125, 95, 90, 105],
+      [45, 110, 95, 115],
+      [60, 105, 80, 75],
+      [45, 65, 110, 95]]
+n_workers = len(bg) # The number of rows.
+n_tasks   = len(bg[0]) # The number of columns.
+
+# The team structure.
+team1 = [0, 2, 4]
+team2 = [1, 3, 5]
+team_max = 2
+
+# The decision variable x_{i,j} is true is worker i is assigned task j and 
+# false otherwise. We now create the boolean decision variables.
+x = [False] * n_workers 
+for i in range(n_workers):
+    t = [False] * n_tasks
+    for j in range(n_tasks):
+        t[j] = model.NewBoolVar(f'x[{i}, {j}]')
+    x[i] = t
+
+# Add the constraints.
+# Each worker is assigned at most one task.
+for i in range(n_workers):
+    model.Add(sum(x[i][j] for j in range(n_tasks)) <= 1)
+# A task is allocated to exactly one worker.
+for j in range(n_tasks):
+    model.Add(sum(x[i][j] for i in range(n_workers)) == 1)
+# A team can have at most team_max tasks assigned to its members.
+model.Add(sum(x[i][j] for i in team1 for j in range(n_tasks)) <= team_max)
+model.Add(sum(x[i][j] for i in team2 for j in range(n_tasks)) <= team_max)
+
+# Build the objective function.
+terms = []
+for i in range(n_workers):
+    for j in range(n_tasks):
+        terms.append(bg[i][j] * x[i][j])
+model.Minimize(sum(terms))
+
+# Solve the problem,
+solver = cp_model.CpSolver()
+status = solver.Solve(model)
+
+if status == cp_model.OPTIMAL or status == cp_model.FEASIBLE:
[…]

### 2020-09-01 optim/or-tools/cp/work-alloc.py

@@ -1,0 +1,60 @@
+#!/bin/python
+
+from ortools.sat.python import cp_model
+
+# A worker can do a task at a particular cost. The problem consists in 
+# allocating all workers to a task they can do at the minimal cost.
+
+# This problem can be solved with an MIP solver but a CP-SAT solver is a
+# better option because the decision variables are boolean.
+
+model = cp_model.CpModel()
+
+# The data for the problem described above can be represented as a bi-partite
+# graph. The adjacency matrix for the graph is:
+bg = [[90, 80, 75, 70],
+      [35, 85, 55, 65],
+      [125, 95, 90, 95],
+      [45, 110, 95, 115],
+      [50, 100, 90, 100]]
+n_workers = len(bg) # The number of rows.
+n_tasks   = len(bg[0]) # The number of columns.
+
+# The decision variable x_{i,j} is true is worker i is assigned task j and 
+# false otherwise. We now create the boolean decision variables.
+x = [False] * n_workers 
+for i in range(n_workers):
+    t = [False] * n_tasks
+    for j in range(n_tasks):
+        t[j] = model.NewBoolVar(f'x[{i}, {j}]')
+    x[i] = t
+
+# Add the constraints.
+# Each worker is assigned at most one task.
+for i in range(n_workers):
+    model.Add(sum(x[i][j] for j in range(n_tasks)) <= 1)
+# A task is allocated to exactly one worker.
+for j in range(n_tasks):
+    model.Add(sum(x[i][j] for i in range(n_workers)) == 1)
+
+# Build the objective function.
+terms = []
+for i in range(n_workers):
+    for j in range(n_tasks):
+        terms.append(bg[i][j] * x[i][j])
+model.Minimize(sum(terms))
+
+# Solve the problem,
+solver = cp_model.CpSolver()
+status = solver.Solve(model)
+
+if status == cp_model.OPTIMAL or status == cp_model.FEASIBLE:
+    print(f'Min cost: {solver.ObjectiveValue()}')
+    print()
+    for i in range(n_workers):
+        for j in range(n_tasks):
+            if solver.BooleanValue(x[i][j]):
+                print(f'Worker {i} assigned to task {j} at cost {bg[i][j]}.')
+else:
+    print('The solver did not find a solution.')
+

### 2020-09-01 optim/or-tools/graph/max-flow.py
Flow problems.

@@ -1,0 +1,34 @@
+#!/bin/python
+
+from ortools.graph import pywrapgraph
+
+def main():
+    # The problem data. A directed edge (u, v) has a capacity c.
+    u = [0, 0, 0, 1, 1, 2, 2, 3, 3]
+    v = [1, 2, 3, 2, 4, 3, 4, 2, 4]
+    c = [20, 30, 10, 40, 30, 10, 20, 5, 20]
+
+    assert len(u) == len(v)
+    assert len(v) == len(c)
+
+    max_flow = pywrapgraph.SimpleMaxFlow()
+    for i in range(len(u)):
+        max_flow.AddArcWithCapacity(u[i], v[i], c[i])
+
+    # Solve the problem, find max flow between nodes 0 and 1.
+    status = max_flow.Solve(0, 4)
+    if status == max_flow.OPTIMAL:
+        print(f'Max flow: {max_flow.OptimalFlow()}')
+        print()
+        print('  Arc  Flow / Capacity')
+        for i in range(max_flow.NumArcs()):
+            print(f'{max_flow.Tail(i)} -> {max_flow.Head(i)} {max_flow.Flow(i)} / {max_flow.Capacity(i)}')
+
+        print(f'Source side min-cut: {max_flow.GetSourceSideMinCut()}')
+        print(f'Sink side min-cut: {max_flow.GetSinkSideMinCut()}')
+    else:
+        print('Unable to compute max-flow.')
+
+if __name__ == '__main__':
+    main()
+

### 2020-09-01 optim/or-tools/graph/min-cost.py

@@ -1,0 +1,38 @@
+#!/bin/python
+
+from ortools.graph import pywrapgraph
+
+def main():
+    # The network consists of edges (s, t) with capacity c and unit costs u.
+    s = [ 0, 0,  1, 1,  1,  2, 2,  3, 4]
+    t = [ 1, 2,  2, 3,  4,  3, 4,  4, 2]
+    c = [15, 8, 20, 4, 10, 15, 4, 20, 5]
+    u = [ 4, 4,  2, 2,  6,  1, 3,  2, 3]
+
+    # Each node can be either a supply node or a demand node. A demand is
+    # depicted by a negative supply.
+    supplies = [20, 0, 0, -5, -15]
+
+    solver = pywrapgraph.SimpleMinCostFlow()
+
+    for i in range(len(s)):
+        solver.AddArcWithCapacityAndUnitCost(s[i], t[i], c[i], u[i])
+
+    for i in range(len(supplies)):
+        solver.SetNodeSupply(i, supplies[i])
+
+    status = solver.Solve()
+    if status == solver.OPTIMAL:
+        print(f'Min cost: {solver.OptimalCost()}')
+        print()
+        print('  Arc  Flow / Capacity  Cost')
+        for i in range(solver.NumArcs()):
+            cost = solver.Flow(i) * solver.UnitCost(i)
+            print(f'{solver.Tail(i)} -> {solver.Head(i)} {solver.Flow(i)} / {solver.Capacity(i)}  {cost}')
+    else:
+        print('The solver could not find a min cost flow.')
+        print(f'status = {solver.StatusName(status)}')
+
+if __name__ == '__main__':
+    main()
+ 

### 2020-09-02 optim/or-tools/tsp/tsp.py
Simple travelling salesman problem.

@@ -1,0 +1,90 @@
+#!/bin/python
+
+from ortools.constraint_solver import routing_enums_pb2
+from ortools.constraint_solver import pywrapcp
+
+def create_data_model():
+    """Stores the data for the problem."""
+    data = {}
+    data['distance_matrix'] = [
+      [0, 2451, 713, 1018, 1631, 1374, 2408, 213, 2571, 875, 1420, 2145, 1972],
+      [2451, 0, 1745, 1524, 831, 1240, 959, 2596, 403, 1589, 1374, 357, 579],
+      [713, 1745, 0, 355, 920, 803, 1737, 851, 1858, 262, 940, 1453, 1260],
+      [1018, 1524, 355, 0, 700, 862, 1395, 1123, 1584, 466, 1056, 1280, 987],
+      [1631, 831, 920, 700, 0, 663, 1021, 1769, 949, 796, 879, 586, 371],
+      [1374, 1240, 803, 862, 663, 0, 1681, 1551, 1765, 547, 225, 887, 999],
+      [2408, 959, 1737, 1395, 1021, 1681, 0, 2493, 678, 1724, 1891, 1114, 701],
+      [213, 2596, 851, 1123, 1769, 1551, 2493, 0, 2699, 1038, 1605, 2300, 2099],
+      [2571, 403, 1858, 1584, 949, 1765, 678, 2699, 0, 1744, 1645, 653, 600],
+      [875, 1589, 262, 466, 796, 547, 1724, 1038, 1744, 0, 679, 1272, 1162],
+      [1420, 1374, 940, 1056, 879, 225, 1891, 1605, 1645, 679, 0, 1017, 1200],
+      [2145, 357, 1453, 1280, 586, 887, 1114, 2300, 653, 1272, 1017, 0, 504],
+      [1972, 579, 1260, 987, 371, 999, 701, 2099, 600, 1162, 1200, 504, 0],
+    ]  # yapf: disable
+    data['num_vehicles'] = 1
+    data['depot'] = 0
+    data['codes'] = {0: 'New York', 1: 'Los Angeles', 2: 'Chicago', \
+    3: 'Minneapolis', 4: 'Denver', 5: 'Dallas', 6: 'Seattle', 7: 'Boston', \
+    8: 'San Francisco', 9: 'St. Louis', 10: 'Houston', 11: 'Phoenix', \
+    12: 'Salt Lake City'}
+
+    return data
+
+def city(mgr, curr, codes):
+    """ Get's city from its index. """
[…]

### 2020-09-02 optim/or-tools/tsp/tsp.py
Minor refactoring.

@@ -34,7 +34,7 @@ def city(mgr, curr, codes):
     """ Get's city from its index. """
     return codes[mgr.IndexToNode(curr)]
 
-def print_solution(mgr, routing, solution, codes):
+def build_route(mgr, routing, solution, codes):
     """ Prints the route. """
     print(f'Distance covered: {solution.ObjectiveValue()} miles.')
     curr = routing.Start(0)
@@ -49,10 +49,16 @@ def print_solution(mgr, routing, solution, codes):
         route_distance += routing.GetArcCostForVehicle(prev, curr, 0)
 
     cities.append(city(mgr, curr, codes))
+
+    return {"route": cities, "distance": route_distance}
+
+def print_route(results):
+    cities = results["route"]
+    route_distance = results["distance"]
     print('Route:')
     print(' -> '.join([c for c in cities]))
     print(f'Route distance: {route_distance} miles')
-
+    
 def main():
     data = create_data_model()
 
@@ -83,7 +89,8 @@ def main():
     solution = routing.SolveWithParameters(params)
 
     if solution:
-        print_solution(manager, routing, solution, data['codes'])
+        results = build_route(manager, routing, solution, data['codes'])
+        print_route(results)
 
 if __name__ == '__main__':
     main()

### 2020-09-04 mortality-models/draft.tex
WIP 04Sep20

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}

\DeclareMathOperator{\pr}{P}

\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behaviour of a member of a 
population one can often make statistical statements with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

The death of an individual in a population or the decay of a radioactive atom
is an event happening to a member of the population. Likewise, the failure of
a machine, an occurrence of an event, the duration of a certain phase in an
economy are all different manifestations of the same phenomenon in which 
something happens that significantly alters the future of the member of the
population. Death terminates an individual's existence, radioactive decay
changes the atom, failure renders a machine unusable and economies change 
after occurrences of certain cataclysmic events. Mortality of an individual
is akin to reliability of a machine. Life of an individual is similar to
the duration of a certain phase of business cycle. As a result, the 
theoretical tools used to study these phenomena are similar to each other. 
Mortality analysis in actuarial science is called survival analysis is 
statistics, reliability theory in engineering, duration analysis in economics 
and event history analysis in sociology.

In this article we shall review the development mathematical models to study
mortality in large enough populations of people.

\section{de Moivre's law}\label{s1}
An annuity is a lifelong payment made to an individual by a company. The 
company would want to estimate how much annuity its customers will consume
before they cease to exist. In order to estimate its yearly obligations, the
company would need to know what proportion of its customers will survive
past various points in the future. It to this problem that de Moivre 
\cite{de1731annuities} addresses his treatise. In order to estimate the value
of annuities one must consider the interest earned by the company's assets and
the probability of survival of its customers. In de Moivre's days the interest
rates were controlled by law. Therefore, he focused on estimating the 
probability of survival of the company's customers.

Developed in the 18th century, de Moivre's model was indeed quite simple. 
He assumed that there is a certain age, $\omega$ beyond which no human 
survives. Further, as time goes by the probability of a person's survival
decreases. The probability that a new born person will survive at least
$x$ years was proposed to be
\begin{equation}\label{s1e1}
s(x) = 1 - \frac{x}{\omega}, 
\end{equation}
where $0 \le x \le \omega$. The function $s$, called the survival function,
has the following characteristics
\begin{enumerate}
\item It is a decreasing function of $x$.
\item $s(x = 0) = 1$ and $s(x = \omega) = 0$. In between these age limits, it
is a linear function of $x$.
\end{enumerate}
de Moivre proposed the use of different $\omega$'s for different age groups, 
making his model piece-wise linear. We will examine a few immediate 
consequences of de Moivre's model and while doing so we will introduce the
notation common in actuarial literature \cite{jordan1967society}. The symbol
$(x)$ denotes life aged $x$. If $k$ is an estimate of the size of a population
born together then the number of people alive at age $x$ is $l_x = ks(x)$.
The number of people who will die in the interval $x$ to $x + 1$ is $d_x = l_x
- l_{x+1}$. The probability that $x$ will survive for another $n$ years is
\begin{equation}\label{s1e2}
{}_np_x = \frac{l_{x+n}}{l_x} = \frac{s(x+n)}{s(x)}.
\end{equation}
The probability that a person will die within next $n$ years of $x$ is
\begin{equation}\label{s1e3}
{}_nq_x = 1 - {}_np_x.
\end{equation}
It is common to denote ${}_1p_x$ as $p_x$ and ${}_nq_x$ as $q_x$. The `force
of mortality' is defined as
\begin{equation}\label{s1e4}
\mu(x) = -\frac{s^\prime(x)}{s(x)}, 
\end{equation}
where a $s^\prime$ denotes the derivative of $s$ with respect to $x$. The 
force of mortality is also called the hazard function in the statistical
literature. For de Moivre's law of mortality, it is easy to confirm that
\begin{eqnarray}
{}_np_x &=& \frac{\omega - (x + n)}{\omega - x} \label{s1e5} \\
{}_nq_x &=& \frac{n}{\omega - x} \label{s1e6} \\
\mu(x) &=& \frac{1}{\omega - x} \label{s1e7}
\end{eqnarray}
de Moivre himself did not consider his law to be an accurate description of
human mortality. He emphasized in his treatise that it was an approximation
useful to calculate the annuities. His piecewise linear approximation was a
valuable device to carry out practical computations in the eighteenth century.

\bibliography{draft}
\bibliographystyle{plain}
\end{document}

### 2020-09-04 p5/ode-2.py

@@ -1,0 +1,47 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Sep  3 19:39:33 2020
+
+@author: ajoshi
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import odeint
+
+def f1(h, alpha = 3*np.pi/4):
+  [x, y] = h
+  xdot = 0.5*(np.tanh(0.75*(x*np.cos(alpha)-y*np.sin(alpha))) - x)
+  ydot = 0.5*(np.tanh(0.75*(x*np.sin(alpha)+y*np.cos(alpha))) - y)
+
+  return [xdot, ydot]
+
+# Create the grid in the domain [-1, 1] x [-1, 1].
+lo, hi = -0.5, 0.5
+X, Y = np.meshgrid(np.linspace(lo, hi, 21), np.linspace(lo, hi, 21))
+u, v = np.zeros(X.shape), np.zeros(Y.shape)
+
+for i in range(X.shape[0]):
+  for j in range(X.shape[1]):
+    xp = f1([X[i, j], Y[i, j]])
+    u[i,j] = xp[0]
+    v[i,j] = xp[1]
+   
+Q = plt.quiver(X, Y, u, v, color='r')
+plt.title(r'$\alpha = 3\pi/4$')
+plt.xlabel('$x$')
+plt.ylabel('$y$')
+plt.xlim([lo, hi])
+plt.ylim([lo, hi])
+x_inits = [[0.5, 0.5], [0.25, 0.5], [0, 0.5], [-0.25, 0.5], [-0.5, 0.5], \
+          [-0.5, 0.25], [-0.5, 0], [-0.5, -0.25], [-0.5, -0.5], \
+          [-0.25, -0.5], [0, -0.5], [0.25, -0.5], [0.5, -0.5], \
+          [0.5, -0.25], [0.5, 0], [0.5, 0.25]]
+
+for x_init in x_inits:  
+  times = np.linspace(0, 50, 2000)  
+  xs = odeint(f1, x_init, times)
+  plt.plot(xs[:,0], xs[:,1], '-', color = 'black') # path
+  plt.plot([xs[0,0]], [xs[0,1]], 'o', markersize = 2) # start
+  plt.plot([xs[-1,0]], [xs[-1,1]], 's', markersize = 2) # end

### 2020-09-04 p5/ode-3.py

@@ -1,0 +1,45 @@
+# -*- coding: utf-8 -*-
+"""
+Spyder Editor
+
+This is a temporary script file.
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import odeint
+
+def f2(h, alpha = np.pi/3.8):
+  [x, y] = h
+  xdot = 0.5*(np.tanh(0.75*(x*np.cos(alpha)-y*np.sin(alpha))) - x)
+  ydot = 0.5*(np.tanh(0.75*(x*np.sin(alpha)+y*np.cos(alpha))) - y)
+
+  return [xdot, ydot]
+
+# Create the grid in the domain [-1, 1] x [-1, 1].
+lo, hi = -0.5, 0.5
+X, Y = np.meshgrid(np.linspace(lo, hi, 21), np.linspace(lo, hi, 21))
+u, v = np.zeros(X.shape), np.zeros(Y.shape)
+
+for i in range(X.shape[0]):
+  for j in range(X.shape[1]):
+    xp = f2([X[i, j], Y[i, j]])
+    u[i,j] = xp[0]
+    v[i,j] = xp[1]
+   
+Q = plt.quiver(X, Y, u, v, color='r')
+plt.title(r'$\alpha = \pi/6$')
+plt.xlabel('$x$')
+plt.ylabel('$y$')
+plt.xlim([lo, hi])
+plt.ylim([lo, hi])
+x_inits = [[0.5, 0.5], [0.25, 0.5], [0, 0.5], [-0.25, 0.5], [-0.5, 0.5], \
+          [-0.5, 0.25], [-0.5, 0], [-0.5, -0.25], [-0.5, -0.5], \
+          [-0.25, -0.5], [0, -0.5], [0.25, -0.5], [0.5, -0.5], \
+          [0.5, -0.25], [0.5, 0], [0.5, 0.25]]
+
+for x_init in x_inits:  
+  times = np.linspace(0, 50, 2000)  
+  xs = odeint(f2, x_init, times)
+  plt.plot(xs[:,0], xs[:,1], '-', color = 'black') # path
+  plt.plot([xs[0,0]], [xs[0,1]], 'o', markersize = 2) # start
+  plt.plot([xs[-1,0]], [xs[-1,1]], 's', markersize = 2) # end

### 2020-09-05 mortality-models/draft.tex
WIP 05Sep2020

@@
 \usepackage{amsmath, amssymb, amsthm, amsfonts}
-
+\numberwithin{equation}{section}
 \DeclareMathOperator{\pr}{P}
@@
 force of mortality is also called the hazard function in the statistical
-literature. For de Moivre's law of mortality, it is easy to confirm that
+literature. The hazard function or the `force of mortality' is an individual's
+susceptibility to die. Its reciprocal is the individual's resistance to death. 
+de Moivre's law proposes that a person's resistance to die varies linearly over
+time. That is,
+\begin{equation}\label{s1e5}
+\nu(x) = \frac{1}{\mu(x)} = \omega - x.
+\end{equation}
+If $x_0 = a, x_1 = a + d, x_2 = a + 2d, ..., x_n = a + nd$ then it is clear 
+that $\nu(x_n)$ form an algebraic progression. This point may seem to be just a 
+mathematical characteristic at this stage. We shall see in the next section that
+it marked a departure from de Moivre's model to a more realistic one. For de 
+Moivre's law of mortality, it is easy to confirm that
 \begin{eqnarray}
-{}_np_x &=& \frac{\omega - (x + n)}{\omega - x} \label{s1e5} \\
-{}_nq_x &=& \frac{n}{\omega - x} \label{s1e6} \\
-\mu(x) &=& \frac{1}{\omega - x} \label{s1e7}
+{}_np_x &=& \frac{\omega - (x + n)}{\omega - x} \label{s1e6} \\
+{}_nq_x &=& \frac{n}{\omega - x} \label{s1e7} \\
+\mu(x) &=& \frac{1}{\omega - x} \label{s1e8}
 \end{eqnarray}
@@
 
+\section{Gompertz's law}\label{s2}
+An immediate consequence of de Moivre's law is that an individual's resistance
+to death diminishes algebraically as time goes by. The English actuary 
+Benjamin Gompertz observed \cite{gompertz1825xxiv} that the resistance to 
+death decreases geometrically instead. If one considers small enough time 
+intervals then at the end of every interval a person's resistance to death 
+reduces by the same proportion. Expressed mathematically, it means that
+\begin{equation}\label{s2e1}
+\frac{d\nu}{dx} = -k\nu(x),
+\end{equation}
+where the function $\nu$ was defined in equation \eqref{s1e5}. It immediately
+follows that
+\begin{equation}\label{s2e2}
+\log\nu(x) = -kx + \log B,
+\end{equation}
+where $B$ is a constant of integration. The `force of mortality' or the
+hazard function is therefore
+\begin{equation}\label{s2e3}
+\mu(x) = Be^{kx}.
+\end{equation}
+The constant $B$ is called baseline mortality and $k$ is called the 
+senescent component.
+An individual's susceptibility to die thus increases exponentially with age.
+From equation \eqref{s1e4} and \eqref{s2e3}, the differential equation for
+the survival function is
+\begin{equation}\label{s2e4}
+\frac{s^\prime(x)}{s(x)} = Be^{kx}.
+\end{equation}
+The solution of this ordinary differential equation is
+\begin{equation}\label{s2e5}
+s(x) = C\exp\left(\frac{B}{k}\exp(kx)\right),
+\end{equation}
+where $C$ is another constant of integration. The constants $B$ and $C$ in
+equation \eqref{s2e5} are determined from the mortality data of the population
+under study. Gompertz suggested that his formula is applicable for the age
+group between $10-15$ years of age up to $55-60$. In order to extend it beyond
+the age of $60$, Gompertz recommended that a different set of constants $B$
+and $C$ be used.
+
+Gompertz was aware that death in a population has at least two factors. One
+of them is the deterioration in the individual's resistance to death. The 
+other one is chance alone. The first factor depends on the individual's age
+while the second factor is constant throughout. However, in the formulation
+of his law of mortality, Gompertz considered only the first factor.
+
+\section{Makeham's law}\label{s3}
+William Makeham, the English mathematician and actuary, added a term to
+Gompertz's law of resistance to death which the latter had acknowledged but
+not included. Makeham's modification \cite{makeham1867law} of equation 
+\eqref{s2e3} is
+\begin{equation}\label{s3e1}
+\mu(x) = A + Be^{kx},
+\end{equation}
+where $A$ is a constant independent of the person's age. Makeham's survival
+function is
+\begin{equation}\label{s3e2}
+s(x) = C\exp\left(Ax + \frac{B}{k}\exp(kx)\right).
+\end{equation}
+Makeham's survival function has three constants $A, B$ and $C$ and therefore
+can be fitted to emprically observed mortality data more effectively than
+Gompertz's model. Makeham's law was often applied to mortality tables in the
+age range beyond $20$ years to almost end of life\cite{jordan1967society}.
 \bibliography{draft}

### 2020-09-07 magnetic-fields/solution.tex
Electrodynamics

\documentclass{article}
\usepackage{amsmath, amsthm, amssymb, amsfonts}
\title{Two problems in Electrodynamics}
\begin{document}
\begin{enumerate}
\item Consider a finite solenoid of length $L$ and cross-sectional area 
$A = \pi R^3$. Let there be $n$ turns of the wire per unit length and 
let $I$ be the current passing through them.

We know that $\nabla\cdot\vec{B}$ is always zero. Imagine a hemisphere whose
base is perpendicular to the solenoid's axis and intersects it. Let the
hemisphere be of radius $a$. Let $H$ denote the surface of the hemisphere
and $V$ denote its volume. Then,
\[
\int_V \nabla\cdot\vec{B}dV = 0 \Rightarrow \int_H \vec{B}\cdot\hat{n}da = 0,
\]
where $\hat{n}$ is the unit outward normal. We can write the surface integral 
as the sum over the base $H_1$ and the curved surface $H_2$. Thus,
\[
\int_{H_1}\vec{B}\cdot\hat{n}da = -\int_{H_2}\vec{B}\cdot\hat{n}da.
\]
As $a \rightarrow \infty$, the field drops as $1/a^3$ while the surface area
rises as $a^2$. Effectively, the integrand drops as $1/a$. Thus, as $a 
\rightarrow \infty$, the integrand on the right hand side of the previous
equation vanishes and so does the integral. Therefore, as $a \rightarrow 
\infty$,
\[
\int_{H_1}\vec{B}\cdot\hat{n}da = 0.
\]
The integrand in this integral is also called the magnetic flux. As $\vec{B}
= 0$ we know that the field lines inside the solenoid are in a direction
opposite to that outside. If $\Phi_i$ is the flux due to the field inside the
solenoid and $\Phi_o$ is that due to the field outside then the previous
equation can be written as 
\[
\Phi_i + \Phi_o = 0.
\]
Further, we know that $\Phi_i = \mu_0 nIA$. Therefore,
\[
\Phi_o = -\mu_0 nIA.
\]
We now consider a square on the surface $H_1$ whose centre lies on the 
axis of the solenoid. If $B$ is the field in this square then it contributes
a flus $BL^2$. The portion of $H_1$ beyond this square does not contribute
significantly because the field drops off as $r^{-3}$. Thus, close enough
to the solenoid the magnitude of the field is
\[
B_{\text{close}} \approx \frac{\mu_0 nIA}{L^2},
\]
and far away from the solenoid it is essentially zero.

\item Given that a charge $Q$ is uniformly distributed on a sphere of radius
$R$. Therefore, its surface charge density is
\begin{equation}\label{p2e1}
\sigma = \frac{Q}{4\pi R^2}.
\end{equation}
The sphere is rotating at a uniform angular velocity of $\vec{\omega}$. We want
to find the magnetic field at an arbitrary point, say $P$. We choose to set
the origin of our coordinate axes to coincide with the centre of the sphere and
we align the $z$ axis to lie along the line $OP$. Let the angle between the
constant vector $\vec{\omega}$ and the $z$ axis by $\psi$. Note that $\psi$
is a constant in our problem. We also align our $x$-axis so that $\vec{\omega}$
lies in the $xz$ plane. We can then write it as
\begin{equation}\label{p2e2}
\vec{\omega} = \omega\sin\psi\hat{i} + \omega\cos\psi\hat{k}.
\end{equation}

The rotation of a charged sphere gives rise to a current density $\vec{J} = 
\sigma\vec{v}$, where $\sigma$, the charge density is given by equation 
\eqref{p2e1} and $\vec{v}$ is the velocity of the point on the sphere at 
which we are measuring $\vec{J}$. Since $\vec{v} = \vec{\omega}\times\vec{r}$
for a uniformly rotating sphere,
\[
\vec{J} = \sigma\vec{\omega} \times \vec{r}.
\]
From equation \eqref{p2e2} and the fact that $\vec{r} = R\sin\theta\cos\phi
\hat{i} + R\sin\theta\sin\phi\hat{j} + R\cos\theta\hat{k}$, we get
\begin{eqnarray}
\vec{J}&=&\sigma R\omega\Big(-\cos\psi\sin\theta\sin\phi\hat{i} \nonumber \\
 & & (\cos\psi\sin\theta \cos\phi - \sin\psi\cos\theta)\hat{j} + \nonumber  \\
 & & \sin\psi\sin\theta\sin\phi\hat{k}\Big)\label{p2e3}
\end{eqnarray}
In order to find the magnetic field due to this current distribution, we 
first find the vector potential. If $\vec{s}$ is the position vector of
the point $P$ then 
\begin{equation}\label{p2e4}
\vec{A}(\vec{s}) = \frac{\mu_0}{4\pi}\int\frac{\vec{J}}{|\vec{s} - \vec{x}|}
da,
\end{equation}
where $\vec{x}$ is a point on the sphere and $da = R^2\sin\theta d\theta d\phi$
is the `area element' on the sphere's surface. From equations \eqref{p2e3}
and \eqref{p2e4} it is clear that we have three integrals to evaluate, one for 
each component. Thus,
\begin{eqnarray}
A_x &=& -\frac{\mu_0}{4\pi}\int_{S}\frac{\sigma R\omega\cos\psi\sin\theta\sin\phi}
{|\vec{s} - \vec{x}|} R^2\sin\theta d\theta d\phi \label{p2e5} \\
A_y &=& \frac{\mu_0}{4\pi}\int_{S}\frac{\sigma R\omega(\cos\psi\sin\theta \cos\phi - \sin\psi\cos\theta)}
{|\vec{s} - \vec{x}|} R^2\sin\theta d\theta d\phi \label{p2e6} \\
A_z &=& \frac{\mu_0}{4\pi}\int_{S}\frac{\sigma R\omega\sin\psi\sin\theta\sin\phi\sin\theta\sin\phi}
{|\vec{s} - \vec{x}|} R^2\sin\theta d\theta d\phi \label{p2e7} 
\end{eqnarray}
The subscript $S$ indicates that the integration is over the surface of the
sphere. Now, the $\theta$ integral goes from $0$ to $\pi$ and the $\phi$ 
integral goes from $0$ to $2\pi$. Since
\begin{eqnarray*}
\int_0^{2\pi}\sin\phi d\phi &=& 0 \\
\int_0^{2\pi}\cos\phi d\phi &=& 0
\end{eqnarray*}
the only surviving integral among equations \eqref{p2e5} to \eqref{p2e6} is
\[
A_y = -\frac{\mu_0}{4\pi}\int_0^\pi 2\pi 
\frac{\sigma R\omega(\sin\psi\cos\theta)}{|\vec{s} - \vec{x}|} R^2\sin\theta d\theta 
\]
The vector potential is
\begin{equation}\label{p2e8}
\vec{A}(\vec{s}) = -\frac{\mu_0 R^3\sigma\omega\sin\psi}{2}
\int_0^\pi \frac{\cos\theta\sin\theta d\theta}{|\vec{s} - \vec{x}|}\hat{j}.
\end{equation}
From the geometry of the problem it is clear that
\[
|\vec{s} - \vec{x}| = \sqrt{s^2 + R^2 - 2Rs\cos\theta}
\]
so that
\begin{equation}\label{p2e9}
\vec{A}(\vec{s}) = -\frac{\mu_0 R^3\sigma\omega\sin\psi}{2}
\int_0^\pi \frac{\cos\theta\sin\theta d\theta}
{\sqrt{s^2 + R^2 - 2Rs\cos\theta}}\hat{j}.
\end{equation}

In order to carry out this integration, put $u = \cos\theta$ so that
\begin{equation}\label{p2e10}
\vec{A}(\vec{s}) = -\frac{\mu_0 R^3\sigma\omega\sin\psi}{2}
\int_{-1}^1\frac{udu}{\sqrt{R^2 + s^2 - 2Rsu}}\hat{j}
\end{equation}
The evaluation of this integral needs care. We first evaluate the corresponding
indefinite integral,
\[
I = \int \frac{udu}{\sqrt{R^2 + s^2 - 2Rsu}}
\]
and introduce the variable $v^2 = R^2 + s^2 - 2Rsu$ so that
\[
du = -\frac{vdv}{Rs}
\]
and hence
\begin{eqnarray*}
I &=& -\int \frac{v}{Rs}\frac{R^2 + s^2 - v^2}{2Rs}\frac{dv}{v} \\
 &=& -\int (R^2 + s^2 - v^2)dv \\
 &=& -\frac{1}{2R^2s^2}\left((R^2+s^2)v - \frac{v^3}{3}\right) \\
 &=& -\frac{v}{2R^2s^2}\left(R^2+s^2 - \frac{R^2+s^2-2Rsu}{3}\right) \\
 &=& -\frac{v}{3R^2s^2}(R^2 + s^2 + Rsu) \\
 &=& -\frac{R^2+s^2+Rsu}{3R^2s^2}\sqrt{R^2 + s^2 - 2Rsu}
\end{eqnarray*}
Substituting this in equation \eqref{p2e10} we get
\begin{equation}\label{p2e11}
\vec{A} = \frac{\mu_0R^3\sigma\omega\sin\psi}{2}
\left[\frac{R^2+s^2+Rsu}{3R^2s^2}\sqrt{R^2 + s^2 - 2Rsu}\right]_{-1}^1\hat{j}.
\end{equation}
The square root takes a value that depends on whether $s > R$ or $s < R$. We
also observe that $-\omega s \sin\psi\hat{j} = \vec{\omega}\times\vec{s}$ so 
that
\begin{equation}\label{p2e12}
\vec{A}(\vec{s}) = \begin{cases}
\frac{\mu_0 R\sigma}{3}(\vec{\omega} \times \vec{s}) \text{ if } s < R \\
\frac{\mu_0 R^4\sigma}{3s^2}(\vec{\omega} \times \vec{s}) \text{ if } s > R
\end{cases}
\end{equation}

We now rotate the axes so that $\vec{\omega}$ is along the $z$-axis whence
\begin{eqnarray}
\vec{A}(r, \theta, \phi)&=& \frac{\mu_0R\omega\sigma}{3}r\sin\theta\hat{\phi} \text{ if } r \le R \nonumber \\
 &=& \frac{\mu_0R^4\omega\sigma}{3}\frac{\sin\theta}{r^2}\hat{\phi} \text{ if } r \ge R \label{p2e13}
\end{eqnarray}
Therefore, the magnetic field is
\begin{eqnarray}
\vec{B}(\vec{r}) &=& \frac{2}{3}\mu_0 R\sigma\vec{\omega} \text{ if } r \le R \nonumber \\
 &=& \frac{\mu_0R^4\omega\sigma}{3}\left(\frac{2\cos\theta}{r^3}\hat{r} + \frac{\sin\theta}{r^3}\hat{\theta}\right) \text{ if } r \ge R \label{p2e14}
\end{eqnarray}

We are now in a position to answer the four questions related to this
problem.
\begin{enumerate}
\item[(a)] Since $\vec{J} = \sigma\vec{v} = \sigma\vec{\omega}\times\vec{r}
= \vec{M}\times\vec{r}$, where the constant vector $\vec{M} = 
\sigma\vec{\omega}$. In order to get the desired form of $\vec{J} = 
\vec{M} \times \Theta$, we need to find a function $\Theta$ such 
that $\nabla\Theta = \vec{r}$. One such choice is $\Theta(x, y, z) = 
x + y + z$ or $\Theta(x - R, y - R, z - R) = x - R + y - R + z - R$, 
both of which that the same gradient.
\item[(b)] The magnetic moment is also defined through the relation,
\[
\vec{A} = \frac{\mu_0}{4\pi} \frac{\vec{m}\times\vec{r}}{r^3}.
\]
Comparing this with the second of equations \eqref{p2e12} we get
\[
\vec{m} = \frac{4\pi}{3}R^4\sigma\vec{\omega}.
\]
\item[(c)] The vector potential is given by equation \eqref{p2e13}.
\item[(d)] The magnetic field is given by equation \eqref{p2e14}.
\end{enumerate}

\end{enumerate}
\end{document}

### 2020-09-07 mortality-models/draft.tex
Electrodynamics

@@
 
-\section{Makeham's law}\label{s3}
+\section{Makeham's law and its generalisations}\label{s3}
 William Makeham, the English mathematician and actuary, added a term to
@@
 age range beyond $20$ years to almost end of life\cite{jordan1967society}. 
+In 1890\cite{makeham1890further}, Makeham proposed yet another modification to 
+his law by adding a linear term to the hazard function or the force of 
+mortality. Makeham thought that in addition to the two factors spotted by
+Gompertz there is another one that grows linearly with age. It was proposed to 
+be
+\begin{equation}\label{s3e3}
+\mu(x) = A_1 + A_2x + Be^{kx}.
+\end{equation}
+The resulting survival function is
+\begin{equation}\label{s3e4}
+s(x) = C\exp\left(A_1x + \frac{A_2}{2}x^2 + \frac{B}{k}\exp(kx)\right).
+\end{equation}
+Makeham's law, like its predecessor, Gompertz's law was used to construct
+mortality tables of adults. Their simplicity allowed them to be used 
+effectively up to as late as the 1940s. Gompertz law was used to construct
+the 1937 Standard Annuity Table and Makeham's law for 1940 Annuity table
+\cite{jordan1967society}. Both laws assumed that the force of mortality 
+increases monotonically with time. However, it is commonly observed that 
+infant mortality contradicts this assumption. In order to extend Makeham's
+law of 1860 to infantile ages, Wilhem Lazarus \cite{lazarus1867ueber} added
+an exponentially decreasing term. His law expressed the force of mortality
+as 
+\begin{equation}\label{s3e5}
+\mu(x) = B_1 e^{-k_1x} + A + B_2e^{k_2 x}.
+\end{equation}
+The constants $B_1$ and $B_2$ are positive numbers. The corresponding survival 
+function is
+\begin{equation}\label{s3e6}
+s(x) = 
+C\exp\left(-\frac{B_1}{k_1}e^{-k_1x} + Ax + \frac{B_2}{k_2}e^{k_2x}\right)
+\end{equation}
+
+The way infants' susceptibility to die of diseases upsets Gompertz's and 
+Makeham's assumption of monotonicity of the law of mortality, do does the 
+spike in the death rate of young adults. Wars were frequent in Europe of the 
+eighteenth century and the victims were disproportionately the young recruits.
+Even otherwise, young adults are more prone to death by accidents. In order
+to take into account this mortality hump in the young, Thiele 
+\cite{thiele1871mathematical} introduced a gaussian term to Lazarus' model.
+His law of mortality is
+\begin{equation}\label{s3e7}
+\mu(x) = B_1 e^{-k_1x} + Ae^{-a_1(x - a_2)^2} + B_2e^{k_2 x}.
+\end{equation}
+All the constants in this equation, namely $A, a_1, a_2, B_1, k_1, B_2$ and 
+$k_2$ are positive numbers. The presence of the gaussian term prevents us
+from writing a closed form expression of the survival function. However, the
+ordinary differential equation for the survival function can be readily 
+integrated numerically. Alternatively, if the initial conditions are available,
+one can write its solution in terms of the error function.
+
+We observe that all generalisations of Makeham's laws introduce additional
+constants to the force of mortality and the hazard function. These constants
+provide additional `degrees of freedom' to the functions used to fit the 
+mortality data, resulting in a better fit.
+
+Forfar, McCutcheon and Wilkie proposed a far wider generalisation of the
+Gompertz and Makeham's laws. They defined a class of models, called GM models
+after their inventors, with the mathematical form
+\begin{equation}\label{s3e8}
+\mu(x) = \sum_{i=0}^{r-1}\alpha_i x^i + 
+\exp\left(\sum_{i=1}^{s-1}\beta_i x^i\right).
+\end{equation}
+They also used the convention that if $r = 0$ then the polynomial terms are 
+absent and if $s = 0$ then the exponential terms do not appear. They called
+the law of force of mortality of equation \eqref{s3e8} as $\mathrm{GM}(r,s)$.
+It is easy to check that Gompertz's law is $\mathrm{GM}(0, 2)$, Makeham's
+first law is $\mathrm{GM}(1,2)$ and Makeham's second law is $\mathrm{GM}(2,2)$.
+The $\mathrm{GM}(0, 2)$, $\mathrm{GM}(2,2)$ and $\mathrm{GM}(1,3)$ are used
+by the Continuous Mortality Investigation Bureau\cite{pitacco2016high} in the 
+United Kingdom.
+
+\section{Other mortality laws prior to the 20th century}\label{s4}
+de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
+ones in vogue in the nineteenth century and prior to it. In this section we
+mention a few notable ones without commenting on their merits. Lambert's
+survival function \cite{lambert1776dottrina} was
+\begin{equation}\label{s4e1}
+s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
+\end{equation}
+It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
+quadratic survival function
+\begin{equation}\label{s4e2}
+s(x) = c - bx - ax^2.
+\end{equation}
+Young proposed a polynomial of degree $40$ as the survival function in 
+\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
+extended Babbage's model to polynomials of higher degree. Moser
+\cite{moser1839gesetze} proposed a polynomial survival function with 
+fractional powers of $x$. It has five degrees of freedom and had the form
+\begin{equation}\label{s4e3}
+s(x) = 1 - ax^{1/4} _ bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
+\end{equation}
+All these models were attempts to fit the mortality data with a continuous 
+function. They were not rooted in demographic observations like the models
+of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
+disuse after Gompertz proposed his model and others improved upon it.
 \bibliography{draft}

### 2020-09-07 mortality-models/draft.tex
WIP 07Sep20

@@
 disuse after Gompertz proposed his model and others improved upon it.
+
+\section{Heligman-Pollard model}\label{s5}
+The term `odds' is a well-defined term in probability theory. When we are 
+dealing with an experiment with two outcomes - that something happens or 
+it does not - then it is common in certain fields to express the 
+probabilities in terms of odds.  For example, if $p$ is the probability 
+that an event will happen and $q = 1 - p$ is the probability that it will 
+not happen then the odds of the experiment is $p/q$.
+
+Recall from equations \eqref{s1e2} and \eqref{s1e3} that $p_x$ is the 
+probability that life aged $x$ will survive for another year and $q_x$ is
+the probability of the complimentary event. Therefore, the odds of death in 
+this case are $q_x/p_x$.  Heligman and Pollard \cite{heligman1980age} proposed 
+a model that has the algebraic structure of Thiele's mortality law but 
+applied to odds of survival. Mathematically it is expressed as
+\begin{equation}\label{s5e1}
+\frac{q_x}{p_x} = A^{(x+B)^C}+
+D\exp\left(-E\log^2\left(\frac{x}{F}\right)\right) + GH^x.
+\end{equation}
+The model has eight parameters, $A, B, C, D, E, F, G$ and $H$. Of these, $A,
+B, C, D \in (0, 1)$, $E > 0$, $F \in (15, \omega)$, $G \in (0, 1)$ and $H
+> 0$. The constant $\omega$ is the hump maximum and it can be between $55$
+and $100$. The Heligman-Pollard model was first developed for the Australian
+population post the second world war. An advantage of Heligman-Pollard model 
+is in biological interpretation of each of the three terms on the right hand 
+side of equation \eqref{s5e1}.
+
+The first term on the right hand side of equation \eqref{s5e1} represent 
+infantile mortality. It is a fast declining exponential representing the 
+rapid fall in mortality during infancy. $A$ is approximately equal to $q_1$ 
+and is a measure of the mortality rate of a one year old child. $C$ is a 
+measure of how quickly the new-born adjusts to its surroundings and gains 
+immunity. A higher $C$ indicates the fast decline in infant mortality as the
+child passes through infancy. When $B = 0$, $q_0 = 0.5$ irrespective of the
+values taken by $A$ and $C$. The value of $q_0$ is closer to $q_1$ if $B$ is
+higher for a given value of $C$. $B$ is thus an indicator of the central 
+tendancy of infant mortality.
+
+The second term is gaussian in the logarithm of age. It represent the accident
+mortality for young adults of both sexes and maternal mortality in young 
+females. It covers the `accident hump` observed in all demographic data. It
+is prominently observed between ages $10$ and $40$. The parameter $D$ controls 
+the severity of accident hump, $E$ determines the spread of the hump and $F$ 
+the location.  Using a statistical analogy, $E$ is a measure of spread and 
+$F$ that of the central tendency.
+
+The third term traces its roots to the Gompertz model. It represents the
+exponential increase in the deterioration of the body. The parameter $G$ 
+represents the senescent mortality and $H$ the increase in the rate of that
+process.
+
 \bibliography{draft}

### 2020-09-08 mortality-models/draft.tex
WIP 08Sep20

@@
 \DeclareMathOperator{\pr}{P}
+\renewcommand{\vec}[1]{\mathbf{#1}}
 
@@
 \section{Introduction}\label{s0}
-Although it is very difficult to predict the behaviour of a member of a 
-population one can often make statistical statements with much greater 
+Although it is very difficult to predict the behaviour of a member of a large
+population one can often make statistical statements about the same population 
+with much greater 
 confidence. For example, it is impossible to predict when a particular 
@@
 many atoms will decay in a long enough time interval. Likewise, it is not
-possible to predict the time of death of an individual although one reliably
+possible to predict the time of death of an individual although one can reliably
 predict how many people from a large enough group will survive past a certain
@@
 population. Death terminates an individual's existence, radioactive decay
-changes the atom, failure renders a machine unusable and economies change 
+transforms the atom, failure renders a machine unusable and economies change 
 after occurrences of certain cataclysmic events. Mortality of an individual
@@
 the duration of a certain phase of business cycle. As a result, the 
-theoretical tools used to study these phenomena are similar to each other. 
+theoretical tools used to study these outwardly different phenomena are similar 
+to each other. 
 Mortality analysis in actuarial science is called survival analysis is 
@@
 
+John Graunt, an English haberdasher, was perhaps the world's first demographer
+and epidemiologist. In spite of the rudimentary mathematical tools available
+in the seventeenth century, he was able to publish the first life tables. They
+were based on the `Bills of Mortality' published in the parishes of London
+every Thursday. The data was not best suited for constructing life tables but
+that was the only one available in Graunt's time. Graunt admitted the 
+inadequacy of his approach. Edmund Halley, the man after whom a comet is 
+named, could get demographic data of a much better quality from the city of
+Wroclaw in Poland. It had the age and sex of the dead individuals. Halley
+used this data to build life tables of a greater accuracy than Graunt's. He
+also demonstrated how to use the life tables to compute annuities. Graunt's
+and Halley's work was entirely empirical, keeping with the times in which
+it was done. Using their life tables was also quite laborious. With the 
+invention of Calculus in the late seventeenth century and the development of 
+the rudimentary ideas of a mathematical function, it became possible to build
+theoretical models of human mortality.
+
+Traditionally, mortality models studied existing demographic data to produce
+mortality tables. They described mortality at a fixed point in time. The pace 
+of technology development was quite slow in the times when the first mortality
+models were proposed. Therefore, the mortality data and the models describing
+them appeared to be like a natural characteristic of the human species. They
+sufficed the needs of the fledgling insurance industry to compute projections
+of annuities.
+
+However, the advancement in health science since the early years of the 
+twentieth century, the relative peace in the western world since the second 
+world war and abundance of food in the rich world resulted in a change in
+historical mortality patterns. Pension funds started realizing that annuities
+were being drawn for longer than expected periods. As the population in the
+developed countries started greying the pension funds had the additional 
+problem of a higher annunity obligations and a lower contribution from newer
+workers. Pension funds now have a keen interest in understanding how mortality 
+rates will evolve in the future. They now want to forecast mortality than just
+produce models of the current mortality.
+
+Demographic forecasting predates mortality forecasting. Economists like Thomas
+Malthus predicted that the human population will grow to such an extent that
+there will not be enough to feed everyone. This dire prediction meant that the
+politicians and the economists had a keen desire to predict how populations
+will grow and whether there will be enough food for everyone. As a result, many 
+statistical methods developed in Demography were readily available for mortality
+forecasting. This need for forecasting was also coeval with the developments
+in the modern theory of mathematical statistics and the foundations of
+probability. It is not a coincidence that the early mortality models were
+mathematical functions while the later ones are stochastic models.
+
+Mortality models, being mathematical functions, produce a single curve for
+a given set of its parameters. On the other hand, mortality forecasts are a
+collection of curves or a family of probablity distributions. Their output
+is accompanied with appropriate confidence intervals to indicate the 
+uncertainty in their reported values.
+
 In this article we shall review the development mathematical models to study
-mortality in large enough populations of people.
+and forecast mortality in large enough populations of people.
 
@@
 \begin{equation}\label{s4e3}
-s(x) = 1 - ax^{1/4} _ bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
+s(x) = 1 - ax^{1/4} + bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
 \end{equation}
@@
 
+\section{Lee-Carter model}\label{s6}
+The problem of forecasting population changes and composition is of interest
+to both the demographers and the pension fund managers. Ronald Lee and 
+Lawrence Carter's motivation for developing a model to forecast mortality was
+the observation that the human life expectancy rose from $47$ years to $75$ 
+years in the United States in the $88$ years starting from the year $1900$.
+Should this trend continue unabated in the future, the human life expectancy
+will touch $100$ years by the year $2065$ upsetting the Social Security 
+Administration's calculations based on the long-term life expectancy of about
+$80.5$ years. The wide difference between Lee and Carter's 
+\cite{lee1992modeling} forecasts and the assumptions of the pension funds 
+made the actuaries turn their attention to forecasting than modelling alone.
+
+Unlike the previous models which tried to build formulae that mimicked the
+behaviour of human mortality in various stages of life, Lee and Carter used
+the statistical techniques of time series for their projection into the future.
+They excluded factors representing advancement in medicine or the behavioural
+and social changes on mortality. Their time series model extrapolated the 
+long-term trends seen in demographic data over several decades. They did not
+find it necessary to assume that the human life cannot extend beyond a certain
+limit. Nothing analogous to the $\omega$ in de Moivre's model would be found
+in Lee and Carter's.
+
+The mathematical form of Lee and Carter's model is
+\begin{equation}\label{s6e1}
+\log{M}(x, t) = {a}_x + {b}_x{k}_t + \epsilon_{x, t},
+\end{equation}
+where the coefficients ${a}_x$ and ${b}_x$ depend on the age, the
+`index' ${k}_t$ depends on time and ${M}(x, t)$ is the mean death
+rate for age $x$ in time $t$. Since
+\[
+\frac{d}{dt}{M}(x, t) = {b}_x\frac{dk_t}{dt},
+\]
+${b}_x$ captures the the changes in mortality rates due to changes in
+the index with time. A positive value of a component of ${b}_x$ indicates
+falling mortality while a negative value indicates the opposite. The term
+$\epsilon_{x, t}$ is an `error term' with mean $0$ and variance that depends
+on the age. It captures the factors that influence mortality but are not
+explicitly included in the model.
+
+The model of equation \eqref{s6e1} does not have a unique solution. If 
+${a}$, ${b}$ and ${k}$ form one solution then for any number $c$,
+the triple ${a} - c{b}, {b}, {k} + c$ is also a solution. In
+fact, even ${a}, c{b}, {k}/c$ is a solution if $c \ne 0$. This phenomenon is
+also called `identifiability problem'\cite{plat2009stochastic}.
+
+Fitting the Lee-Carter model consists in finding the parameters ${a}_x,
+{b}_x$ and ${k}_t$ for a given matrix of mortality rates ${M}(x, t)$.
+The model is solved using the singular value decomposition (SVD). The algorithm
+is described in appendix A to Lee and Carter's paper\cite{lee1992modeling} 
+introducing the model.
+
+Lee and Carter's forecast of life expectancy for the American population was 
+$86.05$ by the year $2065$. The Actuary of the Social Security Administration,
+on the other hand, forecast it as $80.5$. In $2020$ the observed value is 
+$78.87$, suggesting that Lee and Carter's forecast might be closer to the 
+reality in the year $2065$.
+
+The advantages of Lee-Carter model lie in its robustness, simplicity and 
+the fact that it fits well over wide ranges of age. It is also observed to be
+able to fit small non-linearities in the mortality curve quite well. Its 
+shortcomings are its lack of smoothness of age effects in small populations, 
+inability to work with cohorts, inability to cope with improvements varying 
+across times and a possible underestimation of uncertainty. It also has a 
+limited versatality owing to it being a one-factor model.
+
+\section{Age-Period-Cohort models}\label{s7}
+Demographic data is often analysed by taking into account the age, period 
+and the cohort of individuals. The analysis need not be for mortality alone.
+It could be for incidence of diseases or even economic consumption. Consumption
+patterns do depend on age of the individual. But when they are observed over
+a long enough duration they are also found to depend on the period and the 
+cohort. For instance, certain goods may not have been available or would have
+been expensive in a certain period and therefore we not consumed by many. It
+is also well-known that different age groups of people have different tastes.
+The age-period-cohort (APC) models are a class of additive models where the
+predictor is a sum of three time effects, age, period and cohort. APC models
+are thus able to take into account the impact of the changes in the society 
+and the available technology.
+
+Period analysis alone is quite useful to understanding the effect of wars,
+pandemics, revolutions or other upheavals on a population. However, they
+capture the effect of the times in which people are living. Behavioural
+changes in people are better captured by considering cohorts. The 
+characteristics of a cohort are an aggregate of the characteristics of its 
+members\cite{willekens1991age}. Modern APC models require longitudnal data
+to understand the life processes as they evolve. They do not consider just
+the event of the death of an individual but the process of dying whose 
+culmination is death. They take the view that the mortality risks in a 
+population are better determined by an understanding of morbidity in it
+\cite{van1990determinants}.
+
+The APC framework is used widely in the social sciences. In the context of
+mortality modelling and forecasting it started being used while answering the
+questions \cite{willekens1991age}:
+\begin{itemize}
+\item Does mortality show a greater regularity when viewed at the level of a
+cohort or for a particular period? The answer to this question is important
+for mortality forecasting.
+
+Although a full-fledged APC framework arose relatively recently, the importance
+of cohort analysis was not missed by early demographers and actuaries. Derrick
+\cite{derrick1927observations} observed in 1927 that the ratio of mortality
+for one cohort to that of another was constant for all ages above $10$ years.
+It suggested that it may be possible to forecast the mortality of an individual
+based on the cohort in which he falls. It was also observed that the mortality
+rates for the generation that was young during the years of the second world
+war were overestimated for obvious reasons. This once again indicates the 
+importance and the role of cohorts in modelling mortality.
+
+\item Do events early in life affect the experiences later in life? This 
+question can be answered only in the context of a cohort analysis.
+
+While studing the death rates in England, Scotland and Sweden, Kermack
+\cite{kermack1934death} observed that the differences in mortality was mostly
+because of the environmental conditions experienced in the first fifteen years
+of life. It was not a result of the other influences in later life. The 
+study also uncovered the fact that an improvement in the conditions around
+child-birth boosted survival rates of new-borns. It proposed that the health
+of the mother has a positive impact on the health of the child and its chances
+to live longer. Preston and de Walle \cite{preston1978urban} came to similar
+conclusion for the French mortality data. This idea was further supported by 
+Caselli and Capocaccia's study\cite{caselli1989age} on Italian demographics 
+that infant mortality had a weakening effect on eventual mortality. However, 
+this observation is not universally accepted. Other studies indicate that a 
+high infantile mortality results in lower eventual mortality because the 
+`stronger' individuals stay long enough to live long enough 
+\cite{manton1981methods}. Lest one concludes that only childhood environment
+determines mortality rate later in life, Horiuchi reported 
+\cite{horiuchi1983long} that the malnutrition and physical privation during the
+second world war wreaked the health of those who lived through it and affected
+their mortality when they became old.
+\end{itemize}
+
+APC studies are plagued by what is called the `identification problem' common
+to all cohort analysis. It is the problem of assigning an individual to the
+right cohort when the a) the cohorts are adjacent in time, b) the individual 
+is on the border line of two adjacent cohorts and c) the characteristics of
+the cohorts are quite distinct. The problem can be mitigated by following
+the directions suggested in Willekens and Scherbov's \cite{willekens1991age}
+analysis.
+
+APC model can be expressed as a generalized linear model in which we predict
+the mortality rate from age, period and cohort as predictors. We usually use
+the variables in the data that are the most appropriate proxy for age, period
+and cohort. Once the model is build from the available data, we can use it to 
+forecast mortality for a newer set of predictors. It is assumed that the 
+mortality rate can be described as a Poisson process. This assumption is found 
+to be valid for societies with low death rates. It may not hold good when 
+the society is experiencing a war or a pandemic.
+
+\section{Cairns-Blake-Dowd model}\label{s8}
+Cairns, Blake and Dowd proposed a model \cite{cairns2006two} to study the
+evolution of mortality beyond sixty years of age and its impact on longevity
+risk. It was a two factor stochastic model. Of these, the first factor affects
+the mortality rate at all ages in an identical manner and the second one 
+affects the advanced years to a greater extent. Their mortality curve is
+defined in terms of forward survival probability $\pr(t, T_0, T_1, x)$. It is
+the probability measured at $t$ that an individual aged $x$ at time $0$,
+alive at $T_0$ will survive until time $T_1 > T_0$. The curve is represented
+by the equation
+\begin{equation}\label{s8e1}
+1 - p(t+1,t,t+1,x) = 
+\frac{e^{A_1(t+1)+A_2(t+1)(x+t)}}{1+e^{A_1(t+1)+A_2(t+1)(x+t)}},
+\end{equation}
+where $A_1(u)$ and $A_2(u)$ are stochastic processes. If $q(p, T_0, T_1, x)
+= 1 - p(t, T_0, T_1, x)$ then we can readily write equation \eqref{s8e1} in
+the form
+\begin{equation}\label{s8e2}
+\log\left(\frac{q(t,T_0,T_1,x)}{p(t,T_0,T_1,x)}\right) = 
+A_1(t+1) + A_2(t+1)(x+t).
+\end{equation}
+The log-odds of probability of death is thus a linear function of age and the
+coefficients of this relation vary with time. The model uses data from 
+different cohorts at a common point in time. Therefore, the coefficients 
+$A_1(t)$ and $A_2(t)$ are interpreted as factors. The original model of Cairns,
+Blake and Dowd used ordinary least square regression to estimate the factors.
+The authors plotted the trend of the factors for the period between $1961$ and
+$2002$. The factor $A_1(t)$ decreases with time indicating an improvement in
+the mortality. On the other hand, $A_2(t)$ is observed to increase with time.
+This suggests that the improvements in mortality are felt to a lesser extent 
+for older people. In order to forecast the factors at times in future, the 
+authors model $A(t) = (A_1(t), A_2(t))^T$ as a random walk with a drift. That
+is,
+\begin{equation}\label{s8e3}
+A(t+1) - A(t) = \mu + CZ(t+1),
+\end{equation}
+where $\mu$ is a constant vector representing the mean increment, $C$ is a 
+constant $2 \times 2$ upper triangular matrix and $Z(t)$ is a standardised
+two dimensional gaussian random variable. The authors also propose a criterion
+for a good model of this kind. They introduce the idea of `biological 
+reasonability'. In general, a biologically reasonable model will have the 
+first component of $\mu$ negative and the second component positive. Further,
+the mortality rates for older cohorts should normally be greater than the 
+younger cohorts.
+
+Practitioners favour Cairns-Blake-Dowd model for its robustness and the ease
+with which it allows inclusion of parameter uncertainty. It is also more 
+versatile than the Lee-Carter model because it has an additional factor. 
+However, like the Lee-Carter model it does not allow analysis of cohort effects.
+Although it models a wide range of advanced years, it does not fit the 
+mortality data as well as Lee-Carter model.
+
+\section{Stochastic cohort models}\label{s9}
+The stochastic models of Lee-Carter and Cairns-Blake-Dowd were unable to
+incorporate cohort effects. To remedy this shortcoming, Renshaw and Haberman
+\cite{renshaw2006cohort} proposed a modification of the Lee-Carter model that 
+includes cohorts. It can be expressed mathematically as
+\begin{equation}\label{s9e1}
+\log{M}(x, t) = {a}_x + {b}_x{k}_t + {c}_x\gamma_{t-x},
+\end{equation}
+where the symbols have the same meaning as in Lee-Carter model. In the new
+term, third on the right hand side, $\gamma_{t-x}$ captures the cohort effect.
+It is a function of $t - x$, the (approximate) year of birth and ${c}_x$
+is just the regression coefficient of this term. Although the model fitted the
+male mortality data in England and Wales quite well it turned to not as 
+robust as other models. The lack of robustness is believed to be related to
+the shape of the likelihood function. It has more than one local maximum.
+Therefore, when we change the age or the year range the optimizer ends up
+settling on a different maximum each time\cite{cairns2008modelling}. Currie
+simplified the Renshaw-Haberman model to
+\begin{equation}\label{s9e2}
+\log{M}(x, t) = {a}_x + {k}_t + \gamma_{t-x}.
+\end{equation}
+The simplification solved the problem of robustness but made the fit poorer
+\cite{plat2009stochastic}. Plat introduced a model to address the deficiencies
+of all the previous models while retaining their benefits 
+\cite{plat2009stochastic}. Its mathematical structure is
+\begin{equation}\label{s9e3}
+\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + 
+\kappa^3_t(\bar{x} - x)^+ + \gamma_{t-x},
+\end{equation}
+where $(\bar{x} - x)^+ = \max(\bar{x} - x, 0)$ and $\bar{x}$ is the mean age
+in the sample range. The term $a_x$ plays a 
+similar role as it did in Lee-Carter model. It controls the general shape
+of the mortality curvve. Of the four stochastic factors, $\kappa_t^1$
+represents the change in the level of mortality across all ages, $\kappa_t^2$
+represents the change in the level of mortality between ages, $\kappa_t^3$
+captures the differences in mortality behaviour in the early adulthood stage
+and $\gamma_{t-x}$ captures the cohort effect. If one is interested in 
+modelling only the advanced ages then one need not worry about the `accident
+hump' of young adulthood and use the simplified model,
+\begin{equation}\label{s9e4}
+\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + \gamma_{t-x},
+\end{equation}
+Like the Lee-Carter model, this one too suffers from the `identifiability
+problem'. For constants $c_1, c_2, d$, the set of parameters
+\begin{eqnarray*}
+\bar{\gamma}_{t-x} &=& \gamma_{t-x} + c_1 + c_2(t - x) \\
+\bar{\kappa}_t^1 &=& \kappa^1_t + c_1 - d\bar{x}c_2 \\
+\bar{\kappa}_t^2 &=& \kappa^2_t + dc_2 \\
+\bar{a}_x &=& a_x + (1 - d)c_2
+\end{eqnarray*}
+also lead the the same values for $\log M(x, t)$. The problem is mitigated 
+by requiring that 
+\begin{eqnarray*}
+\sum_{i=x_0}^{x_1} \gamma_i &=& 0 \\
+\sum_{i=x_0}^{x_1} i\gamma_i &=& 0 \\
+\sum_t\kappa_t^3 &=& 0,
+\end{eqnarray*}
+where $x_0$ and $x_1$ are the earliest and the latest years of birth to which
+the cohort effect is fitted. 
 \bibliography{draft}

### 2020-09-08 mortality-models/draft.tex
Some more updates.

@@
 \DeclareMathOperator{\pr}{P}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-
+\linespread{1.3}
 \title{A Review of Mortality Models}
@@
 
-In this article we shall review the development mathematical models to study
-and forecast mortality in large enough populations of people.
+Before we start reviewing the deterministic and stochastic mortality models
+let us describe certain characteristic features of human mortality. Although
+the life expectancy at birth has significantly improved in the post second
+world war years the maximum life span seems to be pretty constant throughout 
+history. A new born child faces a great risk of dying as it adjusts itself
+to its new environment, warding off pathogens and fighting diseases. As time 
+goes by the child starts developing immunity against many germs and its
+chances of dying begin to drop. This drop continues until early adulthood 
+when there is a sudden jump in mortality. It happens because of aggressive
+behaviour in the late teens in both sexes and because of maternal mortality
+in females. The spike in mortality eventually dwindles into a monotonic
+increase in mortality as senescence catches up. We will see that there is 
+a rich collection of mathematical models describing these characteristics
+of human mortality. The basic structure of the mortality behaviour also
+suggests that it might be profitable to fit a different model for different
+stages of human life. The earliest models indeed did not cover childhood or
+take into account the spike in mortality starting late teens. They were 
+simple mathematical functions that fit a portion of the mortality curve that
+was of most relevance to the insurance companies. Later models continued to
+be deterministic in nature but had more number of parameters that could be
+fitted for a given empirical data. However, it was not always possible to 
+ascribe an unambiguous biological meaning to all parameters. Further, their
+complexity was also controlled by the limitations of the computing technology
+of their period.
+
+We will sketch the historical development of mortality model, both 
+deterministic and stochastic, starting from de Moivre's model in the next
+section. We also define basic terms commonly used in actuarial science and 
+demonstrate how to evaluate their expressions in simple cases. In section 3
+we describe the Gompertz's model, the first realistic mortality model that 
+was used for a very long time in the industry. The success of Gompertz model
+lay in the number of successors it produced. Of these, we cover Makekam's 
+model and its generalisations in section 4. In section 5 we mention a few
+other mortality models used in European countries but whose significance today
+lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
+was the first one that could be fitted to the entire mortality curve, starting
+from infancy to very advanced years. We describe it in section 6. Starting
+section 7 we turn our attention to stochastic models, starting with the
+celebrated Lee-Carter model. An interesting class of models used across social
+sciences is the Age-Period-Cohort models. We review a few of them relevant
+to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
+and finally we consider stochastic models with cohort effects in section 10.
 

### 2020-09-09 mortality-models/draft.tex
WIP 9-Sep-2020

@@
 \section{Introduction}\label{s0}
-Although it is very difficult to predict the behaviour of a member of a large
+Although it is very difficult to predict the behavior of a member of a large
 population one can often make statistical statements about the same population 
@@
 were being drawn for longer than expected periods. As the population in the
-developed countries started greying the pension funds had the additional 
-problem of a higher annunity obligations and a lower contribution from newer
+developed countries started graying the pension funds had the additional 
+problem of a higher annuity obligations and a lower contribution from newer
 workers. Pension funds now have a keen interest in understanding how mortality 
@@
 a given set of its parameters. On the other hand, mortality forecasts are a
-collection of curves or a family of probablity distributions. Their output
+collection of curves or a family of probability distributions. Their output
 is accompanied with appropriate confidence intervals to indicate the 
@@
 when there is a sudden jump in mortality. It happens because of aggressive
-behaviour in the late teens in both sexes and because of maternal mortality
+behavior in the late teens in both sexes and because of maternal mortality
 in females. The spike in mortality eventually dwindles into a monotonic
@@
 a rich collection of mathematical models describing these characteristics
-of human mortality. The basic structure of the mortality behaviour also
+of human mortality. The basic structure of the mortality behavior also
 suggests that it might be profitable to fit a different model for different
@@
 lay in the number of successors it produced. Of these, we cover Makekam's 
-model and its generalisations in section 4. In section 5 we mention a few
+model and its generalizations in section 4. In section 5 we mention a few
 other mortality models used in European countries but whose significance today
@@
 company would need to know what proportion of its customers will survive
-past various points in the future. It to this problem that de Moivre 
-\cite{de1731annuities} addresses his treatise. In order to estimate the value
+past various points in time in the future. It to this problem that de Moivre 
+\cite{de1731annuities} addressed his treatise. In order to estimate the value
 of annuities one must consider the interest earned by the company's assets and
@@
 human mortality. He emphasized in his treatise that it was an approximation
-useful to calculate the annuities. His piecewise linear approximation was a
+useful to calculate the annuities. His piece-wise linear approximation was a
 valuable device to carry out practical computations in the eighteenth century.
@@
 equation \eqref{s2e5} are determined from the mortality data of the population
-under study. Gompertz suggested that his formula is applicable for the age
+under study. It is common to introduce a constant $c = e^k$ and write equation
+\eqref{s2e5} as
+\begin{equation}\label{s2e6}
+s(x) = C\exp\left(\frac{B}{k}c^x\right),
+\end{equation}
+Gompertz suggested that his formula is applicable for the age
 group between $10-15$ years of age up to $55-60$. In order to extend it beyond
@@
 while the second factor is constant throughout. However, in the formulation
-of his law of mortality, Gompertz considered only the first factor.
+of his law of mortality, Gompertz considered only the first factor. Gompertz
+could have fitted simpler polynomial functions to the mortality data available
+in his times. However, instead of trying to fit a function to the data he
+developed his model based on an observation root in biology that organisms
+deteriorate with age in a geometric proportion. Gompertz model was not only a
+better fit to data than de Moivre's but it also started the tradition of
+finding a biological explanation for every term in a mortality model. It is
+important to note that Gompertz considered death as just a biological 
+phenomenon. His law is best applicable to populations well protected from 
+wars or pestilence. It is, therefore, an upper limit to longevity in a
+population. In that sense, it served to be a useful estimate of the insurance
+companies' liabilities. The success of Gompertz's law in practical calculations
+ensured that for several decades newer mortality laws were only an improvement
+over it. In the next section, we shall consider a few of them.
 
-\section{Makeham's law and its generalisations}\label{s3}
+\section{Makeham's law and its generalizations}\label{s3}
 William Makeham, the English mathematician and actuary, added a term to
@@
 Makeham's survival function has three constants $A, B$ and $C$ and therefore
-can be fitted to emprically observed mortality data more effectively than
-Gompertz's model. Makeham's law was often applied to mortality tables in the
+can be fitted to empirically observed mortality data more effectively than
+Gompertz's model. If we write $c = e^k$ then we can write equation \eqref{s3e2}
+in its customary form
+\[
+s(x) = C\exp\left(Ax + \frac{B}{k}c^x\right).
+\]
+Makeham's law was often applied to mortality tables in the
 age range beyond $20$ years to almost end of life\cite{jordan1967society}. 
@@
 increases monotonically with time. However, it is commonly observed that 
-infant mortality contradicts this assumption. In order to extend Makeham's
-law of 1860 to infantile ages, Wilhem Lazarus \cite{lazarus1867ueber} added
-an exponentially decreasing term. His law expressed the force of mortality
-as 
+infant mortality contradicts this assumption. This is the reason why Gompertz's
+and Makeham's laws were applied only to populations over $20$ years of age.
+
+Mortality decreases from the birth of the child until the onset of adulthood. 
+In order to extend Makeham's law of 1860 to infantile ages, Wilhelm Lazarus 
+\cite{lazarus1867ueber} added an exponentially decreasing term. His law 
+expressed the force of mortality as 
 \begin{equation}\label{s3e5}
@@
 The way infants' susceptibility to die of diseases upsets Gompertz's and 
-Makeham's assumption of monotonicity of the law of mortality, do does the 
+Makeham's assumption of monotonicity of the law of mortality, so does the 
 spike in the death rate of young adults. Wars were frequent in Europe of the 
 eighteenth century and the victims were disproportionately the young recruits.
-Even otherwise, young adults are more prone to death by accidents. In order
+Even otherwise, young adults are more prone to death by accidents. Maternal
+mortality too was quite high in the eighteenth century. In order
 to take into account this mortality hump in the young, Thiele 
-\cite{thiele1871mathematical} introduced a gaussian term to Lazarus' model.
+\cite{thiele1871mathematical} introduced a Gaussian term to Lazarus' model.
 His law of mortality is
@@
 All the constants in this equation, namely $A, a_1, a_2, B_1, k_1, B_2$ and 
-$k_2$ are positive numbers. The presence of the gaussian term prevents us
+$k_2$ are positive numbers. The presence of the Gaussian term prevents us
 from writing a closed form expression of the survival function. However, the
@@
 
-We observe that all generalisations of Makeham's laws introduce additional
+We observe that all generalizations of Makeham's laws introduce additional
 constants to the force of mortality and the hazard function. These constants
@@
 
-Forfar, McCutcheon and Wilkie proposed a far wider generalisation of the
-Gompertz and Makeham's laws. They defined a class of models, called GM models
-after their inventors, with the mathematical form
+Forfar, McCutcheon and Wilkie \cite{forfar1988graduation} proposed a far wider 
+generalization of the Gompertz and Makeham's laws. They defined a class of 
+models, called GM models after their inventors, with the mathematical form
 \begin{equation}\label{s3e8}
@@
 
-\section{Other mortality laws prior to the 20th century}\label{s4}
+Gompertz's model and its descendants considered in this section had simple
+algebraic forms, each term of which could be interpreted in biological
+terms. However, they were not always easy to fit on available mortality
+data. The presence of exponentials also made them numerically unstable. 
+Moreover, they covered only a limited number of sources of mortality.
+
+\section{Other deterministic mortality laws}\label{s4}
 de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
@@
 \end{equation}
+The original mortality laws of Gompertz and Makeham were applicable only
+for adults. Oppermann proposed \cite{oppermann1870graduation} a law for people 
+younger than $20$ years of age. It was
+\begin{equation}\label{s4e4}
+\mu(x) = \frac{a}{\sqrt{x}} + b + cx^{1/3},
+\end{equation}
+where $a, b, c$ are parameter selected to fit the data. Perks 
+\cite{perks1932some} proposed two models of the form
+\begin{eqnarray}
+\mu_1(x) &=& \frac{A + Bc^x}{1 + Dc^x} \label{s4e5} \\
+\mu_2(x) &=& \frac{A + Bc^x}{Kc^{-x} + 1 + Dc^x} \label{s4e6}
+\end{eqnarray}
+The denominators in these forms flatten the exponential increase in Gompertz
+form, especially at advanced age. The mortality law of equation \eqref{s4e5}
+was used to graduate table of annuitant lives in the United Kingdom. Beard
+\cite{beard1971some} wrote his law in terms of the probability that a life 
+aged $x$ will die in the next year. It was
+\begin{equation}\label{s4e7}
+q_x = A + \frac{Bc^x}{Ex^{-2x} + 1 + Dc^x},
+\end{equation}
+where $A, B, c, D, E$ are parameters selected to fit the mortality data.
+Barnett's formula appears on page 141 of \cite{joint1974considerations}. It
+fits data using four parameters and has the mathematical form
+\begin{equation}\label{s4e8}
+q_x = \frac{A - Hx - Bc^x}{1 + A - Hx + Bc^x}.
+\end{equation}
+Wilkie \cite{wilkie1976international} generalized this form to
+\begin{equation}\label{s4e9}
+q_x = \frac{f(x)}{1 + f(x)},
+\end{equation}
+where
+\begin{equation}\label{s4e10}
+f(x) = \exp\left(\sum_{i=1}^s\alpha_i x^{i-1}\right).
+\end{equation}
 All these models were attempts to fit the mortality data with a continuous 
-function. They were not rooted in demographic observations like the models
-of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
-disuse after Gompertz proposed his model and others improved upon it.
+function. Many were increasingly complicated functions that were selected on
+the sole merit that they fitted the observed data well. They were not rooted in 
+demographic observations like the models of Gompertz, Makeham, Lazarus or 
+Thiele. As a result, they soon fell into disuse after stochastic models came on
+the scene.
 
@@
 dealing with an experiment with two outcomes - that something happens or 
-it does not - then it is common in certain fields to express the 
-probabilities in terms of odds.  For example, if $p$ is the probability 
-that an event will happen and $q = 1 - p$ is the probability that it will 
-not happen then the odds of the experiment is $p/q$.
+it does not - then it is common to express the probabilities in terms of odds. 
+For example, if $p$ is the probability that an event will happen and $q = 1 - p$
+is the probability that it will not happen then the odds of the experiment is 
+$p/q$. The usage of odds instead of probabilities is very common in gambling or
+betting.
 
@@
 higher for a given value of $C$. $B$ is thus an indicator of the central 
-tendancy of infant mortality.
+tendency of infant mortality.
 
-The second term is gaussian in the logarithm of age. It represent the accident
+The second term is Gaussian in the logarithm of age. It represent the accident
 mortality for young adults of both sexes and maternal mortality in young 
@@
 
+It is possible to write Heligman-Pollard model in the form
+\begin{equation}\label{s5e2}
+q_x = \frac{f(x)}{1 + f(x)},
+\end{equation}
+where
+\begin{equation}\label{s5e3}
+f(x) = A^{(x+B)^C} + D\exp\left(-E\left(\log\frac{x}{F}\right)^2\right) + 
+GH^x.
+\end{equation}
+A variation of the function $f$ in to
+\begin{equation}\label{s5e4}
+f(x) = A^{(x+B)^C} + D\exp\left(-E\left(\log\frac{x}{F}\right)^2\right) + 
+\frac{GH^x}{1 + GH^x}
+\end{equation}
+leads to the second form of Heligman-Pollard model. The two functions defined
+in equations \eqref{s5e2} and \eqref{s5e3} are not too dissimilar for most
+part of human mortality values. Two more variations of the Heligman-Pollard
+model are available in literature\cite{heligman1980age}.
+
+The advantages of Heligman-Pollard model \cite{umar2019modeling} lie in its
+smoothness, interpolability, parsimony, analytic manipulability and in 
+easy comparison with other models. Like all deterministic models, it models
+mortality at a fixed point time for all ages. As a result it used to construct
+age-period tables. Its shortcomings are because of the numerical instability
+in the algorithms used to find the parameters. It also has a wide range of
+parameters all producing a similar fit.
+
 \section{Lee-Carter model}\label{s6}
@@
 Administration's calculations based on the long-term life expectancy of about
-$80.5$ years. The wide difference between Lee and Carter's 
-\cite{lee1992modeling} forecasts and the assumptions of the pension funds 
-made the actuaries turn their attention to forecasting than modelling alone.
+$80.5$ years. The wide difference between the predictions of the linear 
+increase in life expectancy 
+\cite{lee1992modeling} and the assumptions of the pension funds 
+made the actuaries turn their attention to forecasting than modeling alone.
 
 Unlike the previous models which tried to build formulae that mimicked the
-behaviour of human mortality in various stages of life, Lee and Carter used
+behavior of human mortality in various stages of life, Lee and Carter used
 the statistical techniques of time series for their projection into the future.
-They excluded factors representing advancement in medicine or the behavioural
+They excluded factors representing advancement in medicine or the behavioral
 and social changes on mortality. Their time series model extrapolated the 
@@
 across times and a possible underestimation of uncertainty. It also has a 
-limited versatality owing to it being a one-factor model.
+limited versatility owing to it being a one-factor model.
 
 \section{Age-Period-Cohort models}\label{s7}
-Demographic data is often analysed by taking into account the age, period 
+Demographic data is often analyzed by taking into account the age, period 
 and the cohort of individuals. The analysis need not be for mortality alone.
@@
 pandemics, revolutions or other upheavals on a population. However, they
-capture the effect of the times in which people are living. Behavioural
+capture the effect of the times in which people are living. Behavioral
 changes in people are better captured by considering cohorts. The 
 characteristics of a cohort are an aggregate of the characteristics of its 
-members\cite{willekens1991age}. Modern APC models require longitudnal data
+members\cite{willekens1991age}. Modern APC models require longitudinal data
 to understand the life processes as they evolve. They do not consider just
@@
 The APC framework is used widely in the social sciences. In the context of
-mortality modelling and forecasting it started being used while answering the
+mortality modeling and forecasting it started being used while answering the
 questions \cite{willekens1991age}:
@@
 war were overestimated for obvious reasons. This once again indicates the 
-importance and the role of cohorts in modelling mortality.
+importance and the role of cohorts in modeling mortality.
 
@@
 
-While studing the death rates in England, Scotland and Sweden, Kermack
+While studying the death rates in England, Scotland and Sweden, Kermack
 \cite{kermack1934death} observed that the differences in mortality was mostly
@@
 where $\mu$ is a constant vector representing the mean increment, $C$ is a 
-constant $2 \times 2$ upper triangular matrix and $Z(t)$ is a standardised
-two dimensional gaussian random variable. The authors also propose a criterion
+constant $2 \times 2$ upper triangular matrix and $Z(t)$ is a standardized
+two dimensional Gaussian random variable. The authors also propose a criterion
 for a good model of this kind. They introduce the idea of `biological 
@@
 
-Practitioners favour Cairns-Blake-Dowd model for its robustness and the ease
+Practitioners favor Cairns-Blake-Dowd model for its robustness and the ease
 with which it allows inclusion of parameter uncertainty. It is also more 
@@
 similar role as it did in Lee-Carter model. It controls the general shape
-of the mortality curvve. Of the four stochastic factors, $\kappa_t^1$
+of the mortality curve. Of the four stochastic factors, $\kappa_t^1$
 represents the change in the level of mortality across all ages, $\kappa_t^2$
 represents the change in the level of mortality between ages, $\kappa_t^3$
-captures the differences in mortality behaviour in the early adulthood stage
+captures the differences in mortality behavior in the early adulthood stage
 and $\gamma_{t-x}$ captures the cohort effect. If one is interested in 
-modelling only the advanced ages then one need not worry about the `accident
+modeling only the advanced ages then one need not worry about the `accident
 hump' of young adulthood and use the simplified model,
@@
 the cohort effect is fitted.
+
+\section{Conclusion}
+We traced the development of mortality models from the earliest attempts of
+Graunt and Halley in the seventeenth century to the latest stochastic models
+of the twentieth century. In the four centuries spanning these models, we
+saw how the models addressed the business needs of their times with the
+mathematical methods available to their inventors. The earliest models either
+predate or are coeval to the development of Calculus. They were built on the
+most rudimentary data. Although their form was crude, their aim was very clear.
+They aspired to predict the liabilities of insurance companies. As mathematical
+ideas started to get sophisticated, so did the mortality models. By the
+eighteenth century, the concept of function was understood my most 
+mathematicians at an intuitive level. The idea of probability was still not
+within the ambit of mainstream mathematics. As a result, mortality models of
+those days were written as functions of human age. The key characteristics
+of human mortality did not change drastically enough to warrant anything more
+than a more flexible functional form. However, from the closing decades of the
+nineteenth century, advances in health sciences and developments of vaccines 
+started to prevent many deaths. The developed world also saw a long lull in 
+large scale conflict after the second world war got over. The triumph of
+medical sciences became a cause of worry to the actuaries. They spotted the
+emergence of significant longevity risk to their businesses. They noticed the
+rapid rise in human life expectancy leading to more people drawing pensions
+for more years. This is when the need to forecast mortality arose. The 
+foundations of probability theory were laid in 1930s and stochastic processes
+were put on a firm mathematical footing by the 1950s. Actuaries and 
+statisticians took advantage of these developments to build stochastic models
+to forecast the future course of mortality curves. Their results were not
+in the form of a deterministic curve but a probability distribution with
+well-defined confidence intervals. The availability of cheap computing also
+spurred a growth in this area. Stochastic models began to include cohort 
+analysis. With the advent of newer forms of finance and sophisticated
+financial engineering, mortality models began to be coupled with design of
+financial instruments to hedge risks. This is indeed a very long way since the
+publication of 'Bills of Mortality' every Thursday in the parishes of London.
 \bibliography{draft}

### 2020-09-13 mortality-models/draft.tex
WIP 13Sep20

@@
 \usepackage{amsmath, amssymb, amsthm, amsfonts}
+\usepackage{url}
 \numberwithin{equation}{section}
 \DeclareMathOperator{\pr}{P}
+\DeclareMathOperator{\E}{E}
 \linespread{1.3}
@@
 Therefore, when we change the age or the year range the optimizer ends up
-settling on a different maximum each time\cite{cairns2008modelling}. Currie
+settling on a different extremum each time\cite{cairns2008modelling}. Currie
 simplified the Renshaw-Haberman model to
@@
 
+The Plat model is a synthesis of several stochastic mortality models aimed to
+retain their advantages and eliminating their shortcomings. Its four stochastic
+factors ensure that model is not trivial and yet it is not too complex. It
+retains the correlation structure between age groups, making it a valuable
+tool for solvency calculations. The model is applicable over the full range of 
+human mortality. It captures the cohort effect and has a robust numerical 
+behavior. Another valuable feature of the model is that one can develop its
+risk-neutral version to price longevity derivatives.
+
+\section{Cause-of-Death models}\label{s11}
+All mortality models considered so far have considered only the occurrence of
+death without examining the reason. Unlike most of the animal world, death in
+humans is not solely due to natural causes. The cause-of-death models split the
+aggregate mortality data into a number of causes and model mortality for each
+cause separately. It as important to understand the key drivers of human 
+mortality as the aggregate mortality rate. It has been observed that deaths
+due to cardiovascular diseases have reduced in past few decades. Death rate
+due to neoplasm have also begun to drop. However, deaths due to dementia in 
+advanced ages and drug-abuse in the younger ages have risen. Features like this
+do not emerge by modeling the aggregate death rate. Boumezoued and others
+\cite{alexandre2019modeling} used Human Mortality Database and Human Cause-of-
+death database to construct a cause-of-death mortality model for the American
+population. They define $\mu_{x,t,i}$, the death rate for an age group $x$ due
+to cause $i$ in year $t$ as
+\begin{equation}\label{s11e1}
+\mu_{x,t,i} = \frac{D_{x,t,i}}{E_{x,t}},
+\end{equation}
+where $D_{x,t,i}$ is the number of deaths by cause $i$ in year $t$ of an 
+individual aged $x$ last birthday. $E_{x,t}$ is the total life duration in the
+year $t$ of individuals ages $x$ last birthday. It is also called the 
+`exposure-to-risk'. Note that the exposure is not indexed by $i$ and is an
+aggregate metric. The authors use the competing risks framework according to
+which every cause of death has a cause-specific lifetime associated with  it.
+Thus if $\tau^A$ is the lifetime under cause $A$ and $\tau^B$ is the lifetime
+under all causes except $A$ then the survival function at age $a$ is defined
+as the probability that lifetimes under all clauses will be higher than $a$,
+Thus,
+\begin{equation}\label{s11e2}
+S(a) = \pr(\tau > a) = \pr(\tau^A>a,\tau^B>a) = 
+\exp\left(-\int_0^a\mu(y)dt\right),
+\end{equation}
+where
+\begin{equation}\label{s11e3}
+\tau = \min(\tau^A, \tau^B)
+\end{equation}
+If the lifetimes under cause $A$ is independent of the lifetime under all causes
+then one can write the survival function as
+\begin{equation}\label{s11e4}
+S(a) = \pr(\tau^A>a)\pr(\tau^B>a)=
+\exp\left(-\int_0^a\mu_A(y)dt\right)\exp\left(-\int_0^a\mu_B(y)dt\right),
+\end{equation}
+In equations \eqref{s11e2} and \eqref{s11e4}, $\mu$ is the aggregate mortality
+rate and $\mu_A$ and $\mu_B$ are cause-specific mortality rates. There are 
+several ways to build cause of death forecasts. One can associate a link
+between the causes and a number of clinical factors and use the latter for
+forecasting. However, doing so needs a deep expertise in several medical fields
+and the model needs a large amount of medical data. Another way is to use
+the multivariate Lee-Carter framework to get the join distribution of cause-
+specific death rates. The model is defined by the equation
+\begin{equation}\label{s11e5}
+\log\mu_{x,t,i} = \alpha_{x,i} + \beta_{x,i}\kappa_{t,i},
+\end{equation}
+where $\alpha_{x,i}$ is the static, cause-specific age structure, $\kappa_{t,i}$
+is the cause-specific mortality and $\beta_{x,i}$ is the sensitivity of the
+age class $x$ to the cause $i$.
+
+Cause-of-death models help us understand the interaction and joint-dynamics
+of the causes of death. They allow us to understand the severity of death-
+related pathologies. Further, if certain causes of death are prevented by
+improvements in healthcare or better policies then one can take that into 
+account while building aggregate mortality.
+
+\section{Generalized linear models}\label{s12}
+The ordinary least square linear regression assumes that the residuals are
+normally distributed. The expected value of the response is a linear
+combination of the predictors. Therefore a constant change in the predictor
+results in a proportional change in the respose, a characteristic of a linear
+relationship. However, often times this is not true. A generalized linear 
+model expresses a certain function of the response as a linear model. That is,
+instead of fitting a linear model between a response $y$ and the predictors $X$,
+we git a linear model between a function $g(y)$ of the response and the
+predictors. The function $g$ is called the link function. This method,
+developed by Nelder and Wedderburn \cite{nelder1972generalized} proposes a 
+general structure for several types of regression including the traditional
+ordinary-least-squares, the logistic and the Poisson regressions.
+
+If the probablity mass or density function of a random variable $Y$ can be
+written as
+\begin{equation}\label{s12e1}
+f_Y(y, \theta; \phi) = 
+\exp\left(\frac{y(\theta) - b(\theta)}{\alpha(\phi)} + c(y, \phi)\right)
+\end{equation}
+where $a, b, c$ are functions of their arguments and $\theta, \phi$ are
+parameters then $Y$ is said to belong to an exponential family of distributions.The normal, binomial, Poisson, exponential and gamma distributions are all
+members of the exponential family. A generalized linear model assumes that
+the responses $Y_1, \ldots, Y_n$ are independent and identically distributed
+with the distribution from the exponential family.
+
+A generalized linear model consists of 1) a distribution from the exponential
+family, 2) a linear predictor $\eta = X\beta$ and 3) a link function $g$
+such that $\E(Y|X) = g^{-1}(\eta)$.
+
+If we write $B = e^{k_1}$ in equation \eqref{s3e3} describing the force of
+mortality of Gompertz' law then we can take logarithm of both sides to get
+\begin{equation}\label{e12e2}
+\log\mu(x) = k_1 + kx.
+\end{equation}
+The right hand side of this equation is a linear function in $x$ while the left
+hand side is a logarithm. This is precisely the form of generalized linear
+model with logarithmic link function and a Poisson distrubution. If the errors
+in Gompertz's law are modeled in terms of a binomial distribution then the
+link function is a complimentary log-log function \cite{currie2013fitting} 
+defined as
+\begin{equation}\label{e12e3}
+f(x) = \log(-\log(1 - x)),
+\end{equation}
+where $0 < x < 1$. Mortality laws expressed as sigmoid functions of the form
+\begin{equation}\label{e12e4}
+q_x = \frac{e^{\theta_0 + \theta_1 x}}{1 + e^{\theta_0 + \theta_1 x}},
+\end{equation}
+can be easily transformed to the form
+\begin{equation}\label{e12e5}
+\frac{q_x}{1 - q_x} = e^{\theta_0 + \theta_1 x}
+\end{equation}
+or, taking logarithm of both sides,
+\begin{equation}\label{e12e6}
+\log\frac{q_x}{1 - q_x} = \theta_0 + \theta_1 x.
+\end{equation}
+Once again we observe that the right hand side is a linear function of the
+predictor. The function on the left hand is called a logit function and it is
+a legitimate link function for a generalized linear model. In fact, this link
+function is used to implement the standard logistic regression in applied
+statistics. It is possible to show that the Age-Period-Cohort models and the
+Cairns-Blake-Dowd models can also be expressed as generalized linear models. 
+The algebra involved to demonstrate this fact is a bit involved and we refer
+to the paper by Currie \cite{currie2013fitting} for the details.
+
+Generalized linear model gives the theoretical structure to several mortality
+models and also suggests a practical implementation. It is important to note
+that generalized linear models were not developed to build newer mortality
+forecasting models. Rather, deterministic mortality models were cast as 
+generalized linear models and that permitted practitioners to use a standard
+algorithm to fit their data. Unlike ordinary least squares regression which
+has an exact analytical solution, none exists for the generalized linear models.
+They must be solved numerically using an optimizer. The optimizer is guaranteed
+to find a global minimum if its objective function is convex. When it is not
+it is not uncommon for the optimizer to settle on a local minimum. We mentioned
+this problem in the case of the Renshaw-Haberman model in section \ref{s9}.
+
+\section{Spline models}\label{s13}
+The Lee-Carter model of equation \eqref{s6e1} and the Renshaw-Haberman model
+of equation \eqref{s9e1} contain multiplicative terms of predictors. It is 
+not possible to cast them as generalized linear models. However, it is possible
+to express them as generalized nonlinear models with a link function. Both
+these models can be expressed in terms of Poisson distribution and a logarithmic
+link function. Like their linear counterparts, generalized non-linear models
+do not have an exact analytical solution. Their numerical solution needs an
+optimizer. However, it is not always possible to find an optimal solution.
+Currie reports that \cite{currie2013fitting} among the $25$ simulations he
+tried to fit these models on UK data, in six cases the model did not converge
+and in five other cases the optimizer was unable to fit the model.
+
+Splines were used to draw a smooth curve passing through a set of points. They
+were initially used by draftsmen who represented the points by heavy leads and
+the splines by flexible wooden strips. This idea was borrowed into numerical
+analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
+of the twentieth century. A set of polynomials were used to approximate a
+non-linear function where each polynomial fitted a portion of the function
+being approximated. It was possible to show that under certian conditions
+\cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
+piecewise polynomial function. The polynomials were required to blend into each
+other that the boundaries in a smooth way. That is, not only were their values
+required to match but also their derivatives.
+
+Splines can be implemented using the Cox-de Boor recurrence relation
+\cite{de1978practical}
+\begin{equation}\label{s13e1}
+B_{i,0}(x) = \begin{cases}
+1 & \text{ if } x \in [t_i, t_{i+1}] \\
+0 & \text{ otherwise}
+\end{cases}
+\end{equation}
+and
+\begin{equation}\label{s13e2}
+B_{i,j}(x) = \frac{x - t_i}{t_{i+j} - t_i}B_{i,j-1}(x) + 
+\frac{t_{i+j+1}-x}{t_{i+j+1}-t_{i+1}}B_{i+1,j-1}(x).
+\end{equation}
+
 \section{Conclusion}

### 2020-09-14 mortality-models/draft.tex
WIP 14Sep2020

@@
 
-John Graunt, an English haberdasher, was perhaps the world's first demographer
-and epidemiologist. In spite of the rudimentary mathematical tools available
-in the seventeenth century, he was able to publish the first life tables. They
-were based on the `Bills of Mortality' published in the parishes of London
-every Thursday. The data was not best suited for constructing life tables but
-that was the only one available in Graunt's time. Graunt admitted the 
-inadequacy of his approach. Edmund Halley, the man after whom a comet is 
-named, could get demographic data of a much better quality from the city of
-Wroclaw in Poland. It had the age and sex of the dead individuals. Halley
-used this data to build life tables of a greater accuracy than Graunt's. He
-also demonstrated how to use the life tables to compute annuities. Graunt's
-and Halley's work was entirely empirical, keeping with the times in which
-it was done. Using their life tables was also quite laborious. With the 
-invention of Calculus in the late seventeenth century and the development of 
-the rudimentary ideas of a mathematical function, it became possible to build
-theoretical models of human mortality.
-
 Traditionally, mortality models studied existing demographic data to produce
@@
 and finally we consider stochastic models with cohort effects in section 10. 
+The models considered so far have looked at mortality irrespective of the
+cause of death. However, in recent years, death rate due to certain causes
+is falling while those due to others is rising. One can, therefore, forecast
+death rates by cause and aggregate the results. This is the approach taken by
+the cause-of-death models in section \ref{s11}. Several mortality models can
+be written in the form $f(\mu) = L(\theta)$ where $L$ is a linear function of
+the parameters $\theta$ of the model and $f$ is some function of the force
+of mortality that permits us to have the right hand side as $L(\theta)$. The
+statistically rigorous method to fit such models to mortality data is called
+the generalized linear models. Section \ref{s12} discusses them. Not all
+models can be brought in this format. Important examples being Lee-Carter and
+Renshaw-Haberman. One has to take recourse to spline regression in these
+cases. They are the topic of section \ref{s13}.
 
@@
 analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
-of the twentieth century. A set of polynomials were used to approximate a
+of the twentieth century. A set of polynomials was used to approximate a
 non-linear function where each polynomial fitted a portion of the function
@@
 \end{equation}
+These splines are then joined at the knot points $\{t_i: i = 1, \ldots, n\}$.
+Once the knots are given it is easy to find the splines. However, there is
+little guidance on the number of knots. Too few of them lead to under-fitting
+and too many of them lead to over-fitting. O'Sullivan \cite{o1986statistical}
+suggested that one should use a relatively large number of knots and prevent
+over-fitting by levying a penalty on the second derivatives of the splines.
+Eilers and Marx \cite{eilers1996flexible} simplify O'Sullivan's suggestion
+and apply it to regression on B-splines. They applied the penalty to the
+difference between the coefficients of the adjacent splines. The optimal amount
+of smoothing is determined by cross-validation and the Akaike Information 
+Criterion (AIC). Eilers and Marx call splines with penalties as P-splines.
+They demonstrate that P-splines are an extension of the generalized linear
+models. P-spline transforms conserve statistical moments of the data and have
+polynomial curve fits as limits. Currie and others \cite{currie2004smoothing}
+show how to use P-splines to fit mortality forecasting models. If $x_a = (
+x_{a,1},\ldots,x_{a,n_a})^T$ and $x_y=(x_{y,1},\ldots,x_{y,n_y})^T$ are the
+vectors with ages and years, $\mathcal{B}_a = \{B_{a,1}, \ldots, B_{a, c_a}\}$
+and $\mathcal{B}_y = \{B_{y,1},\ldots,B_{y, c_y}\}$ are the cubic B-slines of
+dimensions $c_a$ and $c_y$ then
+\begin{equation}\label{c14e1}
+X = B_y \otimes B_a
+\end{equation}
+is the regression matrix for the P-spline model
+\begin{equation}\label{c14e2}
+M(x,t) = \sum_{i,j}\theta_{i,j}B_{i,j}^{ay}(x, t).
+\end{equation}
+Here $B_a =(B_{a,j}(x_{a,i}))$ is an $n_a \times c_a$ matrix and $B_y = 
+(B_{y,j}(x_{y,i}))$ is an $n_y \times c_y$ matrix. The penalty matrix is given
+as
+\begin{equation}\label{c14e3}
+P = \lambda_aI_{c_y} \otimes D_a^TD_a + \lambda_y D_y^TD_y\otimes I_{c_a},
+\end{equation}
+where $D_a$ is the difference matrix of the rows of M, $D_y$ is the difference
+matrix on the columns of M, and the two $\lambda$s are the smoothing parameters.
+The regression problem is then minimizing the likelihood function
+\begin{equation}\label{c14e4}
+\ell_p = \ell - \frac{1}{2} m^TPm,
+\end{equation}
+where $\ell$ is the default log likelihood function and $m$ is the vector of
+the elements of $M$.
+
+The greatest benefit of splines over polynomial regression is that the former is
+a more stable numerical method, especially when used for forecasting. The
+reason for this stability is the local non-zero support for splines. 
+Polynomials, on the other hand, have a global support. The P-spline regression
+not only fits the mortality data smoothly, it also produces confidence intervals
+to its forecasts enabling us to compute bounds for financial parameters 
+depending on mortality.
 

### 2020-09-15 mortality-models/draft.tex
Spell check and other minor changes.

@@
 mathematical foundations of the theory of probability were laid. The 
-mathematical models they to describe how deaths occur in a given society are
+mathematical models they use to describe how deaths occur in a given society are
 called `mortality models'.
@@
 is an event happening to a member of the population. Likewise, the failure of
-a machine, an occurrence of an event, the duration of a certain phase in an
+a machine, the extinction of a species, the duration of a certain phase in an
 economy are all different manifestations of the same phenomenon in which 
@@
 world war and abundance of food in the rich world resulted in a change in
-historical mortality patterns. Pension funds started realizing that annuities
+historical mortality patterns. Healthcare systems began to come under stress.
+Pension funds started realizing that annuities
 were being drawn for longer than expected periods. As the population in the
@@
 is accompanied with appropriate confidence intervals to indicate the 
-uncertainty in their reported values.
+uncertainty in their reported values. Mortality models describe the mortality
+behavior at a fixed time. They do not predict the changes in mortality over
+long enough durations of the pension funds' interest. Mortality forecasts
+allow us to extrapolate the mortality curves several decades into the future.
 
@@
 lay in the number of successors it produced. Of these, we cover Makekam's 
-model and its generalizations in section 4. In section 5 we mention a few
+model and its generalizations in section \ref{s3}. In section \ref{s4} we 
+mention a few
 other mortality models used in European countries but whose significance today
@@
 was the first one that could be fitted to the entire mortality curve, starting
-from infancy to very advanced years. We describe it in section 6. Starting
-section 7 we turn our attention to stochastic models, starting with the
-celebrated Lee-Carter model. An interesting class of models used across social
-sciences is the Age-Period-Cohort models. We review a few of them relevant
-to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
-and finally we consider stochastic models with cohort effects in section 10. 
+from infancy to very advanced years. We describe it in section \ref{s5}. 
+Starting section \ref{s6} we turn our attention to stochastic models, starting 
+with the celebrated Lee-Carter model. An interesting class of models used 
+across social sciences is the Age-Period-Cohort models. We review a few of them 
+relevant to mortality in section \ref{s7}. In section \ref{s8} we consider the 
+Cairns-Blake-Dowd model and we consider stochastic models with cohort effects in
+ section \ref{s9}. 
 The models considered so far have looked at mortality irrespective of the
@@
 Renshaw-Haberman. One has to take recourse to spline regression in these
-cases. They are the topic of section \ref{s13}.
+cases. They are the topic of section \ref{s13}. We conclude our review in the
+final section.
 
@@
 terms. However, they were not always easy to fit on available mortality
-data. The presence of exponentials also made them numerically unstable. 
+data. The presence of exponential also made them numerically unstable. 
 Moreover, they covered only a limited number of sources of mortality.
@@
 combination of the predictors. Therefore a constant change in the predictor
-results in a proportional change in the respose, a characteristic of a linear
+results in a proportional change in the response, a characteristic of a linear
 relationship. However, often times this is not true. A generalized linear 
@@
 
-If the probablity mass or density function of a random variable $Y$ can be
+If the probability mass or density function of a random variable $Y$ can be
 written as
@@
 hand side is a logarithm. This is precisely the form of generalized linear
-model with logarithmic link function and a Poisson distrubution. If the errors
+model with logarithmic link function and a Poisson distribution. If the errors
 in Gompertz's law are modeled in terms of a binomial distribution then the
@@
 non-linear function where each polynomial fitted a portion of the function
-being approximated. It was possible to show that under certian conditions
+being approximated. It was possible to show that under certain conditions
 \cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
-piecewise polynomial function. The polynomials were required to blend into each
+Pierce's polynomial function. The polynomials were required to blend into each
 other that the boundaries in a smooth way. That is, not only were their values
@@
 vectors with ages and years, $\mathcal{B}_a = \{B_{a,1}, \ldots, B_{a, c_a}\}$
-and $\mathcal{B}_y = \{B_{y,1},\ldots,B_{y, c_y}\}$ are the cubic B-slines of
+and $\mathcal{B}_y = \{B_{y,1},\ldots,B_{y, c_y}\}$ are the cubic B-splines of
 dimensions $c_a$ and $c_y$ then

### 2020-09-21 speed-dating/generate_sample.py
Added two more choices.

@@ -3,15 +3,50 @@
 import random
 import pandas as pd
 
-sample_size = 10
-self_choices = [1, 2, 3, 4]
+sample_size = 30
+self_choices = [1, 2, 3] # The choice 4 is added to the UI but it is same as 3.
 seek_choices = [1, 2, 3]
 
+def choice_2(seek1, p):
+    """Get the second choice of partner.
+
+        Second partner is chosen with a probability p.
+        We ensure that the second choice is not the same as the first one.
+        Return -1 if the second choice is not made.
+    """
+    if random.uniform(0, 1) > p:
+        c = random.choice(seek_choices)
+        if c == seek1:
+            c = (c + 1) % 4
+    else:
+        c = -1
+
+    return c
+
+def choice_3(seek1, seek2, p):
+    """Get the third choice of partner.
+
+        Third partner is chosen with a probability p.
+        We ensure that the third choice is not the same as the first two.
+        Return -1 if the second choice is not made.
+    """
+    if seek2 != -1 and random.uniform(0, 1) > p:
+        c = random.choice(seek_choices)
+        if c == seek1 or c == seek2:
+            c = 6 - (seek1 + seek2)
+    else:
+        c = -1
+
+    return c
+
 self = random.choices(self_choices, k=sample_size)
-seek = random.choices(seek_choices, k=sample_size)
+seek_1 = random.choices(seek_choices, k=sample_size)
+seek_2 = [choice_2(seek_1[i], 0.7) for i in range(sample_size)]
+seek_3 = [choice_3(seek_1[i], seek_2[i], 0.8) for i in range(sample_size)]
 ids  = [i for i in range(1, sample_size + 1)]
 
-df = pd.DataFrame(list(zip(ids, self, seek)), columns=['id', 'self', 'seek'])
+df = pd.DataFrame(list(zip(ids, self, seek_1, seek_2, seek_3)), \
+                  columns=['id', 'self', 'seek_1', 'seek_2', 'seek_3'])
 filename = f'sample_{sample_size}.csv'
 df.to_csv(filename, index=False)
 

### 2020-09-22 p5/stability_rnn.tex
WIP 22Sep2020

@@
 
+\section{GRU's as Dynamical Systems}\label{s11}
+The most distinctive characteristic of a recurrent neural network is that it
+allows previous outputs to be used as inputs. Mathematically, one can express
+them as
+\begin{equation}\label{s11e1}
+h_{t+1} = f_1(h_t, x_t),
+\end{equation}
+where $x_t$ is the input and $h_t$ is the hidden state at `time' $t$. The
+function $f$ is usually non-linear. Equation \eqref{s11e1} is the general form
+of a non-linear difference equation. In the absence of any input, the hidden
+state evolves as
+\begin{equation}\label{s11e2}
+h_{t+1} = f_1(h_t, 0) = f(h_t).
+\end{equation}
+This is an example of an autonomous dynamical system. The evolution of the
+hidden states of recurrent neural networks can thus be considered to be
+trajectories of dynamical systems. One can then use the rich theory of 
+dynamical systems to study the stability of recurrent neural networks.
+
+A gated recurrent unit with the structure
+\begin{eqnarray}
+z_t &=& \sigma(W_zx_t + U_zh_{t-1} + b_z) \label{s11e3} \\
+r_t &=& \sigma(W_rx_t + U_rh_{t-1} + b_r) \label{s11e4} \\
+h_t &=& (1 - z_t)\circ\tanh(W_hx_t + U_h(r_t\circ h_{t-1} + b_h)) + 
+        z_t \circ h_{t-1} \label{s11e5}
+\end{eqnarray}
+where $W_z, W_r, W_h \in \mathbb{R}^{d \times p}$, $U_z, U_r, U_h \in 
+\mathbb{R}^{d \times d}$ are parameter matrices and  $b_z, b_r, b_h \in
+\mathbb{R}^d$ are bias vectors. In the absence of an input the above
+equations become
+\begin{eqnarray}
+z_t &=& \sigma(U_zh_{t-1} + b_z) \label{s11e6} \\
+r_t &=& \sigma(U_rh_{t-1} + b_r) \label{s11e7} \\
+h_t &=& (1 - z_t)\circ\tanh(U_h(r_t\circ h_{t-1} + b_h)) + 
+        z_t \circ h_{t-1} \label{s11e8}
+\end{eqnarray}
+One can consider equations \eqref{s11e6} to \eqref{s11e8} to be discrete
+forms of the system of ODEs \cite{jordan2019gated}
+\begin{eqnarray}
+z(t) &=& \sigma(U_zh(t) + b_z) \label{s11e9} \\
+r(t) &=& \sigma(U_rh(t) + b_r) \label{s11e10} \\
+\dot{h}(t) &=& (1 - z(t))\circ
+	\left(-h(t)+\tanh(U_h(r(t)\circ h(t)) + b_h)\right) \label{s11e11},
+\end{eqnarray}
+where the dot over $h$ denotes its time derivative. We consider a special case
+of these equations with the following values of parameters, (first row of 
+table I in \cite{jordan2019gated},
+\begin{eqnarray}
+U_r &=&
 \bibliographystyle{plain}

### 2020-09-23 p5/stability_rnn.tex
WIP 23Sep2020

@@
 \usepackage{amsmath, amssymb, amsfonts, amsthm}
+\usepackage{graphicx}
 \DeclareMathOperator*{\argmax}{argmax}
@@
 where the dot over $h$ denotes its time derivative. We consider a special case
-of these equations with the following values of parameters, (first row of 
-table I in \cite{jordan2019gated},
+of these equations in the subsections to follow.
+
+\subsection{Example 1}
+We choose the following values of parameters, (first row of table I in 
+\cite{jordan2019gated}),
 \begin{eqnarray}
-U_r &=&
+U_h &=& 3I_2 \label{s11e12} \\
+U_r &=& 0 \label{s11e13} \\
+U_z &=& 0 \label{s11e14} \\
+b_h &=& 0 \label{s11e15} \\
+b_r &=& 0 \label{s11e16} \\
+b_z &=& 0 \label{s11e17} 
+\end{eqnarray}
+For these values $z(t) = r(t) = 1/2(1, 1)^T$ and
+\begin{equation}\label{s11e18}
+\dot{h}(t) = \frac{1}{2}\left(-h(t) + \tanh\left(\frac{3}{2}h(t)\right)\right)
+\end{equation}
+The stable fixed points of this equation are shown in figure \ref{f1}.
+\begin{figure}[h]
+\caption{Stable points}\label{f1}
+\includegraphics[scale=0.5]{Ex1_Stable_points.png}
+\centering
+\end{figure}
+The coloured dots indicate the start of the trajectory while the blue square
+indicates its end. Trajectories seem to converge on the stable points. The
+phase portrait also indicates that the arrows point toward the stable points.
+
+This equation also has an unable point at the origin. The phase portrait has
+arrows diverging from the origin. The trajectories follow these directions.
+\begin{figure}[h]
+\caption{Unstable points}\label{f2}
+\includegraphics[scale=0.5]{Ex1_Unstable_point.png}
+\centering
+\end{figure}
+We will now examine the behaviour of this unstable point for a collection of
+five such GRUs starting their trajectory in the vicinity of the origin.
+In each of these figures the trajectory of the GRUs with attention layer
+is shown as a dashed line ending with a red square. The trajectories of the
+individual GRUs are shown as solid lines ending at the blue square. The
+initial conditions of the GRUs were chosen randomly from a set of $16$ points
+on a square centred at the origin and of side $0.1$. Figure \ref{f3} shows
+the first set of $6$ samples and figure \ref{f4} shows the remaining $6$
+samples.
+
+In each of these samples, the trajectory of the GRUs with attention layer
+never goes farther than the individual GRUs. It does veer away from the 
+origin but at a slower rate. This observation confirms the theoretical analysis
+of the previous sections that an attention layer does not eliminate an 
+instability. However, they indicate that the attention layer mitigates it to
+a certain extent.
+
+\begin{figure}[h]
+\caption{Unstable point with attention layer - set 1}\label{f3}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_0}&
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_1}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_2}&
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_3}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_4}&
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_5}
+\end{array}$
+\end{center}
+\end{figure}
+
+\begin{figure}[h]
+\caption{Unstable point with attention layer - set 2}\label{f4}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_6}&
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_7}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_8}&
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_10}&
+\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
+\end{array}$
+\end{center}
+\end{figure}
+
+\subsection{Example 2}
 \bibliographystyle{plain}

### 2020-09-23 p5/stability_rnn.tex
Final draft

@@
 We follow the contemporary mathematical practice of not using separate
-notation for vectors and scalars. Instead, we define every symbol. Whenver
+notation for vectors and scalars. Instead, we define every symbol. Whenever
 a symbol stands for a vector, it denotes a column vector. $\mathbb{R}$ is 
@@
 certain $x$, say $x = x_0$. Equation \eqref{s3e1} and $y(x_0) = y_0$ is
-called an \emph{inital value problem}.
+called an \emph{initial value problem}.
 
@@
 \end{equation}
-then one can get the value oy $y$ at $t_n + h$ using the formula
+then one can get the value of $y$ at $t_n + h$ using the formula
 \begin{equation}\label{s6e2}
@@
 hidden layers are not connected to each other. The Burrage-Butcher tensor 
-is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semidefinite 
+is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semi-definite 
 unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
@@
 means that \emph{not all} solutions are stable. On the other hand stability 
-means that \emph{all} solutions are stable. An existence of a few coverging 
+means that \emph{all} solutions are stable. An existence of a few converging 
 solutions does not guarantee the stability of the algorithm. However, an 
@@
 The stable fixed points of this equation are shown in figure \ref{f1}.
-\begin{figure}[h]
+\begin{figure}[!htbp]
 \caption{Stable points}\label{f1}
@@
 
-This equation also has an unable point at the origin. The phase portrait has
-arrows diverging from the origin. The trajectories follow these directions.
-\begin{figure}[h]
+This equation also has an unstable point at the origin. The phase portrait 
+has arrows diverging from the origin. The trajectories follow these 
+directions.
+\begin{figure}[!htbp]
 \caption{Unstable points}\label{f2}
@@
 In each of these figures the trajectory of the GRUs with attention layer
-is shown as a dashed line ending with a red square. The trajectories of the
+is shown as a dashed line ending with a red square. The attention layer is 
+connected in the architecture proposed by Ran and others, described previously
+in the first subsection of section \ref{s10}. Equations \eqref{s10e1} to 
+\eqref{s10e4} were used to generate the final output. We used a randomly 
+generated weight matrix $W_H$. Since $W_v$ and $A$ are fixed we chose their
+product to be a fixed, but randomly generated, vector in $\mathbb{R}^2$. The
+weight matrix $W$, whose rows are used in equation \eqref{s10e3} too was 
+selected randomly.
+
+The trajectories of the
 individual GRUs are shown as solid lines ending at the blue square. The
@@
 
-\begin{figure}[h]
+\begin{figure}[!htbp]
 \caption{Unstable point with attention layer - set 1}\label{f3}
@@
 
-\begin{figure}[h]
+\begin{figure}[!htbp]
 \caption{Unstable point with attention layer - set 2}\label{f4}
@@
 \subsection{Example 2}
+We now choose the parameters in the last row in table I of 
+\cite{jordan2019gated}. Thus,
+\begin{eqnarray}
+U_h &=& \begin{pmatrix}2.9665 & -0.4771 \\ 0.4771 & 2.9665\end{pmatrix}
+         \label{s11e19} \\
+U_r &=& 0 \label{s11e20} \\
+U_z &=& 0 \label{s11e21} \\
+b_h &=& 0 \label{s11e22} \\
+b_r &=& 0 \label{s11e23} \\
+b_z &=& 0 \label{s11e24} 
+\end{eqnarray}
+The ODE for these parameters is
+\begin{equation}\label{s11e25}
+\dot{h}(t) = \frac{1}{2}\left(-h(t) + \tanh\left(\frac{1}{2} 
+\begin{pmatrix}2.9665h_x-0.4471h_y \\0.4471h_x+2.9665h_y\end{pmatrix}\right)
+\right)
+\end{equation}
+The phase portrait of this equation is shown in figure \ref{f5}.
+\begin{figure}[!htbp]
+\caption{Unstable points}\label{f5}
+\includegraphics[scale=0.5]{Ex2_stable_points.png}
+\centering
+\end{figure}
+No fixed points are visible in the domain shown in the figure although the
+trajectories seem to be converging asymptotically toward a few lines. The
+phase portrait nevertheless shows the possibility of the origin being an
+unstable point. The derivative of the function seems to be diverging from the
+origin in all directions. We examine the nature of this differential equation
+in the neighbourhood of the origin in figure \ref{f6}.
+\begin{figure}[!htbp]
+\caption{Unstable points}\label{f6}
+\includegraphics[scale=0.5]{Ex2_Unstable_point.png}
+\centering
+\end{figure}
+The trajectories are indeed diverging away from the origin. Trajectories
+starting from initial conditions close to each other end up going very far
+from each other. This is indeed a signature of an instability.
+
+We repeat the experiment we did in the previous example in which we attach
+an attention layer to five GRUs with different initial conditions, all in the
+neighbourhood of the origin. Once again we observe that the trajectory of the
+GRU's with attention layer does not reach a point farther than the trajectories
+of the individual GRUs.
+
+\begin{figure}[!htbp]
+\caption{Unstable point with attention layer - set 1}\label{f7}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_0}&
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_1}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_2}&
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_3}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_4}&
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_5}
+\end{array}$
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\caption{Unstable point with attention layer - set 2}\label{f8}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_6}&
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_7}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_8}&
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_9}
+\end{array}$
+\end{center}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_10}&
+\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_9}
+\end{array}$
+\end{center}
+\end{figure}
+
+\subsection{Remarks on the experiments}
+Recall the idea of stability of an ODE. If $y_1$ and $y_2$ are two solutions
+of the ODE with initial conditions close to each other then if $y_1$ and $y_2$
+remain close to each other for ever then we say that the ODE is stable in a
+neighbourhood of the two initial conditions. On the other hand, if $y_1$ and 
+$y_2$ go very far from each other then it means that a slight change in the
+initial conditions pushes the solutions very far from each other. It is clear
+from figures \ref{f3}, \ref{f4}, \ref{f7} and \ref{f8} that the individual 
+solutions of the ODEs tend to diverge from each other. Coupling a few ODEs 
+(GRUs) to an attention layer does not eliminate this behaviour. The dashed 
+lines indeed go in a different direction in every figure. But their final 
+destinations, marked by a red square, are closer to each other than the
+final destinations of the bare GRUs.
+
 \bibliographystyle{plain}

### 2020-09-23 speed-dating/Participant.hpp
First draft.

@@ -1,0 +1,26 @@
+#ifndef _PARTICIPANT_HPP_
+#define _PARTICIPANT_HPP_
+
+struct Participant {
+    int id;
+    int self;
+    int seek1;
+    int seek2;
+    int seek3;
+
+    Participant() {id = self = seek1 = seek2 = seek3 = 0;}
+    Participant(int i, int sl, int sk1, int sk2, int sk3) : id(i),
+        self(sl), seek1(sk1), seek2(sk2), seek3(sk3) {}
+	bool can_pair(const Participant& p) const;
+};
+
+bool Participant::can_pair(const Participant& p) const {
+	if (this->self == p.seek1 ||
+		this->self == p.seek2 ||
+		this->self == p.seek3)
+		return true;
+	else
+		return false;
+}
+
+#endif

### 2020-09-23 speed-dating/bgl1.cpp

@@ -1,0 +1,63 @@
+#include <iostream>
+#include <utility>
+#include <algorithm>
+#include <boost/graph/graph_traits.hpp>
+#include <boost/graph/adjacency_list.hpp>
+
+using namespace boost;
+typedef adjacency_list<vecS, vecS, bidirectionalS> Graph;
+
+int main(int, char* [], char* [])
+{
+	enum {A, B, C, D, E, N};
+	const int n_vertices = N;
+
+	void show_vertices(const Graph& g);
+	void show_edges(const Graph& g);
+
+	typedef std::pair<int, int> Edge;
+	Edge edges[] = {Edge(A, B), Edge(A, D), Edge(C, A), Edge(D, C),
+						 Edge(C, E), Edge(B, D), Edge(D, E)};
+	const int n_edges = sizeof(edges)/sizeof(edges[0]);
+
+	Graph g(n_vertices);
+	for (int i = 0; i < n_edges; ++i) {
+		add_edge(edges[i].first, edges[i].second, g);
+	}
+
+	std::cout << "Created the graph." << std::endl;
+	show_vertices(g);
+	show_edges(g);
+
+	return 0;
+}
+
+void show_vertices(const Graph &g) 
+{
+	typedef graph_traits<Graph>::vertex_descriptor Vertex;
+	typedef property_map<Graph, vertex_index_t>::type IndexMap;
+
+	IndexMap index = get(vertex_index, g);
+	std::cout << "Vertices: ";
+	typedef graph_traits<Graph>::vertex_iterator vertex_iter;
+	std::pair<vertex_iter, vertex_iter> vp;
+	for (vp = vertices(g); vp.first != vp.second; ++vp.first) {
+		Vertex v = *vp.first;
+		std::cout << index[v] << " " ;
+	}
+	std::cout << std::endl;
+}
+
+void show_edges(const Graph& g)
+{
+	std::cout << "Edges: ";
+	typedef property_map<Graph, vertex_index_t>::type IndexMap;
+	IndexMap index = get(vertex_index, g);
+	graph_traits<Graph>::edge_iterator ei, ei_end;
+	for (boost::tie(ei, ei_end) = edges(g); ei != ei_end; ++ei) {
+		std::cout << "(" << index[source(*ei, g)] 
+				  << "," << index[target(*ei, g)] << ") ";
+	}
+	std::cout << std::endl;
+}
+

### 2020-09-23 speed-dating/bgl2.cpp

@@ -1,0 +1,93 @@
+#include <iostream>
+#include <utility>
+#include <algorithm>
+#include <vector>
+#include <boost/graph/graph_traits.hpp>
+#include <boost/graph/adjacency_list.hpp>
+
+using namespace boost;
+typedef adjacency_list<vecS, vecS, bidirectionalS> Graph;
+
+struct Participant {
+	int id;
+	int self;
+	int seek1;
+	int seek2;
+	int seek3;
+
+	Participant() {id = self = seek1 = seek2 = seek3 = 0;}
+	Participant(int i, int sl, int sk1, int sk2, int sk3) : id(i), 
+		self(sl), seek1(sk1), seek2(sk2), seek3(sk3) {}
+};
+
+typedef std::pair<int, int> Edge;
+
+int main(int, char* [], char* [])
+{
+	Graph build_graph();
+	Graph g = build_graph();
+
+	return 0;
+}
+
+Graph build_graph()
+{
+	std::vector<Participant> get_vertices();
+	std::vector<Participant> vertices = get_vertices();
+	void show_edges(const Graph& g);
+
+	std::vector<Edge> get_edges(const std::vector<Participant>& vertices);
+	std::vector<Edge> edges = get_edges(vertices);
+
+	std::cout << "Got " << vertices.size() << " vertices" << std::endl;
+	std::cout << "Got " << edges.size() << " edges" << std::endl;
+
+	Graph g(vertices.size());
+	for (int i = 0; i < edges.size(); ++i) {
+		add_edge(edges[i].first, edges[i].second, g);
+	}
+
+	std::cout << "Created the graph." << std::endl;
+	show_edges(g);
+
+	return 0;
+}
+
+void show_edges(const Graph& g)
+{
+	std::cout << "Edges: ";
+	typedef property_map<Graph, vertex_index_t>::type IndexMap;
+	IndexMap index = get(vertex_index, g);
+	graph_traits<Graph>::edge_iterator ei, ei_end;
+
+	for (boost::tie(ei, ei_end) = edges(g); ei != ei_end; ++ei) {
+		std::cout << "(" << index[source(*ei, g)]
+				  << "," << index[target(*ei, g)] << ") ";
+	}
+	std::cout << std::endl;
+}
+
+std::vector<Participant> get_vertices()
+{
+	std::vector<Participant> vertices;
+
+	vertices.push_back(Participant(1, 1, 3, -1, -1));
[…]

### 2020-09-23 speed-dating/bgl3.cpp

@@ -1,0 +1,119 @@
+#include <cstdlib>
+#include <fstream>
+#include <iostream>
+#include <sstream>
+#include <string>
+#include <utility>
+#include <vector>
+#include <boost/graph/graph_traits.hpp>
+#include <boost/graph/adjacency_list.hpp>
+#include <boost/graph/max_cardinality_matching.hpp>
+
+using namespace boost;
+typedef adjacency_list<vecS, vecS, bidirectionalS> Graph;
+typedef std::pair<int, int> Edge;
+
+#include <Participant.hpp>
+
+std::vector<Edge> get_edges(const std::vector<Participant>& ps);
+void show_edges(const Graph& g);
+
+int main(int argc, char* argv[], char* envp[])
+{
+	if (argc < 2) {
+		std::cerr << "Name of the CSV file is missing." << std::endl;
+		return -1;
+	}
+
+	std::string filename = argv[1];
+	std::fstream f(filename);
+	std::string line;
+	int fields[5]; // To store values of the fields of a Participant object.
+
+	std::vector<Participant> ps;
+
+	while (getline(f, line)) {
+		std::string token;
+		std::stringstream ss(line);
+		std::vector<std::string> tokens;
+
+		while(getline(ss, token, ',')) {
+			tokens.push_back(token);
+		}
+		int j = 0;
+		for (std::vector<std::string>::const_iterator ci = tokens.begin();
+		     ci != tokens.end();
+			 ++ci) {
+			fields[j++] = std::atoi(ci->c_str());
+		}
+		ps.push_back(Participant(fields[0], fields[1], fields[2],
+   										   fields[3], fields[4]));
+	}
+
+	f.close();
+
+	std::cout << "Read " << ps.size() << " ps." << std::endl;
+
+	Graph g(ps.size());
+	std::vector<Edge> edges = get_edges(ps);
+
+	for (int i = 0; i < edges.size(); ++i) {
+		add_edge(edges[i].first, edges[i].second, g);
+	}
+	std::cout << "Created the graph." << std::endl;
+	show_edges(g);
+
+	std::vector<graph_traits<Graph>::vertex_descriptor> mate(ps.size());
+	edmonds_maximum_cardinality_matching(g, &mate[0]);
+
+	if (true) {
[…]

### 2020-09-23 speed-dating/generate_sample.py

@@ -3,7 +3,7 @@
 import random
 import pandas as pd
 
-sample_size = 30
+sample_size = 5
 self_choices = [1, 2, 3] # The choice 4 is added to the UI but it is same as 3.
 seek_choices = [1, 2, 3]
 

### 2020-09-23 speed-dating/bgl3.cpp
Minor formatting.

@@ -8,12 +8,11 @@
 #include <boost/graph/graph_traits.hpp>
 #include <boost/graph/adjacency_list.hpp>
 #include <boost/graph/max_cardinality_matching.hpp>
+#include <Participant.hpp>
 
 using namespace boost;
 typedef adjacency_list<vecS, vecS, bidirectionalS> Graph;
 typedef std::pair<int, int> Edge;
-
-#include <Participant.hpp>
 
 std::vector<Edge> get_edges(const std::vector<Participant>& ps);
 void show_edges(const Graph& g);
@@ -47,7 +46,7 @@ int main(int argc, char* argv[], char* envp[])
 			fields[j++] = std::atoi(ci->c_str());
 		}
 		ps.push_back(Participant(fields[0], fields[1], fields[2],
-   										   fields[3], fields[4]));
+                                 fields[3], fields[4]));
 	}
 
 	f.close();
@@ -66,21 +65,17 @@ int main(int argc, char* argv[], char* envp[])
 	std::vector<graph_traits<Graph>::vertex_descriptor> mate(ps.size());
 	edmonds_maximum_cardinality_matching(g, &mate[0]);
 
-	if (true) {
-		std::cout << "Found matching of size " << matching_size(g, &mate[0])
-				  << std::endl;
-		std::cout << "The matching is: " << std::endl;
-		graph_traits<Graph>::vertex_iterator vi, vi_end;
-		for(tie(vi, vi_end) = vertices(g); vi != vi_end; ++vi) {
-			if (mate[*vi] != graph_traits<Graph>::null_vertex() &&
-			    *vi < mate[*vi]) {
-				std::cout << "{" << *vi << "," << mate[*vi] << "}" << std::endl;
-			}
+	std::cout << "Found matching of size " << matching_size(g, &mate[0])
+		  << std::endl;
+	std::cout << "The matching is: " << std::endl;
+	graph_traits<Graph>::vertex_iterator vi, vi_end;
+	for(tie(vi, vi_end) = vertices(g); vi != vi_end; ++vi) {
+		if (mate[*vi] != graph_traits<Graph>::null_vertex() &&
+			*vi < mate[*vi]) {
+			std::cout << "{" << *vi << "," << mate[*vi] << "}" << std::endl;
 		}
-		std::cout << std::endl;
-	} else {
-		std::cerr << "Failed to find a matching." << std::endl;
-	}
+	}	
+	std::cout << std::endl;
 
 	return 0;
 }

### 2020-09-26 speed-dating/bgl3.cpp
Recording processing time, allowing header in input file.

@@ -1,5 +1,7 @@
+#include <chrono>
 #include <cstdlib>
 #include <fstream>
+#include <iomanip>
 #include <iostream>
 #include <sstream>
 #include <string>
@@ -24,6 +26,7 @@ int main(int argc, char* argv[], char* envp[])
 		return -1;
 	}
 
+	auto start = std::chrono::high_resolution_clock::now();
 	std::string filename = argv[1];
 	std::fstream f(filename);
 	std::string line;
@@ -31,7 +34,13 @@ int main(int argc, char* argv[], char* envp[])
 
 	std::vector<Participant> ps;
 
+	bool first_line = true;
 	while (getline(f, line)) {
+		if (first_line) {
+			first_line = false;
+			break;
+		}
+
 		std::string token;
 		std::stringstream ss(line);
 		std::vector<std::string> tokens;
@@ -51,7 +60,9 @@ int main(int argc, char* argv[], char* envp[])
 
 	f.close();
 
+#ifdef _DEBUG_
 	std::cout << "Read " << ps.size() << " ps." << std::endl;
+#endif
 
 	Graph g(ps.size());
 	std::vector<Edge> edges = get_edges(ps);
@@ -59,8 +70,10 @@ int main(int argc, char* argv[], char* envp[])
 	for (int i = 0; i < edges.size(); ++i) {
 		add_edge(edges[i].first, edges[i].second, g);
 	}
+#ifdef _DEBUG_
 	std::cout << "Created the graph." << std::endl;
 	show_edges(g);
+#endif
 
 	std::vector<graph_traits<Graph>::vertex_descriptor> mate(ps.size());
 	edmonds_maximum_cardinality_matching(g, &mate[0]);
@@ -77,6 +90,11 @@ int main(int argc, char* argv[], char* envp[])
 	}	
 	std::cout << std::endl;
 
+	auto end = std::chrono::high_resolution_clock::now();
+	double diff = std::chrono::duration_cast<std::chrono::nanoseconds>(end - start).count() * 1e-9;
+
+	std::cout << "Time taken: " << std::fixed << diff << std::setprecision(9) 
+			  << " s for " << ps.size() << " participants." << std::endl;
 	return 0;
 }
 
@@ -103,8 +121,6 @@ std::vector<Edge> get_edges(const std::vector<Participant>& ps)
 		for (int j = i + 1; j < n_ps; ++j) {
[…]

### 2020-09-26 speed-dating/check_matching.py
A program to check matchings.

@@ -1,0 +1,44 @@
+#!/bin/python
+
+import sys
+import pandas as pd
+
+def match(G, n1, n2):
+    m1, m2 = False, False
+
+    if (G.loc[G['id'] == n1]['self'].iloc[0] == G.loc[G['id'] == n2]['seek_1'].iloc[0]) or \
+       (G.loc[G['id'] == n1]['self'].iloc[0] == G.loc[G['id'] == n2]['seek_2'].iloc[0]) or \
+       (G.loc[G['id'] == n1]['self'].iloc[0] == G.loc[G['id'] == n2]['seek_3'].iloc[0]):
+        m1 = True
+
+    if (G.loc[G['id'] == n2]['self'].iloc[0] == G.loc[G['id'] == n1]['seek_1'].iloc[0]) or \
+       (G.loc[G['id'] == n2]['self'].iloc[0] == G.loc[G['id'] == n1]['seek_2'].iloc[0]) or \
+       (G.loc[G['id'] == n2]['self'].iloc[0] == G.loc[G['id'] == n1]['seek_3'].iloc[0]):
+        m2 = True
+
+    return (m2 and m2)
+
+def check(gfile, mfile):
+    G = pd.read_csv(gfile)
+    n_errors = 0
+
+    with open(mfile, 'r') as f:
+        for line in f:
+            nodes = [int(n.strip()) for n in line.replace('{','').replace('}','').split(',')] 
+            n1, n2 = nodes[0], nodes[1]
+            if not match(G, n1, n2):
+                n_errors += 1
+
+    print(f'There were {n_errors} errors.')
+
+def main():
+    if len(sys.argv) < 3:
+        print('Usage: check_matching <sample file> <match file>')
+        exit(-1)
+
+    check(sys.argv[1], sys.argv[2])
+
+if __name__ == '__main__':
+    main()
+
+

### 2020-09-27 neutron-imaging/neutron.tex
Initial draft.

\documentclass{article}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{url, graphicx}

\numberwithin{equation}{section}
\linespread{1.3}
\title{A Review of Neutron Imaging}
\author{To be filled}
\date{27-Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s1}
Nobody suspected the presence of a neutral particle in an atomic nucleus until
the 1930s. When Heisenberg developed matrix mechanics in 1925 and 
Schr\"{o}dinger wrote his celebrated equation, the nucleus was thought to be
made up of positively charged protons and a few negatively charged `internal
electron'. The latter were supposed to be in present in sufficient numbers to
account for the mass and charge of the nucleus. However, this idea became
increasingly untenable with the development of quantum statistical mechanics
and an understanding of Fermions and Bosons. When Oscar Klein 
\cite{klein1929reflexion} applied the newly discovered Dirac equation to the
`internal electrons' in the nucleus he found that they should find it 
extremely easy to escape the potential barrier of the proton's electric field.
These difficulties were finally resolved when Chadwick reported 
\cite{chadwick1932existence}, the existence of an electrically neutral particle, called the neutron, in the atomic nucleus.

\section{Physical properties of a neutron}\label{s2}
The mass of a free neutron is $1.674927471 \times 10^{-27}$ kg
\cite{taylor2007nist}. Its mean square radius is $0.8 \times 10^{-15}$ m
\cite{bogdan2015particles}. It has no electric charge. However, it has
an intrinsic spin of $1/2$ (in units of $\hslash$)
\cite{basdevant2005fundamentals}. As a result, although it is unaffected by 
an electric field, it does respond to a magnetic field. In fact, a neutron 
responds to all four fundamental forces in Nature. Outside of
the nucleus, the neutron is unstable. It decays into a proton, an electron and
an electron neutrino with a half life of about $10$ minutes and $10$ seconds
\cite{Nakamura_2010}. The energy of decay is $0.782$ MeV \cite{heyde2004basic}.
A small fraction, about $0.001$, of free neutrons also produce gamma radiation 
while decaying. An even smaller fraction, about $4$ per million, of neutrons
produce electrons with an energy lesser than $13.6$ eV. As a result, they
stay stay bound to the proton forming a hydrogen atom. The fact that neutrons
decay cannot be ignored while using them for imaging. This feature of the
neutrons distingushes them from the other probes used for imaging like optical
photons, X-rays, gamma rays, infra-red radiation, sound waves and electrons. 
The latter probes are not known to decay.

\section{Sources of neutrons}\label{s3}
A neutron imaging device must have a source of neutrons the way an X-ray
imaging system has an X-ray tube to generate the X-rays. In this section we
will review a few sources of neutrons. Nuclear fission reactors produce a large
number of neutrons. A typical reactor has a fluence rate of the order of 
$10^{15}$ neutrons per $\text{cm}^{-2}$ per second. However, their energies are 
spread over a large range and the neutrons can be used only within the premises
of the reactor. Instead of relying on nuclear fission reactors, one can look
at nuclei that spontaneously decay to produce neutrons. A popular isotope of 
half-life of $2.6$ years is ${}^{252}$Cf. A typical source of this kind 
produces on an average of $10^{8}$ neutrons per second. Other examples of such
nuclei are ${}^{254}$Cf, ${}^{244}$Cm, ${}^{242}$Cm, ${}^{238}$Pu and 
${}^{232}$U.

One can go back to
Chadwick's experiment \cite{chadwick1932existence} to devise yet another source
of neutrons. Chadwick bombarded a slab of Beryllium with alpha radiation 
emitted from decaying Polonium nuclei. The nuclear reaction
\begin{equation}\label{s3e1}
{}_2^4\text{He} + {}_4^9\text{Be} \rightarrow {}_6^{12}\text{C}+{}_0^1\text{n}
\end{equation}
produced the first neutron detected in a laboratory. Some other reactions that 
also yeild neutrons are\cite{turner2008atoms}
\begin{eqnarray}
{}^2_1\text{H} + {}^3_1\text{H}&\rightarrow&{}_2^{4}\text{He}+{}_0^1\text{n}
\label{s3e2} \\
{}^2_1\text{H} + {}^2_1\text{H}&\rightarrow&{}_2^{3}\text{He}+{}_0^1\text{n}
\label{s3e3} \\
{}^2_1\text{H} + {}_6^{12}\text{C}&\rightarrow&{}_7^{13}\text{N}+{}_0^1\text{n}
\label{s3e4} \\
{}^1_1\text{p} + {}_3^1\text{H}&\rightarrow&{}_2^{4}\text{He}+{}_0^1\text{n}
\label{s3e5} \\
{}_1^1\text{p} + {}_3^7\text{Li}&\rightarrow&{}_4^{7}\text{Be}+{}_0^1\text{n}
 \label{s3e6} 
\end{eqnarray}
Each of these processes produce neutrons with a continuous energy spectrum. If
we need neutrons with narrow energy spread then we have to consider nuclear
reactions triggered by photons. Chadwick and Goldhaber 
\cite{chadwick1934nuclear} bombarded Deuterium nuclei with gamma rays of energy
exceeding $2.22$ MeV to trigger the reaction
\begin{equation}\label{s3e7}
{}^2_1\text{H} + \gamma \rightarrow {}^1_1 \text{H} + {}^1_0 n.
\end{equation}
Similarly, a photon of energy more than $1.67$ MeV can disintegrate Beryllium
into two Helium nuclei and a neutron.
\begin{equation}\label{s3e8}
{}^9_4\text{Be} + \gamma \rightarrow 2 {}^4_2\text{He} + {}^1_0 n.
\end{equation}
The $\gamma$-rays themselves come from radioactive ${}^{124}_{51}$Sb which 
spontaneously disintegrates as
\begin{equation}\label{s3e9}
{}^{124}_{51}\text{Sb} \rightarrow {}^{124}_{52}\text{Te} + e^{-1} + \gamma.
\end{equation}
Other commonly used isotopes to generate $\gamma$ radiation are ${}^{24}_{11}$
Na, ${}^{116}_{49}$In, ${}^{140}_{57}$La and ${}^{226}_{88}$Ra. A recent method
of generating neutrons involes spallation of heavy nuclei due to impact of 
very high energy protons. The term spallation means ejection of fragments from
a body due to extraordinarily large stress. In the case of nuclear spallation,
$\text{H}^{-}$ ions are accelerated to energy up to $1$ GeV before passing
them through a foil. The latter process strips the $\text{H}^{-}$ ion of its
two electrons and leaves behind a bare proton. The protons so generated are
made to pass at relativistic speeds into an accumulator ring. They are 
ejected out of the ring at a $60$ Hz into a tank of liquid Hg where they
trigger spallation of Hg nuclei. The greatest benefit of spallation is that
the resulting neutrons do not trigger a chain reaction as they do in the case
of nuclear fission reactors.

Neutrons of energies up to $0.1$ MeV are called `slow' neutrons, `Fast'
neutrons have energies in the range $10$ to $20$ MeV while those with even
higher energies are called relativistic neutrons.

\section{Interaction of neutrons with matter}\label{s4}
\section{Detection of neutrons}\label{s5}
\section{Neutron dosimetry and shielding}\label{s6}
\section{Propagation of neutron beams}\label{s7}
\section{Neutron diffraction by crystals}\label{s8}
\section{2-d neutron imaging}\label{s9}
\section{Neutron tomography}\label{s10}
\section{Imaging of biologically important structures}\label{s11}
\section{Larmor precession based imaging}\label{s12}
\section{Polarization based imaging}\label{s13}
\section{Conclusion}\label{s14}
\bibliography{neutron}
\bibliographystyle{plain}
\end{document}

### 2020-09-28 neutron-imaging/neutron.tex
WIP 29Sep2020

@@
 \section{Interaction of neutrons with matter}\label{s4}
+Neutrons being electrically neutral can travel significant distances in
+condensed matter. The probability $\mu$ that a neutron interacts with matter
+in unit distance is called its linear attenuation coefficient. It depends on
+the energy of the neutron and the properties of the condensed matter through
+which it is traveling. If $N(x)$ is the number of neutron that have traversed
+a distance $x$ in the condensed matter without any interaction then the number
+of neutrons $dN$ that will interact in the distance interval $x$ and $x + dx$
+is
+\begin{equation}\label{s4e1}
+dN = -\mu Ndx.
+\end{equation}
+The solution of this equation is 
+\begin{equation}\label{s4e2}
+N = N_0 e^{-\mu x},
+\end{equation}
+where $N_0 = N(x = 0)$ is the number of neutrons incident on a beam of the
+material. Neutrons interact only with atomic nuclei. In an elastic collision
+they transfer a portion of their kinetic energy to the nuclei. A series of
+elastic collisions leads to slowing of neutrons, also called their moderation.
+Eventually, the kinetic energy of the neutron drops down to the level of the
+thermal energy of the surroundings and it gets absorbed by a nucleus. Neutrons
+can also undergo inelastic collisions with nuclei. In this case, the neutron
+transfers a portion of its energy to the incident nucleus, raising it to an 
+excited state. Even in this case, neutron meets the same fate of losing energy
+up to the level of thermal energy and getting absorbed. Materials rich in 
+protons, that is H${}^+$ ions, show an overwhelmingly large proportion of
+elastic collisions because the bare proton does not have excited states. The
+de-energized neutron is absorbed leading the formation of a Deuterium atom
+\begin{equation}\label{s4e3
+{}_1^1H + {}^1_0 n \rightarrow {}^2_1 H + \gamma.
+\end{equation}
+The energy of the resulting $\gamma$-radiation is $2.22$ MeV. The precise 
+energy of this radiation is used to detect the presence of neutrons.
+In
+materials with other nuclei, a plot of the neutron collision cross sections 
+with neutron energies show pronounced peaks and troughs at the nuclear 
+transition levels. A notable aspect of elastic collisions is that a neutron of
+mass $m_n$ can transfer its kinetic energy to a nucleus of mass $m$ to the
+extent of
+\begin{equation}\label{s4e4}
+Q = \frac{4m_nmE_n}{(m + m_n)^2},
+\end{equation}
+where $E_n$ is the energy of the neutrons. In a material rich with protons,
+$m = m_n$ and $Q = 1$. Light nuclei are thus far more effective in slowing
+neutrons than heavier ones. Elastic collisions of neutrons with heavy nuclei
+change the direction of motion of nuclei while absorbing only a small portion
+of their kinetic energy.
 \section{Detection of neutrons}\label{s5}

### 2020-09-29 neutron-imaging/neutron.tex
WIP29Sep2020

@@
 de-energized neutron is absorbed leading the formation of a Deuterium atom
-\begin{equation}\label{s4e3
+\begin{equation}\label{s4e3}
 {}_1^1H + {}^1_0 n \rightarrow {}^2_1 H + \gamma.
@@
 \section{Detection of neutrons}\label{s5}
+Neutrons are chargeless and their magnetic moment is too low to be used for
+reliably detecting their presence. They also have a small electric dipole 
+moment but it has not been measured in a laboratory. Free neutrons decay to
+produce a proton and an electron but its half life of around $10$ minutes is
+too long to permit the detection of neutrons through its decay products. That
+leaves us with just one option, consider nuclear reactions triggered by the
+neutrons and measure their results to estimate the number and energy of the
+neutrons. Since the imaging applications involve mostly slow neutrons, we will
+focus our attention on their detection.
+
+Slow neutrons are detected by the nuclear reactions they trigger in ${}^{10}_5$
+B, ${}^6_3$Li and ${}^3_2$He. The first two of these produce alpha particles
+while the last one produces hydrogen atom. One can use non-relativistic energy
+and momentum conservation laws to show that the energy of the alpha particle
+in these reactions is \cite{turner2008atoms}
+\begin{equation}\label{s5e1}
+E = \frac{MQ}{m + M},
+\end{equation}
+where $m$ is the mass of the alpha particle, $M$ that of the daughter 
+nucleus and $Q$ is the energy released in the reaction. The resulting alpha
+particles are detected using proportional counters. Proportional counters
+measure only the number of neutrons, not their energy spectrum. The reactions
+used for detection are \cite{crane1991neutron}
+\begin{equation}\label{s5e2}
+{}^{10}_5\text{B}+{}^1_0\text{n} \rightarrow \begin{cases}
+{}^7_3\text{Li}^\star + {}^4_2\text{He} + 2.31\text{ MeV} \\
+{}^7_3\text{Li} + {}^4_2\text{He} + 2.79\text{ MeV}, 
+\end{cases}
+\end{equation}
+where the $\star$ superscript in the first reaction indicates that the nucleus
+is in an excited state. The other reactions are
+\begin{eqnarray}
+{}^6_3\text{Li}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^4_2\text{He} +
+4.78\text{ MeV} \label{s5e3} \\
+{}^3_2\text{He}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^1_1\text{H} +
+0.765\text{ MeV} \label{s5e4} 
+\end{eqnarray}
 \section{Neutron dosimetry and shielding}\label{s6}

### 2020-09-30 speed-dating/bgl3.cpp
Fixed a bug. Re-enabled printing intermediate messages.

@@ -38,7 +38,7 @@ int main(int argc, char* argv[], char* envp[])
 	while (getline(f, line)) {
 		if (first_line) {
 			first_line = false;
-			break;
+			continue;
 		}
 
 		std::string token;
@@ -60,9 +60,7 @@ int main(int argc, char* argv[], char* envp[])
 
 	f.close();
 
-#ifdef _DEBUG_
 	std::cout << "Read " << ps.size() << " ps." << std::endl;
-#endif
 
 	Graph g(ps.size());
 	std::vector<Edge> edges = get_edges(ps);
@@ -70,10 +68,8 @@ int main(int argc, char* argv[], char* envp[])
 	for (int i = 0; i < edges.size(); ++i) {
 		add_edge(edges[i].first, edges[i].second, g);
 	}
-#ifdef _DEBUG_
 	std::cout << "Created the graph." << std::endl;
 	show_edges(g);
-#endif
 
 	std::vector<graph_traits<Graph>::vertex_descriptor> mate(ps.size());
 	edmonds_maximum_cardinality_matching(g, &mate[0]);

### 2020-10-09 neutron-imaging/neutron.tex
WIP 09Oct2020

@@
 made up of positively charged protons and a few negatively charged `internal
-electron'. The latter were supposed to be in present in sufficient numbers to
+electrons'. The latter were supposed to be in present in sufficient numbers to
 account for the mass and charge of the nucleus. However, this idea became
@@
 The mass of a free neutron is $1.674927471 \times 10^{-27}$ kg
-\cite{taylor2007nist}. Its mean square radius is $0.8 \times 10^{-15}$ m
+\cite{taylor2007nist}. At non-relativistic speed $v$, its de Broglie wavelength
+is \cite{anderson2008neutron}
+\begin{equation}\label{s2e1}
+\lambda = \frac{395.6}{v} \text{nm}
+\end{equation}
+making it an ideal source to explore condensed matter at the atomic scale.
+
+Its mean square radius is $0.8 \times 10^{-15}$ m
 \cite{bogdan2015particles}. It has no electric charge. However, it has
@@
 of their kinetic energy.
+
+The interaction of neutrons with matter are used to develop neutron imaging
+techniques. The elastic scattering gives information about the atomic 
+structure. Neutron diffraction techniques have been used to study solids. On
+the other hand, inelastic scattering reveals information about collective
+excitations. Interactions of both kinds give rise to contrast that is used
+for imaging.
+
 \section{Detection of neutrons}\label{s5}
@@
 \end{eqnarray}
+We mentioned previously that the collision between a proton and a neutron is
+always elastic. Further, since their masses are almost equal, it is possible
+for the neutron to transfer all its kinetic energy to the proton. The proton
+is then detected by scintillators. Organic compounds are usually rich in 
+protons and detectors based on this principle use them in crystalline, 
+amorphous or even liquid forms.
+
+Bubble detectors are inexpensive and portable neutron detecting devices. They
+consist of a transparent polymer with a large number of tiny droplets of 
+a superheated liquid. The passage of a charged particle through such droplets
+results in a rapid phase change from liquid to vapor. The transparent droplets
+become visible bubbles which are counted either manually or electronically.
 \section{Neutron dosimetry and shielding}\label{s6}
+Radiation Dosimetry is the measurement of the `dose' of radiation absorbed by
+a biological sample. Dosimetry is important for the safety of people working
+with radiation. Biological materials are rich in protons. When they are
+exposed to neutrons, the elastic collision of the neutrons and the protons
+leads to a transfer of a substantial amount of energy to the protons resulting
+in breaking of chemical bonds. Neutrons damage biological samples more than
+a comparable dose of electrons or $\gamma$ rays. The amount of energy absorbed
+by $1$ kg of a material is called the `absorbed dose'. It is measured in grays,
+where $1$ gray is equal to $1$ Jkg${}^{-1}$. It is measured and specified at 
+every point of the exposed sample. Before getting into the details of neutron
+dosimetry, we introduce a few physical concepts.
+
+The restricted stopping power of a target is the the linear rate of energy of
+energy lost,  $(-dE/dx)_\Delta$, due to collisions alone and in which the
+energy transfer does not exceed $\Delta$. If $\mu$ is the probability per unit
+distance of travel that a neutron collides with a proton and if $W(Q)dQ$ is the 
+probability that a given collision will result in an energy loss between $Q$
+and $Q + dQ$ then
+\begin{equation}\label{s6e1}
+-\left(\frac{dE}{dx}\right)_\Delta = \mu\int_{Q_{\text{min}}}^\Delta QW(Q)dQ,
+\end{equation}
+where $Q_{\text{min}}$ is the minimum energy needed for excitation of the 
+target \cite{turner2008atoms}. The stopping power depends on both the 
+incident radiation and the absorbing sample. Radiation with high restricted
+stopping power is more dangerous than the one with low restricted stopping 
+power. The product of the absorbed dose $D$ and a quality factor $Q$ (not 
+related to the $Q$ in equation \eqref{s6e1} is called the `dose equivalent'
+$H$ of the sample. The absorbed dose is also a point function of the sample.
+That is, it can be specified at every point of the sample. The SI unit of 
+absorbed dose is sievert.
+
+Neutrons are best shielded by material rich in protons like Iron, water, 
+paraffin, concrete and graphite, in decreasing order of attenuation length
+\cite{turner2008atoms}.
 \section{Propagation of neutron beams}\label{s7}

### 2020-10-10 neutron-imaging/neutron.tex
WIP 10Oct2020

@@
 \cite{turner2008atoms}.
+
 \section{Propagation of neutron beams}\label{s7}
+Typical neutron sources have size of the order of $10$ cm in all directions.
+They are enclosed in moderators to produce a quasi-Maxwell distribution
+\cite{anderson2008neutron}. Their brightness is defined as the flux of neutrons
+per solid angle per unit wavelength. The flux itself is the number of neutrons
+crossing a unit area. One can treat the neutrons using the principles of
+geometrical optics. The phase space of analytical mechanics can be carried over
+to optics and is called etendue. Consider an infinitesimal surface element 
+$dS$ with normal $\hat{n}$ pointing in a medium of refractive index $n$. If
+the surface $dS$ is crossed by neutrons confined to a solid angle $d\Omega$
+then the etendue of neutron beam is
+\begin{equation}\label{s7e1}
+dG = n^2\hat{n}\cdot\hat{n}_b dSd\Omega,
+\end{equation}
+where $\hat{n_b}$ is the normal along the axis of the cone of $d\Omega$ 
+\cite{markvart2007thermodynamics}. The way the volume of the phase space is
+conserved, one can also conclude that the volume of an etendue is conserved. 
+The optical analogue of Liouville's theorem places a limit on the efficiency
+of a neutron source \cite{anderson2008neutron}, 
+\cite{markvart2007thermodynamics}. If one want to focus a neutron beam from
+a source of size $w_1$ on a spot of size $w_2 < w_1$ then one must allow the
+beam to diverge in directions normal to the line joining the source and the
+spot.
+
+Neutron beams can also be transported away from the source using wave guides.
+They are channels coated with Nickel which has a critical angle of $0.10^\circ$
+per $\AA$ neutron wavelength. Where higher critical angles than Nickel are 
+needed super-mirrors are used. Critical angles for super-mirrors can be $4$
+to $6$ times more than Nickel. Oftentimes it is beneficial to carry neutrons
+away from the source to prevent their contamination with fast neutrons which
+are also emitted from the source. Allowing the neutrons a longer flight time
+through wave guides also improves the resolution of pulsed-source instruments.
+The latter quantity is proportional to $\Delta t/t$, where $t$ is the total
+time of flight and $\Delta t$ is the time width of the pulse in the moderator.
+
+A neutron source produces neutrons with a wide range of wavelengths. A device
+that allows us to select one among them is called a monochromator. Crystal
+monochromators are based on Bragg's law
+\begin{equation}\label{s7e2}
+n\lambda = 2d\sin\theta,
+\end{equation}
+where $n$ is the `order' of diffraction, $\lambda$ is the wavelength of 
+neutrons, $d$ is the distance between the lattice planes and $\theta$ is the
+angle between the crystal and the incident beam. (The angle of incidence is
+$\pi/2 - \theta$.) Once a crystal is chosen, $d$ is fixed. One can select
+the wavelength of choice by changing $\theta$. The intensity of the diffracted
+beam reduces for higher order of diffractions. Therefore, monochromators us
+the first order beam.
+
+The direction of propagation of neutron beams can be changed by refraction.
+The refractive index of neutron beams is given by the Fermi thin-slab formula
+\cite{warner1985neutron}
+\begin{equation}\label{s7e3}
+n = 1 - \frac{2\pi\rho b}{k_0^2},
+\end{equation}
+where $b$ is the neutron scattering length, $\rho$ is the number density of
+the scattering nuclei and $k_0$ is the propagation constant in free space.
+A refractive index lesser than $1$ indicates that one must use a concave lens
+to focus a neutron beam. Although a neutron is electrically neutral, it does
+have spin. Therefore, it can be deflected using a magnetic field. If 
+$\vec{r}(t)$ is the trajectory of a neutron beam then its deflection in a
+magnetic field $\vec{B}$ is given by
+\begin{equation}\label{s7e4}
+\frac{d^2\vec{r}}{dt^2} = \pm\frac{\mu}{m}\nabla|\vec{B}|,
+\end{equation}
+where $\mu$ is the neutron's magnetic moment and $m$ is its mass.
+
 \section{Neutron diffraction by crystals}\label{s8}

### 2020-10-11 neutron-imaging/neutron.tex
WIP 11Oct2020

@@
 
-\section{Propagation of neutron beams}\label{s7}
+\section{Optics of neutron beams}\label{s7}
 Typical neutron sources have size of the order of $10$ cm in all directions.
@@
 
-\section{Neutron diffraction by crystals}\label{s8}
-\section{2-d neutron imaging}\label{s9}
-\section{Neutron tomography}\label{s10}
-\section{Imaging of biologically important structures}\label{s11}
-\section{Larmor precession based imaging}\label{s12}
-\section{Polarization based imaging}\label{s13}
-\section{Conclusion}\label{s14}
+The existence of spin states of neutrons allows us to describe beams of 
+neutrons by their polarization states. Being a spin-half particle, a neutron
+can exist in one of $|+\rangle$ of $|-\rangle$ states. If a beam has $N_+$
+neutrons in $|+\rangle$ and $N_-$ neutrons in $|-\rangle$ states then its
+polarization is defined to be
+\begin{equation}\label{s7e5}
+P = \frac{N_+ - N-}{N_+ + N_-}.
+\end{equation}
+It is equally common to express beam polarization in terms of flipping ratio
+as
+\begin{equation}\label{s7e6}
+F = \frac{N_+}{N_-} = \frac{1 - P}{1 + P}.
+\end{equation}
+When the nuclear and the magnetic structure factors of a crystal are equal in
+magnitude, the total structure factor vanishes for one of the two spin states.
+As a result, the Bragg reflection from the crystal plance produces a beam with
+almost all neutrons in the remaining polarization states. The most commonly
+used crystal to achieve neutron polarization is the [111] reflection of 
+Heusler alloy \cite{anderson2008neutron}. This technique is most suitable for
+steady state source of neutrons \cite{williams1988polarized}. In the case of
+a pulsed beam of neutrons it is common to use resonance absorption filters.
+When an unpolarized neutron beam whose energy corresponds to a nuclear
+resonance is shone on a target of polarized nuclear spins, one of the 
+neutron spins is absorbed while the other one is transmitted 
+\cite{williams1988polarized}.
+
+In order to use a neutron beam for imaging the neutrons emerging from a source
+must be collimated into a stream of particles traveling mostly along a single
+direction. Being electrically neutral, they cannot be focused using an electric
+field. Their small magnetic moment requires a very large field gradient to 
+change their direction of propagation. A simpler way of achieving collimation
+is to collect neutrons in a tube lined with a material which is opaque to 
+neutrons. Collimators have a small entrance aperture and a larger exit allowing
+a large neutron flux.
+
+\section{2-d neutron imaging}\label{s8}
+2-d neutron imaging is similar to the more commonly known 2-d X-ray imaging.
+When X-rays are incident on a target and when then the intensity of the 
+transmitted beam is analyzed one observes that different portions of the
+latter have different intensities. This is because different portions of the
+transmitted beam have emerged from different parts of the target each one
+having their own attenuation properties. 2-d neutron imaging works in exactly
+the same manner. Attenuation of X-rays depends on the local electronic
+properties of the target while that of neutron depends on the local nuclear
+properties. That is why, the two imaging techniques are often considered to
+provide complementary information about the target. X-rays provide information
+about the bonding or structural properties while neutrons indicate the 
+elemental composition. The set up for 2-d neutron imaging is quite simple. It 
+has a source of neutrons surrounded by a moderator to thermalize them, a 
+collimator to coax the particles along a single direction and a detector to
+visualize the image.
+
+The intensity of the incident and the transmitted beams is given by equation
+\eqref{s4e2}. If $I_0$ and $I$ are the intensities of the incident and the
+transmitted beams then the degree of darkness of the film used to visualize
+the image is
+\begin{equation}\label{s8e1}
+D_e = \ln\frac{I_0}{I}.
+\end{equation}
+The film brightens most where the transmitted intensity is the highest. In the
+case of a digital receptor, the grey level value $G$ is
+\begin{equation}\label{s8e2}
+G = Cn + G_0,
+\end{equation}
+where $C$ is a constant, $n$ is the neutron density and $G_0$ is another 
+constant that indicates the gray level due to noise in the circuits. The 
+presence of $G_0$ suggests that $G \ne 0$ even in the absence of neutron flux.
+This method suffers from the disadvantage of catching signals not only from
+the transmitted neutrons but from other ambient radiation as well. As a result,
+the images often are blurred. Nevertheless, it is used in situation where the
+sharpness of the image is not critical. For example, in the non-destructive
+testing in materials, it sufficient to know if the material has contaminants,
+defects, lubrication etc. It is also used in the investigation of two-phase
+flows, flows through porous media \cite{anderson2008neutron}, detection of
+hydrgenous materials and moisture \cite{richards2004neutron}.
+
+\section{Neutron tomography}\label{s9}
+Tomography is an imaging technique in which the image of a target is constructed
+by combining the images of its slices. The most usage is in medical imaging
+in the form of CT (Computerized Tomography) scans. X-rays are used over a 
+large number of slices of the human body. The `image' produced by each slice
+is combined with others to get complete information of the portion of interest.
+The way 2-d neutron imaging is an adaptation of 2-d X-ray imaging, neutron
+tomography is applying the principles of X-ray tomography to neutron imaging. 
+Consider a target $T$ whose image we are are interested in. We place the origin
+of the cartesian coordinate system at a suitable point in $T$. We consider $T$
+to be composed of a large number of slices whose normal is along the $z$-axis.
+We know from equation \eqref{s4e2} that the attenuation of intensity of an
+incident neutron beams by a slice $S$ is described by the equation
+\begin{equation}\label{s9e1}
+I = I_0\exp\left(-\int_S\mu(x, y)ds\right).
+\end{equation}
+Since the target is not a homogenous object, we cannot consider it to have a
+uniform attenuation $\mu$ as we did in equation \eqref{s4e2}. The parameter
+$s$ is the coordinate along the direction of propagation of neutrons in the
+target. We can rearrange equation \eqref{s9e1} and define a function
+\begin{equation}\label{s9e2}
+p(t, \theta) = \ln\left(\frac{I}{I_0}\right) = -\int_S\mu(x, y)ds.
+\end{equation}
+The coordinate $t$ runs perpendicular to the coordinate $s$. One can write 
+this in terms of an integral that extends over all $s$ as
+\begin{equation}\label{s9e3}
+p(t, \theta) = 
+-\int_{-\infty}^\infty \mu(x, y)\delta(t - x\cos\theta - y\sin\theta)ds.
+\end{equation}
+The function $p(t, \theta)$ is called the Radon transform of the slice. It is
+taken along the line $t = x\cos\theta + y\sin\theta$, expressed in the 
+normal form \cite{gonzales1992digital}. It is the projection of the function
+$f$ on the line $t = x\cos\theta + y\sin\theta$. It can be shown using the
+Fourier Slice Theorem \cite{bracewell1956strip} that a Fourier transform of
+$p(t, \theta)$ is the same as the slice of the 2-d Fourier transform of the
+target. Thus, taking the 2-d inverse Fourier transform of the result of the
+first operation will give us the original function. The number of such
+projections is determined by Shannon's sampling theorem.
+
+Neutron tomography has the same advantages over neutron radiography that CT
+scans has over X-ray images. Neutron radiography gives information about the
+total attenuation of a neutron beam along its path through the target. It 
+cannot give information about the actual distribution of materials across that
+path. One of the earliest use cases of neutron tomography was in finding 
+concentration of hydrogen in metal castings. The presence of hydrogen is an
+indicator of the loss of mechanical properties due to embrittlement. We know
+that the collision cross-section of neutron with hydrogen is very large and
+that neutron imaging is a powerful tool to investigate hydrogenous materials. 
+Neutron tomography techniques were able to give accurate information 
+about hydrogen concentration as low as $200$ ppm \cite{richards2004neutron}.
+Tomographic techniques are also suitable for investigating objects that do not
+have a substantial plane surface like the blade of a turbine or an aircraft 
+engine \cite{richards2004neutron}.
+
+\section{Imaging of biologically important structures}\label{s10}
+\section{Larmor precession based imaging}\label{s11}
+\section{Polarization based imaging}\label{s12}
+\section{Conclusion}\label{s13}
 \bibliography{neutron}

### 2020-10-12 neutron-imaging/neutron.tex
WIP 12Oct2020

@@
 
-\section{Imaging of biologically important structures}\label{s10}
+\section{Phase-contrast imaging}\label{s10}
+When electromagnetic radiation passes from one medium to another it undergoes
+both refraction and attenuation. The phenomena can be described together 
+through the complex refractive index \cite{heald2012classical} where the
+real part describes the refraction and the imaginary part the attenuation. 
+The propagation of neutron beams through media too can be expressed by a 
+complex refractive index. The 2-d imaging and the tomographic techniques
+considered in the previous sections considered images acquired due to 
+attenuation. In this section we shall briefly considered imaging due to phase
+contrast. Equation \eqref{s7e3} indicated that the refractive index of neutrons
+is less than $1$. Therefore, it is customary to express the complex refractive
+index as
+\begin{equation}\label{s10e1}
+n(\vec{x}) = 1 - \delta(\vec{x}) + i\beta(\vec{x}),
+\end{equation}
+where the dependence on $\vec{x}$, the position in the medium, allows for the
+inhomogeneous nature of the real and imaginary parts of the refractive index.
+Aluminum and Silicon have very small values $\beta$. Therefore, they hardly
+attenuate a neutron beam and cannot produce a 2-d image or a tomogram of 
+sufficient contrast. On the other hand, their $\delta$ values are $5$ orders
+of magnitude larger and can yield excellent phase-contrast images 
+\cite{anderson2008neutron}. Titanium and Copper have very similar $\beta$
+values and therefore cannot be differentiated by their attenuation properties.
+Their $\delta$ values are of opposite signs. Titanium with $\delta=-5.0 \times
+10^{-6}$ makes neutron beams behave somewhat like electromagnetic radiation
+by making the real part of the refractive index slightly greater than one
+\cite{anderson2008neutron}.
+
+The phase differences caused by the real part of the refractive index are
+detected by interference \cite{zawisky2004neutron}. A monochromatic beam of
+neutrons is coherently split into a reference beam and an object beam. The 
+object beam is passed through the target and suffers a phase shift (and 
+attenuation). After it emerges from the target it is superposed with the 
+reference beam. The interference between the two is captured in a digital
+neutron camera.
+
+Phase differences can either be recorded as a 2-d image or a tomogram. In the
+latter case, the 3-d distribution of the refractive index of the target is 
+reconstructed. The phase differences can either be represented as the 
+differences in their values or the gradient of phase shift or the laplacian of
+the phase shift \cite{anderson2008neutron}.
+
 \section{Larmor precession based imaging}\label{s11}
 \section{Polarization based imaging}\label{s12}
-\section{Conclusion}\label{s13}
+\section{Imaging of biologically important structures}\label{s13}
+\section{Conclusion}\label{s14}
 \bibliography{neutron}

### 2020-10-13 neutron-imaging/neutron.tex
WIP 13Oct2020

@@
 \section{Larmor precession based imaging}\label{s11}
+A magnetic dipole of moment $\vec{\mu}$ experiences a torque $\vec{\tau} = 
+\vec{\mu} \times \vec{B}$ in a magnetic field $\vec{B}$. If $\gamma$ is the
+gyromagnetic ratio of the dipole then $\vec{\mu} = \gamma\vec{L}$, $\vec{L}$
+being its angular momentum. Since
+\begin{equation}\label{s11e1}
+\frac{d\vec{L}}{dt} = \vec{\tau} = \gamma\vec{L} \times \vec{B},
+\end{equation}
+when $\vec{B}$ is not parallel (or anti-parallel) to $\vec{L}$, the vector 
+$\vec{L}$ precesses about $\vec{B}$ with a frequency $\omega = \gamma B$, 
+called the Larmor frequency. When a polarized beam of neutrons is made to
+pass through a magnetic material, the spin vector of the neutrons undergoes
+Larmor precession in the internal magnetic field of material. The angle of 
+spin changes by an amount that depends on the magnitude of the magnetic field
+and the time the neutron was present in it. The change in the spin orientation
+can be measured by a spin analyzer. Different regions of the material show
+different changes in spin orientation. By mapping the change in spin to grey
+scale one can create a contrast between these regions 
+\cite{kardjilov2011neutron}. This is the principle behind Larmor precession
+based neutron imaging. If the magnetic fields are sufficiently low so that
+the Larmor tilt is never more than $\pm\pi$, quantitative 2-d and 3-d imaging
+is possible. It was used to study the trapping of magnetic flux in type-I 
+superconductors \cite{kardjilov2008three}. This idea can be extended to 
+examine the how a ferromagnet becomes a paramagnet past the Curie temperature.
+When a beam of polarized neutrons falls on a ferromagnet, the randomly oriented
+magnetic domains of the ferromagnet depolarize the beam. When the same material
+is heated, the domains begin to weaken and beyond the Curie temperature they
+disappear altogether. The polarized beam is not transmitted undisturbed. This
+technique can be used to study the change in magnetic properties of materials
+in response to changing stress, pressure or applied electromagnetic fields
+\cite{kardjilov2011neutron}.
+
 \section{Polarization based imaging}\label{s12}

### 2020-10-14 neutron-imaging/neutron.tex
WIP 14Oct2020

@@
 
-\section{Polarization based imaging}\label{s12}
+\section{Spin-echo based imaging}\label{s12}
+Consider a beam of polarized nucleons (neutrons, protons, or nuclei with an odd
+number of nucleons) passes through a region of homogenous magnetic field its
+spin vector tilts with respect to the original direction because of Larmor
+precession. When the nucleons emerge from the region, they all emerge with the
+same tilt. However, as time goes by, different nucleons have a different angle
+of tilt. This happens because of natural spin relaxation and due to 
+inhomogeneities in the surroundings of the nucleons. If the nucleons are 
+subjected to a field that flips the spin by $\pi$ then after a while all spins
+that had dispersed due to inhomogeneities once again orient themselves together
+creating an `echo'. The magnitude of the echo is not the same as that of the
+original beam because of the natural spin relaxation. This is the principle
+behind the 'spin-echo' phenomenon used in nuclear magnetic resonance 
+spectroscopy and magnetic resonance imaging.
+
+In the nuclear spin echo experiment this arrangement is enhanced with an 
+additional region of uniform magnetic field, identical to the first one
+\cite{mezei1972neutron}. When a polarized neutron enters the first region
+of a uniform magnetic field, its spin vector tilts by an angle, say $\varphi$
+radians. Different neutrons experience a different tilt depending on the 
+time they spent in this region, which in turn depends on their velocity. Now
+suppose that the neutron beam is made to pass through a target. Each neutron
+will see a different local magnetic field in the target and will react to it
+in its unique way. The beam of neutrons is further defocused because of these
+inhomogeneities. It is then made to pass through a $\pi$ flipper. The 
+defocusing that had occurred due to inhomogeneities gets a chance to correct
+itself leading to a spin echo. A second region of a uniform magnetic field
+brings all the spins to the direction parallel to the originally polarized
+neutrons. The inelastic scattering of the neutrons in the target make some of
+them lose more energy than others. The second region of homogeneous magnetic
+field fails to bring such neutrons back to their original polarization state.
+The difference in the direction of polarization creates the contrast for
+neutron spin-echo imaging.
+
+Neutron spin-echo imaging is used with small angle neutron scattering in the
+investigation of soft matter. The latter technique provides information about
+the structure. However, the mechanical properties depends on molecular 
+mobilities and constraints of motion \cite{monkenbusch2007high}, where the
+energy differences are of the order of $0.7 \mu$eV. Such small changes can be
+resolved only through neutron spin-echo imaging.
+
 \section{Imaging of biologically important structures}\label{s13}
+Biological targets have an extremely rich mixture of molecules. A knowledge of
+the presence of certain trace elements in the targets reveals important 
+information about its functioning, health and pathology. Few imaging techniques
+can accurately provide information about the elements. The chemical shift 
+imaging in magnetic resonance imaging provides information at the molecular
+level. It provides little information about the elements present. This is where
+neutron imaging comes into the picture. When neutrons undergo an inelastic
+scattering with the nuclei of the target, they nuclei get into an excited
+state. They get back to their ground state after emitting a photon. We now 
+have a detailed knowledge of the nuclear energy levels. By measuring the energy
+of the emitted photon we can guess the nucleus, and thus the element, which
+might have produced it. This is the principle behing the neutron stimulated
+emission computed tomography (NSECT) imaging.
 \section{Conclusion}\label{s14}

### 2020-10-15 neutron-imaging/neutron.tex
Final version.

@@
 decay cannot be ignored while using them for imaging. This feature of the
-neutrons distingushes them from the other probes used for imaging like optical
+neutrons distinguishes them from the other probes used for imaging like optical
 photons, X-rays, gamma rays, infra-red radiation, sound waves and electrons. 
@@
 produced the first neutron detected in a laboratory. Some other reactions that 
-also yeild neutrons are\cite{turner2008atoms}
+also yield neutrons are\cite{turner2008atoms}
 \begin{eqnarray}
@@
 Na, ${}^{116}_{49}$In, ${}^{140}_{57}$La and ${}^{226}_{88}$Ra. A recent method
-of generating neutrons involes spallation of heavy nuclei due to impact of 
+of generating neutrons involves spallation of heavy nuclei due to impact of 
 very high energy protons. The term spallation means ejection of fragments from
@@
 \section{Detection of neutrons}\label{s5}
-Neutrons are chargeless and their magnetic moment is too low to be used for
+Neutrons are charge less and their magnetic moment is too low to be used for
 reliably detecting their presence. They also have a small electric dipole 
@@
 consist of a transparent polymer with a large number of tiny droplets of 
-a superheated liquid. The passage of a charged particle through such droplets
+a super-heated liquid. The passage of a charged particle through such droplets
 results in a rapid phase change from liquid to vapor. The transparent droplets
@@
 the wavelength of choice by changing $\theta$. The intensity of the diffracted
-beam reduces for higher order of diffractions. Therefore, monochromators us
+beam reduces for higher order of diffraction. Therefore, monochromators us
 the first order beam.
@@
 magnitude, the total structure factor vanishes for one of the two spin states.
-As a result, the Bragg reflection from the crystal plance produces a beam with
+As a result, the Bragg reflection from the crystal plane produces a beam with
 almost all neutrons in the remaining polarization states. The most commonly
@@
 The film brightens most where the transmitted intensity is the highest. In the
-case of a digital receptor, the grey level value $G$ is
+case of a digital receptor, the Grey level value $G$ is
 \begin{equation}\label{s8e2}
@@
 flows, flows through porous media \cite{anderson2008neutron}, detection of
-hydrgenous materials and moisture \cite{richards2004neutron}.
+hydrogenous materials and moisture \cite{richards2004neutron}.
 
@@
 Consider a target $T$ whose image we are are interested in. We place the origin
-of the cartesian coordinate system at a suitable point in $T$. We consider $T$
+of the Cartesian coordinate system at a suitable point in $T$. We consider $T$
 to be composed of a large number of slices whose normal is along the $z$-axis.
@@
 \end{equation}
-Since the target is not a homogenous object, we cannot consider it to have a
+Since the target is not a homogeneous object, we cannot consider it to have a
 uniform attenuation $\mu$ as we did in equation \eqref{s4e2}. The parameter
@@
 reconstructed. The phase differences can either be represented as the 
-differences in their values or the gradient of phase shift or the laplacian of
+differences in their values or the gradient of phase shift or the Laplacian of
 the phase shift \cite{anderson2008neutron}.
@@
 \vec{\mu} \times \vec{B}$ in a magnetic field $\vec{B}$. If $\gamma$ is the
-gyromagnetic ratio of the dipole then $\vec{\mu} = \gamma\vec{L}$, $\vec{L}$
+gyro-magnetic ratio of the dipole then $\vec{\mu} = \gamma\vec{L}$, $\vec{L}$
 being its angular momentum. Since
@@
 can be measured by a spin analyzer. Different regions of the material show
-different changes in spin orientation. By mapping the change in spin to grey
+different changes in spin orientation. By mapping the change in spin to Grey
 scale one can create a contrast between these regions 
@@
 Consider a beam of polarized nucleons (neutrons, protons, or nuclei with an odd
-number of nucleons) passes through a region of homogenous magnetic field its
+number of nucleons) passes through a region of homogeneous magnetic field its
 spin vector tilts with respect to the original direction because of Larmor
@@
 of the emitted photon we can guess the nucleus, and thus the element, which
-might have produced it. This is the principle behing the neutron stimulated
+might have produced it. This is the principle behind the neutron stimulated
 emission computed tomography (NSECT) imaging.
+
+In the previous section we made a passing reference to small angle neutron
+scattering as a technique complementary to neutron spin-echo imaging. We will
+now briefly outline the small angle scattering technique. It is used to gather
+structural and functional information about bio-molecules with a resolution
+in the range $1$ to $100$ nm. Biological macro-molecules do not have the
+structural order of a crystal. The scattering of neutrons by the nuclei of 
+such molecules does not produce a sharp diffraction pattern seen with crystals.
+Rather, they just produce an interference pattern. The intensity of the 
+interference pattern depends on the scattering length density. Neutrons have
+a different scattering length for Hydrogen and Deuterium nuclei and this fact
+is used to generate contrast in the images. Either the entire molecule is 
+imaged in a solution of D${}_2$O or a sub-unit of the molecule is marked with
+Deuterium to get the contrast \cite{anderson2008neutron}.
+
 \section{Conclusion}\label{s14}
+In this article we reviewed the physics of neutrons from the perspective of its
+role in probing materials at various scales of resolution. We started by 
+mentioning the physical properties of neutrons, the way they can be produced,
+how they can be detected and how a beam of neutrons can be manipulated. The
+electrical neutrality of neutrons allows them to penetrate deeply in materials.
+They also have a large collision cross-section with protons, making them
+potentially impactful for biological targets. As a result, one must be careful
+in exposing living matter to a beam of neutrons. We covered methods to measure
+dose of neutron radiation and ways to shield neutrons. Optical and electron
+microscopes rely heavily on our ability to manipulate light rays and beams of
+electrons. The electrical neutrality of neutrons and their weak intrinsic
+magnetic moment make it hard to manipulate their beams. Nevertheless, a few
+techniques are available and we covered them in the section on optics.
+
+In the second half of the article we reviewed imaging techniques starting with
+the simplest 2-d imaging in which we just record the intensity of the 
+transmitted beam and interpret it like an X-ray image. We then covered the
+tomographic technique borrowed from X-rays and described its superiority over
+2-d imaging in getting the distribution of nuclei in targets. Neutrons also 
+have a complex refractive index. The imaginary part of the refractive index is
+used to differentiate between nuclei that have similar attenuation properties.
+The magnetic moment of neutrons permit them to be treated like protons in the
+context of nuclear magnetic resonance. Larmor precession based imaging can be 
+used to study the magnetic properties of solids at the level of magnetic 
+domains. Neutron spin-echo is another technique borrowed from the world of 
+nuclear magnetic resonance. It is used to study the dynamic properties of 
+biologically important molecules. We closed the review with a section on the
+unique ways in which neutron imaging is used to investigate macro-molecules
+and detect elements in biological targets.
+
 \bibliography{neutron}

### 2020-10-16 speed-dating/bgl4.cpp
Cleanup

@@ -153,6 +153,7 @@ void previous_matches(const std::string &filename,
 	std::string line;
 
 	while (getline(f, line)) {
+		if (line.size() == 0) break;
 		std::stringstream ss(line);
 		std::string s;
 		getline(ss, s, ',');

### 2020-10-19 optim/or-tools/tsp/drill.py
WIP 19Oct2020

@@ -1,0 +1,125 @@
+#!/bin/python
+
+import math
+
+from ortools.constraint_solver import routing_enums_pb2
+from ortools.constraint_solver import pywrapcp
+
+def create_data_model():
+    """Stores the data for the problem."""
+    data = {}
+    # Locations in block units
+    data['locations'] = [
+        (288, 149), (288, 129), (270, 133), (256, 141), (256, 157), (246, 157),
+        (236, 169), (228, 169), (228, 161), (220, 169), (212, 169), (204, 169),
+        (196, 169), (188, 169), (196, 161), (188, 145), (172, 145), (164, 145),
+        (156, 145), (148, 145), (140, 145), (148, 169), (164, 169), (172, 169),
+        (156, 169), (140, 169), (132, 169), (124, 169), (116, 161), (104, 153),
+        (104, 161), (104, 169), (90, 165), (80, 157), (64, 157), (64, 165),
+        (56, 169), (56, 161), (56, 153), (56, 145), (56, 137), (56, 129),
+        (56, 121), (40, 121), (40, 129), (40, 137), (40, 145), (40, 153),
+        (40, 161), (40, 169), (32, 169), (32, 161), (32, 153), (32, 145),
+        (32, 137), (32, 129), (32, 121), (32, 113), (40, 113), (56, 113),
+        (56, 105), (48, 99), (40, 99), (32, 97), (32, 89), (24, 89),
+        (16, 97), (16, 109), (8, 109), (8, 97), (8, 89), (8, 81),
+        (8, 73), (8, 65), (8, 57), (16, 57), (8, 49), (8, 41),
+        (24, 45), (32, 41), (32, 49), (32, 57), (32, 65), (32, 73),
+        (32, 81), (40, 83), (40, 73), (40, 63), (40, 51), (44, 43),
+        (44, 35), (44, 27), (32, 25), (24, 25), (16, 25), (16, 17),
+        (24, 17), (32, 17), (44, 11), (56, 9), (56, 17), (56, 25),
+        (56, 33), (56, 41), (64, 41), (72, 41), (72, 49), (56, 49),
+        (48, 51), (56, 57), (56, 65), (48, 63), (48, 73), (56, 73),
+        (56, 81), (48, 83), (56, 89), (56, 97), (104, 97), (104, 105),
+        (104, 113), (104, 121), (104, 129), (104, 137), (104, 145), (116, 145),
+        (124, 145), (132, 145), (132, 137), (140, 137), (148, 137), (156, 137),
[…]

### 2020-10-19 p5/gru-al.py

@@ -1,0 +1,94 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import odeint
+import math
+import random
+
+# Helper functions
+def get_random_weights(n):
+    """Generate n random numbers whose total is 1."""
+    R = random.sample(range(1, 11), n)
+    return np.array([r/sum(R) for r in R])
+
+
+def get_random_points(radius, n):
+    """Generate n random points on a circle of given radius centred at the origin."""
+    W = get_random_weights(n)
+    points = [[radius*np.cos(2*np.pi*w), radius*np.sin(2*np.pi*w)] for w in W]
+    
+    return points
+
+def list_mul_2d(M, m):
+    """ M is a list of lists. Multiply each one of them by m."""
+    return [[m * x for x in L] for L in M]
+
+def eu_dist(x1, x2):
+    """Euclidean distance between points x1 and x2 in R^n."""   
+    return np.linalg.norm(x1 - x2)
+
+def get_distance(x_init, xs):
+    return [eu_dist(x_init, x) for x in xs]
+
+Wh = np.array([get_random_weights(3)])
+AWv = np.array([get_random_weights(2)])
+# Initial points on a small circle of radius 0.01.
+r = 0.01
+N = 10
+x_inits = [[r*np.cos(2*n*np.pi/N), r*np.sin(2*n*np.pi/N)] for n in range(1, N + 1)]
+
+# Build 4 x 3 weight matrix
+w1 = get_random_weights(4)
+w2 = get_random_weights(4)
+w3 = get_random_weights(4)
+W = np.matrix([w1, w2, w3]).T
+
+def f(h, t, U, extra = 0):  
+    v = np.matmul(U, h)
+    v0 = v[0]
+    v1 = v[0]
+    [x, y] = h
+    
+    xdot = 0.5*(np.tanh(v0) - x)
+    ydot = 0.5*(np.tanh(v1) - y)
+    
+    return [xdot, ydot]
+
+def get_HW(x, Wh):
+    H = np.matrix([x, x, x]).T
+    print(f'H = {H.shape}')
+    print(f'Wh = {Wh.shape}')
+    return np.matmul(H, Wh.T)
+
+def get_Ml(x, Wh, AWv):
+    hw = np.array(get_HW(x, Wh))
+    print(f'hw = {hw.shape}')
[…]

### 2020-10-19 speed-dating/read_csv.cpp

@@ -1,0 +1,37 @@
+#include <fstream>
+#include <iostream>
+#include <sstream>
+#include <string>
+#include <vector>
+
+int main(int argc, char* argv[], char* envp[])
+{
+	if (argc < 2) {
+		std::cerr << "Name of the CSV file is missing." << std::endl;
+		return -1;
+	}
+
+	std::string filename = argv[1];
+	std::fstream f(filename);
+	std::string line;
+
+	while (getline(f, line)) {
+		std::string token;
+		std::stringstream ss(line);
+		std::vector<std::string> tokens;
+
+		while(getline(ss, token, ',')) {
+			tokens.push_back(token);
+		}
+		for (std::vector<std::string>::const_iterator ci = tokens.begin();
+		     ci != tokens.end();
+			 ++ci) {
+			std::cout << *ci << ", ";
+		}
+		std::cout << std::endl;
+	}
+
+	f.close();
+	return 0;
+}
+

### 2020-10-21 neutron-imaging/neutron.tex
Added another section

@@
 
-\section{Imaging of biologically important structures}\label{s13}
+\section{Neutron imaging of materials}\label{s13}
+Neutrons are sensitive to lighter elements and sometimes they have a 
+significantly different collision cross section for neighboring elements 
+building a contrast between the portions in the images where they are present.
+Neutrons are the only probes that can distingush between isotopes. Their
+electrical neutrality permits them to penetrate deeply and their small magnetic
+moment allows them to interact magnetically with the target. All these
+properties make neutrons an attractive probe to investigate the static and
+dynamic properties of materials \cite{anderson2008neutron}.
+
+The lost foam casting process consists of molten metal replacing the polystyrene
+foam in a cast. Neutrons provide a sharp contrast between the hydrogen-rich
+polystyrene and the metal. It is possible to take frequent images of the 
+casting process to produce a video of how the molten metal displaces the foam
+and settles down in the cast to solidify. One can study the effect of foam
+thickness on the quality of casting. The video images also suggest a flow
+rate that will not trigger turbulence in the molten metal. They provide 
+details of the multiphase flow of the molten metal and the foam that no other
+imaging technique gives.
+
+The mechanical properties of materials depend on the local stress fields in
+them. Many solids have a non-zero local stress built up during its 
+solidification process. These stresses exist in absence of any traction or
+forces of constraints. Their presence can be detected from the strains 
+accompanying them. We mentioned previously that neutron diffraction is a 
+valuable tool to examine crystals. They measure the lattice parameters like
+the distance between lattice planes. If $d_0$ is the expected spacing between
+lattice plans and if $d$ is the observed value then it indicates the presence
+of a strain field of magnitude
+\[
+\varepsilon = \frac{d - d_0}{d_0}.
+\]
+The strain field is a tensor. A complete description of it requires one to
+measure all its components. Of the nine components, only six are independent.
+Once measures the strain at certain orientations $(\theta, \phi)$ and uses
+them to infer the six independent cartesian components of the strain tensor.
+The relationship between strain at an arbitray orientation $(\theta, \phi)$
+and the six cartesian components is
+\begin{eqnarray}
+\varepsilon(\theta, \phi) &=& \varepsilon_{xx}\cos^2\theta\sin^2\phi +
+\varepsilon_{xy}\sin(2\theta)\sin^2\phi \nonumber \\
+ & & + \varepsilon_{yy}\sin^2\theta\sin^2\phi+\varepsilon_{zz}\cos^2\phi 
+\nonumber \\
+ & & + \varepsilon_{xz}\cos\theta\sin(2\phi)+
+\varepsilon_{yz}\sin\theta\sin(2\phi) \label{s13e1}
+\end{eqnarray}
+If we have $\varepsilon(\theta,\phi)$ for six directions then we have six
+equations in six unknowns from which we can determine the entire strain field.
+Once we have the local strain field, we can compute the local stress field 
+using Hooke's law.
+
+The technique to measure local strain and stress described in the previous
+paragraph works well for materials that are largely homegeneous. In the case
+of granular and porous materials, the stress-strain relations are very 
+complicated. They depend on the instantaneous forces as well as past history
+of loading. Unlike the homogeneous materials porous and granular materials 
+have inhomogenous, anisotropic and non-linear stress-strain fields. Traditional
+probes like X-rays cannot penetrate deep inside the material to measure the
+fields. However neutron beams can easily reach depths that X-rays cannot making
+neutrons a valuable tool to measure stresses and strains in granular media.
+
+\section{Imaging of biologically important structures}\label{s14}
 Biological targets have an extremely rich mixture of molecules. A knowledge of
@@
 
-\section{Conclusion}\label{s14}
+\section{Conclusion}\label{s15}
 In this article we reviewed the physics of neutrons from the perspective of its
@@
 nuclear magnetic resonance. It is used to study the dynamic properties of 
-biologically important molecules. We closed the review with a section on the
+biologically important molecules. We then covered the use of neutron probes to 
+measure local stress-strain fields in homogeneous and granular materials. 
+We closed the review with a section on the
 unique ways in which neutron imaging is used to investigate macro-molecules

### 2020-11-01 queueing-theory/resample.py
Initial draft

@@ -1,0 +1,12 @@
+#!/bin/py
+
+# The data in Excel is at a frequency of 30 minutes. I need it at an interval
+# of 5 minutes. This calls for a resampling. The function in this file does
+# it.
+
+def resample(T):
+    T1 = [[n] * 6 for n in T]
+    T2 = [n for sublist in T1 for n in sublist]
+
+    return T2
+

### 2020-11-01 queueing-theory/resample.py
Removed magic numbers.

@@ -4,7 +4,9 @@
 # of 5 minutes. This calls for a resampling. The function in this file does
 # it.
 
-def resample(T):
+def resample(T, available_freq, desired_freq):
+    factor = int(available_freq/desired_freq)
+
     T1 = [[n] * 6 for n in T]
     T2 = [n for sublist in T1 for n in sublist]
 

### 2020-11-03 fdtd/em.tex
WIP 03Nov2020

\documentclass[11pt]{article}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{physics, mathtools, graphicx}
\numberwithin{equation}{section}
\linespread{1.2}
\title{Young's Experiment}
\author{To be added}
\begin{document}
\maketitle
\section{Introduction}
Sir Isaac Newton proposed that light is a stream of tiny particles emitted
from a luminous body. His influence on the world of physicists ensured that
his corpuscular theory of light was favored over the rival wave theory for 
almost two centuries. Thomas Young's double-slit experiment that the most
convincing proof that light behaves like a wave. The experiment consists in
passing light through two slits cut out in a screen and observing the 
illumination on another distant screen. When the slits are wide they create
two separate illuminated regions on the screen. When they are narrow, the
two bright regions fade away and a pattern of alternating dark and bright
fringes appears. Young reasoned that these fringes are a result of constructive
and destructive interference of light waves. Light emerging from narrow slits
diffracts into the geometrical shadow. The interference fringes are produced by
the diffracted waves. Both diffraction and interference are phenomena associated
with waves.

Young's experiment showed that when the slits are wide enough, light indeed
travels in a straight line like Newtonian particles described in Newton's first
law. This is the realm of `geometrical optics'. In this realm, light casts
sharp shadows. However, when the slits are narrow, the shadows are no longer
sharp. A penumbra appears and as the slits narrow down further the penumbra 
evolves into an interference pattern as shown in figure \ref{f1}.

\begin{figure}
\centering
\includegraphics[scale=0.5]{young}
\caption{Young's experiment. Figure from reference \cite{wiki1}}\label{f1}
\end{figure}

Young explained his experiment by proposing that the two slits act as sources
of `secondary waves' which interfere on the screen. The wavelength $\lambda$
of the two sources is identical. When the path difference of the two waves
at a point on the screen is an integral multiple of $\lambda$, the waves
interfere constructively and result in a bright band. On the other hand, when
the path difference is an odd multiple of $\lambda/2$, the waves interfere
destructively, resulting in a dark band.

Most common sources of light produce a range of wavelengths. As a result, the
interference bands produced on the screen are not as sharp as one would expect
were the light monochromatic. They also produce light waves whose phase 
changes randomly at an extremely fast rate. Therefore, if we use two separate
light sources illuminating the two slits, there is no fixed path difference
at any point on the screen and therefore no interference pattern is observed.
In order to overcome this problem, the two slits are always illuminated by 
the same source of light. This ensures that irrespective of the random phase
fluctuations at the source, the path difference is a function of the 
experiment's geometry alone.

If $d$ is the distance between the slits and $D$ is the separation between
the screen that has the slits and the screen on which we observe the 
interference pattern then the distance between two consecutive bright (or
dark) fringes is given by \cite{jenkins1937fundamentals}
\begin{equation}\label{s1e1}
\delta y = \lambda \frac{D}{d},
\end{equation}
where $\lambda$ is the wavelength of light.

Although Young's experiment demonstrated the wave nature of light in 1801, it 
was not clear in at that time what exactly was oscillating. The answer to that
question had to wait six decades until James Clerk Maxwell proposed his 
equations governing the electric and magnetic phenomena. His genius lay in 
adding a `displacement current' term to Ampere's law and realizing that the 
other known laws like Gauss law, Faraday's law, the fact that there are
no magnetic monopoles together can explain all electric and magnetic 
phenomena. An immediate consequence of his equations was his wave equation
which showed the existence of electromagnetic waves. Maxwell proposed that
light is a form of an electromagnetic wave and that gave a theoretical
explanation to Young's experiment.

\section{Maxwell equations}
Maxwell equations, in SI units, are
\begin{eqnarray}
\div\vb{D} &=& \rho_f \label{s2e1} \\
\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s2e2} \\
\div\vb{B} &=& 0 \label{s2e3} \\
\curl\vb{H} &=& \vb{J}_f + \pdv{\vb{D}}{t}, \label{s2e4}
\end{eqnarray}
where $\vb{E}$ is the electric field, $\vb{B}$ is the magnetic field,
$\vb{D}$ is the electric displacement and $\vb{H}$ is the auxilary magnetic
vector. $\rho_f$ is the density of free charges and $\vb{J}_f$ is the 
density of free current. Equation \eqref{s2e1} is Gauss' law. Equation
\eqref{s2e2} is Faraday's law. Equation \eqref{s2e3} is the mathematical
expression of the fact that there are no magnetic monopoles. Equation
\eqref{s2e4} is Ampere's law. The term $\partial\vb{D}/\partial t$ in equation
\eqref{s2e4} is called the displacement current. Adding it to Ampere's law
was Maxwell's correction to the law. We will show that this term is critical
in getting the equation of electromagnetic waves.

The electric displacement and the electric field
are related by the relation
\begin{equation}\label{s2e5}
\vb{D} = \epsilon_0\vb{E} + \vb{P},
\end{equation}
where $\epsilon_0$ is the permittivity of free space and $\vb{P}$ is the
polarization of the medium. It is the number of electric dipoles per unit
volume of the material. The volume of the material is chosen to be small at
a macroscopic scale but large enough at a microscopic scale so that it contains
a very large number of molecules of the material. The magnetic field $\vb{B}$
is related to the field $\vb{H}$ by the relation,
\begin{equation}\label{s2e6}
\vb{H} = \mu_0^{-1}(\vb{B} - \vb{M}),
\end{equation}
where $\mu_0$ is the permeability of free space and $\vb{M}$ is the 
magnetization of the medium. It is the number of magnetic dipoles per unit
volume of the material. The volume of the material is chosen in a manner 
similar to the one used to define $\vb{P}$.

There are no electric or magnetic dipoles in vacuum so that equations
\eqref{s2e5} and \eqref{s2e6} become
\begin{eqnarray}
\vb{D} &=& \epsilon_0\vb{E} \label{s2e7} \\
\vb{H} &=& \mu_0^{-1}\vb{B} \label{s2e8}
\end{eqnarray}
It is also common to write equations \eqref{s2e5} and \eqref{s2e6} as
\begin{eqnarray}
\vb{D} &=& \epsilon\vb{E} \label{s2e9} \\
\vb{H} &=& \mu^{-1}\vb{B}, \label{s2e10}
\end{eqnarray}
where $\epsilon$ and $\mu$ are permittivity and permeability of media. These
equations are applicable to linear, isotropic media. A general form of these
equations is
\begin{eqnarray}
D_i &=& \epsilon_{ij}E_j \label{s2e11} \\
B_i &=& \mu_{ij}H_j \label{s2e12}
\end{eqnarray}
where $\epsilon_{ij}$ is the permittivity tensor and $\mu_{ij}$ is the
permeability tensor of the medium.

In this article, we will consider electromagnetic waves in vacuum where there 
are no free charges or currents. Therefore $\rho_f = 0$ and $\vb{J}_f= 0$.
Maxwell equations, in this case, simplify to
\begin{eqnarray}
\div\vb{E} &=& 0 \label{s2e13} \\
\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s2e14} \\
\div\vb{B} &=& 0 \label{s2e15} \\
\curl\vb{B} &=& \mu_0\epsilon_0\pdv{\vb{E}}{t} \label{s2e16}
\end{eqnarray}
Taking the curl of equation \eqref{s2e14} and using equation \eqref{s2e16}
in the result, we get
\begin{equation}\label{s2e17}
\laplacian\vb{E} = \mu_0\epsilon_0\pdv[2]{\vb{E}}{t}.
\end{equation}
This is the equation of an electromagnetic wave in vacuum. Its speed is
\begin{equation}\label{s2e18}
c_0 = \frac{1}{\sqrt{\epsilon_0\mu_0}}.
\end{equation}
If we take the curl of equation \eqref{s2e16} and use equation \eqref{s2e14},
we get
\begin{equation}\label{s2e19}
\laplacian\vb{B} = \mu_0\epsilon_0\pdv[2]{\vb{B}}{t}.
\end{equation}
Thus, the magnetic field $\vb{B}$ satisfies an identical equation. The key
physical idea governing the wave equations \eqref{s2e17} and \eqref{s2e19}
is that a changing magnetic field produces an electric field (Faraday's law
of equation \eqref{s2e14}) and a changing electric field produces a 
magnetic field (Ampere's law with Maxwell's correction of equation 
\eqref{s2e16}). These two laws together ensure that the electric and magnetic
waves sustain each other and produce an electromagnetic wave.

\section{Electromagnetic waves}\label{s3}
\section{Finite Difference Time Domain Method}\label{s4}

\bibliographystyle{plain}
\bibliography{em}
\end{document}

### 2020-11-04 fdtd/em.tex
WIP 4Nov2020

@@
 \numberwithin{equation}{section}
+
+\newcommand{\opr}{\prime}
 \linespread{1.2}
@@
 
-\section{Maxwell equations}
+\section{Maxwell equations}\label{s2}
 Maxwell equations, in SI units, are
@@
 \section{Electromagnetic waves}\label{s3}
-\section{Finite Difference Time Domain Method}\label{s4}
+Electromagnetic waves in free space are transverse. If we express the solutions
+of equations \eqref{s2e17} and \eqref{s2e19} as
+\begin{eqnarray}
+\vb{E} &=& \vb{E}_0\exp(i(\vb{k}\cdot\vb{r} - \omega t)) \label{s3e1} \\
+\vb{B} &=& \vb{B}_0\exp(i(\vb{k}\cdot\vb{r} - \omega t)) \label{s3e2} 
+\end{eqnarray}
+then equations \eqref{s2e13} and \eqref{s2e15} give
+\begin{eqnarray}
+\vb{k}\cdot\vb{E}_0 &=& 0 \label{s3e3} \\
+\vb{k}\cdot\vb{B}_0 &=& 0. \label{s3e4}
+\end{eqnarray}
+from which we readily get $\vb{k}\cdot\vb{E} = 0$ and $\vb{k}\cdot\vb{B} = 0$.
+Thus, the electric and magnetic fields are perpendicular to the direction of
+propagation in free space. When electromagnetic waves are confined to wave
+guides this is not the case \cite{griffiths2005introduction}. A wave traveling
+through a hollow guide can have either the electric field transverse to the
+direction of propagation or the magnetic field transverse to direction of
+propagation, but not both. The former case is called `Transverse Electric' 
+(TE) mode and the latter case is called `transverse magnetic' (TM) mode. A 
+coaxial cable can support waves with both electric and magnetic fields 
+transverse to the direction of propagation. These modes of propagation are
+called `Transverse Electric (and) Magnetic' (TEM) modes.
+
+\section{Electromagnetic waves in matter}\label{s4}
+We can easily extend the treatment of section \ref{s2} to the case of 
+electromagnetic waves in matter for which the linear relationships of
+equations \eqref{s2e9} and \eqref{s2e10} hold good and which are devoid of 
+free charges and currents. If we use these relations instead of those given 
+by equations \eqref{s2e7} and \eqref{s2e8}, we get the wave equations of the 
+form
+\begin{eqnarray}
+\laplacian\vb{E} &=& \mu\epsilon\pdv[2]{\vb{E}}{t} \label{s4e1} \\
+\laplacian\vb{B} &=& \mu\epsilon\pdv[2]{\vb{B}}{t}. \label{s4e2}
+\end{eqnarray}
+The speed of electromagnetic waves is 
+\begin{equation}\label{s4e3}
+v = \frac{1}{\sqrt{\mu\epsilon}}.
+\end{equation}
+The ratio of the speed of electromagnetic waves in vacuum to that in materials
+is the refractive index
+\begin{equation}\label{s4e4}
+n = \sqrt{\frac{\epsilon}{\epsilon_0}}\sqrt{\frac{\mu}{\mu_0}}.
+\end{equation}
+The permittivity and permeability of materials depends on the frequency of
+electromagnetic fields. As a result, the refractive index is not a constant
+but a frequency dependent quantity. The dependence of $n$ on the frequency
+$\omega$ gives rise to dispersion of electromagnetic waves in media.
+
+Let us now extend our analysis of the propagation of electromagnetic waves
+in media which have free charges and which permit free currents. The current
+density in response to the electric field is
+\begin{equation}\label{s4e5}
+\vb{J}_f = \sigma\vb{E}.
+\end{equation}
+When we use this relation in the equation of continuity of charges,
+\begin{equation}\label{s4e6}
+\pdv{\rho_f}{t} + \div\vb{J}_f = 0,
+\end{equation}
+we get
+\begin{equation}\label{s4e7}
+\pdv{\rho_f}{t} + \frac{\sigma}{\epsilon}\rho_f = 0,
+\end{equation}
+whose solution is
+\begin{equation}\label{s4e8}
+\rho_f(t) = \rho_f(0)\exp\left(-\frac{\sigma}{\epsilon}t\right).
+\end{equation}
+Thus, the density of free charges decays exponentially, the rate of decay being
+faster for materials with higher conductivity. This behavior allows us to
+consider $\rho_f(t)$ to be like a transient and ignore it from our analysis.
+That is, we consider the steady state analysis when $\rho_f(t)$ is almost
+zero. Maxwell equations in the steady state, after using equation \eqref{s4e5}
+become
+\begin{eqnarray}
+\div\vb{E} &=& 0 \label{s4e9} \\
+\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s4e10} \\
+\div\vb{B} &=& 0 \label{s4e11} \\
+\curl\vb{B} &=& \mu\sigma\vb{E} + \mu\epsilon\pdv{\vb{E}}{t}. \label{s4e12}
+\end{eqnarray}
+Carrying out the same manipulations as we did in the case of vacuum, we get
+the wave equations
+\begin{eqnarray}
+\laplacian\vb{E} &=& \mu\epsilon\pdv[2]{\vb{E}}{t} + \mu\sigma\pdv{\vb{E}}{t}
+\label{s4e13} \\
+\laplacian\vb{B} &=& \mu\epsilon\pdv[2]{\vb{B}}{t} + \mu\sigma\pdv{\vb{B}}{t}.
+\label{s4e14} 
+\end{eqnarray}
+These are inhomogeneous wave equations. They admit plane wave solutions of
+the form
+\begin{eqnarray}
+\vb{E}(\vb{r}, t) &=& \vb{E}_0\exp(i(\vb{k}\cdot\vb{r} - \omega t))
+\label{s4e15} \\
+\vb{B}(\vb{r}, t) &=& \vb{B}_0\exp(i(\vb{k}\cdot\vb{r} - \omega t)),
+\label{s4e16} 
+\end{eqnarray}
+where it is understood that only the real part of the quantities have a 
+physical significance. The wave number $k$ is also a complex quantity
+\begin{equation}\label{s4e17}
+k^2 = \mu\epsilon\omega^2 + i\mu\sigma\omega.
+\end{equation}
+One can write
+\begin{equation}\label{s4e18}
+k = k_+ + ik_-,
+\end{equation}
+where
+\begin{equation}\label{s4e19}
+k_\pm = \omega\frac{\epsilon\mu}{2}
+\left(\sqrt{1 + \frac{\sigma^2}{\epsilon^2\omega^2}} \pm 1\right)^{1/2}
+\end{equation}
+The imaginary part $k_-$ of the wave number leads to an attenuation of the
+electromagnetic waves in conducting media. The extent of attenuation is 
+measured by the skin depth
+\begin{equation}\label{s4e20}
+d = \frac{1}{k_-},
+\end{equation}
+It is the depth in the material at which the amplitude of an incident
+electromagnetic wave decays by a factor of $1/e$.
+
+\section{Finite Difference Time Domain Method}\label{s5}
+Finite difference methods are a way to solve differential equations numerically
+by writing them as difference equations. The derivative of a function $f$
+with respect to $x$ at a point $a$ can be approximated by
+\begin{equation}\label{s5e1}
+f^\opr(a) \approx \frac{f(a+h) - f(a)}{h}
+\end{equation}
+when $h$ is sufficiently small. The quantity $f(a+h) - f(a)$ is often called
+the forward difference. A related quantity $f(a) - f(a - h)$ is called the
+backward difference. The central difference is defined as
+\[
+f\left(a + \frac{h}{2}\right) - f\left(a - \frac{h}{2}\right).
+\]
+It is often convenient to approximate the derivative in equation \eqref{s5e1}
+in terms of the central difference as
+\begin{equation}\label{s5e2}
+f^\opr(a) \approx \frac{1}{h}
+\left(f\left(a + \frac{h}{2}\right) - f\left(a - \frac{h}{2}\right)\right).
+\end{equation}
+Once a differential equation is approximated by a difference equation, we can 
+cast it as an equation involving matrices. There are excellent numerical
+libraries to solve matrix equations which can be used to get an approximate
+numerical solution of the differential equation.
+
+The finite-difference time-domain method (FDTD) was developed by Kane Yee
+\cite{yee1966numerical} so solve Maxwell equations numerically. The domain
+in which the solution has to be found is divided into a grid of points at
+which the derivatives of $\vb{E}$ and $\vb{B}$ are evaluated using the
+central difference approximation of \eqref{s5e2}. The grid spans both space
+and time variables. The `time-domain' refers to the fact that the fields are
+analyzed as functions of time as opposed to taking their Fourier transforms
+and writing them as functions of frequency. Since the method involves
+functions of time and their derivatives with respect to time (and space)
+it works for a wide range of frequencies. In fact, the frequency is one of
+the parameters of an FDTD algorithm.
+
+We will illustrate FDTD for a simple one-dimensional problem. We observed
+in section \ref{s2} that an electromagnetic wave is sustained because a 
+changing electric field produces a magnetic field and vice versa. For sake
+of simplicity, let us consider the progagation of electromagnetic waves in
+vacuum. The pair of equations from which the wave equation emerges is
+\begin{eqnarray}
+\curl\vb{H} &=& \epsilon_0\pdv{\vb{E}}{t} \label{s5e3} \\
+\curl\vb{E} &=& -\mu_0\pdv{\vb{H}}{t} \label{s5e4}
+\end{eqnarray}
+Let us assume that the wave propagates in the $z$-direction with the 
+electric field in the $x$-direction and the $\vb{H}$ field in the
+$y$-direction. Equations \eqref{s5e3} and \eqref{s5e4} simplify to
+\begin{eqnarray}
+\pdv{E_x}{t} &=& -\frac{1}{\epsilon_0}\pdv{H_y}{z} \label{s5e5} \\
+\pdv{H_y}{t} &=& -\frac{1}{\mu_0}\pdv{E_x}{z} \label{s5e6}
+\end{eqnarray}
+The fields $E_x$ and $H_y$ are functions of $z$ and $t$. In this case, the
+domain of the problem is a segment along the $z$-axis. We divide it in by a 
+set of grid points and evaluate the fields at each one of them at successive
+instants of time. The numerical solution of the problem thus consists in
+evaluating $E_x(z, t)$ and $H_y(z, t)$ at a discrete set of points. We can
+approximate the derivatives in equations \eqref{s5e5} and \eqref{s5e6} in
+terms of central differences and write
+\begin{eqnarray}
+\frac{E_x(k, n + 1/2) - E_x(k, n - 1/2)}{\Delta t} 
+ &=& -\frac{H_y(k + 1/2, n) - H_y(k - 1/2, n)}{\epsilon_0\Delta x}
+\label{s5e7} \\
+\frac{H_y(k + 1/2, n + 1) - H_y(k + 1/2, n)}{\Delta t} 
+&=&
+-\frac{E_x(k + 1, n + 1/2) - E_x(k - 1, n + 1/2)}{\epsilon_0\Delta x}
+\nonumber \\
+ & & \label{s5e8} 
+\end{eqnarray}
+
+Notes:
+\begin{enumerate}
+\item What is the Finite difference method?
+\item What is FDTD? Why is it useful in solving electromagnetic problems?
+\item Courant-Friedrichs-Lewy stability condition.
+\item Illustration of an example in 1d.
+\end{enumerate}
 

### 2020-11-07 fdtd/em.tex
WIP 08Nov2020

@@
 \begin{equation}\label{s2e18}
-c_0 = \frac{1}{\sqrt{\epsilon_0\mu_0}}.
+c = \frac{1}{\sqrt{\epsilon_0\mu_0}}.
 \end{equation}
@@
 approximate the derivatives in equations \eqref{s5e5} and \eqref{s5e6} in
-terms of central differences and write
+terms of central differences and write \cite{sullivan2013electromagnetic}
 \begin{eqnarray}
@@
 \end{eqnarray}
+We rewrite these equations as
+\begin{eqnarray}
+E_x\left(k ,n+\frac{1}{2}\right) &=& E_x\left(k ,
+n-\frac{1}{2}\right) \nonumber \\
+ & & -\frac{\Delta t}{\epsilon_0\Delta x}\left(\left(
+H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
+\right) \nonumber \\
+ & & \label{s5e9}
+\end{eqnarray}
+and
+\begin{eqnarray}
+H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
+\nonumber \\
+ & & -\frac{\Delta t}{\mu_0\Delta x}\left(
+E_x\left(k + 1, n + \frac{1}{2}\right) - E_x\left(k - 1, n+\frac{1}{2}\right)
+\right) \nonumber \\
+ & & \label{s5e10}
+\end{eqnarray}
+A few observations about these equations:
+\begin{itemize}
+\item $E_x$ is computed at space coordinates $0, 1, \ldots$ and time 
+coordinates $1/2, 3/2, \ldots$.
+\item $H_y$ is computed at space coordinates $1/2, 3/2, \ldots$ and time 
+coordinates $0, 1, \ldots$.
+\item The points at which $E_x$ is computed are interleaved between the points
+at which $H_y$ is computed in both space and time grid.
+\end{itemize}
+When these equations are implemented in code, we compute the fields at 
+\emph{all} points in the domain \emph{at a given instant of time}. We repeat
+this process for the desired number of time steps.
 
-Notes:
-\begin{enumerate}
-\item What is the Finite difference method?
-\item What is FDTD? Why is it useful in solving electromagnetic problems?
-\item Courant-Friedrichs-Lewy stability condition.
-\item Illustration of an example in 1d.
-\end{enumerate}
+There are a few more considerations when we implement these equations in code.
+Equations \eqref{s5e9} and \eqref{s5e10} are very similar except that the
+coefficient of the second term on the right hand side of the two equations
+have $\epsilon_0$ in the first case and $\mu_0$ in the second case. The
+values of these universal constants are $\epsilon_0 = 8.85418782\times 10^{-12}
+$ m$^{-3}$kg$^{-1}$s$^4$A$^2$ and $\mu_0=1.25663706 \times 10^{-6}$ m kg
+s$^{-2}$ A$^{-2}$. The two constants differ by $6$ orders of magnitude. In
+order to bridge this gap, we introduce a new variable
+\begin{equation}\label{s5e11}
+\tilde{E}_x = \sqrt{\frac{\epsilon_0}{\mu_0}}E_x.
+\end{equation}
+In terms of this variable, equations \eqref{s5e9} and \eqref{s5e10} become
+\begin{eqnarray}
+\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
+n-\frac{1}{2}\right) \nonumber \\
+ & & -\frac{\Delta t}{\mu_0\epsilon_0\Delta x}\left(\left(
+H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
+\right) \nonumber \\
+ & & \label{s5e12}
+\end{eqnarray}
+and
+\begin{eqnarray}
+H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
+\nonumber \\
+ & & -\frac{\Delta t}{\mu_0\epsilon_0\Delta x}\left(
+\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
+\right) \nonumber \\
+ & & \label{s5e13}
+\end{eqnarray}
+Recall from equation \eqref{s2e18} that $c = 1/\sqrt{\epsilon_0\mu_0}$ so that
+we can write the two equations above as
+\begin{equation}\label{s5e14}
+\tilde{E}_x = \sqrt{\frac{\epsilon_0}{\mu_0}}E_x.
+\end{equation}
+In terms of this variable, equations \eqref{s5e9} and \eqref{s5e10} become
+\begin{eqnarray}
+\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
+n-\frac{1}{2}\right) \nonumber \\
+ & & -\frac{c \Delta t}{\Delta x}\left(\left(
+H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
+\right) \nonumber \\
+ & & \label{s5e15}
+\end{eqnarray}
+and
+\begin{eqnarray}
+H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
+\nonumber \\
+ & & -\frac{c \Delta t}{\Delta x}\left(
+\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
+\right) \nonumber \\
+ & & \label{s5e16}
+\end{eqnarray}
+While implementing these equations in software we will have to choose certain
+values for $\Delta x$ and $\Delta t$. $\Delta x$ is chosen in such a way that
+there are at least ten points per wave. That is, if we are simulating a wave
+with wavelength $\lambda$ then \cite{sullivan2013electromagnetic}
+\begin{equation}\label{s5e17}
+\Delta x \approx \frac{\lambda}{10}.
+\end{equation}
+The time step $\Delta t$ is determined by the Courant-Friedrichs-Lewy condition
+\cite{cfl}. When we are solving a hyperbolic partial differential equation
+numerically on a grid, we should choose a time step in such a way that it is
+smaller than the time taken for the wave to travel to a neighboring point
+a distance $\Delta x$ away. Electromagnetic waves need a time $\Delta x/c$
+to travel from one grid point to the neighboring grid point. Therefore, we
+choose 
+\begin{equation}\label{s5e18}
+\Delta t = \frac{\Delta x}{2c}.
+\end{equation}
+Using this equation in \eqref{s5e15} and \eqref{s5e16} we get
+\begin{eqnarray}
+\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
+n-\frac{1}{2}\right) \nonumber \\
+ & & -\frac{1}{2}\left(\left(
+H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
+\right) \nonumber \\
+ & & \label{s5e19}
+\end{eqnarray}
+and
+\begin{eqnarray}
+H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
+\nonumber \\
+ & & -\frac{1}{2}\left(
+\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
+\right) \nonumber \\
+ & & \label{s5e20}
+\end{eqnarray}
+Equations \eqref{s5e19} and \eqref{s5e20} can be readily written in software
+to simulate propagation of electromagnetic waves in one dimension.
+
+So far we have considered the propagation of electromagnetic waves in 
+vacuum. We now consider their propagation in dielectric and lossy media. In
+the case of a dielectric medium, the first equation of the pair \eqref{s5e5}
+and \eqref{s5e6} changes to
+\begin{equation}\label{s5e21}
+\pdv{E_x}{t} = -\frac{1}{\epsilon}\pdv{H_y}{z}.
+\end{equation}
+For linear dielectric, one can write
+\begin{equation}\label{s5e22}
+\epsilon = \kappa\epsilon_0,
+\end{equation}
+where $\kappa > 1$ is the relative permittivity, also called the dielectric
+constant. If we follow the development of equations \eqref{s5e19} and
+\eqref{s5e20} from Maxwell equation with the changes indicated in equations
+\eqref{s5e21} and \eqref{s5e22}, we get
+\begin{eqnarray}
+\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
+n-\frac{1}{2\kappa}\right) \nonumber \\
+ & & -\frac{1}{2\kappa}\left(\left(
+H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
+\right) \nonumber \\
+ & & \label{s5e23}
+\end{eqnarray}
+and
+\begin{eqnarray}
+H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
+\nonumber \\
+ & & -\frac{1}{2}\left(
+\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
+\right) \nonumber \\
+ & & \label{s5e24}
+\end{eqnarray}
+Equation \eqref{s5e24} is identical with \eqref{s5e20} while equation
+\eqref{s5e23} differs from \eqref{s5e21} by the presence of dielectric 
+constant in the coefficient of the second term on the right hand side.
+
+In the case of electromagnetic waves propagating in a medium of dielectric
+constant $\kappa$ and electric conductivity $\sigma$, Ampere's law becomes
+\begin{equation}\label{s5e25}
+\curl\vb{H} = \kappa\epsilon_0\pdv{\vb{E}}{t} - \sigma\vb{E},
+\end{equation}
+where we have used Ohm's law \eqref{s4e5}. The one dimensional equations
+for the $H_y$ and $\tilde{E}_x$ (refer to \eqref{s5e11}) fields are
+\begin{eqnarray}
+\pdv{\tilde{E}_x}{t} &=& -\frac{c}{\kappa}\pdv{H_y}{z} - 
+\frac{\sigma}{\kappa\epsilon_0}\tilde{E}_x \label{s5e26} \\
+\pdv{\tilde{H}_y}{t} &=& -c\pdv{\tilde{E}_x}{z}. \label{s5e27}
+\end{eqnarray}
+If we follow the derivation of the difference equation from the differential
+equations and write 
+\[
+\tilde{E}_x(k, n) = \frac{\tilde{E}_x(k, n + 1/2) + \tilde{E}_x(k, n - 1/2)}{2}
+\]
+then we get a pair of equation of which the second one is the same as 
+\eqref{s5e24} but the first one is
+\begin{eqnarray}
+\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \frac{1-\alpha}{1+\alpha}
+\tilde{E}_x\left(k , n-\frac{1}{2\kappa}\right) \nonumber \\
+ & & -\frac{1}{2\kappa(1 + \alpha)}\left(\left(
+H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
+\right), \nonumber \\
+ & & \label{s5e28}
+\end{eqnarray}
+where \cite{sullivan2013electromagnetic}
+\begin{equation}\label{s5e29}
+\alpha = \frac{\sigma\Delta t}{2\kappa\epsilon_0}.
+\end{equation}
+
+\section{Some practical matters}\label{s6}
+\subsection{Absorbing boundary condition}
+\subsection{Perfectly matching layer}
+
+\section{A simple example of FDTD}\label{s7}
 

### 2020-11-08 fdtd/em.tex
Final draft

@@
 where $\vb{E}$ is the electric field, $\vb{B}$ is the magnetic field,
-$\vb{D}$ is the electric displacement and $\vb{H}$ is the auxilary magnetic
+$\vb{D}$ is the electric displacement and $\vb{H}$ is the auxiliary magnetic
 vector. $\rho_f$ is the density of free charges and $\vb{J}_f$ is the 
@@
 changing electric field produces a magnetic field and vice versa. For sake
-of simplicity, let us consider the progagation of electromagnetic waves in
+of simplicity, let us consider the propagation of electromagnetic waves in
 vacuum. The pair of equations from which the wave equation emerges is
@@
 \subsection{Absorbing boundary condition}
+A partial differential equation is not completely specified without its
+boundary conditions. One of the simplest boundary conditions is the 
+absorbing boundary condition. It permits the waves arriving on the boundary
+proceed beyond it without change. The choice of $\Delta t$ made in equation
+\eqref{s5e18} dictates that a wave will take two time steps to travel from
+one grid point to the next one. Therefore, we can easily implement the
+absorbing boundary condition as
+\begin{equation}\label{s6e1}
+\tilde{E}_x(0, n) = \tilde{E}_x(1, n-2).
+\end{equation}
+We do not need a separate equation for the $H_y$ field because it is 
+determined by the $\tilde{E}_x$ field.
+
 \subsection{Perfectly matching layer}
+It is easy to implement an absorbing boundary condition in one-dimensional
+problem because there is only one direction of propagation. In two or more
+dimensions it is difficult to come up with a boundary condition analogous to
+equation \eqref{s6e1} without introducing spurious reflections. In order to
+get over this problem, Berenger \cite{berenger1994perfectly} introduced an
+artificial absorbing layer outside the boundary that strongly attenuated a
+wave incident on it. However, the problem of spurious reflection does not 
+vanish. Instead of the grid boundary reflecting a wave, the absorbing layer
+starts reflecting. Berenger showed that it is possible to choose the material
+characteristics of the absorbing layer in such a way that there are no 
+reflected waves. Such an absorbing layer is called the \emph{perfectly matching
+layer}. In order to derive the equations describing it, we write Maxwell
+equations in two dimensions
+\begin{eqnarray}
+\pdv{\tilde{\vb{D}}}{t} &=& c\curl\vb{H} \label{s6e2} \\
+\pdv{\vb{H}}{t} &=& -c\curl\tilde{\vb{E}} \label{s6e3} \\
+\tilde{\vb{D}}(\omega) &=& \kappa^\ast(\omega)\tilde{\vb{E}}(\omega)
+\label{s6e4}
+\end{eqnarray}
+The complex dielectric constant in equation \eqref{s6e4} is defined as
+\begin{equation}\label{s6e5}
+\kappa^\ast = \kappa - \frac{i\sigma}{\omega\epsilon_0}.
+\end{equation}
+The field $\tilde{\vb{D}}$ is related to the electric displacement as
+\begin{equation}\label{s6e6}
+\tilde{\vb{D}} = c\vb{D} = \frac{1}{\sqrt{\mu_0\epsilon_0}}\vb{D}.
+\end{equation}
+When a wave propagating in one medium with an intrinsic impedance $\eta_1$
+enters another medium with intrinsic impedance $\eta_2$ then its gets
+reflected by a proportion
+\begin{equation}\label{s6e7}
+\Gamma = \frac{\eta_2 - \eta_1}{\eta_2 + \eta_1},
+\end{equation}
+where
+\begin{equation}\label{s6e8}
+\eta_{1,2} = \sqrt{\frac{\mu_{1,2}}{\kappa_{1,2}\epsilon_0}}.
+\end{equation}
+If we assume that the two media are non-magnetic then $\mu_1 = \mu_2 = \mu_0$.
+If we match the characteristic impedance of the absorbing layer with the
+domain then equation \eqref{s6e8} guarantees that there will be no reflection.
+However, we also want to ensure that the waves are strongly attenuated in
+the absorbing layer. In order to find the conditions to make that happen,
+we write equations \eqref{s6e2} to \eqref{s6e4} in component form. We choose 
+the TM waves so that the relevant components are $\tilde{E}_z, H_x$ and $H_y$.
+If we also take a Fourier transform of these equations we get
+\begin{eqnarray}
+i\omega\tilde{D}_x &=& c\left(\pdv{H_y}{x} - \pdv{H_x}{y}\right)\label{s6e9} \\
+\tilde{D}_z(\omega) &=& \kappa^\ast(\omega)\tilde{E}_z(\omega) \label{s6e10} \\
+i\omega H_x &=& -c\pdv{\tilde{E}_z}{y} \label{s6e11} \\
+i\omega H_y &=&  c\pdv{\tilde{E}_z}{x} \label{s6e12}
+\end{eqnarray}
+We now introduce new functions $\tilde{\kappa}, \tilde{\mu}_x$ and
+$\tilde{\mu}_y$ into equations \eqref{s6e9} to \eqref{s6e12} such that
+\begin{eqnarray}
+i\omega\tilde{\kappa}(x)\tilde{\kappa}(y)\tilde{D}_x &=& 
+c\left(\pdv{H_y}{x} - \pdv{H_x}{y}\right)\label{s6e13} \\
+\tilde{D}_z(\omega) &=& \kappa^\ast(\omega)\tilde{E}_z(\omega) \label{s6e14} \\
+i\omega H_x\tilde{\mu}_x(x)\tilde{\mu}_x(y) &=& 
+-c\pdv{\tilde{E}_z}{y} \label{s6e15} \\
+i\omega H_y\tilde{\mu}_y(x)\tilde{\mu}_y(y)&=&  
+c\pdv{\tilde{E}_z}{x} \label{s6e16}
+\end{eqnarray}
+The functions are chosen such that
+\begin{eqnarray}
+\Re(\tilde{\kappa}) &=& 1 \label{s6e17} \\
+\Re(\tilde{\mu}_x) &=& 1 \label{s6e18} \\
+\Re(\tilde{\mu}_y) &=& 1 \label{s6e19}
+\end{eqnarray}
+The functions $\tilde{\mu}_x$ and $\tilde{\mu}_y$ are written as
+\begin{equation}\label{s6e20}
+\tilde{\mu}_i = \mu_i - \frac{\tilde{\sigma}_i}{\omega\mu_0}, i = x, y
+\end{equation}
+and the imaginary part of $\tilde{\kappa}$ is written as
+\begin{equation}\label{s6e21}
+\tilde{\kappa} = -\frac{\tilde{\sigma}}{\omega\epsilon_0}.
+\end{equation}
+These conditions ensure the attenuation of waves in the absorbing layer
+\cite{sullivan2013electromagnetic}.
 
 \section{A simple example of FDTD}\label{s7}
+In this section we will illustrate a simple example of 1-d FDTD simulation
+demonstrating the propagation of a Gaussian pulse. The code is a minor 
+modification of the code available in Houle and Sullivan's book 
+\cite{sullivan2013electromagnetic}. We first import the necessary packages.
+\begin{verbatim}
+import numpy as np
+import matplotlib.pyplot as plt
+\end{verbatim}
+Next we set up the physical parameters of the problem.
+\begin{verbatim}
+ke = 200
+# x-component of normalized electric field.
+ex = np.zeros(ke)
+# y-component of magnetic field.
+hy = np.zeros(ke)
+
+kc = ke // 2  # The centre of the grid.
+t0 = 40       # The mean of the Gaussian pulse.
+spread = 12   # The standard deviation of the pulse.
+
+nsteps = 100
+\end{verbatim}
+In the next code snippet we evaluate the fields at all grid points for a 
+$100$ iterations.
+\begin{verbatim}
+for t in range(1, nsteps + 1):
+  for k in range(1, ke):
+    ex[k] +=0.5 * (hy[k - 1] - hy[k])
+
+  ex[kc] = np.exp(-0.5 * ((t0 - t)/spread)**2)
+
+  for k in range(ke - 1):
+    hy[k] += 0.5 * (ex[k] - ex[k + 1])
+\end{verbatim}
+Finally, we plot the results.
+\begin{verbatim}
+plt.figure(figsize=(8, 8))
+plt.subplot(211)
+plt.plot(ex, color='k', linewidth=1)
+plt.ylabel('E$_x$', fontsize='14')
+plt.xticks(np.arange(0, ke + 1, step=20))
+plt.xlim(0, 200)
+plt.yticks(np.arange(-1, 1.2, step=1))
+plt.ylim(-1.2, 1.2)
+plt.text(100, 0.5, 'T = {}'.format(t), horizontalalignment='center')
+plt.subplot(212)
+plt.plot(hy, color='k', linewidth=1)
+plt.ylabel('H$_y$', fontsize='14')
+plt.xlabel('FDTD cells')
+plt.xticks(np.arange(0, ke + 1, step=20))
+plt.xlim(0, 200)
+plt.yticks(np.arange(-1, 1.2, step=1))
+plt.ylim(-1.2, 1.2)
+plt.subplots_adjust(bottom=0.2, hspace=0.45)
+plt.show()
+\end{verbatim}
+The results of the simulation are
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{fdtd-1}
+\caption{Propagation of a Gaussian pulse}\label{f2}
+\end{figure}
 

### 2020-11-10 fdtd/em.tex
WIP 10Nov2020

@@
 \end{verbatim}
-The results of the simulation are
+The results of the simulation are shown in figure \ref{f2}
 \begin{figure}
@@
 
+\section{An example of a 2d simulation}\label{s8}
+We will extend the code in the previous section to two dimensions. Once again,
+the code is a minor modification of the code in Houle in Sullivan's book
+\cite{sullivan2013electromagnetic}.
+
+We first begin by importing the necessary packages.
+\begin{verbatim}
+import numpy as np
+import matplotlib.pyplot as plt
+\end{verbatim}
+
+Next we set up the physical parameters.
+\begin{verbatim}
+ddx = 0.01        # Cell size.
+c = 3e8           # Speed of light in vacuum.
+dt = ddx / 2*c    # Time step size.
+epsz = 8.854e-12  # Permittivity of free space.
+t0 = 20           # Mean magnitude of the pulse.
+spread = 6        # Spread of the pulse.
+\end{verbatim}
+
+The simulation parameters are:
+\begin{verbatim}
+ie =  60
+je =  60
+ic = ie // 2
+jc =  je // 2
+ez =  np.zeros((ie,  je))
+dz =  np.zeros((ie,  je))
+hx =  np.zeros((ie,  je))
+hy =  np.zeros((ie,  je))
+gaz = np.ones((ie, je))
+nsteps = 50
+
+# Dictionary to track data at various times.
+plotting_points = [
+  {'label': 'a', 'num_steps': 20, 'data_to_plot': None},
+  {'label': 'b', 'num_steps': 30, 'data_to_plot': None},
+  {'label': 'c', 'num_steps': 40, 'data_to_plot': None},
+  {'label': 'd', 'num_steps': 50, 'data_to_plot': None},
+]
+\end{verbatim}
+
+The code to compute the electric and magnetic fields is:
+\begin{verbatim}
+def compute_fields(time_step):
+  for j in range(1, je):
+    for i in range(1, ie):
+      dz[i, j] = dz[i, j] + 0.5 * (hy[i, j] - hy[i - 1, j] - hx[i, j] + hx[i, j - 1])
+
+  dz[ic, jc] = np.exp(-0.5 * ((t0 - time_step) / spread) ** 2)
+
+  for j in range(1, je):
+    for i in range(1, ie):
+      ez[i, j] = gaz[i, j] * dz[i, j]
+
+  for j in range(je - 1):
+    for i in range(ie - 1):
+      hx[i, j] = hx[i, j] + 0.5 * (ez[i, j] - ez[i, j + 1])
+
+  for j in range(je - 1):
+    for i in range(ie - 1):
+      hy[i, j] = hy[i, j] + 0.5 * (ez[i + 1, j] - ez[i, j])
+
+  for plotting_point in plotting_points:
+    if time_step == plotting_point['num_steps']:
+      plotting_point['data_to_plot'] = np.copy(ez)
+\end{verbatim}
+
+We compute the fields for a certain number of time steps, defined in the
+simulation parameters setup.
+\begin{verbatim}
+for t in range(1, nsteps + 1):
+  compute_fields(t)
+\end{verbatim}
+
+Finally, we plot the results.
+\begin{verbatim}
+plt.rcParams['font.size'] = 12
+plt.rcParams['grid.color'] = 'gray'
+plt.rcParams['grid.linestyle'] = 'dotted'
+fig = plt.figure(figsize=(8, 7))
+X, Y = np.meshgrid(range(je), range(ie))
+
+def plot_e_field(ax, data, timestep, label):
+    """3d Plot of E field at a single time step"""
+    ax.set_zlim(0, 1)
+    ax.view_init(elev=20., azim=45)
+    ax.plot_surface(X, Y, data[:, :], rstride=1, cstride=1, 
+                    color='white', edgecolor='black', linewidth=.25)
+    ax.zaxis.set_rotate_label(False)
+    ax.set_zlabel(r' $E_{Z}$', rotation=90, labelpad=10, fontsize=14)
+    ax.set_zticks([0, 0.5, 1])
+    ax.set_xlabel('cm')
+    ax.set_ylabel('cm')
+    ax.set_xticks(np.arange(0, 61, step=20))
+    ax.set_yticks(np.arange(0, 61, step=20))
+    ax.text2D(0.6, 0.7, f"T = {timestep}", transform=ax.transAxes)
+
+    ax.xaxis.pane.fill = ax.yaxis.pane.fill = ax.zaxis.pane.fill = False
+    plt.gca().patch.set_facecolor('white')
+    ax.text2D(-0.2, 0.8, "({})".format(label), transform=ax.transAxes)
+    ax.dist = 11
+
+for subplot_num, plotting_point in enumerate(plotting_points):
+    ax = fig.add_subplot(2, 2, subplot_num + 1, projection='3d')
+    plot_e_field(ax, plotting_point['data_to_plot'],
+    plotting_point['num_steps'],
+    plotting_point['label'])
+    plt.subplots_adjust(bottom=0.05, left=0.10, hspace=0.05)
+
+plt.show()
+\end{verbatim}
+The results of the simulation are shown in figure \ref{f3}.
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{fdtd-2}
+\caption{Propagation of a Gaussian pulse in 2d}\label{f3}
+\end{figure}
 \bibliographystyle{plain}

### 2020-11-11 fdtd/em.tex
Final draft

@@
 dt = ddx / 2*c    # Time step size.
-epsz = 8.854e-12  # Permittivity of free space.
 t0 = 20           # Mean magnitude of the pulse.
@@
 \end{figure}
+
+\section{Simulation of Young's experiment}\label{s9}
+The code snippets in this section show how to simulate Young's experiment
+using FDTD. This one is also a minor modification of the code in Houle in 
+Sullivan's book \cite{sullivan2013electromagnetic}. We start with importing
+all packages needed for the simulation. The first three are needed for the
+numerical computations and generating the intensity plot. The remaining ones
+are needed to store the plots as image files.
+\begin{verbatim}
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as patches
+
+import glob
+from PIL import Image
+import re
+import os
+\end{verbatim}
+
+We define the physical constants used in the problem.
+\begin{verbatim}
+c = 3e8           # The speed of light in vacuum.
+epsr = 1e4        # The relative dielectric permittivity of the screen.
+sigma = 0.3       # The conductivity of the screen.
+epsz = 8.854e-12  # Permittivity of free space in SI units.
+\end{verbatim}
+
+The geometrical parameters of the problem are:
+\begin{verbatim}
+# The extent of the domain. **You can alter these parameters.**
+ie = 200 # vertical limit.
+je = 200 # horizontal limit.
+
+# The central points in either directions.
+ic = ie // 2 - 1
+jc = je // 2 - 1
+
+# The location of the perfectly matching layer.
+ia = 10           # starting vertical position after PML.
+ib = ie - ia - 1  # ending horizontal position before PML.
+ja = 10           # starting horizontal position after PML.
+jb = je - ja - 1  # ending horizontal position before PML.
+
+# Slit geometry. **You can alter these parameters.**
+slit_width = 8 # width of each slit (vertical measure).
+slit_sep = 20 # separation between both slits (vertical measure).
+screen_width = 8 # width of the screen (horizontal measure).
+\end{verbatim}
+
+We now allocate space to store the two dimensional field vectors. We simulate
+a TM wave propagating in vacuum until it hits the screen with two slits in 
+it.
+\begin{verbatim}
+ez = np.zeros((ie, je)) # electric field.
+dz = np.zeros((ie, je)) # electric displacement.
+hx = np.zeros((ie, je)) # x-component of H field.
+hy = np.zeros((ie, je)) # y-component of H field.
+iz = np.zeros((ie, je))
+ihx = np.zeros((ie, je))
+ihy = np.zeros((ie, je))
+
+ez_inc = np.zeros(je) # incident electric field
+hx_inc = np.zeros(je) # x-component of incident H field
+\end{verbatim}
+
+The simulation parameters and certain auxiliary variables are defined in the
+code snippet below.
+\begin{verbatim}
+ddx = 0.01 # Cell size in m
+dt = ddx / 2*c
+gaz = np.ones((ie, je))
+gbz = np.zeros((ie, je))
+
+boundary_low = [0, 0]
+boundary_high = [0, 0]
+# Time for which the source is generated. 
+# **You can alter this parameter.**
+t0 = 10       
+# Number of time steps. 
+# **You can alter this parameter.**
+nsteps = 500
+
+# Vectors used for PML.
+gi2 = np.ones(ie)
+gi3 = np.ones(ie)
+fi1 = np.zeros(ie)
+fi2 = np.ones(ie)
+fi3 = np.ones(ie)
+
+gj2 = np.ones(je)
+gj3 = np.ones(je)
+fj1 = np.zeros(je)
+fj2 = np.ones(je)
+fj3 = np.ones(je)
+
+# Dictionary to keep track of desired points for plotting.
+plotting_points = []
+for num_steps in range(1,nsteps+1):
+    plotting_points.append({'label': 'a',
+                            'num_steps': num_steps,
+                            'data_to_plot': None})
+\end{verbatim}
+
+We now define the functions that do the FDTD computations. The first one
+creates the dielectric profile of the domain. It initializes the portion of the
+domain that has the screen with the screens $\sigma$ and $\kappa$.
+\begin{verbatim}
+def build_dielectric_profile():
+  for j in range(ja, jb): # vertical range
+      for i in range(ia, ib): # horizontal range
+          if (
+              ((i > 0 and i < ic - slit_width/2-slit_sep/2) or
+              (i > ic - slit_sep/2 + slit_width/2 and 
+			   i < ic + slit_sep/2 - slit_width/2) or
+              (i > ic + slit_width/2+slit_sep/2 and i < ie))
+              and (j > jc-screen_width/2 and j < jc+screen_width/2)
+              ): # If position is inside the dielectric
+
+              gaz[i, j] = 1 / (epsr + (sigma * dt / epsz))
+              gbz[i, j] = (sigma * dt / epsz)
+\end{verbatim}
+
+The next function implements the absorbing boundary condition with a perfectly
+matched layer.
+\begin{verbatim}
+def create_pml():
+  npml = ia + 1 # number of points for the PML
+  for n in range(npml):
+      xnum = npml - n
+      xd = npml
+      xxn = xnum / xd
+      xn = 0.33 * xxn ** 3
+
+      gi2[n] = 1 / (1 + xn)
+      gi2[ie - 1 - n] = 1 / (1 + xn)
+      gi3[n] = (1 - xn) / (1 + xn)
+      gi3[ie - 1 - n] = (1 - xn) / (1 + xn)
+
+      gj2[n] = 1 / (1 + xn)
+      gj2[je - 1 - n] = 1 / (1 + xn)
+      gj3[n] = (1 - xn) / (1 + xn)
+      gj3[je - 1 - n] = (1 - xn) / (1 + xn)
+
+      xxn = (xnum - 0.5) / xd
+      xn = 0.33 * xxn ** 3
+
+      fi1[n] = xn
+      fi1[ie - 2 - n] = xn
+      fi2[n] = 1 / (1 + xn)
+      fi2[ie - 2 - n] = 1 / (1 + xn)
+      fi3[n] = (1 - xn) / (1 + xn)
+      fi3[ie - 2 - n] = (1 - xn) / (1 + xn)
+
+      fj1[n] = xn
+      fj1[je - 2 - n] = xn
+      fj2[n] = 1 / (1 + xn)
+      fj2[je - 2 - n] = 1 / (1 + xn)
+      fj3[n] = (1 - xn) / (1 + xn)
+      fj3[je - 2 - n] = (1 - xn) / (1 + xn)
+\end{verbatim}
+
+The fields are computed in
+\begin{verbatim}
+def compute_fields(time_step):
+  # Print progress.
+  if time_step%10 == 0:
+      print("\r",100*time_step/nsteps,'%', end="")
+
+  for j in range(1, je):
+      ez_inc[j] = ez_inc[j] + 0.5 * (hx_inc[j - 1] - hx_inc[j])
+
+  # Absorbing Boundary Conditions
+  ez_inc[0] = boundary_low.pop(0)
+  boundary_low.append(ez_inc[1])
+  ez_inc[je - 1] = boundary_high.pop(0)
+  boundary_high.append(ez_inc[je - 2])
+
+  for j in range(1, je):
+      for i in range(1, ie):
+          dz[i, j] = gi3[i] * gj3[j] * dz[i, j] + \
+                      gi2[i] * gj2[j] * 0.5 * \
+                      (hy[i, j] - hy[i - 1, j] -
+                      hx[i, j] + hx[i, j - 1])
+
+  # Source of electromagnetic waves.
+  ez_inc[1] = (np.sin(2*np.pi*time_step/(2*t0)))**2
+
+  for i in range(ia, ib + 1):
+      dz[i, ja] = dz[i, ja] + 0.5 * hx_inc[ja - 1]
+      dz[i, jb] = dz[i, jb] - 0.5 * hx_inc[jb]
+
+  for j in range(0, je):
+      for i in range(0, ie):
+          ez[i, j] = gaz[i, j] * (dz[i, j] - iz[i, j])
+          iz[i, j] = iz[i, j] + gbz[i, j] * ez[i, j]
+
+  for j in range(0, je - 1):
+      hx_inc[j] = hx_inc[j] + 0.5 * (ez_inc[j] - ez_inc[j + 1])
+
+  for j in range(0, je - 1):
+      for i in range(0, ie - 1):
+          curl_e = ez[i, j] - ez[i, j + 1]
+          ihx[i, j] = ihx[i, j] + curl_e
+          hx[i, j] = fj3[j] * hx[i, j] + fj2[j] * \
+                      (0.5 * curl_e + fi1[i] * ihx[i, j])
+
+  for i in range(ia, ib + 1):
+      hx[i, ja - 1] = hx[i, ja - 1] + 0.5 * ez_inc[ja]
+      hx[i, jb] = hx[i, jb] - 0.5 * ez_inc[jb]
+
+  for j in range(0, je):
+      for i in range(0, ie - 1):
+          curl_e = ez[i, j] - ez[i + 1, j]
+          ihy[i, j] = ihy[i, j] + curl_e
+          hy[i, j] = fi3[i] * hy[i, j] - fi2[i] * \
+                      (0.5 * curl_e + fj1[j] * ihy[i, j])
+
+  for j in range(ja, jb + 1):
+      hy[ia - 1, j] = hy[ia - 1, j] - 0.5 * ez_inc[j]
+      hy[ib, j] = hy[ib, j] + 0.5 * ez_inc[j]
+
+  # Save data at certain points for later plotting
+  for plotting_point in plotting_points:
+      if time_step == plotting_point['num_steps']:
+          plotting_point['data_to_plot'] = np.copy(ez)
+\end{verbatim}
+
+The main simulation is carried out by calling the functions in the previous
+snippets.
+\begin{verbatim}
+build_dielectric_profile()
+create_pml()
+
+print('Starting FDTD...')
+for time_step in range(1, nsteps + 1):
+  compute_fields(time_step)
+\end{verbatim}
+
+The final snippet creates the intensity profile and saves intermediate results
+in image files.
+\begin{verbatim}
+plt.rcParams['font.size'] = 12
+plt.rcParams['grid.color'] = 'gray'
+plt.rcParams['grid.linestyle'] = 'dotted'
+
+print('Starting to save images to make gif...')
+for subplot_num, plotting_point in enumerate(plotting_points):
+    if subplot_num%10 == 0:
+        print("\r",100*subplot_num/nsteps,'%', end="")
+
+    fig = plt.figure()
+    ax = fig.add_subplot(1, 1, 1)
+
+    # Plot the dielectric in white color
+    rect1 = patches.Rectangle((jc-screen_width/2,0),
+                              screen_width,
+                              ic-slit_sep/2-slit_width/2,
+                              linewidth=1,
+                              edgecolor='#ffffff',
+                              facecolor='#ffffff')
+    rect2 = patches.Rectangle((jc-screen_width/2,
+                               ic-slit_sep/2+slit_width/2),
+                              screen_width,
+                              slit_sep-slit_width,
+                              linewidth=1,
+                              edgecolor='#ffffff',
+                              facecolor='#ffffff')
+    rect3 = patches.Rectangle((jc-screen_width/2,
+                               ic+slit_sep/2+slit_width/2),
+                              screen_width,
+                              ic-slit_sep/2-slit_width/2,
+                              linewidth=1,
+                              edgecolor='#ffffff',
+                              facecolor='#ffffff')
+    ax.add_patch(rect1)
+    ax.add_patch(rect2)
+    ax.add_patch(rect3)
+
+    plt.imshow(plotting_point['data_to_plot'],
+               interpolation='bilinear',
+               cmap='gist_heat')
+    plt.savefig('figs_%s.png'%str(subplot_num+1))
+    plt.close();
+
+def tryint(s):
+    try:
+        return int(s)
+    except:
+        return s
+
+def alphanum_key(s):
+    """ Turn a string into a list of string and number chunks.
+        "z23a" -> ["z", 23, "a"]
+    """
+    return [ tryint(c) for c in re.split('([0-9]+)', s) ]
+
+def sort_nicely(l):
+    """ Sort the given list in the way that humans expect.
+    """
+    l.sort(key=alphanum_key)
+    return l
+
+# filepaths
+if not os.path.exists('figs'):
+    print('Folder "figs" already exists. Saving/replacing images inside it...')
+    os.makedirs('figs')
+else:
+    print('Creating folder "figs". Saving images inside it...')
+
+fp_in = "figs_*.png"
+fp_out = "YoungMovie.gif"
+
+img, *imgs = [Image.open(f) for f in sort_nicely(glob.glob(fp_in))]
+# duration: time of each frame in miliseconds
+img.save(fp=fp_out, format='GIF', append_images=imgs,
+         save_all=True, duration=20, loop=0, optimize=True)
+
+x = []
+x1 = 40 # Starting vertical position
+x2 = 160 # Final vertical position
+
+# Construct array with x values, along vertical position from x1 to x2
+for i in range(x1,x2):
+    x.append(i)
+
+# Here we compute an averaged field over two times t1 and t2
+average = 0
+t1 = nsteps - 100
+t2 = nsteps - 1
+y_measure = 150 # Horizontal position where the filed is to be measured
+
+for i in range(t1,t2):
+    average += abs(plotting_points[i]['data_to_plot'][x1:x2,y_measure])
+
+# Plot the averaged field
+ignore = plt.plot(x, average/(t2-t1))
+\end{verbatim}
+
+The intensity profile generated with $500$ steps of simulation are 
+shown in figure \ref{f4}. Figure \ref{f5} shows the interference 
+pattern.
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{fdtd-3}
+\caption{Intensity profile in Young's experiment}\label{f4}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{figs_499}
+\caption{Interference pattern in Young's experiment}\label{f5}
+\end{figure}
+
 \bibliographystyle{plain}

### 2020-11-11 queueing-theory/resample.py
Corrected shebang

@@ -1,4 +1,4 @@
-#!/bin/py
+#!/bin/python
 
 # The data in Excel is at a frequency of 30 minutes. I need it at an interval
 # of 5 minutes. This calls for a resampling. The function in this file does

### 2020-11-11 queueing-theory/resample.py
WIP 11Nov2020

@@ -7,7 +7,7 @@
 def resample(T, available_freq, desired_freq):
     factor = int(available_freq/desired_freq)
 
-    T1 = [[n] * 6 for n in T]
+    T1 = [[n] * factor for n in T]
     T2 = [n for sublist in T1 for n in sublist]
 
     return T2

### 2020-11-12 fdtd/em_young.py
Rewrote the code

@@ -1,0 +1,279 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Nov 12 13:29:03 2020
+
+@author: ajoshi
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as patches # Plot rectangles
+import os
+import time
+
+# Physical parameters of the problem.
+c = 3e8                 # The speed of light in vacuum.
+kappa = 7e3             # The relative dielectric permittivity of the screen.
+sigma = 0.1             # The conductivity of the screen.
+epsilon_0 = 8.854e-12   # Permittivity of free space in SI units.
+
+# Geometrical parameters of the problem. 
+# The extent of the domain. **You can alter these parameters.**
+xe = 100 # Width.
+ye = 100 # Height.
+
+# The central points in either directions.
+yc = ye // 2 - 1
+xc = xe // 2 - 1
+
+# The location of the perfectly matching layer. 
+# PML starts after these limits.
+pml_dn = ye // 20           # starting vertical position after PML.
+pml_up = ye - pml_dn - 1    # ending vertical position before PML.
+pml_lf = xe // 20           # starting horizontal position after PML.
+pml_rt = xe - pml_lf - 1    # ending horizontal position before PML.
+
+# Slit geometry. **You can alter these parameters.**
+slit_height = 2 # height of each slit (vertical measure).
+slit_sep = 15 # separation between both slits (vertical measure).
+screen_thinkness = 2 # thickness of the screen (horizontal measure).
+
+# The field vectors.
+Ez = np.zeros((ye, xe))     # z-component of E field.
+Dz = np.zeros((ye, xe))     # z-component of D field.
+Hx = np.zeros((ye, xe))     # x-component of H field.
+Hy = np.zeros((ye, xe))     # y-component of H field.
+int_Ez = np.zeros((ye, xe)) # Time integral of E_z field.
+int_Hx = np.zeros((ye, xe)) # Time integral of H_x field.
+int_Hy = np.zeros((ye, xe)) # Time integral of H_y field.
+
[…]

### 2020-11-12 fdtd/fdtd-2d.py

@@ -1,0 +1,99 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Nov 10 14:23:02 2020
+
+@author: ajoshi
+"""
+
+import numpy as np
+from matplotlib import pyplot as plt
+
+ie =  60
+je =  60
+ic = int(ie / 2)
+jc =  int(je / 2)
+ez =  np.zeros((ie,  je))
+dz  =  np.zeros((ie,  je))
+hx  =  np.zeros((ie,  je))
+hy  =  np.zeros((ie,  je))
+ddx = 0.01 # Cell size
+dt = ddx / 6e8 # Time step size
+# Create Dielectric Profile
+epsz = 8.854e-12
+# Pulse Parameters
+t0 = 20
+spread = 6
+gaz = np.ones((ie, je))
+nsteps = 50
+
+# Dictionary to keep track of desired points for plotting
+plotting_points = [
+{'label': 'a', 'num_steps': 20, 'data_to_plot': None},
+{'label': 'b', 'num_steps': 30, 'data_to_plot': None},
+{'label': 'c', 'num_steps': 40, 'data_to_plot': None},
+{'label': 'd', 'num_steps': 50, 'data_to_plot': None},
+]
+# Main FDTD Loop
+for time_step in range(1, nsteps + 1):
+# Calculate Dz
+    for j in range(1, je):
+        for i in range(1, ie):
+            dz[i, j] = dz[i, j] + 0.5 * (hy[i, j] - hy[i - 1, j] -
+                                         hx[i, j] + hx[i, j - 1])
+# Put a Gaussian pulse in the middle
+    pulse = np.exp(-0.5 * ((t0 - time_step) / spread) ** 2)
+    dz[ic, jc] = pulse
+# Calculate the Ez field from Dz
+    for j in range(1, je):
+        for i in range(1, ie):
+            ez[i, j] = gaz[i, j] * dz[i, j]
+            # Calculate the Hx field
+    for j in range(je - 1):
+        for i in range(ie - 1):
+            hx[i, j] = hx[i, j] + 0.5 * (ez[i, j] - ez[i, j + 1])
+# Calculate the Hy field
+    for j in range(je - 1):
+        for i in range(ie - 1):
+            hy[i, j] = hy[i, j] + 0.5 * (ez[i + 1, j] - ez[i, j])
+# Save data at certain points for later plotting
+    for plotting_point in plotting_points:
[…]

### 2020-11-12 fdtd/em.tex
WIP 12-Nov-2020.

--- 2020-11-10 fdtd/em.tex
@@
 \end{figure}
+
+\section{Simulation of Young's experiment}\label{s9}
+The code snippets in this section show how to simulate Young's experiment
+using FDTD. This one is also a minor modification of the code in Houle in 
+Sullivan's book \cite{sullivan2013electromagnetic}. We start with importing
+all packages needed for the simulation. The first three are needed for the
+numerical computations and generating the intensity plot. The remaining ones
+are needed to store the plots as image files.
+\begin{verbatim}
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as patches 
+import os
+import time
+\end{verbatim}
+
+We define the physical constants used in the problem.
+\begin{verbatim}
+c = 3e8                 # The speed of light in vacuum.
+kappa = 7e3             # The relative dielectric permittivity of the screen.
+sigma = 0.1             # The conductivity of the screen.
+epsilon_0 = 8.854e-12   # Permittivity of free space in SI units.
+\end{verbatim}
+
+The geometrical parameters of the problem are:
+\begin{verbatim}
+# The extent of the domain. **You can alter these parameters.**
+xe = 100 # Width.
+ye = 100 # Height.
+
+# The central points in either directions.
+yc = ye // 2 - 1
+xc = xe // 2 - 1
+
+# The location of the perfectly matching layer. 
+# PML starts after these limits.
+pml_dn = ye // 20           # starting vertical position after PML.
+pml_up = ye - pml_dn - 1    # ending vertical position before PML.
+pml_lf = xe // 20           # starting horizontal position after PML.
+pml_rt = xe - pml_lf - 1    # ending horizontal position before PML.
+
+# Slit geometry. **You can alter these parameters.**
+slit_height = 2 # height of each slit (vertical measure).
+slit_sep = 15 # separation between both slits (vertical measure).
+screen_thinkness = 2 # thickness of the screen (horizontal measure).
+\end{verbatim}
+
+We now allocate space to store the two dimensional field vectors. We simulate
+a TM wave propagating in vacuum until it hits the screen with two slits in 
+it.
+\begin{verbatim}
+Ez = np.zeros((ye, xe))     # z-component of E field.
+Dz = np.zeros((ye, xe))     # z-component of D field.
+Hx = np.zeros((ye, xe))     # x-component of H field.
+Hy = np.zeros((ye, xe))     # y-component of H field.
+int_Ez = np.zeros((ye, xe)) # Time integral of E_z field.
+int_Hx = np.zeros((ye, xe)) # Time integral of H_x field.
+int_Hy = np.zeros((ye, xe)) # Time integral of H_y field.
+
+Ez_in = np.zeros(xe) # z component of the incident E field.
+Hx_in = np.zeros(xe) # x component of the incident H field.
+\end{verbatim}
+
+The simulation parameters and certain auxiliary variables are defined in the
+code snippet below.
+\begin{verbatim}
+cell_size = 0.01 # Cell size in m
+dt = cell_size / 2*c
+# The dielectric profile.
+dep_1 = np.ones((ye, xe))
+dep_2 = np.zeros((ye, xe))
+# The absorbing boundary conditions.
+lob = [0, 0]
+hib = [0, 0]
+
+# **You can alter this parameter.**
+t0 = 10 # Time for which the source is generated. 
+# **You can alter this parameter.**
+nsteps = 500  # Number of time steps.
+
+# Vectors used for PML.
+defy_2 = np.ones(ye)
+defy_3 = np.ones(ye)
+def2y_1 = np.zeros(ye)
+def2y_2 = np.ones(ye)
+def2y_3 = np.ones(ye)
+
+defx_2 = np.ones(xe)
+defx_3 = np.ones(xe)
+def2x_1 = np.zeros(xe)
+def2x_2 = np.ones(xe)
+def2x_3 = np.ones(xe)
+
+plotting_points = []
+for n in range(1, nsteps+1):
+    plotting_points.append({'label': 'a', 'num_steps': n, 'Ez_data': None}
+\end{verbatim}
+
+We now define the functions that do the FDTD computations. The first one
+creates the dielectric profile of the domain. It initializes the portion of the
+domain that has the screen with the screens $\sigma$ and $\kappa$.
+\begin{verbatim}
+def build_dielectric_profile():
+  for j in range(pml_lf, pml_rt): # vertical range
+      for i in range(pml_dn, pml_up): # horizontal range
+          if point_on_screen(i, j): # If position is inside the dielectric                  
+              dep_1[i, j] = 1 / (kappa + (sigma * dt / epsilon_0))
+              dep_2[i, j] = (sigma * dt / epsilon_0)
+\end{verbatim}
+
+It uses another function that checks if a point is on the screen.
+\begin{verbatim}
+def point_on_screen(i, j):
+    return (((i > 0 and i < yc - slit_height/2 - slit_sep/2) or 
+              (i > yc - slit_sep/2 + slit_height/2 and 
+               i < yc + slit_sep/2 - slit_height/2) or 
+              (i > yc + slit_height/2 + slit_sep/2 and 
+               i < ye)) 
+              and (j > xc - screen_thinkness/2 and 
+                   j < xc + screen_thinkness/2))
+\end{verbatim}
+
+The next function implements the absorbing boundary condition with a perfectly
+matched layer.
+\begin{verbatim}
+def create_pml():
+  for n in range(pml_dn + 1):    
+      # auxp stands for auxiliary parameter, as defined in eqn (3.20) of 
+      # Houle and Sullivan's book.
+      auxp = ((pml_dn + 1 - n) / (pml_dn + 1)) ** 3
+      auxp /= 3
+
+      defy_2[n] = 1 / (1 + auxp)
+      defy_2[ye - 1 - n] = 1 / (1 + auxp)
+      defy_3[n] = (1 - auxp) / (1 + auxp)
+      defy_3[ye - 1 - n] = (1 - auxp) / (1 + auxp)
+
+      defx_2[n] = 1 / (1 + auxp)
+      defx_2[xe - 1 - n] = 1 / (1 + auxp)
+      defx_3[n] = (1 - auxp) / (1 + auxp)
+      defx_3[xe - 1 - n] = (1 - auxp) / (1 + auxp)
+
+      auxp = ((pml_dn + 0.5 - n) / (pml_dn + 1)) ** 3
+      auxp /= 3
+
+      def2y_1[n] = auxp
+      def2y_1[ye - 2 - n] = auxp
+      def2y_2[n] = 1 / (1 + auxp)
+      def2y_2[ye - 2 - n] = 1 / (1 + auxp)
+      def2y_3[n] = (1 - auxp) / (1 + auxp)
+      def2y_3[ye - 2 - n] = (1 - auxp) / (1 + auxp)
+
+      def2x_1[n] = auxp
+      def2x_1[xe - 2 - n] = auxp
+      def2x_2[n] = 1 / (1 + auxp)
+      def2x_2[xe - 2 - n] = 1 / (1 + auxp)
+      def2x_3[n] = (1 - auxp) / (1 + auxp)
+      def2x_3[xe - 2 - n] = (1 - auxp) / (1 + auxp)
+\end{verbatim}
+
+The fields are computed in
+\begin{verbatim}
+def compute_fields(t): 
+  if t % 100 == 0:
+      print('|', end = '')
+  elif t % 10 == 0:
+      print('.', end = '')
+
+  for j in range(1, xe):
+      Ez_in[j] = Ez_in[j] + 0.5 * (Hx_in[j - 1] - Hx_in[j])
+
+  # Absorbing Boundary Conditions
+  Ez_in[0] = lob.pop(0)
+  lob.append(Ez_in[1])  
+  Ez_in[xe - 1] = hib.pop(0)
+  hib.append(Ez_in[xe - 2])
+
+  for j in range(1, xe):
+      for i in range(1, ye):
+          Dz[i, j] = defy_3[i] * defx_3[j] * Dz[i, j] + \
+                      defy_2[i] * defx_2[j] * 0.5 * \
+                      (Hy[i, j] - Hy[i - 1, j] -
+                      Hx[i, j] + Hx[i, j - 1])
+
+  # Source of electromagnetic waves.
+  Ez_in[1] = (np.sin(2*np.pi*t/(2*t0)))**2
+
+  for i in range(pml_dn, pml_up + 1):
+      Dz[i, pml_lf] = Dz[i, pml_lf] + 0.5 * Hx_in[pml_lf - 1]
+      Dz[i, pml_rt] = Dz[i, pml_rt] - 0.5 * Hx_in[pml_rt]
+
+  for j in range(0, xe):
+      for i in range(0, ye):
+          Ez[i, j] = dep_1[i, j] * (Dz[i, j] - int_Ez[i, j])
+          int_Ez[i, j] = int_Ez[i, j] + dep_2[i, j] * Ez[i, j]
+
+  for j in range(0, xe - 1):
+      Hx_in[j] = Hx_in[j] + 0.5 * (Ez_in[j] - Ez_in[j + 1])
+
+  for j in range(0, xe - 1):
+      for i in range(0, ye - 1):
+          curl_e = Ez[i, j] - Ez[i, j + 1]
+          int_Hx[i, j] = int_Hx[i, j] + curl_e
+          Hx[i, j] = def2x_3[j] * Hx[i, j] + def2x_2[j] * \
+                      (0.5 * curl_e + def2y_1[i] * int_Hx[i, j])
+
+  for i in range(pml_dn, pml_up + 1):
+      Hx[i, pml_lf - 1] = Hx[i, pml_lf - 1] + 0.5 * Ez_in[pml_lf]
+      Hx[i, pml_rt] = Hx[i, pml_rt] - 0.5 * Ez_in[pml_rt]
+
+  for j in range(0, xe):
+      for i in range(0, ye - 1):
+          curl_e = Ez[i, j] - Ez[i + 1, j]
+          int_Hy[i, j] = int_Hy[i, j] + curl_e
+          Hy[i, j] = def2y_3[i] * Hy[i, j] - def2y_2[i] * \
+                      (0.5 * curl_e + def2x_1[j] * int_Hy[i, j])
+
+  for j in range(pml_lf, pml_rt + 1):
+      Hy[pml_dn - 1, j] = Hy[pml_dn - 1, j] - 0.5 * Ez_in[j]
+      Hy[pml_up, j] = Hy[pml_up, j] + 0.5 * Ez_in[j]
+
+  for pp in plotting_points:
+      if t == pp['num_steps']:
+          pp['Ez_data'] = np.copy(Ez)
+\end{verbatim}
+
+Although this code is enough to carry out the simulation, we need some
+more functions to plot the results. The first one computes the average
+intensity of the field at a fixed point.
+\begin{verbatim}
+def generate_avg_plot():
+    left = int(ye * 0.2)
+    right = int(ye * 0.8)
+    x = [int(f) for f in np.linspace(left, right - 1, num = right - left)]
+
+    t1 = int(nsteps * 0.8)
+    t2 = nsteps - 1
+    y0 = int(xe * 0.75) # Horizontal position where the filed is to be measured
+
+    average = 0
+    for i in range(t1, t2):
+        average += abs(plotting_points[i]['Ez_data'][left:right, y0])
+
+    plt.plot(x, average/(t2 - t1))
+    plt.savefig('avg_intensity.png')
+    plt.close()
+\end{verbatim}
+
+The next ones saves the field profiles at various time steps.
+\begin{verbatim}
+def create_rectangle(xy, width, height):
+    clr = '#ffffff'
+    return patches.Rectangle(xy, width, height, 
+                             linewidth=1, edgecolor=clr, facecolor=clr)
+
+def save_plots():       
+    print('Saving images.')
+    if not os.path.exists('figs'):
+        os.makedirs('figs')
+
+    pwd = os.getcwd()
+    os.chdir('figs')
+
+    xy1 = (xc - screen_thinkness/2, 0)
+    xy2 = (xc - screen_thinkness/2, yc - slit_sep/2 + slit_height/2)
+    xy3 = (xc - screen_thinkness/2, yc + slit_sep/2 + slit_height/2)
+
+    for sp_no, pp in enumerate(plotting_points):  
+        if sp_no % 100 == 0:
+            print('|', end = '')
+        elif sp_no % 10 == 0:
+            print('.', end = '')
+
+        r1 = create_rectangle(xy1, 
+                              screen_thinkness, 
+                              yc - slit_sep/2 - slit_height/2)
+        r2 = create_rectangle(xy2, 
+                              screen_thinkness, 
+                              slit_sep - slit_height)
+        r3 = create_rectangle(xy3, 
+                              screen_thinkness, 
+                              yc - slit_sep/2 - slit_height/2)
+
+        ax = plt.figure().add_subplot(1, 1, 1)
+        ax.add_patch(r1)
+        ax.add_patch(r2)
+        ax.add_patch(r3)
+
+        plt.imshow(pp['Ez_data'], interpolation='bilinear', cmap='gist_heat')
+        plt.savefig(f'figs_{str(sp_no + 1)}.png')
+        plt.close()
+
+    os.chdir(pwd)
+\end{verbatim}
+We call these functions in the main function.
+\begin{verbatim}
+def main():    
+    build_dielectric_profile()
+    create_pml()
+
+    print('Computing the fields...')
+    t1 = time.time()
+    for t in range(1, nsteps + 1):
+      compute_fields(t)
+
+    t2 = time.time()
+    print()
+    print(f'Simulation took {round(t2 - t1, 2)} s.')
+    generate_avg_plot()
+
+    print('Saving the plots...')
+    t1 = time.time()           
+    save_plots()  
+    t2 = time.time()
+    print()
+    print(f'Saving images took {round(t2 - t1, 2)} s.') 
+\end{verbatim}
+
+The intensity profile generated with $500$ steps of simulation are 
+shown in figure \ref{f4}. Figure \ref{f5} shows the interference 
+pattern.
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{avg_intensity}
+\caption{Intensity profile in Young's experiment}\label{f4}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{figs_500}
+\caption{Interference pattern in Young's experiment}\label{f5}
+\end{figure}
+
 \bibliographystyle{plain}

### 2020-11-12 queueing-theory/analysis-1.R

@@ -1,0 +1,51 @@
+library(tseries)
+data <- read.csv('wait_prob_1.csv')
+
+E.series <- E.series[!is.na(data$E)]
+plot(
+  seq(1, length(E.series)),
+  E.series,
+  cex = 0.1,
+  main = 'Traffic by time',
+  xlab = 'Time',
+  ylab = 'Traffic'
+)
+# The plot indicates that the time series might be stationary. Let's check it
+# with a formal test.
+result.0 <- adf.test(E.series)
+# The p-value is < 0.05 and we can accept the alternative hypothesis that the
+# series is indeed stationary.
+
+result.1 <- acf(E.series, lag = 150)
+p <- which(result.1$acf < 0.5)[1]
+
+result.2 <- pacf(E.series)
+# The partial auto-correlation function plot shows that the function decays only
+# after lag = 14. It also drops after lag = 10. I am inclined to select lag = 10
+# because the rise at lag = 13 seems to be by chance.
+q <- which(result.2$acf < 0.05)[1]
+
+# We did not choose differences of the series. The original series itself was
+# stationary. Therefore,
+d <- 0
+
+# Let's try to fit ARIMA model.
+train.limit <- 1500 # We use first 1500 points for prediction
+result.3 <-
+  arima(E.series[1:train.limit], order = c(p, d, q), method = 'CSS')
+
+result.4 <-
+  predict(result.3, n.ahead = length(E.series) - train.limit)
+E.pred <- result.4$pred
+E.actual <- E.series[(train.limit + 1):length(E.series)]
+
+plot(
+  seq(1, length(E.pred)),
+  E.pred,
+  cex = 0.1,
+  xlab = 'Time',
+  ylab = 'Predicted E',
+  main = 'Traffic prediction'
+)
+points(seq(1, length(E.pred)), E.actual, cex = 0.1, col = 'blue')
+# The plot shows that the model is quite bad.

### 2020-11-12 queueing-theory/analysis-2.R

@@ -1,0 +1,24 @@
+# There are 96 records per day. Ideally, one should not use more than those many
+# points to predict.
+
+library(tseries)
+data <- read.csv('wait_prob_1.csv')
+
+E.series <- data$E[!is.na(data$E)]
+result.0 <- adf.test(E.series)
+result.1 <- acf(E.series, lag = 150)
+p <- which(result.1$acf < 0.05)[1]
+result.2 <- pacf(E.series)
+q <- which(result.2$acf < 0.05)[1]
+d <- 1
+train.limit <- 3 * p 
+result.3 <-
+  arima(E.series[1:train.limit], 
+        order = c(p, d, q), method = 'CSS', optim.control = list(maxit=1000))
+# There is no convergence problem now.
+
+n.ahead <- 5
+result.4 <- predict(result.3, n.ahead = n.ahead)
+E.pred <- result.4$pred
+E.actual <- E.series[(train.limit + 1):(train.limit + n.ahead)]
+plot(E.actual, E.pred, cex = 0.2)

### 2020-11-13 queueing-theory/analysis-3.R
Trying to find the distribution of number of calls and the average handling time.

@@ -1,0 +1,48 @@
+library(MASS)
+library(vcd)
+data <- read.csv('wait_prob_1.csv')
+# Use only weekday's data.
+data <- data[data$isoday_of_week < 6, colnames(data)]
+summ.ncalls <- summary(data$ncalls)
+boxplot(data$ncalls)
+
+# The outliers carry critical information. They are not a result of bad data recording.
+q1 <- summ.ncalls[2]
+q3 <- summ.ncalls[5]
+iqr <- IQR(data$ncalls)
+upper.limit <- q3 + 1.5 * iqr
+lower.limit <- max(q1 - 1.5 * iqr, 0)
+
+clean.ncalls <- data$ncalls[lower.limit < data$ncalls & data$ncalls < upper.limit]
+truehist(clean.ncalls)
+
+truehist(data$ncalls, xlab = '# calls per 5m')
+lines(density(data$ncalls))
+# We should not bother about the data with no calls. The summary does not indicate
+# serious outliers, although the box plot shows a few. 
+
+Y <- data$ncalls[data$ncalls > 0]
+boxplot(Y)
+truehist(Y)
+lines(density(Y))
+
+Y.hat <- rpois(n = length(clean.ncalls), lambda = mean(Y))
+
+chisq.test(clean.ncalls, Y.hat, simulate.p.value = TRUE)
+# p-value of 0.5187 suggests that we cannot reject the null hypothesis that the
+# two populations are similar. So this test suggests that Y is indeed Poisson
+# distributed.
+M <- max(max(clean.ncalls), max(Y.hat))
+plot(
+  clean.ncalls[order(clean.ncalls)],
+  Y.hat[order(Y.hat)],
+  xlim = c(0, M),
+  ylim = c(0, M),
+  xlab = 'Actual',
+  ylab = 'Simulated',
+  main = 'Fit',
+  cex = 0.1
+)
+abline(0, 1, col = 'blue')
+# The fit is not bad. Let's assume that the incoming call data is indeed Poisson
+# distributed.

### 2020-11-13 queueing-theory/analysis-4.R

@@ -1,0 +1,60 @@
+library(MASS)
+data <- read.csv('wait_prob_1.csv')
+# Use only weekday's data.
+data <- data[data$isoday_of_week < 6, colnames(data)]
+
+# Examination of average time.
+summ.avg.time <- summary(data$Avg.time)
+q1 <- summ.avg.time[2]
+q3 <- summ.avg.time[5]
+iqr <- IQR(data$Avg.time)
+upper.limit <- q3 + 1.5 * iqr
+lower.limit <- max(q1 - 1.5 * iqr, 0)
+
+clean.avg.time <- data$Avg.time[lower.limit < data$Avg.time & data$Avg.time < upper.limit]
+truehist(clean.avg.time)
+
+Y <- clean.avg.time[order(clean.avg.time)]
+mean.Y <- mean(Y)
+check_fit <- function(k) {
+  lambda <- k / mean.Y
+  simulated.Y <- rgamma(n = length(Y),
+                        shape = k,
+                        rate = lambda)
+  order.Y.hat <- order(simulated.Y)
+  main.title <- paste('k = ', k)
+  x.max <- max(Y)
+  y.max <- max(simulated.Y)
+  plot(
+    Y,
+    simulated.Y[order.Y.hat],
+    cex = 0.1,
+    xlim = c(0, x.max),
+    ylim = c(0, y.max),
+    xlab = 'Y',
+    ylab = 'Y.hat',
+    main = main.title
+  )
+  abline(0, 1, col = 'blue')
+}
+
+check_fit(1)
+check_fit(2)
+check_fit(3)
+check_fit(4)
+check_fit(5)
+check_fit(6)
+check_fit(7)
+check_fit(8)
+check_fit(9)
+check_fit(10)
+check_fit(11)
+check_fit(12)
+
+# I choose k = 11 as the best fit because (1) The blue line is almost diagonal.
+# This indicates that the range of data on the two axes is almost the same; (2)
+# the data is almost along the blue line except at the extremities. (3) The
+# upper extremity fits better than the lower extremity.
+# 
+# Thus average time obeys a gamma distribution with parameters shape = 12 and
+# rate = 12/mean(data$Avg.time)

### 2020-11-13 optim/amber-india/problem-1.R
Linear and nonlinear optimization problems.

@@ -1,0 +1,33 @@
+library(lpSolve)
+
+advertisement <- c(1.3e6, 6e5, 5e5)
+constraints <-
+  matrix(
+    data = c(
+      9e4, 3e4, 4e4, # Planning cost.
+      3e5, 1.5e5, 1e5, # Advertising cost.
+      -1.2e6, -0.2e6, 0, # Young people.
+      -0.5e6, -0.2e6, -0.2e6, # Middle-aged people.
+      -0, -4e4, -1.2e5, # Coupons.
+      1, 0, 0 # Available slots.s
+    ),
+    nrow = 6,
+    ncol = 3,
+    byrow = TRUE
+  )
+
+const.rhs <- c(1e6, 4e6, -5e6, -5e6, -1.49e6, 5)
+
+stopifnot(length(const.rhs) == nrow(constraints))
+
+lp.results <-
+  lp(direction = "min",
+     objective.in = advertisement,
+     const.mat = constraints,
+     const.dir = rep("<=", nrow(constraints)),
+     const.rhs = const.rhs)
+if (lp.results$status == 0) {
+  print(paste('Objective value =', round(lp.results$objval, 0)))
+  optimal.x <- lp.results$solution
+  print(paste('Solution: x1 =', optimal.x[1], ' x[2] =', optimal.x[2], ' x[3] =', optimal.x[3]))
+}

### 2020-11-13 optim/amber-india/problem-2.R

@@ -1,0 +1,63 @@
+library(nloptr)
+
+# https://stackoverflow.com/questions/31431575/minimization-with-r-nloptr-package-multiple-equality-constraints
+
+eval_f <- function(x) {
+  objective <- 0.1 * x[1] * x[1] - 1.13 * x[1] + 0.4 +
+    0.002 * x[2] * x[2] - 0.124 * x[2] - 0.14 +
+    0.00321 * x[3] * x[3] - 0.706 * x[3] + 0.09
+  gradient <-
+    c(0.2 * x[2] - 1.13, 0.004 * x[2] - 0.124, 0.00641 * x[3] - 0.706)
+  
+  return(list('objective' = objective, 'gradient' = gradient))
+}
+
+eval_g_ineq <- function(x) {
+  constraints <- rbind(
+    9e4 * x[1] + 3e4 * x[2] + 4e4 * x[3] - 1e6,
+    3e5 * x[1] + 1.5e5 * x[2] + 1e5 * x[3] - 4e6,-1.2e6 *
+      x[1] - 0.2e6 * x[2] + 5e6,
+    -0.5e6 * x[1] - 0.2e6 * x[2] - 0.2e6 * x[3] + 5e6,
+    -4e4 * x[2] - 1.2e5 * x[3] + 1.49e6,
+    x[1] - 5
+  )
+  gradient <- rbind(
+    c(9e4, 3e4, 4e4),
+    c(3e5, 1.5e5, 1e5),
+    c(-1.2e6,-0.2e6, 0),
+    c(-0.5e6,-0.2e6,-0.2e6),
+    c(0,-4e6,-1.2e5),
+    c(1, 0, 0)
+  )
+  return(list("constraints" = constraints, "jacobian" = gradient))
+}
+
+x0 <- c(4, 0, 14)
+# lower and upper bounds of control
+lb <- c(0, 0, 0)
+ub <- c(Inf, Inf, Inf)
+local_opts <- list("algorithm" = "NLOPT_LD_MMA",
+                   "xtol_rel" = 1.0e-8)
+opts <- list(
+  "algorithm" = "NLOPT_LD_AUGLAG",
+  "xtol_rel" = 1.0e-8,
+  "maxeval" = 10000,
+  "local_opts" = local_opts
+)
+res <- nloptr(
+  x0 = x0,
+  eval_f = eval_f,
+  lb = lb,
+  ub = ub,
+  eval_g_ineq = eval_g_ineq,
+  opts = opts
+)
+
[…]

### 2020-11-22 revenue-forecast/analysis-1.R
Models for segment 1.

@@ -1,0 +1,75 @@
+seg1 <- read.csv('segment-1.csv')
+seg1$month <- substr(seg1$Date, 1, 2)
+
+analyze_model <- function(formula.str, show.plot = TRUE) {
+  x <- seg1$ARPU[1:200]
+  y <- seg1$Segment.1.MRR[1:200]
+  m <- lm(as.formula(formula.str))
+  
+  newdata <- data.frame(x = seg1$ARPU[201:nrow(seg1)])
+  mrr.predicted  <- predict(m, newdata = newdata)
+  mrr.actual  <- seg1$Segment.1.MRR[201:nrow(seg1)]
+  
+  if (show.plot) {
+    plot(
+      mrr.actual,
+      mrr.predicted,
+      cex = 0.1,
+      xlab = 'actual',
+      ylab = 'predicted',
+      main = formula.str
+    )
+    abline(0, 1, col = 'blue')
+  }
+  
+  m
+}
+# Linear model.
+m <- analyze_model('y ~ x')
+summary(m)
+plot(m)
+
+
+# Linear model with cosine term.
+m <- analyze_model('y ~ x + cos(x)')
+summary(m)
+plot(m)
+
+# Quadratic model with cosine term.
+m <- analyze_model('y ~ x + I(x^2) + cos(x)')
+summary(m)
+plot(m)
+
+# From the three models fitted above, it is clear that the second model is the
+# best.
+
+# Can one find monthly MRR by aggregating the prediction?
+m <- analyze_model('y ~ x + cos(x)', show.plot = FALSE)
+newdata <- data.frame(x = seg1$ARPU[201:nrow(seg1)])
+mrr.predicted  <- predict(m, newdata = newdata)
+
+prediction <-
+  data.frame(month = seg1$month[201:nrow(seg1)], mrr.predicted = mrr.predicted)
+agg.prediction <-
+  aggregate(mrr.predicted ~ month, data = prediction, sum)
+agg.actual <-
+  aggregate(Segment.1.MRR ~ month, data = seg1[201:nrow(seg1),], sum)
+
+agg.comparison <- merge(agg.actual, agg.prediction)
+colnames(agg.comparison)[2] <- 'mrr.actual'
+agg.comparison$pct_error <-
+  (agg.comparison$mrr.predicted - agg.comparison$mrr.actual) / agg.comparison$mrr.actual
+
+agg.comparison
+# NO. The percentage error is quite high.
+
[…]

### 2020-11-22 revenue-forecast/analysis-2.R
Changed the name of a variable,

@@ -1,0 +1,75 @@
+seg <- read.csv('segment-2.csv')
+seg$month <- substr(seg$Date, 1, 2)
+
+plot(
+  seg$ARPU,
+  seg$Segment.2.MRR,
+  cex = 0.1,
+  xlab = 'ARPU',
+  ylab = 'MRR',
+  main = 'Segment 2'
+)
+
+# A linear model is not going to fit well.
+# 
+plot(
+  seg$ARPU,
+  log(seg$Segment.2.MRR),
+  cex = 0.1,
+  xlab = 'ARPU',
+  ylab = 'log(MRR)',
+  main = 'Segment 2'
+)
+
+plot(seq(1:nrow(seg)), seg$ARPU, cex = 0.1, xlab = 'time', ylab = 'ARPU', main = 'Segment 2')
+plot(seq(1:nrow(seg)), seg$Segment.2.MRR, cex = 0.1, xlab = 'time', ylab = 'MRR', main = 'Segment 2')
+
+analyze_model <- function(formula.str, show.plot = TRUE) {
+  x <- seg$ARPU[1:200]
+  y <- seg$Segment.2.MRR[1:200]
+  m <- lm(as.formula(formula.str))
+  
+  newdata <- data.frame(x = seg$ARPU[201:nrow(seg)])
+  mrr.predicted  <- predict(m, newdata = newdata)
+  mrr.actual  <- seg$Segment.2.MRR[201:nrow(seg)]
+  
+  if (show.plot) {
+    plot(
+      mrr.actual,
+      mrr.predicted,
+      cex = 0.1,
+      xlab = 'actual',
+      ylab = 'predicted',
+      main = formula.str
+    )
+    abline(0, 1, col = 'blue')
+  }
+  
+  m
+}
+
+# Quadratic model.
+m <- analyze_model('y ~ x + I(x^2)')
+summary(m)
+plot(m)
+
+# Quadratic model.
+m <- analyze_model('y ~ x + I(x^2)')
+summary(m)
+plot(m)
+
+# Quadratic + exponential.
+m <- analyze_model('y ~ x + I(x^2) + exp(x)')
+summary(m)
+plot(m)
+
+# Trignometric + exponential
+m <- analyze_model('y ~ I(x^2) +  cos(x) + exp(x)')
+summary(m)
+plot(m)
+
+# Trignometric + exponential
+m <- analyze_model('y ~ I(x^2) + cos(x) + exp(x)')
+summary(m)
+plot(m)
+

### 2020-11-22 revenue-forecast/analysis-1.R
Changed the name of a variable.

@@ -1,14 +1,14 @@
-seg1 <- read.csv('segment-1.csv')
-seg1$month <- substr(seg1$Date, 1, 2)
+seg <- read.csv('segment-1.csv')
+seg$month <- substr(seg$Date, 1, 2)
 
 analyze_model <- function(formula.str, show.plot = TRUE) {
-  x <- seg1$ARPU[1:200]
-  y <- seg1$Segment.1.MRR[1:200]
+  x <- seg$ARPU[1:200]
+  y <- seg$Segment.1.MRR[1:200]
   m <- lm(as.formula(formula.str))
   
-  newdata <- data.frame(x = seg1$ARPU[201:nrow(seg1)])
+  newdata <- data.frame(x = seg$ARPU[201:nrow(seg)])
   mrr.predicted  <- predict(m, newdata = newdata)
-  mrr.actual  <- seg1$Segment.1.MRR[201:nrow(seg1)]
+  mrr.actual  <- seg$Segment.1.MRR[201:nrow(seg)]
   
   if (show.plot) {
     plot(
@@ -45,15 +45,15 @@ plot(m)
 
 # Can one find monthly MRR by aggregating the prediction?
 m <- analyze_model('y ~ x + cos(x)', show.plot = FALSE)
-newdata <- data.frame(x = seg1$ARPU[201:nrow(seg1)])
+newdata <- data.frame(x = seg$ARPU[201:nrow(seg)])
 mrr.predicted  <- predict(m, newdata = newdata)
 
 prediction <-
-  data.frame(month = seg1$month[201:nrow(seg1)], mrr.predicted = mrr.predicted)
+  data.frame(month = seg$month[201:nrow(seg)], mrr.predicted = mrr.predicted)
 agg.prediction <-
   aggregate(mrr.predicted ~ month, data = prediction, sum)
 agg.actual <-
-  aggregate(Segment.1.MRR ~ month, data = seg1[201:nrow(seg1),], sum)
+  aggregate(Segment.1.MRR ~ month, data = seg[201:nrow(seg), ], sum)
 
 agg.comparison <- merge(agg.actual, agg.prediction)
 colnames(agg.comparison)[2] <- 'mrr.actual'
@@ -64,12 +64,17 @@ agg.comparison
 # NO. The percentage error is quite high.
 
 # Build a separate model for monthly aggregates.
-seg1.agg <- aggregate(cbind(Segment.1.MRR, ARPU) ~ month, data = seg1, sum)
[…]

### 2020-11-22 revenue-forecast/analysis-2.R
LOESS performs best.

@@ -11,65 +11,102 @@ plot(
 )
 
 # A linear model is not going to fit well.
-# 
+#
 plot(
-  seg$ARPU,
-  log(seg$Segment.2.MRR),
+  seg$Total.paid.customers,
+  seg$Segment.2.MRR,
   cex = 0.1,
-  xlab = 'ARPU',
-  ylab = 'log(MRR)',
+  xlab = 'Total paid customers',
+  ylab = 'MRR',
   main = 'Segment 2'
 )
 
-plot(seq(1:nrow(seg)), seg$ARPU, cex = 0.1, xlab = 'time', ylab = 'ARPU', main = 'Segment 2')
-plot(seq(1:nrow(seg)), seg$Segment.2.MRR, cex = 0.1, xlab = 'time', ylab = 'MRR', main = 'Segment 2')
+# Likely to get a better model with total paid customers.
 
 analyze_model <- function(formula.str, show.plot = TRUE) {
-  x <- seg$ARPU[1:200]
+  x <- seg$Total.paid.customers[1:200]
   y <- seg$Segment.2.MRR[1:200]
   m <- lm(as.formula(formula.str))
   
-  newdata <- data.frame(x = seg$ARPU[201:nrow(seg)])
+  newdata <- data.frame(x = seg$Total.paid.customers[201:nrow(seg)])
   mrr.predicted  <- predict(m, newdata = newdata)
   mrr.actual  <- seg$Segment.2.MRR[201:nrow(seg)]
   
   if (show.plot) {
     plot(
+      newdata$x,
       mrr.actual,
-      mrr.predicted,
       cex = 0.1,
-      xlab = 'actual',
-      ylab = 'predicted',
-      main = formula.str
+      xlab = 'Total paid customers',
+      ylab = 'MRR',
+      main = formula.str,
+      type = 'l'
     )
-    abline(0, 1, col = 'blue')
+    lines(newdata$x, mrr.predicted, col = 'blue')
+    legend(
+      'topleft',
+      legend = c('actual', 'predicted'),
+      col = c('black', 'blue'),
+      lty = c(1, 1),
+      bty = 'n'
+    )
   }
+  
+  print(summary(m))
+  plot(m, cex = 0.1, main = formula.str)
   
   m
 }
 
-# Quadratic model.
-m <- analyze_model('y ~ x + I(x^2)')
-summary(m)
-plot(m)
+analyze_loess <- function(formula.str, show.plot = TRUE) {
[…]

### 2020-11-22 revenue-forecast/analysis-3.R
LOESS on segment 1.

@@ -1,0 +1,38 @@
+seg <- read.csv('segment-1.csv')
+seg$month <- substr(seg$Date, 1, 2)
+
+analyze_loess <- function(formula.str, show.plot = TRUE) {
+  x <- seg$ARPU[1:200]
+  y <- seg$Segment.1.MRR[1:200]
+  m <- loess(as.formula(formula.str), control=loess.control(surface="direct"))
+  
+  newdata <- data.frame(x = seg$ARPU[201:nrow(seg)])
+  mrr.predicted  <- predict(m, newdata = newdata)
+  mrr.actual  <- seg$Segment.1.MRR[201:nrow(seg)]
+  
+  if (show.plot) {
+    plot(
+      newdata$x,
+      mrr.actual,
+      cex = 0.1,
+      xlab = 'ARPU',
+      ylab = 'MRR',
+      main = formula.str,
+      type = 'l'
+    )
+    lines(newdata$x, mrr.predicted, col = 'blue')
+    legend(
+      'topleft',
+      legend = c('actual', 'predicted'),
+      col = c('black', 'blue'),
+      lty = c(1, 1),
+      bty = 'n'
+    )
+  }
+  
+  print(summary(m))
+  
+  m
+}
+
+m <- analyze_loess('y ~ x')

### 2020-11-23 revenue-forecast/analysis-4.R
Some more analysis.

@@ -1,0 +1,93 @@
+# Segment 1: time series analysis of ARPU.
+library(tseries)
+seg <- read.csv('segment-1.csv')
+seg$month <- substr(seg$Date, 1, 2)
+
+arpu <- seg$ARPU
+plot(
+  arpu,
+  cex = 0.1,
+  xlab = 'time',
+  ylab = 'ARPU',
+  main = 'Linear model'
+)
+time <- seq(1, length(arpu))
+m1 <- lm(arpu ~ time)
+abline(m1, col = 'blue')
+# The linear model is not bad but one should try something better.
+
+m2 <- lm(arpu ~ time + I(time ^ 2))
+summary(m2)
+arpu.pred <- predict(m2, newdata = data.frame(time = time))
+plot(
+  arpu,
+  arpu.pred,
+  cex = 0.1,
+  xlab = 'Actual ARPU',
+  ylab = 'Predicted ARPU',
+  main = 'Performance of quadratic model'
+)
+abline(0, 1)
+# The quadratic model fits best for daily data.
+
+future.times <- seq(from = 366, to = 366 + 90)
+future.arpu  <-
+  predict(m2, newdata = data.frame(time = future.times))
+
+all.arpu <- c(arpu, future.arpu)
+plot(
+  all.arpu,
+  cex = 0.1,
+  xlab = 'time',
+  ylab = 'ARPU',
+  main = 'Actual + Predicted ARPU'
+)
+
+# Analysis of monthly data.
+
+seg.agg <- aggregate(ARPU ~ month, data = seg, sum)
+m.arpu <- seg.agg$ARPU
+plot(
+  m.arpu,
+  cex = 0.1,
+  xlab = 'time',
+  ylab = 'Monthly ARPU',
+  main = 'Linear model for monthly data'
+)
+time <- seq(1, nrow(seg.agg))
+m3 <- lm(m.arpu ~ time)
+abline(m3, col = 'blue')
+# Not a great fit.
+
+adf.test(m.arpu)
+# The time series is not stationary.
+adf.test(diff(m.arpu))
+# This too is not stationary.
+# Therefore, no time series analysis can be done. We will go with the 20% increase
+# rule.
+
+# How does data aggregated from prediction of daily series look like?
+jan.arpu <- sum(future.arpu[1:31])
+feb.arpu <- sum(future.arpu[32:59])
+mar.arpu <- sum(future.arpu[60:91])
+
[…]

### 2020-11-23 fdtd/em_young_0.py
WIP 23Nov2020.

@@ -1,0 +1,261 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Nov 12 09:47:47 2020
+
+@author: ajoshi
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as patches
+import time
+import os
+
+# Physical parameters of the problem.
+c = 3e8                 # The speed of light in vacuum.
+kappa = 1e4             # The relative dielectric permittivity of the screen.
+sigma = 0.3             # The conductivity of the screen.
+epsilon_0 = 8.854e-12   # Permittivity of free space in SI units.
+    
+# Geometrical parameters of the problem. 
+# The extent of the domain. **You can alter these parameters.**
+ye = 200 # Extent along y-axis,
+xe = 200 # Extent along x-axis
+
+# The central points in either directions.
+yc = ye // 2 - 1
+xc = xe // 2 - 1
+
+# The location of the perfectly matching layer.
+y_pml_0 = 10                # starting vertical position after PML.
+y_pml_1 = ye - y_pml_0 - 1  # ending horizontal position before PML.
+x_pml_0 = 10                # starting horizontal position after PML.
+x_pml_1 = xe - x_pml_0 - 1  # ending horizontal position before PML.
+
+# Slit geometry. **You can alter these parameters.**
+slit_ht  = 8    # The height of each slit.
+slit_sep = 20   # The separation between both slits.
+slit_wt  = 8    # The width of the slit.
+
+# The field vectors.
+Ez = np.zeros((ye, xe)) # E field.
+Dz = np.zeros((ye, xe)) # D field.
+Hx = np.zeros((ye, xe)) # x-component of the H field.
+Hy = np.zeros((ye, xe)) # x-component of the H field.
+Ez_inc = np.zeros(xe)   # z-component of the incident E field.
+Hx_inc = np.zeros(xe)   # x-component of the incident H field.
+
+# Simulation parameters and data.
+cell_size = 0.01 # Cell size in m
+dt = cell_size / 2*c
+# Dielectric factors are the coefficients of time integrals of E_z field.
[…]

### 2020-11-23 queueing-theory/analysis-5.R

@@ -1,0 +1,36 @@
+library(queueing)
+library(MASS)
+
+data <- read.csv('wait_prob_1.csv')
+# Use only weekday's data.
+data <- data[data$isoday_of_week < 6, colnames(data)]
+
+# Examination of average time.
+summ.avg.time <- summary(data$Avg.time)
+q1 <- summ.avg.time[2]
+q3 <- summ.avg.time[5]
+iqr <- IQR(data$Avg.time)
+upper.limit <- q3 + 1.5 * iqr
+lower.limit <- max(q1 - 1.5 * iqr, 0)
+
+clean.avg.time <- data$Avg.time[lower.limit < data$Avg.time & data$Avg.time < upper.limit]
+
+X <- data$ncalls[data$ncalls > 0]
+Y <- clean.avg.time[order(clean.avg.time)]
+lambda <- mean(X)
+mu <- 12/mean(Y)
+
+# Try a queueing model.
+model.1 <- NewInput.MM1(lambda = lambda, mu = mu)
+CheckInput(model.1)
+qm.1 <- QueueingModel(model.1)
+print(summary(qm.1))
+
+main.title <- 'Distribution of waiting times'
+fw <- qm.1$FW
+fwq <- qm.1$FWq
+
+y.label <- 'F_w, F_{w_q}'
+curve(fw, from = 0, to = 1, xlab = 't', ylab = y.label, main = main.title, col='black')
+curve(fwq, from = 0, to = 1, col = 'blue')
+legend('bottomright', legend = c('Fw', 'F_{w_q}'), cols = c('black', 'blue'))

### 2020-11-23 queueing-theory/fa.py

@@ -1,0 +1,60 @@
+import math
+
+def prob_of_wait(E, m):
+    """
+    E:      traffic
+    m:      # agents.
+    Reference: https://en.wikipedia.org/wiki/Erlang_(unit)
+    """ 
+    p = 1
+               
+    if m > E:   
+        try:
+            numerator = E**m/math.factorial(m) * m/(m - E)
+            denominator = 0
+
+            for i in range(m):
+                denominator += E**i/math.factorial(i)
+
+            denominator += numerator
+
+            p = numerator/denominator
+        except OverflowError:
+            print(f'Overflow due to E = {E}, m = {m}')
+        
+    return p
+
+def calculate_ASA(wait_prob, mu, nagents, E):
+    a_large_number = 1000
+    
+    if nagents > E:
+        return wait_prob * mu/(nagents - E) * time_window_size
+    else:
+        return a_large_number
+    
+def find_nagents(E, m):
+    start = m
+             
+    if not math.isnan(E):
+        threshold = 0.8
+        max_iter = m
+        count = 0
+        while (prob_of_wait(E, start) > threshold) and (count < max_iter):
+            start += 1
+            count += 1   
+            
+        if count == max_iter:
+            print(f'Max iter reached for E = {E}, m = {m}.')
+        
+    return start
+
+def find_nagents_1(E, m):
+    start = m
+             
+    if not math.isnan(E):
+        threshold = 0.8
+        while prob_of_wait(E, start) > threshold:
+            start += 1
+            
+    return start
+    :q

### 2020-11-23 queueing-theory/final-model/ts-analysis.R

@@ -1,0 +1,25 @@
+data <- read.csv('wait_prob_1.csv')
+calls <- data$ncalls
+d <- ts(calls, frequency = 160)
+plot(d)
+# The series does not appear to have seasonality.
+# 
+library(tseries)
+adf.test(calls)
+# The small p value suggests that the time series does not have a unit root.
+# That is, it does not have a trend.
+
+acf(calls)
+pacf(calls)
+# The ACF first falls below 0.05 at lag = 28. The PACF first falls below 0.05
+# at lag = 5. Given the long lag, it might help to check if the first difference
+# of the series has a better behavior.
+acf(diff(calls))
+pacf(diff(calls))
+# The acf plot suggests that p = 1. The fact that we considered first difference
+# suggests that d = 1. The pacf plot isn't that encouraging.
+sample <- calls[1:20]
+fit.0 <- arima(diff(sample), c(1, 1, 0))
+summary(fit.0)
+pred.0 <- predict(fit.0, n.ahead = 10)
+

### 2020-11-23 revenue-forecast/analysis-5.R

@@ -1,0 +1,41 @@
+# Segment 2: time series analysis of 'Total paid customers'.
+seg <- read.csv('segment-2.csv')
+seg$month <- substr(seg$Date, 1, 2)
+
+tpc <- seg$Total.paid.customers
+mrr <- seg$Segment.2.MRR
+adf.test(tpc)
+# The series is not stationary.
+adf.test(diff(tpc))
+# The first differences are stationary.
+acf(diff(tpc), main = '1st difference of TPC', lag.max = 100)
+#
+pacf(diff(tpc), lag.max = 100)
+m1 <-
+  arima(
+    tpc,
+    order = c(12, 0, 0),
+    method = "ML",
+    optim.control = list(maxit = 1000)
+  )
+future.tpc <- predict(m1, n.ahead = 90)
+all.tpc <- c(tpc, future.tpc$pred)
+plot(
+  tpc,
+  xlim = c(0, length(all.tpc)),
+  ylim = c(0, max(all.tpc)),
+  col = 'blue',
+  cex = 0.1,
+  xlab = 'time',
+  ylab = 'TPC',
+  main = 'Current and future TPC'
+)
+points(seq(from = 366, to = length(all.tpc)),
+       future.tpc$pred,
+       col = 'red',
+       cex = 0.1)
+
+adf.test(mrr)
+acf(mrr, lag.max = 120, main = 'MRR') # ACF persists for a very long time, 116 lags
+pacf(mrr, main = 'MRR') # Does not look like MA process.
+m2 <- arima(mrr, order = c(100, 0, 0), method = "ML", optim.control = list(maxit=1000))

### 2020-11-24 optim/opt-talk/optimization.Rmd
Up to LP.

---
title: "Optimization"
author: "Amey Joshi"
date: "11/23/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Contents

- What is optimization?
- Convex sets
- Convex programming
- Linear programming
- Quadratic programming

## What is optimization?

- To optimize means to make most of what is available.
- One likes to 
  - **maximize** returns,
  - **minimize** risks,
  - **minimize** cost,
  - **maximize** profit,
  - find the **best** route from here to there.
- Sometimes we have to achieve some of these simultaneously,
  - **Maximize** returns but **minimize** risk.
- Many times one has to work under constraints.
  - **Maximize** returns but don't invest more than $5\%$ of my funds in a single instrument,
  - **Maximize** profits but operate within a given inventory of raw materials.

## A general optimization problem

- minimize $f(x)$ subject to
  - $g_i(x) \ge 0$, $i = 1, \ldots, m$,
  - $h_i(x) = 0$, $i = 1, \ldots, n$.
- $f, g_i, h_i$ are arbitrary functions.
- $f$ is called the **objective function** while $g_i$ and $h_i$ are called
**constraints**.
- $x$ is an $n$-dimensional vector. That is, $x = (x_1, \ldots, x_n)$.
- This problem is usually quite difficult. 
- It has been attempted since at least $300$ years and has ushered many new concepts
in Mathematics.
- (Unfortunately) We are not mathematicians.
  - We will talk in the mathematical language but we will focus on the ideas, not
  the technicalities.
  - We will consider the optimization problem in simpler settings.

## Some notation

- Mathematics is a language. Here's some vocabulary.
- $\mathbb{N}$ denotes the set of all natural numbers $0, 1, \ldots$.
- $\mathbb{Z}$ denotes the set of all integers $\ldots, -2, -1, 0, 1, 2, \ldots$.
- $\mathbb{R}$ denotes the set of all real numbers.
- $\mathbb{R}^n$ is a set of all $n$-dimensional vectors whose components are real
numbers. They contain elements like $x = (x_1, \ldots, x_n)$.
- If $S$ is a set and it contains an element $a$, we write it as $a \in S$.

## Convex sets

- Suppose $x, y \in S$ then the set of points $\lambda x + (1 - \lambda)y$, where
$0 \le \lambda \le 1$ is a real number is called the set of all *convex combinations*
of $x$ and $y$.
  - In plain language, this set has all the points on the straight line joining $x$ and $y$.
- A set $S$ is called a convex set if convex combinations of all pairs of its points
lies in the set.
  - In plain language, if you take any two points in the set and draw a straight 
  line joining them then the line should lie entirely in the set.
- Examples of convex sets:
  - A circle, a square, an ellipse, a sphere, a box.
- Examples of sets that are not convex.
  - A circle with a hole, heart-shaped set, a torus (medu-vade)

## Convex functions - 1

- A function $f$ is a rule that maps **every** member of a set $D$ to a **unique**
member of a set $R$. It is written as $f: D \mapsto R$. The set $D$ is called the
domain of the function and the set $R$ the range.
- Examples:
  - A rule that maps an integer to its square. Its domain is $\mathbb{Z}$ and its
  range is $\mathbb{N}$.
- Let $f:\mathbb{R}^m \mapsto \mathbb{R}^n$. $f$ is convex if 
$||f(\lambda x + (1 - \lambda y))|| \le \lambda ||f(x)|| + (1 - \lambda) ||f(y)||$.
- What is $||f(x)||$? $f(x) \in \mathbb{R}^n$. It is a vector. $||f(x)||$ is the
magnitude of the vector.
  - The magnitude of a vector $y = (y_1, \ldots, y_n)$ is $||y|| = \sqrt{y_1^2 + \cdots + y_n^2}$.
- If $f$ is convex, $-f$ is concave.
- A function may be convex in some region, concave in others.
- Choose $\lambda = 1/2$. If $\bar{x} = (x + y)/2$, $\overline{f(x)} = (f(x) + f(y))/2$ then
$||f(\bar{x})|| \le ||\overline{f(x)}||$.
  - Function of average is at most the average of the function.

## Convex functions - 2
```{r fig.width=12}
par(mfrow = c(1, 2))
x = seq(from = 0.1, to = 5, length.out = 100)

plot(x, (x-2)^2, xlab = 'x', ylab = expression((x-2)^2), main = 'Convex function', type = 'l')
plot(x, log(x), xlab = 'x', ylab = 'log(x)', main = 'Concave function', type = 'l')
```

- In a convex function, a chord lies above the tangent. In a concave function it is it the other way round.

## Convex functions - 3
- Why are we so concerned about convex functions?
  - For a convex function, a local minimum is the global minimum.
  - If you find a minimum you can be assured that it is **the** minimum. You don't have to search any further.
- An optimization problem consists of
  - A set $F$ of feasible points.
  - A cost function $c: F \mapsto \mathbb{R}$.
- The solution of the optimization problem consists in finding $x^\ast \in F$ such that
$c(x^\ast) \le c(x)$ for all $x \in F$.
- A convex optimization problem is the one in which the set $F$ and the function $c$
are convex.
- In general, convex problems are much easier to solve than non-convex ones.
- CPLEX solves convex problems.

## Linear Programming
- Programming?
  - This is a term from the 1940s when a computer meant a person who computes and
  a program mean a plan.
  - In particular, it referred to the US army's training and logistic plans. George
  Dantzing, the man who developed the simplex algorithm was working on military
  problems during WW-II.
- It has the same meaning in 'dynamic programming', a style of building algorithms
that saves intermediate results in tables.
- In a linear programming problem the objective function and the constraints are
**linear functions**.
  - Constraints are in the form of lines/planes/hyperplanes.
  - Feasible region is a convex polygon/polyhedron/polytope.
- An optimum, if it exists, is at the corner of the polytope or along an edge.

## Linear Programming - Enough of theory
- The 'hello world' of optimization is 'the diet problem', a simple version of which is:
```{r}
library(knitr)
diet.data <- data.frame(food = c('Trout', 'CB Sandwich', 'Burrito', 'Hamburger'),
                        Vit.A = c(203, 90, 270, 500),
                        Vit.C = c(92, 84, 80, 90),
                        Vit.D = c(100, 230, 512, 210),
                        calories= c(600, 350, 250, 500))
needs <- data.frame(nutrition = c('Vit.A', 'Vit.C', 'Vit.D'), amount = c(2000, 300, 430))
kables(list(kable(diet.data, caption = 'Foods'),
kable(needs, caption = 'Nutritional needs')))

```

- Minimize your calorie intake but ensure that you get minimum nutrition of 
every kind.
- How to solve it? Let $x_1, x_2, x_3, x_4$ be the amounts of foods consumed.
  - The total calories consumed is $600x_1 + 350x_2 + 250x_3 + 500x_4$,
  - Vit.A consumed is $203x_1 + 90x_2 + 270x_3 + 500x_4$,
  - Vit.C consumed is $92x_1 + 84x_2 + 80x_3 + 90x_4$,
  - Vit.D consumed is $600x_1 + 350x_2 + 250x_3 + 500x_4$.
- Natural to assume $x_1 \ge 0, x_2 \ge 0, x_3 \ge 0, x_4 \ge 0$.

## The diet problem - 1
```{r echo=TRUE}
library(lpSolve)
tot.calories <- c(600, 350, 250, 500)
nutrition <- matrix(data = c(203, 90, 270, 500, 92, 84, 80, 90, 100, 230, 512, 210), byrow = TRUE, nrow = 3)
needs <- c(2000, 300, 430)
food <- c('Trout', 'CB Sandwich', 'Burrito', 'Hamburger')
stopifnot(length(needs) == nrow(nutrition))
lp.results <-
  lp(direction = "min",
     objective.in = tot.calories,
     const.mat = nutrition,
     const.dir = rep(">=", nrow(nutrition)),
     const.rhs = needs)
if (lp.results$status == 0) {
  print(paste('Calories consumed =', round(lp.results$objval, 0)))
  X <- lp.results$solution
  print(paste('Solution:', food[1], ":", X[1], ',', food[2], ':', X[2], ',', food[3], ':', X[3], ',', food[4], ':', X[4]))
}
```

Even professional mathematicians don't order $7.407$ Burritos.

## The diet problem - 2
```{r echo=TRUE}
lp.results <-
  lp(direction = "min",
     objective.in = tot.calories,
     const.mat = nutrition,
     const.dir = rep(">=", nrow(nutrition)),
     const.rhs = needs,
     int.vec = c(1, 2, 3, 4)) # Insist that the solution must be an integer.
if (lp.results$status == 0) {
  print(paste('Calories consumed =', round(lp.results$objval, 0)))
  X <- lp.results$solution
  print(paste('Solution:', food[1], ":", X[1], ',', food[2], ':', X[2], ',', food[3], ':', X[3], ',', food[4], ':', X[4]))
}
```
- Adding the last argument is to the function `lp` is a trivial change to the code but
a significant change to the problem.
  - Integer linear programming is NP-hard.

## Quadratic programming

### 2020-11-24 optim/opt-talk/optimization.Rmd
First draft

@@
 
-## Quadratic programming
+## Quadratic programming - 1
+- The problem does not get complicated if the objective function is a quadratic form.
+That is, it can be written as $$Q = \frac{1}{2}x^T\Sigma x. \tag{1}$$ where $x \in \mathbb{R}^n$
+and $\Sigma$ is an $n \times n$ symmetric matrix.
+- Consider the problem of allocating a certain amount of money into asset classes with the objective of
+minimizing the risk while still ensuring a certain return. It is equivalent to finding out what fraction 
+$x_i$ of the available money should be invested in an asset 1 ≤ i ≤ N . Let the mean returns of the $N$ assets be
+$r_1 , . . . , r_N$ and let $\Sigma$ be the covariance matrix of the returns.
+- The problem then is to minimize the quantity $Q$ in equation (1) subject to
+  - $x_1r_1 + \cdots + x_nr_n \ge r$, where $r$ is that minimum return desired.
+  - $x_1 + \cdots + x_n = 1$, that is, the entire money is invested and
+  - $x_1, \ldots, x_n \ge 0$, that is, there is not shorting.
+
+## Quadratic programming - 2
+- We simulate the returns and the variance-covariance matrix of the returns of
+a portfolio of $N = 5$ assets.
+- Generating a random covariance matrix is a little involved process for it has to be positive semidefinite.
+We use the fact that if Q is an orthogonal matrix and D a diagonal matrix with all non-negative entries
+then QT DQ is positive semi-definite. We can extract an orthogonal matrix from any matrix by its QR
+decomposition.
+
+```{r echo=TRUE}
+set.seed(19471508)
+N <- 5 # number of assets
+mu <- rnorm(n = N, mean = 6, sd = 2) # mean returns
+
+# Randomly generated covariance matrix.
+M <- matrix(data = rnorm(N*N), nrow = N, ncol = N)
+Q <- qr.Q(qr(M))
+C <- Q %*% diag(seq(from = 1, to = 5)) %*% t(Q) # The variance-covariance matrix.
+```
+
+## Quadratic programming - 3
+- We solve the quadratic program.
+
+```{r echo=TRUE}
+library(quadprog)
+k <- rep(1, N)
+r <- 6.5 # Minimum return
+b <- c(r, 1, -1, rep(0, 5))
+A <- rbind(mu, k, -k, diag(rep(1, N)))
+qp.results <- solve.QP(Dmat = C, dvec = rep(0, N), Amat = t(A), bvec = b)
+s <- round(qp.results$solution, 4)
+names(s) <- paste(sep = "", "asset", seq(from = 1, to = N))
+cat("The optimal asset allocation is:\n")
+print(s)
+cat(paste(sep = "", "The risk of the portfolio is ", round(qp.results$value, 4), ".\n"))
+```
+
+## Closing remarks
+- You might feel that optimization is too technical and dry. Yet,
+  - Several problems in physics and engineering are cast as (nonlinear) optimization problems.
+  - Optimization is at the heart of many AI/ML problems. You may have heard of 
+  terms like gradient descent,conjugate gradient descent, batch gradient descent, 
+  XGBOOST, ADABOOST etc. Every one of them is an optimization algorithm.
+  - Even traditional statistical problems like ordinary least square regression
+  and logistic regression are solved with an optimizer. You **minimize** an
+  error function in each of them.
+- Linear optimization might not be glamorous but the simplex method is the 
+commercially most successful algorithm. Some examples:
+  - How to transport $m$ items from $n$ factories to $l$ warehouses in the 
+  *cheapest* way?
+  - How to match $m$ people to $n$ jobs to get the *maximal* matching?
+  - How to build a schedule of buses so that they are idle for the *least* time?
+  - How to design a tour of $n$ cities that is the *cheapest*? (This problem
+  is computationally very expensive.)

### 2020-11-24 fdtd/summary.tex
Summary of the article.

\documentclass[11pt]{article}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{physics, mathtools, graphicx}
\newcommand{\opr}{\prime}
\linespread{1.2}
\title{Young's experiment}
\author{TBD}
\date{12-Nov-2020}
\begin{document}
\maketitle
Young's experiment is a landmark in the history of physics. It provided a firm
evidence in favor of the wave theory of light. The experiment consists of
allowing light to pass through two narrow slits, a distance $d$ apart, in a 
screen and letting it on another screen a distance $D$ from the first screen.
Young observed a pattern of alternating bright and dark bands on the second 
screen and concluded that it was because of constructive and destructive
interference of waves. The nature of the waves was not known in Young's time.
It was only in the later part of the nineteenth century that light was
understood as an electromagnetic wave. This article describes Young's experiment
, derives the electromagnetic wave equation from Maxwell's equations, describes
their nature in material media and solves them numerically to demonstrate the
formation of interference fringes in a numerical simulation.

The numerical simulation of the interference phenomenon is carried out using
the finite difference time domain technique. The derivatives in each one of 
Maxwell's equations are written as central differences on a carefully selected
grid of points spanning the domain of integration. The electric and the magnetic
field are evaluated at alternating points. The finite size of the domain of
integration is taken into account by the absorbing boundary conditions. We
first develop them in the simpler case of one-dimensional problem and extend
it to two-dimensional problems using the perfectly matched layers. We
demonstrate the finite difference time domain technique first in a simple 
situation where we consider the propagation of a gaussian light pulse. We stop
the simulation well before the light reaches the boundary so that we do not
have to worry about the boundary conditions. We build on these simple examples
to simulate Young's experiment where we take into account the boundary
conditions by implementing a perfectly matched layer.

Numerical algorithms, especially the ones solving hyperbolic partial
differential equations, can be unstable. We explain the 
Courant-Friedrichs-Lewy condition that ensures that the time-step of simulation
is chosen in such a way that it is shorter than the time it takes for the 
wave to reach a neighboring grid point.
\end{document}

### 2020-11-25 optim/amber-india/problem-1.R
Added comments.

@@ -1,6 +1,24 @@
 library(lpSolve)
 
+# The objective is to maximize the exposure through three advertisement channels.
+# If x[1], x[2], x[3] are number of ads in TV, magazines and newspapers then the
+# exposure is 1.3e6 x[1] + 6e5 x[2] + 5e5 x[3].
 advertisement <- c(1.3e6, 6e5, 5e5)
+
+# The constraints are:
+# [1] Planning cost should be at most 1 MM. Therefore, 
+#     9e4x[1] + 3e4x[2] + 4e4x[3] <= 1e6.
+# [2] Advertising cost should be at most 4 MM. Therefore, 
+#     3e5x[1] + 1.5e5x[2] + 1e5x[3] <= 4e6.
+# [3] At least 5MM young people should be reached. 
+#     1.2e6x[1] + 0.2e6x[2] + 0x[3] >= 5e6 or
+#     -1.2e6x[1] - 0.26x[2] - 0x[3] <= -5e6.
+# [4] At least 5MM middle-aged people should be reached.
+#     -0.5e6x[1] - 0.2e6x[2] - 0.2e6x[3] <= -5e6
+# [5] Exhaust last year's coupons.
+#     -0x[1] - 4e4x[2] - 1.2e5x[3] <= -1.49e6
+# [6] There are only five slots on TV.
+#      x[1] + 0.x[2] + 0.x[3] <= 5.
 constraints <-
   matrix(
     data = c(
@@ -9,7 +27,7 @@ constraints <-
       -1.2e6, -0.2e6, 0, # Young people.
       -0.5e6, -0.2e6, -0.2e6, # Middle-aged people.
       -0, -4e4, -1.2e5, # Coupons.
-      1, 0, 0 # Available slots.s
+      1, 0, 0 # Available slots.
     ),
     nrow = 6,
     ncol = 3,
@@ -29,5 +47,7 @@ lp.results <-
 if (lp.results$status == 0) {
   print(paste('Objective value =', round(lp.results$objval, 0)))
   optimal.x <- lp.results$solution
-  print(paste('Solution: x1 =', optimal.x[1], ' x[2] =', optimal.x[2], ' x[3] =', optimal.x[3]))
+  print(paste('Solution: x1 =', round(optimal.x[1], 2),
+              ' x[2] =', round(optimal.x[2], 2), 
+              ' x[3] =', round(optimal.x[3], 2)))
 }

### 2020-11-25 optim/amber-india/problem-2.R

@@ -3,6 +3,8 @@ library(nloptr)
 # https://stackoverflow.com/questions/31431575/minimization-with-r-nloptr-package-multiple-equality-constraints
 
 eval_f <- function(x) {
+  # nloptr always minimizes the objective function. We want to maximize the
+  # profit. Therefore, we seek to minimize negative of profit.
   objective <- 0.1 * x[1] * x[1] - 1.13 * x[1] + 0.4 +
     0.002 * x[2] * x[2] - 0.124 * x[2] - 0.14 +
     0.00321 * x[3] * x[3] - 0.706 * x[3] + 0.09
@@ -13,14 +15,16 @@ eval_f <- function(x) {
 }
 
 eval_g_ineq <- function(x) {
+  # These are the constraints from the previous part.
   constraints <- rbind(
     9e4 * x[1] + 3e4 * x[2] + 4e4 * x[3] - 1e6,
-    3e5 * x[1] + 1.5e5 * x[2] + 1e5 * x[3] - 4e6,-1.2e6 *
-      x[1] - 0.2e6 * x[2] + 5e6,
+    3e5 * x[1] + 1.5e5 * x[2] + 1e5 * x[3] - 4e6,
+    -1.2e6 * x[1] - 0.2e6 * x[2] + 5e6,
     -0.5e6 * x[1] - 0.2e6 * x[2] - 0.2e6 * x[3] + 5e6,
     -4e4 * x[2] - 1.2e5 * x[3] + 1.49e6,
     x[1] - 5
   )
+  # Gradient of the constraints.
   gradient <- rbind(
     c(9e4, 3e4, 4e4),
     c(3e5, 1.5e5, 1e5),
@@ -32,7 +36,9 @@ eval_g_ineq <- function(x) {
   return(list("constraints" = constraints, "jacobian" = gradient))
 }
 
+# Initial values, we start from the solution to the previous problem.
 x0 <- c(4, 0, 14)
+
 # lower and upper bounds of control
 lb <- c(0, 0, 0)
 ub <- c(Inf, Inf, Inf)
@@ -44,6 +50,7 @@ opts <- list(
   "maxeval" = 10000,
   "local_opts" = local_opts
 )
+
 res <- nloptr(
   x0 = x0,
   eval_f = eval_f,
@@ -56,8 +63,12 @@ res <- nloptr(
 obj.val <- -res$objective * 1e6 # The lhs of quadratic equations is in MM USD.
[…]

### 2020-11-28 optim/amber-india/problem-1.R
Corrected typos

@@ -39,7 +39,7 @@ const.rhs <- c(1e6, 4e6, -5e6, -5e6, -1.49e6, 5)
 stopifnot(length(const.rhs) == nrow(constraints))
 
 lp.results <-
-  lp(direction = "min",
+  lp(direction = "max",
      objective.in = advertisement,
      const.mat = constraints,
      const.dir = rep("<=", nrow(constraints)),

### 2020-11-28 optim/amber-india/problem-2.R

@@ -8,8 +8,9 @@ eval_f <- function(x) {
   objective <- 0.1 * x[1] * x[1] - 1.13 * x[1] + 0.4 +
     0.002 * x[2] * x[2] - 0.124 * x[2] - 0.14 +
     0.00321 * x[3] * x[3] - 0.706 * x[3] + 0.09
+  # This is the gradient of the objective function.
   gradient <-
-    c(0.2 * x[2] - 1.13, 0.004 * x[2] - 0.124, 0.00641 * x[3] - 0.706)
+    c(0.2 * x[1] - 1.13, 0.004 * x[2] - 0.124, 0.00642 * x[3] - 0.706)
   
   return(list('objective' = objective, 'gradient' = gradient))
 }
@@ -37,7 +38,8 @@ eval_g_ineq <- function(x) {
 }
 
 # Initial values, we start from the solution to the previous problem.
-x0 <- c(4, 0, 14)
+# x0 <- c(4, 0, 14)
+x0 <- c(1, 19, 9)
 
 # lower and upper bounds of control
 lb <- c(0, 0, 0)

### 2020-11-28 optim/opt-talk/insofe.tex
WIP 28Nov2020

\documentclass{beamer}
\usefonttheme[onlymath]{serif}
\usepackage{graphicx}
\newcommand{\sor}{\mathbf{R}}
\usetheme{Boadilla}
\title{Optimisation Methods in Business Analytics}
\author{Amey Joshi}
\date{\today}
\begin{document}
\begin{frame}
\titlepage
\end{frame}
\begin{frame}
\frametitle{The mathematics of decision-making}
\begin{itemize}
\item To optimise is to find the `best' among a number of possibilities.
\item For example, when we select a school for our children, we consider 
several criteria:
\begin{itemize}
\item The quality of education,
\item The distance from home,
\item The peer group,
\item Extra-curricular activities, sports etc.,
\item The cost.
\end{itemize}
\item We choose the school that \textbf{best} meets \textbf{all} the criteria.
\item We follow similar process to find the right job, car, life-partner or
home. A large number of business decisions follow a similar pattern.
\item Optimisation is the mathematical model behind decision-making.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{A brief history of optimisation}
\begin{itemize}
\item Kepler conceived and used the `secretary problem' to select his second
wife.
\item Newton was interested in finding a shape that experiences least 
resistance while moving in air - a study that led to the development of the
Calculus of Variations.
\item The method of least squares by Gauss and Legendre in the 19th century.
\item Until the 20th century the problems were mostly of a scientific nature.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Optimisation outside science}
\begin{itemize}
\item The socialist states required a model to describe their centralised 
economic planning. Leontif developed an input-output model, an early linear
program (LP).
\item Kantorovich put the input-output model in the form of an LP.
\item Stigler worked on the problem of designing the cheapest diet that met
the nutritional needs of US soldiers in the 1930s.
\item Dantzig developed Simplex algorithm to solve LP.
\begin{itemize}
\item Perhaps the world's most `commercially successful' algorithm.
\item Used in all spheres of life - shipping, scheduling, matching, industrial
processes, military operations, transport.
\item Mysterious behaviour - theory warns that it can be very slow, in
practice it is quite fast.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The optimisation problem}
\begin{itemize}
\item Minimise $f: \sor^n \rightarrow \sor$ subject to
\begin{itemize}
\item $g_i(x) \ge 0$, $i = 1, \ldots, n$ and $g_i: \sor^m \rightarrow \sor$,
\item $h_j(x) = 0$, $j = 1, \ldots, m$ and $h_j: \sor^m \rightarrow \sor$.
\end{itemize}
\item The function $f$ to be minimised is called the \textbf{objective}
function.
\item The variables $x = (x_1, \ldots, x_n) \in \sor^n$ to be searched for
the minimum are called the \textbf{decision} variables.
\item The functions $g_i, h_j$ are called \textbf{constraints}.
\item The form is not restrictive.
\begin{itemize}
\item To maximise $f$, minimise $-f$.
\item A constraint of the form $g(x) <= a$ can be written as $-g(x) + a \ge 0$.
\end{itemize}
\item The general optimisation problem is very difficult to solve.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Convex sets}
\begin{itemize}
\item If you select any two points in a set $S$, join them by a straight line and
if the line lies entirely in $S$ then the set is convex.
\begin{itemize}
\item A circle, a polygon, a sphere are all convex.
\item A circle with a hole, a torus, a heart-shaped figure are not convex.
\end{itemize}
\end{itemize}
\begin{figure}
\includegraphics[scale=0.3]{cardioid}
\caption{A non-convex set}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Convex functions}
\begin{itemize}
\item $f$ is a convex function if 
\begin{equation}
f\left(\frac{x_1 + x_2}{2}\right) \le \frac{f(x_1) + f(x_2)}{2},
\end{equation}
for all $x_1, x_2$. Function of average is $\le$ average of the function.
\end{itemize}
\begin{figure}
\includegraphics[scale=0.3]{f1}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Why the fuss about convexity?}
\begin{itemize}
\item If the objective function $f$ and the constraints $g_i, f_j$ in an
optimisation problem are convex then it is called a \textbf{convex optimisation}
problem.
\item Convexity is the boundary between tractable and intractable problems.
\item A minimum of a convex function is \textbf{the} global minimum.
\end{itemize}
\begin{figure}
\includegraphics[scale=0.3]{many}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Linear programming}
\begin{itemize}
\item If the objective function $f$ and the constraints $g_i, f_j$ in an
optimisation problem are linear then it is a \textbf{linear programming}
problem.
\item Linear? Fine. What is `programming'? A relic of the 1940s.
\item The set of all points $x$ which satisfy the constraints is called the
feasible set $F$.
\item The optimal point $x^\ast \in F$ is the one for which $f(x^\ast) \le
f(x)$ for all $x \in F$.
\item The feasible set is a polygon in 2d or a polyhedron in 3d or a convex
polytope in higher dimensions.
\item The point $x^\ast$ is \textbf{always} on the corner of the polytope.
\begin{itemize}
\item Simplex algorithm: hop from corner to corner in search of the minimum.
\item Interior point methods: cut through the interior of the polytope to
reach the minimum.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The diet problem}
\begin{itemize}
\item A military problem from the 1930s. There are $77$ food items. For each
one of them we have data about the number of calories and amount of Ca, Fe,
vitamins A, B1, B2, B3 and C. 
\item We are also given how many calories and nutrients are needed by an
average man, every day.
\item \textbf{The problem}: Find the cheapest combinations of foods that meets
calorific and nutritional needs.
\item George Stigler and his team took 120 person days to solve it using
heuristic methods and desk calculators.
\item Simplex method implemented in Google OR tools on Google Colaboratory
took $0.5$ ms.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Integer linear programming}
\begin{itemize}
\item What if you insist that some or all decision variables are integers?
\item The problem ceases to be convex. 
\begin{itemize}
\item If $m$ and $n$ are two integers in a set and if $m + n$ is odd then
clearly $(m + n)/2$ is not a member of the set.
\end{itemize}
\item Integer linear programming is $NP$-hard.
\item Solved using 
\begin{itemize}
\item Cutting planes method;
\item Branch and bound method;
\item Branch and cut.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Examples of LP - Transportation Problem - 1}
\begin{itemize}
\item There are $m$ warehouses and $n$ outlets. Each warehouse has a certain 
amount of material available and each outlet has a certain demand for the 
material. Every pair of a warehouse and an outlet has a certain cost to 
transport the material.

\item Find the cheapest way to transport goods from the warehouses to the outlets
so that every outlet's demands are met and no warehouse supplies more than
it has.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Examples of LP - Transportation Problem - 1}
\begin{itemize}
\item Decision variables: $x_{ij}$ the amount of material shipped from warehouse
$i$ to outlet $j$.
\item Objective function: 
\begin{equation}\nonumber
\sum_{i=1}^m\sum_{j=1}^n c_{ij}x_{ij}
\end{equation}
\item Constraints:
\begin{itemize}
\item If $a_1, \ldots, a_m$ are the amounts available in the warehouses then
\begin{equation}\nonumber
\sum_{j=1}^n x_{ij} \le a_i.
\end{equation}
\item If $d_1, \ldots, d_n$ are the demands of each outlet then
\begin{equation}\nonumber
\sum_{i=1}^m x_{ij} \le d_j.
\end{equation}
\item Each $x_{ij}$ should be $\ge 0$.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Some more examples of LP}
\begin{itemize}
\item We are given a lorry with a certain capacity in terms of the weight it 
can carry. We are also given a set of items. Each item has a weight, a value 
and a quantity. The quantity is the number of items of that kind. Fill the 
lorry with a selection of items such that the value is maximized. (Knapsack
problem)

\item Consider a rail network connecting two cities by way of a number of 
intermediate cities, where each link of the network has a number assigned to it 
representing its capacity. Assuming a steady state condition, find a maximal 
flow from one given city to the other. (Max-flow problem)

\item There are $m$ agents and $n$ tasks. An agent can do one or more tasks.
Find an assignment of the agents such that maximum number of tasks are completed
and no task is done be more than one agent. (Assignment problem)

\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Beyond LP}
\begin{itemize}
\item When the objective function is a quadratic form we have the quadratic 
programming problem. Used in portfolio optimisation.
\item Even OLS regression is an optimisation problem.
\item Logistic regression uses MLE solved by an optimiser.
\item Gradient descent, Adaptive gradient descent, congugate gradient descent,
xgboost are all optimisation algorithms used in ML algorithms. All of them 
minimize an error term.
\item Several optimisation algorithms are inspired by nature. For example,
simulated annealing, genetic algorithms, swarm-based algorithms etc.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{To sum up ...}
\begin{itemize}
\item Optimisation mimicks human thinking; no surprise that so many AI
algorithms use optimisation underneath;
\item The boundary between easy and hard problem is determined by convexity;
\item Linear programming is far more pervasive than we sometimes imagine.
\end{itemize}
\end{frame}
\end{document}

### 2020-11-28 optim/opt-talk/optimization.Rmd
WIP 28Nov2020

@@
 **constraints**.
-- $x$ is an $n$-dimensional vector. That is, $x = (x_1, \ldots, x_n)$.
+- $x$ is an $n$-dimensional vector. That is, $x = (x_1, \ldots, x_n)$. These are the **decision variables**.
 - This problem is usually quite difficult. 
@@
   - A circle, a square, an ellipse, a sphere, a box.
-- Examples of sets that are not convex.
-  - A circle with a hole, heart-shaped set, a torus (medu-vade)
+- Examples of sets that are [not convex](https://en.wikipedia.org/wiki/Convex_set).
+  - A circle with a hole, heart-shaped set, a [torus](https://en.wikipedia.org/wiki/Torus).
 
@@
 member of a set $R$. It is written as $f: D \mapsto R$. The set $D$ is called the
-domain of the function and the set $R$ the range.
+domain of the function and the set $R$ the co-domain.
 - Examples:
   - A rule that maps an integer to its square. Its domain is $\mathbb{Z}$ and its
-  range is $\mathbb{N}$.
+  codomain is $\mathbb{N}$.
 - Let $f:\mathbb{R}^m \mapsto \mathbb{R}^n$. $f$ is convex if 
@@
 - An optimization problem consists of
-  - A set $F$ of feasible points.
+  - A set $F$ of feasible points. A feasible point is the one that satisfies all constraints.
   - A cost function $c: F \mapsto \mathbb{R}$.
@@
 - It has the same meaning in 'dynamic programming', a style of building algorithms
-that saves intermediate results in tables.
+that save intermediate results in tables.
 - In a linear programming problem the objective function and the constraints are
@@
   - Constraints are in the form of lines/planes/hyperplanes.
-  - Feasible region is a convex polygon/polyhedron/polytope.
+  - Feasible region is a convex polygon/[polyhedron](https://en.wikipedia.org/wiki/Polyhedron)/polytope.
 - An optimum, if it exists, is at the corner of the polytope or along an edge.

### 2020-11-28 optim/opt-talk/cardioid.py

@@ -1,0 +1,12 @@
+#!/bin/python
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+theta = np.linspace(0, 2*np.pi, 1000)
+a = 3
+r = 3 + 3 * np.cos(theta)
+
+plt.polar(theta, r)
+plt.fill(theta, r)
+plt.savefig('cardioid.png')

### 2020-11-28 optim/opt-talk/diet-lp/write_to_csv.py

@@ -1,0 +1,107 @@
+import csv
+
+def main():
+  data = [
+    ['Commodity', 'Unit', 'Price (cents)', 'Calories', 'Protein (g)', 'Ca (g)',
+	'Fe (mg)', 'Vit A (IU)', 'Thiamine (mg)', 'Riboflavin (mg)', 'Niacin (mg)',
+	'Vit C (mg)'],
+    ['Wheat Flour (Enriched)', '10 lb.', 36, 44.7, 1411, 2, 365, 0, 55.4, 33.3, 441, 0],
+    ['Macaroni', '1 lb.', 14.1, 11.6, 418, 0.7, 54, 0, 3.2, 1.9, 68, 0],
+    ['Wheat Cereal (Enriched)', '28 oz.', 24.2, 11.8, 377, 14.4, 175, 0, 14.4, 8.8, 114, 0],
+    ['Corn Flakes', '8 oz.', 7.1, 11.4, 252, 0.1, 56, 0, 13.5, 2.3, 68, 0],
+    ['Corn Meal', '1 lb.', 4.6, 36.0, 897, 1.7, 99, 30.9, 17.4, 7.9, 106, 0],
+    ['Hominy Grits', '24 oz.', 8.5, 28.6, 680, 0.8, 80, 0, 10.6, 1.6, 110, 0],
+    ['Rice', '1 lb.', 7.5, 21.2, 460, 0.6, 41, 0, 2, 4.8, 60, 0],
+    ['Rolled Oats', '1 lb.', 7.1, 25.3, 907, 5.1, 341, 0, 37.1, 8.9, 64, 0],
+    ['White Bread (Enriched)', '1 lb.', 7.9, 15.0, 488, 2.5, 115, 0, 13.8, 8.5, 126, 0],
+    ['Whole Wheat Bread', '1 lb.', 9.1, 12.2, 484, 2.7, 125, 0, 13.9, 6.4, 160, 0],
+    ['Rye Bread', '1 lb.', 9.1, 12.4, 439, 1.1, 82, 0, 9.9, 3, 66, 0],
+    ['Pound Cake', '1 lb.', 24.8, 8.0, 130, 0.4, 31, 18.9, 2.8, 3, 17, 0],
+    ['Soda Crackers', '1 lb.', 15.1, 12.5, 288, 0.5, 50, 0, 0, 0, 0, 0],
[…]

### 2020-11-28 optim/opt-talk/diet-mip.py

@@ -1,0 +1,144 @@
+import time
+from ortools.linear_solver import pywraplp
+
+def main():
+  # Commodity, Unit, 1939 price (cents), Calories, Protein (g), Calcium (g), Iron (mg),
+  # Vitamin A (IU), Thiamine (mg), Riboflavin (mg), Niacin (mg), Ascorbic Acid (mg)
+  data = [
+    ['Wheat Flour (Enriched)', '10 lb.', 36, 44.7, 1411, 2, 365, 0, 55.4, 33.3, 441, 0],
+    ['Macaroni', '1 lb.', 14.1, 11.6, 418, 0.7, 54, 0, 3.2, 1.9, 68, 0],
+    ['Wheat Cereal (Enriched)', '28 oz.', 24.2, 11.8, 377, 14.4, 175, 0, 14.4, 8.8, 114, 0],
+    ['Corn Flakes', '8 oz.', 7.1, 11.4, 252, 0.1, 56, 0, 13.5, 2.3, 68, 0],
+    ['Corn Meal', '1 lb.', 4.6, 36.0, 897, 1.7, 99, 30.9, 17.4, 7.9, 106, 0],
+    ['Hominy Grits', '24 oz.', 8.5, 28.6, 680, 0.8, 80, 0, 10.6, 1.6, 110, 0],
+    ['Rice', '1 lb.', 7.5, 21.2, 460, 0.6, 41, 0, 2, 4.8, 60, 0],
+    ['Rolled Oats', '1 lb.', 7.1, 25.3, 907, 5.1, 341, 0, 37.1, 8.9, 64, 0],
+    ['White Bread (Enriched)', '1 lb.', 7.9, 15.0, 488, 2.5, 115, 0, 13.8, 8.5, 126, 0],
+    ['Whole Wheat Bread', '1 lb.', 9.1, 12.2, 484, 2.7, 125, 0, 13.9, 6.4, 160, 0],
+    ['Rye Bread', '1 lb.', 9.1, 12.4, 439, 1.1, 82, 0, 9.9, 3, 66, 0],
+    ['Pound Cake', '1 lb.', 24.8, 8.0, 130, 0.4, 31, 18.9, 2.8, 3, 17, 0],
+    ['Soda Crackers', '1 lb.', 15.1, 12.5, 288, 0.5, 50, 0, 0, 0, 0, 0],
+    ['Milk', '1 qt.', 11, 6.1, 310, 10.5, 18, 16.8, 4, 16, 7, 177],
[…]

### 2020-11-28 optim/opt-talk/diet.py

@@ -1,0 +1,141 @@
+from __future__ import print_function
+from ortools.linear_solver import pywraplp
+
+def main():
+  # Commodity, Unit, 1939 price (cents), Calories, Protein (g), Calcium (g), Iron (mg),
+  # Vitamin A (IU), Thiamine (mg), Riboflavin (mg), Niacin (mg), Ascorbic Acid (mg)
+  data = [
+    ['Wheat Flour (Enriched)', '10 lb.', 36, 44.7, 1411, 2, 365, 0, 55.4, 33.3, 441, 0],
+    ['Macaroni', '1 lb.', 14.1, 11.6, 418, 0.7, 54, 0, 3.2, 1.9, 68, 0],
+    ['Wheat Cereal (Enriched)', '28 oz.', 24.2, 11.8, 377, 14.4, 175, 0, 14.4, 8.8, 114, 0],
+    ['Corn Flakes', '8 oz.', 7.1, 11.4, 252, 0.1, 56, 0, 13.5, 2.3, 68, 0],
+    ['Corn Meal', '1 lb.', 4.6, 36.0, 897, 1.7, 99, 30.9, 17.4, 7.9, 106, 0],
+    ['Hominy Grits', '24 oz.', 8.5, 28.6, 680, 0.8, 80, 0, 10.6, 1.6, 110, 0],
+    ['Rice', '1 lb.', 7.5, 21.2, 460, 0.6, 41, 0, 2, 4.8, 60, 0],
+    ['Rolled Oats', '1 lb.', 7.1, 25.3, 907, 5.1, 341, 0, 37.1, 8.9, 64, 0],
+    ['White Bread (Enriched)', '1 lb.', 7.9, 15.0, 488, 2.5, 115, 0, 13.8, 8.5, 126, 0],
+    ['Whole Wheat Bread', '1 lb.', 9.1, 12.2, 484, 2.7, 125, 0, 13.9, 6.4, 160, 0],
+    ['Rye Bread', '1 lb.', 9.1, 12.4, 439, 1.1, 82, 0, 9.9, 3, 66, 0],
+    ['Pound Cake', '1 lb.', 24.8, 8.0, 130, 0.4, 31, 18.9, 2.8, 3, 17, 0],
+    ['Soda Crackers', '1 lb.', 15.1, 12.5, 288, 0.5, 50, 0, 0, 0, 0, 0],
[…]

### 2020-11-28 optim/opt-talk/functions.py

@@ -1,0 +1,14 @@
+#!/bin/python
+
+import matplotlib.pyplot as plt
+import numpy as np
+
+x = np.linspace(0, 5, 100)
+y1 = np.power(x - 2, 2)
+y2 = np.log(10*x + 1)
+
+plt.plot(x, y1, label = r"$y = (x - 2)^2$" + " Convex")
+plt.plot(x, y2, label = r"$y = \log(10x + 1)$" + " Concave")
+plt.legend()
+plt.title("Convex and concave functions")
+plt.savefig("f1.png")

### 2020-11-28 optim/opt-talk/many.py

@@ -1,0 +1,12 @@
+#!/bin/python
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+x = np.linspace(0, 25, 100)
+y = np.sqrt(x) * np.cos(x) + x/3
+plt.plot(x, y)
+plt.xlabel("x")
+plt.ylabel("y")
+plt.title("A function with several extrema")
+plt.savefig("many.png")

### 2020-11-29 optim/amber-india/prob3.tex
The third problem.

\documentclass{article}
\begin{document}
\begin{enumerate}
\item
Let $s$ be the selling price, $c$ be the cost price, $u$ be the units in stock
and $D$ be the random demand. Then the profit is
\begin{equation}
p = s \min(u, D) - cu.
\end{equation}

\item The decision variable is $u$, the units stocked.

\item The optimal number of units is
\begin{equation}
u = F^{-1}\left(\frac{s - c}{s}\right),
\end{equation}
where $F$ is the cumulative distribution function of the random demand.

In the problem $p = 12$ dollars. The cost per glass is $75/25 = 3$ 
dollars. Therefore, 
\begin{equation}
u = F^{-1}\left(\frac{12 - 3}{12}\right) = F^{-1}(0.75).
\end{equation}
The demand $D \sim U(200, 500)$ so that $u = 200 + 300 \times 0.75 = 425$.

Alternatively you can use the R function 
\begin{verbatim}
u <- qunif(0.75, 200, 500)
\end{verbatim}
to get the same answer.

The optimal profit is $(12 - 3) \times 425/25 = 153$ dollars.
\end{enumerate}
\end{document}

### 2020-12-02 optim/opt-talk/insofe.tex
Spell check

@@
 and a quantity. The quantity is the number of items of that kind. Fill the 
-lorry with a selection of items such that the value is maximized. (Knapsack
+lorry with a selection of items such that the value is maximised. (Knapsack
 problem)
@@
 \item Logistic regression uses MLE solved by an optimiser.
-\item Gradient descent, Adaptive gradient descent, congugate gradient descent,
+\item Gradient descent, Adaptive gradient descent, conjugate gradient descent,
 xgboost are all optimisation algorithms used in ML algorithms. All of them 
-minimize an error term.
+minimise an error term.
 \item Several optimisation algorithms are inspired by nature. For example,
@@
 \begin{itemize}
-\item Optimisation mimicks human thinking; no surprise that so many AI
+\item Optimisation mimics human thinking; no surprise that so many AI
 algorithms use optimisation underneath;

### 2021-03-09 la/c1.tex
Finished the section on fields.

@@
 The set $(Z_m, \oplus)$ is an abelian group. However, $(Z_m, \odot)$
-has an inverse if and only if $m$ is prime.
+has an inverse if and only if $m$ is prime. If $m$ is prime then the 
+remainders when $\alpha\dot\beta$ are divided by $m$ are in the range
+$0, \ldots, m- 1$. If it is not then the remainders may not include $1$
+when one of $\alpha$ of $\beta$ is a divisor of $m$. As a result, in
+this latter case, some elements may not have an inverse.
+\begin{enumerate}
+\item The table of additions modulo $5$ is
+\[
+\begin{array}{c|c c c c c}
++ & 0 & 1 & 2 & 3 & 4 \\
+\hline 
+0 & 0 & 1 & 2 & 4 & 4 \\
+1 & 1 & 2 & 3 & 4 & 0 \\
+2 & 2 & 3 & 4 & 0 & 1 \\
+3 & 3 & 4 & 0 & 1 & 2 \\
+4 & 4 & 0 & 1 & 2 & 3
+\end{array}
+\]
+From this table it we observe that $1 + 4 = 0$ so that $-1 = 4$.
+\item The table of multiplication modulo $7$ is
+\[
+\begin{array}{c|c c c c c c c c}
+\cdot & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\
+\hline
+0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+1 & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\
+2 & 0 & 2 & 4 & 6 & 1 & 3 & 5 \\
+3 & 0 & 3 & 6 & 2 & 5 & 1 & 4 \\
+4 & 0 & 4 & 1 & 5 & 2 & 6 & 3 \\
+5 & 0 & 5 & 3 & 1 & 6 & 4 & 2 \\
+6 & 0 & 6 & 5 & 4 & 3 & 2 & 1 
+\end{array}
+\]
+From this table we read $3 \times 5 = 1$ so that $1/3 = 5$.
 \end{enumerate}
+
+\item Let
+\[
+x_m = \begin{cases}
+1 & \text{ when } m = 0 \\
+x_{m-1} + 1 & \text{ otherwise}
+\end{cases}
+\]
+Suppose that $x_m = 0$ and the field is not finite. Then $x_{km} = 0$ for all
+$k = 1, 2, \ldots$ or that $x_{km - 1} + 1 = 0$ for all $k = 1, 2, \ldots$ or
+that $x_{km - 1} = -1$ for all $k = 1, 2, \ldots$. However, the additive
+inverse has to be unique.
+
+If the field is finite then we proved in the previous exercise that $m$ must
+be prime.
+
+\item The set $Q(\sqrt{2})$ is a field but $Z(\sqrt{2})$ is not. The latter
+set does not have a multiplicative inverse.
+
+\item The set of all polynomials with integer coefficients is not a field.
+The reciprocal of $x^2 + x - 2$ is calculated as
+\begin{eqnarray*}
+\frac{1}{x^2+x-2} &=& -\frac{1}{3}\left(\frac{1}{1-x} - 
+                      \frac{1}{2}\frac{1}{1+x/2}\right) \\
+ &=& -\frac{1}{6} + \frac{x}{4} - \frac{7}{8}x^2 + \frac{5}{16}x^3 + \cdots.
+\end{eqnarray*}
+Clearly, the reciprocal is not a polynomial of integer coefficients.
+
+The set of all polynomials with real coefficients is a field. By the 
+fundamental theorem of algebra, any polynomial of degree $n$ has $n$ roots,
+some of which may be complex. Further, the complex roots always come in pairs
+of conjugates. If all roots are real then the polynomial definitely has a
+reciprocal with real coefficients. We will show that even if the polynomial has
+complex roots, the resulting reciprocal is real.
+
+Since the complex roots come in pairs, the reciprocal will have factors like
+\[
+\frac{1}{(x - a)(x - \bar{a})} = \frac{i}{2\Im(a)}\left(\frac{1}{x - \bar{a}}
+- \frac{1}{x - a}\right).
+\]
+We can write this as
+\begin{eqnarray*}
+\frac{1}{(x - a)(x - \bar{a})} &=& \frac{i}{2\Im(a)}\left[\left(\frac{1}{a} + 
+\frac{1}{\bar{a}}\right) - x\left(\frac{1}{a^2} + \frac{1}{\bar{a}^2}\right) 
++ x^2\left(\frac{1}{a^3} + \frac{1}{\bar{a}^3}\right) + \cdots\right] \\
+ &=& \frac{i}{2\Im(a)}\sum_{k \ge 0}(-x)^k
+     \left(\frac{1}{a^k} + \frac{1}{\bar{a}^k}\right) \\
+ &=& \frac{i}{2\Im(a)}\sum_{k \ge 0}(-x)^k\frac{a^k + \bar{a}^k}{|a|^{2k}}
+\end{eqnarray*}
+If $a = x + iy$ then
+\[
+a^k + \bar{a}^k = \sum_{l=0}^k\binom{k}{l}x^{l}y^{k-l}(i^{k-1} + (-i)^{k-l}),
+\]
+which is always real.
+
+\item Consider the pair $(\alpha, \beta)$ of real numbers.
+\begin{enumerate}
+\item Let addition and multiplication be defined as
+\begin{eqnarray*}
+(\alpha, \beta) + (\gamma, \delta) &=& (\alpha + \gamma, \beta + \delta) \\
+(\alpha, \beta)(\gamma, \delta) &=& (\alpha\gamma, \beta\delta)
+\end{eqnarray*}
+Then $(0, 0)$ is the additive identity and $(1, 1)$ the multiplicative. 
+Therefore, the multiplicative inverse of $(\alpha, \beta)$ is defined with the
+equation
+\[
+(\alpha, \beta)(\gamma, \delta) = (1, 1).
+\]
+Clearly, the inverse is not defined for $(\alpha, 0)$ and $(0, \beta)$.
+
+\item If the addition and multiplication are defined as
+\begin{eqnarray*}
+(\alpha, \beta) + (\gamma, \delta) &=& (\alpha + \gamma, \beta + \delta) \\
+(\alpha, \beta)(\gamma, \delta) &=& (\alpha\gamma - \beta\delta,
+                                     \alpha\delta + \beta\gamma)
+\end{eqnarray*}
+then the operations are same as that for complex numbers, a field.
+
+\item If $\alpha$ and $\beta$ are complex numbers then the first scheme fails
+to form a field for the same reasons. In the second scheme, consider the
+product $(1, i)(1, -i) = (0, 0)$. Thus, a product of two non-zero numbers is
+zero. This is the same problem that we had in $Z_m$ when $m$ was not a prime.
+However, if non-zero elements $a, b$ belong to a field then $ab \ne 0$.
+\end{enumerate}
+
+\end{enumerate}

### 2021-04-11 com/c1.tex
Initial draft.

\chapter{Functions on Euclidean Space}\label{c1}
\section{Norm and Inner Product}\label{c1s1}
$\sor^n$ has points $x = (x^1, \ldots, x^n)$ such that if $\alpha \in \sor$
then $\alpha x = (\alpha x^1, \ldots, \alpha x^n)$ is also a point in $\sor^n$
and so is $x + y = (x^1 + y^1, \ldots x^n + y^n)$. $\sor^n$ is a linear vector
space. 
\begin{defn}\label{c1s1d1}
The norm of a vector $x \in \sor^n$ is defined as
\[
\norm{x} = \left((x^1)^2 + \cdots + (x^n)^2\right)^{1/2}.
\]
\end{defn}

\begin{defn}\label{c1s1d2}
The inner product of vectors $x, y \in \sor$ is defined as
\[
\ip{x}{y} = \abs{\sum_{i=1}^n x^i y^i }.
\]
\end{defn}

\begin{lem}\label{c1s1l1}
$\norm{x}^2 = \ip{x}{x}$.
\end{lem}

\begin{lem}\label{c1s1l2}
$\ip{x}{y} = \ip{y}{x}$.
\end{lem}

\begin{lem}\label{c1s1l3}
$\ip{\alpha x}{y} = \alpha\ip{y}{x}$ and $\ip{x}{\alpha y} = \alpha\ip{x}{y}$
for all $\alpha \in \sor$.
\end{lem}

\begin{lem}\label{c1s1l4}
$\ip{x}{y_1 + y_2} = \ip{x}{y_1} + \ip{x}{y_2}$ and $\ip{x_1 + x_2}{y} = 
\ip{x_1}{y} + \ip{x_2}{y}$.
\end{lem}

\begin{lem}[Polarisation identity]\label{c1s1l5}
\[
\ip{x}{y} = \frac{\norm{x + y}^2 - \norm{x - y}^2}{4}.
\]
\end{lem}
\begin{proof}
Follows immediately from lemmas \ref{c1s1l1} and \ref{c1s1l2}.
\end{proof}

The norm has the following properties.
\begin{thm}\label{c1s1t1}
If $x, y \in \sor^n$ and $\alpha \in \sor$ then
\begin{enumerate}
\item $\norm{x} \ge 0$ and $\norm{x} = 0$ iff $x = 0$.
\item Cauchy-Schwarz inequality
\[
\ip{x}{y} \le \norm{x}\norm{y}.
\]
the equality is true if $x = \lambda y$ for some $\lambda \in \sor$.
\item $\norm{x + y} \le \norm{x} + \norm{y}$.
\item $\norm{\alpha x} = \abs{\alpha}\norm{x}$ for all $\alpha \in \sor$.
\end{enumerate}
\end{thm}
\begin{proof}
$\norm{x}$ is the positive square root of the sum of non-negative reals. 
Therefore it is non-negative. If $x = 0$ then clearly $\norm{x} = 0$. If
$\norm{x} = 0$ then the sum of $n$ non-negative numbers is zero, which is
possible only if all of them are zero.

Consider $\norm{x - \lambda y}^2 = \ip{x - \lambda y}{x - \lambda y}$
so that
\[
\norm{x - \lambda y}^2 = \ip{x}{x} - 2\lambda\ip{x}{y}  + \lambda^2\ip{y}{y}
\]
We showed previously that $\norm{x - \lambda y} \ge 0$ and that it is zero
iff $x - \lambda y = 0$. If $\norm{x - \lambda y}^2 > 0$ then then the quadratic
on the rhs has no real roots, in which case the discriminant
\[
4\lambda^2\ip{x}{y}^2 - 4\lambda^2\norm{x}^2\norm{y}^2 < 0
\]
from which it follows that
\[
\ip{x}{y} < {\norm{x}}{\norm{y}}.
\]
If, on the other hand, $\norm{x - \lambda y} = 0$ then the quadratic has real
roots in which case $\ip{x}{y} \ge {\norm{x}}{\norm{y}}$. However, if $\norm{x -
\lambda y} = 0$ then $x = \lambda y$ and then $\ip{x}{y} = {\norm{x}}
{\norm{y}}$. In either case, $\ip{x}{y} \le {\norm{x}} {\norm{y}}$. Now suppose
that $\ip{x}{y} = \norm{x}\norm{y}$ then the discriminant of the quadratic
is zero which indicates that real roots exist and that $x = \lambda y$.

$\norm{x + y}^2 = \ip{x+y}{x+y} = \norm{x}^2 + 2\ip{x}{y} + \norm{y}^2$. Using
the Cauchy-Schwarz inequality, $\norm{x + y}^2 \le \norm{x}^2 + 2\norm{x}
\norm{y} + \norm{y}^2 = (\norm{x} + \norm{y})^2$.

The fourth statement follows immediately from the definitions of the scalar
product and the norm.
\end{proof}

### 2021-04-12 com/c1.tex
WIP 12Apr2021

@@
 \end{proof}
+
+\begin{lem}\label{c1s1l6}
+$\norm{x} = \norm{-x}$.
+\end{lem}
 
@@
 \end{proof}
+
+\subsection{Problems}
+\begin{enumerate}
+\item We will prove by induction on $n$. If $n = 1$, $x = (x^1)$ and $\norm{x}
+= \sqrt{(x^1)^2} \le \abs{x^1}$, as we always choose the positive square root.
+Assume that the statement is true for all $n < k$ and consider
+\[
+\norm{x}^2 = \sum_{i=1}^{k-1}(x^i)^2 + (x^k)^2 \le 
+\left(\sum_{i=1}^{k-1}\abs{x^i}\right)^2 + \abs{x^k}^2 = 
+\left(\sum_{i=1}^k\abs{x^i}^2\right)^2.
+\]
+
+\item When is $\norm{x + y} = \norm{x} + \norm{y}$? Only when $x = \lambda y$ 
+for $\lambda \ge 0$.
+
+\item $\norm{x - y} = \norm{x + (-y)} \le \norm{x} + \norm{-y} = \norm{x} + 
+\norm{y}$. Equality holds when $x = \lambda y$ with $\lambda \le 0$.
+
+\item $\norm{x} = \norm{x - y + y} \le \norm{x - y} + \norm{y}$ so that
+$\norm{x} - \norm{y} \le \norm{x - y}$. We can similarly show that $\norm{y} -
+\norm{x} \le \norm{y - x} = \norm{x - y}$.
+
+\item The length of one side of a triangle never exceeds the sum of lengths of
+the other two sides.
+
+\item Given that $f$ and $g$ are (real-valued) integrable on $[a, b]$. 
+\begin{enumerate}
+\item[(a)] Then
+\[
+\int_a^b (f - \lambda g)^2 = \int_a^b f^2 - 2\lambda\int_a^b fg + \lambda^2
+\int_a^b g^2.
+\]
+The expression is non-negative. If suppose it is zero then $f - \lambda g = 0$
+almost everywhere. If $f$ and $g$ are continuous then $f = \lambda g$, that is
+the two functions are linearly dependent, in which case
+\[
+\int_a^b fg = \left(\int_a^b f^2\right)^{1/2}\left(\int_a^b g^2\right)^{1/2}.
+\]
+On the other hand, if the expression
+is positive then the quadratic
+\[
+\lambda^2 \int_a^b g^2 - 2\lambda\int_a^b fg + \int_a^b f^2
+\]
+has no real roots, which is possible only when the discriminant
+\[
+4\lambda^2 \left(\int_a^b fg\right)^2 - 
+4\lambda^2\left(\int_a^b f^2\right)\left(\int_a^b g^2\right) < 0
+\]
+
+\item[(b)] In the case of equality, $f = \lambda g$ only if both functions are
+continuous. Otherwise, $f = \lambda g$ a.e. and the two are \emph{not} linearly
+dependent.
+
+\item[(c)] Part 2 of theorem \ref{c1s1t1} is a special case of this statement
+because a norm defined as a sum has properties similar to an integral except
+that when the norm is zero the vector is also zero.
+\end{enumerate}
+
+\item
+\begin{enumerate}
+\item[(a)] If $T$ is inner product preserving then $\ip{T(x)}{T(y)}=\ip{x}{y}$.
+Choose $x = y$. Then $\ip{T(x)}{T(x)} = \ip{x}{y} \Rightarrow \norm{T(x)}^2 =
+\norm{x}^2$ or that $T$ is norm preserving.
+
+To prove the converse, use the polarisation identity (lemma \ref{c1s1l5})
+\[
+\ip{T(x)}{T(y)} = \frac{\norm{T(x) + T(y)}^2 - \norm{T(x) - T(y)}^2}{4}.
+\]
+Since $T$ is linear, $T(x \pm y) = T(x) \pm T(y)$ so that
+is inner product preserving as well.
+\[
+\ip{T(x)}{T(y)} = \frac{\norm{T(x + y)}^2 - \norm{T(x - y)}^2}{4}.
+\]
+Since $T$ is norm preserving,
+\[
+\ip{T(x)}{T(y)} = \frac{\norm{x + y}^2 - \norm{x - y}^2}{4} = \ip{x}{y}.
+\]
+
+\begin{rem}
+Every inner product preserving transformation is a norm-preserving 
+transformation but the converse is true only if the transformation is linear.
+The set of norm preserving transformations is, in general, bigger than the
+set of inner-product preserving transformations.
+\end{rem}
+
+\item[(b)] Let $T$ be a norm-preserving linear transformation and let if 
+possible $T(x_1) = T(x_2)$. Then $0 = \norm{T(x_1) - T(x_2)} = 
+\norm{T(x_1 - x_2)} = \norm{x_1 - x_2}$ which implies that $x_1 = x_2$.
+
+Thus, $T(e_i) \ne T(e_j)$ for all $i \ne j$ and $T$ is also onto. As a result,
+$T^{-1}$ exists. To show that $T^{-1}$ is of the same sort, it suffices to show
+that $T^{-1}$ is linear and norm preserving. To show that it is linear, let
+$T(x_1) = y_1, T(x_2) = y_2$ and consider
+\[
+T^{-1}(y_1 + y_2) = T^{-1}(T(x_1) + T(x_2)).
+\]
+Since $T$ is linear, the right hand side is $T^{-1}(T(x_1 + x_2)) = x_1 + x_2
+= T^{-1}(y_1) + T^{-1}(y_2)$. Similarly, $T^{-1}(cy_1) = T^{-1}(cT(x_1)) = 
+T^{-1}T(cx_1) = cx_1 = cT^{-1}(y_1)$. To show that it is norm-preserving,
+consider $\norm{T^{-1}(y_1)} = \norm{T^{-1}T(x_1)} = \norm{x_1}$. But $T$ is
+norm-preserving so that we have $\norm{T^{-1}(y_1)}=\norm{T(x_1)}=\norm{y_1}$,
+making $T^{-1}$ norm-preserving.
+\end{enumerate}
+\end{enumerate}
+
+\subsection{Some theorems used to solve the problems}
+\begin{thm}\label{c1s1t2}
+If $f$ is a real-valued continuous function on $[a, b]$ and 
+\[
+\int_a^b f^2 = 0
+\]
+then $f = 0$.
+\end{thm}
+\begin{proof}
+The integrand is non-negative and continuous throughout $[a, b]$. Let it take 
+a positive value $c$ for some $x_0 \in [a, b]$. Then there is neighbourhood
+$(x_0 - \delta, x_0 + \delta$ in which the integrand is positive and hence
+\[
+\int_{x_0 - \delta}^{x_0 + \delta} f^2 > 0.
+\]
+But
+\[
+\int_a^b f^2 \ge \int_{x_0 - \delta}^{x_0 + \delta} f^2 > 0,
+\]
+a contradiction. 
+\end{proof}
+
+\begin{thm}\label{c1s1t3}
+If $f$ is a real-valued integrable function on $[a, b]$ and 
+\[
+\int_a^b f^2 = 0
+\]
+then $f = 0$ a.e..
+\end{thm}
+\begin{proof}
+Suppose $f = 0$ a.e. is not true. Then create sets 
+\begin{eqnarray*}
+B_0 &=& \left\{x \in [a, b]: f^2 \ge 1\right\} \\
+B_n &=& \left\{x \in [a, b] : \frac{1}{n-1} > f^2 \ge \frac{1}{n}\right\}
+\end{eqnarray*}
+where $n > 1$ is an integer. The sets $\{B_n\}$ are disjoint and the integral is
+just the sum of measures of each of these sets. When $f = 0$ a.e. is not true
+at least one of these sets, say $B_k$, will have a positive measure, in which
+case
+\[
+\int_a^b f^2 > \frac{\mu(B_k)}{k} > 0,
+\]
+a contradiction.
+
+\end{proof}

### 2021-04-14 com/c1.tex
WIP 14Apr2021

@@
 \chapter{Functions on Euclidean Space}\label{c1}
-\section{Norm and Inner Product}\label{c1s1}
+\section{Norm and inner product}\label{c1s1}
 $\sor^n$ has points $x = (x^1, \ldots, x^n)$ such that if $\alpha \in \sor$
@@
 \end{enumerate}
+
+\item 
+\begin{enumerate}
+\item[(a)] If $T$ is norm preserving then it is also inner-product preserving.
+Therefore, it leaves the quantity
+\[
+\frac{\ip{x}{y}}{\norm{x}\norm{y}}
+\]
+invariant.
+
+\item[(b)] There is a mistake in the problem statement. $T$ is angle preserving
+if $\lambda_1 = \ldots = \lambda_n = \lambda$. In this case, let
+\begin{eqnarray*}
+a &=& \alpha_1 x_1 + \cdots + \alpha_n x_n \\
+b &=& \beta_1 x_1 + \cdots + \beta_n x_n 
+\end{eqnarray*}
+Then
+\begin{eqnarray*}
+\norm{a}^2 &=& \ip{a}{a} = \sum_{i,j=1}^n\alpha_i\alpha_j\ip{x_i}{x_j} \\
+\norm{b}^2 &=& \ip{b}{b} = \sum_{i,j=1}^n\beta_i\beta_j\ip{x_i}{x_j} \\
+\ip{a}{b} &=& \sum_{i,j=1}^n\alpha_i\beta_j\ip{x_i}{x_j}
+\end{eqnarray*}
+and
+\begin{eqnarray*}
+\norm{T(a)}^2 = \sum_{i,j=1}^n\lambda^2\alpha_i\alpha_j\ip{x_i}{x_j} \\
+\norm{T(b)}^2 = \sum_{i,j=1}^n\lambda^2\beta_i\beta_j\ip{x_i}{x_j} \\
+\ip{T(a)}{T(b)} = \sum_{i,j=1}^n\lambda^2\alpha_i\beta_j\ip{x_i}{x_j}
+\end{eqnarray*}
+so that
+\[
+\angle(T(a),T(b)) = \angle(a, b).
+\]
+
+To spot this mistake, consider the $2$-dimensional case with $T(x_1) = x_1,
+T(x_2) = -x_2$.
+
+Let us now examine the converse. Let $T(x_i) = \lambda_i x_i$. Introduce the
+variables
+\begin{eqnarray*}
+A_{ij} &=& \alpha_i\alpha_j\ip{x_i}{x_j} \\
+B_{ij} &=& \beta_i\beta_j\ip{x_i}{x_j} \\
+C_{ij} &=& \alpha_i\beta_j\ip{x_i}{x_j}
+\end{eqnarray*}
+so that
+\begin{eqnarray*}
+\angle(a, b) &=& \frac{\sum_{r,s}C_{rs}}
+                 {\sqrt{\sum_{k,l}A_{kl}}\sqrt{\sum_{m,n}B_{mn}}} \\
+(\angle(a, b))^2 &=& \frac{\sum_{rstu}C_{rs}C_{tu}}
+                 {\sum_{k,l}A_{kl}\sum_{m,n}B_{mn}}  \\
+(\angle(T(a), T(b)))^2 &=& \frac{\sum_{rstu}\lambda_r\lambda_s\lambda_t\lambda_u
+    C_{rs}C_{tu}}
+    {\sum_{k,l}\lambda_k\lambda_l A_{kl}\sum_{m,n}\lambda_m\lambda_n B_{mn}} 
+\end{eqnarray*}
+When $T$ is angle preserving,
+\[
+\sum_{k,l,m,n,r,s,t,u}\lambda_k\lambda_l\lambda_m\lambda_n A_{kl}B_{mn}C_{rs}
+C_{tu} = 
+\sum_{k,l,m,n,r,s,t,u}\lambda_r\lambda_s\lambda_t\lambda_u A_{kl}B_{mn}C_{rs}
+C_{tu}
+\]
+or
+\[
+\sum_{k,l,m,n,r,s,t,u}(\lambda_k\lambda_l\lambda_m\lambda_n - 
+\lambda_r\lambda_s\lambda_t\lambda_u)A_{kl}B_{mn}C_{rs}C_{tu} = 0.
+\]
+This equation is true for all vectors $a, b$, that is all possible values of 
+$A_{kl}, B_{mn}, C_{rs}$ and $C_{tu}$. Therefore, we must have
+\[
+\lambda_k\lambda_l\lambda_m\lambda_n = \lambda_r\lambda_s\lambda_t\lambda_u
+\]
+for all $k, l, m, n, r, s, t, u$. In particular, it is true for $k = l = m = n$
+and $r = s = t = u$ so that $\lambda_k^4 = \lambda_r^4$ or $\lambda_k = \pm
+\lambda_r$.
+\end{enumerate}
 \end{enumerate}
@@
 \end{proof}
+
+\section{Subsets of euclidean sets}\label{c1s2}
+\begin{thm}[Heine-Borel]
+$[a, b]$ is compact.
+\end{thm}
+\begin{proof}
+Let $\mathcal{O}$ be an open cover of $[a, b]$ and define the set
+\[
+X = \{x : [a, x] \text{ has a finite open cover}\}.
+\]
+The set $X$ is non-empty because $a \in X$. It is also bounded. Therefore, it
+has a supremum, say $L$.
+
+$L \in X$. Since $L = \sup X$, for any $\epsilon > 0$, $L - \epsilon \in X$.
+Choose $\epsilon$ such that it is the radius of one of the member of 
+$\mathcal{O}$ that covers $L$. But this means that $[a, L]$ has a finite cover
+and hence $L \in X$.
+
+$L = b$. If it did not then consider an open set of radius $\epsilon$ that
+covers $L$. Then $L + \epsilon$ will also be in $X$ contradicting the fact
+that $L = \sup X$.
+\end{proof}

### 2021-04-15 com/c1.tex
WIP 16Apr2021

@@
 \end{enumerate}
+
+\item The transformation
+\[
+T = \begin{bmatrix}\cos\theta&\sin\theta\\-\sin\theta&\cos\theta\end{bmatrix}
+\]
+is a rotation by an angle $\theta$ therefore it is angle-preserving. 
+Nevertheless, one can formally confirm its nature by checking if it is norm-
+preserving. Consider 
+\[
+y = T(x) = \begin{bmatrix}x^1\cos\theta + x^2\sin\theta \\ 
+-x^1\sin\theta + x^2\cos\theta\end{bmatrix}
+\]
+so that
+\begin{eqnarray*}
+\norm{y} &=& (x^1)^2\cos^2\theta + (x^2)^2\sin^2\theta + 2x^1x^2\sin\theta
+\cos\theta \\
+ & & + (x^1)^2\sin^2\theta + (x^2)^2\cos^2\theta - 2x^1x^2\sin\theta
+\cos\theta \\
+ &=& (x^1)^2 + (x^2)^2 \\
+ &=&  \norm{x}.
+\end{eqnarray*}
+
+\item $T$ is a fixed linear transformation. Let $h = (h^1, \ldots, h^m)$,
+then the $j$th component of $T(h)$ is $T(h)^j = T_{ij}h^j$ and the norm of
+$T(h)$ is
+\[
+\norm{T(h)} = \sum_{j=1}^m (T_{ij}h^j)^2
+\]
+Let $M$ be the largest among $\abs{T_{ij}}$ so that
+\[
+\norm{T(h)} \le M \sum_{j=1}^m (h^j)^2 = M\norm{h}.
+\]
+\begin{rem} This exercise shows that a linear transformation does not lead to
+a blow up.
+\end{rem}
+
+\item Quite straightforward.
+
+\item $T : \sor^n \rightarrow (\sor^n)^\ast$ is such that $T(x) = \varphi_x$.
+Consider $T(x_1 + x_2) = \varphi_{x_1 + x_2}$ so that
+\begin{eqnarray*}
+T(x_1 + x_2)(y) &=& \varphi_{x_1+x_2}(y) \\
+ &=& \ip{x_1+x_2}{y} \\
+ &=& \ip{x_1}{y} + \ip{x_2}{y} \\
+ &=& \varphi_{x_1}(y) + \varphi_{x_2}(y) \\
+ &=& T(x_1)(y) + T(x_2)(y) \\
+ &=& (T(x_1) + T(x_2))(y)
+\end{eqnarray*}
+\begin{rem}
+$\varphi_x$ is called a linear functional. The space of all linear functionals,
+the dual space $(\sor^n)^\ast$ is a linear space so is the set of 
+transformations $T(x) = \varphi_x$.
+\end{rem}
+
+\item Follows immediately from the fact that $\norm{x+y}^2 = \ip{x+y}{x+y}$ 
+and the linearity of the inner product.
 \end{enumerate}
@@
 \section{Subsets of euclidean sets}\label{c1s2}
-\begin{thm}[Heine-Borel]
+\begin{lem}\label{c1s2l1}
+Every set $X$ has an open cover.
+\end{lem}
+\begin{proof}
+$\{\cup_x B_\epsilon(x) : x \in X, \epsilon > 0\}$ is an open cover, where 
+$B_\epsilon(x)$ is an open ball of radius $\epsilon$ centred at $x$.
+\end{proof}
+
+\begin{thm}[Heine-Borel]\label{c1s2t1}
 $[a, b]$ is compact.
@@
 \end{proof}
+
+\begin{lem}\label{c1s2l2}
+If $B \subset \sor^m$ is a compact set then for any $x \in \sor^n$, the set
+$\{x\} \times B$ is compact.
+\end{lem}
+\begin{proof}
+The finite open cover $A_1, \ldots, A_k$ of $B$ can be extended to $\{x\} \times
+A_1, \ldots, \{x\} \times A_k$, which is a finite open cover of $\{x\}\times B$.
+\end{proof}
+
+\begin{thm}\label{c1s2t2}
+If $B \subset \sor^m$ is a compact set and $\mathcal{O}$ is an open cover of
+$\{x\} \times B$, for an $x \in \sor^n$, then there exists a subset $U$ of
+$\sor^n$, containing $x$, such that $U \times B$ is covered by a finite number
+of members of $\mathcal{O}$.
+\end{thm}
+\begin{proof}
+From \eqref{c1s2l1} we can assume that $\mathcal{O}$ is a finite set. Consider
+$\{x\} \times B$. Let $\{x\} \times W_1, \ldots \{x\} \times W_k$ be its cover.
+Each one of these has an open rectangle $Q_1 \times R_1, \ldots R_k \times Q_k$
+such that $Q_1, \ldots Q_k$ are subsets of $\sor^n$ and $R_1, \ldots, R_k$ are
+subsets of $\sor^m$. Let $U = Q_1 \cap \cdots \cap Q_k$. Then $U \times B$ is
+covered by $Q_1 \times R_1, \ldots, Q_k \times R_k$.
+\end{proof}
+
+This theorem assures that one can always extend the finite open cover of a 
+subset of $\sor^m$ to an open cover of a subset of $\sor^n \times \sor^m$.
+
+\begin{prop}\label{c1s2p1}
+If $A \subset \sor^n$ and $B \subset \sor^m$ are compact then $A \times B 
+\subset \sor^n \times \sor^m$ is compact.
+\end{prop}
+\begin{proof}
+If $U_1, \ldots, U_k$ is an open cover of $A$ and $V_1, \ldots, V_l$ that of
+$B$ then the sets $U_i \times V_j$ where $i = 1, \ldots, k$ and $j = 1, \ldots,
+l$ is a finite open cover of $A \times B$.
+\end{proof}
+
+As a result of theorem \ref{c1s2t1} and proposition \ref{c1s2p1} we have
+\begin{lem}\label{c1s2l3}
+A closed bounded set of $\sor^n$ is compact.
+\end{lem}
+
+\subsection{Some additional lemmas}
+\begin{lem}\label{c1s2l4}
+A set if open if every element of it lies in a open ball entirely in the set.
+\end{lem}
+\begin{proof}
+By definition, a set is open if every element of it lies in an open hyper-
+rectangle entirely in the set. Choose the rectangle to be a hypercube of side 
+$a$ centered at the element. Then we can enclose a ball of radius $a/2$ with 
+centre coinciding the hypercube's centre.
+\end{proof}
+
+\subsection{Problems}
+\begin{enumerate}
+\item It is easy to check that a union, countable or not, of open sets is open.
+Now consider a finite intersection of open sets. If $x \in U_1, \ldots, U_k$
+then there are open balls $B^1_{\epsilon_1}(x), \ldots, B^k_{\epsilon_k}(x)$
+entirely in $U_1, \ldots, U_k$. The ball $B_\epsilon(x)$ where $\epsilon = 
+\min(\epsilon_1, \ldots, \epsilon_k)$ lies entirely in $U_1 \cap \cdots U_k$,
+making it an open set.
+
+Infinite unions may not be open because the infimum of $\epsilon_1, \epsilon_2,
+\ldots$ may be zero leaving us with a singleton set after taking an intersection
+and a singleton set is always closed.
+
+As an example consider the open sets is $U_n = (-1/n, 1/n)$, $n \in \soi^+$.
+Their intersection is $\{0\}$.
+
+\begin{rem}
+That is why a topological space is defined as the pair $(X, \tau)$, where $\tau$
+is a collection of subsets of $X$ such that
+\begin{enumerate}
+\item $X$ and $\varnothing$ belong to $\tau$.
+\item An arbitrary union of sets of $\tau$ belong to $\tau$.
+\item A \underline{finite} intersection of sets of $\tau$ belong to $\tau$.
+\end{enumerate}
+We want a topological space to be closed under operations on open sets.
+\end{rem}
+
+\item Proved as lemma \ref{c1s2l4}.
+
+\item 
+\begin{enumerate}
+\item[(a)] $U = \{x \in \sor^n : \norm{x} \le 1\}$. The interior is the open
+ball $B_1(0)$, the boundary is the sphere $S^2$ and the exterior is everything
+outside it.
+\item[(b)] $S^2 = \{x \in \sor^n : \norm{x} = 1\}$. There is no interior. 
+Neither is there any exterior. $S^2$ is its own boundary.
+\item[(c)] $U = \{x \in \sor^n : x^i \in \soq\}$. There is no neighbourhood of 
+$x \in U$ with points only in $U$. The same is true for the set $\sor^n - U$.
+Therefore, $U$ is neither open nor closed. It has no interior or exterior and
+it is its own boundary.
+\end{enumerate}
+
+\item If $A$ was defined as $(0, 1) \times (0, 1)$ and there was no mention of
+the boundary then a diagonal line would have met the restrictions of the 
+problem.
+
+If $A = [0, 1] \times [0, 1]$ then let $i_n, j_n = 2^{-n}$ for all $n \in \son$.
+Then the points $H=\{0, i_n, 1 - i_n, 1\}$ divide $[0, 1]$ in $n$ intervals of
+equal length. Same is true for the points $V = \{0, j_n, 1 - j_n, 1\}$. The
+set $H \times V$ defines a grid of equally sized cells of the closed rectangle
+ $A$. Choose one point from each of these cells.
+
+\item $A$ is an open set and is a subset of $[0, 1]$. The boundary of $A$
+contains points all neighbourhoods of which contain some points in $A$ and
+some in its exterior. If $E$ is the exterior of $A$ and $\partial A$ is its
+boundary then $E \cap \partial A = [0, 1] - A$.
+
+It suffices to show that $E$ is empty. Let $x \in E$. Then there exists a
+neighbourbood $(x - \epsilon, x + \epsilon)$ such that there is no point of
+$A$ in it. But that is impossible because there is at least one rational number
+in this neighbourhood. Thus, $E = \varnothing$ and hence $\partial A = [0, 1]
+- A$.
+
+\end{enumerate}

### 2021-04-16 com/c1.tex
WIP 16Apr

@@
 
+\begin{defn}\label{c1s2d1}
+A point $x$ is a limit point of a set $X$ is $B_\epsilon(x)$ contains a point
+of $X$ for all $\epsilon > 0$.
+\end{defn}
+
+\begin{lem}\label{c1s2l5}
+If $X$ is a closed set then it contains all its limit points.
+\end{lem}
+\begin{proof}
+Let $x_0$ be a limit point of $X$ and $x_0 \notin X$. Since $X$ is closed,
+$\sor^n - X$ is open. Since $x_0$ is a limit point of $X$, every neighbourhood
+of $x_0$ has points of $X$. Therefore, there is no neighbourhood of $x_0$ with
+points in $\sor^n - X$, a contradiction.
+\end{proof}
+
 \subsection{Problems}
@@
 
+\item Let $X$ be a compact set in $\sor^n$ and let $U_1, \ldots, U_k$ be its
+finite, open cover. Each $U_i$ is of the form $(a_i^1, b_i^1) \times \cdots
+\times (a_i^n, b_i^n)$. Let $a_0^1 = \min(a_1^1, \ldots, a_k^1), b_0^1 = \min(
+b_1^1, \ldots, b_k^1), \ldots a_0^n = \min(a_1^n, \ldots, a_k^n), b_0^n = 
+\min(b_1^n, \ldots, b_k^n)$ so that the set $X$ is bounded by $[a_0^1, b_0^1]
+\times \cdots \times [a_0^n, b_0^n]$.
+
+To show that $X$ is closed consider an arbitrary limit point $x_0$ of $X$. Let,
+if possible, $x_0$ not be in $X$. Consider a family of open sets
+\[
+U_n = \{\sor^n - B_{n^{-1}}(x_0) : n \in \soi^+\}.
+\]
+This is an open cover of $X$. Since $X$ is compact, there exists $m > 0$ such
+that $U_1, \ldots, U_m$ cover $X$. Therefore, the points in $B_{m^{-1}}(x_0)$
+do not belong to $X$, which implies that $x_0$ is not a limit point of $X$.
+
+\item
+\begin{enumerate}
+\item $A$ is closed and $y \notin A$. Therefore, $y$ is not a limit point of 
+$A$. Therefore, there exists $d > 0$ such that $B_d(y)$ lies outside $A$. 
+That is, for all $x \in A$, $\norm{x - y} > d$.
+
+\item Consider the set $\{||x - y|| : x \in A, y \in B\}$. It is bounded below
+and therefore has an infimum, say $d_0 \ge 0$. Let $x_0 \in A$ and $y_0 \in B$
+be such that $\norm{x_0 - y_0} = d_0$. Consider the set of points $\{x(\lambda)
+= x_0 + \lambda(y_0 - x_0) : \lambda \in [0, 1]\}$. This is the straight line 
+joining $x_0$ and $y_0$. There exists $\mu \in [0, 1]$ such that $x(\mu)$
+is neither in $A$ nor in $B$. For if not, there would be two possibilities:
+\begin{itemize}
+\item $x(\lambda) \in A$ for all $\lambda \le \mu$ and $x(\lambda) \in B$ for
+all $\lambda > \mu$. The sets $U_n = \{B_{n^{-1}}(x_0 + \mu(y_0 - x_0)) : n 
+\in \soi^+\}$ form an open cover of $B$. Since $B$ is compact, there exists 
+$m > 0$ such that $U_m$ does not cover $B$. Therefore, for $\mu < \lambda < 
+\mu + m^{-1}$, $x(\lambda) \notin B$, contradicting the assumption that $
+x(\lambda) \in B$ for all $\lambda > \mu$.
+
+\item $x(\lambda) \in A$ for all $\lambda < \mu$ and $x(\lambda) \in B$ for
+all $\lambda \ge \mu$. This is ruled out because if $x(\mu)$ is a limit point
+of $A$ and $A$ is closed.
+\end{itemize}
+
+Since $x(\mu) \notin A$ and $x(\mu)$ is also not a limit point of $A$, there 
+exists $d_1 > 0$ such that $B_{d_1}(x(\mu))$ and $A$ are disjoint so that for 
+all $x \in A$, $\norm{x - x(\mu)} > d_1$.
+
+Since $x(\mu) \notin B$ and $B$ is compact, there exists $d_2 > 0$ such that
+$B_{d_2}(x(\mu)) \cap B = \varnothing$ so that for all $y \in B$, $\norm{y -
+x(\mu)} > d_2$.
+
+Therefore, $\norm{x - y} \le \norm{x - x(\mu)} + \norm{x(\mu) - y} < d_1+d_2$.
+
+\item Consider the sets $A = \{(x^1, x^2) : x^1 x^2 \ge 1, x^2 > 0\}$ and $B = 
+\{(x^1, x^2) : x^1 x^2 \le -1, x^2 > 0\}$. $A$ is the region about the hyperbola
+$x^1x^2 = 1$ and $B$ is the region above the hyperbola $x^1x^2 = -1$. The two
+sets are closed because they contain all their limit points. However, neither
+is compact. As $x^2 \rightarrow \infty$, the two sets get closer to each other
+and there is no $d > 0$ such that their elements are at least a distance $d$
+apart.
 \end{enumerate}
+
+\end{enumerate}

### 2021-04-17 com/c1.tex
WIP 17Apr2021

@@
 
+\item Given that $U$ is open, $C$ is compact, $C \subset U$ so that, $U^c$ is 
+closed $C \cap U^c = \varnothing$. Therefore, (from the previous exercise),
+there exits $d > 0$ such that for all $x \in C, y \in U^c$, $\norm{x - y} \ge
+d$.
+
+Define the collection $\{B_{d/2}(x) | x \in U\}$. It is an open cover of $U$
+and $C \subset U$. Since $C$ is compact, there is finite subcover, say $
+B_{d/2}(x_1) \cup \cdots \cup B_{d/2}(x_n)$ that covers $C$.
+
+Let 
+\[
+D = \overline{\cup_{i=1}^n B_{d/2}(x_i)}
+\]
+then clearly $C \subset D \subset U$. Further, if $X$ is an open set then 
+$X$ lies in the interior of $\overline{X}$.
 \end{enumerate}
+
+\section{Functions and continuity}\label{c1s3}
+\begin{defn}\label{c1s3d1}
+Consider a function $f: \sor^n \rightarrow \sor^m$. Then
+\[
+\lim_{x \rightarrow a}f(x) = b
+\]
+if for any $\epsilon > 0$, there exists $\delta > 0$ such that for all $0 < 
+\norm{x - a} \le \delta$, $\norm{f(x) -b} \le \epsilon$.
+\end{defn}
+\begin{rem}
+$f$ need not be defined at $x = a$ for its limit to be defined at $x = a$.
+\end{rem}
+
+\begin{defn}\label{c1s3d2}
+Consider a function $f: \sor^n \rightarrow \sor^m$. $f$ is said to be continuous
+at $x = a$ if for any $\epsilon > 0$, there exists $\delta > 0$ such that 
+$\norm{x - a} \le \delta \Rightarrow \norm{f(x) - f(a)} \le epsilon$.
+\end{defn}
+\begin{rem}
+$f$ must be defined at $x = a$ for it to be continuous at $x = a$.
+\end{rem}
+
+\begin{thm}\label{c1s3t1}
+If $A \subset \sor^n$ and $f: A \rightarrow \sor^m$. Then $f$ is continuous
+on $A$ if and only if for every open set $U \subset \sor^m$, there is an open 
+set $V \subset \sor^n$ sich that $f^{-1}(U) = V \cap A$.
+\end{thm}
+\begin{proof}
+Assume that $f$ is continuous in $A$. Let $U$ be an open set in $\sor^m$.
+Let $a \in f^{-1}(U)$. By continuity of $f$ in A, for any $\epsilon > 0$, there
+exists $\delta > 0$ such that $x \in B_\delta(a) \Rightarrow f(x) \in B_\epsilon
+(f(a))$. Now the open ball $B_\delta(a)$ need not lie entirely in $A$. Therefore, we say that for any $B_\epsilon(f(a))$ there exists a set $A \cap B_\delta(a)$
+such that $x \in A \cap B_\delta(a) \Rightarrow f(x) \in B_\epsilon(f(a))$. This
+is true for all points $a \in A$.
+
+To prove the converse, let $U = B_\epsilon(y)$, where $y \in \sor^m$. Then
+there exists an open set $V \subset \sor^n$ such that $f^{-1}(U) = V \cap A$.
+Choose $\delta$ such that $B_\delta(f^{-1}(y)) \subset V \cap A$. This proves
+the continuity of $f$.
+\end{proof}
+
+\begin{rem}
+The greatest significance of this theorem is that one can gets an equivalent
+definition of continuity that does not require limits.
+\end{rem}
+
+Continuous functions have several useful properties.
+\begin{thm}\label{c1s3t2}
+If $f: A \rightarrow \sor^m$ is continuous where $A \subset \sor^n$ and $A$
+is compact then $f(A)$ is also compact.
+\end{thm}
+\begin{proof}
+Since $A$ is compact, it has a finite open cover. Since $f$ is continuous, it
+maps an open set in $A$ to an open set in $f(A)$. Thus, $f(A)$ can be covered
+with a finite number of open sets.
+\end{proof}
+
+\begin{rem}
+What about the converse? Functions can be many-to-one. Continuity guarantees
+that the pre-image is open but it cannot guarantee that it is finite. $f(x) =
+\tanh(x)$, $x \in \sor$ is continuous but $\atanh([-1, 1])$ is not compact.
+\end{rem}
+
+\begin{defn}\label{c1s3d3}
+Let $f: A \rightarrow \sor$ be bounded. If
+\begin{eqnarray*}
+M(a, f, \epsilon) &=& \sup\{f(x) : x \in B_\epsilon(a) \cap A\} \\
+m(a, f, \epsilon) &=& \inf\{f(x) : x \in B_\epsilon(a) \cap A\} 
+\end{eqnarray*}
+then the oscillation of $f$ at $a$ is
+\[
+o(f, a) = \lim_{\epsilon \rightarrow 0} (M(a, f, \epsilon) - m(a, f, \epsilon)).
+\]
+\end{defn}
+
+\begin{lem}\label{c1s3l1}
+$o(f, a)$ always exists.
+\end{lem}
+\begin{proof}
+Since $f$ is bounded, the infimum and supremum exist.
+\end{proof}
+
+\begin{thm}\label{c1s3t3}
+A bounded function $f$ is continuous at $a$ if and only if $o(f, a) = 0$.
+\end{thm}
+\begin{proof}
+If $f$ is continuous at $a$ then $f(x) \in B_{\epsilon}(f(a))$ or $M(x, f, a)
+= f(a) + \epsilon$ and $m(x, f, a) = f(a) - \epsilon$ from which it follows
+that $o(f, a) = 0$.
+
+Conversely, if $o(f, a) = 0$ then as $\delta \rightarrow 0$, $M(a,f,\delta)
+- m(a, f, \delta) < \epsilon \Rightarrow f(x) \in (f(a) - \epsilon/2, f(a) + 
+\epsilon/2)$.
+\end{proof}
+
+\subsection{Problems}
+\begin{enumerate}
+\item If 
+\[
+\lim_{x \rightarrow a} f^i(x) = b^i
+\]
+for all $i = 1, \ldots, m$ then it is clear that $f(x) = (f^1(x), \ldots,
+f^m(x)) \rightarrow (b^1, \ldots, b^m) = b$ as $x \rightarrow a$.
+
+To prove the converse, note that 
+\[
+\norm{f(x) - b}^2 = \sum_{i=1}^m (f^i(x) - b^i)^2 < \epsilon^2 \Rightarrow
+\abs{f^i(x) - b^i} < \epsilon.
+\]
+
+\item If $f^i$ is continuous at $a$ then $\abs{f^i(x) - f^i(a)} < \epsilon/
+\sqrt{m}$ and
+\[
+\norm{f(x) - f(a)}^2 = \sum_{i=1}^m \abs{f^i(x) - f^i(a)}^2 < \epsilon^2
+\Rightarrow \norm{f(x) - f(a)} < \epsilon.
+\]
+
+Conversely, 
+\[
+\norm{f(x) - f(a)}^2 < \epsilon^2 \Rightarrow \abs{f^i(x) - f^i(a)}
+^2 < \epsilon^2 \Rightarrow \abs{f^i(x) - f^i(a)} < \epsilon.
+\]
+
+\item Let $T: \sor^n \rightarrow \sor^m$ be a linear transformation. Consider
+$T(x + h) - T(x) = T(h) \le M\norm{h}$, by problem 10 of section \ref{c1s1}.
+Therefore, if $\norm{h} < \epsilon/M$ then $\norm{T(x + h) - T(x)} < \epsilon$.
+
+\item The set $A$ is the region in the first quadrant between the $x$-axis and
+the parabola $y = x^2$.
+\begin{enumerate}
+\item A straightline that does not cut through $A$ surely has an interval
+aroung the origin in $R^2 - A$ because the entire line lies in $R^2 - A$. A 
+line that cuts $A$ still has an interval around the origin in $R^2 - A$
+because for small $x$, $y = x^2 < mx$ and therefore the line is entirely in $R^2
+- A$.
+
+\item $f$ is the indicator function for $A$. For $h \in \sor^2$, define 
+$g_h: \sor \rightarrow \sor = f(th)$. Thus, $g_h(t) = 1$ if $th \in A$ else it
+is zero. To show that $g_h(t)$ is continuous at $t = 0$, consider an interval
+$(-\delta, \delta)$. Any line through the origin has an interval around the 
+origin entirely in $R^2 - A$ on which $g_h(t)$ takes the value $0$. Therefore,
+$g_h$ as a function of $t$ is continuous at the origin.
+
+However $f(h)$ is not continuous because it takes a value $1$ if we approach
+$0$ from the positive $x$ axis and it takes a value $0$ if we approach from
+any other side.
+\end{enumerate}
+
+\item The function $f(x) = \norm{x - a}$ is continuous.
+
+\item $A$ is not closed. Therefore, there is at least one limit point of $A$,
+say $x_0$, that is not part of $A$. The function $f = \norm{x - x_0}^{-1}$ is
+an example of a continuous but unbounded function on $A$.
+
+\item Since $A$ is compact and $f: A \rightarrow \sor$ is continuous, $f(A)$
+is compact. Since $f(A) \subset \sor$, its compactness means that $f(A)$ is 
+also closed and bounded. Therefore $\sup(A), \inf(A) \in f(A)$ so that
+there are $x_m, x_M \in A$ such that $f(x_m) = \inf(A)$ and $f(x_M) = \sup(A)$.
+
+\item $f:[a, b] \rightarrow \sor$ is an increasing function. Without loss of
+generality, assume that $a < x_1 < \ldots, < x_n < b$. Now, 
+\[
+o(f, x_i) = \lim_{\delta \rightarrow 0}(M(x_i, f, \delta) - m(x_i, f, \delta))
+\]
+so that
+\[
+\sum_{i=1}^n o(f, x_i) = \lim_{\delta \rightarrow 0} \sum_{i=1}^n
+(M(x_i, f, \delta) - m(x_i, f, \delta))
+\]
+Since $f$ is increasing,
+\begin{eqnarray*}
+M(x_i, f, \delta) &\le& f\left(\frac{x_i + x_{i+1}}{2}\right) \\
+m(x_i, f, \delta) &\ge& f\left(\frac{x_{i-1} + x_i}{2}\right) 
+\end{eqnarray*}
+for any $\delta > 0$. Therefore, 
+\[
+\sum_{i=1}^n o(f, x_i) \le \sum_{i=1}^n f\left(\frac{x_i + x_{i+1}}{2}\right) -
+f\left(\frac{x_{i-1} + x_i}{2}\right) = f\left(\frac{b + x_n}{2}\right) - 
+f\left(\frac{a + x_1}{2}\right).
+\]
+The result follows from the fact that 
+\begin{eqnarray*}
+f(a) &<& f\left(\frac{a + x_1}{2}\right) \\
+f(b) &>& f\left(\frac{b + x_n}{2}\right).
+\end{eqnarray*}
+
+\end{enumerate}

### 2021-05-08 complex-analysis/c1.tex
Complex numbers

\chapter{Complex Numbers}\label{c1}
\section{Hierarchy of numbers}\label{c1s1}
The set of natural numbers is a monoid. When extended to the set of integers
it becomes a group under addition. In fact, it is also a ring. The set of
rational numbers is a field. But it is not a complete field. The set of real
numbers is a complete field which also has an order. However, the set $\sor[x]$
, of all polynomials with real coefficients, is not algebraically closed. That 
is, some polynomials do not have real roots. The set of complex numbers is an
algebraically closed, complete field. However, it does not have an order. More
on that later.

\section{Construction of complex numbers}\label{c1s2}
There are several ways to construct the set of complex numbers. We will consider
the one that is the simplest. A real number signifies a displacement along a 
line. A complex number does the same on a plane. Therefore, one expects the
complex number to be made up of two real numbers. Let us consider the set of 
all ordered pairs $(x, y)$ of real numbers with the following operations defined
on them.
\begin{itemize}
\item $(x_1, y_1) + (x_2, y_2) = (x_1 + x_2, y_1 + y_2)$,
\item $(x_1, y_1)(x_2, y_2) = (x_1x_2 - y_1y_2, x_1y_2 + x_2y_1)$.
\end{itemize}
It is easy to check that the set of ordered pairs with these definitions is a
field. It is also easy to convince oneself that $(x, 0)$ has all the properties
of the real numbers. If we denote $(0, 1)$ as $i$ and write $(x, y) = x + iy$
then all the properties above can be derived if we assume $i^2 = -1$.

We will now show that there is no order relation in $\soc$. An order relation
$<$ on a field $F$ has the following properties.
\begin{itemize}
\item for any $x, y \in F$ exactly one of $x < y, x = y, x > y$ is true.
\item if $x < y$ and $y < z$ then $x < z$.
\item if $x < y$ and $u < v$ then $x + u < y + v$.
\item if $0 < x < u$ and $0 < y < v$ then $xy < uv$.
\end{itemize}
Let, if possible, $\soc$ be ordered. We know that $i \ne 0$. If $i > 0$ then
$i^2 = -1 > 0$ is a contradiction. If $i < 0$ then $-i > 0$ and $(-i)^2 = -1 
> 0$ another contradiction.

Since a complex number can be represented as a point on the plane, we can
as well represent it in the polar form $(r\cos\theta, r\sin\theta)$ or
$r\cos\theta + ir\sin\theta = r(\cos\theta + i\sin\theta)$. The number $r$
is easily seen to be the positive square root $\sqrt{x^2 + y^2}$ and it is
unique for a given complex number. It is called the \emph{modulus} of the
complex number. The angle $\theta$ is called the \emph{argument} and it is
clearly multivalued. All angles $\theta + 2\pi n$, $n \in \soi$ correspond to
the same complex number. The multi-valuedness of the argument has many 
interesting consequences to the nature of functions of complex variables. The
modulus of a complex number $z = x + iy$ is denoted by $\abs{z}$ and the
argument by $\Arg{z}$. The value of $\theta$ in the range $[0, 2\pi)$ is called
the \emph{principle value} of the argument. (An equivalent convention is to let
it be in the range $(-\pi, \pi]$).

We know the Taylor series expansions of the real functions $\sin$ and $\cos$,
\begin{eqnarray*}
\sin\theta &=& \theta - \frac{\theta^3}{3!} + \frac{\theta^5}{5!} - \ldots \\
\cos\theta &=& 1 + \frac{\theta^2}{2!} - \frac{\theta^4}{4!} + \ldots.
\end{eqnarray*}
From these expressions we can easily infer that
\begin{equation}\label{c1s2e1}
\cos\theta + i\sin\theta = e^{i\theta}.
\end{equation}
It is called \emph{Euler's formula}. It allows us to represent a complex number
as $z = (x, y) = x + iy = r\exp(i\theta)$. The last representation simplifies
many calculations and has some immediate interesting consequences. First of all
$e^{i\pi} = -1$ so that $\log(-1) = i\pi$. Logarithms of negative numbers are
imaginary. They not only exist but they are multi-valued. $\log(-1) = i\pi + 
2\pi in$ for $n \in \soi$. Since $e^{2\pi i + 2\pi im} = 1$, it is easy to see 
that 
\[
1^{1/n} = \exp\left(i\frac{2\pi}{n} + i\frac{2\pi m}{n}\right).
\]
If we let $m$ range from $0$ to $n - 1$ we get $n$ different roots of unity.

The conjugate of a complex number $z = (x, y) = x + iy = re^{i\theta}$ is
defined as $\bar{z} = (x, -y) = x - iy = re^{-i\theta}$.

\section{A few inequalities}\label{c1s3}
Since $\abs{z} = r = \sqrt{x^2 + y^2}$, it follows that
\begin{eqnarray}
\abs{x} &\le& \abs{z} \label{c1s3e1} \\
\abs{y} &\le& \abs{z} \label{c1s3e2}
\end{eqnarray}
Now consider
\[
\abs{z_1 + z_2}^2 = (z_1 + z_2)\overline{(z_1 + z_2)} = \abs{z_1}^2 + \abs{z_2}^2
 + z_1\bar{z}_2 + \bar{z}_1z_2.
\]
We can combine the last two terms,
\[
\abs{z_1 + z_2}^2 = \abs{z_1}^2 + \abs{z_2} + 2\re(z_1z_2) \le 
\abs{z_1}^2 + \abs{z_2}^2 + 2\abs{z_1z_2},
\] 
where we used \eqref{c1s3e1} to get the last inequality. Since $\abs{z_1z_2}
= \abs{z_1}\abs{z_2}$, we get 
\begin{equation}\label{c1s3e3}
\abs{z_1 + z_2} \le \abs{z_1} + \abs{z_2}.
\end{equation}
We can use the same reasoning to conclude that
\begin{equation}\label{c1s3e4}
\abs{z_1 - z_2} \ge \abs{z_1} - \abs{z_2}.
\end{equation}
We next prove the Cauchy-Schwarz inequality,
\begin{equation}\label{c1s3e5}
\abs{z_1\bar{w}_1 + \cdots + z_n\bar{w}_n}^2 \le \sum_{i=1}^n\abs{z_i}^2
\sum_{i=1}^n\abs{w_i}^2.
\end{equation}
We can consider $z = (z_1, \ldots, z_n)$ and $w = (w_1, \ldots, w_n)$ to be 
elements of $\soc^n$. We can define the inner product in $\soc^n$ as
\begin{equation}\label{c1s3e6}
(z, w) = \sum_{i=1}^n z_i\bar{w}_i.
\end{equation}
We note the following properties of the inner product.
\begin{eqnarray}
(z, w) &=& \overline{(w, z)} \label{c1s3e7} \\
(\alpha z, w) &=& \alpha(z, w) \label{c1s3e8} \\
(z, \alpha w) &=& \bar{\alpha}(z, w), \label{c1s3e9}
\end{eqnarray} 
where $\alpha \in \soc$.

In terms of the inner product, the Cauchy-Schwarz inequality is
\begin{equation}\label{c1s3e10}
(z, w)^2 \le (z, z)(w, w).
\end{equation}
In order to prove this, consider the vector
\[
u = z - \frac{(z, w)}{(w, w)}w.
\]
Then
\[
(u, u) = (z, z) - \frac{\overline{(z, w)}}{(w, w)}(z, w) - 
\frac{(z, w)}{(w, w)}(w, z) + \frac{\abs{(z, w)}^2}{(w, w)^2}(w, w).
\]
Using properties \eqref{c1s3e7} to \eqref{c1s3e9} we get
\[
(u, u) = (z, z) - \frac{\abs{(z, w)}^2}{(w, w)}
\]
Since $(u, u) \ge 0$ is always true, equation \eqref{c1s3e10} follows. Since
the modulus of a complex number and its conjugate are the same, we can as well
write equation \eqref{c1s3e5} as
\begin{equation}\label{c1s3e11}
\abs{z_1w_1 + \cdots + z_nw_n}^2 \le \sum_{i=1}^n\abs{z_i}^2
\sum_{i=1}^n\abs{w_i}^2.
\end{equation}

{\color{red}{Add Lagrange's identity}}.

Before leaving this section we note that the inner product in $\soc$ is just
the product $z_1\bar{z}_2$. In the representation using ordered pair of reals,
\[
(z_1, z_2) = (x_1, y_1)(x_2, -y_2) = (x_1x_2 + y_1y_2, x_1y_2 - x_2y_1)
\]
If $(x_1, y_1)$ and $(x_2, y_2)$ were 2-dimensional vectors then the real part
of $(z_1, z_2)$ is their dot product and the imaginary part is the magnitude of
the cross product. This idea is exploited to develop the exterior algebra and
differential forms.
